### 2016-07-19 paper/ms.tex
adding skeleton paper

% Copyright 2015-2016 Dan Foreman-Mackey and the co-authors listed below.

\documentclass[manuscript, letterpaper]{aastex6}

\pdfoutput=1

\include{vc}
\usepackage{microtype}

\usepackage{url}
\usepackage{amssymb,amsmath}
\usepackage{natbib}
\usepackage{multirow}
\bibliographystyle{aasjournal}

% ----------------------------------- %
% start of AASTeX mods by DWH and DFM %
% ----------------------------------- %

\setlength{\voffset}{0in}
\setlength{\hoffset}{0in}
\setlength{\textwidth}{6in}
\setlength{\textheight}{9in}
\setlength{\headheight}{0ex}
\setlength{\headsep}{\baselinestretch\baselineskip} % this is 2 lines in ``manuscript''
\setlength{\footnotesep}{0in}
\setlength{\topmargin}{-\headsep}
\setlength{\oddsidemargin}{0.25in}
\setlength{\evensidemargin}{0.25in}

\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{colorlinks = false}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6\\
With modifications by David W. Hogg and Daniel Foreman-Mackey
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother

% Section spacing:
\makeatletter
\let\origsection\section
\renewcommand\section{\@ifstar{\starsection}{\nostarsection}}
\newcommand\nostarsection[1]{\sectionprelude\origsection{#1}}
\newcommand\starsection[1]{\sectionprelude\origsection*{#1}}
\newcommand\sectionprelude{\vspace{1em}}
\let\origsubsection\subsection
\renewcommand\subsection{\@ifstar{\starsubsection}{\nostarsubsection}}
\newcommand\nostarsubsection[1]{\subsectionprelude\origsubsection{#1}}
\newcommand\starsubsection[1]{\subsectionprelude\origsubsection*{#1}}
\newcommand\subsectionprelude{\vspace{1em}}
\makeatother

\widowpenalty=10000
\clubpenalty=10000

\sloppy\sloppypar

% ------------------ %
% end of AASTeX mods %
% ------------------ %

% Projects:
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\kepler}{\project{Kepler}}

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}

\newcommand{\figureref}[1]{\ref{fig:#1}}
\newcommand{\Figure}[1]{Figure~\figureref{#1}}
\newcommand{\figurelabel}[1]{\label{fig:#1}}

\newcommand{\Table}[1]{Table~\ref{tab:#1}}
\newcommand{\tablelabel}[1]{\label{tab:#1}}

\renewcommand{\eqref}[1]{\ref{eq:#1}}
\newcommand{\Eq}[1]{Equation~(\eqref{#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\eqref{#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}

\newcommand{\sectionname}{Section}
\newcommand{\sectref}[1]{\ref{sect:#1}}
\newcommand{\Sect}[1]{\sectionname~\sectref{#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\sectalt}[1]{\sectref{#1}}
\newcommand{\App}[1]{Appendix~\sectref{#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

\newcommand{\T}{\ensuremath{\mathrm{T}}}
\newcommand{\dd}{\ensuremath{\,\mathrm{d}}}
\newcommand{\unit}[1]{{\ensuremath{\,\mathrm{#1}}}}
\newcommand{\bvec}[1]{{\ensuremath{\boldsymbol{#1}}}}

% TO DOS
\newcommand{\todo}[3]{{\color{#2}\emph{#1}: #3}}
\newcommand{\dfmtodo}[1]{\todo{DFM}{red}{#1}}
\newcommand{\agoltodo}[1]{\todo{Agol}{blue}{#1}}

% \shorttitle{}
% \shortauthors{}
% \submitted{Submitted to \textit{The Astrophysical Journal}}

\begin{document}

\title{%
Fast and scalable Gaussian process modeling of stellar variability
\vspace{-3\baselineskip}  % OMG AASTEX6 IS SO BROKEN
}

\newcounter{affilcounter}
% \altaffiltext{1}{}

\setcounter{affilcounter}{1}
\edef \uw {\arabic{affilcounter}}\stepcounter{affilcounter}
\altaffiltext{\uw}       {Astronomy Department, University of Washington,
                          Seattle, WA, 98195, USA}

\edef \sagan {\arabic{affilcounter}}\stepcounter{affilcounter}
\altaffiltext{\sagan}{Sagan Fellow}

\author{%
    Eric~Agol\altaffilmark{\uw} and
    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan}
}

\begin{abstract}

Our Gaussian Processes are fast.

\end{abstract}

\keywords{%
% methods: data analysis
% ---
% methods: statistical
% ---
% catalogs
% ---
% planetary systems
% ---
% stars: statistics
}

\section{Introduction}

This is the literature \citep{Ambikasaran:2015}.

\section{Summary}

\vspace{1.5em}
All of the code used in this project is available from
\url{https://github.com/dfm/GenRP} under the MIT open-source software
license.
This code (plus some dependencies) can be run to re-generate all of the
figures and results in this paper; this version of the paper was generated
with git commit \texttt{\githash} (\gitdate).

\acknowledgments
It is a pleasure to thank
...
for helpful contributions to the ideas and code presented here.

This research made use of the NASA \project{Astrophysics Data System} and the
NASA Exoplanet Archive.
The Exoplanet Archive is operated by the California Institute of Technology,
under contract with NASA under the Exoplanet Exploration Program.

This paper includes data collected by the \kepler\ mission. Funding for the
\kepler\ mission is provided by the NASA Science Mission directorate.
We are grateful to the entire \kepler\ team, past and present.

These data were obtained from the Mikulski Archive for Space Telescopes
(MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST is provided by the NASA Office of Space Science via grant
NNX13AC07G and by other grants and contracts.

\facility{Kepler}
\software{%
    % \project{ceres} \citep{Agarwal:2016},
    % \project{corner.py} \citep{Foreman-Mackey:2016},
    % \project{emcee} \citep{Foreman-Mackey:2013},
    % \project{george} \citep{Ambikasaran:2016},
	% \project{matplotlib} \citep{Hunter:2007},
	% \project{numpy} \citep{Van-Der-Walt:2011},
	% \project{scipy} \citep{Jones:2001}.
}

\appendix

\bibliography{genrp}

\end{document}

### 2016-08-03 paper/ms.tex
moving paper over from sharelatex

@@
 {\noindent\footnotesize\textsc{\@journalinfo}}\par
-{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6\\
-With modifications by David W. Hogg and Daniel Foreman-Mackey
+{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6 with modifications
 }\par\vspace*{-\baselineskip}\vspace*{0.625in}
@@
 \author{%
-    Eric~Agol\altaffilmark{\uw} and
-    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan}
+    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan} and
+    Eric~Agol\altaffilmark{\uw}
 }
@@
 
-Our Gaussian Processes are fast.
+We derive a fast, $O(N)$, Gaussian Process approach that involves Lorentzian
+kernels.  This basis set can approximate commonly used kernels, and is a good
+description of stellar variability.
 
@@
 
-This is the literature \citep{Ambikasaran:2015}.
+\section{Outline}
+
+\begin{enumerate}
+\item Non-parametric modeling of time series with correlated noise using Gaussian processes
+\item Assumptions:\\
+  a). Stationary noise (although this can be broken).\\
+  b). Gaussian
+\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
+   curves; Solar curves; short-cadence.
+\item Solution(s):\\
+  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
+  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
+  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
+\item Problem: HODLR is tricky to use; CARMA requires certain relations between
+  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
+\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
+   $O(N*p^2)$, where p is number of Lorentzians.
+\item Generalized Ambikasarn for complex coefficients (Hermitian).
+\item Performance:\\
+  a). Show $O(N*p^2)$ scaling\\
+  b). Show comparison to HODLR.\\
+  c). Show comparison to wavelet (ala Carter \& Winn).\\
+  d). Compare with CARMA Kallman filter(?).
+\item Various methods/formulae:\\
+  a). Likelihood computation.\\
+  b). Generating GP with particular covariance [yet to solve]. [ ]\\
+  c). Inferring particular components.\\
+  d). Computing derivatives of GPs (as used in RV methods).\\
+  e). Forecasting/interpolating.\\
+  f). Inferring confidence intervals.\\
+  g). Derivative of likelihood function with respect to kernel parameters.\\
+  h). Non-stationary GP (can the coefficients vary with time?).\\
+  i). Multi-band (each element of extended matrix becomes a matrix).
+
+\item Generalization to non-stationary noise.
+\item Example application: TYC 3559\\
+  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
+  b). Analysis of solar data.
+
+\item Possible future applications:\\
+  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
+      with Lorentzians, not Gaussians.\\
+  b). Multi-waveband GPs.\\
+  c). RV interpolation.\\
+  d). Non-parametric phase functions.
+
+\item Appendix:
+  a). Description/summary of mathematics.\\
+  b). Description of code.
+\end{enumerate}
+
+\subsection{Derivative of likelihood with respect to Kernel parameters}
+
+The derivative of the likelihood can be computed in $O(N)$ operations as follows.
+
+The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
+\begin{equation}
+    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
+    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
+\end{equation}
+where $^\prime$ means the derivative wrt $x_i$.
+The kernel may be written as:
+\begin{equation}
+    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
+    \begin{array}{cc}
+       w_i + \sum_{l=1}^p \alpha_l & j = i \\
+       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
+    \end{array}
+    \right.
+\end{equation}
+Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
+
+Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
+\begin{eqnarray}
+    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
+    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
+     \left\{
+    \begin{array}{cc}
+       1 & j = i \\
+        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
+    \end{array}
+    \right.\\
+    \frac{\partial K_{ij}}{\partial \beta_k} &=&
+     \left\{
+    \begin{array}{cc}
+       0  & j = i \\
+        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
+    \end{array}
+    \right.
+\end{eqnarray}
+The first term in the derivative of the likelihood can be computed in $O(N)$
+evaluations, while the second term can be computed in three steps:
+\begin{itemize}
+\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
+\item Next, create an e
+\end{itemize}
+Writing this term out:
 
@@
 for helpful contributions to the ideas and code presented here.
+
+EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
+NASA Astrobiology Institute’s Virtual Planetary Laboratory, supported
+by NASA under cooperative agreement NNH05ZDA001C.
 

### 2016-09-02 paper/ms.tex
Future apps added

@@
   c). RV interpolation.\\
-  d). Non-parametric phase functions.
+  d). Non-parametric phase functions.\\
+  e). Better calibration of the flicker log(g) relation.\\
+  f). Measurement of rotational periods - gyrochronology.\\
+  g). Doppler shifts in oscillating EBs.
 
@@
 
+Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
+follow closely the notation and equations given in \citet{Ambikasaran2015}.  We start
+with the observation that a Lorentzian correlation function may be decomposed into the sum of two
+exponential functions:
+\begin{eqnarray}
+K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
+ &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
+\end{eqnarray}
+where $\beta = \beta_R + i\beta_I$ is the 
+complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
+is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.
+
+Now, the Lorentzian kernal may be included in the GRP algorithm 
+using the two exponential kernels with $\beta_1 = \beta$ and $\beta_2 = \beta^*$.  This 
+requires the algorithm to use complex arithmetic, which has several disadvantages:
+1).\ complex arithmetic is more expensive by a factor of 2-3; 2).\ the complex conjugate equation
+is redundant as it turns out that $l_2 = l_1^*$, $r_2 = r_1^*$, and $x_i = x_i^*$ (i.e.\ $x_i$ is
+real), meaning that the same equations are being solved twice for the real and complex components; 
+and 3).\ if the derivative of the likelihood is computed with automatic
+differentiation (AD), most AD algorithms require real variables.
+
+Given these drawbacks, we instead modify the GRP algorithm to use a combination of two real
+components rather than two complex components;  this addresses all three of these drawbacks
+simultaneously.  The one cost is that the banded extended matrix has a bandwidth which is
+larger by one, both above and below the diagonal.
+
+Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
+find the real and imaginary components of the complex equationi for $\beta$, and 
+discard the $\beta^*$ equation, which is redundant.  This gives four additional equations for
+each $\beta_j$ with non-zero imaginary component, while the kernel components with real
+$\beta_j$ still only have two additional components in the extended matrix.  We will enumerate
+the number of real $\beta$'s as $p_0$, while the total number of $\beta$'s is still $p$.
+This gives a total number of equations to be solved with the extended matrix of $N_{ex}= 
+(4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data points.
+
+For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
+auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
+r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
+imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
+matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
+so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric 
+set of equations, and since the values of these variables are discarded in the end, their sign
+is irrelevant).  The length of the vectors $l_k$ and $r_k$
+is the number of kernel components, $p$;  we denote the individual elements of these vectors
+as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
+to be the real values of $\beta$, and for these we do not include the imaginary component
+equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
+components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
+ = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
+\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
+Equation (61) becomes:
+\begin{equation}
+% See 9/2/2016 notes
+\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
+\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
+\end{equation}
+for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61).
+
+Equation (60) has real and imaginary components:
+\begin{eqnarray}
+\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
+\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
+\end{eqnarray}
+with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
+\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.
+
+Finally, equation (59) has real and imaginary components:
+\begin{align}
+-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
+ &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
+\end{align}
+with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
+we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
+imaginary components of equations (59) and (60) as well for these, which reduces the total number
+of equations to solve.
+
+With these additional equations, we can solve for the likelihood of sums of full Lorentzian kernels 
+in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables 
+for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the 
+cross terms between the real and imaginary variables.  The off-diagonal components are $2(p+1)$ above 
+and below the diagonal, for a total bandwidth of $4p+5$; in the case in which $p_0=p$ so that the
+imaginary components of $\beta$ are zero, the off-diagonal components revert to $p+1$, giving
+a bandwidth of $2p+3$.
+
+As with the GRP method, we can buid an extended matrix which is real and symmetric, and which may be 
+solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix 
+may be computed from the LU decomposition, and it is related to the determinant of the kernel by a 
+factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
+We arrange the auxiliary variables with real values of 
+$\beta$ first, followed by the complex variables, yielding the vector
+$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
+x_2, ..., x_N]$.
+
+{\bf Should I give some examples of this?  Some pseudocode? Some colored matrices like Figures 1-2?}
+
 \bibliography{genrp}

### 2016-09-03 paper/ms.tex
Figure added to Appendix ala Ambikasaran (2015).

@@
     % \project{emcee} \citep{Foreman-Mackey:2013},
-    % \project{george} \citep{Ambikasaran:2016},
+    % \project{george} \citep{Ambikasaran2016},
 	% \project{matplotlib} \citep{Hunter:2007},
@@
 
-{\bf Should I give some examples of this?  Some pseudocode? Some colored matrices like Figures 1-2?}
+The matrix has a similar structure as the real case constructed by \citep{Ambikasaran2015}, but the
+imaginary and real components mix (this is due to the representation of imaginary numbers as
+$2\times 2$ matrices).  Figure \ref{matrix_structure} shows an example of the structure of the extended
+matrix for two Lorentzians with non-zero $\omega$ values.  This is analagous to the real case with
+$p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
+that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran2015}.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[scale=0.175]{./twoterm.eps}
+\hspace{0.2in}
+\includegraphics[scale=1.0]{./images/colorcode.eps}
+\end{center}
+\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
+the example shown in \citet{Ambikasaran2015}.}
+\label{matrix_structure}
+\end{figure}
+
+{\bf Should I give some examples of this?  Some pseudocode?}
 

### 2016-09-03 julia/single_lorentzian/test_lorentzian.jl

@@ -70,8 +70,8 @@ beta_imag = [omega1,omega2] #,omega3,omega4]
 #beta = [complex(0.4,omega1)]
 #aex_full,bex_full,equations_full,variables_full = compile_matrix_full(alpha,beta_real,beta_imag,w0,t,y)
 #aex_symm01,bex_symm01,equations_symm01,variables_symm01 = compile_matrix_symm01(alpha,beta_real,beta_imag,w0,t,y)
-aex_symm_full,bex_symm_full,equations_symm_full,variables_symm_full = compile_matrix_symm_full(alpha,beta_real,beta_imag,w0,t,y)
-aex_symm,bex_symm,equations_symm,variables_symm = compile_matrix_symm(alpha,beta_real,beta_imag,w0,t,y)
+aex_symm_full,bex_symm_full,equations_symm_full,variables_symm_full,aex_factor_symm_full,aex_color_symm_full = compile_matrix_symm_full(alpha,beta_real,beta_imag,w0,t,y)
+aex_symm,bex_symm,equations_symm,variables_symm,aex_factor_symm = compile_matrix_symm(alpha,beta_real,beta_imag,w0,t,y)
 #aex_real,bex_real,al_small_real,indx_real,log_like,logdetofa = compile_matrix(alpha,beta_real,beta_imag,w0,t,y)
 
 p = length(alpha)
@@ -159,3 +159,23 @@ println("Expected : ",p0+2(p-p0+1))
 #println("Agreement: ",agreement)
 #return
 #end
+
+# Print out the color matrix tex code:
+
+f = open("twoterm.tex", "w")
+println(f,"\\documentclass{standalone}")
+println(f,"\\usepackage{tikz}")
+println(f,"\\begin{document}")
+println(f,"\\begin{tikzpicture}")
+
+for i=1:nex0
+  for j=1:nex0
+    if aex_color_symm_full[i,j] != ""
+       println(f,"\\draw [fill = ",aex_color_symm_full[i,j],"](",i,",",nex0+1-j,") rectangle (",i+1,",",nex0+1-(j+1),");")
+    end
+  end
+end
+println(f,"\\end{tikzpicture}")
+println(f,"\\end{document}")
+close(f)
+

### 2016-09-04 paper/ms.tex
Moved figure; updated ms.tex

@@
 Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
-find the real and imaginary components of the complex equationi for $\beta$, and 
-discard the $\beta^*$ equation, which is redundant.  This gives four additional equations for
-each $\beta_j$ with non-zero imaginary component, while the kernel components with real
-$\beta_j$ still only have two additional components in the extended matrix.  We will enumerate
-the number of real $\beta$'s as $p_0$, while the total number of $\beta$'s is still $p$.
-This gives a total number of equations to be solved with the extended matrix of $N_{ex}= 
-(4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data points.
+find the real and imaginary components of the complex equation for $\beta$ by taking the
+sum and difference of the $\beta$ and $\beta^*$ equations (divided by 2 and $2i$, respectively), 
+and then discard the $\beta^*$ equation, which is redundant.  This keeps the same number of
+additional equations for each $\beta_j$ with non-zero imaginary component (four additional
+equations in the extended matrix), but converts the complex equations to real, thus avoiding
+complex arithmetic, while the kernel components with real $\beta_j$ still only have two 
+additional components in the extended matrix.  The extended matrix adds an extra band above
+and below the diagonal when the $\beta$ and $\beta^*$ equations are combined since this
+mixes the components of the original equations which are shifted by one column from one
+another.  We will enumerate the number of real $\beta$'s as $p_0$, while the total number 
+of $\beta$'s is still $p$.  This gives a total number of equations to be solved with the 
+extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data 
+points.
 
@@
 \hspace{0.2in}
-\includegraphics[scale=1.0]{./images/colorcode.eps}
+\includegraphics[scale=1.0]{./colorcode.eps}
 \end{center}

### 2016-09-06 paper/ms.tex
Minor edits of appendix

@@
 of $\beta$'s is still $p$.  This gives a total number of equations to be solved with the 
-extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data 
+extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $N$ is the number of data 
 points.
@@
 Equation (61) becomes:
-\begin{equation}
+\begin{equation} \label{Amb61}
 % See 9/2/2016 notes
@@
 \end{equation}
-for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61).
+for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
+in \citet{Ambikasaran2015}.
 
-Equation (60) has real and imaginary components:
-\begin{eqnarray}
+Equation (60) in \citet{Ambikasaran2015} has real and imaginary components:
+\begin{eqnarray} \label{Amb60}
 \gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
@@
 
-Finally, equation (59) has real and imaginary components:
-\begin{align}
+Finally, equation (59) in \citet{Ambikasaran2015} has real and imaginary components:
+\begin{align} \label{Amb59}
 -&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
@@
 we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
-imaginary components of equations (59) and (60) as well for these, which reduces the total number
+imaginary components of equations (59) and (60) in \citet{Ambikasaran2015} as well for these, which reduces the total number
 of equations to solve.
@@
 for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the 
-cross terms between the real and imaginary variables.  The off-diagonal components are $2(p+1)$ above 
-and below the diagonal, for a total bandwidth of $4p+5$; in the case in which $p_0=p$ so that the
-imaginary components of $\beta$ are zero, the off-diagonal components revert to $p+1$, giving
-a bandwidth of $2p+3$.
+cross terms between the real and imaginary variables.  The off-diagonal components are $p_0+2(p-p_0)+2$ above 
+and below the diagonal, for a total bandwidth of $2p_0+4(p-p_0)+5$. In the case in which $p_0=p$, so 
+that the imaginary components of each $\beta$ are zero, the off-diagonal components revert to $p+1$ 
+above and below the diagonal, giving a bandwidth of $2p+3$ (the same as the GRP algorithm).
 
-As with the GRP method, we can buid an extended matrix which is real and symmetric, and which may be 
+As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be 
 solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix 
@@
 $x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
-x_2, ..., x_N]$.
+x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [b_1/2, \{0; j=1,...,p_0+2(p-p_0)\}, b_2/2, ..., b_N/2]$.
+The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
+\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2p_0+4(p-p_0)+5) \times N_{ex}$.
+The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
+with a banded algorithm.
 

### 2016-09-06 julia/single_lorentzian/compile_matrix_full.jl

@@ -35,6 +35,7 @@ for k=2:n
   end
   for j=1:(p-p0)
     variables[(k-2)*(4(p-p0)+2p+1)+1+p0+(j-1)*2+1] = string(" r^R_{",k,",",j,"}")
+    println((k-2)*(4(p-p0)+2p+1)+1+p0+(j-1)*2+2)
     variables[(k-2)*(4(p-p0)+2p+1)+1+p0+(j-1)*2+2] = string(" r^I_{",k,",",j,"}")
   end
   for j=1:p0

### 2016-09-06 julia/transpose/time_julia_final.jl

@@ -60,8 +60,9 @@ for it=1:nnt
   logdeta_final= compile_matrix_symm(alpha_final,beta_real_final,beta_imag_final,w0,t,nex_final,aex_final,al_small_final,indx_final)
   time_compute_final[it] = toq();
   tic()
-#  @code_warntype   compute_likelihood(p_final,y,aex_final,al_small_final,indx_final,logdeta_final)
-  log_like_final= compute_likelihood(p_final,p0_final,y,aex_final,al_small_final,indx_final,logdeta_final)
+  bex = zeros(Float64,nex_final)
+#  @code_warntype compute_likelihood(p_final,y,aex_final,al_small_final,indx_final,logdeta_final,bex)
+  log_like_final= compute_likelihood(p_final,p0_final,y,aex_final,al_small_final,indx_final,logdeta_final,bex)
   time_likelihood_final[it] = toq();
   println(n," final:   ",time_compute_final[it]," ",time_likelihood_final[it])
 

### 2016-09-09 paper/ms.tex

@@
 
+\section{Solving for log likelihood with extended matrix}
+
+Here we summarize \citet{Ambikasaran2015}, and then point to appendix
+for how to do this with oscillating kernels.
+
+\section{Training}
+
+May use derivative of likelihood + BFGS-B to optimize kernel parameters.
+
+Q: How do we determine how many kernels to use?
+
+\section{Prediction/interpolation}
+
+Precition/interpolation involves constructing a matrix of covariances
+between the (noisy) training set and the times at which we wish to
+predict or interpolate data points.   If our training times are given by
+${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
+mean values at the predicted times are given by (RW 2.23):
+\begin{eqnarray}
+\bar f_{*,k} &=& \sum_j K(t_{*,k},t_j) b_j,\\
+&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
+&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
+y_k &=& \sum_j K(t_j,t_k) b_j,
+\end{eqnarray}
+where the latter equation may be solved for ${\bf b}$ using the extended
+matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
+matrix $K({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
+of $MN$ multiplications is required to obtain the predicted values.
+If $M$ is large, this can become quickly prohibitive.  It turns out
+that the structure of the sum of exponential kernels may be exploited
+to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
+i.e.\ a linear scaling with the total number of data points, which can 
+be much fewer if $M$ is large.
+
+The procedure works as follows:
+\begin{itemize}
+\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
+\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
+compute:
+\begin{eqnarray} \label{prediction}
+\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
+\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} b_j\\
+\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} b_j,
+\end{eqnarray}
+where we have separated out the training times before and after the predicted time $t_{*,k}$.
+
+\item Next, we update the coefficients recursively:
+\begin{eqnarray}
+\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} b_j\right]\\
+\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} b_j\right].
+\end{eqnarray}
+With these updates, we can then use equation \ref{prediction} to compute the expected
+mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
+$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
+so this procedure avoids a heavy computational burden.
+
+The variance may be handled in a similar manner, but with $p^2$ coefficients
+to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
+\begin{eqnarray}
+cov(f_{*,k}) &=&  K(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+&=& K(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
+\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
+\end{eqnarray}
+The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
+\end{itemize}
+
+Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
+then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
+just solve for the difference in fewer operations? {\bf TBD}
+
 \acknowledgments
@@
 EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
-NASA Astrobiology Institute’s Virtual Planetary Laboratory, supported
+NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
 by NASA under cooperative agreement NNH05ZDA001C.

### 2016-10-04 paper/ms.tex
minor notes in outline

@@
   c). Inferring particular components.\\
-  d). Computing derivatives of GPs (as used in RV methods).\\
-  e). Forecasting/interpolating.\\
-  f). Inferring confidence intervals.\\
-  g). Derivative of likelihood function with respect to kernel parameters.\\
-  h). Non-stationary GP (can the coefficients vary with time?).\\
-  i). Multi-band (each element of extended matrix becomes a matrix).
+  d). TBD Computing derivatives of GPs (as used in RV methods).\\
+  e). TBD Forecasting/interpolating.\\
+  f). TBD Inferring confidence intervals.\\
+  g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
+  h). TBD Non-stationary GP (can the coefficients vary with time?).\\
+  i). TBD Multi-band (each element of extended matrix becomes a matrix).
 
-\item Generalization to non-stationary noise.
 \item Example application: TYC 3559\\
   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
-  b). Analysis of solar data.
+  b). Simulated datasets - hyperparameter/PSD inference.\\
+  c). Analysis of solar data?
 

### 2016-10-05 paper/ms.tex
Added some intro, refs

--- 2016-09-09 paper/ms.tex
@@
 
+Stars are variable, for better or worse:  their variability provides information about the properties
+of stars, but also results in noise that must be overcome in detecting and characterizing their
+planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
+goals:  1). to analyze the frequency spectrum of stellar variability to beter understand stellar
+properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
+flux variability and spectral variations.
+
+A promising technique for addressing both of these goals is the Gaussian Process formalism 
+\citep{Rasmussen, Gibson}.  The basis of this method is to assume that stellar variability
+is random, but that the random variability is correlated in time, and the nature of the
+correlations is completely specified by a correlation matrix.  In its simplest form, the 
+correlation matrix is modeled with a time-independent autocorrelation function (although 
+the time-independence can be relaxed), including a white-noise component that may vary
+with time, and the autocorrelation function is described by a functional form that may be 
+simply parameterized by a relatively small number of parameters, sometimes referred to as 
+`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
+describe the autocorrelation function, thus inferring properties of the stellar variabilty
+which may be used to characterize properties of the star, such as the rotation period,
+the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
+
 \section{Outline}
@@
   c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
+  d).
+KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp.
+1775-1784, 2015
 \item Problem: HODLR is tricky to use; CARMA requires certain relations between
@@
    $O(N*p^2)$, where p is number of Lorentzians.
+  - Sum of Lorentzians is a good description of stellar variability:
+     - granulation has expoential (or sums of exponential) noise
+     - asteroseismic obeys sums of Lorentzians
+     - periodic variability due to spots/plages/etc. can also be
+       approximated by Lorentzian(s)
 \item Generalized Ambikasarn for complex coefficients (Hermitian).
@@
   i). Multi-band (each element of extended matrix becomes a matrix).
-
+  j). Approximating various kernels: Matern, exp(sin^2), Gaussian, etc.
 \item Generalization to non-stationary noise.
@@
   b). Analysis of solar data.
-
+  c). CO2 Mauna Kea
 \item Possible future applications:\\

### 2016-10-05 julia/test_sp500.jl

@@ -1,0 +1,207 @@
+using Optim
+using PyPlot
+using StatsBase
+using ForwardDiff
+include("compile_matrix_symm.jl")
+include("compute_likelihood.jl")
+include("bandec_trans.jl")
+include("banbks_trans.jl")
+include("regress.jl")
+
+function test_sp500()
+
+function log_like_derivative_wrapper(p0,p,x,t,y)
+  n = length(t)
+  nex = (4(p-p0)+2p0+1)*(n-1)+1
+  m1 = 2(p-p0)+p0+2
+  if p0 == p
+    m1 = p0+ 1
+  end
+  width = 2m1+1
+  indx::Vector{Int64} = collect(1:nex)
+#  aex = zeros(Number,width,nex)
+#  al_small= zeros(Number,m1,nex)
+#  bex = zeros(Number,nex)
+
+  function log_like(x::Vector)
+    w0 = exp(x[1])
+    alpha = exp(x[2:p+1])
+    beta_real = exp(x[p+2:2p+1])
+    beta_imag = zeros(eltype(x),p)
+    if p > p0
+      beta_imag[p0+1:p] = exp(x[2p+2:3p-p0+1])
+    end
+    aex = zeros(eltype(x),width,nex)
+    al_small= zeros(eltype(x),m1,nex)
+    bex = zeros(eltype(x),nex)
+    logdeta = compile_matrix_symm(alpha,beta_real,beta_imag,w0,t,nex,aex,al_small,indx)
+    log_like = compute_likelihood(p,p0,y,aex,al_small,indx,logdeta,bex)
+#    println("alpha: ",alpha," beta_real: ",beta_real," white: ",w0," log_like: ",log_like)
+    return -log_like
+  end
+
+## Optimize the likelihood:
+  xdiff = copy(x)
+## Now perturb x:
+#  dlogdx = zeros(eltype(x),length(x))
+#  log_like0=log_like(x)
+#  println(log_like0,x)
+#  log_like0=log_like(x)
+#  x0 = copy(x)
+#  for i=1:length(x)
+#    x    = copy(x0)
+#    if (x[i] == 0.) then
+#      h = 1e-6
+#    else
+#      h = 1e-6*x[i]
+#    end
+#    x[i] = x[i]+h
+#    println(i," ",x,size(x))
+#    log_like_h = log_like(x)
+#    dlogdx[i]=(log_like_h-log_like0)/h
+#  end
[…]

### 2016-10-18 paper/ms.tex
More work on the introduction.

@@
 planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
-goals:  1). to analyze the frequency spectrum of stellar variability to beter understand stellar
+goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to beter understand stellar
 properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
@@
 A promising technique for addressing both of these goals is the Gaussian Process formalism 
-\citep{Rasmussen, Gibson}.  The basis of this method is to assume that stellar variability
-is random, but that the random variability is correlated in time, and the nature of the
+\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
+is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
 correlations is completely specified by a correlation matrix.  In its simplest form, the 
@@
 the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
+The goal of planet detection includes measuring the Doppler shift of the star, or measuring
+the decrement of flux as a planet transits a star.  To properly account for the impact of 
+stellar variability on the planet parameters requires treating correlations in the data, and 
+to do so also requires evaluating the correlation matrix.
+
+The size of the correlation matrix, $N_{elements}$ scales as the square of the length of the time-series,
+$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become 
+prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
+its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
+= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
+using long time series datsets, such as the \texttt{Kepler} dataset $N_{time} \approx 10^{5-6}$,
+\texttt{Spitzer} time series with $N_{time} \approx 10^5$, and solar time series with
+$N_{time} \approx 10^7$.  So faster means of solving for the likelihood is required in these
+cases.
+
+One promising solution is to use approximate techiques based upon the fact that correlation
+matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
+algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the 
+solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
+solver is complex and can be challenging to use with general choices of kernel.  Another
+promising approximate technique is to use a uniform resampling of the dataset, and then
+rely on the structure of Toeplitz matrices, which require uniformly space time series,
+to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
+the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
+this approach has the drawback of being approximate.  A third approach is to use a kernel
+based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
+model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
+is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
+of the Lorentzians are required to have a specific relation.  In addition, these
+techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}
+
+The approach we explore in this paper is based upon a method developed by Press \&
+Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
+matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
+The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
+They observed how a combination of two exponential kernels would have a simliar scaling,
+but they did not demonstrate how to extend this to more than two kernel components.
+
+A significant advance was made by generalizing the Press-Rybicki method to an
+arbitrary number of exponential kernels by utilizing the fact that the elements of
+such matrices can be written in terms of the products of components of two vectors,
+which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
+Rybicki-Press' (\texttt{GenRP}) approach
+has many advantages:  the computation time and storage both scale as $O(N_{time})$,
+the coefficients of the kernels need not be stationary, the computation of the
+likelihood is exact, and the algorithm is simple to implement, and thus is robust.
+Unlike the \texttt{CARMA} model, there is no restriction on the relation between
+the coefficients (except that the kernel needs to be positive definite).  Unlike
+\texttt{HODLR} and \texttt{KISS-GP} the method is exact.  The main drawback of
+this approach is that the sum of exponential kernels (with real coefficients) can 
+only represent a simplified range power spectra of auto-correlated time series
+that can be expressed the sum of Lorentzians that peak at zero frequency.
+
+However, this drawback turns out to not be significant:  the exponential kernel
+can have a complex exponential coefficients (plus their complex-conjugates), and 
+thus approximate an arbitrary autocorrelation function with enough terms, including
+quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
+a {\it very} good approximation of stellar variability spectra:  the granulation
+noise due to stellar convection may be expressed in terms of zero-frequency
+Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
+Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
+approach to kernels of the form:
+\begin{equation}
+K(\tau) = a_0 \delta(\tau) + \sum_j a_{j=1}^{N_{kernel}} e^{-\beta_R \tau} \cos{(\beta_I \tau)},
+\end{equation}
+where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
+in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
+$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
+and zero otherwise.  We show that
+the extended matrix that embeds this multi-component kernel can be written
+in real notation, which speeds up the evaluation of the likelihood function,
+and allows optimization software to use automatic differential in computing
+derivatives of the likelihood function.
+
+The plan of the paper is...
 
@@
 \begin{enumerate}
-\item Non-parametric modeling of time series with correlated noise using Gaussian processes
-\item Assumptions:\\
-  a). Stationary noise (although this can be broken).\\
-  b). Gaussian
-\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
-   curves; Solar curves; short-cadence.
-\item Solution(s):\\
-  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
-  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
-  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
-  d).
+%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
+%\item Assumptions:\\
+%  a). Stationary noise (although this can be broken).\\
+%  b). Gaussian
+%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
+%   curves; Solar curves; short-cadence.
+%\item Solution(s):\\
+%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
+%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
+%  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
+%  d).
 KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp.
 1775-1784, 2015
-\item Problem: HODLR is tricky to use; CARMA requires certain relations between
-  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
-\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
-   $O(N*p^2)$, where p is number of Lorentzians.
-  - Sum of Lorentzians is a good description of stellar variability:
-     - granulation has expoential (or sums of exponential) noise
-     - asteroseismic obeys sums of Lorentzians
-     - periodic variability due to spots/plages/etc. can also be
-       approximated by Lorentzian(s)
-\item Generalized Ambikasarn for complex coefficients (Hermitian).
+%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
+%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
+%\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
+%   $O(N*p^2)$, where p is number of Lorentzians.
+%  - Sum of Lorentzians is a good description of stellar variability:
+%     - granulation has expoential (or sums of exponential) noise
+%     - asteroseismic obeys sums of Lorentzians
+%     - periodic variability due to spots/plages/etc. can also be
+%       approximated by Lorentzian(s)
+%\item Generalized Ambikasarn for complex coefficients (Hermitian).
 \item Performance:\\
@@
   i). Multi-band (each element of extended matrix becomes a matrix).
-  j). Approximating various kernels: Matern, exp(sin^2), Gaussian, etc.
+  j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
 \item Generalization to non-stationary noise.
@@
   c). CO2 Mauna Kea
+  d). TBD Computing derivatives of GPs (as used in RV methods).\\
+  e). TBD Forecasting/interpolating.\\
+  f). TBD Inferring confidence intervals.\\
+  g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
+  h). TBD Non-stationary GP (can the coefficients vary with time?).\\
+  i). TBD Multi-band (each element of extended matrix becomes a matrix).
+
+\item Example application: TYC 3559\\
+  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
+  b). Simulated datasets - hyperparameter/PSD inference.\\
+  c). Analysis of solar data?
+
 \item Possible future applications:\\

### 2016-10-18 paper/ms.tex

@@
 
-\subsection{Derivative of likelihood with respect to Kernel parameters}
+\section{Generalized Press-Rybicki}
 
-The derivative of the likelihood can be computed in $O(N)$ operations as follows.
+The general problem to be solved for Gaussian Process analysis of stellar time series is to evaluate
+the likelihood function,
+\begin{equation}
+{\cal{L}} = \ln{p(\mathbf{y}\vert {\mathbf{t},\bvec{\theta}},\bvec{\alpha})} = -\frac{1}{2}\ln{\vert \mathbf{K}(\bvec{\alpha}) \vert}
+- \frac{N_{time}}{2} \ln{(2\pi)}
+    - \frac{1}{2} {\mathbf r} (\bvec{\theta} ) ^T \mathbf{K}(\bvec{\alpha})^{-1} \mathbf{r}(\bvec{\theta}),
+\end{equation}
+where $\mathbf{y}$ are the measured time series (for example fluxes or radial velocities) at times
+$\mathbf{t}$, while the data are modeled with a function $m(\mathbf{t};\bvec{\theta})$ (for example,
+a transit model or Keplerian orbital model) which leaves residuals $\mathbf{r}(\bvec{\theta}) =
+\mathbf{y} - m(\mathbf{t};\bvec{\theta})$. % Note: I'm using the same notation as the peerless project. -EA
+The vector $\bvec{\alpha}$ specifies the parameters of the kernel, while the vector $\bvec{\theta}$
+specifies the parameters of the physical model.  In addition to this likelihood, a prior may be placed on 
+the values of these model parameters (either the physical or kernel); the log of this prior may be
+added to this log likelihood function.  Note that the kernel
+parameters may also have a physical meaning, but they describe a noisy process, and thus affect
+the correlated variability of stellar noise rather than a deterministic model as encapsulated in
+$m(\mathbf{t};\bvec{\theta})$.
 
-The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
-\begin{equation}
-    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
-    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
-\end{equation}
-where $^\prime$ means the derivative wrt $x_i$.
-The kernel may be written as:
-\begin{equation}
-    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
-    \begin{array}{cc}
-       w_i + \sum_{l=1}^p \alpha_l & j = i \\
-       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
-    \end{array}
-    \right.
-\end{equation}
-Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
+We express the kernel as the sum of exponential functions...
 
-Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
-\begin{eqnarray}
-    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
-    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
-     \left\{
-    \begin{array}{cc}
-       1 & j = i \\
-        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
-    \end{array}
-    \right.\\
-    \frac{\partial K_{ij}}{\partial \beta_k} &=&
-     \left\{
-    \begin{array}{cc}
-       0  & j = i \\
-        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
-    \end{array}
-    \right.
-\end{eqnarray}
-The first term in the derivative of the likelihood can be computed in $O(N)$
-evaluations, while the second term can be computed in three steps:
-\begin{itemize}
-\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
-\item Next, create an e
-\end{itemize}
-Writing this term out:
+%\subsection{Derivative of likelihood with respect to Kernel parameters}
+%
+%The derivative of the likelihood can be computed in $O(N)$ operations as follows.
+%
+%The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
+%\begin{equation}
+%    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
+%    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
+%\end{equation}
+%where $^\prime$ means the derivative wrt $x_i$.
+%The kernel may be written as:
+%\begin{equation}
+%    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
+%    \begin{array}{cc}
+%       w_i + \sum_{l=1}^p \alpha_l & j = i \\
+%       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
+%    \end{array}
+%    \right.
+%\end{equation}
+%Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
+%
+%Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
+%\begin{eqnarray}
+%    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
+%    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
+%     \left\{
+%    \begin{array}{cc}
+%       1 & j = i \\
+%        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
+%    \end{array}
+%    \right.\\
+%    \frac{\partial K_{ij}}{\partial \beta_k} &=&
+%     \left\{
+%    \begin{array}{cc}
+%       0  & j = i \\
+%        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
+%    \end{array}
+%    \right.
+%\end{eqnarray}
+%The first term in the derivative of the likelihood can be computed in $O(N)$
+%evaluations, while the second term can be computed in three steps:
+%\begin{itemize}
+%\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
+%\item Next, create an e
+%\end{itemize}
+%Writing this term out:
 

### 2016-10-20 paper/ms.tex
Editing section describing our version of GenRP

@@
 \usepackage{microtype}
+\usepackage{amsmath}
 
@@
 Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
-approach to kernels of the form:
-\begin{equation}
-K(\tau) = a_0 \delta(\tau) + \sum_j a_{j=1}^{N_{kernel}} e^{-\beta_R \tau} \cos{(\beta_I \tau)},
-\end{equation}
-where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
-in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
-$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
-and zero otherwise.  We show that
+approach to kernels with complex exponentials, which when combined with the
+complex conjugate leads to damped harmonic auto-correlation functions which
+are flexible enough to describe a wide range of variability, and may still
+be solved in $O(N_{time}$ operations.
+
+We show that
 the extended matrix that embeds this multi-component kernel can be written
@@
 
-We express the kernel as the sum of exponential functions...
+The kernel is modeled as a sum of damped harmonics,
+\begin{equation}
+K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
+\end{equation}
+where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay 
+in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
+$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
+and zero otherwise.
+
+Now, the damped harmonic kernel components may be rewritten as:
+\begin{equation}
+ a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
+\end{equation}
+where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
+matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran2015}.
+The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
+but otherwise the mathematical form of the extended matrix is identical, but with twice as many
+exponential kernel components as the real case.
+
+Although this is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
+in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
+arithmetic.  In addition, it turns out that the complex conjugate component, $\beta^* = \beta_R - i\beta_I$, in 
+the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
+component $\beta = \beta_R + i \beta_I$;  thus the computation is redundant.  Finally, complex
+arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
+which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
+to compute this explicitly!).
+
+Consequently, we use the fact that $\exp{i\beta_I \tau} =  \cos{\beta_I \tau} + i \sin{\beta_I \tau}$ to
+rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
+may be replaced by a $2\times 2$ matrix, $ \begin{pmatrix} 0 & 1 \\ -1 & 0\end{pmatrix}$, so that
+$\cos{\beta_I \tau} + i \sin{\beta_I \tau}$ may be rewritten as the rotation matrix
+$\begin{pmatrix} \cos{\beta_I \tau} & \sin{\beta_I \tau} \\ -\sin{\beta_I \tau} & \cos{\beta_I \tau}\end{pmatrix}$.
+We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex 
+equations, resulting in a real extended matrix that is the same size as the complex matrix.
+This transformation leads to about a factor of two in computational savings, but has the
+advantage of allowing the automatic differentiation due to the use of real arithmetic.
+
+\subsection{Kernel components}
 
@@
 
-\appendix
+\appendix \label{appendixa}
 

### 2016-10-20 paper/ms.tex
Added discussion of kernel approximation.

@@
 \usepackage{microtype}
-\usepackage{amsmath}
 
@@
 in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
-$a_0$ is a white-noise component with $\delta(\tau) = 1$ for $\tau=0$,
-and zero otherwise.
+$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
+the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
+time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
+of the measured times if the correlated components are zero.
 
@@
 
-Consequently, we use the fact that $\exp{i\beta_I \tau} =  \cos{\beta_I \tau} + i \sin{\beta_I \tau}$ to
+Consequently, we use the fact that $\exp{(i\beta_I \tau)} =  \cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ to
 rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
-may be replaced by a $2\times 2$ matrix, $ \begin{pmatrix} 0 & 1 \\ -1 & 0\end{pmatrix}$, so that
-$\cos{\beta_I \tau} + i \sin{\beta_I \tau}$ may be rewritten as the rotation matrix
-$\begin{pmatrix} \cos{\beta_I \tau} & \sin{\beta_I \tau} \\ -\sin{\beta_I \tau} & \cos{\beta_I \tau}\end{pmatrix}$.
+may be replaced by a $2\times 2$ anti-symmetric matrix,
+\begin{equation}
+\left(\begin{tabular}{ll}
+0 & 1 \\
+-1 & 0
+\end{tabular}\right),
+\end{equation}
+while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$, 
+so that $\cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ may be rewritten as the rotation matrix
+\begin{equation}
+\left(\begin{tabular}{ll} 
+$\cos{(\beta_I \tau)}$ & $\sin{(\beta_I \tau)}$ \\ 
+$-\sin{(\beta_I \tau)}$ & $\cos{(\beta_I \tau)}$ 
+\end{tabular}\right).
+\end{equation}
+The algebraic properties of these matrices are identical to the complex numbers, but only involve
+real components.
 We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex 
@@
 This transformation leads to about a factor of two in computational savings, but has the
-advantage of allowing the automatic differentiation due to the use of real arithmetic.
+additional advantage of allowing the automatic differentiation due to the use of real arithmetic.
 
-\subsection{Kernel components}
+\section{Kernel components}
+
+A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
+plus the white noise component:
+\begin{equation}
+K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
+\end{equation}
+Note that this encompasses all smooth, time-staionary kernels, and generally $c_1 \le 0$ as
+correlated noise initially decreases with separation in time.  We wish to approximate this
+as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
+can carry out the approximation in terms of sums of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)}$,
+which has a Taylor series expansion:
+\begin{equation}
+\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} = 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 -
+\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
+\end{equation}
+which may be continued to any order in $\tau$ as necessary.
+If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
+requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
+chosen to match the coefficients $c_j$.  Not that for $N=1$ we only require a damped exponential
+kernel with $\beta_I = 0$.
+
+Another way to view this is in the power spectrum of the Gaussian Process, which is the
+Fourier Transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
+is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
+width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
+power spectrum is: {\color{red} Check that this is true.}
+\begin{equation}
+P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
+\end{equation}
+where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
+that a smooth distribution may be approximated arbitrarily precisely by summing together
+Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
+a peak in the power spectrum is narrow at a specific frequency, then a small value of 
+$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
+of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
+can be arranged to cancel in the wings, giving features in the power spectrum that are
+narrower in frequency.
+
+The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
+kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
+$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
+used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
+commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
+used to approximate each of these kernels, with various degrees of success:
+\begin{itemize}
+\item Constant kernel
+\item The squared exponential kernel has the behavior of declining exponentially at large
+separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
+approximate a Gaussian kernel.  For example, to 
+\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
+\begin{equation}
+K_{3/2}(\tau,\tau_0)
+= (1+x)e^{-x},
+\end{equation} 
+where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This may be
+well approximated by the sum of two exponential kernels: 
+\begin{equation}
+K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
+\end{equation} 
+where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
+large values of $b$, the error on this scales as $2/(e^2b) \approx 0.27/b$; for
+$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
+have opposite signs, which may lead to numerical instability for values of $b$ that
+are too large.  One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
+$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
+have this property, which leads to qualitatively different noise properties on small
+timescales due to the sharpness of the kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
+turns out to be an excellent description of stellar granulation, and thus the approximate
+formula is simultaneously physically relevant and computationally convenient.
+\item Matern 5/2 kernel.
+\item Cosine kernel
+\item exp-sin-sq kernel
+\end{itemize}
 

### 2016-10-21 paper/ms.tex
Added discussion of Matern 3/2 kernel

@@
 Stars are variable, for better or worse:  their variability provides information about the properties
-of stars, but also results in noise that must be overcome in detecting and characterizing their
+of stars, but also results in noise that must be accounted for in detecting and characterizing their
 planetary systems.  When analyzing astronomical times series of stars, then, there are generally two 
-goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to beter understand stellar
+goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
 properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
@@
 the time-independence can be relaxed), including a white-noise component that may vary
-with time, and the autocorrelation function is described by a functional form that may be 
+with each exposure, and the autocorrelation function is described by a functional form that may be 
 simply parameterized by a relatively small number of parameters, sometimes referred to as 
@@
 which may be used to characterize properties of the star, such as the rotation period,
-the asteroseismic frequencies, or the `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
-The goal of planet detection includes measuring the Doppler shift of the star, or measuring
-the decrement of flux as a planet transits a star.  To properly account for the impact of 
+the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
+The goals of planet detection include measuring the Doppler shift of the star, measuring
+the decrement of flux as a planet transits a star, and/or measuring the times of transit to
+look for dynamical interactions between planets.  To properly account for the impact of 
 stellar variability on the planet parameters requires treating correlations in the data, and 
@@
 
-The size of the correlation matrix, $N_{elements}$ scales as the square of the length of the time-series,
+The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
 $N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become 
@@
 = N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
-using long time series datsets, such as the \texttt{Kepler} dataset $N_{time} \approx 10^{5-6}$,
-\texttt{Spitzer} time series with $N_{time} \approx 10^5$, and solar time series with
-$N_{time} \approx 10^7$.  So faster means of solving for the likelihood is required in these
+using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
+{\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
+$N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
 cases.
@@
 solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
-solver is complex and can be challenging to use with general choices of kernel.  Another
+solver is complicated and can be challenging to use with general choices of kernel.  Another
 promising approximate technique is to use a uniform resampling of the dataset, and then
@@
 The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
-They observed how a combination of two exponential kernels would have a simliar scaling,
+They also indicated that a combination of two exponential kernels would have a simliar scaling,
 but they did not demonstrate how to extend this to more than two kernel components.
 
-A significant advance was made by generalizing the Press-Rybicki method to an
+A significant advance was made in generalizing the Press-Rybicki method to an
 arbitrary number of exponential kernels by utilizing the fact that the elements of
@@
 the coefficients (except that the kernel needs to be positive definite).  Unlike
-\texttt{HODLR} and \texttt{KISS-GP} the method is exact.  The main drawback of
-this approach is that the sum of exponential kernels (with real coefficients) can 
-only represent a simplified range power spectra of auto-correlated time series
-that can be expressed the sum of Lorentzians that peak at zero frequency.
+\texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback 
+of this approach is that the sum of exponential kernels with real coefficients can 
+only represent a limited range of power spectra of auto-correlated time series:
+those which can be expressed the sum of Lorentzians that peak at zero frequency.
 
 However, this drawback turns out to not be significant:  the exponential kernel
-can have a complex exponential coefficients (plus their complex-conjugates), and 
-thus approximate an arbitrary autocorrelation function with enough terms, including
-quasi-periodic kernels.  It turns out that the sum of Lorentzian components is
-a {\it very} good approximation of stellar variability spectra:  the granulation
-noise due to stellar convection may be expressed in terms of zero-frequency
-Lorentzians, the asteroseismic oscillations may be expressed as non-zero frequency
-Lorentizians, as can stellar rotation.  In this paper we extend the \texttt{GenRP} 
-approach to kernels with complex exponentials, which when combined with the
-complex conjugate leads to damped harmonic auto-correlation functions which
-are flexible enough to describe a wide range of variability, and may still
+can also have a complex exponential coefficients (plus their complex-conjugates), and 
+thus approximate an almost arbitrary autocorrelation function with enough terms, most
+importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian 
+components is a {\it very} good approximation of stellar variability spectra:  the 
+activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
+Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
+Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be 
+parameterized this way as well.  In this paper 
+we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when 
+combined with the complex conjugate leads to damped harmonic auto-correlation functions 
+which are flexible enough to describe a wide range of variability, and may still
 be solved in $O(N_{time}$ operations.
 
-We show that
-the extended matrix that embeds this multi-component kernel can be written
+We show that the extended matrix that embeds this multi-component kernel can be written
 in real notation, which speeds up the evaluation of the likelihood function,
@@
 approximate a Gaussian kernel.  For example, to
+
 \item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
 \begin{equation}
-K_{3/2}(\tau,\tau_0)
-= (1+x)e^{-x},
+K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
 \end{equation} 
-where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This may be
-well approximated by the sum of two exponential kernels: 
+where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
+$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
+\begin{equation}
+\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
+\end{equation}
+where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels: 
 \begin{equation}
@@
 where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
-large values of $b$, the error on this scales as $2/(e^2b) \approx 0.27/b$; for
+large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
 $b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
 have opposite signs, which may lead to numerical instability for values of $b$ that
-are too large.  One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
+are large.
+
+One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
 $\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
 have this property, which leads to qualitatively different noise properties on small
-timescales due to the sharpness of the kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
-turns out to be an excellent description of stellar granulation, and thus the approximate
+timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
+turns out to be an excellent description of stellar activity and granulation, and thus the approximate
 formula is simultaneously physically relevant and computationally convenient.  
-\item Matern 5/2 kernel.
+The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
+while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
+spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
+%\item Matern 5/2 kernel.
 \item Cosine kernel
-\item exp-sin-sq kernel
+\item exp-sin-square kernel
 \end{itemize}

### 2016-10-21 paper/ms.tex

@@
 \begin{itemize}
-\item Constant kernel
-\item The squared exponential kernel has the behavior of declining exponentially at large
-separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
+\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
+correlations in data, and can be represented with $\beta =  0$.
+\item {\bf Exponential kernel.} This is also known as a Mat\'ern 1/2 kernel, and is exactly
+represented with $\beta_I=0$.
+\item The squared exponential kernel has the behavior of declining steeply at large
+time separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
 approximate a Gaussian kernel.  For example, to
@@
 %\item Matern 5/2 kernel.
-\item Cosine kernel
+\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
 \item exp-sin-square kernel

### 2016-10-22 paper/ms.tex
Added discussion of Gaussian, Matern 5/2

@@
 represented with $\beta_I=0$.
-\item The squared exponential kernel has the behavior of declining steeply at large
-time separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
-approximate a Gaussian kernel.  For example, to
-
+\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
+\item exp-sin-square kernel
+\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $e^{-(\tau/\tau_0)^2/2}$, 
+has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.  
+An infinite number of cosine kernels may be used to approximate a Gaussian kernel, but any
+finite combination of cosine kernels leads to a negative amplitude at finite $\tau$.  Thus,
+in general it is difficult to model a squared exponential with sums of damped cosine terms due
+to the heavier tail.  However, the Mat\'ern kernel may be used to approximate it on short
+timescales.
+\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
+\begin{equation}
+C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
+\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} ({\sqrt{(8p+4)}x)^{p-i},
+\end{equation} 
+where $p$ is a positive integer.  In the limit $p\rightarrow \infty, this becomes the
+squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
+values of $p$, the Mat\'ern kernel may be approximated well by the sums of
+exponential kernels, described next.
 \item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
@@
 spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
-%\item Matern 5/2 kernel.
-\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
-\item exp-sin-square kernel
+\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given
+by
+\begin{equation}
+C_{5/2}(x) = \sigma^2 (1+y+y^2/3)e^{-y},
+\end{equation}
+where $y = \sqrt{5}x$.  The Fourier transform is:
+\begin{equation}
+\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
+\end{equation}
+This kernel may be approximated by the sum of three exponential
+kernels, and has a steeper dependence at large frequencies, $f^{-6}$.  We aren't aware of a stellar
+power spectrum that requires such a steep spectrum, and thus we forgo implementation of this kernel.
 \end{itemize}

### 2016-10-24 paper/ms.tex

--- 2016-10-21 paper/ms.tex
@@
 correlations in data, and can be represented with $\beta =  0$.
-\item {\bf Exponential kernel.} This is also known as a Mat\'ern 1/2 kernel, and is exactly
+\item {\bf Exponential kernel.} This is exactly
 represented with $\beta_I=0$.
-\item The squared exponential kernel has the behavior of declining steeply at large
-time separation, and has $c_1$ = 0.  An infinite number of exponential kernels may be used to
-approximate a Gaussian kernel.  For example, to
-
+\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
+\item exp-sin-square kernel
+\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
+has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
+Nevertheless, it may be well approximated by the sum of three exponential kernels:
+\begin{eqnarray}
+G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
+ &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
+ &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
+\end{eqnarray}
+which has an accuracy of better than $<4\times 10^{-4}$ for all z.
+%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
+\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
+\begin{equation}
+C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
+\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
+\end{equation} 
+where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
+squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
+values of $p$, the Mat\'ern kernel may be approximated well by the sum of
+$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
+$p=1$ is relevant to stellar variability, which we describe next.
 \item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
@@
 timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
-turns out to be an excellent description of stellar activity and granulation, and thus the approximate
+turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
 formula is simultaneously physically relevant and computationally convenient.  
@@
 spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
-%\item Matern 5/2 kernel.
-\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
-\item exp-sin-square kernel
+\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
+\begin{equation}
+C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
+\end{equation}
+where $y = \sqrt{5}x$.  The Fourier transform is:
+\begin{equation}
+\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
+\end{equation}
+This kernel may be approximated well by the sum of four exponential kernels
+\begin{eqnarray}
+K_{5/2}(y) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
+ &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
+\end{eqnarray}
+which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
+The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
+power spectrum that requires a steeper spectrum than this, we forgo approximating further Mat\'ern kernels.
 \end{itemize}
+
+A standard practice in constructing kernels for solving problems with Gaussian Process
+is to add or multiply the preceding standard kernel components.  In the case of damped cosine
+kernels, the product of two kernels is: 
+\begin{eqnarray}
+K_j(\tau)K_k(\tau) &=&
+\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
+&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k}) +K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
+\end{eqnarray}
+which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
+construct new kernels from damped cosine kernels.
 

### 2016-10-24 paper/ms.tex
Added discussion of exponential-sine-squared kernel

@@
 \item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
-\item exp-sin-square kernel
 \item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
@@
 power spectrum that requires a steeper spectrum than this, we forgo approximating further Mat\'ern kernels.
+\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
+which has been used to model quasi-periodic stellar variability due to star spots.   It is
+given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$, which may be expanded as a series in $\cos{(2kw)}$,
+with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:  
+\begin{equation}
+E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
+\end{equation}
+with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
+$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
+$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
+$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
+$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
+This approximation is accurate to better than 0.5\% for $\Gamma < 3$.
 \end{itemize}
+
+{\color{red} Should we include some plots of these kernels, with the approximations?}
 
@@
 construct new kernels from damped cosine kernels.
+
+The upshot is that the damped cosine is an expressive basis kernel that may be used
+to represent a wide range of autocorrelation functions relevant for stellar variability.
+We recommend using a series of sums of damped cosines (in which some parameters may be 
+fixed to specified values to represent particular kernels) to model different amplitudes
+and timescales of variability.  If an initial estimate of the power spectrum or
+autocorrelation function is available, then these may be fit with this series to
+initialize the coefficients of the damped cosine kernel components.
+Since a Gaussian Process
+must be positive definite, we recommend modeling the logarithm of $a_j$,
+$\beta_{R,j}$, and $\beta_{I,j}$ to enforce this to be the case.  One caveat is
+that to produce certain kernels, a negative value of $a_j$ is needed (such
+as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
+two coefficients may be sampled from a logarithm, which the individual
+values are allowed to be negative;  this will enforce the positive definite
+constraint, but may allow for other constraints, such as an autocorrelation
+function which has a derivative of zero at $\tau = 0$.
 

### 2016-10-24 paper/expsin2_approx.jl

@@ -1,0 +1,177 @@
+# Try to approximate exponential-sine-squared kernel:
+using PyPlot
+using Optim
+
+function expsin2_approx()
+nexp = 5
+ngamma = 100
+coeff_grid = zeros(ngamma,nexp)
+#gamma_grid = logspace(-2,0.5,ngamma)
+gamma_grid = linspace(0.01,3.0,ngamma)
+ntime = 200
+time = collect(linspace(0,2pi,ntime))
+
+gamma = 1.0
+gaussian=exp(-gamma.*sin(time).^2)
+
+function compute_chi(x)
+model = zeros(ntime)
+for k=1:nexp
+  model += x[k].*cos((k-1)*2.*time)
+end
+return sum((gaussian-model).^2)
+end  
+
+npar = nexp
+
+for i=1:ngamma
+gamma = gamma_grid[i]
+gaussian=exp(-gamma.*sin(time).^2)
+
+abest=zeros(npar)
+abest = [0.6,0.2,0.01,0.05,.025,.025]
+
+atrial = zeros(npar)
+for k=1:npar
+  atrial[k] = abest[k]
+end
+model = zeros(ntime)
+for k=1:nexp
+  model += atrial[k].*cos((k-1)*2.*time)
+end
+clf()
+plot(time,gaussian)
+plot(time,model)
+#chibest=sum((gaussian-model).^2)
+chibest=compute_chi(atrial)
+
+println("Initial chi-square: ",chibest)
+#read(STDIN,Char)
+
+result = optimize(compute_chi, atrial, BFGS(), OptimizationOptions(autodiff = true))
+
+println(Optim.minimizer(result),Optim.minimum(result))
+abest= Optim.minimizer(result)
+clf()
+plot(time,gaussian)
+model = zeros(ntime)
+for k=1:nexp
+  model += abest[k].*cos((k-1)*2.*time)
+end
+plot(time,model)
+plot(time,(gaussian-model))
+chi=sum((gaussian-model).^2)
+println("New minimum: ",chi," ",abest," ",std(gaussian-model)," ",maximum(abs(gaussian-model)))
+#read(STDIN,Char)
+coeff_grid[i,:] = abest
+end
+
+#coeff1 = log10(vec(coeff_grid[:,1]))
+
+#function coeff1_fit(x)
+## Fit the first coefficient:
[…]

### 2016-10-25 paper/ms.tex
Added kernel approximation plot; cleaned up notation.

@@
 \item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
-correlations in data, and can be represented with $\beta =  0$.
-\item {\bf Exponential kernel.} This is exactly
-represented with $\beta_I=0$.
-\item {\bf Cosine kernel.} The cosine kernel may be represented exactly with $\beta_R=0$.
+correlations in data, and can be represented with $\beta_R = \beta_I =  0$.
+\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
+represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
+\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$, 
+may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
 \item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
@@
 $p=1$ is relevant to stellar variability, which we describe next.
-\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ kernel is given by: 
+\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by: 
 \begin{equation}
-K_{3/2}(\tau,\tau_0) = (1+x)e^{-x},
+K_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
 \end{equation} 
-where $x = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
+where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
 $\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
@@
 \begin{equation}
-K_{3/2}(\tau,\tau_0) \approx (1-b)e^{-x}+be^{-x\frac{b-1}{b}},
+K_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
 \end{equation} 
 where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
-large values of $b$, the maximum error on this scales as $2/(e^2b) \approx 0.27/b$; for
-$b=10$, the error in this approximation is $<3$\%.  Note that the two exponential terms
+large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
+$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
 have opposite signs, which may lead to numerical instability for values of $b$ that
@@
 \begin{equation}
-C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
+K_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
 \end{equation}
-where $y = \sqrt{5}x$.  The Fourier transform is:
+where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
 \begin{equation}
@@
 \begin{eqnarray}
-K_{5/2}(y) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
+K_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
  &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
@@
 The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
-power spectrum that requires a steeper spectrum than this, we forgo approximating further Mat\'ern kernels.
+power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
 \item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
 which has been used to model quasi-periodic stellar variability due to star spots.   It is
-given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$, which may be expanded as a series in $\cos{(2kw)}$,
+given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be 
+expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
 with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:  
@@
 $c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
-This approximation is accurate to better than 0.5\% for $\Gamma < 3$.
+This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
+deviation of 0.1\%.
 \end{itemize}
 
-{\color{red} Should we include some plots of these kernels, with the approximations?}
+Examples of the foregoing kernels and their approximations are shown in Figure \ref{kernel_approx}.
+It is clear from this figure that these standard kernels are well approximated by the sums
+of between two and five damped cosine kernels, demonstrating that these form an adequate set
+for modeling a wide range of kernel behaviors.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{kernel_approx.png}
+\caption{Approximation of various kernels with the sums of damped cosines with $\tau_0=1$. 
+(a). Kernels that may be expressed exactly: constant (dark grey), exponential (light grey), 
+cosine (grey), and damped cosine (salmon).
+(b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
+damped cosine kernels (magenta). (c). The approximation of the exponential-sine-squared
+kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
+$K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
+is approximated with $b=100$. }
+\label{kernel_approx}
+\end{center}
+\end{figure}
 

### 2016-10-25 paper/expsin2_approx.jl

@@ -154,7 +154,7 @@ println("Fit to fifth coefficient")
 read(STDIN,Char)
 
 # Now test with random numbers:
-ntest = 100
+ntest = 10000
 dev = zeros(ntest)
 for i=1:ntest
   time = rand()*2pi

### 2016-10-25 paper/plot_approx.jl

@@ -1,0 +1,96 @@
+using PyPlot
+# Set up a time grid for the plot:
+ntime = 1000
+time = collect(linspace(0,2pi,ntime))
+
+# exp-sine-squared:
+x1 = [0.24998970200450718,0.7528029456942368,0.9604772168743339,0.6433934213976025]
+x2 = [0.4508506501130407,-0.44533765457178637,1.0810460905919905,1.1810193449724977]
+x3 = [0.20659251082649457,-0.20736062231716337,1.659019056376522,0.21064077694469815]
+x4 = [0.07284486969656653,-0.0729380659597069,2.431485661375288,0.047122405324731394]
+x5 = [0.019204025483424515,-0.01921291003735305,3.2797651575680282,0.011419793258824862]
+
+gamma = 1.0
+model_expsin2_g1=zeros(ntime)
+model_expsin2_g1 += x1[1]+x1[2].*exp(-gamma.^x1[3].*x1[4])
+model_expsin2_g1 += (x2[1]+x2[2].*exp(-gamma.^x2[3].*x2[4])).*cos(2.*time)
+model_expsin2_g1 += (x3[1]+x3[2].*exp(-gamma.^x3[3].*x3[4])).*cos(4.*time)
+model_expsin2_g1 += (x4[1]+x4[2].*exp(-gamma.^x4[3].*x4[4])).*cos(6.*time)
+model_expsin2_g1 += (x5[1]+x5[2].*exp(-gamma.^x5[3].*x5[4])).*cos(8.*time)
+expsin2_g1 = exp(-gamma.*sin(time).^2)
+
+gamma = 3.0
+model_expsin2_g3=zeros(ntime)
+model_expsin2_g3 += x1[1]+x1[2].*exp(-gamma.^x1[3].*x1[4])
+model_expsin2_g3 += (x2[1]+x2[2].*exp(-gamma.^x2[3].*x2[4])).*cos(2.*time)
+model_expsin2_g3 += (x3[1]+x3[2].*exp(-gamma.^x3[3].*x3[4])).*cos(4.*time)
+model_expsin2_g3 += (x4[1]+x4[2].*exp(-gamma.^x4[3].*x4[4])).*cos(6.*time)
+model_expsin2_g3 += (x5[1]+x5[2].*exp(-gamma.^x5[3].*x5[4])).*cos(8.*time)
+expsin2_g3 = exp(-gamma.*sin(time).^2)
+
+# Gaussian (squared-exponential):
[…]

### 2016-10-26 paper/plot_approx.jl
Made changes to kernel figure.

@@ -1,7 +1,7 @@
 using PyPlot
 # Set up a time grid for the plot:
-ntime = 1000
-time = collect(linspace(0,2pi,ntime))
+ntime = 10000
+time = collect(linspace(0,4pi,ntime))
 
 # exp-sine-squared:
 x1 = [0.24998970200450718,0.7528029456942368,0.9604772168743339,0.6433934213976025]
@@ -52,14 +52,14 @@ fix,axes = subplots(2,2)
 
 ax = axes[1]
 ax[:set_title]("(a). Simple, exact kernels")
-ax[:plot](time,exp(-time),alpha=0.5,label=L"$e^{-t}$",linewidth=4,color="silver")
 ax[:plot](time,1.+0.*time,alpha=0.5,label=L"$1$",linewidth=4,color="black")
-ax[:plot](time,cos(time),alpha=1.0,alpha=0.5,label=L"$\cos{t}$",color="grey",linewidth=4)
-ax[:plot](time,exp(-time).*cos(time),alpha=1.0,alpha=0.5,label=L"$e^{-t}\cos{t}$",color="coral",linewidth=4)
-ax[:legend](loc="lower left")
+ax[:plot](time,exp(-time/3),alpha=0.5,label=L"$e^{-t/3}$",linewidth=4,color="c")
+ax[:plot](time,cos(time),alpha=1.0,alpha=0.5,label=L"$\cos{(t)}$",color="olive",linewidth=4)
+ax[:plot](time,exp(-time./3).*cos(time),alpha=1.0,alpha=0.5,label=L"$e^{-t/3}\cos{(t)}$",color="coral",linewidth=4)
+ax[:legend](loc="lower center",fontsize=10)
 ax[:set_xlabel]("Time")
 ax[:set_ylabel]("Dimensional ACF")
-ax[:axis]([0,2pi,-1.1,1.1])
+ax[:axis]([0,4pi,-1.1,1.1])
 
 ax = axes[3]
 ax[:set_title]("(b). Gaussian")
@@ -76,7 +76,7 @@ ax[:plot](time,model_expsin2_g1,alpha=1.0,label=L"$E(w,1)$, approx",ls="dashed",
 ax[:plot](time,expsin2_g3,alpha=0.25,label=L"$E(w,3)$, exact",linewidth=4,color="g")
[…]

### 2016-10-27 paper/ms.tex
Made changes to kernel figure.
Revised Gaussian fit; drafted code for simulating GP in O(N)

@@
 \end{equation}
-Note that this encompasses all smooth, time-staionary kernels, and generally $c_1 \le 0$ as
-correlated noise initially decreases with separation in time.  We wish to approximate this
+Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
+correlated noise initially declines with separation in time.  We wish to approximate this
 as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
-can carry out the approximation in terms of sums of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)}$,
+can carry out the approximation in terms of sums of damped cosine kernel components
+\begin{equation}
+K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
+\end{equation}
 which has a Taylor series expansion:
-\begin{equation}
-\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} = 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 -
+\begin{eqnarray}
+\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\c &-&
 \frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
-\end{equation}
+\end{eqnarray}
 which may be continued to any order in $\tau$ as necessary.
@@
 requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
-chosen to match the coefficients $c_j$.  Not that for $N=1$ we only require a damped exponential
+chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
 kernel with $\beta_I = 0$.
 
-Another way to view this is in the power spectrum of the Gaussian Process, which is the
-Fourier Transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
+Another way to view this is in the power spectrum of the gaussian process, which is the
+Fourier transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
 is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
@@
 of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
-can be arranged to cancel in the wings, giving features in the power spectrum that are
-narrower in frequency.
+can be arranged to cancel in the wings, giving features in the power spectrum that have
+narrower wings.
 
-The functional form of $K_j$ may be used to approximate some commonly used Kernel components: radial
-kernels can be approximated with $\beta_I=0$, while periodic kernels can be approximated by
-$\beta_I \ne 0$.  The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
+The functional form of $K_j$ may be used to approximate some commonly used Kernel components.
+The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
 used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
 commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be 
-used to approximate each of these kernels, with various degrees of success:
+used to approximate each of these kernels to better than 1\% with various numbers of damped
+cosine kernel components:
 \begin{itemize}
-\item {\bf Constant kernel} A constant kernel component can be used to capture long-timescale
-correlations in data, and can be represented with $\beta_R = \beta_I =  0$.
+\item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
+correlations in data, and can be represented exactly with $\beta_R = \beta_I =  0$.
 \item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
@@
 has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
-Nevertheless, it may be well approximated by the sum of three exponential kernels:
+Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
+which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of a
+Lorentzian.  Nevertheless, the squared exponential kernel may be well approximated by the sum of 
+three exponential kernels:
 \begin{eqnarray}
-G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
- &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
- &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
+%G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
+% &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
+% &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
+G(z) &\approx& 2.670143 e^{-1.500258 z} \cos{(0.346731 z)} \\
+ &-& 0.458705 e^{-4.125588 z} \cos{(1.390912 z)}\\
+ &-& 1.211438 e^{-1.744602 z} \cos{(1.798436 z)},
 \end{eqnarray}
-which has an accuracy of better than $<4\times 10^{-4}$ for all z.
+which has an accuracy of better than $<2.5\times 10^{-4}$ for all z.
+Note that the last three coefficients depend on the first six as we have constrained
+the first three terms of the Taylor series to match that of a Gaussian exactly, $1-z^2/2$.
+% [2.6701433354886324,1.5002583976327974,-0.3467305042550739,-0.4587049531317026,4.125587583864702,1.3909117346609636] -1.2114383823569297 1.7446017341309372 1.7984359925567543 
 %[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
@@
 values of $p$, the Mat\'ern kernel may be approximated well by the sum of
-$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel, while
+$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel,
+$C_{1/2}(x) = \sigma^2 e^{-x}$ ($p=0$), while
 $p=1$ is relevant to stellar variability, which we describe next.
@@
 \end{equation} 
-where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier Transform\footnote{Using the normalization convention: 
+where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier transform\footnote{Using the normalization convention: 
 $\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
@@
 \includegraphics[width=\textwidth]{kernel_approx.png}
-\caption{Approximation of various kernels with the sums of damped cosines with $\tau_0=1$. 
-(a). Kernels that may be expressed exactly: constant (dark grey), exponential (light grey), 
-cosine (grey), and damped cosine (salmon).
+\caption{Approximation of various kernels with the sums of damped cosines.
+(a). Kernels that may be expressed exactly: constant (dark grey), exponential (cyan), 
+cosine (olive), and damped cosine (salmon).
 (b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
@@
 $K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
-is approximated with $b=100$. }
+is approximated with $b=100$. The exact kernels are solid, while approximate are dashed.}
 \label{kernel_approx}
@@
 
-A standard practice in constructing kernels for solving problems with Gaussian Process
+A standard practice in constructing kernels for solving problems with Gaussian processes
 is to add or multiply the preceding standard kernel components.  In the case of damped cosine
@@
 \frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
-&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k}) +K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
+&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
+&+&\left. K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
 \end{eqnarray}
@@
 as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
-two coefficients may be sampled from a logarithm, which the individual
+two coefficients may be sampled from a logarithm, while the individual
 values are allowed to be negative;  this will enforce the positive definite
@@
 \section{Summary}
+
+Although we have in mind application of this fast method to stellar variability,
+the method is general for one-dimensional GP problems, and may be applied to
+other problems.    Within astrophysics, correlated noise (due to the environment,
+detector, or modeling uncertainty) may be present in
+gravitational wave time series, and so Gaussian processes may be a way to address
+this problem \citep{Moore2016}.  Accreting black holes show time series which
+may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
+motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,1995PhRvL..74.1060R}.
+This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
+measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
+and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.
 

### 2016-10-27 julia/simulate_exp_gp.jl

@@ -1,0 +1,29 @@
+function simulate_exp_gp(t,alpha,beta,ndev)
+#
+# Function for simulating a GP based upon an exponential correlation
+# matrix using an analytic form for the Cholesky decomposition and
+# a computation in O(N) operations.  The autocorrelation function is:
+#
+#    K_ij = alpha * exp(-beta*(|t_i-t_j|))
+#
+# Requirements:
+#  - alpha should be a real positive number.
+#  - The times t *must* be sorted in order from least to greatest.
+#  - beta may be complex
+#  - ndev: normal deviates drawn from N(0,1) with the same length as time vector t.
+# Output:
+#  - data is a GP drawn from this correlation function with length nt
+#
+nt = length(t)
+data = zeros(eltype(beta),nt)
+gamma = zeros(eltype(beta),nt)
+for i=1:nt-1
+  gamma[i] = exp(-beta*(t[i+1]-t[i]))
+  data[i] = sqrt(1.0-gamma[i]^2)*sqrt(alpha)*ndev[i]
+end
+data[nt] = sqrt(alpha)*ndev[nt]
+for i=nt-1:-1:1
+  data[i] += gamma[i]*data[i+1]
+end
+return data
+end

### 2016-10-27 julia/test_sim_gp.jl

@@ -1,0 +1,10 @@
+using PyPlot
+
+include("simulate_exp_gp.jl")
+nt = 100000
+t = collect(linspace(0,10,nt))
+beta = 0.3
+alpha = 1.0
+ndev = randn(nt)
+data = simulate_exp_gp(t,alpha,beta,ndev)
+plot(t,data)

### 2016-10-27 paper/gaussian_approx3.jl

@@ -1,0 +1,99 @@
+# Try to approximate squared-exponential kernel:
+using PyPlot
+using Optim
+
+function gaussian_approx3()
+
+ntime = 1000
+time = collect(linspace(0,10,ntime))
+gaussian=exp(-time.^2/2.)
+
+function compute_chi(x)
+model = zeros(ntime)
+c1 = 1.0
+c2 = 0.0
+c3 = 1.0
+for k=1:nexp
+  model += x[1+(k-1)*3].*exp(-x[2+(k-1)*3].*time).*cos(x[3+(k-1)*3].*time)
+  c1 -= x[1+(k-1)*3]
+  c2 -= x[1+(k-1)*3]*x[2+(k-1)*3]
+  c3 += x[1+(k-1)*3]*(x[2+(k-1)*3]^2-x[3+(k-1)*3]^2)
+end
+c2 = c2/c1
+c3 = sqrt((c3+c1*c2^2)/c1)
+#model += c1.*exp(-c2.*time).*cos(x[nexp*3+1].*time)
+model += c1.*exp(-c2.*time).*cos(c3.*time)
+return sum((gaussian-model).^2)
+end  
+
+nexp = 2
+#nexp = 2
+#nexp = 4
+npar = 3*nexp
+abest=zeros(npar)
+#abest[1:12]=[1./6.,0.1,sqrt(3.),1./6.,0.1,sqrt(5.),1./3.,0.1,0.1,1./3.,0.1,0.2]
+# Here are best parameters so far:
+#New minimum: 855 2.522483789529161 [0.09778291672462784,1.5548632770098678,0.32192744041675075,0.596959899332403,0.3132368383113517,-0.18163977156887445,0.2670528045472698,-1.075894228060323]
+#abest=[0.09778291672462784,1.5548632770098678,0.32192744041675075,0.596959899332403,0.3132368383113517,-0.18163977156887445,0.2670528045472698,-1.075894228060323]
+#New minimum: 4.2559231434310284e-7 [1.4707280328402002,1.3556988707754538,1.9133776307579318,-0.7331091089111925,3.6344847969483203,1.3207054708438177,-2.5676669507901395,1.4654679150660068,-1.8772555559415482,2.8301497749105042,1.5683246010038931,0.32031045862921226] 2.0325193117951805e-5
[…]

### 2016-10-27 paper/plot_approx.jl

@@ -29,7 +29,8 @@ model_expsin2_g3 += (x5[1]+x5[2].*exp(-gamma.^x5[3].*x5[4])).*cos(8.*time)
 expsin2_g3 = exp(-gamma.*sin(time).^2)
 
 # Gaussian (squared-exponential):
-abest=[2.7518074963750734,1.5312499407621147,-0.34148888178746056,-0.3546256365636662,4.753504292338284,2.042571639590954,-1.3968344706386124,1.7967431129812566,-1.7641684073510928]
+#abest=[2.7518074963750734,1.5312499407621147,-0.34148888178746056,-0.3546256365636662,4.753504292338284,2.042571639590954,-1.3968344706386124,1.7967431129812566,-1.7641684073510928]
+abest=[2.670143,1.500258, 0.346731,-0.458705,4.125588, 1.390912, -1.211438,1.744602, 1.798436]
 gaussian=exp(-time.^2/2.)
 nexp = 3
 
@@ -54,10 +55,10 @@ ax = axes[1]
 ax[:set_title]("(a). Simple, exact kernels")
 ax[:plot](time,1.+0.*time,alpha=0.5,label=L"$1$",linewidth=4,color="black")
 ax[:plot](time,exp(-time/3),alpha=0.5,label=L"$e^{-t/3}$",linewidth=4,color="c")
-ax[:plot](time,cos(time),alpha=1.0,alpha=0.5,label=L"$\cos{(t)}$",color="olive",linewidth=4)
-ax[:plot](time,exp(-time./3).*cos(time),alpha=1.0,alpha=0.5,label=L"$e^{-t/3}\cos{(t)}$",color="coral",linewidth=4)
+ax[:plot](time,cos(time),alpha=1.0,alpha=0.5,label=L"$\cos{(\tau)}$",color="olive",linewidth=4)
+ax[:plot](time,exp(-time./3).*cos(time),alpha=1.0,alpha=0.5,label=L"$e^{-t/3}\cos{(\tau)}$",color="coral",linewidth=4)
 ax[:legend](loc="lower center",fontsize=10)
-ax[:set_xlabel]("Time")
+ax[:set_xlabel](L"$\tau/\tau_0$")
 ax[:set_ylabel]("Dimensional ACF")
 ax[:axis]([0,4pi,-1.1,1.1])
 
[…]

### 2016-11-07 paper/ms.tex
working on CI

@@
 
-We derive a fast, $O(N)$, Gaussian Process approach that involves Lorentzian
-kernels.  This basis set can approximate commonly used kernels, and is a good
-description of stellar variability.
+We present a scalable method for Gaussian Process regression in one dimension
+with a specific emphasis on large astronomical time-series data sets.
+This method can be applied to any Gaussian Process model where the spectral
+density can be expressed as any general mixture of Lorentzian functions.
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{kernel_approx.png}
+\includegraphics[width=\textwidth]{./approximate_kernels/kernel_approx.pdf}
 \caption{Approximation of various kernels with the sums of damped cosines.
@@
 \begin{center}
-\includegraphics[scale=0.175]{./twoterm.eps}
+\includegraphics[scale=0.175]{./twoterm.pdf}
 \hspace{0.2in}
-\includegraphics[scale=1.0]{./colorcode.eps}
+\includegraphics[scale=1.0]{./colorcode.pdf}
 \end{center}

### 2016-11-07 .ci/travis.sh

@@ -1,0 +1,31 @@
+#!/bin/bash -x
+
+# Download the requested version of Eigen
+mkdir -p eigen
+cd eigen
+wget --quiet "http://bitbucket.org/eigen/eigen/get/${EIGEN_VERSION}.tar.gz"
+tar -xf ${EIGEN_VERSION}.tar.gz --strip-components 1
+cd ..
+
+# If testing C++, deal with that here
+if [[ $TEST_LANG == cpp ]]
+then
+  cd cpp
+  cmake . -DEIGEN_CHECK_INCLUDE_DIRS=../eigen
+  make
+  cd ..
+  return
+fi
+
+# Conda Python
+hash -r
+conda config --set always_yes yes --set changeps1 no
+conda update -q conda
+conda info -a
+conda create --yes -n test python=$PYTHON_VERSION numpy=$NUMPY_VERSION Cython setuptools pytest
+source activate test
+
+# Build the extension
+cd python
+CXX=g++ python setup.py build_ext -I../eigen --inplace
+cd ..

### 2016-11-08 paper/ms.tex
some words in the intro

@@
 
-Stars are variable, for better or worse:  their variability provides information about the properties
-of stars, but also results in noise that must be accounted for in detecting and characterizing their
-planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
-goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
-properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
-flux variability and spectral variations.
+Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
+models for time-series analysis.
+For GP modeling, a functional form is chosen to describe the autocovariance
+of the data and the parameters of this function are fit for or marginalized.
+In the astrophysical literature, GPs have been used to model stochastic
+variability in light curves of stars (CITE), active galactic nuclei (CITE),
+and X-ray binaries (CITE).
+They have also been used as models for the cosmic microwave background (CITE),
+correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
+residuals caused by model inconsistencies (CITE + better words).
+While these models are widely applicable, their use has been limited, in
+practice, by the computational cost and scaling.
+In general, the cost of computing a GP likelihood scales as the third power of
+the number of data points $\mathcal{O}(N^3)$ and in the current era of large
+time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
+observations each~---~this cost is prohibitive.
 
-A promising technique for addressing both of these goals is the Gaussian Process formalism
-\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
-is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
-correlations is completely specified by a correlation matrix.  In its simplest form, the
-correlation matrix is modeled with a time-independent autocorrelation function (although
-the time-independence can be relaxed), including a white-noise component that may vary
-with each exposure, and the autocorrelation function is described by a functional form that may be
-simply parameterized by a relatively small number of parameters, sometimes referred to as
-`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
-describe the autocorrelation function, thus inferring properties of the stellar variabilty
-which may be used to characterize properties of the star, such as the rotation period,
-the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
-The goals of planet detection include measuring the Doppler shift of the star, measuring
-the decrement of flux as a planet transits a star, and/or measuring the times of transit to
-look for dynamical interactions between planets.  To properly account for the impact of
-stellar variability on the planet parameters requires treating correlations in the data, and
-to do so also requires evaluating the correlation matrix.
+In this paper, we present a class of GP models that enable likelihood
+calculations that scale linearly with the number of data points
+$\mathcal{O}(N)$ for one dimensional data sets.
+This method is a generalization of a method developed by
+\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
+year old paper \citep{Rybicki:1995}.
+For this method to be applicable, the data must be one-dimensional and the
+covariance function must be written as a mixture of Lorentzian functions in
+Fourier space.
+However, there is no further constraint on the data or the model.
+In particular, the measurements don't need to be evenly spaced and the
+uncertainties can be heteroscedastic.
+This method is especially appealing compared to other similar methods~--~we
+will return to these below~--~because it is exact, flexible, robust, simple,
+and fast.
 
-The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
-$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
-prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
-its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
-= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
-using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
-{\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
-$N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
-cases.
+In the following pages, we will motivate the general problem of GP regression,
+describe the previously published scalable method \citep{Rybicki:1995,
+Ambikasaran:2015} and our generalization, and demonstrate the model's
+application on various real and simulated data sets.
+Alongside this paper, we have released efficient and well-tested
+implementations of this method written in \project{C++}, \project{Python}, and
+\project{Julia}.
+These implementations are available online at \project{GitHub}
+\url{https://github.com/dfm/GenRP} and \project{Zenodo} \dfmtodo{add zenodo
+archive}.
 
-One promising solution is to use approximate techiques based upon the fact that correlation
-matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
-algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
-solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
-solver is complicated and can be challenging to use with general choices of kernel.  Another
-promising approximate technique is to use a uniform resampling of the dataset, and then
-rely on the structure of Toeplitz matrices, which require uniformly space time series,
-to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
-the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
-this approach has the drawback of being approximate.  A third approach is to use a kernel
-based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
-model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
-is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
-of the Lorentzians are required to have a specific relation.  In addition, these
-techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}
+% DFM: I'm going to redistribute the following in other sections.
 
-The approach we explore in this paper is based upon a method developed by Press \&
-Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
-matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
-The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
-They also indicated that a combination of two exponential kernels would have a simliar scaling,
-but they did not demonstrate how to extend this to more than two kernel components.
+% Stars are variable, for better or worse:  their variability provides information about the properties
+% of stars, but also results in noise that must be accounted for in detecting and characterizing their
+% planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
+% goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
+% properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
+% flux variability and spectral variations.
 
-A significant advance was made in generalizing the Press-Rybicki method to an
-arbitrary number of exponential kernels by utilizing the fact that the elements of
-such matrices can be written in terms of the products of components of two vectors,
-which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
-Rybicki-Press' (\texttt{GenRP}) approach
-has many advantages:  the computation time and storage both scale as $O(N_{time})$,
-the coefficients of the kernels need not be stationary, the computation of the
-likelihood is exact, and the algorithm is simple to implement, and thus is robust.
-Unlike the \texttt{CARMA} model, there is no restriction on the relation between
-the coefficients (except that the kernel needs to be positive definite).  Unlike
-\texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
-of this approach is that the sum of exponential kernels with real coefficients can
-only represent a limited range of power spectra of auto-correlated time series:
-those which can be expressed the sum of Lorentzians that peak at zero frequency.
+% A promising technique for addressing both of these goals is the Gaussian Process formalism
+% \citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
+% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
+% correlations is completely specified by a correlation matrix.  In its simplest form, the
+% correlation matrix is modeled with a time-independent autocorrelation function (although
+% the time-independence can be relaxed), including a white-noise component that may vary
+% with each exposure, and the autocorrelation function is described by a functional form that may be
+% simply parameterized by a relatively small number of parameters, sometimes referred to as
+% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
+% describe the autocorrelation function, thus inferring properties of the stellar variabilty
+% which may be used to characterize properties of the star, such as the rotation period,
+% the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
+% The goals of planet detection include measuring the Doppler shift of the star, measuring
+% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
+% look for dynamical interactions between planets.  To properly account for the impact of
+% stellar variability on the planet parameters requires treating correlations in the data, and
+% to do so also requires evaluating the correlation matrix.
 
-However, this drawback turns out to not be significant:  the exponential kernel
-can also have a complex exponential coefficients (plus their complex-conjugates), and
-thus approximate an almost arbitrary autocorrelation function with enough terms, most
-importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
-components is a {\it very} good approximation of stellar variability spectra:  the
-activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
-Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
-Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
-parameterized this way as well.  In this paper
-we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
-combined with the complex conjugate leads to damped harmonic auto-correlation functions
-which are flexible enough to describe a wide range of variability, and may still
-be solved in $O(N_{time}$ operations.
+% The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
+% $N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
+% prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
+% its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
+% = N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
+% using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
+% {\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
+% $N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
+% cases.
 
-We show that the extended matrix that embeds this multi-component kernel can be written
-in real notation, which speeds up the evaluation of the likelihood function,
-and allows optimization software to use automatic differential in computing
-derivatives of the likelihood function.
+% One promising solution is to use approximate techiques based upon the fact that correlation
+% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
+% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
+% solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
+% solver is complicated and can be challenging to use with general choices of kernel.  Another
+% promising approximate technique is to use a uniform resampling of the dataset, and then
+% rely on the structure of Toeplitz matrices, which require uniformly space time series,
+% to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
+% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
+% this approach has the drawback of being approximate.  A third approach is to use a kernel
+% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
+% model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
+% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
+% of the Lorentzians are required to have a specific relation.  In addition, these
+% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}
 
-The plan of the paper is...
+% The approach we explore in this paper is based upon a method developed by Press \&
+% Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
+% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
+% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
+% They also indicated that a combination of two exponential kernels would have a simliar scaling,
+% but they did not demonstrate how to extend this to more than two kernel components.
+
+% A significant advance was made in generalizing the Press-Rybicki method to an
+% arbitrary number of exponential kernels by utilizing the fact that the elements of
+% such matrices can be written in terms of the products of components of two vectors,
+% which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
+% Rybicki-Press' (\texttt{GenRP}) approach
+% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
+% the coefficients of the kernels need not be stationary, the computation of the
+% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
+% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
+% the coefficients (except that the kernel needs to be positive definite).  Unlike
+% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
+% of this approach is that the sum of exponential kernels with real coefficients can
+% only represent a limited range of power spectra of auto-correlated time series:
+% those which can be expressed the sum of Lorentzians that peak at zero frequency.
+
+% However, this drawback turns out to not be significant:  the exponential kernel
+% can also have a complex exponential coefficients (plus their complex-conjugates), and
+% thus approximate an almost arbitrary autocorrelation function with enough terms, most
+% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
+% components is a {\it very} good approximation of stellar variability spectra:  the
+% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
+% Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
+% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
+% parameterized this way as well.  In this paper
+% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
+% combined with the complex conjugate leads to damped harmonic auto-correlation functions
+% which are flexible enough to describe a wide range of variability, and may still
+% be solved in $O(N_{time}$ operations.
+
+% We show that the extended matrix that embeds this multi-component kernel can be written
+% in real notation, which speeds up the evaluation of the likelihood function,
+% and allows optimization software to use automatic differential in computing
+% derivatives of the likelihood function.
+
+% The plan of the paper is...
 

### 2016-11-09 docs/conf.py
building some docs

@@ -1,0 +1,62 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import os
+import sys
+
+d = os.path.dirname
+sys.path.insert(0, d(d(os.path.abspath(__file__))))
+import genrp  # NOQA
+
+extensions = [
+    "sphinx.ext.autodoc",
+    "sphinx.ext.intersphinx",
+    "sphinx.ext.mathjax",
+]
+intersphinx_mapping = {
+    'python': ('https://docs.python.org/3/', None),
+    'numpy': ('https://docs.scipy.org/doc/numpy/', None),
+    'scipy': ('https://docs.scipy.org/doc/scipy/reference/', None),
+    'astropy': ('http://docs.astropy.org/en/stable/', None),
+    'emcee': ('http://dan.iel.fm/emcee/current/', None)
+}
+
+templates_path = ["_templates"]
+source_suffix = ".rst"
+master_doc = "index"
+
+# General information about the project.
+project = "GenRP"
+author = "Dan Foreman-Mackey, Eric Agol, & contributors"
+copyright = "2016, " + author
+
+version = genrp.__version__
+release = genrp.__version__
+
+exclude_patterns = ["_build"]
+pygments_style = "sphinx"
+
+# Readthedocs.
+on_rtd = os.environ.get("READTHEDOCS", None) == "True"
+if not on_rtd:
+    import sphinx_rtd_theme
+    html_theme = "sphinx_rtd_theme"
+    html_theme_path = [sphinx_rtd_theme.get_html_theme_path()]
+
+html_context = dict(
+    display_github=True,
+    github_user="dfm",
+    github_repo="GenRP",
+    github_version="master",
+    conf_py_path="/docs/",
+    # script_files=[
+    #     "_static/jquery.js",
+    #     "_static/underscore.js",
+    #     "_static/doctools.js",
+    #     "//cdn.mathjax.org/mathjax/latest/MathJax.js"
+    #     "?config=TeX-AMS-MML_HTMLorMML",
+    #     "_static/js/analytics.js",
+    # ],
+)
+html_static_path = ["_static"]
+html_show_sourcelink = False

### 2016-11-10 .ci/travis-linux.sh
making a change to the paper

@@ -1,10 +1,13 @@
 #!/bin/bash -x
 
 # Install conda
-# http://conda.pydata.org/docs/travis.html#the-travis-yml-file
-wget http://repo.continuum.io/miniconda/Miniconda-latest-Linux-x86_64.sh -O miniconda.sh
-bash miniconda.sh -b -p $HOME/miniconda
-export PATH="$HOME/miniconda/bin:$PATH"
+if [[ $TEST_LANG == python ]]
+then
+  # http://conda.pydata.org/docs/travis.html#the-travis-yml-file
+  wget http://repo.continuum.io/miniconda/Miniconda-latest-Linux-x86_64.sh -O miniconda.sh
+  bash miniconda.sh -b -p $HOME/miniconda
+  export PATH="$HOME/miniconda/bin:$PATH"
+fi
 
 # Install Python dependencies
 source "$( dirname "${BASH_SOURCE[0]}" )"/travis.sh

### 2016-11-29 paper/ms.tex
building some docs
making a change to the paper

@@
 
+%
+% New Outline:
+% 1. Introduction
+% 2. Gaussian processes
+%   - GPs in general
+%   - RP method
+%   - Ambikasaran method
+%   - Generalization
+% 3. Implementation and performance
+%   - Implementation considerations: band solver vs. sparse, parameterization
+%   - Benchmark numbers
+%   - Scaling with N and M
+%   - Description of APIs
+% 4. Examples
+%   - Simulated data: a simple exact GenRP simulation, a GenRP simulation with
+%     structure, a few standard kernels, a AGW kernel
+%   - An AGN light curve
+%   - An RR Lyrae
+%   - A kepler light curve
+%   - An asteroseismic target
+% 5. Comparisons to other methods
+% 6. Summary
+%   - Possible applications
+%
+
+\section{Gaussian processes in one-dimension}
+
+\citep{Rasmussen:2006}
+
+\subsection{Gaussian processes with Lorentzian power spectra}
+
+\citep{Rybicki:1995, Ambikasaran:2015}
+
+\subsection{Generalization to quasi-periodic covariance}
+
+Our method.
+
+\section{Implementation \& performance}
+
+\subsection{Implementation considerations}
+
+Solvers.
+Banded vs.\ sparse.
+
+\subsection{Benchmarks \& scaling}
+
+\subsection{Parameterization \& API}
+
+\section{Examples}
+
+\subsection{Simulated data}
+
+\begin{enumerate}
+\item an OU process
+\item a full GenRP process
+\item Some standard kernels: Matern, Exp-squared, etc.
+\item A KISS-GP process
+\end{enumerate}
+
+\subsection{Real data}
+
+\begin{enumerate}
+\item AGN
+\item RR Lyrae
+\item Kepler with simulated transit
+\item Asteroseismic target
+\end{enumerate}
+
+\section{Comparisons to other methods}
+
+Toeplitz, KISS-GP, CARMA, HODLR.
+
+Limitations of GenRP: one-dimension, stationary, etc.
+
+\section{Summary}
+
 % DFM: I'm going to redistribute the following in other sections.
@@
 
-\section{Outline}
-
-\begin{enumerate}
+% \section{Outline}
+% \begin{enumerate}
 %\item Non-parametric modeling of time series with correlated noise using Gaussian processes
@@
 %\item Generalized Ambikasarn for complex coefficients (Hermitian).
-\item Performance:\\
-  a). Show $O(N*p^2)$ scaling\\
-  b). Show comparison to HODLR.\\
-  c). Show comparison to wavelet (ala Carter \& Winn).\\
-  d). Compare with CARMA Kallman filter(?).
-\item Various methods/formulae:\\
-  a). Likelihood computation.\\
-  b). Generating GP with particular covariance [yet to solve]. [ ]\\
-  c). Inferring particular components.\\
-  d). Computing derivatives of GPs (as used in RV methods).\\
-  e). Forecasting/interpolating.\\
-  f). Inferring confidence intervals.\\
-  g). Derivative of likelihood function with respect to kernel parameters.\\
-  h). Non-stationary GP (can the coefficients vary with time?).\\
-  i). Multi-band (each element of extended matrix becomes a matrix).
-  j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
-\item Generalization to non-stationary noise.
-\item Example application: TYC 3559\\
-  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
-  b). Analysis of solar data.
-  c). CO2 Mauna Kea
-  d). TBD Computing derivatives of GPs (as used in RV methods).\\
-  e). TBD Forecasting/interpolating.\\
-  f). TBD Inferring confidence intervals.\\
-  g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
-  h). TBD Non-stationary GP (can the coefficients vary with time?).\\
-  i). TBD Multi-band (each element of extended matrix becomes a matrix).
+% \item Performance:\\
+%   a). Show $O(N*p^2)$ scaling\\
+%   b). Show comparison to HODLR.\\
+%   c). Show comparison to wavelet (ala Carter \& Winn).\\
+%   d). Compare with CARMA Kallman filter(?).
+% \item Various methods/formulae:\\
+%   a). Likelihood computation.\\
+%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
+%   c). Inferring particular components.\\
+%   d). Computing derivatives of GPs (as used in RV methods).\\
+%   e). Forecasting/interpolating.\\
+%   f). Inferring confidence intervals.\\
+%   g). Derivative of likelihood function with respect to kernel parameters.\\
+%   h). Non-stationary GP (can the coefficients vary with time?).\\
+%   i). Multi-band (each element of extended matrix becomes a matrix).
+%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
+% \item Generalization to non-stationary noise.
+% \item Example application: TYC 3559\\
+%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
+%   b). Analysis of solar data.
+%   c). CO2 Mauna Kea
+%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
+%   e). TBD Forecasting/interpolating.\\
+%   f). TBD Inferring confidence intervals.\\
+%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
+%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
+%   i). TBD Multi-band (each element of extended matrix becomes a matrix).
 
-\item Example application: TYC 3559\\
-  a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
-  b). Simulated datasets - hyperparameter/PSD inference.\\
-  c). Analysis of solar data?
+% \item Example application: TYC 3559\\
+%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
+%   b). Simulated datasets - hyperparameter/PSD inference.\\
+%   c). Analysis of solar data?
 
-\item Possible future applications:\\
-  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
-      with Lorentzians, not Gaussians.\\
-  b). Multi-waveband GPs.\\
-  c). RV interpolation.\\
-  d). Non-parametric phase functions.\\
-  e). Better calibration of the flicker log(g) relation.\\
-  f). Measurement of rotational periods - gyrochronology.\\
-  g). Doppler shifts in oscillating EBs.
+% \item Possible future applications:\\
+%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
+%       with Lorentzians, not Gaussians.\\
+%   b). Multi-waveband GPs.\\
+%   c). RV interpolation.\\
+%   d). Non-parametric phase functions.\\
+%   e). Better calibration of the flicker log(g) relation.\\
+%   f). Measurement of rotational periods - gyrochronology.\\
+%   g). Doppler shifts in oscillating EBs.
 
-\item Appendix:
-  a). Description/summary of mathematics.\\
-  b). Description of code.
-\end{enumerate}
+% \item Appendix:
+%   a). Description/summary of mathematics.\\
+%   b). Description of code.
+% \end{enumerate}
 

### 2016-12-16 paper/ms.tex
words about parameterization
adding notation for the Lorentzian limit and a notebook checking the parameters
Added computation of SHO, ACF of spot models

@@
 % The approach we explore in this paper is based upon a method developed by Press \&
-% Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
+% Rybicki twenty years ago \citep{Rybicki:1995}.  They observed that a correlation
 % matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
@@
 % such matrices can be written in terms of the products of components of two vectors,
-% which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
+% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
 % Rybicki-Press' (\texttt{GenRP}) approach
@@
 %  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
-%  c). Tri-diagonal (Press \& Rybicki) \citep{1995PhRvL..74.1060R}
+%  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
 %  d).
@@
 %  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
-%\item Solution: Ambikasaran \citep{Ambikasaran2015} shows how sum of exponential kernels can be solved in order
+%\item Solution: Ambikasaran \citep{Ambikasaran:2015} shows how sum of exponential kernels can be solved in order
 %   $O(N*p^2)$, where p is number of Lorentzians.
@@
 where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
-matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran2015}.
+matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
 The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
@@
 may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
-motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,1995PhRvL..74.1060R}.
+motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,Rybicki:1995}.
 This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
@@
 
-Here we summarize \citet{Ambikasaran2015}, and then point to appendix
+Here we summarize \citet{Ambikasaran:2015}, and then point to appendix
 for how to do this with oscillating kernels.
@@
 Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
-follow closely the notation and equations given in \citet{Ambikasaran2015}.  We start
+follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
 with the observation that a Lorentzian correlation function may be decomposed into the sum of two
@@
 for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
-in \citet{Ambikasaran2015}.
-
-Equation (60) in \citet{Ambikasaran2015} has real and imaginary components:
+in \citet{Ambikasaran:2015}.
+
+Equation (60) in \citet{Ambikasaran:2015} has real and imaginary components:
 \begin{eqnarray} \label{Amb60}
@@
 with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
-\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.
-
-Finally, equation (59) in \citet{Ambikasaran2015} has real and imaginary components:
+\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.
+
+Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
 \begin{align} \label{Amb59}
@@
 we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
-imaginary components of equations (59) and (60) in \citet{Ambikasaran2015} as well for these, which reduces the total number
+imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
 of equations to solve.
@@
 
-The matrix has a similar structure as the real case constructed by \citep{Ambikasaran2015}, but the
+The matrix has a similar structure as the real case constructed by \citep{Ambikasaran:2015}, but the
 imaginary and real components mix (this is due to the representation of imaginary numbers as
@@
 $p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
-that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran2015}.
+that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran:2015}.
 
@@
 \caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
-the example shown in \citet{Ambikasaran2015}.}
+the example shown in \citet{Ambikasaran:2015}.}
 \label{matrix_structure}

### 2016-12-16 paper/approximate_kernels/two_components.jl

@@ -1,0 +1,131 @@
+using PyPlot
+using Polynomials
+
+# Compute PSD for a two-component Lorentzian with
+# both complex coefficients.  Check that PSD is positive
+# and plot it.
+
+#psd_check(x)
+x = rand(8)
+# x=[ar1,ai1,br1,bi1,ar2,ai2,br2,bi2]
+
+ar1=x[1]
+ai1=x[2]
+br1=x[3]
+bi1=x[4]
+ar2=x[5]
+ai2=x[6]
+br2=x[7]
+bi2=x[8]
+# Compute d=0 case 10% of time:
+if rand() < 0.1
+  bi2 = (ar1*br1+ar2*br2-ai1*bi1)/ai2
+end
+
+d = ar1*br1 - ai1*bi1 + ar2*br2 - ai2*bi2
+
+a = (2(br2^2-bi2^2)*(ar1*br1-ai1*bi1)+2(br1^2-bi1^2)*(ar2*br2-ai2*bi2)
+ +(bi1^2+br1^2)*(ai1*bi1+ar1*br1)+(bi2^2+br2^2)*(ai2*bi2+ar2*br2))
+
+b = (2(br2^2-bi2^2)*(bi1^2+br1^2)*(ai1*bi1+ar1*br1)
++2(br1^2-bi1^2)*(bi2^2+br2^2)*(ai2*bi2+ar2*br2)
++(bi2^2+br2^2)^2*(ar1*br1-ai1*bi1)
++(bi1^2+br1^2)^2*(ar2*br2-ai2*bi2))
+
+c = ((bi2^2+br2^2)^2*(bi1^2+br1^2)*(ai1*bi1+ar1*br1)
++(bi1^2+br1^2)^2*(bi2^2+br2^2)*(ai2*bi2+ar2*br2))
+
+if abs(d) < eps()
+  d = 0.0
+  println("Reached quadratic test")
+  disc = b^2-4a*c
+  if disc < 0
+# there is no z_max, so set it to a negative value:
+    zmax = -1.0
+  else
+    if a > 0
+      zmax = (-b+sqrt(disc))/(2*a)
+    else
+      zmax = (-b-sqrt(disc))/(2*a)
+    end
+  end
+else
+
+  a /= d
+  b /= d
+  c /= d
+
+  delta = 18a*b*c-4a^3*c+a^2*b^2-4b^3-27c^2
+
+  if delta >= 0
+    disc = sqrt(a^2-3b)
[…]

### 2016-12-18 paper/ms.tex
Described SHO; added placeholder for Sturm's theorem.

@@
 %   - Generalization
+%   - Relation to stochastically-driven, damped simple harmonic oscillator
 % 3. Implementation and performance
 %   - Implementation considerations: band solver vs. sparse, parameterization
+%   - Ensuring positive-definite covariance matrix: Sturm's theorem
 %   - Benchmark numbers
@@
 
+The general ACF is given by
+\begin{equation}
+K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
+\end{equation}
+which has a power spectrum given by
+\begin{equation}
+P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
+\end{equation}
+The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
+Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
+$a_rb_r-a_ib_i \ge 0$.
+
+\subsection{Relation to stochastically-driven, damped simple harmonic oscillator}
+
+Intrinsic stellar variability consists primarily of random fluctuations driven by 
+convection, oscillations excited by convection, and rotational variability.  These first
+two types of variability may be well-described by a model of a damped simple-harmonic oscillator
+driven by stochastic variability \citep{1990ApJ...364..699A}.  The governing differential equation is
+\begin{equation}
+y^{\prime\prime}+ \frac{\omega_0}{Q} y^\prime + \omega_0^2 y = \epsilon,
+\end{equation}
+where $y(t)$ is a displacement, $^{\prime}$ represents a time derivative,
+$\omega_0$ is the frequency of an undamped oscillator, $Q$ is the quality factor
+of the oscillation, and $\epsilon(t)$ is a stochastic driving force.
+
+In the limit of an infinite time series and white random forcing, the noise-free power 
+spectrum of this equation is given by:
+\begin{equation}
+P(\omega) = P_0 \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
+\end{equation}
+where $P_0$ is a normalization constant.
+
+This power spectrum matches that of our basis kernel if $\omega_0^2 = b_r^2+b_i^2$,
+$a_rb_r-a_ib_i = 0$, $a_ib_i+a_rb_r=\omega_0^2$, and $4b_r^2 = \omega_0^2/Q^2$.
+The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
+frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
+in Figure \ref{sho_psd} for various values of $Q$.  This power spectrum has
+a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)}$, which is
+greater than zero for $Q > 1/\sqrt{2}$.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{./approximate_kernels/sho.png}
+\caption{Normalized power spectra of the stochastically-driven, damped simple
+harmonic oscillator.}
+\label{sho_psd}
+\end{center}
+\end{figure}
+
+This power spectrum has several limits of interest:
+\begin{itemize}
+\item $Q=1/2$:  This results in a power spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
+This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
+autocovariance function:
+\begin{equation}
+K(\tau) = a_r e^{-b_r \vert\tau\vert} \left(\cos{(b_i \vert\tau\vert)} + \frac{b_r}{b_i}\sin{(b_i \vert\tau\vert)}\right),
+\end{equation}
+in the limit that $b_i \rightarrow \infty$, for which the second term has a coefficients that diverges,
+but with large values of $b_i$, this may approximate the Mat\'ern kernel well.  Note that this
+power spectrum peaks at $\omega=0$.
+\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
+form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
+dependence
+\begin{equation}
+K(\tau) = a_r e^{-b_r\vert\tau\vert} \left(\cos{(b_i\vert\tau\vert)}+\sin{(b_i\vert\tau\vert)}\right).
+\end{equation}
+Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
+peak at zero frequency; such a low quality oscillator doesn't display even quasi-periodic behavior.
+\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.  The
+power spectrum approaches
+\begin{equation}
+P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
+\end{equation}
+where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
+a width of $1/(2Qx_{max})$ and a peak height $\propto Q^2$.
+\end{itemize}
+
+The stochastically driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic stellar variability, with
+low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing asteroseismic variability.
+
 \section{Implementation \& performance}
@@
 Banded vs.\ sparse.
+
+\subsection{Ensuring positive definiteness: Sturm's theorem}
+
+Relation of positive power spectrum to positive definite covariance matrix.
+Power spectrum of a sum of kernels.
+Sturm's theorem applied to numerator of power spectrum.
 

### 2016-12-18 paper/ms.tex
Matched notation

@@
 
+Parameters:
+\begin{eqnarray}\eqlabel{kernel}
+k(\tau) &=& \sum_{j=1}^p \left[
+    a_j\,\exp\left(-c_j\,\tau\right)\,\cos\left(-d_j\,\tau\right) +
+    b_j\,\exp\left(-c_j\,\tau\right)\,\sin\left(-d_j\,\tau\right)
+\right]
+\end{eqnarray}
+where $\tau = |\Delta t|$.
+Equivalently, this can be written as
+\begin{eqnarray}
+k(\tau) &=& \sum_{j=1}^p \left[
+    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau\right) +
+    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau\right)
+\right]\quad.
+\end{eqnarray}
+
+The power spectral density is computed by taking the Fourier transform of
+\eq{kernel} to find
+\begin{eqnarray}
+S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
+\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
+{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
+\end{eqnarray}
+
+For some $b_j=0$, the PSD for the $j$-th component becomes
+\begin{eqnarray}
+S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
+    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
+    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
+\right]
+\end{eqnarray}
+a pair of Lorentzian distributions with width $c_j$ centered on
+$\omega = \pm d_j$.
+
 \citep{Rybicki:1995, Ambikasaran:2015}
@@
 
-The general ACF is given by
-\begin{equation}
-K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
-\end{equation}
-which has a power spectrum given by
-\begin{equation}
-P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
-\end{equation}
-The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
-Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
-$a_rb_r-a_ib_i \ge 0$.
+%The general ACF is given by
+%\begin{equation}
+%K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
+%\end{equation}
+%which has a power spectrum given by
+%\begin{equation}
+%P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
+%\end{equation}
+%The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
+%Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
+%$a_rb_r-a_ib_i \ge 0$.
 
@@
 
-This power spectrum matches that of our basis kernel if $\omega_0^2 = b_r^2+b_i^2$,
-$a_rb_r-a_ib_i = 0$, $a_ib_i+a_rb_r=\omega_0^2$, and $4b_r^2 = \omega_0^2/Q^2$.
+% Check to see that signs match
+This power spectrum matches that of our basis kernel if $\omega_0^2 = c_j^2+d_j^2$,
+$a_jc_j-b_jd_j = 0$, $b_jd_j+a_jc_j=\omega_0^2$, and $4c_j^2 = \omega_0^2/Q^2$.
 The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
@@
 \begin{equation}
-K(\tau) = a_r e^{-b_r \vert\tau\vert} \left(\cos{(b_i \vert\tau\vert)} + \frac{b_r}{b_i}\sin{(b_i \vert\tau\vert)}\right),
+K(\tau) = a_j e^{-c_j \vert\tau\vert} \left(\cos{(d_j \vert\tau\vert)} + \frac{c_j}{d_j}\sin{(d_j \vert\tau\vert)}\right),
+% Check to see that signs match in this equation.
 \end{equation}

### 2016-12-19 paper/ms.tex
Minor edits

@@
 % Check to see that signs match
-This power spectrum matches that of our basis kernel if $\omega_0^2 = c_j^2+d_j^2$,
-$a_jc_j-b_jd_j = 0$, $b_jd_j+a_jc_j=\omega_0^2$, and $4c_j^2 = \omega_0^2/Q^2$.
+This power spectrum matches that of our basis kernel if
+\begin{eqnarray}
+\omega_0^2 &=& c_j^2+d_j^2,\\
+a_jc_j-b_jd_j &=& 0,\\
+\omega_0^2&=& b_jd_j+a_jc_j ,\\
+\omega_0^2/Q^2&=& 4c_j^2.
+\end{eqnarray}
 The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
 frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
-in Figure \ref{sho_psd} for various values of $Q$.  This power spectrum has
-a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)}$, which is
+in Figure \ref{sho_psd} for various values of $Q$.  
+Note that the oscillation frequency that appears in the kernel, $d_j$, differs
+from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
+frequency, $c_j$.  This power spectrum has
+a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
 greater than zero for $Q > 1/\sqrt{2}$.
@@
 
-The stochastically driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic stellar variability, with
-low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing asteroseismic variability.
+The stochastically-driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic 
+stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing 
+asteroseismic variability.
 
@@
 \item Asteroseismic target
+\item Stellar rotational variability
 \end{enumerate}

### 2016-12-19 paper/ms.tex

@@
 %\begin{equation}
-%K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
+%k(\tau) = a_r e^{-b_r \tau} \cos{(b_r \tau)} + a_i e^{-b_r \tau} \sin{(b_r \tau)}.
 %\end{equation}
@@
 \begin{equation}
-P(\omega) = P_0 \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
+P(\omega) = P_0 \sqrt{\frac{2}{\pi}} \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
 \end{equation}
@@
 \begin{eqnarray}
-\omega_0^2 &=& c_j^2+d_j^2,\\
-a_jc_j-b_jd_j &=& 0,\\
-\omega_0^2&=& b_jd_j+a_jc_j ,\\
-\omega_0^2/Q^2&=& 4c_j^2.
-\end{eqnarray}
-The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
+a_j &=& P_0  \omega_0 Q\\
+b_j &=& a_j \left(4Q^2-1\right)^{-1/2}\\
+c_j &=& \frac{\omega_0}{2Q}\\
+d_j &=& \omega_0 \left(1-\frac{1}{4Q^2}\right)^{1/2},
+%\omega_0^2 &=& c_j^2+d_j^2,\\
+%a_jc_j-b_jd_j &=& 0,\\
+%\omega_0^2&=& b_jd_j+a_jc_j ,\\
+%\omega_0^2/Q^2&=& 4c_j^2.
+\end{eqnarray}
+for $Q \ge \frac{1}{2}$ and $j=p=1$.  This gives a kernel with dependence
+\begin{eqnarray}\eqlabel{sho_kernel}
+%k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
+k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
+\lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
+\tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
+\end{eqnarray}
+The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
 frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
@@
 \begin{center}
-\includegraphics[width=\textwidth]{./approximate_kernels/sho.png}
+\includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
 \caption{Normalized power spectra of the stochastically-driven, damped simple
-harmonic oscillator.}
+harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
 \label{sho_psd}
@@
 \begin{itemize}
-\item $Q=1/2$:  This results in a power spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
+\item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power 
+spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
 This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
@@
 \begin{equation}
-K(\tau) = a_j e^{-c_j \vert\tau\vert} \left(\cos{(d_j \vert\tau\vert)} + \frac{c_j}{d_j}\sin{(d_j \vert\tau\vert)}\right),
+k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
 % Check to see that signs match in this equation.
 \end{equation}
-in the limit that $b_i \rightarrow \infty$, for which the second term has a coefficients that diverges,
-but with large values of $b_i$, this may approximate the Mat\'ern kernel well.  Note that this
-power spectrum peaks at $\omega=0$.
+in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second 
+term has a coefficient that diverges, but with large values of $d_j$, this may approximate the 
+Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
 \item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
@@
 \begin{equation}
-K(\tau) = a_r e^{-b_r\vert\tau\vert} \left(\cos{(b_i\vert\tau\vert)}+\sin{(b_i\vert\tau\vert)}\right).
-\end{equation}
+k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
+\end{equation}
+corresponding to $\phi = \pi/4$.
 Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
-peak at zero frequency; such a low quality oscillator doesn't display even quasi-periodic behavior.
-\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.  The
-power spectrum approaches
+peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
+\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak. 
+The kernel approaches
+\begin{equation}
+k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
+\end{equation}
+for $Q \gg 1$.  The power spectrum approaches
 \begin{equation}
@@
 where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
-a width of $1/(2Qx_{max})$ and a peak height $\propto Q^2$.
+a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
 \end{itemize}
 
-The stochastically-driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic 
-stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing 
-asteroseismic variability.
+The stochastically-driven, damped simple-harmonic oscillator can then describe a wide range of intrinsic 
+stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q \gg 1$ representing 
+asteroseismic variability.  Taking the sum over oscillators with various $Q, P_0,$ and $\omega_0$ can
+give an accurate accounting of the power spectrum of stellar variability with $Q \ge \frac{1}{2}$ and
+$P_0 >0$.  As this kernel is exactly described by the exponential kernel, this functional form
+is tractable to solving in $\mathcal{O}(N)$ operations, as described next.
 
@@
 
-Relation of positive power spectrum to positive definite covariance matrix.
-Power spectrum of a sum of kernels.
-Sturm's theorem applied to numerator of power spectrum.
+The power spectrum is computed from taking the square of the Fourier transform of the data
+time series; hence, the power-spectrum must be non-negative.  In constructing a kernel,
+if any of the parameters $a_j$ or $b_j$ are negative, then it is possible for the power
+spectrum to go negative, which violates the definition of the power spectrum.  Consequently,
+if any of these coefficients is negative, it is necessary to check that the power spectrum
+is still non-negative.  Note that a non-negative power-spectrum does not require the auto-correlation
+function to be positive for all $\tau$.  The positive power spectrum is related to the
+positive eigenvalues of the covariance matrix which are necessary for a positive-definite matrix
+in limiting cases \citep{messerschmitt2006autocorrelation}.  We find empirically that requiring an everywhere
+positive power spectrum results in positive eigenvalues for the covariance matrix, and
+so we describe here how to ensure a positive power spectrum using Sturm's theorem.
+%Relation of positive power spectrum to positive definite covariance matrix.
+%Power spectrum of a sum of kernels.
+
+In the case of $p$ complex Lorentzians we can check for negative values of the PSD by solving for the
+roots of the power spectrum, abbreviating with $z = \omega^2$:
+\begin{equation}
+P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
+\end{equation}
+where
+\begin{eqnarray}
+q_j &=& a_jc_j-b_jd_j\\
+r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
+s_j &=& 2(c_j^2-d_j^2)\\
+t_j &=& (c_j^2+d_j^2)^2.
+\end{eqnarray}
+The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
+\begin{equation}
+P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
+\end{equation}
+which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
+which may be solved exactly for the roots.
+
+A procedure based upon Sturm's theorem \citep{Dorrie1965} allows one to determine whether there are any 
+real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative 
+$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing 
+$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.
+
+We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
+give us the signs of these polynomials evaluated at $z=0$.
+Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial 
+as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where 
+$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.
+
+With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each 
+of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign 
+changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given 
+by $N_{+}=\sigma(0)-\sigma(\infty)$.
+
+We have checked that this procedure works for a wide range of parameters, and we find that it robustly 
+matches the number of positive real roots which we evaluated numerically.
+
+The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic 
+manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
+likelihood may be set to zero.
+%Sturm's theorem applied to numerator of power spectrum.
 
@@
 \begin{equation}
-K(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
+k(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
 \end{equation}
@@
 \begin{equation}
-K(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
+k(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
 \end{equation}
@@
 which may be continued to any order in $\tau$ as necessary.
-If an approximation to $K(\tau)$ is desired in a Taylor series to some high order $N$, this
+If an approximation to $k(\tau)$ is desired in a Taylor series to some high order $N$, this
 requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
@@
 correlations in data, and can be represented exactly with $\beta_R = \beta_I =  0$.
-\item {\bf Exponential kernel.} This kernel, $K(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
+\item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
 represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
-\item {\bf Cosine kernel.} The cosine kernel, $K(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
+\item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
 may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
@@
 \begin{eqnarray}
-\bar f_{*,k} &=& \sum_j K(t_{*,k},t_j) b_j,\\
+\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
 &=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
 &=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
-y_k &=& \sum_j K(t_j,t_k) b_j,
+y_k &=& \sum_j k(t_j,t_k) b_j,
 \end{eqnarray}
@@
 matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
-matrix $K({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
+matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
 of $MN$ multiplications is required to obtain the predicted values.
@@
 \begin{eqnarray}
-cov(f_{*,k}) &=&  K(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
-&=& K(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
-\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
+\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
 &=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
@@
 \begin{eqnarray}
-K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
+k(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
  &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
@@
 the example shown in \citet{Ambikasaran:2015}.}
-\label{matrix_structure}
+\figurelabel{matrix_structure}
 \end{figure}

### 2016-12-21 paper/ms.tex
Reran exp-squared fit using Sturm's theorem; updated paper

@@
 
-A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
-plus the white noise component:
-\begin{equation}
-k(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \vert\tau\vert^j.
-\end{equation}
-Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
-correlated noise initially declines with separation in time.  We wish to approximate this
-as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
-can carry out the approximation in terms of sums of damped cosine kernel components
-\begin{equation}
-K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
-\end{equation}
-which has a Taylor series expansion:
-\begin{eqnarray}
-\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\c &-&
-\frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 + \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
-\end{eqnarray}
-which may be continued to any order in $\tau$ as necessary.
-If an approximation to $k(\tau)$ is desired in a Taylor series to some high order $N$, this
-requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
-chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
-kernel with $\beta_I = 0$.
-
-Another way to view this is in the power spectrum of the gaussian process, which is the
-Fourier transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
-is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
-width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
-power spectrum is: {\color{red} Check that this is true.}
-\begin{equation}
-P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
-\end{equation}
-where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
-that a smooth distribution may be approximated arbitrarily precisely by summing together
-Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
-a peak in the power spectrum is narrow at a specific frequency, then a small value of
-$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
-of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
-can be arranged to cancel in the wings, giving features in the power spectrum that have
-narrower wings.
-
-The functional form of $K_j$ may be used to approximate some commonly used Kernel components.
+%A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
+%plus the white noise component:
+%\begin{equation}
+%k(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \tau^j.
+%\end{equation}
+%Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
+%correlated noise initially declines with separation in time.  We wish to approximate this
+%as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
+%can carry out the approximation in terms of sums of damped cosine kernel components
+%\begin{equation}
+%K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
+%\end{equation}
+%which has a Taylor series expansion:
+%\begin{eqnarray}
+%\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\
+%- \frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 &+& \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
+%\end{eqnarray}
+%which may be continued to any order in $\tau$ as necessary.
+%If an approximation to $k(\tau)$ is desired in a Taylor series to some high order $N$, this
+%requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
+%chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
+%kernel with $\beta_I = 0$.
+%
+%Another way to view this is in the power spectrum of the gaussian process, which is the
+%Fourier transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
+%is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
+%width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
+%power spectrum is: {\color{red} Check that this is true.}
+%\begin{equation}
+%P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
+%\end{equation}
+%where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
+%that a smooth distribution may be approximated arbitrarily precisely by summing together
+%Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
+%a peak in the power spectrum is narrow at a specific frequency, then a small value of
+%$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
+%of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
+%can be arranged to cancel in the wings, giving features in the power spectrum that have
+%narrower wings.
+
+The functional form of $k_j$ may be used to approximate some commonly used Kernel components.
 The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
@@
 Lorentzian.  Nevertheless, the squared exponential kernel may be well approximated by the sum of
-three exponential kernels:
+four exponential kernels:
 \begin{eqnarray}
-%G(z) &\approx& 2.751807 e^{-1.531250 z} \cos{(0.341489 z)} \\
-% &-& 0.354626 e^{-4.753504 z} \cos{(2.042572 z)}\\
-% &-& 1.396834 e^{-1.796743 z} \cos{(1.764168 z)},
-G(z) &\approx& 2.670143 e^{-1.500258 z} \cos{(0.346731 z)} \\
- &-& 0.458705 e^{-4.125588 z} \cos{(1.390912 z)}\\
- &-& 1.211438 e^{-1.744602 z} \cos{(1.798436 z)},
-\end{eqnarray}
-which has an accuracy of better than $<2.5\times 10^{-4}$ for all z.
-Note that the last three coefficients depend on the first six as we have constrained
-the first three terms of the Taylor series to match that of a Gaussian exactly, $1-z^2/2$.
-% [2.6701433354886324,1.5002583976327974,-0.3467305042550739,-0.4587049531317026,4.125587583864702,1.3909117346609636] -1.2114383823569297 1.7446017341309372 1.7984359925567543
-%[2.751807496353032,1.5312499408182438,-0.34148888180984543,-0.35462563655835033,4.753504292338051,2.0425716395901157,-1.3968344706423566,1.7967431129750422,-1.7641684073374964
+G(z) &\approx& 63.119512 e^{-1.479579 z} -32.173615 e^{-1.593691 z} -29.425156 e^{-1.388929 z}\\
+&+& e^{-1.383928 z} \left[-0.519196\cos{(1.883396 z)} +0.285478 \sin{(1.883396 z)}\right] \\
+\end{eqnarray}
+which has an accuracy of better than $<3\times 10^{-3}$ for all z, and the power spectrum has
+been constrained to be positive.
 \item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
@@
 \begin{equation}
-K_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
+k_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
 \end{equation}
@@
 \begin{equation}
-\hat K_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
+\hat k_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
 \end{equation}
@@
 \begin{equation}
-K_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
+k_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
 \end{equation}
@@
 \begin{equation}
-K_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
+k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
 \end{equation}
@@
 \begin{equation}
-\hat K_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
+\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
 \end{equation}
@@
 \begin{eqnarray}
-K_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
+k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
  &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
@@
 kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
-$K_{3/2}$ (yellow/brown) and $K_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
+$k_{3/2}$ (yellow/brown) and $k_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
 is approximated with $b=100$. The exact kernels are solid, while approximate are dashed.}
@@
 \begin{eqnarray}
-K_j(\tau)K_k(\tau) &=&
+k_j(\tau)k_k(\tau) &=&
 \frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
-&=& \frac{1}{2} \left(K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
-&+&\left. K_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
+&=& \frac{1}{2} \left(k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
+&+&\left. k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
 \end{eqnarray}
@@
 \hspace{0.2in}
-\includegraphics[scale=1.0]{./colorcode.pdf}
+\includegraphics[scale=0.85]{./colorcode.pdf}
 \end{center}

### 2016-12-21 paper/approximate_kernels/gaussian_approx7.jl

@@ -1,0 +1,90 @@
+# Try to approximate squared-exponential kernel:
+using PyPlot
+using Optim
+using Roots
+using Polynomials
+include("sturms_theorem.jl")
+
+#function gaussian_approx6()
+
+ntime = 1000
+time = collect(linspace(0,10,ntime))
+gaussian=exp(-time.^2/2.)
+
+function gauss_model(x)
+model = zeros(ntime)
+for k=1:nexp
+  if k > 1
+    x[2+(k-1)*4]=0.0
+    x[4+(k-1)*4]=0.0
+  end
+  model += x[1+(k-1)*4].*exp(-x[3+(k-1)*4].*time).*cos(x[4+(k-1)*4].*time)
+  model += x[2+(k-1)*4].*exp(-x[3+(k-1)*4].*time).*sin(x[4+(k-1)*4].*time)
+end
+return model
+end
+
+function compute_chi(x)
+model = gauss_model(x)
+sum_coeff = zeros(x)
+# Compute Taylor-series coefficients:
+for k=1:nexp
+  sum_coeff[1] += x[1+(k-1)*4]
+  sum_coeff[2] +=-x[1+(k-1)*4]*x[3+(k-1)*4]+x[2+(k-1)*4]*x[4+(k-1)*4]
+  sum_coeff[3] += x[1+(k-1)*4]*x[3+(k-1)*4]^2-2*x[2+(k-1)*4]*x[3+(k-1)*4]*x[4+(k-1)*4]-x[1+(k-1)*4]*x[4+(k-1)*4]^2
+  sum_coeff[4] +=-x[1+(k-1)*4]*x[3+(k-1)*4]^3+3*x[2+(k-1)*4]*x[3+(k-1)*4]^2*x[4+(k-1)*4]+3*x[1+(k-1)*4]*x[3+(k-1)*4]*x[4+(k-1)*4]^2-x[2+(k-1)*4]*x[4+(k-1)*4]^3
+end
+
+return sum((gaussian-model).^2) +100.*((sum_coeff[1]-1.0)^2+sum_coeff[2]^2+(sum_coeff[3]+1.0)^2+sum_coeff[4]^2)+(sturms_theorem(x) > 0)*1e10
+end  
+
+nexp = 4
+npar = 4*nexp
+
[…]

### 2016-12-21 paper/approximate_kernels/plot_approx.jl

@@ -30,13 +30,16 @@ expsin2_g3 = exp(-gamma.*sin(time).^2)
 
 # Gaussian (squared-exponential):
 #abest=[2.7518074963750734,1.5312499407621147,-0.34148888178746056,-0.3546256365636662,4.753504292338284,2.042571639590954,-1.3968344706386124,1.7967431129812566,-1.7641684073510928]
-abest=[2.670143,1.500258, 0.346731,-0.458705,4.125588, 1.390912, -1.211438,1.744602, 1.798436]
+#abest=[2.670143,1.500258, 0.346731,-0.458705,4.125588, 1.390912, -1.211438,1.744602, 1.798436]
+abest = [-0.519196,0.285478,1.383928,1.83396, 63.119512,0.0,1.479579,0.0, -32.173615,0.0,1.593691,0.0, -29.425156,0.0,1.388929,0.0]
+
 gaussian=exp(-time.^2/2.)
-nexp = 3
+nexp = 4
 
 model_gaussian = zeros(ntime)
 for k=1:nexp
-  model_gaussian += abest[1+(k-1)*3].*exp(-abest[2+(k-1)*3].*time).*cos(abest[3+(k-1)*3].*time)
+  model_gaussian += abest[1+(k-1)*4].*exp(-abest[3+(k-1)*4].*time).*cos(abest[4+(k-1)*4].*time)
+  model_gaussian += abest[2+(k-1)*4].*exp(-abest[3+(k-1)*4].*time).*sin(abest[4+(k-1)*4].*time)
 end
 
 # Matern 5/2:

### 2016-12-22 paper/ms.tex
Revised notation; Lorentzian -> damped sinusoid.

@@
 This method can be applied to any Gaussian Process model where the spectral
-density can be expressed as any general mixture of Lorentzian functions.
+density can be expressed as any general mixture of damped sinusoid functions.
 
@@
 For this method to be applicable, the data must be one-dimensional and the
-covariance function must be written as a mixture of Lorentzian functions in
-Fourier space.
+covariance function must be written as a mixture of damped sinusoid functions.
 However, there is no further constraint on the data or the model.
@@
 
-In the case of $p$ complex Lorentzians we can check for negative values of the PSD by solving for the
+In the case of $p$ damped sinusoids we can check for negative values of the PSD by solving for the
 roots of the power spectrum, abbreviating with $z = \omega^2$:
@@
 
-The kernel is modeled as a sum of damped harmonics,
-\begin{equation}
-k(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
-\end{equation}
-where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay
-in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
-$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
-the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
-time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
-of the measured times if the correlated components are zero.
-
-Now, the damped harmonic kernel components may be rewritten as:
-\begin{equation}
- a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
-\end{equation}
-where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
-matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
-The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
-but otherwise the mathematical form of the extended matrix is identical, but with twice as many
-exponential kernel components as the real case.
-
-Although this is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
+%The kernel is modeled as a sum of damped harmonics,
+%\begin{equation}
+%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
+%\end{equation}
+%where $\tau > 0$ is a time-offset between data points, $c_j$ represents the decay
+%in correlation with time and $d_j$ represents oscillations of period $P = 2\pi/d_j$, and
+%$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
+%the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
+%time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
+%of the measured times if the correlated components are zero.
+
+%Now, the damped harmonic kernel components may be rewritten as:
+%\begin{equation}
+% a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
+%\end{equation}
+%where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
+%matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
+%The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
+%but otherwise the mathematical form of the extended matrix is identical, but with twice as many
+%exponential kernel components as the real case.
+
+Although the complex kernel is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
 in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
-arithmetic.  In addition, it turns out that the complex conjugate component, $\beta^* = \beta_R - i\beta_I$, in
+arithmetic.  In addition, it turns out that the complex conjugate component, $(c_j+id_j)^* = c_j - id_j$, in
 the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
-component $\beta = \beta_R + i \beta_I$;  thus the computation is redundant.  Finally, complex
+component $c_j + i d_j$;  thus the computation is redundant.  Finally, complex
 arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
@@
 
-Consequently, we use the fact that $\exp{(i\beta_I \tau)} =  \cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ to
+Consequently, we use the fact that $\exp{(id_j\tau)} =  \cos{(d_j\tau)} + i \sin{(d_j \tau)}$ to
 rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
@@
 while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$,
-so that $\cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ may be rewritten as the rotation matrix
+so that $\cos{(d_j\tau)} + i \sin{(d_j\tau)}$ may be rewritten as the rotation matrix
 \begin{equation}
 \left(\begin{tabular}{ll}
-$\cos{(\beta_I \tau)}$ & $\sin{(\beta_I \tau)}$ \\
-$-\sin{(\beta_I \tau)}$ & $\cos{(\beta_I \tau)}$
+$\cos{(d_j\tau)}$ & $\sin{(d_j\tau)}$ \\
+$-\sin{(d_j\tau)}$ & $\cos{(d_j \tau)}$
 \end{tabular}\right).
@@
 \item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
-correlations in data, and can be represented exactly with $\beta_R = \beta_I =  0$.
+correlations in data, and can be represented exactly with $c_j= d_j =  0$.
 \item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
-represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
+represented with $d_j=0$, $c_j= \tau_0^{-1}$.
 \item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
-may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
+may be represented exactly with $c_j=0$ and $d_j= \frac{2\pi}{P}$.
 \item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
@@
 Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
-which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of a
-Lorentzian.  Nevertheless, the squared exponential kernel may be well approximated by the sum of
-four exponential kernels:
+which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of the
+power spectrum of a damped sinuosoid.  Nevertheless, the squared exponential kernel may be well approximated 
+by the sum of four exponential kernels:
 \begin{eqnarray}
@@
 k_j(\tau)k_k(\tau) &=&
-\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
-&=& \frac{1}{2} \left(k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
-&+&\left. k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
+\frac{1}{2} e^{-(c_j+c_k) \tau} \left((a_ja_k+b_jb_k)\cos{\left[(d_j-d_k)\tau\right]} + (a_jb_k-a_kb_j)\sin{\left[(d_j-d_k)\tau\right]} \right. \\
+ &+&\left.(a_ja_k-b_jb_k)\cos{\left[(d_j+d_k)\tau\right]} -(a_jb_k-a_kb_j)\sin{\left[(d_j+d_k)\tau\right]}\right),
 \end{eqnarray}
-which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
-construct new kernels from damped cosine kernels.
-
-The upshot is that the damped cosine is an expressive basis kernel that may be used
-to represent a wide range of autocorrelation functions relevant for stellar variability.
-We recommend using a series of sums of damped cosines (in which some parameters may be
+which is the sum of two damped sinusoidal kernels.  Thus, only addition is necessary to
+construct new kernels from the product of damped sinuosidal kernels.
+
+The upshot is that the damped sinusoid is an expressive basis kernel that may be used
+to represent a wide range of autocorrelation functions relevant for stellar variability, and
+more.
+We recommend using a series of sums of damped sinusoids (in which some parameters may be
 fixed to specified values to represent particular kernels) to model different amplitudes
@@
 autocorrelation function is available, then these may be fit with this series to
-initialize the coefficients of the damped cosine kernel components.
+initialize the coefficients of the damped sinusoid kernel components.
 Since a Gaussian Process
-must be positive definite, we recommend modeling the logarithm of $a_j$,
-$\beta_{R,j}$, and $\beta_{I,j}$ to enforce this to be the case.  One caveat is
-that to produce certain kernels, a negative value of $a_j$ is needed (such
+must be positive definite, we recommend modeling the logarithm of $a_j$, $b_j$,
+$c_j$, and $d_j$ to enforce this to be the case.  One caveat is
+that to produce certain kernels, a negative value of $a_j$ or $b_j$ is needed (such
 as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
@@
 
+Outside of astronomy, this technique may have application to seismology \citep{Robinson1967},
+
 \vspace{1.5em}
@@
 
-\section{Prediction/interpolation}
-
-Precition/interpolation involves constructing a matrix of covariances
-between the (noisy) training set and the times at which we wish to
-predict or interpolate data points.   If our training times are given by
-${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
-mean values at the predicted times are given by (RW 2.23):
-\begin{eqnarray}
-\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
-&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
-&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
-y_k &=& \sum_j k(t_j,t_k) b_j,
-\end{eqnarray}
-where the latter equation may be solved for ${\bf b}$ using the extended
-matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
-matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
-of $MN$ multiplications is required to obtain the predicted values.
-If $M$ is large, this can become quickly prohibitive.  It turns out
-that the structure of the sum of exponential kernels may be exploited
-to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
-i.e.\ a linear scaling with the total number of data points, which can
-be much fewer if $M$ is large.
-
-The procedure works as follows:
-\begin{itemize}
-\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
-\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
-compute:
-\begin{eqnarray} \label{prediction}
-\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
-\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} b_j\\
-\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} b_j,
-\end{eqnarray}
-where we have separated out the training times before and after the predicted time $t_{*,k}$.
-
-\item Next, we update the coefficients recursively:
-\begin{eqnarray}
-\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} b_j\right]\\
-\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} b_j\right].
-\end{eqnarray}
-With these updates, we can then use equation \ref{prediction} to compute the expected
-mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
-$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
-so this procedure avoids a heavy computational burden.
-
-The variance may be handled in a similar manner, but with $p^2$ coefficients
-to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
-\begin{eqnarray}
-cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
-&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
-\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
-&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
-\end{eqnarray}
-The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
-\end{itemize}
-
-Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
-then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
-just solve for the difference in fewer operations? {\bf TBD}
+%\section{Prediction/interpolation}
+%
+%Precition/interpolation involves constructing a matrix of covariances
+%between the (noisy) training set and the times at which we wish to
+%predict or interpolate data points.   If our training times are given by
+%${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
+%mean values at the predicted times are given by (RW 2.23):
+%\begin{eqnarray}
+%\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
+%&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
+%&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
+%y_k &=& \sum_j k(t_j,t_k) h_j,
+%\end{eqnarray}
+%where the latter equation may be solved for ${\bf h}$ using the extended
+%matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
+%matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
+%of $MN$ multiplications is required to obtain the predicted values.
+%If $M$ is large, this can become quickly prohibitive.  It turns out
+%that the structure of the sum of exponential kernels may be exploited
+%to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
+%i.e.\ a linear scaling with the total number of data points, which can
+%be much fewer if $M$ is large.
+%
+%The procedure works as follows:
+%\begin{itemize}
+%\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
+%\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
+%compute:
+%\begin{eqnarray} \label{prediction}
+%\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
+%\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} h_j\\
+%\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} h_j,
+%\end{eqnarray}
+%where we have separated out the training times before and after the predicted time $t_{*,k}$.
+%
+%\item Next, we update the coefficients recursively:
+%\begin{eqnarray}
+%\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} h_j\right]\\
+%\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} h_j\right].
+%\end{eqnarray}
+%With these updates, we can then use equation \ref{prediction} to compute the expected
+%mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
+%$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
+%so this procedure avoids a heavy computational burden.
+%
+%The variance may be handled in a similar manner, but with $p^2$ coefficients
+%to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
+%\begin{eqnarray}
+%cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+%&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
+%\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
+%&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
+%\end{eqnarray}
+%The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
+%\end{itemize}
+%
+%Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
+%then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
+%just solve for the difference in fewer operations? {\bf TBD}
 
@@
 follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
-with the observation that a Lorentzian correlation function may be decomposed into the sum of two
+with the observation that a damped sinusoid correlation function may be decomposed into the sum of two
 exponential functions:
 \begin{eqnarray}
-k(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
- &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
+k_j(\tau) &=& a_j e^{-c_j\tau} \cos{(-d_j\tau)} + b_j e^{-c_j\tau} \sin{(-d_j \tau)}\\
+ &=& \frac{1}{2}(a_j+ib_j)\left[e^{-(c_j+id_j) \tau}+e^{-(c_j-id_j) \tau}\right],
 \end{eqnarray}
-where $\beta = \beta_R + i\beta_I$ is the
-complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
-is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.
-
-Now, the Lorentzian kernal may be included in the GRP algorithm
-using the two exponential kernels with $\beta_1 = \beta$ and $\beta_2 = \beta^*$.  This
+where $c_j + id_j$ is the complex kernel parameter and $d_j = 2\pi/P$
+is the frequency of oscillation of the kernel with period $P$.
+
+Now, the damped sinusoid kernal may be included in the GRP algorithm
+using the above two exponential kernels.  This
 requires the algorithm to use complex arithmetic, which has several disadvantages:
@@
 
-Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
-find the real and imaginary components of the complex equation for $\beta$ by taking the
-sum and difference of the $\beta$ and $\beta^*$ equations (divided by 2 and $2i$, respectively),
-and then discard the $\beta^*$ equation, which is redundant.  This keeps the same number of
-additional equations for each $\beta_j$ with non-zero imaginary component (four additional
+Rather than using $c_j+id_j$ and its complex conjugate as components of the kernel, we instead
+find the real and imaginary components of the complex equation by taking the
+sum and difference of the $c_j+id_j$ and $c_j-id_j$ equations (divided by 2 and $2i$, respectively),
+and then discard the $c_j-id_j$ equation, which is redundant.  This keeps the same number of
+additional equations for each $(c_j,d_j)$ with non-zero imaginary component (four additional
 equations in the extended matrix), but converts the complex equations to real, thus avoiding
-complex arithmetic, while the kernel components with real $\beta_j$ still only have two
+complex arithmetic, while the kernel components with $d_j=0$ still only have two
 additional components in the extended matrix.  The extended matrix adds an extra band above
-and below the diagonal when the $\beta$ and $\beta^*$ equations are combined since this
+and below the diagonal when the $c_j+id_j$ and $c_j-id_j$ equations are combined since this
 mixes the components of the original equations which are shifted by one column from one
-another.  We will enumerate the number of real $\beta$'s as $p_0$, while the total number
-of $\beta$'s is still $p$.  This gives a total number of equations to be solved with the
+another.  We will enumerate the number of $d_j=0$ cases as $p_0$, while the total number
+of $c_j$'s is still $p$.  This gives a total number of equations to be solved with the
 extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $N$ is the number of data
@@
 
-For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
+For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
 auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
@@
 as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
-to be the real values of $\beta$, and for these we do not include the imaginary component
+to be the kernels with $d_j=0$, and for these we do not include the imaginary component
 equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
-components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
- = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
-\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
+components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_p+id_p) t_{k,k+1})}\right]^T
+ = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
+\cos{(d_j t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-c_j t_{k,k+1}}\sin{(d_j t_{k,k+1})}$.
 Equation (61) becomes:
@@
 % See 9/2/2016 notes
-\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
-\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
-\end{equation}
-for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61)
+\sum_{j=1}^p \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^p \left[
+\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{h_k}{2},
+\end{equation}
+for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
 in \citet{Ambikasaran:2015}.
@@
 \begin{align} \label{Amb59}
--&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
- &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
+-&r_{k,j}^R + \frac{a_j}{2} x_k + \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
+ &r_{k,j}^I + \frac{b_j}{2} x_k + \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
 \end{align}
@@
 
-With these additional equations, we can solve for the likelihood of sums of full Lorentzian kernels
+With these additional equations, we can solve for the likelihood of sums of damped sinusoid
 in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables
-for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the
+for each damped sinusoid with non-zero $\omega$, and due to having a slightly larger bandwidth due to the
 cross terms between the real and imaginary variables.  The off-diagonal components are $p_0+2(p-p_0)+2$ above
 and below the diagonal, for a total bandwidth of $2p_0+4(p-p_0)+5$. In the case in which $p_0=p$, so
-that the imaginary components of each $\beta$ are zero, the off-diagonal components revert to $p+1$
+that $d_j=0$, the off-diagonal components revert to $p+1$
 above and below the diagonal, giving a bandwidth of $2p+3$ (the same as the GRP algorithm).
@@
 factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
-We arrange the auxiliary variables with real values of
-$\beta$ first, followed by the complex variables, yielding the vector
+We arrange the auxiliary variables with the $d_j=0$
+terms first, followed by the complex variables, yielding the vector
 $x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
-x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [b_1/2, \{0; j=1,...,p_0+2(p-p_0)\}, b_2/2, ..., b_N/2]$.
+x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [h_1/2, \{0; j=1,...,p_0+2(p-p_0)\}, h_2/2, ..., h_N/2]$.
 The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
@@
 imaginary and real components mix (this is due to the representation of imaginary numbers as
-$2\times 2$ matrices).  Figure \ref{matrix_structure} shows an example of the structure of the extended
-matrix for two Lorentzians with non-zero $\omega$ values.  This is analagous to the real case with
+$2\times 2$ matrices).  Figure \ref{fig:matrix_structure} shows an example of the structure of the extended
+matrix for two damped sinusoids with non-zero $\omega$ values.  This is analagous to the real case with
 $p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows

### 2016-12-26 paper/ms.tex
an intro to gps

@@
 % Projects:
-\newcommand{\project}[1]{\textsl{#1}}
+\newcommand{\project}[1]{\textsf{#1}}
 \newcommand{\kepler}{\project{Kepler}}
+\newcommand{\genrp}{\project{genrp}}
 
@@
 
-\section{Gaussian processes in one-dimension}
-
-\citep{Rasmussen:2006}
+\section{Gaussian processes}
+
+Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
+models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
+covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
+parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
+respectively.
+Under this model, the log-likelihood of observing a dataset
+\begin{eqnarray}
+\bvec{y} = \left(\begin{array}{ccccc}
+    y_1\quad && \cdots\quad && y_N
+\end{array}\right)^\T
+\end{eqnarray}
+at coordinates
+\begin{eqnarray}
+X = \left(\begin{array}{ccccc}
+    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
+\end{array}\right)^\T
+\end{eqnarray}
+is
+\begin{eqnarray}\eqlabel{gp-likelihood}
+\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
+    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
+        \bvec{r}_\bvec{\theta}
+    -\frac{1}{2}\ln\det K_\bvec{\alpha}
+    - \frac{N}{2} \ln{(2\pi)}
+\end{eqnarray}
+where
+\begin{eqnarray}
+    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
+    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
+    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
+\end{array}\right)^\T
+\end{eqnarray}
+is the vector of residuals and the elements of the covariance matrix $K$ are
+given by $[K_\bvec{\alpha}]_{ij} = k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$.
+The maximum likelihood values for the parameters $\bvec{\theta}$ and
+$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
+maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
+$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
+CITE}.
+Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
+can be obtained by multiplying the likelihood by a prior
+$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
+\dfmtodo{CITE}) algorithm to sample from the posterior probability density.
+
+The application of GP models is generally limited to small datasets because
+the computational cost of computing the inverse and determinant of the matrix
+$K_\bvec{\alpha}$ scales as the cube of the number of data points $N$,
+$\mathcal{O}(N^3)$.
+This means that for large datasets, every evaluation of the likelihood will
+quickly become computationally intractable.
+In this case, standard non-linear optimization or MCMC will no longer be
+practical inference methods.
+
+In the following Section, we present a method of improving this scaling that
+we call the \genrp\ method.
+The \genrp\ method requires using a specific model for the covariance
+$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ and it has several limitations.
+The method can only be applied to one-dimensional datasets.
+When we say ``one-dimensional'' here, it means that the \emph{input
+coordinates} $\bvec{x}_i$ are scalar, $\bvec{x}_i \equiv t_i$.\footnote{We are
+using $t$ as the input coordinate because one-dimensional GPs are often
+applied to time series data but this isn't a real restriction and the \genrp\
+method can be applied to \emph{any} one-dimensional dataset.}
+Furthermore, the covariance function for the \genrp\ method is ``stationary''.
+This means that the function $k_\bvec{\alpha}(t_i,\,t_j)$ is only a function
+of $\tau_{ij} \equiv |t_i - t_j|$.
+
+\section{The genrp model}
 

### 2016-12-26 cpp/include/genrp/carma.h

@@ -6,14 +6,10 @@
 #include <complex>
 #include <Eigen/Dense>
 
+#include "genrp/utils.h"
+
 namespace genrp {
 namespace carma {
-
-template <typename T1, typename T2>
-inline bool isclose (const T1& a, const T2& b) {
-  using std::abs;
-  return (abs(a - b) <= 1e-6);
-}
 
 Eigen::VectorXcd roots_from_params (const Eigen::VectorXd& params) {
   unsigned n = params.rows();
@@ -83,17 +79,18 @@ void get_genrp_coeffs (
   bool is_conj;
   std::complex<double> term1, term2, full_term;
   for (size_t k = 0; k < p_; ++k) {
-    term1 = 0.0;
-    term2 = 0.0;
-    for (size_t l = 0; l < q_ + 1; ++l) {
-      term1 += beta_[l] * pow(arroots_[k], l);
-      term2 += beta_[l] * pow(-arroots_[k], l);
+    term1 = log(beta_[0]);
+    term2 = log(beta_[0]);
+    for (size_t l = 1; l < q_ + 1; ++l) {
+      term1 = _logsumexp(term1, log(beta_[l]) + std::complex<double>(l) * log(arroots_[k]));
+      term2 = _logsumexp(term2, log(beta_[l]) + std::complex<double>(l) * log(-arroots_[k]));
     }
-    full_term = -sigma_ * sigma_ * term1 * term2 / arroots_[k].real();
+    full_term = 2.0 * log(sigma_) + term1 + term2 - log(-arroots_[k].real());
     for (size_t l = 0; l < p_; ++l) {
       if (l != k)
-        full_term /= (arroots_[l] - arroots_[k]) * (std::conj(arroots_[l]) + arroots_[k]);
+        full_term -= log(arroots_[l] - arroots_[k]) + log(std::conj(arroots_[l]) + arroots_[k]);
     }
+    full_term = exp(full_term);
 
     // Check for and discard conjugate pairs.
     if (isclose(full_term.imag(), 0.0) && isclose(arroots_[k].imag(), 0.0)) {

### 2016-12-26 cpp/include/genrp/poly.h

@@ -50,8 +50,8 @@ Eigen::Matrix<T, Eigen::Dynamic, 1> polyrem (const Eigen::Matrix<T, Eigen::Dynam
       p = m - n + 1;
   using std::abs;
   T d, scale = T(1.0) / v[0];
-  Eigen::Matrix<T, Eigen::Dynamic, 1> q = Eigen::Matrix<T, Eigen::Dynamic, 1>::Zero(std::max(p, 1)),
-                  r = u;  // This makes a copy!
+  Eigen::Matrix<T, Eigen::Dynamic, 1> q(std::max(p, 1)), r = u; // This makes a copy!
+  q.setConstant(T(0.0));
   for (int k = 0; k < p; ++k) {
     d = scale * r[k];
     q[k] = d;
@@ -59,11 +59,11 @@ Eigen::Matrix<T, Eigen::Dynamic, 1> polyrem (const Eigen::Matrix<T, Eigen::Dynam
   }
   int strt;
   for (strt = 0; strt < m; ++strt) {
-    if (abs(r[strt]) >= 1e-10) {
-      break;
+    if (abs(r[strt]) >= 1e-8) {
+      return r.tail(m + 1 - strt);
     }
   }
-  return r.tail(m + 1 - strt);
+  return r;
 }
 
 template <typename T>

### 2016-12-26 cpp/include/genrp/solvers/band.h

@@ -74,12 +74,12 @@ int BandSolver<T>::compute (
   using std::abs;
 
   this->computed_ = false;
-  // bool flag = check_coefficients(
-  //   alpha_real, beta_real,
-  //   alpha_complex_real, alpha_complex_imag,
-  //   beta_complex_real, beta_complex_imag
-  // );
-  // if (!flag) return SOLVER_NOT_COMPUTED;
+  bool flag = check_coefficients(
+    alpha_real, beta_real,
+    alpha_complex_real, alpha_complex_imag,
+    beta_complex_real, beta_complex_imag
+  );
+  if (!flag) return SOLVER_NOT_COMPUTED;
   if (x.rows() != diag.rows()) return SOLVER_DIMENSION_MISMATCH;
 
   // Save the dimensions for later use

### 2016-12-26 cpp/include/genrp/utils.h

@@ -1,12 +1,24 @@
 #ifndef _GENRP_UTILS_H_
 #define _GENRP_UTILS_H_
 
+#include <cmath>
 #include <vector>
 #include <Eigen/Core>
 
 #include "genrp/poly.h"
 
 namespace genrp {
+
+template <typename T1, typename T2>
+inline bool isclose (const T1& a, const T2& b) {
+  using std::abs;
+  return (abs(a - b) <= 1e-6);
+}
+
+template <typename T>
+T _logsumexp (const T& a, const T& b) {
+  return b + log(T(1.0) + exp(a - b));
+}
 
 template <typename Derived>
 bool check_coefficients (

### 2016-12-26 cpp/src/carma_comp.cc

@@ -1,4 +1,5 @@
 #include <iostream>
+#include <cmath>
 #include <complex>
 #include <sys/time.h>
 #include <Eigen/Core>
@@ -42,12 +43,28 @@ int main (int argc, char* argv[])
   Eigen::MatrixXd compute_times(max_terms, 3);
   compute_times.setConstant(0.0);
   for (size_t nterms = 1; nterms <= max_terms; ++nterms) {
-    Eigen::VectorXd carma_arparams = Eigen::VectorXd::Random(nterms),
-                    carma_maparams = Eigen::VectorXd::Random(nterms-1),
+    Eigen::VectorXd carma_arparams,
+                    carma_maparams,
                     alpha_real, beta_real, alpha_complex_real, alpha_complex_imag,
                     beta_complex_real, beta_complex_imag;
 
     compute_times(nterms - 1, 0) = nterms;
+
+    bool is_ok = false;
+    while (!is_ok) {
+      // Resample the parameters until good ones are chosen
+      carma_arparams = Eigen::VectorXd::Random(nterms);
+      carma_maparams = Eigen::VectorXd::Random(nterms-1);
+      carma_arparams.array() += 1.0;
+      carma_maparams.array() += 1.0;
+      carma_arparams.array() /= 2.0 * nterms;
+      carma_maparams.array() /= 2.0 * nterms;
+      genrp::carma::CARMASolver carma_solver(0.0, carma_arparams, carma_maparams);
+      carma_solver.get_genrp_coeffs(alpha_real, beta_real,
+        alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag);
+
+      is_ok = genrp::check_coefficients(alpha_real, beta_real, alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag);
+    }
 
     // Compute using the CARMA model.
     strt = get_timestamp();
@@ -65,33 +82,48 @@ int main (int argc, char* argv[])
 
     // Compute using the genrp model.
     genrp::BandSolver<double> solver;
+    int flag = -1;
     strt = get_timestamp();
     for (size_t i = 0; i < niter; ++i) {
-      solver.compute(alpha_real, beta_real, alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag, x, diag);
[…]

### 2016-12-26 cpp/src/test_poly.cc
fixing stupid rem bug

@@ -58,6 +58,7 @@ int main (int argc, char* argv[])
   c.resize(1);
   c << 0.75;
   d = genrp::polyrem(a, b);
+  std::cout << c.transpose() << std::endl << d.transpose() << std::endl;
   ASSERT_ALL_CLOSE("polyrem1", c, d);
   d = genrp::polyrem(b, a);
   ASSERT_ALL_CLOSE("polyrem2", b, d);

### 2016-12-26 paper/ms.tex
fixing stupid rem bug
a few more words about kernel parameters

@@
 is the vector of residuals and the elements of the covariance matrix $K$ are
-given by $[K_\bvec{\alpha}]_{ij} = k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$.
+given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
@@
 The \genrp\ method requires using a specific model for the covariance
-$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ and it has several limitations.
+$k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
 The method can only be applied to one-dimensional datasets.
 When we say ``one-dimensional'' here, it means that the \emph{input
-coordinates} $\bvec{x}_i$ are scalar, $\bvec{x}_i \equiv t_i$.\footnote{We are
+coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
 using $t$ as the input coordinate because one-dimensional GPs are often
@@
 Furthermore, the covariance function for the \genrp\ method is ``stationary''.
-This means that the function $k_\bvec{\alpha}(t_i,\,t_j)$ is only a function
-of $\tau_{ij} \equiv |t_i - t_j|$.
+This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
+of $\tau_{nm} \equiv |t_n - t_m|$.
 
@@
 
-\subsection{Gaussian processes with Lorentzian power spectra}
-
-Parameters:
-\begin{eqnarray}\eqlabel{kernel}
-k(\tau) &=& \sum_{j=1}^p \left[
-    a_j\,\exp\left(-c_j\,\tau\right)\,\cos\left(-d_j\,\tau\right) +
-    b_j\,\exp\left(-c_j\,\tau\right)\,\sin\left(-d_j\,\tau\right)
+To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
+of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
+function is
+\begin{eqnarray}
+k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
+\end{eqnarray}
+where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
+$\delta_{nm}$ is the Dirac delta, and $\bvec{\alpha} = (a,\,c)$.
+The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
+the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
+with a small number of operations for each data point.
+Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
+mixtures of exponentials
+\begin{eqnarray}
+k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
+    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
+\end{eqnarray}
+In this case, the inverse becomes more complicated but \eq{gp-likelihood} can
+still be evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number
+of components in the mixture and $N$ is still the number of data points.
+
+It turns out that this kernel function can be made even more general by
+introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
+In this case, the covariance function becomes
+\begin{eqnarray}\eqlabel{genrp-kernel-complex}
+k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
+    \sum_{j=1}^N &&\left[
+    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
+        \right. \nonumber\\
+    &&+\left.
+    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
 \right]
 \end{eqnarray}
-where $\tau = |\Delta t|$.
-Equivalently, this can be written as
-\begin{eqnarray}
-k(\tau) &=& \sum_{j=1}^p \left[
-    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau\right) +
-    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau\right)
-\right]\quad.
-\end{eqnarray}
-
-The power spectral density is computed by taking the Fourier transform of
-\eq{kernel} to find
+and, for this function, \eq{gp-likelihood} can still be solved with
+$\mathcal{O}(J^2\,N)$ operations.
+The details of this method and a few implementation considerations are
+discussed in \app{implementation}.
+
+\eq{genrp-kernel-complex} can also be expressed as
+\begin{eqnarray}\eqlabel{genrp-kernel}
+k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
+    \sum_{j=1}^J &&\left[
+    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
+        \right.\nonumber\\
+    &&+ \left.
+    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
+\right] \quad.
+\end{eqnarray}
+The Fourier transform of this covariance function is the power spectral
+density (PSD) of the model and it is given by
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-For some $b_j=0$, the PSD for the $j$-th component becomes
+The physical interpretation of this model isn't immediately obvious and we
+will return to a more general discussion of the physical intuition in a moment
+but we can start with one useful special case of this model.
+If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
+term of \eq{genrp-kernel} becomes
+\begin{eqnarray}
+k_j(\tau_{nm}) =
+    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
+\end{eqnarray}
+and the PSD for the this component is
 \begin{eqnarray}
@@
     \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
-\right]
-\end{eqnarray}
-a pair of Lorentzian distributions with width $c_j$ centered on
-$\omega = \pm d_j$.
-
-\citep{Rybicki:1995, Ambikasaran:2015}
-
-\subsection{Generalization to quasi-periodic covariance}
-
-Our method.
+\right] \quad.
+\end{eqnarray}
+This PSD is the sum of two Lorentzian or Cauchy distributions with width
+$c_j$ centered on $\omega = \pm d_j$.
+This model can be interpreted intuitively as a quasiperiodic oscillator with
+amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
+d_j$.
 
@@
 
-\appendix \label{appendixa}
+\appendix
+
+\section{Implementation details}\sectlabel{implementation}
 

### 2016-12-26 paper/ms.tex
edits for SHO

@@
 \end{eqnarray}
+
 The physical interpretation of this model isn't immediately obvious and we
@@
 
-%The general ACF is given by
-%\begin{equation}
-%k(\tau) = a_r e^{-b_r \tau} \cos{(b_r \tau)} + a_i e^{-b_r \tau} \sin{(b_r \tau)}.
-%\end{equation}
-%which has a power spectrum given by
-%\begin{equation}
-%P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
-%\end{equation}
-%The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
-%Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
-%$a_rb_r-a_ib_i \ge 0$.
-
-\subsection{Relation to stochastically-driven, damped simple harmonic oscillator}
-
-Intrinsic stellar variability consists primarily of random fluctuations driven by
-convection, oscillations excited by convection, and rotational variability.  These first
-two types of variability may be well-described by a model of a damped simple-harmonic oscillator
-driven by stochastic variability \citep{1990ApJ...364..699A}.  The governing differential equation is
-\begin{equation}
-y^{\prime\prime}+ \frac{\omega_0}{Q} y^\prime + \omega_0^2 y = \epsilon,
-\end{equation}
-where $y(t)$ is a displacement, $^{\prime}$ represents a time derivative,
-$\omega_0$ is the frequency of an undamped oscillator, $Q$ is the quality factor
-of the oscillation, and $\epsilon(t)$ is a stochastic driving force.
-
-In the limit of an infinite time series and white random forcing, the noise-free power
-spectrum of this equation is given by:
-\begin{equation}
-P(\omega) = P_0 \sqrt{\frac{2}{\pi}} \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
-\end{equation}
-where $P_0$ is a normalization constant.
+Another special case of the \genrp\ model of great physical interest is a
+stochastically driven, damped simple harmonic oscillator.
+The differential equation for this system is
+\begin{equation}
+    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
+    + \omega_0^2\right]\, y(t) = \epsilon(t)
+\end{equation}
+where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
+quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
+force.
+In the limit of an infinite time series and white random forcing, the PSD of
+this equation is given by \citep{Anderson:1990}
+\begin{equation}
+S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
+    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
+\end{equation}
+where $S_0$ is a normalization constant.
 

### 2016-12-28 .ci/travis.sh
words about sho

@@ -15,6 +15,7 @@ then
     echo "Building the paper..."
     export GENRP_BUILDING_PAPER=true
     source "$( dirname "${BASH_SOURCE[0]}" )"/setup-texlive.sh
+    return
   fi
   export GENRP_BUILDING_PAPER=false
   return

### 2016-12-28 paper/ms.tex
words about sho
sho with Q < 1/2
sign error
more words about sho and moving section

@@
 density (PSD) of the model and it is given by
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{genrp-psd}
 S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
@@
 \end{eqnarray}
-
 The physical interpretation of this model isn't immediately obvious and we
 will return to a more general discussion of the physical intuition in a moment
-but we can start with one useful special case of this model.
+but we can start with a discussion of some useful special cases of this model.
+
 If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
@@
 
-Another special case of the \genrp\ model of great physical interest is a
-stochastically driven, damped simple harmonic oscillator.
-The differential equation for this system is
-\begin{equation}
-    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
-    + \omega_0^2\right]\, y(t) = \epsilon(t)
-\end{equation}
-where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
-quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
-force.
-In the limit of an infinite time series and white random forcing, the PSD of
-this equation is given by \citep{Anderson:1990}
-\begin{equation}
-S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
-    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
-\end{equation}
-where $S_0$ is a normalization constant.
-
-% Check to see that signs match
-This power spectrum matches that of our basis kernel if
-\begin{eqnarray}
-a_j &=& P_0  \omega_0 Q\\
-b_j &=& a_j \left(4Q^2-1\right)^{-1/2}\\
-c_j &=& \frac{\omega_0}{2Q}\\
-d_j &=& \omega_0 \left(1-\frac{1}{4Q^2}\right)^{1/2},
-%\omega_0^2 &=& c_j^2+d_j^2,\\
-%a_jc_j-b_jd_j &=& 0,\\
-%\omega_0^2&=& b_jd_j+a_jc_j ,\\
-%\omega_0^2/Q^2&=& 4c_j^2.
-\end{eqnarray}
-for $Q \ge \frac{1}{2}$ and $j=p=1$.  This gives a kernel with dependence
-\begin{eqnarray}\eqlabel{sho_kernel}
-%k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
-k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
-\lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
-\tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
-\end{eqnarray}
-The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
-frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
-in Figure \ref{sho_psd} for various values of $Q$.
-Note that the oscillation frequency that appears in the kernel, $d_j$, differs
-from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
-frequency, $c_j$.  This power spectrum has
-a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
-greater than zero for $Q > 1/\sqrt{2}$.
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
-\caption{Normalized power spectra of the stochastically-driven, damped simple
-harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
-\label{sho_psd}
-\end{center}
-\end{figure}
-
-This power spectrum has several limits of interest:
-\begin{itemize}
-\item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
-spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
-This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
-autocovariance function:
-\begin{equation}
-k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
-% Check to see that signs match in this equation.
-\end{equation}
-in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
-term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
-Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
-\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
-form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
-dependence
-\begin{equation}
-k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
-\end{equation}
-corresponding to $\phi = \pi/4$.
-Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
-peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
-\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
-The kernel approaches
-\begin{equation}
-k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
-\end{equation}
-for $Q \gg 1$.  The power spectrum approaches
-\begin{equation}
-P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
-\end{equation}
-where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
-a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
-\end{itemize}
-
-The stochastically-driven, damped simple-harmonic oscillator can then describe a wide range of intrinsic
-stellar variability, with low $Q \approx 1$ representing granulation noise, and high $Q \gg 1$ representing
-asteroseismic variability.  Taking the sum over oscillators with various $Q, P_0,$ and $\omega_0$ can
-give an accurate accounting of the power spectrum of stellar variability with $Q \ge \frac{1}{2}$ and
-$P_0 >0$.  As this kernel is exactly described by the exponential kernel, this functional form
-is tractable to solving in $\mathcal{O}(N)$ operations, as described next.
+Similarly, setting both $b_j$ and $d_j$ to zero, we get a Ornstein--Uhlenbeck
+process
+\begin{eqnarray}
+k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
+\end{eqnarray}
+with the PSD
+\begin{eqnarray}
+S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
+    \frac{1}{1+\left(\frac{\omega}{c_j}\right)^2} \quad.
+\end{eqnarray}
 
@@
 \subsection{Parameterization \& API}
+
+\section{Genrp as a model of stellar variations}
+
+Another special case of the \genrp\ model of great physical interest is a
+stochastically driven simple harmonic oscillator.
+The differential equation for this system is
+\begin{equation}
+    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
+    + \omega_0^2\right]\, y(t) = \epsilon(t)
+\end{equation}
+where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
+quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
+force.
+In the limit of an infinite time series and white random forcing, the PSD of
+this equation is given by \citep{Anderson:1990}
+\begin{equation}\eqlabel{sho-psd}
+S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
+    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
+\end{equation}
+where $S_0$ is a normalization constant.
+The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
+\begin{eqnarray}
+a_j &=& S_0\,\omega_0\,Q \\
+b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
+c_j &=& \frac{\omega_0}{2\,Q}\\
+d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
+\end{eqnarray}
+for $Q \ge \frac{1}{2}$.
+For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
+terms with parameters
+\begin{eqnarray}
+a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
+        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
+b_{j\pm} &=& 0 \\
+    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
+d_{j\pm} &=& 0 \quad.
+\end{eqnarray}
+It is interesting to note that, because of the damping, the characteristic
+oscillation frequency in this model $d_j$, for any finite quality factor $Q$,
+is not equal to the frequency of the undamped oscillator $\omega_0$.
+
+The power spectrum in \eq{sho-psd} has several limits of physical interest:
+\begin{itemize}
+
+{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
+\begin{eqnarray}
+S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
+\end{eqnarray}
+A model like this has been used to model the granulation is Solar data
+\citep{Michel:2009}}
+
+{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
+\begin{eqnarray}
+S(\omega) =
+\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
+\end{eqnarray}
+with the corresponding covariance function
+\begin{eqnarray}
+k(\tau) &=& \lim_{f \to 0}\,
+    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
+    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
+&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1-\omega_0\,\tau] \quad.
+\end{eqnarray}
+This covariance function is also known as the Mat\'ern-3/2 function
+\citep{Rasmussen:2006}.
+}
+
+{\item Finally, in the limit $Q \to \infty$, \eq{sho-psd} becomes
+\begin{eqnarray}
+S(\omega) =
+\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2-1\right]^2} \quad.
+\end{eqnarray}}
+
+\end{itemize}
+
+% This gives a kernel with dependence
+% \begin{eqnarray}\eqlabel{sho_kernel}
+% %k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
+% k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
+% \lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
+% \tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
+% \end{eqnarray}
+% The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
+% frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
+% in Figure \ref{sho_psd} for various values of $Q$.
+% Note that the oscillation frequency that appears in the kernel, $d_j$, differs
+% from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
+% frequency, $c_j$.  This power spectrum has
+% a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
+% greater than zero for $Q > 1/\sqrt{2}$.
+
+% \begin{figure}[!htbp]
+% \begin{center}
+% \includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
+% \caption{Normalized power spectra of the stochastically-driven, damped simple
+% harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
+% \label{sho_psd}
+% \end{center}
+% \end{figure}
+
+% This power spectrum has several limits of interest:
+% \begin{itemize}
+% \item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
+% spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
+% This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
+% autocovariance function:
+% \begin{equation}
+% k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
+% % Check to see that signs match in this equation.
+% \end{equation}
+% in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
+% term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
+% Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
+% \item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
+% form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
+% dependence
+% \begin{equation}
+% k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
+% \end{equation}
+% corresponding to $\phi = \pi/4$.
+% Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
+% peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
+% \item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
+% The kernel approaches
+% \begin{equation}
+% k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
+% \end{equation}
+% for $Q \gg 1$.  The power spectrum approaches
+% \begin{equation}
+% P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
+% \end{equation}
+% where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
+% a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
+% \end{itemize}
+
+The stochastically-driven, damped simple-harmonic oscillator can then describe
+a wide range of intrinsic stellar variability, with low $Q \approx 1$
+representing granulation noise, and high $Q \gg 1$ representing asteroseismic
+variability.
+Taking the sum over oscillators with various $Q, S_0,$ and $\omega_0$ can give
+an accurate accounting of the power spectrum of stellar variability with $Q
+\ge \frac{1}{2}$ and $S_0 >0$.
+As this kernel is exactly described by the exponential kernel, this functional
+form is tractable to solving in $\mathcal{O}(N)$ operations, as described
+next.
 

### 2016-12-29 paper/ms.tex
changing the title
adding new SHO figure and editing

@@
 \title{%
-Fast and scalable Gaussian process modeling of stellar variability
+Fast and scalable Gaussian process modeling with applications
+to astronomical time series
 \vspace{-3\baselineskip}  % OMG AASTEX6 IS SO BROKEN
@@
 \end{eqnarray}
-In this case, the inverse becomes more complicated but \eq{gp-likelihood} can
-still be evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number
-of components in the mixture and $N$ is still the number of data points.
+In this case, the inverse will be dense but \eq{gp-likelihood} can still be
+evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number of
+components in the mixture and $N$ is still the number of data points.
 
@@
 and the PSD for the this component is
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{lorentz-psd}
 S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
@@
 Another special case of the \genrp\ model of great physical interest is a
-stochastically driven simple harmonic oscillator.
+stochastically-driven simple harmonic oscillator.
 The differential equation for this system is
@@
 with the corresponding covariance function
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{approx-matern}
 k(\tau) &=& \lim_{f \to 0}\,
@@
     \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
-&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1-\omega_0\,\tau] \quad.
+&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
 \end{eqnarray}
@@
 \citep{Rasmussen:2006}.
+This suggests that the Mat\'ern-3/2 covariance can be well approximated using
+the \genrp\ framework with a small value of $f$ in \eq{approx-matern} but we
+caution that this could also lead to numerical issues with the solver.
 }
 
-{\item Finally, in the limit $Q \to \infty$, \eq{sho-psd} becomes
-\begin{eqnarray}
-S(\omega) =
-\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2-1\right]^2} \quad.
+{\item Finally, in the limit of large $Q$, the model approaches a high
+    quality oscillation with frequency $\omega_0$ and covariance function
+\begin{eqnarray}
+k(\tau) \approx
+    \frac{1}{2}\,S_0\,\omega_0\,Q\,
+    \exp\left(-\frac{\omega_0\,\tau}{2\,Q}\right)\,
+    \cos\left(\omega_0\,\tau\right) \quad.
 \end{eqnarray}}
@@
 \end{itemize}
-
-% This gives a kernel with dependence
-% \begin{eqnarray}\eqlabel{sho_kernel}
-% %k(\tau) &=& P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \left[\cos{\lambda}+\frac{1}{\left(4Q^2-1\right)^{1/2}}
-% k(\tau) &=&  P_0  \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \cos{\left(\lambda-\phi\right)},\\
-% \lambda &=& \left(1-\frac{1}{4Q^2}\right)^{1/2}\omega_0\tau,\\
-% \tan{\phi} &=& \left(4Q^2-1\right)^{-1/2}.
-% \end{eqnarray}
-% The power spectrum can be normalized by $P_0\omega_0 Q$ in amplitude and $\omega_0$ in
-% frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
-% in Figure \ref{sho_psd} for various values of $Q$.
-% Note that the oscillation frequency that appears in the kernel, $d_j$, differs
-% from that of the simple-harmonic oscillator, $\omega_0$, due to the damping
-% frequency, $c_j$.  This power spectrum has
-% a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)} = d_j$, which is
-% greater than zero for $Q > 1/\sqrt{2}$.
-
-% \begin{figure}[!htbp]
-% \begin{center}
-% \includegraphics[width=0.75\textwidth]{./approximate_kernels/sho.png}
-% \caption{Normalized power spectra of the stochastically-driven, damped simple
-% harmonic oscillator. The dotted black line shows the peak frequency, $d_j = \omega_0 (1-1/(2Q)^2)^{1/2}$.}
-% \label{sho_psd}
-% \end{center}
-% \end{figure}
-
-% This power spectrum has several limits of interest:
-% \begin{itemize}
-% \item $Q=1/2$:  This corresponds to a critically damped harmonic oscillator, which results in a power
-% spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
-% This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
-% autocovariance function:
-% \begin{equation}
-% k(\tau) = a_j e^{-c_j \tau} \left[\cos{(d_j \tau)} + \frac{c_j}{d_j}\sin{(d_j \tau)}\right],
-% % Check to see that signs match in this equation.
-% \end{equation}
-% in the limit that $d_j \rightarrow \infty$, corresponding to $\phi=\pi/2$, for which the second
-% term has a coefficient that diverges, but with large values of $d_j$, this may approximate the
-% Mat\'ern kernel well.  Note that this power spectrum peaks at $\omega=0$.
-% \item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
-% form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
-% dependence
-% \begin{equation}
-% k(\tau) = a_j e^{-c_j\tau} \left[\cos{(d_j\tau)}+\sin{(d_j\tau)}\right],
-% \end{equation}
-% corresponding to $\phi = \pi/4$.
-% Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
-% peak at zero frequency; such a low quality oscillator doesn't even display quasi-periodic behavior.
-% \item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.
-% The kernel approaches
-% \begin{equation}
-% k(\tau) = P_0 \omega_0 Q e^{-\omega_0\tau/(2Q)} \cos{\left(\omega_0\tau\right)},
-% \end{equation}
-% for $Q \gg 1$.  The power spectrum approaches
-% \begin{equation}
-% P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
-% \end{equation}
-% where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
-% a width of $1/(Qx_{max})$ and a peak height $\propto Q^2$.
-% \end{itemize}
-
-The stochastically-driven, damped simple-harmonic oscillator can then describe
-a wide range of intrinsic stellar variability, with low $Q \approx 1$
-representing granulation noise, and high $Q \gg 1$ representing asteroseismic
-variability.
-Taking the sum over oscillators with various $Q, S_0,$ and $\omega_0$ can give
-an accurate accounting of the power spectrum of stellar variability with $Q
-\ge \frac{1}{2}$ and $S_0 >0$.
-As this kernel is exactly described by the exponential kernel, this functional
-form is tractable to solving in $\mathcal{O}(N)$ operations, as described
-next.
+\Figure{sho} shows a plot of the PSD for these limits and several other values
+of $Q$.
+This figure demonstrates that for $Q \le 1/2$, the model has no oscillatory
+behavior and that for large $Q$, the shape of the PSD near the peak frequency
+approaches a Lorentzian.
+
+These special cases demonstrate that the stochastically-driven simple harmonic
+oscillator provides a physically motivated model that is flexible enough to
+describe a wide range of stellar variations.
+Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
+model for asteroseismic oscillations.
+In practice, we will take a sum over oscillators with different values of $Q$,
+$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
+stellar time series.
+Since this kernel is exactly described by the exponential kernel, the
+likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
+measurements in $\mathcal{O}(N)$ operations using the \genrp\ method described
+in the previous section.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/sho.pdf}
+\caption{The power spectrum of a stochastically-driven simple harmonic
+    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
+    factor $Q$.
+    For comparison, the dashed line shows the Lorentzian function from
+    \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
+    $S(d_j) = Q^2 = 100$.
+    \figurelabel{sho}}
+\end{center}
+\end{figure}
 

### 2016-12-29 paper/figures/plot_setup.py

@@ -1,0 +1,31 @@
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+from cycler import cycler
+from matplotlib import rcParams
+from savefig import monkey_patch
+
+__all__ = ["setup", "COLORS", "SQUARE_FIGSIZE"]
+
+def setup():
+    monkey_patch()
+
+    rcParams["font.size"] = 16
+    rcParams["font.family"] = "sans-serif"
+    rcParams["font.sans-serif"] = ["Computer Modern Sans"]
+    rcParams["text.usetex"] = True
+    rcParams["text.latex.preamble"] = r"\usepackage{cmbright}"
+    rcParams["figure.autolayout"] = True
+    rcParams["axes.prop_cycle"] = cycler("color", (
+        "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
+        "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
+    ))  # d3.js color cycle
+
+COLORS = dict(
+    DATA="k",
+    MODEL_1="#1f77b4",
+    MODEL_2="#ff7f0e",
+)
+
+SQUARE_FIGSIZE = (4, 4)

### 2016-12-29 paper/figures/sho.py

@@ -1,0 +1,36 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as plt
+from plot_setup import setup, SQUARE_FIGSIZE
+
+setup()
+
+def sho_psd(Q, x):
+    x2 = x*x
+    return 1.0 / ((x2 - 1)**2 + x2 / Q**2)
+
+def lorentz_psd(Q, x):
+    return Q**2 * (1.0 / ((x - 1)**2 * (2*Q)**2 + 1) +
+                   1.0 / ((x + 1)**2 * (2*Q)**2 + 1))
+
+fig, ax = plt.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+x = 10**np.linspace(-1.1, 1.1, 5000)
+
+for Q_name, Q in [("1/2", 0.5), ("1/\\sqrt{2}", 1./np.sqrt(2)),
+                  ("2", 2.0), ("10", 10.0)]:
+    ax.plot(x, sho_psd(Q, x), label="$Q = {0}$".format(Q_name), lw=1.5)
+
+ax.plot(x, lorentz_psd(10.0, x), "--k")
+
+ax.set_xscale("log")
+ax.set_yscale("log")
+ax.set_xlim(x.min(), x.max())
+ax.set_ylim(2e-4, 200.0)
+ax.legend(loc=3, fontsize=11)
+ax.set_xlabel("$\omega/\omega_0$")
+ax.set_ylabel("$S(\omega) / S(0)$")
+fig.savefig("sho.pdf", bbox_inches="tight")

### 2016-12-30 paper/ms.tex
typo
another typo
acks
Small corrections; p -> J; p_0 -> J_0

--- 2016-12-28 paper/ms.tex
@@
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
-    \sum_{j=1}^N &&\left[
+    \sum_{j=1}^J &&\left[
     \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
@@
 \begin{eqnarray}\eqlabel{genrp-psd}
-S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
+S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
 \frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
@@
 This model can be interpreted intuitively as a quasiperiodic oscillator with
-amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
+amplitude $A_j = a_j$, quality factor $Q_j = c_j$ \agoltodo{quality factor should
+be dimensionless}, and period $P_j = 2\,\pi /
 d_j$.
@@
 
-In the case of $p$ damped sinusoids we can check for negative values of the PSD by solving for the
+In the case of $J$ damped sinusoids we can check for negative values of the PSD by solving for the
 roots of the power spectrum, abbreviating with $z = \omega^2$:
 \begin{equation}
-P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
+P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
 \end{equation}
@@
 \begin{equation}
-P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
-\end{equation}
-which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
+P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
+\end{equation}
+which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
 which may be solved exactly for the roots.
@@
 %\begin{equation}
-%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
+%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^J a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
 %\end{equation}
@@
 while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
-spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,2009A&A...495..979M}.
+spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,Michel:2009}.
 \item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
@@
 %    \begin{array}{cc}
-%       w_i + \sum_{l=1}^p \alpha_l & j = i \\
-%       \sum_{l=1}^p \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
+%       w_i + \sum_{l=1}^J \alpha_l & j = i \\
+%       \sum_{l=1}^J \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
 %    \end{array}
@@
 %\end{equation}
-%Since this equation is the sum of $p$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
+%Since this equation is the sum of $J$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
 %
@@
 %that the structure of the sum of exponential kernels may be exploited
-%to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
+%to obtain the predicted mean in of order $J(M+N)+J^2N$ multiplications,
 %i.e.\ a linear scaling with the total number of data points, which can
@@
 %With these updates, we can then use equation \ref{prediction} to compute the expected
-%mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
-%$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
+%mean at $t_{*,k}$ with an additional $J$ operations.  Since there are $N$ values of
+%$t_k$, the number of operations needed to update $\epsilon$ is of order $J(N+M)$, and
 %so this procedure avoids a heavy computational burden.
 %
-%The variance may be handled in a similar manner, but with $p^2$ coefficients
+%The variance may be handled in a similar manner, but with $J^2$ coefficients
 %to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
@@
 mixes the components of the original equations which are shifted by one column from one
-another.  We will enumerate the number of $d_j=0$ cases as $p_0$, while the total number
-of $c_j$'s is still $p$.  This gives a total number of equations to be solved with the
-extended matrix of $N_{ex}= (4(p-p_0)+2p_0+1)\times(N-1)+1$, where $N$ is the number of data
+another.  We will enumerate the number of $d_j=0$ cases as $J_0$, while the total number
+of $c_j$'s is still $J$.  This gives a total number of equations to be solved with the
+extended matrix of $N_{ex}= (4(J-J_0)+2J_0+1)\times(N-1)+1$, where $N$ is the number of data
 points.
@@
 is irrelevant).  The length of the vectors $l_k$ and $r_k$
-is the number of kernel components, $p$;  we denote the individual elements of these vectors
-as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
+is the number of kernel components, $J$;  we denote the individual elements of these vectors
+as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
 to be the kernels with $d_j=0$, and for these we do not include the imaginary component
 equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
-components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_p+id_p) t_{k,k+1})}\right]^T
+components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_J+id_J) t_{k,k+1})}\right]^T
  = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
@@
 % See 9/2/2016 notes
-\sum_{j=1}^p \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^p \left[
+\sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
 \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{h_k}{2},
 \end{equation}
-for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
+for $j \in \{1,...,J\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
 in \citet{Ambikasaran:2015}.
@@
 \end{eqnarray}
-with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
+with $j \in \{1,...,J\}$ and $k \in \{1,...,N-1\}$
 \citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.
@@
 \end{align}
-with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
+with $j \in \{1,...,J\}$ and $k \in \{2,...,N\}$.  Note that for the $J_0$ real components
 we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
@@
 for each damped sinusoid with non-zero $\omega$, and due to having a slightly larger bandwidth due to the
-cross terms between the real and imaginary variables.  The off-diagonal components are $p_0+2(p-p_0)+2$ above
-and below the diagonal, for a total bandwidth of $2p_0+4(p-p_0)+5$. In the case in which $p_0=p$, so
-that $d_j=0$, the off-diagonal components revert to $p+1$
-above and below the diagonal, giving a bandwidth of $2p+3$ (the same as the GRP algorithm).
+cross terms between the real and imaginary variables.  The off-diagonal components are $J_0+2(J-J_0)+2$ above
+and below the diagonal, for a total bandwidth of $2J_0+4(J-J_0)+5$. In the case in which $J_0=J$, so
+that $d_j=0$, the off-diagonal components revert to $J+1$
+above and below the diagonal, giving a bandwidth of $2J+3$ (the same as the GRP algorithm).
 
@@
 terms first, followed by the complex variables, yielding the vector
-$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
-x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [h_1/2, \{0; j=1,...,p_0+2(p-p_0)\}, h_2/2, ..., h_N/2]$.
+$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,J_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=J_0+1,...,J\},
+x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [h_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, h_2/2, ..., h_N/2]$.
 The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
-\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2p_0+4(p-p_0)+5) \times N_{ex}$.
+\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
 The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
@@
 matrix for two damped sinusoids with non-zero $\omega$ values.  This is analagous to the real case with
-$p=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
+$J=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
 that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran:2015}.
@@
 \end{center}
-\caption{Pictorial description of the extended sparse matrix where $N=10$, $p_0=0$, and $p=2$, following
+\caption{Pictorial description of the extended sparse matrix where $N=10$, $J_0=0$, and $J=2$, following
 the example shown in \citet{Ambikasaran:2015}.}

### 2016-12-30 paper/simulate_GP/cholesky.jl

@@ -1,8 +1,9 @@
 alpha = rand()+im*rand()
 beta = rand()+im*rand()
+#beta = 0.0+im*rand()
 
 # Compute covariance matrix:
-nt = 11
+nt = 5
 t = linspace(0,nt-1,nt)
 
 cov = zeros(Complex{Float64},nt,nt)
@@ -29,3 +30,39 @@ ccov=chol(cov)
 #end
 #
 #covinvchol=chol(covinv)
+
+# Check that cholesky decomposition is correct:
+*(ctranspose(ccov),ccov)-cov
+
+# Now try constructing:
+a=zeros(Complex64,nt)
+b=zeros(Complex64,nt)
+v=zeros(Complex64,nt)
+matrix = zeros(Complex64,nt,nt)
+for i=1:nt
+  a[i]=sqrt(alpha)*exp(beta*t[i])
+  b[i]=sqrt(alpha)*exp(-beta*t[i])
+  v[i]=b[i]/a[i]
+end
+d = Diagonal(a)
+l = zeros(Complex64,nt,nt)
+lam=zeros(Complex64,nt,nt)
+for i=1:nt
+  for j=1:i
+    l[i,j]=1.0+im*0.0
+  end
+  if i < nt
+    lam[i,i] = sqrt(v[i]-v[i+1])
+  else
+    lam[i,i] = sqrt(v[i])
+  end
+  for j=1:nt
+    if j >= i
+      matrix[i,j] = a[i]*b[j]
+    else
+      matrix[i,j] = a[j]*b[i]
+    end
+  end
+end
+cholm = *(d,*(l,lam))
+

### 2016-12-30 paper/ms.tex
fixing merge conflict

@@
 with the corresponding covariance function
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{approx-matern}
 k(\tau) &=& \lim_{f \to 0}\,
-    \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
+    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
     \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
-&=& \frac{1}{2}\,S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
+&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
 \end{eqnarray}
@@
 \citep{Rasmussen:2006}.
+This suggests that the Mat\'ern-3/2 covariance can be well approximated using
+the \genrp\ framework with a small value of $f$ in \eq{approx-matern} but we
+caution that this could also lead to numerical issues with the solver.
 }
 
-{\item Finally, in the limit $Q \to \infty$, \eq{sho-psd} becomes
-\begin{eqnarray}
-S(\omega) =
-\sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2-1\right]^2} \quad.
+{\item Finally, in the limit of large $Q$, the model approaches a high
+    quality oscillation with frequency $\omega_0$ and covariance function
+\begin{eqnarray}
+k(\tau) \approx
+    S_0\,\omega_0\,Q\,
+    \exp\left(-\frac{\omega_0\,\tau}{2\,Q}\right)\,
+    \cos\left(\omega_0\,\tau\right) \quad.
 \end{eqnarray}}
@@
 It is a pleasure to thank
-...
+Sivaram Ambikasaran,
+Ruth Angus,
+Will Farr,
+and
+David W.\ Hogg
 for helpful contributions to the ideas and code presented here.

### 2016-12-31 paper/ms.tex
fixing #36

@@
 This model can be interpreted intuitively as a quasiperiodic oscillator with
-amplitude $A_j = a_j$, quality factor $Q_j = c_j$ \agoltodo{quality factor should
-be dimensionless}, and period $P_j = 2\,\pi /
-d_j$.
+amplitude $A_j = a_j$, quality factor $Q_j = d_j\,(2\,c_j)^{-1}$, and period
+$P_j = 2\,\pi\,{d_j}^{-1}$.
 

### 2016-12-31 paper/ms.tex
Dirac -> Kronecker
Added ACF figure for SHO section.

@@
 where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
-$\delta_{nm}$ is the Dirac delta, and $\bvec{\alpha} = (a,\,c)$.
+$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
 The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
@@
 \begin{center}
-\includegraphics[width=0.75\textwidth]{figures/sho.pdf}
-\caption{The power spectrum of a stochastically-driven simple harmonic
+\includegraphics[width=0.5\textwidth]{figures/sho.pdf}
+\includegraphics[width=0.5\textwidth]{figures/acf_sho.pdf}
+\caption{(top) The power spectrum of a stochastically-driven simple harmonic
     oscillator (\eqalt{sho-psd}) plotted for several values of the quality
@@
     \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
-    $S(d_j) = Q^2 = 100$.
+    $S(d_j) = Q^2 = 100$. (bottom) The corresponding autocorrelation functions.
     \figurelabel{sho}}

### 2016-12-31 paper/ms.tex
merging

@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{figures/sho.pdf}
-\includegraphics[width=0.5\textwidth]{figures/acf_sho.pdf}
-\caption{(top) The power spectrum of a stochastically-driven simple harmonic
+\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
+\caption{(left) The power spectrum of a stochastically-driven simple harmonic
     oscillator (\eqalt{sho-psd}) plotted for several values of the quality
@@
     \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
-    $S(d_j) = Q^2 = 100$. (bottom) The corresponding autocorrelation functions.
+    $S(d_j) = Q^2 = 100$.
+    (right) The corresponding autocorrelation functions with the same colors.
     \figurelabel{sho}}

### 2017-01-04 paper/ms.tex

@@
 \end{eqnarray}
+
+These identities yield a kernel of the form
+\begin{equation}
+k_{SHO}(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
+\cosh{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
+2 (1+\omega_0 \tau), & Q = 1/2\\
+\cos{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
+\end{cases}
+\end{equation}
+where $\eta = \vert 1-(4Q^2)^{-1}\vert$.
 It is interesting to note that, because of the damping, the characteristic
-oscillation frequency in this model $d_j$, for any finite quality factor $Q$,
+oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
 is not equal to the frequency of the undamped oscillator $\omega_0$.

### 2017-01-05 python/genrp/kernels.py
reverting S_0 definition

@@ -384,58 +384,56 @@ class SHOTerm(Kernel):
         return "SHOTerm({0.log_S0}, {0.log_Q}, {0.log_w0})".format(self)
 
     @property
-    def p_real(self):
-        Q = math.exp(self.log_Q)
-        return 2 if Q < 0.5 else 0
+    def p_real(self, log_half=math.log(0.5)):
+        return 2 if self.log_Q < log_half else 0
 
     @property
-    def p_complex(self):
-        Q = math.exp(self.log_Q)
-        return 0 if Q < 0.5 else 1
+    def p_complex(self, log_half=math.log(0.5)):
+        return 0 if self.log_Q < log_half else 1
 
     @property
-    def alpha_real(self):
-        Q = math.exp(self.log_Q)
-        if Q >= 0.5:
+    def alpha_real(self, log_half=math.log(0.5)):
+        if self.log_Q >= log_half:
             return np.empty(0)
+        Q = math.exp(self.log_Q)
         f = 1.0 / math.sqrt(1.0 - 4.0 * Q**2)
         return 0.5*math.exp(self.log_S0+self.log_w0)*Q*np.array([1.0+f, 1.0-f])
 
     @property
-    def beta_real(self):
-        Q = math.exp(self.log_Q)
-        if Q >= 0.5:
+    def beta_real(self, log_half=math.log(0.5)):
+        if self.log_Q >= log_half:
             return np.empty(0)
+        Q = math.exp(self.log_Q)
         f = math.sqrt(1.0 - 4.0 * Q**2)
         return 0.5*math.exp(self.log_w0)/Q*np.array([1.0-f, 1.0+f])
 
     @property
-    def alpha_complex_real(self):
-        Q = math.exp(self.log_Q)
-        if Q < 0.5:
+    def alpha_complex_real(self, log_half=math.log(0.5)):
+        if self.log_Q < log_half:
             return np.empty(0)
+        Q = math.exp(self.log_Q)
         return np.array([math.exp(self.log_S0 + self.log_w0) * Q])
 
     @property
-    def alpha_complex_imag(self):
-        Q = math.exp(self.log_Q)
-        if Q < 0.5:
+    def alpha_complex_imag(self, log_half=math.log(0.5)):
+        if self.log_Q < log_half:
             return np.empty(0)
+        Q = math.exp(self.log_Q)
         return np.array([math.exp(self.log_S0+self.log_w0)*Q /
[…]

### 2017-01-06 paper/ms.tex
Corrected definition of \eta.
figure path
Fixed paths to figure in ms.tex
tiny typographical change
proposal: re-define S_0
comment on definition of S_0
reverting S_0 definition
Started writing discussion section.

@@
 \begin{equation}
-k_{SHO}(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
-\cosh{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
+k(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
+\cosh{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
 2 (1+\omega_0 \tau), & Q = 1/2\\
-\cos{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
+\cos{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
 \end{cases}
 \end{equation}
-where $\eta = \vert 1-(4Q^2)^{-1}\vert$.
+where $\eta = \vert 1-(4Q^2)^{-1}\vert^{1/2}$.
 It is interesting to note that, because of the damping, the characteristic
@@
 A model like this has been used to model the granulation is Solar data
-\citep{Michel:2009}}
+\citep{Michel:2009}}, which has a kernel of
+\begin{equation}
+k(\tau) = S_0 \omega_0e^{-\frac{1}{\sqrt{2}}\omega_0\tau} \cos{\left(\frac{\omega_0\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
+\end{equation}
 
@@
     For comparison, the dashed line shows the Lorentzian function from
-    \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
-    $S(d_j) = Q^2 = 100$.
+    \eq{lorentz-psd} with $\omega_0/c_j = 2\,Q = 20$ and normalized so that
+    $S(d_j)/S(0) = Q^2 = 100$.
     (right) The corresponding autocorrelation functions with the same colors.
@@
 Q: How do we determine how many kernels to use?
+
+\section{Discussion}
+
+There are two primary applications we imagine for the \genrp formalism: 1). modeling
+astrophysical variability to infer the properties of the object that is varying; 2).
+account for astrophysical variability as a source of noise when trying to detect 
+additional phenomena; 3). interpolate or extrapolate variations to future or missing
+times.  Some examples of the first are determining the characteristic
+variability timescale of quasars, measuring the asteroseismic variations of stars,
+or detection of quasi-periodic variability in a high-energy source.  Examples of the
+second are detecting transiting exoplanets and correcting for stellar activity in
+radial velocity measurements. Examples of the third are measuring time delays in 
+multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.
+
+Our background is in studying transiting exoplanets, which has only recently begun
+to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
+One early successful attempt at accounting for correlated noise was the wavelet approach
+of Carter \& Winn (2010).  This approach allowed for a power-law spectrum for the
+noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
+physically-accurate description of stellar variability.  Their approach runs in
+$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
+particular type of noise overpredicts the power spectrum at low frequency compared
+to the granulation power spectrum.  It also cannot describe quasi-periodic noise.
 
@@
 
+% Reprint Ambikasaran equations (59)-(61)
+
 For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
@@
 % See 9/2/2016 notes
+% Use n rather than k for {1...N} - use g, h & u, v?
+% \gamma & \phi - real & imaginary 
 \sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
-\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{h_k}{2},
+\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{y_k}{2},
 \end{equation}
@@
 \begin{center}
-\includegraphics[scale=0.175]{./twoterm.pdf}
+\includegraphics[scale=0.175]{figures/twoterm.pdf}
 \hspace{0.2in}
-\includegraphics[scale=0.85]{./colorcode.pdf}
+\includegraphics[scale=0.85]{figures/colorcode.pdf}
 \end{center}

### 2017-01-06 paper/ms.tex

@@
 \end{equation}
-where $S_0$ is a normalization constant.
+where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
+= \sqrt{2/\pi}\,S_0$.
 The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
@@
 \begin{equation}
-k(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
-\cosh{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
-2 (1+\omega_0 \tau), & Q = 1/2\\
-\cos{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
+k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
+\begin{cases}
+    \cosh{(\eta\,\omega_0\,\tau)} +
+        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
+    2\,(1+\omega_0\,\tau), & Q = 1/2\\
+    \cos{(\eta\,\omega_0\,\tau)} +
+        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
 \end{cases}
 \end{equation}
-where $\eta = \vert 1-(4Q^2)^{-1}\vert^{1/2}$.
+where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
 It is interesting to note that, because of the damping, the characteristic
@@
 \begin{equation}
-k(\tau) = S_0 \omega_0e^{-\frac{1}{\sqrt{2}}\omega_0\tau} \cos{\left(\frac{\omega_0\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
+k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
+    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
 \end{equation}
@@
     For comparison, the dashed line shows the Lorentzian function from
-    \eq{lorentz-psd} with $\omega_0/c_j = 2\,Q = 20$ and normalized so that
-    $S(d_j)/S(0) = Q^2 = 100$.
+    \eq{lorentz-psd} with $c_j = \omega_0/2\,Q = 1/20$ and normalized so that
+    $S(d_j)/S(0) = 100$.
     (right) The corresponding autocorrelation functions with the same colors.

### 2017-01-06 paper/ms.tex
Added to discussion of applications.

@@
 
-There are two primary applications we imagine for the \genrp formalism: 1). modeling
-astrophysical variability to infer the properties of the object that is varying; 2).
-account for astrophysical variability as a source of noise when trying to detect 
-additional phenomena; 3). interpolate or extrapolate variations to future or missing
+There are three primary applications we envision for the \genrp formalism: 1). modeling
+the variability of an astrophysical system to infer it properties; 2).
+accounting for astrophysical variability as a source of noise when trying to detect 
+additional phenomena; 3). interpolating or extrapolat variability to future or missing
 times.  Some examples of the first are determining the characteristic
@@
 
-Our background is in studying transiting exoplanets, which has only recently begun
+\subsection{Application to stellar variability, transiting exoplanets}
+
+Our background is in studying transiting exoplanets, a field which has only recently begun
 to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
-One early successful attempt at accounting for correlated noise was the wavelet approach
-of Carter \& Winn (2010).  This approach allowed for a power-law spectrum for the
+One promising early attempt at accounting for correlated noise was the wavelet approach
+of Carter \& Winn (2010).  Their technique allowed for a power-law spectrum for the
 noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
-physically-accurate description of stellar variability.  Their approach runs in
+physically-accurate, description of stellar variability.  Their approach runs in
 $\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
@@
 to the granulation power spectrum.  It also cannot describe quasi-periodic noise.
+
+Further progress was made by parameterizing the covariance matrix with simple,
+analytic functions that describe the autocorrelation function.  The parameters
+of these functions can then be optimized to best match the observed variability
+pattern of the residuals (after subtracting a transit model).  The disadvantage of
+this approach is that the functions used are chosen somewhat arbitrarily (e.g.\
+the exponential-squared function), and the Cholesky decomposition of the covariance
+matrix for computing the likelihood takes $\cal{O}(N^3)$ operations which prohibits 
+application of this technique to large datasets.
+
+Given the drawbacks of these approaches, the \genrp formalism allows both a fast
+computation of the likelihood in $\cal{O}(N)$ time, as well as a functional form
+that accurately describes stellar variability due to the relation to the simple
+harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
+oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
+are obtained, the effects of stellar variability will more dramatically impact the
+correct inference of planetary transit parameters, and so we expect that \genrp
+will be important for transit timing, transit spectroscopy, Doppler beaming,
+phase functions, and more.
 

### 2017-01-06 paper/ms.tex
Modified notation in the appendix.

@@
 
-There are three primary applications we envision for the \genrp formalism: 1). modeling
-the variability of an astrophysical system to infer it properties; 2).
+There are three primary applications we envision for the \genrp\ formalism: 1).\ modeling
+the variability of an astrophysical system to infer it properties; 2).\
 accounting for astrophysical variability as a source of noise when trying to detect 
-additional phenomena; 3). interpolating or extrapolat variability to future or missing
+additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
 times.  Some examples of the first are determining the characteristic
-variability timescale of quasars, measuring the asteroseismic variations of stars,
+variability timescale of a quasar, measuring the asteroseismic variation of a star,
 or detection of quasi-periodic variability in a high-energy source.  Examples of the
@@
 physically-accurate, description of stellar variability.  Their approach runs in
-$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
+${\cal O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
 particular type of noise overpredicts the power spectrum at low frequency compared
@@
 the exponential-squared function), and the Cholesky decomposition of the covariance
-matrix for computing the likelihood takes $\cal{O}(N^3)$ operations which prohibits 
+matrix for computing the likelihood takes ${\cal O}(N^3)$ operations which prohibits 
 application of this technique to large datasets.
 
-Given the drawbacks of these approaches, the \genrp formalism allows both a fast
-computation of the likelihood in $\cal{O}(N)$ time, as well as a functional form
+Given the drawbacks of these approaches, the \genrp\ formalism allows both a fast
+computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
 that accurately describes stellar variability due to the relation to the simple
@@
 are obtained, the effects of stellar variability will more dramatically impact the
-correct inference of planetary transit parameters, and so we expect that \genrp
+correct inference of planetary transit parameters, and so we expect that \genrp\
 will be important for transit timing, transit spectroscopy, Doppler beaming,
@@
 For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
-auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
-r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
+auxiliary vectors $l_n = g_n - i h_n$, with $n \in \{1,2,...,N\}$ and $l_1 = 0$; $r_n =
+u_n - i v_n$, with $n \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
 imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
@@
 set of equations, and since the values of these variables are discarded in the end, their sign
-is irrelevant).  The length of the vectors $l_k$ and $r_k$
+is irrelevant).  The length of the vectors $l_n$ and $r_n$
 is the number of kernel components, $J$;  we denote the individual elements of these vectors
-as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
+as $g_{n,j}+ih_{n,j}$ and $u_{n,j}+iv_{n,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
 to be the kernels with $d_j=0$, and for these we do not include the imaginary component
-equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
-components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_J+id_J) t_{k,k+1})}\right]^T
- = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
-\cos{(d_j t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-c_j t_{k,k+1}}\sin{(d_j t_{k,k+1})}$.
+equations, which are unnecessary and only add extra computation time.  We also define vectors $\phi$ and $\psi$ with
+complex components as well:  $\gamma_n = \phi_n + i\psi_n = \left[\exp{(-(c_1+id_1) t_{n,n+1})} ... \exp{(-(c_J+id_J) t_{n,n+1})}\right]^T
+$.  The individual elements are given by $\phi_{n,j} = e^{-c_j t_{n,n+1}}
+\cos{(d_j t_{n,n+1})}$ and $\psi_{n,j} = -e^{-c_j t_{n,n+1}}\sin{(d_j t_{n,n+1})}$.
 Equation (61) becomes:
@@
 % \gamma & \phi - real & imaginary 
-\sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
-\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{y_k}{2},
-\end{equation}
-for $j \in \{1,...,J\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
+\sum_{j=1}^J \left(\frac{a_j}{2} g_{n,j} +\frac{b_j}{2} h_{n,j}\right)+ \frac{d}{2} x_n + \sum_{j=1}^J \left[
+\phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j}\right] = \frac{y_n}{2},
+\end{equation}
+for $j \in \{1,...,J\}$ and $n \in \{1,...,N\}$, where $y_j$ is the $j$th data point.  There is no imaginary component to equation (61)
 in \citet{Ambikasaran:2015}.
@@
 \begin{eqnarray} \label{Amb60}
-\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
-\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
-\end{eqnarray}
-with $j \in \{1,...,J\}$ and $k \in \{1,...,N-1\}$
+\phi_{n,j} g_{n,j} +\psi_{n,j} h_{n,j} + \phi_{n,j} x_n - g_{n+1,j} &=& 0,\\
+\psi_{n,j} g_{n,j} -\phi_{n,j} h_{n,j} + \psi_{n,j} x_n + h_{n+1,j} &=& 0,
+\end{eqnarray}
+with $j \in \{1,...,J\}$ and $n \in \{1,...,N-1\}$
 \citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.
@@
 Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
-\begin{align} \label{Amb59}
--&r_{k,j}^R + \frac{a_j}{2} x_k + \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
- &r_{k,j}^I + \frac{b_j}{2} x_k + \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
-\end{align}
-with $j \in \{1,...,J\}$ and $k \in \{2,...,N\}$.  Note that for the $J_0$ real components
-we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
+\begin{eqnarray} \label{Amb59}
+-u_{n,j} + \frac{a_j}{2} x_n + \phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j} &=&0,\\
+ v_{n,j} + \frac{b_j}{2} x_n + \psi_{n,j} u_{n+1,j} - \phi_{n,j} v_{n+1,j} &=&0,
+\end{eqnarray}
+with $j \in \{1,...,J\}$ and $n \in \{2,...,N\}$.  Note that for the $J_0$ real components
+we can drop the variables $v_{n,j}$ and $h_{n,j}$ as these are zero, and we can drop the
 imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
@@
 terms first, followed by the complex variables, yielding the vector
-$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,J_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=J_0+1,...,J\},
-x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [h_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, h_2/2, ..., h_N/2]$.
+$x_{ex} = [x_1, \{u_{2,j},g_{2,j};j=1,...,J_0\}, \{u_{2,j},v_{2,j},g_{2,j},h_{2,j};j=J_0+1,...,J\},
+x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [y_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, y_2/2, ..., y_N/2]$.
 The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},

### 2017-01-08 paper/figures/rotation.py
include constraints on rotation period

@@ -125,7 +125,7 @@ while np.any(m):
 sampler = emcee.EnsembleSampler(nwalkers, ndim, log_prob)
 pos, _, _ = sampler.run_mcmc(pos, 250)
 sampler.reset()
-pos, _, _ = sampler.run_mcmc(pos, 500)
+pos, _, _ = sampler.run_mcmc(pos, 1000)
 
 # Compute the model predictions
 gp.set_parameter_vector(ml_params)
@@ -176,7 +176,7 @@ ax.set_yscale("log")
 ax.set_xscale("log")
 ax.set_xlim(f[0], f[-1])
 ax.set_ylim(2e-4, 2e-1)
-ax.set_xlabel("$\omega\,[\mathrm{day}^{-1}]$")
+ax.set_xlabel("$\omega\,[\mathrm{days}^{-1}]$")
 ax.set_ylabel("$S(\omega)$")
 ax.annotate("power spectrum", xy=(1, 1), xycoords="axes fraction",
             ha="right", va="top", xytext=(-5, -5), textcoords="offset points",
@@ -190,7 +190,7 @@ ax.plot(tau, q[1], color=color, lw=1.5)
 ax.plot(tau, acf, "--k", lw=1.5)
 ax.set_xlim(tau[0], tau[-1])
 ax.set_ylim(0, 0.135)
-ax.set_xlabel(r"$\tau\,[\mathrm{day}]$")
+ax.set_xlabel(r"$\tau\,[\mathrm{days}]$")
 ax.set_ylabel(r"$k(\tau)$")
 ax.annotate("covariance function", xy=(1, 1), xycoords="axes fraction",
             ha="right", va="top", xytext=(-5, -5), textcoords="offset points",
@@ -209,3 +209,19 @@ ax.annotate("simulated light curve", xy=(1, 1), xycoords="axes fraction",
             fontsize=12)
 
 fig.savefig("rotation.pdf", bbox_inches="tight", dpi=300)
+plt.close(fig)
+
+# Plot the period constraint
+period_samps = np.exp(sampler.flatchain[:, 2])
+fig, ax = plt.subplots(1, 1, figsize=SQUARE_FIGSIZE)
+ax.hist(period_samps, 40, histtype="step", color=color)
+ax.yaxis.set_major_locator(plt.NullLocator())
+ax.set_xlim(2.9, 4.3)
+ax.set_xlabel("rotation period [days]")
+fig.savefig("rotation-period.pdf", bbox_inches="tight", dpi=300)
+
[…]

### 2017-01-08 paper/ms.tex
words about asterosesmology

@@
 {\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{granulation-psd}
 S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
@@
 
-\section{Examples}
-
-\subsection{Simulated data}
+\section{Examples with simulated data}
 
@@
 
-\subsection{Real data}
+\section{Examples with real data}
+
+\subsection{Asteroseismic oscillations}
+
+The asteroseismic oscillations of thousands of stars were measured using light
+curves from the \kepler\ mission (\dfmtodo{cite}) and asteroseismology is a
+key science driver for many of the upcoming large scale photometric surveys
+(\dfmtodo{cite}).
+Most asteroseismic analyses have been limited to relatively high
+signal-to-noise oscillations because the standard methods based on statistics
+of the empirical periodogram of the data cannot be used to formally propagate
+the measurement uncertainties to the constraints on physical parameters
+(\dfmtodo{CITE}) and more sophisticated methods are computationally expensive
+and they scale poorly to the current state-of-the-art datasets (\dfmtodo{CITE
+Brewer etc.}).
+
+\genrp\ alleviates these problems by providing a physically motivated
+probabilistic model that can be evaluated efficiently even for large datasets.
+In practice, we will model the star as a mixture of stochastically driven
+simple harmonic oscillators where the amplitudes and frequencies of the
+oscillations are computed using a physical model and evaluate the probability
+of the observed dataset using a Gaussian Process with a PSD given by a sum of
+terms given by \eq{sho-psd}.
+This gives us a method of computing the likelihood function for the parameters
+of the physical model in $\mathcal{O}(N)$ operations and this can be combined
+with standard non-linear optimization or posterior sampling methods to infer
+constraints on the parameters for the given dataset.
+
+To demonstrate this method, we will use a very simple heuristic model based on
+\dfmtodo{CITE} where the model PSD is given by a mixture of 6 components with
+amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
+several nuisance parameters.
+The first term is used to capture the granulation ``background'' using
+\eq{granulation-psd} with two free parameters $S_g$ and $\omega_g$.
+The remaining 5 terms are given by \eq{sho-psd} where $Q$ is a nuisance
+parameter shared between terms and the frequencies are given by
+\begin{eqnarray}
+\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
+\end{eqnarray}
+and the amplitudes are given by
+\begin{eqnarray}
+S_{0,\,j} =
+    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
+\end{eqnarray}
+where $j$ is an integer running from $-2$ to 2 and $\epsilon$, $A$, and $W$
+are shared nuisance parameters.
+This model could be easily extended to include small frequency splitting and
+$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
+like $\log g$.
+
+\dfmtodo{Add plots etc.}
 
 \begin{enumerate}
+\item Asteroseismic target
+\item Stellar rotation variability
+\item Kepler with simulated transit
 \item AGN
-\item RR Lyrae
-\item Kepler with simulated transit
-\item Asteroseismic target
-\item Stellar rotational variability
 \end{enumerate}

### 2017-01-09 cpp/include/genrp/carma.h
making solver picklable

@@ -7,6 +7,7 @@
 #include <Eigen/Dense>
 
 #include "genrp/utils.h"
+#include "genrp/exceptions.h"
 
 namespace genrp {
 namespace carma {
@@ -166,7 +167,7 @@ void predict (double yerr) {
   variance_ = yerr * yerr + tmp.real();
 
   // Check the variance value for instability.
-  if (variance_ < 0.0) throw 1;
+  if (variance_ < 0.0) throw carma_exception();
 };
 
 void update_state (double y) {

### 2017-01-09 cpp/include/genrp/exceptions.h

@@ -1,0 +1,22 @@
+#ifndef _GENRP_EXCEPTIONS_H_
+#define _GENRP_EXCEPTIONS_H_
+
+#include <exception>
+
+namespace genrp {
+
+struct carma_exception : public std::exception {
+  const char * what () const throw () {
+    return "CARMA model encountered an instability";
+  }
+};
+
+struct compute_exception : public std::exception {
+  const char * what () const throw () {
+    return "you must call 'compute' first";
+  }
+};
+
+}; // namespace genrp
+
+#endif // _GENRP_EXCEPTIONS_H_

### 2017-01-09 cpp/include/genrp/modeling.h

@@ -1,0 +1,35 @@
+#ifndef _GENRP_MODELING_H_
+#define _GENRP_MODELING_H_
+
+#include <string>
+
+namespace genrp {
+
+class Parameter {
+  Parameter () : frozen_(false) {};
+  Parameter (std::string name) : name_(name), frozen_(false) {};
+  Parameter (std::string name, double value, bool frozen = false) : name_(name), frozen_(frozen), value_(value) {};
+
+  const std::string& get_name () const { return name_; };
+  void set_name (std::string name) { name_ = name; };
+  double get_value () const { return value_; };
+  void set_value (double value) { value_ = value; };
+
+  bool is_frozen () const { return frozen_; };
+  void freeze () { frozen_ = true; };
+  void thaw () { frozen_ = false; };
+
+private:
+  std::string name_;
+  bool frozen_;
+  double value_;
+};
+
+class Kernel {
+  virtual p_real () const { return 0; };
+
+};
+
+};
+
+#endif

### 2017-01-09 cpp/include/genrp/solver/band.h

@@ -2,10 +2,12 @@
 #define _GENRP_SOLVER_BAND_H_
 
 #include <cmath>
+#include <tuple>
 #include <iostream>
 #include <Eigen/Core>
 
 #include "genrp/utils.h"
+#include "genrp/exceptions.h"
 #include "genrp/banded.h"
 #include "genrp/solver/solver.h"
 
@@ -48,7 +50,7 @@ public:
   using Solver<T>::compute;
   using Solver<T>::solve;
 
-private:
+protected:
   Eigen::Matrix<T, Eigen::Dynamic, Eigen::Dynamic> a_, al_;
   Eigen::VectorXi ipiv_;
 
@@ -235,12 +237,8 @@ int BandSolver<T>::compute (
 
 template <typename T>
 void BandSolver<T>::solve (const Eigen::MatrixXd& b, T* x) const {
-  if (b.rows() != this->n_) {
-    throw SOLVER_DIMENSION_MISMATCH;
-  }
-  if (!(this->computed_)) {
-    throw SOLVER_NOT_COMPUTED;
-  }
+  if (b.rows() != this->n_) throw std::domain_error("dimension mismatch");
+  if (!(this->computed_)) throw compute_exception();
   int nrhs = b.cols();
 
   BLOCKSIZE

### 2017-01-09 cpp/include/genrp/solver/direct.h

@@ -5,6 +5,7 @@
 #include <Eigen/Dense>
 
 #include "genrp/utils.h"
+#include "genrp/exceptions.h"
 #include "genrp/solver/solver.h"
 
 namespace genrp {
@@ -81,12 +82,8 @@ int DirectSolver<T>::compute (
 
 template <typename T>
 void DirectSolver<T>::solve (const Eigen::MatrixXd& b, T* x) const {
-  if (b.rows() != this->n_) {
-    throw SOLVER_DIMENSION_MISMATCH;
-  }
-  if (!(this->computed_)) {
-    throw SOLVER_NOT_COMPUTED;
-  }
+  if (b.rows() != this->n_) throw std::domain_error("dimension mismatch");
+  if (!(this->computed_)) throw compute_exception();
   int nrhs = b.cols();
 
   Eigen::Matrix<T, Eigen::Dynamic, Eigen::Dynamic>

### 2017-01-09 cpp/include/genrp/solver/solver.h

@@ -2,7 +2,8 @@
 #define _GENRP_SOLVER_SOLVER_H_
 
 #include <Eigen/Core>
-#include "genrp/poly.h"
+
+#include "genrp/exceptions.h"
 
 namespace genrp {
 namespace solver {
@@ -35,7 +36,11 @@ public:
 
   Eigen::Matrix<T, Eigen::Dynamic, Eigen::Dynamic> solve (const Eigen::MatrixXd& b) const;
   T dot_solve (const Eigen::VectorXd& b) const;
-  T log_determinant () const;
+  T log_determinant () const {
+    if (!(this->computed_)) throw compute_exception();
+    return log_det_;
+  };
+  bool computed () const { return computed_; };
 
   // Helpers
   int compute (
@@ -73,14 +78,10 @@ public:
 
 template <typename T>
 T Solver<T>::dot_solve (const Eigen::VectorXd& b) const {
+  if (!(this->computed_)) throw compute_exception();
   Eigen::Matrix<T, Eigen::Dynamic, 1> out(n_);
   solve(b, out.data());
   return b.transpose().cast<T>() * out;
-}
-
-template <typename T>
-T Solver<T>::log_determinant () const {
-  return log_det_;
 }
 
 template <typename T>

### 2017-01-09 python/genrp/genrp.cpp

@@ -7,6 +7,27 @@
 
 namespace py = pybind11;
 
+class PicklableBandSolver : public genrp::solver::BandSolver<double> {
+public:
+  auto serialize () const {
+    return std::make_tuple(this->computed_, this->n_, this->p_real_, this->p_complex_, this->log_det_,
+                           this->a_, this->al_, this->ipiv_);
+  };
+
+  void deserialize (bool computed, int n, int p_real, int p_complex, double log_det,
+                    Eigen::MatrixXd a, Eigen::MatrixXd al, Eigen::VectorXi ipiv) {
+    this->computed_ = computed;
+    this->n_ = n;
+    this->p_real_ = p_real;
+    this->p_complex_ = p_complex;
+    this->log_det_ = log_det;
+    this->a_ = a;
+    this->al_ = al;
+    this->ipiv_ = ipiv;
+  };
+
+};
+
 PYBIND11_PLUGIN(_genrp) {
   py::module m("_genrp", "GenRP extension");
 
@@ -77,12 +98,10 @@ PYBIND11_PLUGIN(_genrp) {
     }
   );
 
-  py::class_<genrp::solver::BandSolver<double> > solver(m, "Solver");
+  py::class_<PicklableBandSolver > solver(m, "Solver");
   solver.def(py::init<>());
 
-  solver.def("compute",
-    [](
-      genrp::solver::BandSolver<double>& solver,
+  solver.def("compute", [](PicklableBandSolver& solver,
       const Eigen::VectorXd& alpha_real,
       const Eigen::VectorXd& beta_real,
       const Eigen::VectorXd& alpha_complex_real,
@@ -90,46 +109,56 @@ PYBIND11_PLUGIN(_genrp) {
       const Eigen::VectorXd& beta_complex_real,
       const Eigen::VectorXd& beta_complex_imag,
       const Eigen::VectorXd& x,
-      const Eigen::VectorXd& diag
-    ) {
-      return solver.compute(
-        alpha_real,
-        beta_real,
-        alpha_complex_real,
-        alpha_complex_imag,
-        beta_complex_real,
-        beta_complex_imag,
-        x,
-        diag
-      );
-    }
-  );
+      const Eigen::VectorXd& diag) {
+    return solver.compute(
+      alpha_real,
+      beta_real,
+      alpha_complex_real,
+      alpha_complex_imag,
+      beta_complex_real,
+      beta_complex_imag,
+      x,
+      diag
[…]

### 2017-01-09 python/genrp/tests.py

@@ -5,11 +5,16 @@ from __future__ import division, print_function
 import pytest
 import numpy as np
 
+try:
+    import cPickle as pickle
+except ImportError:
+    import pickle
+
 from . import Solver, GP, kernels
 from ._genrp import get_kernel_value
 
 __all__ = ["test_invalid_parameters", "test_log_determinant", "test_solve",
-           "test_nyquist_singularity"]
+           "test_pickle", "test_nyquist_singularity"]
 
 
 @pytest.mark.skip(reason="solver no longer checks for ordering")
@@ -60,6 +65,7 @@ def test_solve(seed=42):
     np.random.seed(seed)
     t = np.sort(np.random.rand(500))
     diag = np.random.uniform(0.1, 0.5, len(t))
+    b = np.random.randn(len(t))
 
     alpha_real = np.array([1.3, 1.5])
     beta_real = np.array([0.5, 0.2])
@@ -69,6 +75,11 @@ def test_solve(seed=42):
     beta_complex_imag = np.array([1.0])
 
     solver = Solver()
+    with pytest.raises(RuntimeError):
+        solver.log_determinant()
+    with pytest.raises(RuntimeError):
+        solver.dot_solve(b)
+
     solver.compute(
         alpha_real, beta_real, alpha_complex_real, alpha_complex_imag,
         beta_complex_real, beta_complex_imag, t, diag
@@ -78,13 +89,45 @@ def test_solve(seed=42):
         beta_complex_real, beta_complex_imag, t[:, None] - t[None, :]
     )
     K[np.diag_indices_from(K)] += diag
-    b = np.random.randn(len(t))
     assert np.allclose(solver.solve(b).T, np.linalg.solve(K, b))
 
     b = np.random.randn(len(t), 5)
-    print(b)
-    print(solver.solve(b))
     assert np.allclose(solver.solve(b), np.linalg.solve(K, b))
+
+
+def test_pickle(seed=42):
+    np.random.seed(seed)
+    t = np.sort(np.random.rand(500))
+    diag = np.random.uniform(0.1, 0.5, len(t))
+    y = np.sin(t)
+
+    alpha_real = np.array([1.3, 1.5])
+    beta_real = np.array([0.5, 0.2])
+    alpha_complex_real = np.array([1.0])
+    alpha_complex_imag = np.array([0.1])
+    beta_complex_real = np.array([1.0])
+    beta_complex_imag = np.array([1.0])
+
+    def compare(solver1, solver2):
[…]

### 2017-01-10 paper/figures/rotation.py
more modeling interface tweaks

@@ -14,7 +14,7 @@ from plot_setup import setup, SQUARE_FIGSIZE, COLORS
 import genrp
 from genrp import terms
 
-np.random.seed(42)
+np.random.seed(123)
 setup()
 
 # Define the custom kernel
@@ -53,7 +53,13 @@ yerr *= 1e3 / mean
 
 # Set up the GP model
 kernel = RotationTerm(
-    np.log(np.var(y)), np.log(10), np.log(2.0), np.log(1.0)
+    np.log(np.var(y)), np.log(10), np.log(2.0), np.log(1.0),
+    bounds=[
+        np.log(np.var(y) * np.array([0.01, 100])),
+        np.log([np.max(np.diff(t)), (t.max() - t.min())]),
+        np.log([3*np.median(np.diff(t)), 0.5*(t.max() - t.min())]),
+        [-5.0, np.log(5.0)],
+    ]
 )
 gp = genrp.GP(kernel, mean=np.median(y))
 gp.compute(t, yerr)
@@ -65,12 +71,7 @@ def neg_log_like(params, y, gp):
 
 # Optimize with random restarts
 initial_params = gp.get_parameter_vector()
-bounds = [
-    np.log(np.var(y) * np.array([0.01, 100])),
-    np.log([np.max(np.diff(t)), (t.max() - t.min())]),
-    np.log([3*np.median(np.diff(t)), 0.5*(t.max() - t.min())]),
-    [-5.0, np.log(5.0)],
-]
+bounds = gp.get_parameter_bounds()
 best = (np.inf, initial_params)
 for i in range(10):
     p0 = np.array([np.random.uniform(*b) for b in bounds])

### 2017-01-10 python/genrp/genrp.py

@@ -103,6 +103,14 @@ class GP(Model):
             raise ValueError("dimension mismatch")
         return np.ascontiguousarray(y, dtype=float)
 
+    def log_prior(self):
+        lp = (
+            self.mean.log_prior() +
+            self.log_white_noise.log_prior() +
+            self.kernel.log_prior()
+        )
+        return lp if np.isfinite(lp) else -np.inf
+
     def log_likelihood(self, y, _const=np.log(2*np.pi)):
         y = self._process_input(y)
         resid = y - self.mean.get_value(self._t)
@@ -225,6 +233,14 @@ class GP(Model):
             self.kernel.parameter_vector
         ))
 
+    @property
+    def parameter_bounds(self):
+        return list(chain(
+            self.mean.parameter_bounds,
+            self.log_white_noise.parameter_bounds,
+            self.kernel.parameter_bounds
+        ))
+
     @parameter_vector.setter
     def parameter_vector(self, v):
         i = self.mean.full_size
@@ -256,6 +272,16 @@ class GP(Model):
 
     def thaw_parameter(self, name):
         self._apply_to_parameter("thaw_parameter", name)
+
+    def freeze_all_parameters(self):
+        self.mean.freeze_all_parameters()
+        self.log_white_noise.freeze_all_parameters()
+        self.kernel.freeze_all_parameters()
+
+    def thaw_all_parameters(self):
+        self.mean.thaw_all_parameters()
+        self.log_white_noise.thaw_all_parameters()
+        self.kernel.thaw_all_parameters()
 
     def get_parameter(self, name):
         return self._apply_to_parameter("get_parameter", name)

### 2017-01-10 python/genrp/modeling.py

@@ -11,11 +11,16 @@ class Model(object):
 
     parameter_names = tuple()
 
-    def __init__(self, *args):
+    def __init__(self, *args, **kwargs):
         self.unfrozen_mask = np.ones(self.full_size, dtype=bool)
         self.dirty = True
         if len(args):
             self.parameter_vector = args
+        self.parameter_bounds = kwargs.get(
+            "bounds", [(None, None) for _ in range(self.full_size)])
+        if len(self.parameter_bounds) != self.full_size:
+            raise ValueError("the number of bounds must equal the number of "
+                             "parameters")
 
     def get_value(self, x):
         raise NotImplementedError("overloaded by subclasses")
@@ -60,6 +65,13 @@ class Model(object):
                      for p, f in zip(self.parameter_names, self.unfrozen_mask)
                      if f)
 
+    def get_parameter_bounds(self, include_frozen=False):
+        if include_frozen:
+            return self.parameter_bounds
+        return list(p
+                    for p, f in zip(self.parameter_bounds, self.unfrozen_mask)
+                    if f)
+
     def get_parameter_vector(self, include_frozen=False):
         if include_frozen:
             return self.parameter_vector
@@ -82,6 +94,12 @@ class Model(object):
         i = self.get_parameter_names(include_frozen=True).index(name)
         self.unfrozen_mask[i] = True
 
+    def freeze_all_parameters(self):
+        self.unfrozen_mask[:] = False
+
+    def thaw_all_parameters(self):
+        self.unfrozen_mask[:] = True
+
     def get_parameter(self, name):
         i = self.get_parameter_names(include_frozen=True).index(name)
         return self.get_parameter_vector(include_frozen=True)[i]
@@ -91,6 +109,14 @@ class Model(object):
         v = self.get_parameter_vector(include_frozen=True)
         v[i] = value
         self.set_parameter_vector(v, include_frozen=True)
+
+    def log_prior(self):
+        for p, b in zip(self.parameter_vector, self.parameter_bounds):
+            if b[0] is not None and p < b[0]:
+                return -np.inf
+            if b[1] is not None and p > b[1]:
+                return -np.inf
+        return 0.0
 
 
 class ConstantModel(Model):

### 2017-01-11 paper/figures/plot_setup.py
emcee3 for rotation example [ci skip]

@@ -6,7 +6,7 @@ from cycler import cycler
 from matplotlib import rcParams
 from savefig import monkey_patch
 
-__all__ = ["setup", "COLORS", "SQUARE_FIGSIZE", "HORIZONTAL_FIGSIZE"]
+__all__ = ["setup", "get_figsize", "COLORS", "SQUARE_FIGSIZE"]
 
 def setup():
     monkey_patch()
@@ -28,5 +28,7 @@ COLORS = dict(
     MODEL_2="#ff7f0e",
 )
 
-SQUARE_FIGSIZE = (4, 4)
-HORIZONTAL_FIGSIZE = (8, 4)
+def get_figsize(rows=1, cols=1):
+    return (4 * cols, 4 * rows)
+
+SQUARE_FIGSIZE = get_figsize(1, 1)

### 2017-01-11 paper/figures/rotation.py

@@ -3,13 +3,13 @@
 
 from __future__ import division, print_function
 
-import emcee
+import emcee3
 import fitsio
 import numpy as np
 import matplotlib.pyplot as plt
 from scipy.optimize import minimize
 
-from plot_setup import setup, SQUARE_FIGSIZE, COLORS
+from plot_setup import setup, get_figsize, COLORS
 
 import genrp
 from genrp import terms
@@ -70,6 +70,7 @@ def neg_log_like(params, y, gp):
     return -gp.log_likelihood(y)
 
 # Optimize with random restarts
+print("Running optimization with randomized restarts...")
 initial_params = gp.get_parameter_vector()
 bounds = gp.get_parameter_bounds()
 best = (np.inf, initial_params)
@@ -85,7 +86,6 @@ for i in range(10):
         ))
 gp.set_parameter_vector(best[1])
 ml_params = np.array(best[1])
-y_samp = gp.sample(t)
 
 # Do the MCMC
 def log_prob(params):
@@ -95,6 +95,7 @@ def log_prob(params):
     return gp.log_likelihood(y)
 
 # Initialize
+print("Running MCMC sampling...")
 ndim = len(ml_params)
 nwalkers = 32
 pos = ml_params + 1e-5 * np.random.randn(nwalkers, ndim)
@@ -106,10 +107,9 @@ while np.any(m):
     m = ~np.isfinite(lp)
 
 # Sample
-sampler = emcee.EnsembleSampler(nwalkers, ndim, log_prob)
-pos, _, _ = sampler.run_mcmc(pos, 250)
-sampler.reset()
-pos, _, _ = sampler.run_mcmc(pos, 1000)
+sampler = emcee3.Sampler()
+ensemble = emcee3.Ensemble(emcee3.SimpleModel(log_prob), pos)
+sampler.run(ensemble, 1250, progress=True)
 
 # Compute the model predictions
 gp.set_parameter_vector(ml_params)
@@ -122,20 +122,21 @@ tau = np.linspace(0, 4*period, 5000)
 acf = gp.kernel.get_value(tau)
 
 # Compute the sample predictions
-samples = sampler.flatchain
-samples = samples[np.random.randint(len(samples), size=1000)]
-psds = np.empty((len(samples), len(omega)))
-acfs = np.empty((len(samples), len(tau)))
-for i, s in enumerate(samples):
[…]

### 2017-01-11 paper/ms.tex
making solver picklable
adding astero figure to document

@@
 \end{eqnarray}
-where $j$ is an integer running from $-2$ to 2 and $\epsilon$, $A$, and $W$
+where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
 are shared nuisance parameters.
@@
 \dfmtodo{Add plots etc.}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
+\caption{A comparison between the empirical PSD and the posterior inference of
+the PSD as a mixture of stochastically driven simple harmonic oscillators.
+(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
+    on the full four year baseline of the mission.
+    The light gray curve shows the raw periodogram and the black curve has
+    been smoothed with a Gaussian filter with \dfmtodo{SOME WIDTH}.
+(middle) The same periodogram computed using about a month of data.
+(bottom) The inferred power spectrum under the mixture of SHOs model described
+    in the text.
+    The black line shows the median of posterior PSD and the gray contours
+    show the 68\% credible region.
+    \figurelabel{astero}}
+\end{center}
+\end{figure}
+
+\subsection{Stellar rotation}
+
+For many stars it is possible
 

### 2017-01-12 paper/ms.tex
adding rotation figures

@@
 (middle) The same periodogram computed using about a month of data.
-(bottom) The inferred power spectrum under the mixture of SHOs model described
-    in the text.
+(bottom) The power spectrum inferred using the mixture of SHOs model described
+    in the text and only one month of \kepler\ data.
     The black line shows the median of posterior PSD and the gray contours
@@
 
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
+\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
+    from the inference shown in \Figure{astero} compared to the published
+    value (error bar) based on the full \kepler\ dataset \dfmtodo{CITE}.
+    \figurelabel{astero-corner}}
+\end{center}
+\end{figure}
+
 \subsection{Stellar rotation}
 
-For many stars it is possible
+Another source of variability that can be measured from time series
+measurements of stars is rotation.
+The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
+quasiperiodic variations in photometric or spectroscopic observations
+(\dfmtodo{CITE}).
+It has been demonstrated that for light curves with nearly uniform sampling,
+the empirical autocorrelation function provides a reliable estimate of the
+rotation period of a star (\dfmtodo{CITE}) and that a GP model with a
+quasiperiodic covariance function can be used to make probabilistic
+measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
+The covariance function used for this type of analysis has the form
+\begin{eqnarray}\eqlabel{sine2}
+k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
+    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
+\end{eqnarray}
+where $P$ is the period of the oscillation.
+The key difference between this function and other quasiperiodic kernels is
+that it is everywhere positive.
+We can construct a simple \genrp\ covariance function with similar properties
+as follows
+\begin{eqnarray}\eqlabel{rot-kernel}
+k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
+    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
+\right]
+\end{eqnarray}
+for $a>0$, $b>0$, and $c>0$.
+The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
+but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
+as a drop-in replacement for a substantial gain in computational efficiency.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/rotation.pdf}
+\caption{Inferred constraints on a quasiperiodic GP model using the covariance
+    function in \eq{rot-kernel} and one quarter of \kepler\ data.
+(left) The \kepler\ data (black points) and the maximum likelihood model
+    prediction (blue curve).
+    The solid blue line shows the predictive mean and the blue contours show
+    the predictive standard deviation.
+(center) Inferred constraints on the model PSD.
+    The dashed line shows the maximum likelihood PSD, the blue solid line
+    shows the median of posterior PSD, and the blue contours show the 68\%
+    credible region.
+(right) Inferred constraints on the model covariance function from
+    \eq{rot-kernel}.
+    The dashed line shows the maximum likelihood model, the blue solid line
+    shows the median of posterior, and the blue contours show the 68\%
+    credible region.
+    \figurelabel{rotation}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/rotation-period.pdf}
+\caption{The posterior constraint on the rotation period of KIC~1430163 using
+    the dataset and model from \Figure{rotation}.
+    The period is the parameter $P$ in \eq{rot-kernel} and this figure shows
+    the posterior distribution marginalized over all other nuisance parameters
+    in \eq{rot-kernel}.
+    This is consistent with the published rotation period made using the
+    autocorrelation function and the full \kepler\ data (\dfmtodo{CITE}).
+    \figurelabel{rotation-period}}
+\end{center}
+\end{figure}
 

### 2017-01-12 paper/ms.tex
consistent bibitems

@@
 
-The power spectrum is computed from taking the square of the Fourier transform of the data
-time series; hence, the power-spectrum must be non-negative.  In constructing a kernel,
-if any of the parameters $a_j$ or $b_j$ are negative, then it is possible for the power
-spectrum to go negative, which violates the definition of the power spectrum.  Consequently,
-if any of these coefficients is negative, it is necessary to check that the power spectrum
-is still non-negative.  Note that a non-negative power-spectrum does not require the auto-correlation
-function to be positive for all $\tau$.  The positive power spectrum is related to the
-positive eigenvalues of the covariance matrix which are necessary for a positive-definite matrix
-in limiting cases \citep{messerschmitt2006autocorrelation}.  We find empirically that requiring an everywhere
-positive power spectrum results in positive eigenvalues for the covariance matrix, and
-so we describe here how to ensure a positive power spectrum using Sturm's theorem.
+The power spectrum is computed from taking the square of the Fourier transform
+of the data time series; hence, the power-spectrum must be non-negative.  In
+constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
+then it is possible for the power spectrum to go negative, which violates the
+definition of the power spectrum.
+Consequently, if any of these coefficients is negative, it is necessary to
+check that the power spectrum is still non-negative.  Note that a non-negative
+power-spectrum does not require the auto-correlation
+function to be positive for all $\tau$.  The positive power spectrum is
+related to the positive eigenvalues of the covariance matrix which are
+necessary for a positive-definite matrix in limiting cases
+\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
+positive power spectrum results in positive eigenvalues for the covariance
+matrix, and so we describe here how to ensure a positive power spectrum using
+Sturm's theorem.
 %Relation of positive power spectrum to positive definite covariance matrix.
@@
 
-A procedure based upon Sturm's theorem \citep{Dorrie1965} allows one to determine whether there are any
+A procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
 real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
@@
 like $\log g$.
+
+KIC~11615890 \citep{Pinsonneault:2014}
+\begin{eqnarray}
+    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
+    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz}
+\end{eqnarray}
 
@@
 % A promising technique for addressing both of these goals is the Gaussian Process formalism
-% \citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
+% \citep{Rasmussen2006,Gibson:2012}.  The basis of this method is to assume that stellar variability
 % is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
@@
 % which may be used to characterize properties of the star, such as the rotation period,
-% the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
+% the asteroseismic frequencies, or granulation noise known as `flicker'
+% \citep{Aerts2010,Noyes:1984,Bastien:2013}.
 % The goals of planet detection include measuring the Doppler shift of the star, measuring
@@
 % algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
-% solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
+% solver is only approximate \citep{Ambikasaran:2013,Ambikasaran:2016}.  In addition, the
 % solver is complicated and can be challenging to use with general choices of kernel.  Another
@@
 % rely on the structure of Toeplitz matrices, which require uniformly space time series,
-% to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
+% to obtain approximate solutions of the likelihood function \citep{Wilson:2015},
 % the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
@@
 % based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
-% model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
+% model \citep[aka \texttt{CARMA};][]{Kelly:2014}.  This approach can model a stationary power-spectrum that
 % is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
@@
 % activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
-% Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
-% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
+% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
+% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
 % parameterized this way as well.  In this paper
@@
 %\item Solution(s):\\
-%  a). HODLR \citep{Ambikasaran2013,Ambikasaran2016}\\
-%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly2014}\\
+%  a). HODLR \citep{Ambikasaran:2013,Ambikasaran:2016}\\
+%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly:2014}\\
 %  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
 %  d).
-KISS-GP \citep{WilsonNickisch2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp.
+KISS-GP \citep{Wilson:2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp.
 1775-1784, 2015
@@
 % \item Possible future applications:\\
-%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
+%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{Wilson:2013}, but
 %       with Lorentzians, not Gaussians.\\
@@
 formula is simultaneously physically relevant and computationally convenient.
-The power spectrum of a single exponential scales as $f^{-2}$ at high frequency,
-while the Mat\'ern kernel scales as $f^{-4}$ which more closely matches the power
-spectrum due to activity and granulation in the Sun \citep{2004A&A...414.1139A,Michel:2009}.
+The power spectrum of a single exponential scales as $f^{-2}$ at high
+frequency, while the Mat\'ern kernel scales as $f^{-4}$ which more
+closely matches the power spectrum due to activity and granulation in
+the Sun \citep{Aigrain:2004,Michel:2009}.
 \item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
@@
 gravitational wave time series, and so Gaussian processes may be a way to address
-this problem \citep{Moore2016}.  Accreting black holes show time series which
-may be modeled by correlated noise \citep{Kelly2014};  indeed, this was the
-motivation for the original technique developed by Rybicki \& Press \citep{1992ApJ...398..169R,Rybicki:1995}.
-This approach may be broadly used for characterizing quasar variability \citep{2010ApJ...721.1014M},
-measuring time lags with reverberation mapping \citep{2011ApJ...735...80Z},
-and modeling time delays in multiply-imaged gravitationally-lensed systems \citep{1998ApJ...507..108P}.
-
-Outside of astronomy, this technique may have application to seismology \citep{Robinson1967},
+this problem \citep{Moore:2016}.  Accreting black holes show time series which
+may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
+motivation for the original technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
+This approach may be broadly used for characterizing quasar variability
+\citep{MacLeod:2010}, measuring time lags with reverberation mapping
+\citep{Zu:2011}, and modeling time delays in multiply-imaged
+gravitationally-lensed systems \citep{Press:1998}.
+
+Outside of astronomy, this technique may have application to seismology
+\citep{Robinson:1967},
 
@@
     % \project{emcee} \citep{Foreman-Mackey:2013},
-    % \project{george} \citep{Ambikasaran2016},
+    % \project{george} \citep{Ambikasaran:2016},
 	% \project{matplotlib} \citep{Hunter:2007},
@@
 As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
-solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix
+solved with a banded solver, as in \citet{Press:1992}.  The determinant of this extended matrix
 may be computed from the LU decomposition, and it is related to the determinant of the kernel by a

### 2017-01-12 paper/ms.tex

@@
 \newcommand{\kepler}{\project{Kepler}}
-\newcommand{\genrp}{\project{genrp}}
+\newcommand{\celerite}{\project{celerite}}
 
@@
 These implementations are available online at \project{GitHub}
-\url{https://github.com/dfm/GenRP} and \project{Zenodo} \dfmtodo{add zenodo
+\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
 archive}.
@@
 % 4. Examples
-%   - Simulated data: a simple exact GenRP simulation, a GenRP simulation with
+%   - Simulated data: a simple exact celerite simulation, a celerite simulation with
 %     structure, a few standard kernels, a AGW kernel
@@
 In the following Section, we present a method of improving this scaling that
-we call the \genrp\ method.
-The \genrp\ method requires using a specific model for the covariance
+we call the \celerite\ method.
+The \celerite\ method requires using a specific model for the covariance
 $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
@@
 using $t$ as the input coordinate because one-dimensional GPs are often
-applied to time series data but this isn't a real restriction and the \genrp\
+applied to time series data but this isn't a real restriction and the \celerite\
 method can be applied to \emph{any} one-dimensional dataset.}
-Furthermore, the covariance function for the \genrp\ method is ``stationary''.
+Furthermore, the covariance function for the \celerite\ method is ``stationary''.
 This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
@@
 
-\section{The genrp model}
+\section{The celerite model}
 
@@
 In this case, the covariance function becomes
-\begin{eqnarray}\eqlabel{genrp-kernel-complex}
+\begin{eqnarray}\eqlabel{celerite-kernel-complex}
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
@@
 
-\eq{genrp-kernel-complex} can also be expressed as
-\begin{eqnarray}\eqlabel{genrp-kernel}
+\eq{celerite-kernel-complex} can also be expressed as
+\begin{eqnarray}\eqlabel{celerite-kernel}
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
@@
 density (PSD) of the model and it is given by
-\begin{eqnarray}\eqlabel{genrp-psd}
+\begin{eqnarray}\eqlabel{celerite-psd}
 S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
@@
 If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
-term of \eq{genrp-kernel} becomes
+term of \eq{celerite-kernel} becomes
 \begin{eqnarray}
@@
 
-\section{Genrp as a model of stellar variations}
-
-Another special case of the \genrp\ model of great physical interest is a
+\section{celerite as a model of stellar variations}
+
+Another special case of the \celerite\ model of great physical interest is a
 stochastically-driven simple harmonic oscillator.
@@
 = \sqrt{2/\pi}\,S_0$.
-The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
+The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
 \begin{eqnarray}
@@
 for $Q \ge \frac{1}{2}$.
-For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
+For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
 terms with parameters
@@
 This suggests that the Mat\'ern-3/2 covariance can be well approximated using
-the \genrp\ framework with a small value of $f$ in \eq{approx-matern} but we
+the \celerite\ framework with a small value of $f$ in \eq{approx-matern} but we
 caution that this could also lead to numerical issues with the solver.
@@
 likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
-measurements in $\mathcal{O}(N)$ operations using the \genrp\ method described
+measurements in $\mathcal{O}(N)$ operations using the \celerite\ method described
 in the previous section.
@@
 \item an OU process
-\item a full GenRP process
+\item a full celerite process
 \item Some standard kernels: Matern, Exp-squared, etc.
@@
 
-\genrp\ alleviates these problems by providing a physically motivated
+\celerite\ alleviates these problems by providing a physically motivated
 probabilistic model that can be evaluated efficiently even for large datasets.
@@
 that it is everywhere positive.
-We can construct a simple \genrp\ covariance function with similar properties
+We can construct a simple \celerite\ covariance function with similar properties
 as follows
@@
 
-Limitations of GenRP: one-dimension, stationary, etc.
+Limitations of celerite: one-dimension, stationary, etc.
 
@@
 % which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
-% Rybicki-Press' (\texttt{GenRP}) approach
+% Rybicki-Press' (\texttt{celerite}) approach
 % has many advantages:  the computation time and storage both scale as $O(N_{time})$,
@@
 % the coefficients (except that the kernel needs to be positive definite).  Unlike
-% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
+% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{celerite} method is exact.  The main drawback
 % of this approach is that the sum of exponential kernels with real coefficients can
@@
 % parameterized this way as well.  In this paper
-% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
+% we extend the \texttt{celerite} approach to kernels with complex exponentials, which when
 % combined with the complex conjugate leads to damped harmonic auto-correlation functions
@@
 
-Although the complex kernel is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
+Although the complex kernel is a straightforward extension to the \texttt{celerite} formalism, there is a disadvantage
 in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
@@
 All of the code used in this project is available from
-\url{https://github.com/dfm/GenRP} under the MIT open-source software
+\url{https://github.com/dfm/celerite} under the MIT open-source software
 license.
@@
 
-There are three primary applications we envision for the \genrp\ formalism: 1).\ modeling
+There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
 the variability of an astrophysical system to infer it properties; 2).\
@@
 
-Given the drawbacks of these approaches, the \genrp\ formalism allows both a fast
+Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
 computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
@@
 are obtained, the effects of stellar variability will more dramatically impact the
-correct inference of planetary transit parameters, and so we expect that \genrp\
+correct inference of planetary transit parameters, and so we expect that \celerite\
 will be important for transit timing, transit spectroscopy, Doppler beaming,
@@
 
-\bibliography{genrp}
+\bibliography{celerite}
 

### 2017-01-15 paper/ms.tex
some notes about implementation

@@
 \usepackage{url}
-\usepackage{amssymb,amsmath}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{natbib}
@@
 % ----------------------------------- %
+
+% Matrix fix:
+% http://tex.stackexchange.com/questions/317824/letter-c-appearing-inside-pmatrix-environment-with-aastex
+\makeatletter
+\def\env@matrix{\hskip -\arraycolsep % taken from amsmath.sty lines 895ff
+  \let\@ifnextchar\new@ifnextchar
+  \array{*{\c@MaxMatrixCols}c}}
+\makeatother
+
+% Column spacing in matrix
+% http://tex.stackexchange.com/questions/275725/adjusting-separation-between-matrix-entries
+\setlength\arraycolsep{25pt}
 
@@
 function is
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{kernel-simple}
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
@@
 \section{Implementation \& performance}
+
+\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
+elements are given by
+\begin{eqnarray}
+K_{nm} = k(\tau_{nm})
+\end{eqnarray}
+for $k(\tau_{nm})$ given by equation \eq{kernel-simple} could be computed
+efficiently by taking advantage of the structure of this covariance function
+and \citet{Ambikasaran:2015} generalized this computation to apply to the full
+mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
+efficient method for computing the determinant of $K$.
+
+\subsection{The algorithm}
+
+To provide some insight for this method, we will start by working through the
+simplest possible example.
+In this case, we'll assume that we have three data points
+$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
+uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
+like to compute the likelihood of these data under a GP model with the
+covariance function
+\begin{eqnarray}
+k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
+\end{eqnarray}
+This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
+and $d=0$.
+To demonstrate motivate this method, we will write out the full system of
+equations that we must solve to apply the inverse of $K$.
+In matrix notation, this can be written as
+\begin{eqnarray}
+K\,\bvec{x} &=& \bvec{y} \\
+\begin{pmatrix}
+    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
+    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
+    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
+\end{pmatrix}\,
+\begin{pmatrix}
+    x_1 \\ x_2 \\ x_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ y_2 \\ y_3
+\end{pmatrix} \quad.
+\end{eqnarray}
+Now, if we introduce the variables
+\begin{eqnarray}
+    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
+\end{eqnarray}
+where $u_{N} = 0$, and
+\begin{eqnarray}\eqlabel{algo-first}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
+\end{eqnarray}
+where $g_{0} = 0$,
+the system of equations can be rewritten as
+\begin{eqnarray}
+(a+{\sigma_1}^2)\,x_1 + e^{c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{c\,t_2}\,u_2 &=& y_2 \\
+a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
+\end{eqnarray}
+The system defined by \eq{algo-first} through \eq{algo-last} can be
+rewritten as a matrix equation to show the benefit that this seemingly trivial
+reformulation provides:
+\begin{eqnarray}
+\begin{pmatrix}
+    a+{\sigma_1}^2 & e^{c\,t_1} & 0 & 0 & 0 & 0 & 0 \\
+    e^{c\,t_1} & 0 & -1 & 0 & 0 & 0 & 0 \\
+    0 & -1 & 0 & e^{-c\,t_2} & a & 0 & 0 \\
+    0 & 0 & e^{-c\,t_2} & a+{\sigma_2}^2 & e^{c\,t_2} & 0 & 0 \\
+    0 & 0 & a & e^{c\,t_2} & 0 & -1 & 0 \\
+    0 & 0 & 0 & 0 & -1 & 0 & e^{-c\,t_3} \\
+    0 & 0 & 0 & 0 & 0 & e^{-c\,t_3} & a+{\sigma_3}^2 \\
+\end{pmatrix}\,
+\begin{pmatrix}
+    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
+\end{pmatrix} \quad.
+\end{eqnarray}
 

### 2017-01-15 paper/ms.tex
typo in extended matrix
words about implementation

@@
 
-\subsection{The algorithm}
+\subsection{An example}
 
@@
 In matrix notation, this can be written as
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{impl-matrix}
 K\,\bvec{x} &=& \bvec{y} \\
@@
 \begin{eqnarray}
-(a+{\sigma_1}^2)\,x_1 + e^{c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{c\,t_2}\,u_2 &=& y_2 \\
+(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
 a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
@@
 \begin{pmatrix}
-    a+{\sigma_1}^2 & e^{c\,t_1} & 0 & 0 & 0 & 0 & 0 \\
-    e^{c\,t_1} & 0 & -1 & 0 & 0 & 0 & 0 \\
-    0 & -1 & 0 & e^{-c\,t_2} & a & 0 & 0 \\
-    0 & 0 & e^{-c\,t_2} & a+{\sigma_2}^2 & e^{c\,t_2} & 0 & 0 \\
-    0 & 0 & a & e^{c\,t_2} & 0 & -1 & 0 \\
-    0 & 0 & 0 & 0 & -1 & 0 & e^{-c\,t_3} \\
-    0 & 0 & 0 & 0 & 0 & e^{-c\,t_3} & a+{\sigma_3}^2 \\
+    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
+    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
+    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
+    0 & 0 & 0 & 0 & -1 & 0 & a \\
+    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
 \end{pmatrix}\,
@@
     y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
-\end{pmatrix} \quad.
-\end{eqnarray}
+\end{pmatrix}\nonumber
+\end{eqnarray}
+where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
+system and rewrite \eq{impl-matrix} as
+\begin{eqnarray}
+    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
+\end{eqnarray}
+Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
+started with, it is now has a sparse banded structure that can be exploited to
+solve the system efficiently.
+In particular, band matrix solvers are available that can perform an
+LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
+operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
+general~--~and we can take advantage of these algorithms to solve our system
+exactly because the target vector $\bvec{x}$ is a subset of the elements of
+$\bvec{x}_\mathrm{ext}$.
+
+In the following section, we will discuss this method more generally but it's
+worth noting a few important facts that can already be seen in this example.
+First, the fundamental reason why this matrix $K$ can be solved efficiently is
+the following property of exponentials
+\begin{eqnarray}
+    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
+    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
+    e^{-c\,(t_3 - t_1)}
+\end{eqnarray}
+and it is important to note that this property does not extend to other common
+covariance functions like the ``exponential-squared'' function
+\begin{eqnarray}
+    k(\tau) \propto e^{-c\,\tau^2} \quad.
+\end{eqnarray}
+Furthermore, our derivation of the extended matrix requires that the data
+points be monotonically sorted in time.
+Neither of these properties will be satisfied in general for multidimensional
+inputs and all of our following discussion will assume a sorted
+one-dimensional dataset.
+
+\subsection{The algorithm}
+
+\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
+extended matrix formalism in practice.
+First, even if the covariance function is a mixture of exponentials, the
+extended matrix will still be banded with bandwidth \dfmtodo{check}.
+Second, \citet{Ambikasaran:2015} proved that the determinant of
+$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
 

### 2017-01-16 paper/ms.tex
my version of the extended system of equations

@@
 
-\subsection{The algorithm}
-
 \citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
@@
 First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with bandwidth \dfmtodo{check}.
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components $J$.
 Second, \citet{Ambikasaran:2015} proved that the determinant of
 $K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
+This means that we can use this extended matrix formalism to compute the
+marginalized likelihood in $\mathcal{O}(N\,J^2)$ operations.
+
+\subsection{The algorithm}
+
+In this section, we generalize the method from the previous section to the
+covariance function given by \eq{celerite-kernel-complex}.
+This derivation follows \citet{Ambikasaran:2015} but it includes explicit
+treatment of complex parameters.
+
+\begin{eqnarray}
+\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
+\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,x_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,g_{n-1,j} - \psi_{n,j}\,x_n - \psi_{n,j}\,h_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + \frac{a_j}{2}\,x_n + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} - \frac{b_j}{2}\,x_n - \psi_{n+1,j}\,v_{n+1,j}
+\end{eqnarray}
+
+\begin{eqnarray}
+\sum_{j=1}^J \left[\frac{a_j}{2}\,g_{n,j}+\frac{b_j}{2}\,h_{n,j}\right]
++ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
++ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
+    &=& y_n
+\end{eqnarray}
 

### 2017-01-16 paper/figures/matrix.py
adding new matrix figure

@@ -1,0 +1,111 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+from plot_setup import setup, get_figsize
+
+setup()
+
+J = 2
+width = 2*J + 1 + 4*J + 1 + 2*J
+height = 1 + 4*J
+
+block = np.zeros((height, width), dtype=int)
+
+# y_n
+block[0, 2*J] = 1  # diag
+block[0, :2*J:2] = 2  # a/2
+block[0, 1:2*J:2] = 3  # b/2
+block[0, 2*J+1:4*J+1:2] = 4  # phi
+block[0, 2*J+2:4*J+1:2] = 5  # psi
+
+# g_{n,j}
+for j in range(J):
+    block[1+2*j, 2*j] = 4  # phi
+    block[1+2*j, 2*j+1] = 5  # psi
+    block[1+2*j, 2*J] = 4  # phi
+    block[1+2*j, 4*J+2*j+1] = 7  # -1
+
+# h_{n,j}
+for j in range(J):
+    block[2+2*j, 2*j] = 5  # psi
+    block[2+2*j, 2*j+1] = 6  # -phi
+    block[2+2*j, 2*J] = 5  # psi
+    block[2+2*j, 4*J+2*j+2] = 8  # 1
+
+# u_{n,j}
+for j in range(J):
+    block[1+2*J+2*j, 6*J+1] = 2  # a/2
+    block[1+2*J+2*j, 6*J+2+2*j] = 4  # phi
+    block[1+2*J+2*j, 6*J+3+2*j] = 5  # psi
+    block[1+2*J+2*j, 2*J+2*j+1] = 7  # -1
+
+# v_{n,j}
+for j in range(J):
+    block[2+2*J+2*j, 6*J+1] = 3  # b/2
+    block[2+2*J+2*j, 6*J+2+2*j] = 5  # psi
+    block[2+2*J+2*j, 6*J+3+2*j] = 6  # -phi
+    block[2+2*J+2*j, 2*J+2*j+2] = 8  # 1
+
+# Combine the blocks
+N = 5
+full_dim = (1 + 4*J) * (N-1) + 1
+full = np.zeros((full_dim, full_dim), dtype=int)
+
+full[:height, :6*J+2] = block[:, 2*J:]
+
+for n in range(1, N):
[…]

### 2017-01-16 paper/figures/plot_setup.py

@@ -6,7 +6,16 @@ from cycler import cycler
 from matplotlib import rcParams
 from savefig import monkey_patch
 
-__all__ = ["setup", "get_figsize", "COLORS", "SQUARE_FIGSIZE"]
+__all__ = ["setup", "get_figsize", "COLORS", "COLOR_CYCLE", "SQUARE_FIGSIZE"]
+
+COLOR_CYCLE = (
+    "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
+    "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
+)
+
+# s = ("1f77b4aec7e8ff7f0effbb782ca02c98df8ad62728ff98969467bdc5b0d58c564bc49c9"
+#      "4e377c2f7b6d27f7f7fc7c7c7bcbd22dbdb8d17becf9edae5")
+# COLOR_CYCLE = ["#" + s[i*6:(i+1)*6] for i in range(len(s)//6)]
 
 def setup():
     monkey_patch()
@@ -17,10 +26,7 @@ def setup():
     rcParams["text.usetex"] = True
     rcParams["text.latex.preamble"] = r"\usepackage{cmbright}"
     rcParams["figure.autolayout"] = True
-    rcParams["axes.prop_cycle"] = cycler("color", (
-        "#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b",
-        "#e377c2", "#7f7f7f", "#bcbd22", "#17becf",
-    ))  # d3.js color cycle
+    rcParams["axes.prop_cycle"] = cycler("color", COLOR_CYCLE)
 
 COLORS = dict(
     DATA="k",

### 2017-01-17 paper/figures/matrix.py
incorrect factors of 2 fixed

@@ -92,8 +92,8 @@ ax.set_yticks([])
 
 names = [
     r"${\sigma_n}^2 + \sum a_j$",
-    r"$a_j / 2$",
-    r"$b_j / 2$",
+    r"$a_j$",
+    r"$b_j$",
     r"$\phi_{n,j}$",
     r"$\psi_{n,j}$",
     r"$-\phi_{n,j}$",

### 2017-01-18 paper/ms.tex
adding new matrix figure
new matrix figure in ms
incorrect factors of 2 fixed
adding transit example

--- 2017-01-15 paper/ms.tex
@@
 First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with bandwidth \dfmtodo{check}.
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components $J$.
 Second, \citet{Ambikasaran:2015} proved that the determinant of
 $K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
+This means that we can use this extended matrix formalism to compute the
+marginalized likelihood in $\mathcal{O}(N\,J^2)$ operations.
+
+\subsection{The algorithm}
+
+In this section, we generalize the method from the previous section to the
+covariance function given by \eq{celerite-kernel}.
+This derivation follows \citet{Ambikasaran:2015} but it includes explicit
+treatment of complex parameters.
+
+In the case of the full \celerite\ covariance function
+(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
+analogy to the $u_n$ and $g_n$ that we introduced in the previous section
+\begin{eqnarray}\eqlabel{full-system-first}
+\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
+\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,x_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,x_n - \psi_{n,j}\,g_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,x_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,x_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
+\end{eqnarray}
+with the boundary conditions
+\begin{eqnarray}
+    g_{0,j} = 0 \quad, \quad
+    h_{0,j} = 0 \quad, \quad
+    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
+    v_{N,j} = 0
+\end{eqnarray}
+for all $j$.
+Using these variables and some algebra, we find that the following expression
+\begin{eqnarray}\eqlabel{full-system-last}
+\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
++ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
++ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
+    &=& y_n
+\end{eqnarray}
+is equivalent to the target matrix equation
+\begin{eqnarray}
+K\,\bvec{x} &=& \bvec{y} \quad.
+\end{eqnarray}
+\eq{full-system-first} through \eq{full-system-last} define a banded matrix
+equation in the ``extended'' space and, as before, this can be used to solve
+for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(J^2\,N)$ operations.
+\Figure{matrix} shows a pictorial representation of the sparsity pattern of
+the extended matrix $K_\mathrm{ext}$.
+Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
+$\bvec{x}_\mathrm{ext}$ and $\bvec{y}_\mathrm{ext}$ are defined schematically
+as
+\begin{eqnarray}
+{\bvec{x}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    x_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & x_2 & u_{2,j} & \cdots &
+    h_{N-1,j} & x_N
+\end{array}\right)
+\end{eqnarray}
+and
+\begin{eqnarray}
+{\bvec{y}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    y_1 & 0 & 0 & 0 & 0 & y_2 & 0 & \cdots &
+    0 & y_N
+\end{array}\right) \quad.
+\end{eqnarray}
+
+After constructing the extended matrix (using a compact storage format), the
+extended matrix can be factorized using a LU-decomposition\footnote{Even
+though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
+Cholesky solver cannot be used for increased efficiency.} routine optimized
+for band matrices.
+This decomposition can be used to compute the determinant of $K$ and to solve
+$K^{-1}\,\bvec{y}$.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
+\caption{A pictorial representation of the sparse extended matrix
+$K_\mathrm{ext}$ with $N=5$ and $J=2$.
+Each colored block corresponds to a non-zero entry in the matrix as described
+in the legend.
+    \figurelabel{matrix}}
+\end{center}
+\end{figure}
 

### 2017-01-18 paper/figures/transit_model.py

@@ -1,0 +1,50 @@
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import transit
+import numpy as np
+from celerite import terms, modeling
+
+__all__ = ["RotationTerm", "TransitModel"]
+
+
+class RotationTerm(terms.Term):
+    parameter_names = ("log_amp", "log_timescale", "log_period", "log_factor")
+
+    def get_real_coefficients(self):
+        f = np.exp(self.log_factor)
+        return (
+            np.exp(self.log_amp) * (1.0 + f) / (2.0 + f),
+            np.exp(-self.log_timescale),
+        )
+
+    def get_complex_coefficients(self):
+        f = np.exp(self.log_factor)
+        return (
+            np.exp(self.log_amp) / (2.0 + f),
+            0.0,
+            np.exp(-self.log_timescale),
+            2*np.pi*np.exp(-self.log_period),
+        )
+
+class TransitModel(modeling.Model):
+    parameter_names = ("mean_flux", "log_period", "log_ror", "log_duration",
+                       "t0", "impact", "q1", "q2")
+
+    def __init__(self, texp, *args, **kwargs):
+        self.texp = texp
+        super(TransitModel, self).__init__(*args, **kwargs)
+
+    def get_value(self, t):
+        system = transit.SimpleSystem(
+            period=np.exp(self.log_period),
+            ror=np.exp(self.log_ror),
+            duration=np.exp(self.log_duration),
+            t0=self.t0,
+            impact=self.impact,
+            q1=self.q1,
+            q2=self.q2
+        )
+        lc = system.light_curve(t, texp=self.texp)
+        return 1e3 * (lc - 1.0) + self.mean_flux

### 2017-01-18 paper/figures/transit_sample.py

@@ -1,0 +1,181 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import os
+import sys
+import corner
+import emcee3
+import pickle
+import fitsio
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.optimize import minimize
+
+import celerite
+from plot_setup import setup, get_figsize
+from transit_model import RotationTerm, TransitModel
+
+setup()
+np.random.seed(42)
+
+data, hdr = fitsio.read("data/kplr001430163-2013011073258_llc.fits",
+                        header=True)
+texp = float(hdr["INT_TIME"] * hdr["NUM_FRM"]) / 60. / 60. / 24.
+
+N = 500
+m = data["SAP_QUALITY"] == 0
+m &= np.isfinite(data["TIME"])
+m &= np.isfinite(data["PDCSAP_FLUX"])
+t = np.ascontiguousarray(data["TIME"][m], dtype=np.float64)[:N]
+y = np.ascontiguousarray(data["PDCSAP_FLUX"][m], dtype=np.float64)[:N]
+yerr = np.ascontiguousarray(data["PDCSAP_FLUX_ERR"][m], dtype=np.float64)[:N]
+t -= t.min()
+
+# Build the true model
+true_model = TransitModel(
+    texp,
+    0.0,
+    np.log(4.0),    # period
+    np.log(0.015),  # Rp / Rs
+    np.log(0.5),    # duration
+    0.5*t.max(),    # t_0
+    0.5,            # impact
+    0.5,            # q_1
+    0.5,            # q_2
+)
+true_params = np.array(true_model.get_parameter_vector())
+
+# Inject the transit into the data
+true_transit = 1e-3*true_model.get_value(t) + 1.0
+y *= true_transit
+
+# Normalize the data
+med = np.median(y)
+y = (y / med - 1.0) * 1e3
+yerr *= 1e3 / med
+
+# Set up the GP model
+mean = TransitModel(
+    texp,
+    0.0,
+    np.log(4.0),
+    np.log(0.015),
+    np.log(0.5),
+    0.5*t.max(),
+    0.5,
+    0.5,
+    0.5,
+    bounds=[
+        (-0.5, 0.5),
+        np.log([3.9, 4.1]),
+        (np.log(0.005), np.log(0.1)),
[…]

### 2017-01-19 paper/ms.tex
adding transit figures to ms
switching direction of Matern approximation
including both matern approximations
Minor edits to ms.tex

@@
 It turns out that this kernel function can be made even more general by
-introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
+introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$,
+and taking the real component of this complex kernel.
 In this case, the covariance function becomes
@@
 $\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
-uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
+variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
 like to compute the likelihood of these data under a GP model with the
@@
 \begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{x} &=& \bvec{y} \\
+K\,\bvec{x} &=& \bvec{y}, \\
 \begin{pmatrix}
@@
 This means that we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(N\,J^2)$ operations.
+marginalized likelihood in $\mathcal{O}(J^2\,N)$ operations.
 
@@
 This derivation follows \citet{Ambikasaran:2015} but it includes explicit
-treatment of complex parameters.
+treatment of complex parameters, and their complex conjugates.
 
@@
 The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{sho-complex}
 a_j &=& S_0\,\omega_0\,Q \\
@@
 terms with parameters
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{sho-real}
 a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
         \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
-b_{j\pm} &=& 0 \\
-    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
-d_{j\pm} &=& 0 \quad.
+b_{j\pm} &=& 0 \nonumber\\
+    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
+    \nonumber\\
+d_{j\pm} &=& 0 \quad. \nonumber
 \end{eqnarray}
@@
 \end{eqnarray}
-with the corresponding covariance function
+with the corresponding covariance function (using
+\eqalt{celerite-kernel} and \eqalt{sho-real})
+\begin{eqnarray}\eqlabel{approx-matern}
+k(\tau) &=& \lim_{f \to 0}\,
+    \frac{1}{2}\,S_0\,\omega_0\,
+    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
+          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
+    \right] \\
+&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
+\end{eqnarray}
+or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
 \begin{eqnarray}\eqlabel{approx-matern}
@@
     \figurelabel{rotation-period}}
+\end{center}
+\end{figure}
+
+\subsection{Exoplanet transit fitting}
+
+KIC~1430163 with simulated transit.
+\Figure{transit-ml} and \Figure{transit-corner}.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/transit-ml.pdf}
+\caption{\figurelabel{transit-ml}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/transit-corner.pdf}
+\caption{\figurelabel{transit-corner}}
 \end{center}

### 2017-01-20 paper/ms.tex
normalization of astero plots

@@
 It turns out that this kernel function can be made even more general by
-introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$,
-and taking the real component of this complex kernel.
+introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
+$c_j \to c_j\pm i\,d_j$.
 In this case, the covariance function becomes

### 2017-01-20 paper/figures/astero_plot.py

@@ -27,6 +27,7 @@ uHz_conv = 1e-6 * 24 * 60 * 60
 # Save the current state of the GP and data
 with open("astero-{0}.pkl".format(kicid), "rb") as f:
     gp, fit_y, freq, power_all, power_some = pickle.load(f)
+measurement_var = np.median(gp._yerr**2)
 
 backend = emcee3.backends.HDFBackend("astero-{0}.h5".format(kicid))
 
@@ -67,6 +68,7 @@ plt.close(fig)
 
 # Trim the burnin
 samples = backend.get_coords(discard=burnin, flat=True)
+log_probs = backend.get_log_probability(discard=burnin, flat=True)
 
 # Compute the model predictions
 time_grid = np.linspace(0, 1.4, 5000)
@@ -78,6 +80,10 @@ for i, j in enumerate(np.random.randint(len(samples), size=n)):
     gp.set_parameter_vector(s)
     psds[i] = gp.kernel.get_psd(2*np.pi*freq)
     acors[i] = gp.kernel.get_value(time_grid)
+
+# Get the median modes
+gp.set_parameter_vector(samples[np.argmax(log_probs)])
+peak_freqs = gp.kernel.get_freqs()
 
 # Plot the predictions
 fig, (ax1, ax2) = plt.subplots(1, 2, figsize=get_figsize(1, 2))
@@ -119,23 +125,27 @@ plt.close(fig)
 # Make comparison plot
 fig, axes = plt.subplots(3, 1, sharex=True, figsize=get_figsize(2.5, 2))
 
+factor = 1.  # / (2*np.pi)**2
 freq_uHz = freq / uHz_conv
-axes[0].plot(freq_uHz, np.sqrt(power_all), "k", alpha=0.3, rasterized=True)
-axes[0].plot(freq_uHz, np.sqrt(gaussian_filter(power_all, 5)), "k",
+axes[0].plot(freq_uHz, power_all * factor, "k", alpha=0.3, rasterized=True)
+axes[0].plot(freq_uHz, gaussian_filter(power_all, 5) * factor, "k",
              rasterized=True)
 
-axes[1].plot(freq_uHz, np.sqrt(power_some), "k", alpha=0.3, rasterized=True)
-axes[1].plot(freq_uHz, np.sqrt(gaussian_filter(power_some, 5)), "k",
+axes[1].plot(freq_uHz, power_some * factor, "k", alpha=0.3, rasterized=True)
[…]

### 2017-01-20 paper/figures/astero_term.py

@@ -34,6 +34,11 @@ class AsteroTerm(terms.Term):
             np.append(beta, 0.5*omega/Q*np.sqrt(4*Q*Q-1)),
         )
 
+    def get_freqs(self):
+        j = np.arange(-self.nterms, self.nterms+1, 1)
+        delta = j*np.exp(self.log_delta_nu) + self.epsilon
+        return np.exp(self.log_nu_max) + delta
+
     def log_prior(self):
         lp = super(AsteroTerm, self).log_prior()
         if not np.isfinite(lp):

### 2017-02-02 python/celerite/modeling.py
updating some examples

@@ -222,6 +222,14 @@ class ModelSet(Model):
     def thaw_parameter(self, name):
         self._apply_to_parameter("thaw_parameter", name)
 
+    def freeze_all_parameters(self):
+        for model in self.models.values():
+            model.freeze_all_parameters()
+
+    def thaw_all_parameters(self):
+        for model in self.models.values():
+            model.thaw_all_parameters()
+
     def get_parameter(self, name):
         return self._apply_to_parameter("get_parameter", name)
 

### 2017-02-02 python/celerite/terms.py

@@ -142,7 +142,7 @@ class ComplexTerm(Term):
     def log_prior(self):
         # Constraint required for term to be positive definite. Can be relaxed
         # with multiple terms but must be treated carefully.
-        if self.log_a + self.log_c < self.log_b + self.log_d:
+        if self.fit_b and self.log_a + self.log_c < self.log_b + self.log_d:
             return -np.inf
         return super(ComplexTerm, self).log_prior()
 

### 2017-02-02 python/examples/simulated/agw.py

@@ -12,8 +12,8 @@ from scipy.optimize import minimize
 import george
 from george import kernels
 
-import genrp
-from genrp.kernels import SHOTerm
+import celerite
+from celerite import terms
 
 def savefig(name, fig=None):
     if fig is None:
@@ -27,8 +27,9 @@ t = np.sort(np.random.uniform(0, 5, 100))
 yerr = np.random.uniform(0.1, 0.2, len(t))
 
 true_freq = 2.0
+true_q = 10.0
 kernel = 0.5*kernels.ExpSquaredKernel(2.0)
-kernel += 1.0*kernels.ExpSquaredKernel(100.0) * \
+kernel += 1.0*kernels.ExpSquaredKernel(true_q) * \
     kernels.CosineKernel(period=1.0 / true_freq)
 true_gp = george.GP(kernel)
 K = true_gp.get_matrix(t)
@@ -36,24 +37,33 @@ K[np.diag_indices_from(K)] += yerr**2
 y = np.random.multivariate_normal(np.zeros_like(t), K)
 
 # Set up the fit
-kernel = SHOTerm(0.0, 0.0, 0.0)
-kernel += SHOTerm(0.0, 0.0, 0.0)
-kernel += SHOTerm(0.0, 0.0, 0.0)
-fit_gp = genrp.GP(kernel, log_white_noise=0.0, fit_white_noise=True)
+# bounds = [(-8.0, 8.0), (-8.0, 8.0), (-8.0, 8.0)]
+# kernel = SHOTerm(0.0, -0.5 * np.log(2), 0.0, bounds=bounds)
+# kernel.freeze_parameter("log_Q")
+# kernel += SHOTerm(0.0, 0.0, 0.0, bounds=bounds)
+# kernel += SHOTerm(0.0, 0.0, 0.0, bounds=bounds)
+kernel = terms.RealTerm(0.0, 0.0, bounds=[(-8, 8), (-8, 8)])
+kernel += terms.ComplexTerm(0.0, 0.0, 0.0, 0.0,
+                            bounds=[(-8, 8), (-8, 8), (-8, 8), (-8, 8)])
+fit_gp = celerite.GP(kernel)
+fit_gp.compute(t, yerr)
 
-def log_likelihood(params):
+def nll(params):
     fit_gp.set_parameter_vector(params)
-    fit_gp.compute(t, yerr)
-    return fit_gp.log_likelihood(y)
[…]

### 2017-02-10 paper/ms.tex
sho typo
updating some examples
cleaning up a bit [ci skip]

@@
 where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
-= \sqrt{2/\pi}\,S_0$.
+= \sqrt{2/\pi}\,S_0\,Q^2$.
 The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
@@
 \end{figure}
-
-\begin{enumerate}
-\item Asteroseismic target
-\item Stellar rotation variability
-\item Kepler with simulated transit
-\item AGN
-\end{enumerate}
 
@@
 
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[scale=0.175]{figures/twoterm.pdf}
-\hspace{0.2in}
-\includegraphics[scale=0.85]{figures/colorcode.pdf}
-\end{center}
-\caption{Pictorial description of the extended sparse matrix where $N=10$, $J_0=0$, and $J=2$, following
-the example shown in \citet{Ambikasaran:2015}.}
-\figurelabel{matrix_structure}
-\end{figure}
-
 {\bf Should I give some examples of this?  Some pseudocode?}

### 2017-02-10 celerite/plot_setup.py

@@ -23,14 +23,15 @@ COLORS = dict(
     MODEL_2="#ff7f0e",
 )
 
-def setup():
+def setup(auto=False):
     monkey_patch()
     rcParams["font.size"] = 16
     rcParams["font.family"] = "sans-serif"
     rcParams["font.sans-serif"] = ["Computer Modern Sans"]
     rcParams["text.usetex"] = True
     rcParams["text.latex.preamble"] = r"\usepackage{cmbright}"
-    # rcParams["figure.autolayout"] = True
+    if auto:
+        rcParams["figure.autolayout"] = True
     rcParams["axes.prop_cycle"] = cycler("color", COLOR_CYCLE)
 
 def get_figsize(rows=1, cols=1):

### 2017-02-10 paper/figures/astero_plot.py

@@ -11,9 +11,9 @@ from scipy.ndimage.filters import gaussian_filter
 
 import emcee3
 
-from plot_setup import setup, get_figsize, COLORS
+from celerite.plot_setup import setup, get_figsize, COLORS
 
-setup()
+setup(auto=True)
 np.random.seed(42)
 
 # Helpers

### 2017-02-10 paper/figures/astero_sample.py

@@ -21,9 +21,9 @@ import celerite
 from celerite import modeling
 
 from astero_term import AsteroTerm
-from plot_setup import setup, get_figsize, COLORS
+from celerite.plot_setup import setup, get_figsize, COLORS
 
-setup()
+setup(auto=True)
 
 def format_filename(name):
     base = "astero-{0}-".format(kicid)

### 2017-02-10 paper/figures/benchmark.py

@@ -51,9 +51,9 @@ from scipy.linalg import cho_factor
 fig, ax = plt.subplots(1, 1, figsize=plot_setup.get_figsize(1, 1))
 ax.plot(N, N / N[-1] * 2.0, "k", label="$\mathcal{O}(N)$")
 ax.plot(N, times[:, 0], ".-", label="compute")
-ax.plot(N, times[:, 1], ".-", label="log likelihood")
+ax.plot(N, times[:, 1], ".--", label="log likelihood")
 m = np.isfinite(times[:, 2])
-ax.plot(N[m], times[:, 2][m], ".-", label="Cholesky")
+ax.plot(N[m], times[:, 2][m], ".:", label="Cholesky")
 ax.set_xscale("log")
 ax.set_yscale("log")
 ax.set_xlim(N.min(), N.max())

### 2017-02-10 paper/figures/carma_comp.py

@@ -1,0 +1,81 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+from timer import benchmark
+
+from celerite.solver import Solver, CARMASolver
+from celerite import plot_setup
+plot_setup.setup()
+
+# Simulate a "dataset"
+np.random.seed(42)
+t = np.sort(np.random.rand(2**11))
+yerr = np.random.uniform(0.1, 0.2, len(t))
+y = np.sin(t)
+
+ps = 2 ** np.arange(8)
+js = np.empty_like(ps)
+times = np.empty((len(ps), 2))
+times[:] = np.nan
+
+for i, p in enumerate(ps):
+    arparams = np.random.randn(p)
+    maparams = np.random.randn(p - 1)
+
+    times[i, 0] = benchmark("""
+carma_solver = CARMASolver(0.0, arparams, maparams)
+carma_solver.log_likelihood(t, y, yerr)
+""", "from __main__ import arparams, maparams, t, y, yerr, CARMASolver")
+
+    carma_solver = CARMASolver(0.0, arparams, maparams)
+    params = list(carma_solver.get_celerite_coeffs())
+    params += [t, yerr**2]
+    celerite_solver = Solver()
+
+    # Hack to deal with extra terms
+    js[i] = len(params[0]) + 2*len(params[2])
+    diff = js[i] - p
+    if diff > 0:
+        m = diff // 2
+        if m > 0:
+            for j in range(2, 6):
+                params[j] = params[j][:-m]
+            js[i] = len(params[0]) + 2*len(params[2])
+            diff = js[i] - p
+        if diff > 0:
+            for j in range(2):
+                params[j] = params[j][:-diff]
+            js[i] = len(params[0]) + 2*len(params[2])
+
+    times[i, 1] = benchmark("""
+celerite_solver.compute(*params)
+celerite_solver.dot_solve(y)
+""", "from __main__ import params, celerite_solver, y")
+
+    print(p, js[i], len(params[0]), len(params[2]), times[i])
+
+a = np.log(ps)
+b = np.log(times[:, 0])
+print(np.polyfit(a, b, 1))
+
+m = np.isfinite(times[:, 1])
+a = np.log(js[m])
[…]

### 2017-02-10 paper/figures/kernel_psd.py

@@ -1,49 +1,0 @@
-#!/usr/bin/env python
-# -*- coding: utf-8 -*-
-
-from __future__ import division, print_function
-import numpy as np
-import matplotlib.pyplot as plt
-
-def kernel(terms, tau):
-    k = np.zeros_like(tau)
-    t = np.abs(tau)
-    for a, b, c, d in terms:
-        k += np.exp(-c*t)*(a*np.cos(d*t) + b*np.sin(d*t))
-    return k
-
-def psd(terms, omega):
-    w2 = omega ** 2
-    p = np.zeros_like(w2)
-    for a, b, c, d in terms:
-        p += np.sqrt(2.0 / np.pi) * (
-            (a*c+b*d)*(c**2+d**2)+(a*c-b*d)*w2
-        ) / (
-            w2**2 + 2.0*(c**2-d**2)*w2+(c**2+d**2)**2
-        )
-    return p
-
-tau = np.linspace(0, 5, 5000)
-t = np.linspace(0, 3, 500)
-omega = np.linspace(0, 10, 5000)
-models = [
-    [(1.0, 0.0, 1.0, 0.0), ],
-    [(1.0, 0.0, 1.0, 2.5), ],
-    [(1.0, 0.2, 1.0, 5.0), ],
-    [(0.5, 0.2, 1.0, 5.0), (0.6, 0.0, 1.0, 0.0), ],
-]
-
-fig, axes = plt.subplots(1, 3, figsize=(12, 4))
-
-for terms in models:
-    axes[0].plot(tau, kernel(terms, tau))
-    axes[1].plot(omega, psd(terms, omega))
-
-    np.random.seed(42)
-    K = kernel(terms, t[:, None] - t[None, :])
-    y = np.random.multivariate_normal(np.zeros_like(t), K)
-    axes[2].plot(t, y)
-
-axes[1].set_yscale("log")
-
-fig.savefig("kernel_psd.png")

### 2017-02-10 paper/figures/matrix.py

@@ -89,7 +89,7 @@ c = [(0.89411765336990356, 0.10196078568696976, 0.10980392247438431),
      (0.60000002384185791, 0.60000002384185791, 0.60000002384185791),
      (0.89411765336990356, 0.10196078568696976, 0.10980392247438431)]
 cmap = matplotlib.colors.ListedColormap(["white"] + list(c), name="cmap")
-fig, ax = plt.subplots(1, 1, figsize=get_figsize(2, 2))
+fig, ax = plt.subplots(1, 1, figsize=get_figsize(2.3, 2.3))
 ax.pcolor(full, cmap=cmap, vmin=0, vmax=len(c))
 
 # Plot the edges
@@ -122,4 +122,4 @@ ax.plot([0.25, 0.25, 8.5, 8.5, 0.25],
         full_dim - np.array([0.25, 16.75, 16.75, 0.25, 0.25]),
         "k")
 
-fig.savefig("matrix.pdf")
+fig.savefig("matrix.pdf", bbox_inches="tight")

### 2017-02-10 paper/figures/sho.py

@@ -5,9 +5,9 @@ from __future__ import division, print_function
 
 import numpy as np
 import matplotlib.pyplot as plt
-from plot_setup import setup, HORIZONTAL_FIGSIZE
+from celerite.plot_setup import setup, HORIZONTAL_FIGSIZE
 
-setup()
+setup(auto=True)
 
 def sho_psd(Q, x):
     x2 = x*x

### 2017-02-10 paper/figures/transit_sample.py

@@ -14,10 +14,10 @@ import matplotlib.pyplot as plt
 from scipy.optimize import minimize
 
 import celerite
-from plot_setup import setup, get_figsize
+from celerite.plot_setup import setup, get_figsize
 from transit_model import RotationTerm, TransitModel
 
-setup()
+setup(auto=True)
 np.random.seed(42)
 
 data, hdr = fitsio.read("data/kplr001430163-2013011073258_llc.fits",

### 2017-02-11 celerite/__init__.py
allow runtime LAPACK choices

@@ -8,10 +8,12 @@ except NameError:
     __CELERITE_SETUP__ = False
 
 if not __CELERITE_SETUP__:
-    __all__ = ["Solver", "GP", "terms"]
+    __all__ = ["Solver", "GP", "terms",
+               "__library_version__", "__with_lapack__"]
 
-    from . import terms
+    from . import terms, solver
     from .celerite import GP
-    from .solver import Solver, get_library_version
+    from .solver import Solver
 
-    __library_version__ = get_library_version()
+    __library_version__ = solver.get_library_version()
+    __with_lapack__ = solver.with_lapack()

### 2017-02-11 celerite/celerite.py

@@ -2,9 +2,10 @@
 
 from __future__ import division, print_function
 import math
+import logging
 import numpy as np
 
-from .solver import Solver
+from .solver import Solver, with_lapack
 from .modeling import ModelSet, ConstantModel
 
 __all__ = ["GP"]
@@ -36,11 +37,24 @@ class GP(ModelSet):
     def __init__(self,
                  kernel,
                  mean=0.0, fit_mean=False,
-                 log_white_noise=-float("inf"), fit_white_noise=False):
+                 log_white_noise=-float("inf"), fit_white_noise=False,
+                 use_lapack=None):
         self.solver = None
         self._computed = False
         self._t = None
         self._y_var = None
+
+        if use_lapack is None:
+            if with_lapack():
+                coeffs = kernel.coefficients
+                nterms = len(coeffs[0]) + 2 * len(coeffs[2])
+                use_lapack = (nterms >= 8)
+            else:
+                use_lapack = False
+        if use_lapack and not with_lapack():
+            use_lapack = False
+            logging.warn("celerite was not compiled with lapack support")
+        self._use_lapack = use_lapack
 
         # Build up a list of models for the ModelSet
         models = [("kernel", kernel)]
@@ -131,7 +145,7 @@ class GP(ModelSet):
         self._yerr = np.empty_like(self._t)
         self._yerr[:] = yerr
         if self.solver is None:
-            self.solver = Solver()
+            self.solver = Solver(self._use_lapack)
         (alpha_real, beta_real, alpha_complex_real, alpha_complex_imag,
          beta_complex_real, beta_complex_imag) = self.kernel.coefficients
         self.solver.compute(

### 2017-02-11 celerite/solver.cpp

@@ -17,13 +17,16 @@ namespace py = pybind11;
 //
 class PicklableBandSolver : public celerite::solver::BandSolver<double> {
 public:
+  PicklableBandSolver (bool use_lapack = false) : celerite::solver::BandSolver<double>(use_lapack) {};
+
   auto serialize () const {
-    return std::make_tuple(this->computed_, this->n_, this->p_real_, this->p_complex_, this->log_det_,
+    return std::make_tuple(this->use_lapack_, this->computed_, this->n_, this->p_real_, this->p_complex_, this->log_det_,
                            this->a_, this->al_, this->ipiv_);
   };
 
-  void deserialize (bool computed, int n, int p_real, int p_complex, double log_det,
-                    Eigen::MatrixXd a, Eigen::MatrixXd al, Eigen::VectorXi ipiv) {
+  void deserialize (bool use_lapack, bool computed, int n, int p_real, int p_complex,
+                    double log_det, Eigen::MatrixXd a, Eigen::MatrixXd al, Eigen::VectorXi ipiv) {
+    this->use_lapack_ = use_lapack;
     this->computed_ = computed;
     this->n_ = n;
     this->p_real_ = p_real;
@@ -48,6 +51,13 @@ need to call these directly. This interface was built using `pybind11
 )delim");
 
   m.def("get_library_version", []() { return CELERITE_VERSION_STRING; }, "The version of the linked C++ library");
+  m.def("with_lapack", []() {
+#ifdef WITH_LAPACK
+    return true;
+#else
+    return false;
+#endif
+  }, "Was celerite compiled with LAPACK support");
 
   m.def("get_kernel_value",
     [](
@@ -169,6 +179,7 @@ method must always be called first.
 
 )delim");
   solver.def(py::init<>());
+  solver.def(py::init<bool>());
 
   solver.def("compute", [](PicklableBandSolver& solver,
       const Eigen::VectorXd& alpha_real,
@@ -335,20 +346,20 @@ Returns:
   });
 
   solver.def("__setstate__", [](PicklableBandSolver& solver, py::tuple t) {
-    if (t.size() != 8)
-    throw std::runtime_error("Invalid state!");
+    if (t.size() != 9) throw std::runtime_error("Invalid state!");
 
-    new (&solver) PicklableBandSolver();
[…]

### 2017-02-11 cpp/include/celerite/exceptions.h

@@ -23,6 +23,12 @@ struct dimension_mismatch : public std::exception {
   }
 };
 
+struct no_lapack : public std::exception {
+  const char * what () const throw () {
+    return "celerite was not compiled with LAPACK support";
+  }
+};
+
 }; // namespace celerite
 
 #endif // _CELERITE_EXCEPTIONS_H_

### 2017-02-11 cpp/include/celerite/solver/band.h

@@ -10,10 +10,6 @@
 #include "celerite/banded.h"
 #include "celerite/solver/solver.h"
 
-#ifndef MIN_LAPACK_WIDTH
-#define MIN_LAPACK_WIDTH 8
-#endif
-
 namespace celerite {
 namespace solver {
 
@@ -36,7 +32,7 @@ namespace solver {
 template <typename T>
 class BandSolver : public Solver<T> {
 public:
-  BandSolver () : Solver<T>() {};
+  BandSolver (bool use_lapack = false) : Solver<T>(), use_lapack_(use_lapack) {};
   ~BandSolver () {};
 
   void build_matrix (
@@ -121,6 +117,7 @@ public:
   using Solver<T>::solve;
 
 protected:
+  bool use_lapack_;
   Eigen::Matrix<T, Eigen::Dynamic, Eigen::Dynamic> a_;
   Eigen::Matrix<T, Eigen::Dynamic, Eigen::Dynamic> al_;
   Eigen::VectorXi ipiv_;
@@ -315,7 +312,9 @@ int BandSolver<T>::compute (
 
   int offset_factor = 1;
 #ifdef WITH_LAPACK
-  if (width >= MIN_LAPACK_WIDTH) offset_factor = 2;
+  if (use_lapack_) offset_factor = 2;
+#else
+  if (use_lapack_) throw no_lapack();
 #endif
 
   // Build the extended matrix.
@@ -328,7 +327,7 @@ int BandSolver<T>::compute (
 
   // Reshape the working arrays
 #ifdef WITH_LAPACK
-  if (width < MIN_LAPACK_WIDTH)
+  if (!use_lapack_)
 #endif
   al_.resize(width, dim_ext);
   ipiv_.resize(dim_ext);
@@ -336,7 +335,7 @@ int BandSolver<T>::compute (
   // Factorize the sparse matrix
   int nothing;
 #ifdef WITH_LAPACK
-  if (width >= MIN_LAPACK_WIDTH)
+  if (use_lapack_)
     band_factorize(dim_ext, width, width, a_, ipiv_);
   else
 #endif
@@ -345,14 +344,13 @@ int BandSolver<T>::compute (
 // Compute the determinant
   T ld = T(0.0);
 #ifdef WITH_LAPACK
-  if (width >= MIN_LAPACK_WIDTH)
+  if (use_lapack_)
     for (int i = 0; i < dim_ext; ++i) ld += log(abs(a_(2*width, i)));
[…]

### 2017-02-11 tests/test_celerite.py

@@ -10,15 +10,24 @@ try:
 except ImportError:
     import pickle
 
+import celerite
 from celerite import Solver, GP, terms
-from celerite.solver import get_kernel_value, CARMASolver
+from celerite.solver import get_kernel_value, CARMASolver, with_lapack
 
 __all__ = ["test_carma", "test_log_determinant", "test_solve", "test_dot",
            "test_pickle", "test_build_gp", "test_log_likelihood",
            "test_predict", "test_nyquist_singularity"]
 
 
-def test_carma(seed=42):
+lapack_switch = pytest.mark.parametrize(
+    "use_lapack",
+    [False,
+     pytest.mark.skipif(celerite.__with_lapack__ is False, True,
+                        reason="LAPACK support not compiled")]
+)
+
+@lapack_switch
+def test_carma(use_lapack, seed=42):
     np.random.seed(seed)
     t = np.sort(np.random.uniform(0, 5, 100))
     yerr = 0.1 + np.zeros_like(t)
@@ -29,7 +38,7 @@ def test_carma(seed=42):
     carma_ll = carma_solver.log_likelihood(t, y, yerr)
     params = carma_solver.get_celerite_coeffs()
 
-    solver = Solver()
+    solver = Solver(use_lapack)
     solver.compute(
         params[0], params[1], params[2], params[3], params[4], params[5],
         t, yerr**2
@@ -37,18 +46,17 @@ def test_carma(seed=42):
     celerite_ll = -0.5*(
         solver.dot_solve(y) + solver.log_determinant() + len(t)*np.log(2*np.pi)
     )
-
     assert np.allclose(carma_ll, celerite_ll)
 
 
 def _test_log_determinant(alpha_real, beta_real, alpha_complex_real,
                           alpha_complex_imag, beta_complex_real,
-                          beta_complex_imag, seed=42):
+                          beta_complex_imag, use_lapack, seed=42):
     np.random.seed(seed)
     t = np.sort(np.random.rand(5))
     diag = np.random.uniform(0.1, 0.5, len(t))
 
-    solver = Solver()
+    solver = Solver(use_lapack)
     solver.compute(
         alpha_real, beta_real, alpha_complex_real, alpha_complex_imag,
[…]

### 2017-02-13 paper/ms.tex
allow runtime LAPACK choices
fixing #60
adding @ruthangus to author list

@@
 \setcounter{affilcounter}{1}
-\edef \uw {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\uw}       {Astronomy Department, University of Washington,
-                          Seattle, WA, 98195, USA}
 
@@
 
+\edef \uw {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\uw}{Astronomy Department, University of Washington,
+                   Seattle, WA}
+
+\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\simons}{Simons Fellow}
+
+\edef \columbia {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\columbia}{Department of Astronomy, Columbia University,
+                         New York, NY}
+
 \author{%
-    Daniel~Foreman-Mackey\altaffilmark{\uw,\sagan} and
-    Eric~Agol\altaffilmark{\uw}
+    Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw},
+    Eric~Agol\altaffilmark{\uw}, and
+    Ruth~Angus\altaffilmark{\simons,\columbia}
 }
@@
 Now, if we introduce the variables
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{algo-first}
     u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
@@
 where $u_{N} = 0$, and
-\begin{eqnarray}\eqlabel{algo-first}
+\begin{eqnarray}
     g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
@@
 Banded vs.\ sparse.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/benchmark.pdf}
+\caption{Benchmark\figurelabel{benchmark}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/carma_comp.pdf}
+\caption{Benchmark\figurelabel{carma-comp}}
+\end{center}
+\end{figure}
 
@@
 Sivaram Ambikasaran,
-Ruth Angus,
 Will Farr,

### 2017-02-13 celerite/__init__.py
adding LAPACK_VARIANT option to build

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.1"
+__version__ = "0.1.2.dev0"
 
 try:
     __CELERITE_SETUP__

### 2017-02-13 celerite/build.py

@@ -7,6 +7,7 @@ import re
 import sys
 import logging
 import tempfile
+from numpy.distutils.system_info import get_info
 
 import setuptools
 from setuptools.command.build_ext import build_ext as _build_ext
@@ -169,21 +170,19 @@ class build_ext(_build_ext):
             ext.extra_compile_args = opts
 
         # Link to numpy's LAPACK if available
-        info = None
-        for ext in self.extensions:
-            if not any(k[0] == "WITH_LAPACK" for k in ext.define_macros):
-                continue
-            if info is None:
-                import pprint
-                import numpy.__config__ as npconf
-                info = npconf.get_info("blas_opt_info")
-                print("Found LAPACK linking info:")
-                pprint.pprint(info)
-            for k, v in info.items():
-                try:
-                    setattr(ext, k, getattr(ext, k) + v)
-                except TypeError:
-                    continue
+        variant = os.environ.get("LAPACK_VARIANT", None)
+        if variant is not None and variant.lower() != "none":
+            info = get_info(variant)
+            if not len(info):
+                logging.warn("LAPACK info for variant '{0}' not found")
+                info = get_info("blas_opt")
+            for ext in self.extensions:
+                for k, v in info.items():
+                    try:
+                        setattr(ext, k, getattr(ext, k) + v)
+                    except TypeError:
+                        continue
+                ext.define_macros += [("WITH_LAPACK", None)]
 
         # Run the standard build procedure.
         _build_ext.build_extension(self, ext)

### 2017-02-15 paper/ms.tex
adding LAPACK_VARIANT option to build
adding sparse solver back in

@@
 In this case, the inverse will be dense but \eq{gp-likelihood} can still be
-evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number of
-components in the mixture and $N$ is still the number of data points.
+evaluated in $\mathcal{O}(J^2\,N)$ operations \citep{Anderson:1999} where $J$
+is the number of components in the mixture and $N$ is still the number of data
+points.
 
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{figures/carma_comp.pdf}
-\caption{Benchmark\figurelabel{carma-comp}}
+\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_default.pdf}
+\caption{Benchmark\figurelabel{benchmark-all}}
 \end{center}

### 2017-02-15 celerite/solver.cpp

@@ -177,6 +177,9 @@ Returns:
 
 )delim");
 
+  //
+  // ------ BAND ------
+  //
   py::class_<PicklableBandSolver> solver(m, "Solver", R"delim(
 A thin wrapper around the C++ BandSolver class
 
@@ -370,7 +373,9 @@ Returns:
     );
   });
 
-  // CARMA
+  //
+  // ------ CARMA ------
+  //
   py::class_<celerite::carma::CARMASolver> carma_solver(m, "CARMASolver", R"delim(
 A thin wrapper around the C++ CARMASolver class
 )delim");
@@ -388,6 +393,126 @@ A thin wrapper around the C++ CARMASolver class
     );
   });
 
+
+  //
+  // ------ SPARSE ------
+  //
+  py::class_<celerite::solver::SparseSolver<double> > sparse_solver(m, "SparseSolver", R"delim(
+A thin wrapper around the C++ SparseSolver class
+)delim");
+  sparse_solver.def(py::init<>());
+
+  sparse_solver.def("compute", [](celerite::solver::SparseSolver<double>& solver,
+      const Eigen::VectorXd& alpha_real,
+      const Eigen::VectorXd& beta_real,
+      const Eigen::VectorXd& alpha_complex_real,
+      const Eigen::VectorXd& alpha_complex_imag,
+      const Eigen::VectorXd& beta_complex_real,
+      const Eigen::VectorXd& beta_complex_imag,
+      const Eigen::VectorXd& x,
+      const Eigen::VectorXd& diag) {
+    return solver.compute(
+      alpha_real,
+      beta_real,
+      alpha_complex_real,
+      alpha_complex_imag,
+      beta_complex_real,
+      beta_complex_imag,
+      x,
+      diag
+    );
+  },
+  R"delim(
+Assemble the extended matrix and perform the banded LU decomposition
+
+Args:
+    alpha_real (array[j_real]): The coefficients of the real terms.
+    beta_real (array[j_real]): The exponents of the real terms.
+    alpha_complex_real (array[j_complex]): The real part of the
+        coefficients of the complex terms.
+    alpha_complex_imag (array[j_complex]): The imaginary part of the
+        coefficients of the complex terms.
+    beta_complex_real (array[j_complex]): The real part of the
+        exponents of the complex terms.
+    beta_complex_imag (array[j_complex]): The imaginary part of the
+        exponents of the complex terms.
+    x (array[n]): The _sorted_ array of input coordinates.
[…]

### 2017-02-15 cpp/include/celerite/solver/sparse.h

@@ -1,0 +1,322 @@
+#ifndef _CELERITE_SOLVER_SPARSE_H_
+#define _CELERITE_SOLVER_SPARSE_H_
+
+#include <cmath>
+#include <Eigen/Core>
+#include <Eigen/Sparse>
+
+#include "celerite/utils.h"
+#include "celerite/exceptions.h"
+#include "celerite/solver/solver.h"
+
+namespace celerite {
+namespace solver {
+
+#define BLOCKSIZE_BASE                              \
+  int width = p_real + 2 * p_complex + 2,           \
+      block_size = 2 * p_real + 4 * p_complex + 1,  \
+      dim_ext = block_size * (n - 1) + 1;           \
+  if (p_complex == 0) width = p_real + 1;
+
+#define BLOCKSIZE                                   \
+  int p_real = this->p_real_,                       \
+      p_complex = this->p_complex_,                 \
+      n = this->n_;                                 \
+  BLOCKSIZE_BASE
+
+/// The class provides an alternative solver that uses Eigen's SparseLU
+///
+/// The `celerite::solver::SparseSolver::compute` method must be called before
+/// most of the other methods.
+template <typename T>
+class SparseSolver : public Solver<T> {
+public:
+  SparseSolver () : Solver<T>() {};
+  ~SparseSolver () {};
+
+  void build_matrix (
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& alpha_real,
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& beta_real,
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& alpha_complex_real,
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& alpha_complex_imag,
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& beta_complex_real,
+    const Eigen::Matrix<T, Eigen::Dynamic, 1>& beta_complex_imag,
+    const Eigen::VectorXd& x,
+    Eigen::SparseMatrix<T>& A
+  );
+
+  /// Compute the extended matrix and perform the SparseLU decomposition
+  ///
+  /// @param alpha_real The coefficients of the real terms.
+  /// @param beta_real The exponents of the real terms.
[…]

### 2017-02-15 cpp/src/test_solvers.cc

@@ -3,6 +3,7 @@
 
 #include "celerite/solver/direct.h"
 #include "celerite/solver/band.h"
+#include "celerite/solver/sparse.h"
 
 #define DO_TEST(NAME, VAR1, VAR2)                            \
 {                                                            \
@@ -12,6 +13,7 @@
   delta = std::abs(base - comp);                             \
   if (delta > 1e-10) {                                       \
     std::cerr << "Test failed: '" << #NAME << "' - error: " << delta << std::endl; \
+    return 1;                                                \
   } else                                                     \
     std::cerr << "Test passed: '" << #NAME << "' - error: " << delta << std::endl; \
 }
@@ -52,28 +54,37 @@ int main (int argc, char* argv[])
 
   // The times need to be sorted.
   std::sort(x.data(), x.data() + x.size());
-  std::cerr << x << "\n\n";
 
   // Compute the y values.
   y = sin(x.array());
 
   celerite::solver::DirectSolver<double> direct;
   celerite::solver::BandSolver<double> band;
+  celerite::solver::SparseSolver<double> sparse;
 
   direct.compute(alpha_real, beta_real, x, yerr2);
   band.compute(alpha_real, beta_real, x, yerr2);
+  sparse.compute(alpha_real, beta_real, x, yerr2);
   DO_TEST(band_real_log_det, direct.log_determinant(), band.log_determinant())
+  DO_TEST(sparse_real_log_det, sparse.log_determinant(), band.log_determinant())
   DO_TEST(band_real_dot_solve, direct.dot_solve(y), band.dot_solve(y))
+  DO_TEST(sparse_real_dot_solve, band.dot_solve(y), sparse.dot_solve(y))
 
   band.compute(alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag, x, yerr2);
   direct.compute(alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag, x, yerr2);
-  DO_TEST(band_complex_dot_solve, direct.dot_solve(y), band.dot_solve(y))
+  sparse.compute(alpha_complex_real, alpha_complex_imag, beta_complex_real, beta_complex_imag, x, yerr2);
   DO_TEST(band_complex_log_det, direct.log_determinant(), band.log_determinant())
[…]

### 2017-02-16 paper/ms.tex
some edits to the paper

@@
 
-%
-% New Outline:
-% 1. Introduction
-% 2. Gaussian processes
-%   - GPs in general
-%   - RP method
-%   - Ambikasaran method
-%   - Generalization
-%   - Relation to stochastically-driven, damped simple harmonic oscillator
-% 3. Implementation and performance
-%   - Implementation considerations: band solver vs. sparse, parameterization
-%   - Ensuring positive-definite covariance matrix: Sturm's theorem
-%   - Benchmark numbers
-%   - Scaling with N and M
-%   - Description of APIs
-% 4. Examples
-%   - Simulated data: a simple exact celerite simulation, a celerite simulation with
-%     structure, a few standard kernels, a AGW kernel
-%   - An AGN light curve
-%   - An RR Lyrae
-%   - A kepler light curve
-%   - An asteroseismic target
-% 5. Comparisons to other methods
-% 6. Summary
-%   - Possible applications
-%
-
 \section{Gaussian processes}
@@
 
-In the following Section, we present a method of improving this scaling that
-we call the \celerite\ method.
-The \celerite\ method requires using a specific model for the covariance
-$k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
-The method can only be applied to one-dimensional datasets.
+In the following Section, we present a method of substantially improving this
+scaling in many circumstances.
+We call our method and its implementations \celerite.\footnote{The name
+\celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
+of light in a vacuum.} The \celerite\ method requires using a specific model
+for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has
+several limitations but, in subsequent sections, we demonstrate that it can be
+used to increase the computational efficiency of many astronomical data
+analysis problems.
+The main limitation of this method is that it can only be applied to
+one-dimensional datasets.
 When we say ``one-dimensional'' here, it means that the \emph{input
@@
 To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
-of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
-function is
+of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
+operations when the covariance function is given by
 \begin{eqnarray}\eqlabel{kernel-simple}
@@
 $\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
-The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
+The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
 the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
@@
 In this case, the inverse will be dense but \eq{gp-likelihood} can still be
-evaluated in $\mathcal{O}(J^2\,N)$ operations \citep{Anderson:1999} where $J$
-is the number of components in the mixture and $N$ is still the number of data
-points.
+evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
+in the mixture and $N$ is still the number of data points.
 
@@
 and, for this function, \eq{gp-likelihood} can still be solved with
-$\mathcal{O}(J^2\,N)$ operations.
+$\mathcal{O}(N)$ operations.
 The details of this method and a few implementation considerations are
-discussed in \app{implementation}.
-
-\eq{celerite-kernel-complex} can also be expressed as
+discussed in the following \sectionname\ but we will first discuss some
+properties of this covariance function.
+
+By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
+and cosine functions, we can see the autocorrelation structure is defined by a
+mixture of quasiperiodic oscillators
 \begin{eqnarray}\eqlabel{celerite-kernel}
@@
 \end{eqnarray}
-The Fourier transform of this covariance function is the power spectral
-density (PSD) of the model and it is given by
+The Fourier transform\footnote{Here and throughout we have defined the Fourier
+transform of the function $f(t)$ as $F(\omega)=(2\,\pi)^{-1/2}\,
+\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
+function is the power spectral density (PSD) of the process and it is given by
 \begin{eqnarray}\eqlabel{celerite-psd}
@@
 In this case, we'll assume that we have three data points
-$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
+$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
 variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
@@
 To demonstrate motivate this method, we will write out the full system of
-equations that we must solve to apply the inverse of $K$.
+equations that we must solve to apply the inverse of $K$ in order to compute
+the first term of \eq{gp-likelihood}.
 In matrix notation, this can be written as
@@
 This means that we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(J^2\,N)$ operations.
+marginalized likelihood in $\mathcal{O}(N)$ operations.
 
@@
 equation in the ``extended'' space and, as before, this can be used to solve
-for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(J^2\,N)$ operations.
+for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(N)$ operations.
 \Figure{matrix} shows a pictorial representation of the sparsity pattern of
@@
 for band matrices.
-This decomposition can be used to compute the determinant of $K$ and to solve
-$K^{-1}\,\bvec{y}$.
+This decomposition can be used to compute the determinant of $K$, solve
+$K^{-1}\,\bvec{y}$, and subsequently calculate the marginalized likelihood in
+\eq{gp-likelihood}.
 
@@
 
-\subsection{Implementation considerations}
-
-Solvers.
-Banded vs.\ sparse.
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{figures/benchmark.pdf}
-\caption{Benchmark\figurelabel{benchmark}}
-\end{center}
-\end{figure}
+\subsection{Implementation considerations \& scaling}
+
+The extended system defined in the previous section is sparse with fewer than
+a few percent non-zero entries and band structure.
+In this \sectionname, we empirically investigate the performance and scaling
+of three different algorithms for solving this extended systems:
+
+\begin{enumerate}
+
+\item A simple algorithm for computing the LU decomposition for banded
+    matrices using Gaussian elimination \citep{Press:1992, Press:2007},
+
+\item The general banded LU decomposition implementation from
+    LAPACK\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs} methods
+        from LAPACK.} \citep{Anderson:1999} using optimized BLAS routines, and
+
+\item A general sparse LU solver that exploits the sparsity but not the band
+    structure \citep{Guennebaud:2010}.
+
+\end{enumerate}
+
+NNZ: $(10\,J + 1)\,N - 4\,J + 1$
 
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_default.pdf}
-\caption{Benchmark\figurelabel{benchmark-all}}
+\caption{Benchmark\figurelabel{benchmark}}
 \end{center}
@@
 
-\subsection{Benchmarks \& scaling}
-
-\subsection{Parameterization \& API}
-
 \section{celerite as a model of stellar variations}
 
-Another special case of the \celerite\ model of great physical interest is a
+\dfmtodo{Add more intro to this section.}
+
+A special case of the \celerite\ model of great physical interest is a
 stochastically-driven simple harmonic oscillator.
@@
 \item a full celerite process
-\item Some standard kernels: Matern, Exp-squared, etc.
-\item A KISS-GP process
+\item some sort of QP oscillation
 \end{enumerate}
@@
 % derivatives of the likelihood function.
-
-% The plan of the paper is...
-
-% \section{Outline}
-% \begin{enumerate}
-%\item Non-parametric modeling of time series with correlated noise using Gaussian processes
-%\item Assumptions:\\
-%  a). Stationary noise (although this can be broken).\\
-%  b). Gaussian
-%\item Problem: can't handle large datasets - e.g. entire Kepler long-cadence; Spitzer light
-%   curves; Solar curves; short-cadence.
-%\item Solution(s):\\
-%  a). HODLR \citep{Ambikasaran:2013,Ambikasaran:2016}\\
-%  b). CARMA (Lorentzian power spectra, but with limitations) \citep{Kelly:2014}\\
-%  c). Tri-diagonal (Press \& Rybicki) \citep{Rybicki:1995}
-%  d).
-KISS-GP \citep{Wilson:2015} %Kernel Interpolation for Scalable Structured Gaussian Processes (KISS-GP) Andrew Wilson, Hannes Nickisch Proceedings of The 32nd International Conference on Machine Learning, pp.
-1775-1784, 2015
-%\item Problem: HODLR is tricky to use; CARMA requires certain relations between
-%  coefficients of Lorentzians (right?); P\&R only works for up to 2 components + white noise
-%\item Solution: Ambikasaran \citep{Ambikasaran:2015} shows how sum of exponential kernels can be solved in order
-%   $O(N*p^2)$, where p is number of Lorentzians.
-%  - Sum of Lorentzians is a good description of stellar variability:
-%     - granulation has expoential (or sums of exponential) noise
-%     - asteroseismic obeys sums of Lorentzians
-%     - periodic variability due to spots/plages/etc. can also be
-%       approximated by Lorentzian(s)
-%\item Generalized Ambikasarn for complex coefficients (Hermitian).
-% \item Performance:\\
-%   a). Show $O(N*p^2)$ scaling\\
-%   b). Show comparison to HODLR.\\
-%   c). Show comparison to wavelet (ala Carter \& Winn).\\
-%   d). Compare with CARMA Kallman filter(?).
-% \item Various methods/formulae:\\
-%   a). Likelihood computation.\\
-%   b). Generating GP with particular covariance [yet to solve]. [ ]\\
-%   c). Inferring particular components.\\
-%   d). Computing derivatives of GPs (as used in RV methods).\\
-%   e). Forecasting/interpolating.\\
-%   f). Inferring confidence intervals.\\
-%   g). Derivative of likelihood function with respect to kernel parameters.\\
-%   h). Non-stationary GP (can the coefficients vary with time?).\\
-%   i). Multi-band (each element of extended matrix becomes a matrix).
-%   j). Approximating various kernels: Matern, exp(sin$^2$), Gaussian, etc.
-% \item Generalization to non-stationary noise.
-% \item Example application: TYC 3559\\
-%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
-%   b). Analysis of solar data.
-%   c). CO2 Mauna Kea
-%   d). TBD Computing derivatives of GPs (as used in RV methods).\\
-%   e). TBD Forecasting/interpolating.\\
-%   f). TBD Inferring confidence intervals.\\
-%   g). Autodiff: Derivative of likelihood function with respect to kernel parameters.\\
-%   h). TBD Non-stationary GP (can the coefficients vary with time?).\\
-%   i). TBD Multi-band (each element of extended matrix becomes a matrix).
-
-% \item Example application: TYC 3559\\
-%   a). Entire long-cadence 17-quarter Kepler dataset (~65k data points).\\
-%   b). Simulated datasets - hyperparameter/PSD inference.\\
-%   c). Analysis of solar data?
-
-% \item Possible future applications:\\
-%   a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{Wilson:2013}, but
-%       with Lorentzians, not Gaussians.\\
-%   b). Multi-waveband GPs.\\
-%   c). RV interpolation.\\
-%   d). Non-parametric phase functions.\\
-%   e). Better calibration of the flicker log(g) relation.\\
-%   f). Measurement of rotational periods - gyrochronology.\\
-%   g). Doppler shifts in oscillating EBs.
-
-% \item Appendix:
-%   a). Description/summary of mathematics.\\
-%   b). Description of code.
-% \end{enumerate}
-
-\section{Generalized Press-Rybicki}
-
-The general problem to be solved for Gaussian Process analysis of stellar time series is to evaluate
-the likelihood function,
-\begin{equation}
-{\cal{L}} = \ln{p(\mathbf{y}\vert {\mathbf{t},\bvec{\theta}},\bvec{\alpha})} = -\frac{1}{2}\ln{\vert \mathbf{K}(\bvec{\alpha}) \vert}
-- \frac{N_{time}}{2} \ln{(2\pi)}
-    - \frac{1}{2} {\mathbf r} (\bvec{\theta} ) ^T \mathbf{K}(\bvec{\alpha})^{-1} \mathbf{r}(\bvec{\theta}),
-\end{equation}
-where $\mathbf{y}$ are the measured time series (for example fluxes or radial velocities) at times
-$\mathbf{t}$, while the data are modeled with a function $m(\mathbf{t};\bvec{\theta})$ (for example,
-a transit model or Keplerian orbital model) which leaves residuals $\mathbf{r}(\bvec{\theta}) =
-\mathbf{y} - m(\mathbf{t};\bvec{\theta})$. % Note: I'm using the same notation as the peerless project. -EA
-The vector $\bvec{\alpha}$ specifies the parameters of the kernel, while the vector $\bvec{\theta}$
-specifies the parameters of the physical model.  In addition to this likelihood, a prior may be placed on
-the values of these model parameters (either the physical or kernel); the log of this prior may be
-added to this log likelihood function.  Note that the kernel
-parameters may also have a physical meaning, but they describe a noisy process, and thus affect
-the correlated variability of stellar noise rather than a deterministic model as encapsulated in
-$m(\mathbf{t};\bvec{\theta})$.
-
-%The kernel is modeled as a sum of damped harmonics,
-%\begin{equation}
-%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^J a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
-%\end{equation}
-%where $\tau > 0$ is a time-offset between data points, $c_j$ represents the decay
-%in correlation with time and $d_j$ represents oscillations of period $P = 2\pi/d_j$, and
-%$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
-%the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
-%time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
-%of the measured times if the correlated components are zero.
-
-%Now, the damped harmonic kernel components may be rewritten as:
-%\begin{equation}
-% a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
-%\end{equation}
-%where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
-%matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
-%The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
-%but otherwise the mathematical form of the extended matrix is identical, but with twice as many
-%exponential kernel components as the real case.
-
-Although the complex kernel is a straightforward extension to the \texttt{celerite} formalism, there is a disadvantage
-in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
-arithmetic.  In addition, it turns out that the complex conjugate component, $(c_j+id_j)^* = c_j - id_j$, in
-the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
-component $c_j + i d_j$;  thus the computation is redundant.  Finally, complex
-arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
-which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
-to compute this explicitly!).
-
-Consequently, we use the fact that $\exp{(id_j\tau)} =  \cos{(d_j\tau)} + i \sin{(d_j \tau)}$ to
-rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
-may be replaced by a $2\times 2$ anti-symmetric matrix,
-\begin{equation}
-\left(\begin{tabular}{ll}
-0 & 1 \\
--1 & 0
-\end{tabular}\right),
-\end{equation}
-while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$,
-so that $\cos{(d_j\tau)} + i \sin{(d_j\tau)}$ may be rewritten as the rotation matrix
-\begin{equation}
-\left(\begin{tabular}{ll}
-$\cos{(d_j\tau)}$ & $\sin{(d_j\tau)}$ \\
-$-\sin{(d_j\tau)}$ & $\cos{(d_j \tau)}$
-\end{tabular}\right).
-\end{equation}
-The algebraic properties of these matrices are identical to the complex numbers, but only involve
-real components.
-We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex
-equations, resulting in a real extended matrix that is the same size as the complex matrix.
-This transformation leads to about a factor of two in computational savings, but has the
-additional advantage of allowing the automatic differentiation due to the use of real arithmetic.
 
@@
 
-%\subsection{Derivative of likelihood with respect to Kernel parameters}
-%
-%The derivative of the likelihood can be computed in $O(N)$ operations as follows.
-%
-%The derivative of the log likelihood with respect to model parameters ${\bf x}$ is given by:
-%\begin{equation}
-%    \frac{\partial \ln{\cal L}}{\partial x_i} = (\ln{\cal L})^\prime = -\frac{1}{2}\sum_i \frac{K_{ii}^\prime}{K_{ii}}
-%    - \frac{1}{2} {\bf y}^T {\bf K}^{-1} {\bf K}^\prime {\bf K}^{-1} {\bf y},
-%\end{equation}
-%where $^\prime$ means the derivative wrt $x_i$.
-%The kernel may be written as:
-%\begin{equation}
-%    K_{ij}({\bf \alpha},{\bf \beta}) = \left\{
-%    \begin{array}{cc}
-%       w_i + \sum_{l=1}^J \alpha_l & j = i \\
-%       \sum_{l=1}^J \alpha_l \exp{(-\beta_l |t_i-t_j|)}  & j \ne i
-%    \end{array}
-%    \right.
-%\end{equation}
-%Since this equation is the sum of $J$ semi-separable matrices, then applying the product rule to compute the derivative will give a single semi-separable matrix for the derivative with respect to each kernel parameter.
-%
-%Now, letting ${\bf x} = (w,{\bf \alpha},{\bf \beta})$, then the derivative may be written as:
-%\begin{eqnarray}
-%    \frac{\partial K_{ii}}{\partial w} &=& \delta_{ij}\\
-%    \frac{\partial K_{ij}}{\partial \alpha_k} &=&
-%     \left\{
-%    \begin{array}{cc}
-%       1 & j = i \\
-%        \exp{(-\beta_k |t_i-t_j|)}  & j \ne i
-%    \end{array}
-%    \right.\\
-%    \frac{\partial K_{ij}}{\partial \beta_k} &=&
-%     \left\{
-%    \begin{array}{cc}
-%       0  & j = i \\
-%        -\alpha_k|t_i-t_j|\exp{(-\beta_k |t_i-t_j|)}  & j \ne i
-%    \end{array}
-%    \right.
-%\end{eqnarray}
-%The first term in the derivative of the likelihood can be computed in $O(N)$
-%evaluations, while the second term can be computed in three steps:
-%\begin{itemize}
-%\item First, solve ${\bf z_1} = {\bf K}^{-1}{\bf y}$ with the GRP solver.
-%\item Next, create an e
-%\end{itemize}
-%Writing this term out:
-
 \section{Summary}
@@
 
-\vspace{1.5em}
-All of the code used in this project is available from
-\url{https://github.com/dfm/celerite} under the MIT open-source software
-license.
-This code (plus some dependencies) can be run to re-generate all of the
-figures and results in this paper; this version of the paper was generated
-with git commit \texttt{\githash} (\gitdate).
-
-\section{Solving for log likelihood with extended matrix}
-
-Here we summarize \citet{Ambikasaran:2015}, and then point to appendix
-for how to do this with oscillating kernels.
-
-\section{Training}
-
-May use derivative of likelihood + BFGS-B to optimize kernel parameters.
-
-Q: How do we determine how many kernels to use?
-
-\section{Discussion}
-
 There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
@@
 multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.
-
-\subsection{Application to stellar variability, transiting exoplanets}
 
@@
 
+\vspace{1.5em}
+All of the code used in this project is available from
+\url{https://github.com/dfm/celerite} under the MIT open-source software
+license.
+This code (plus some dependencies) can be run to re-generate all of the
+figures and results in this paper; this version of the paper was generated
+with git commit \texttt{\githash} (\gitdate).
+
 \acknowledgments
@@
 \software{%
-    % \project{ceres} \citep{Agarwal:2016},
-    % \project{corner.py} \citep{Foreman-Mackey:2016},
-    % \project{emcee} \citep{Foreman-Mackey:2013},
-    % \project{george} \citep{Ambikasaran:2016},
-	% \project{matplotlib} \citep{Hunter:2007},
-	% \project{numpy} \citep{Van-Der-Walt:2011},
-	% \project{scipy} \citep{Jones:2001}.
+     \project{corner.py} \citep{Foreman-Mackey:2016},
+     \project{Eigen} \citep{Guennebaud:2010},
+     \project{emcee} \citep{Foreman-Mackey:2013},
+     \project{george} \citep{Ambikasaran:2016},
+     \project{LAPACK} \citep{Anderson:1999},
+     \project{matplotlib} \citep{Hunter:2007},
+     \project{numpy} \citep{Van-Der-Walt:2011},
+     \project{transit} \citep{Foreman-Mackey:2016a},
+     \project{scipy} \citep{Jones:2001}.
 }
 
-\appendix
-
-\section{Implementation details}\sectlabel{implementation}
-
-Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
-follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
-with the observation that a damped sinusoid correlation function may be decomposed into the sum of two
-exponential functions:
-\begin{eqnarray}
-k_j(\tau) &=& a_j e^{-c_j\tau} \cos{(-d_j\tau)} + b_j e^{-c_j\tau} \sin{(-d_j \tau)}\\
- &=& \frac{1}{2}(a_j+ib_j)\left[e^{-(c_j+id_j) \tau}+e^{-(c_j-id_j) \tau}\right],
-\end{eqnarray}
-where $c_j + id_j$ is the complex kernel parameter and $d_j = 2\pi/P$
-is the frequency of oscillation of the kernel with period $P$.
-
-Now, the damped sinusoid kernal may be included in the GRP algorithm
-using the above two exponential kernels.  This
-requires the algorithm to use complex arithmetic, which has several disadvantages:
-1).\ complex arithmetic is more expensive by a factor of 2-3; 2).\ the complex conjugate equation
-is redundant as it turns out that $l_2 = l_1^*$, $r_2 = r_1^*$, and $x_i = x_i^*$ (i.e.\ $x_i$ is
-real), meaning that the same equations are being solved twice for the real and complex components;
-and 3).\ if the derivative of the likelihood is computed with automatic
-differentiation (AD), most AD algorithms require real variables.
-
-Given these drawbacks, we instead modify the GRP algorithm to use a combination of two real
-components rather than two complex components;  this addresses all three of these drawbacks
-simultaneously.  The one cost is that the banded extended matrix has a bandwidth which is
-larger by one, both above and below the diagonal.
-
-Rather than using $c_j+id_j$ and its complex conjugate as components of the kernel, we instead
-find the real and imaginary components of the complex equation by taking the
-sum and difference of the $c_j+id_j$ and $c_j-id_j$ equations (divided by 2 and $2i$, respectively),
-and then discard the $c_j-id_j$ equation, which is redundant.  This keeps the same number of
-additional equations for each $(c_j,d_j)$ with non-zero imaginary component (four additional
-equations in the extended matrix), but converts the complex equations to real, thus avoiding
-complex arithmetic, while the kernel components with $d_j=0$ still only have two
-additional components in the extended matrix.  The extended matrix adds an extra band above
-and below the diagonal when the $c_j+id_j$ and $c_j-id_j$ equations are combined since this
-mixes the components of the original equations which are shifted by one column from one
-another.  We will enumerate the number of $d_j=0$ cases as $J_0$, while the total number
-of $c_j$'s is still $J$.  This gives a total number of equations to be solved with the
-extended matrix of $N_{ex}= (4(J-J_0)+2J_0+1)\times(N-1)+1$, where $N$ is the number of data
-points.
-
-% Reprint Ambikasaran equations (59)-(61)
-
-For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
-auxiliary vectors $l_n = g_n - i h_n$, with $n \in \{1,2,...,N\}$ and $l_1 = 0$; $r_n =
-u_n - i v_n$, with $n \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
-imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
-matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
-so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
-set of equations, and since the values of these variables are discarded in the end, their sign
-is irrelevant).  The length of the vectors $l_n$ and $r_n$
-is the number of kernel components, $J$;  we denote the individual elements of these vectors
-as $g_{n,j}+ih_{n,j}$ and $u_{n,j}+iv_{n,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
-to be the kernels with $d_j=0$, and for these we do not include the imaginary component
-equations, which are unnecessary and only add extra computation time.  We also define vectors $\phi$ and $\psi$ with
-complex components as well:  $\gamma_n = \phi_n + i\psi_n = \left[\exp{(-(c_1+id_1) t_{n,n+1})} ... \exp{(-(c_J+id_J) t_{n,n+1})}\right]^T
-$.  The individual elements are given by $\phi_{n,j} = e^{-c_j t_{n,n+1}}
-\cos{(d_j t_{n,n+1})}$ and $\psi_{n,j} = -e^{-c_j t_{n,n+1}}\sin{(d_j t_{n,n+1})}$.
-Equation (61) becomes:
-\begin{equation} \label{Amb61}
-% See 9/2/2016 notes
-% Use n rather than k for {1...N} - use g, h & u, v?
-% \gamma & \phi - real & imaginary
-\sum_{j=1}^J \left(\frac{a_j}{2} g_{n,j} +\frac{b_j}{2} h_{n,j}\right)+ \frac{d}{2} x_n + \sum_{j=1}^J \left[
-\phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j}\right] = \frac{y_n}{2},
-\end{equation}
-for $j \in \{1,...,J\}$ and $n \in \{1,...,N\}$, where $y_j$ is the $j$th data point.  There is no imaginary component to equation (61)
-in \citet{Ambikasaran:2015}.
-
-Equation (60) in \citet{Ambikasaran:2015} has real and imaginary components:
-\begin{eqnarray} \label{Amb60}
-\phi_{n,j} g_{n,j} +\psi_{n,j} h_{n,j} + \phi_{n,j} x_n - g_{n+1,j} &=& 0,\\
-\psi_{n,j} g_{n,j} -\phi_{n,j} h_{n,j} + \psi_{n,j} x_n + h_{n+1,j} &=& 0,
-\end{eqnarray}
-with $j \in \{1,...,J\}$ and $n \in \{1,...,N-1\}$
-\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran:2015}.
-
-Finally, equation (59) in \citet{Ambikasaran:2015} has real and imaginary components:
-\begin{eqnarray} \label{Amb59}
--u_{n,j} + \frac{a_j}{2} x_n + \phi_{n,j} u_{n+1,j} + \psi_{n,j} v_{n+1,j} &=&0,\\
- v_{n,j} + \frac{b_j}{2} x_n + \psi_{n,j} u_{n+1,j} - \phi_{n,j} v_{n+1,j} &=&0,
-\end{eqnarray}
-with $j \in \{1,...,J\}$ and $n \in \{2,...,N\}$.  Note that for the $J_0$ real components
-we can drop the variables $v_{n,j}$ and $h_{n,j}$ as these are zero, and we can drop the
-imaginary components of equations (59) and (60) in \citet{Ambikasaran:2015} as well for these, which reduces the total number
-of equations to solve.
-
-With these additional equations, we can solve for the likelihood of sums of damped sinusoid
-in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables
-for each damped sinusoid with non-zero $\omega$, and due to having a slightly larger bandwidth due to the
-cross terms between the real and imaginary variables.  The off-diagonal components are $J_0+2(J-J_0)+2$ above
-and below the diagonal, for a total bandwidth of $2J_0+4(J-J_0)+5$. In the case in which $J_0=J$, so
-that $d_j=0$, the off-diagonal components revert to $J+1$
-above and below the diagonal, giving a bandwidth of $2J+3$ (the same as the GRP algorithm).
-
-As with the GRP method, we can build an extended matrix which is real and symmetric, and which may be
-solved with a banded solver, as in \citet{Press:1992}.  The determinant of this extended matrix
-may be computed from the LU decomposition, and it is related to the determinant of the kernel by a
-factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
-We arrange the auxiliary variables with the $d_j=0$
-terms first, followed by the complex variables, yielding the vector
-$x_{ex} = [x_1, \{u_{2,j},g_{2,j};j=1,...,J_0\}, \{u_{2,j},v_{2,j},g_{2,j},h_{2,j};j=J_0+1,...,J\},
-x_2, ..., x_N]$.  The right hand side is given by: $y_{ex} = [y_1/2, \{0; j=1,...,J_0+2(J-J_0)\}, y_2/2, ..., y_N/2]$.
-The extended matrix, $A_{ex}$, is constructed by inserting the coefficients of equations \ref{Amb61},
-\ref{Amb60}, and \ref{Amb59} into a matrix of dimensions $(2J_0+4(J-J_0)+5) \times N_{ex}$.
-The solution is found by solving $A_{ex} x_{ex} = y_{ex}$ with LU decomposition and back-substitution
-with a banded algorithm.
-
-The matrix has a similar structure as the real case constructed by \citep{Ambikasaran:2015}, but the
-imaginary and real components mix (this is due to the representation of imaginary numbers as
-$2\times 2$ matrices).  Figure \ref{fig:matrix_structure} shows an example of the structure of the extended
-matrix for two damped sinusoids with non-zero $\omega$ values.  This is analagous to the real case with
-$J=2$ which also has four additional equations for each row of the original kernel.  This matrix shows
-that there is an extra band above and below the diagonal \citep[compare with Figure 2 of][]{Ambikasaran:2015}.
-
-{\bf Should I give some examples of this?  Some pseudocode?}
-
 \bibliography{celerite}

### 2017-02-17 celerite/modeling.py
allow bounds dict with named parameters

@@ -37,12 +37,22 @@ class Model(object):
         self.unfrozen_mask = np.ones(self.full_size, dtype=bool)
         self.dirty = True
 
-        # Pull out the recognized keyword arguments
-        self.parameter_bounds = kwargs.pop(
-            "bounds", [(None, None) for _ in range(self.full_size)])
+        # Deal with bounds
+        self.parameter_bounds = []
+        bounds = kwargs.pop("bounds", dict())
+        try:
+            # Try to treat 'bounds' as a dictionary
+            for name in self.parameter_names:
+                self.parameter_bounds.append(bounds.get(name, (None, None)))
+        except AttributeError:
+            # 'bounds' isn't a dictionary - it had better be a list
+            self.parameter_bounds = list(bounds)
         if len(self.parameter_bounds) != self.full_size:
             raise ValueError("the number of bounds must equal the number of "
                              "parameters")
+        if not all(len(b) == 2 for b in bounds):
+            raise ValueError("the bounds for each parameter must have the "
+                             "format: '(min, max)'")
 
         # Parameter values can be specified as arguments or keywords
         if len(args):

### 2017-02-17 paper/figures/astero_plot.py

@@ -7,7 +7,6 @@ import pickle
 import corner
 import numpy as np
 import matplotlib.pyplot as plt
-from scipy.ndimage.filters import gaussian_filter
 
 import emcee3
 

### 2017-02-19 paper/ms.tex
allow bounds dict with named parameters
implementing a kernel product

@@
-% Copyright 2015-2016 Dan Foreman-Mackey and the co-authors listed below.
+% Copyright 2015-2017 Dan Foreman-Mackey and the co-authors listed below.
 
@@
 likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
-measurements in $\mathcal{O}(N)$ operations using the \celerite\ method described
-in the previous section.
+measurements in $\mathcal{O}(N)$ operations using the \celerite\ method
+described in the previous section.
 
@@
 
-\begin{enumerate}
-\item an OU process
-\item a full celerite process
-\item some sort of QP oscillation
-\end{enumerate}
+\subsection{Recovery of a celerite process}
+
+In this first example, we will simulate a dataset using a known \celerite\
+process and fit it with \celerite\ to demonstrate that valid inferences can be
+made in this idealized case.
+
+\subsection{Inferences with the ``wrong'' model}
+
+For this example, we will simulate two datasets using known GP models with
+kernels outside of the support of a \celerite\ process.
+We will use these simulations to demonstrate that, at least in these cases,
+valid inferences can be made about the physical parameters of the model even
+if the effective covariance function is known to be ``wrong''.
 

### 2017-02-19 celerite/celerite.py

@@ -407,13 +407,19 @@ class GP(ModelSet):
                                                  .get_value(x1))
         return K
 
-    def sample(self, x, tiny=1e-12, size=None):
+    def sample(self, x=None, diag=None, include_diagonal=False, size=None):
         """
         Sample from the prior distribution over datasets
 
         Args:
-            x (array[n]): The independent coordinates where the observations
-                should be made.
+            x (Optional[array[n]]): The independent coordinates where the
+                observations should be made. If ``None``, the coordinates used
+                in the last call to ``compute`` will be used.
+            diag (Optional[array[n] or float]): If provided, this will be
+                added to the diagonal of the covariance matrix.
+            include_diagonal (Optional[bool]): Should the white noise and/or
+                ``yerr`` terms be included on the diagonal?
+                (default: ``False``)
             size (Optional[int]): The number of samples to draw.
 
         Returns:
@@ -421,7 +427,8 @@ class GP(ModelSet):
             distribution over datasets.
 
         """
-        K = self.get_matrix(x, include_diagonal=True)
-        K[np.diag_indices_from(K)] += tiny
+        K = self.get_matrix(x, include_diagonal=include_diagonal)
+        if diag is not None:
+            K[np.diag_indices_from(K)] += diag
         sample = np.random.multivariate_normal(np.zeros_like(x), K, size=size)
         return self.mean.get_value(x) + sample

### 2017-02-19 celerite/terms.py

@@ -2,7 +2,7 @@
 
 from __future__ import division, print_function
 import numpy as np
-from itertools import chain
+from itertools import chain, product
 
 from .modeling import Model, ModelSet
 from .solver import get_kernel_value, get_psd_value, check_coefficients
@@ -90,6 +90,12 @@ class Term(Model):
     def __radd__(self, b):
         return TermSum(b, self)
 
+    def __mul__(self, b):
+        return TermProduct(self, b)
+
+    def __rmul__(self, b):
+        return TermProduct(b, self)
+
     def get_real_coefficients(self):
         """
         Get the arrays ``alpha_real`` and ``beta_real``
@@ -165,6 +171,67 @@ class Term(Model):
         return pars
 
 
+class TermProduct(Term, ModelSet):
+
+    def __init__(self, k1, k2):
+        super(TermProduct, self).__init__([("k1", k1), ("k2", k2)])
+
+    def __repr__(self):
+        return " * ".join(map("{0}".format, (self.models["k1"],
+                                             self.models["k2"])))
+
+    @property
+    def terms(self):
+        return [self]
+
+    def get_all_coefficients(self):
+        c1 = self.models["k1"].get_all_coefficients()
+        nr1, nc1 = len(c1[0]), len(c1[2])
+        c2 = self.models["k2"].get_all_coefficients()
+        nr2, nc2 = len(c2[0]), len(c2[2])
+
+        # First compute real terms
+        nr = nr1 * nr2
+        ar = np.empty(nr)
+        cr = np.empty(nr)
+        gen = product(zip(c1[0], c1[1]), zip(c2[0], c2[1]))
+        for i, ((aj, cj), (ak, ck)) in enumerate(gen):
+            ar[i] = aj * ak
+            cr[i] = cj + ck
+
+        # Then the complex terms
+        nc = nr1 * nc2 + nc1 * nr2 + 2 * nc1 * nc2
+        ac = np.empty(nc)
+        bc = np.empty(nc)
+        cc = np.empty(nc)
+        dc = np.empty(nc)
+
+        # real * complex
+        gen = product(zip(c1[0], c1[1]), zip(*(c2[2:])))
+        gen = chain(gen, product(zip(c2[0], c2[1]), zip(*(c1[2:]))))
[…]

### 2017-02-19 tests/test_celerite.py

@@ -4,6 +4,7 @@ from __future__ import division, print_function
 
 import pytest
 import numpy as np
+from itertools import product
 
 try:
     import cPickle as pickle
@@ -31,6 +32,30 @@ method_switch = pytest.mark.parametrize(
                            reason="Sparse support not sparse"),
     ]
 )
+
+
+def test_kernel(seed=42):
+    np.random.seed(seed)
+    t = np.sort(np.random.uniform(0, 5, 100))
+    tau = t[:, None] - t[None, :]
+
+    k1 = terms.RealTerm(log_a=0.1, log_c=0.5)
+    k2 = terms.ComplexTerm(0.2, -3.0, 0.5, 0.01)
+    k3 = terms.SHOTerm(1.0, 0.2, 3.0)
+
+    K1 = k1.get_value(tau)
+    K2 = k2.get_value(tau)
+    K3 = k3.get_value(tau)
+
+    assert np.allclose((k1 + k2).get_value(tau), K1 + K2)
+    assert np.allclose((k3 + k2).get_value(tau), K3 + K2)
+    assert np.allclose((k1 + k2 + k3).get_value(tau), K1 + K2 + K3)
+
+    for (a, b), (A, B) in zip(product((k1, k2, k3, k1+k2, k1+k3, k2+k3),
+                                      (k1, k2, k3)),
+                              product((K1, K2, K3, K1+K2, K1+K3, K2+K3),
+                                      (K1, K2, K3))):
+        assert np.allclose((a * b).get_value(tau), A*B)
 
 @method_switch
 def test_carma(method, seed=42):

### 2017-02-20 celerite/terms.py
fixing typo in ms expression for the product of two terms

@@ -8,7 +8,8 @@ from .modeling import Model, ModelSet
 from .solver import get_kernel_value, get_psd_value, check_coefficients
 
 __all__ = [
-    "Term", "TermSum", "RealTerm", "ComplexTerm", "SHOTerm", "Matern32Term"
+    "Term", "TermProduct", "TermSum",
+    "RealTerm", "ComplexTerm", "SHOTerm", "Matern32Term",
 ]
 
 

### 2017-02-20 paper/ms.tex
fixing typo in ms expression for the product of two terms
adding figurues to simulated data examples section

@@
     \frac{1}{1+\left(\frac{\omega}{c_j}\right)^2} \quad.
+\end{eqnarray}
+
+It's worth noting that the product of two terms of the form found inside the
+sum in \eq{celerite-kernel} can also be re-written as a sum with updated
+parameters
+\begin{eqnarray}
+k_j(\tau) \, k_k(\tau) =
+    e^{-\tilde{c}\,\tau}\,[
+        a_+\,\cos(d_+\,\tau) + b_+\,\sin(d_+\,\tau) +
+        a_-\,\cos(d_-\,\tau) + b_-\,\sin(d_-\,\tau)
+    ]
+\end{eqnarray}
+where
+\begin{eqnarray}
+    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
+    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
+    \tilde{c} &=& c_j + c_k \\
+    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
 \end{eqnarray}
@@
 
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
+\caption{(left) A simulated dataset.
+    (right) The inferred PSD~--~the blue contours encompass 68\% of the
+    posterior mass~--~compared to the true PSD (dashed black line).
+    \figurelabel{simulated-correct}}
+\end{center}
+\end{figure}
+
 \subsection{Inferences with the ``wrong'' model}
 
-For this example, we will simulate two datasets using known GP models with
-kernels outside of the support of a \celerite\ process.
-We will use these simulations to demonstrate that, at least in these cases,
-valid inferences can be made about the physical parameters of the model even
-if the effective covariance function is known to be ``wrong''.
+For this example, we simulate a dataset using a known GP model with a kernel
+outside of the support of a \celerite\ process.
+This means that the true autocorrelation of the process can never be correctly
+represented by the model that we're using to fit but we will use this example
+to demonstrate that, at least in this case, valid inferences can still be made
+about the physical parameters of the model.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
+\caption{(left) A simulated dataset.
+    (right) The inferred period of the process. The true period is indicated
+    by the vertical orange line, the posterior inference using the correct
+    model is shown as the blue dashed histogram, and the inference made using
+    the ``wrong'' effective model is shown as the black histogram.
+    \figurelabel{simulated-wrong}}
+\end{center}
+\end{figure}
 

### 2017-02-21 paper/ms.tex
a few words about the 'wrong' example [ci skip]

@@
 \newcommand{\celerite}{\project{celerite}}
+\newcommand{\emcee}{\project{emcee}}
 
@@
 parameters
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{product-rule}
 k_j(\tau) \, k_k(\tau) =
@@
 These identities yield a kernel of the form
-\begin{equation}
+\begin{equation}\eqlabel{sho-kernel}
 k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
@@
 about the physical parameters of the model.
+
+For this example, the data are simulated from a quasiperiodic GP with the
+kernel
+\begin{eqnarray}\eqlabel{sim-wrong-true}
+k_\mathrm{true} (\tau) = \alpha\,
+    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
+    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
+\end{eqnarray}
+where $P_\mathrm{true}$ is the fundamental period of the process.
+This autocorrelation structure yields the power spectrum
+\begin{eqnarray}
+S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
+    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
+        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
+    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
+        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
+\right]
+\end{eqnarray}
+that for large $\omega$ falls off exponentially.
+When compared to \eq{celerite-psd}~--~that for large $\omega$ goes as
+$\omega^{-4}$, at most~--~it is clear that a \celerite\ model can never
+perfectly reproduce the structure of this process.
+That being said, we will demonstrate that rigorous inferences can be made
+about $P_\mathrm{true}$ even with an effective model.
+The left panel of \Figure{simulated-wrong} shows the simulated dataset.
+We then fit this simulated data using the product of two SHO terms
+(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
+1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
+We note that using \eq{product-rule}, the product of two \celerite\ terms can
+also be expressed as a \celerite\ term.
+We apply log-uniform priors on all the parameters and use \emcee\
+\citep{Foreman-Mackey:2013} to sample the posterior probability for all of the
+parameters.
+The inferred distribution for the parameter $P$ is shown in the right panel of
+\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
+the inferences made using the correct model (\eqalt{sim-wrong-true}).
+The inference made using this effective \celerite\ model are indistinguishable
+from the inferences made using the correct model but substantially less
+computation time is required for the \celerite\ inference.
 

### 2017-02-21 paper/ms.tex
a few more words about astero example [ci skip]

@@
 \newcommand{\kepler}{\project{Kepler}}
+\newcommand{\tess}{\project{TESS}}
 \newcommand{\celerite}{\project{celerite}}
@@
 made in this idealized case.
+The simulated dataset is shown in the left panel of \Figure{simulated-correct}
+and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
+$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
+The true PSD is shown as a dashed line in the right panel of
+\Figure{simulated-correct}.
+We applied log-uniform priors to all of the parameters and used \emcee\
+\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
+the marginalized posterior inference of the PSD.
+This inference is shown in the right panel of \Figure{simulated-correct} as a
+blue contour indicating 68\% of the posterior mass.
+It is clear from this figure that, as expected, the inference correctly
+reproduces the true PSD.
 
@@
 
+In this section we will demonstrate several use cases of \celerite\ when
+applied to real datasets.
+Each of these examples touches on an active area of research so we will limit
+our examples to be qualitative in nature and make no claim of optimality but
+we hope that these examples will encourage interested readers to investigate
+the applicability of \celerite\ to their research.
+
+All of the following examples show time domain datasets with a clear bias in
+favor of large homogeneous photometric surveys but these methods can similarly
+be applied to spectroscopy, where wavelength~--~instead of time~--~is the
+independent coordinate and other one-dimensional domains.
+It is also possible to cast some two-dimensional problems in the \celerite\
+framework (\dfmtodo{CITE Ian C}).
+
 \subsection{Asteroseismic oscillations}
@@
 To demonstrate this method, we will use a very simple heuristic model based on
-\dfmtodo{CITE} where the model PSD is given by a mixture of 6 components with
+\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
 amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
@@
 \eq{granulation-psd} with two free parameters $S_g$ and $\omega_g$.
-The remaining 5 terms are given by \eq{sho-psd} where $Q$ is a nuisance
+The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
 parameter shared between terms and the frequencies are given by
@@
 
-KIC~11615890 \citep{Pinsonneault:2014}
+To demonstrate the applicability of this model, we apply it to infer the
+asteroseismic parameters of the giant star KIC~11615890 observed by the
+\kepler\ Mission.
+The goal of this example is to show that, even for a low signal-to-noise
+dataset with a short baseline, it is possible to infer asteroseismic
+parameters with formal uncertainties that are consistent with the parameters
+inferred with a much larger dataset.
+Looking forward to \tess, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
+using only one month of \kepler\ data and compare our results to the results
+inferred from the full 4 year baseline of the mission.
+For KIC~11615890, the published asteroseismic parameters based on the full
+dataset are \citep{Pinsonneault:2014}
 \begin{eqnarray}
     \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
-    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz}
-\end{eqnarray}
-
-\dfmtodo{Add plots etc.}
+    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
+\end{eqnarray}
+We randomly select a month-long segment of \kepler\ data, initialize our
+\celerite\ model using a grid search in the parameter space, and then use
+\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
+the full set of parameters.
+\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
+and $\Delta\nu$ compared to the results from the literature.
+\dfmtodo{Add some more discussion here.}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
+\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
+    from the inference shown in \Figure{astero} compared to the published
+    value (error bar) based on the full \kepler\ dataset \dfmtodo{CITE}.
+    \figurelabel{astero-corner}}
+\end{center}
+\end{figure}
 
@@
     \figurelabel{astero}}
-\end{center}
-\end{figure}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
-\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
-    from the inference shown in \Figure{astero} compared to the published
-    value (error bar) based on the full \kepler\ dataset \dfmtodo{CITE}.
-    \figurelabel{astero-corner}}
 \end{center}

### 2017-02-21 paper/ms.tex
some words about transit example [ci skip]

@@
 \subsection{Exoplanet transit fitting}
+
+In this example, we inject the signal of a simulated exoplanet transit into a
+real \kepler\ light curve and then demonstrate that we can recover the true
+physical parameters of the exoplanet while modeling the stellar variability
+using \celerite.
+This example is different from all the previous examples because in this case,
+we are uninterested in the inferred parameters of the covariance model.
+Instead, we're interested in inferring constraints on the parameters of the
+mean model.
+In \eq{gp-likelihood} these parameters are called $\bvec{\theta}$ and in this
+example, the mean function $\mu_\bvec{\theta}(t)$ is a limb-darkened transit
+light curve \dfmtodo{CITE MA} parameterized by a period $P$, a transit
+duration $T$, a phase or epoch $t_0$, an impact parameter $b$, the radius
+of the planet in units of the stellar radius $R_P/R_\star$, and several
+parameters describing the limb-darkening profile of the star \dfmtodo{CITE}.
 

### 2017-02-21 paper/figures/astero_plot.py

@@ -144,6 +144,9 @@ q = np.percentile(uHz_conv/(2*np.pi)*psds, [16, 50, 84], axis=0)
 axes[2].fill_between(freq_uHz, q[0], q[2], color="k", alpha=0.3,
                      rasterized=True)
 axes[2].plot(freq_uHz, q[1], "k", alpha=0.8, rasterized=True)
+for t in gp.kernel.terms:
+    p = t.get_psd(2*np.pi*freq)*uHz_conv/(2*np.pi)
+    axes[2].plot(freq_uHz, p, ":k", alpha=0.8, rasterized=True)
 axes[2].axhline(white_noise_some)
 
 labels = [

### 2017-02-22 paper/ms.tex
removing kernel approximation section [come back here if you want to remember it!]

@@
 % derivatives of the likelihood function.
-
-\section{Kernel components}
-
-%A stationary, smooth kernel may generally be expressed as a Taylor series for the $\tau \ne 0$ component,
-%plus the white noise component:
-%\begin{equation}
-%k(\tau) = a_0 \delta(\tau) +  \sum_{j=0}^\infty c_j \tau^j.
-%\end{equation}
-%Note that this encompasses all smooth, time-stationary kernels, and generally $c_1 \le 0$ as
-%correlated noise initially declines with separation in time.  We wish to approximate this
-%as the sum of damped cosine kernels, and since this general kernel is symmetric in $\tau$, we
-%can carry out the approximation in terms of sums of damped cosine kernel components
-%\begin{equation}
-%K_j(\tau,a_j,\beta_{R,j},\beta_{I,j}) = a_j \exp{(\beta_{R,j} \tau)}\cos{(\beta_{I,j} \tau)},
-%\end{equation}
-%which has a Taylor series expansion:
-%\begin{eqnarray}
-%\exp{(\beta_R \tau)}\cos{(\beta_I \tau)} &=& 1 - \beta_R \tau + \frac{1}{2}\left(\beta_R^2 - \beta_I^2\right)\tau^2 \\
-%- \frac{1}{6} \left(\beta_R^3-3\beta_R\beta_I^2\right)\tau^3 &+& \frac{1}{24} \left(\beta_R^4+\beta_I^4-6\beta_R^2\beta_I^2\right) \tau^4+ O(\tau^5),
-%\end{eqnarray}
-%which may be continued to any order in $\tau$ as necessary.
-%If an approximation to $k(\tau)$ is desired in a Taylor series to some high order $N$, this
-%requires combining $N/2$ components of $K_j(\tau,a_j,\beta_{R,j},\beta_{I,j})$ with $\beta_{R,j}$ and $\beta_{I,j}$
-%chosen to match the coefficients $c_j$.  Note that for $N=1$ we only require a damped exponential
-%kernel with $\beta_I = 0$.
-%
-%Another way to view this is in the power spectrum of the gaussian process, which is the
-%Fourier transform of the autocorrelation function.  The power spectrum of the $K_j$ kernel
-%is a Lorentzian, for which the frequency is given by $\omega = \beta_{I,j}$, while the
-%width of the Lorentzian is determined by $\beta_{R,j}$.  The functional form of this
-%power spectrum is: {\color{red} Check that this is true.}
-%\begin{equation}
-%P(\vert f\vert) = a_j \beta_{R,j} \frac{1}{\beta_{R,j}^2 + (2\pi f - \beta_{I,j})^2},
-%\end{equation}
-%where $f$ is the frequency in Hz.  The counterpart to the Taylor series is the observation
-%that a smooth distribution may be approximated arbitrarily precisely by summing together
-%Lorentzians with various values of $(a_j,\beta_{R,j},\beta_{I,j})$.  In the case that
-%a peak in the power spectrum is narrow at a specific frequency, then a small value of
-%$\beta_{R,j}$ gives a narrow Lorentzian that may approximate that peak.  If the wings
-%of the distribution decay faster than $f^{-2}$, then differences between Lorentzians
-%can be arranged to cancel in the wings, giving features in the power spectrum that have
-%narrower wings.
-
-The functional form of $k_j$ may be used to approximate some commonly used Kernel components.
-The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
-used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
-commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be
-used to approximate each of these kernels to better than 1\% with various numbers of damped
-cosine kernel components:
-\begin{itemize}
-\item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
-correlations in data, and can be represented exactly with $c_j= d_j =  0$.
-\item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
-represented with $d_j=0$, $c_j= \tau_0^{-1}$.
-\item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
-may be represented exactly with $c_j=0$ and $d_j= \frac{2\pi}{P}$.
-\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
-has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
-Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
-which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of the
-power spectrum of a damped sinuosoid.  Nevertheless, the squared exponential kernel may be well approximated
-by the sum of four exponential kernels:
-\begin{eqnarray}
-G(z) &\approx& 63.119512 e^{-1.479579 z} -32.173615 e^{-1.593691 z} -29.425156 e^{-1.388929 z}\\
-&+& e^{-1.383928 z} \left[-0.519196\cos{(1.883396 z)} +0.285478 \sin{(1.883396 z)}\right] \\
-\end{eqnarray}
-which has an accuracy of better than $<3\times 10^{-3}$ for all z, and the power spectrum has
-been constrained to be positive.
-\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
-\begin{equation}
-C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
-\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
-\end{equation}
-where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
-squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
-values of $p$, the Mat\'ern kernel may be approximated well by the sum of
-$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel,
-$C_{1/2}(x) = \sigma^2 e^{-x}$ ($p=0$), while
-$p=1$ is relevant to stellar variability, which we describe next.
-\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by:
-\begin{equation}
-k_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
-\end{equation}
-where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier transform\footnote{Using the normalization convention:
-$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
-\begin{equation}
-\hat k_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
-\end{equation}
-where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels:
-\begin{equation}
-k_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
-\end{equation}
-where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
-large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
-$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
-have opposite signs, which may lead to numerical instability for values of $b$ that
-are large.
-
-One advantage of the Mat\'ern 3/2 kernel is that it is smooth across
-$\tau=0$ since its derivative is zero at $\tau=0$.  The exponential kernel {\it does not}
-have this property, which leads to qualitatively different noise properties on small
-timescales due to the sharpness of the exponential kernel at $\tau=0$.  The Mat\'ern 3/2 kernel
-turns out to be similar to the observed properties of stellar activity and granulation, and thus the approximate
-formula is simultaneously physically relevant and computationally convenient.
-The power spectrum of a single exponential scales as $f^{-2}$ at high
-frequency, while the Mat\'ern kernel scales as $f^{-4}$ which more
-closely matches the power spectrum due to activity and granulation in
-the Sun \citep{Aigrain:2004,Michel:2009}.
-\item {\bf Mat\'ern 5/2 kernel.}  The Mat\'ern 5/2 kernel is given by
-\begin{equation}
-k_{5/2}(t) = C_{5/2}(y) = \sigma^2 (1+y+y^2/3)e^{-y},
-\end{equation}
-where $y = \sqrt{5}\vert \tau/\tau_0\vert$.  The Fourier transform is:
-\begin{equation}
-\hat k_{5/2}(\omega) = \frac{2\tau_0 10^{5/2}}{3\pi^{1/2}} (5+\tau_0^2\omega^2)^{-3}.
-\end{equation}
-This kernel may be approximated well by the sum of four exponential kernels
-\begin{eqnarray}
-k_{5/2}(\tau) &\approx& -147.6965573 e^{-1.0097610 y} + 85.8875034 e^{-1.0621734 y}\\
- &-&4.6540017 e^{-0.8353745 y} + 67.4630744 e^{-0.9160394 y},
-\end{eqnarray}
-which is accurate to $<4\times 10^{-6}$ for $y \le 10$.
-The Fourier transform has a steeper dependence at large frequencies, $f^{-6}$, and as we are not aware of a stellar
-power spectrum that requires a steeper spectrum than this, we forego approximating further Mat\'ern kernels.
-\item {\bf Exponential-sine-square kernel.}  This kernel is a periodic, positive valued kernel
-which has been used to model quasi-periodic stellar variability due to star spots.   It is
-given by $E(w,\Gamma) = e^{-\Gamma \sin^2{w}}$ where $w = \frac{\pi}{P}\vert \tau \vert$, which may be
-expanded as a series in $\cos{(2kw)} = \cos{\frac{2\pi}{P} k \vert \tau \vert)}$,
-with $k \in \mathbb{N}$.  We have expanded this to $k=4$, and fit for the coefficients, giving:
-\begin{equation}
-E(w,\Gamma) \approx \sum_{k=0}^4 \left(c_{k,1}+c_{k,2} e^{-c_{k,4}\Gamma^{c_{k,3}}}\right) \cos{(2kw)},
-\end{equation}
-with $c_{0,1}=0.24999, c_{0,2}=0.752803, c_{0,3}=0.960477, c_{0,4}=0.643393$ for $k=0$,
-$c_{1,1}=0.450851, c_{1,2}=-0.445338, c_{1,3}=0.960477, c_{1,4}= 1.18102$ for $k=1$,
-$c_{2,1}=0.206593, c_{2,2}=-0.207361, c_{2,3}=1.65902, c_{2,4}= 0.210641$ for $k=2$,
-$c_{3,1}=0.0728449, c_{3,2}=-0.0729381, c_{3,3}=2.43149, c_{3,4}= 0.0471224$ for $k=3$, and
-$c_{4,1}=0.019204, c_{4,2}=-0.0192129, c_{4,3}=3.27977, c_{4,4}= 0.0114198$ for $k=4$.
-This approximation is accurate to better than 0.6\% for $\Gamma < 3$, with a standard
-deviation of 0.1\%.
-\end{itemize}
-
-Examples of the foregoing kernels and their approximations are shown in Figure \ref{kernel_approx}.
-It is clear from this figure that these standard kernels are well approximated by the sums
-of between two and five damped cosine kernels, demonstrating that these form an adequate set
-for modeling a wide range of kernel behaviors.
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{./approximate_kernels/kernel_approx.pdf}
-\caption{Approximation of various kernels with the sums of damped cosines.
-(a). Kernels that may be expressed exactly: constant (dark grey), exponential (cyan),
-cosine (olive), and damped cosine (salmon).
-(b). Approximation of the squared-exponential (Gaussian) kernel (pink) with three
-damped cosine kernels (magenta). (c). The approximation of the exponential-sine-squared
-kernel with $\Gamma =1$ (blue) and $\Gamma = 3$ (green). (d). Two Mat\'ern kernels,
-$k_{3/2}$ (yellow/brown) and $k_{5/2}$ (violet).  The Mat\'ern 3/2 kernel
-is approximated with $b=100$. The exact kernels are solid, while approximate are dashed.}
-\label{kernel_approx}
-\end{center}
-\end{figure}
-
-A standard practice in constructing kernels for solving problems with Gaussian processes
-is to add or multiply the preceding standard kernel components.  In the case of damped cosine
-kernels, the product of two kernels is:
-\begin{eqnarray}
-k_j(\tau)k_k(\tau) &=&
-\frac{1}{2} e^{-(c_j+c_k) \tau} \left((a_ja_k+b_jb_k)\cos{\left[(d_j-d_k)\tau\right]} + (a_jb_k-a_kb_j)\sin{\left[(d_j-d_k)\tau\right]} \right. \\
- &+&\left.(a_ja_k-b_jb_k)\cos{\left[(d_j+d_k)\tau\right]} -(a_jb_k-a_kb_j)\sin{\left[(d_j+d_k)\tau\right]}\right),
-\end{eqnarray}
-which is the sum of two damped sinusoidal kernels.  Thus, only addition is necessary to
-construct new kernels from the product of damped sinuosidal kernels.
-
-The upshot is that the damped sinusoid is an expressive basis kernel that may be used
-to represent a wide range of autocorrelation functions relevant for stellar variability, and
-more.
-We recommend using a series of sums of damped sinusoids (in which some parameters may be
-fixed to specified values to represent particular kernels) to model different amplitudes
-and timescales of variability.  If an initial estimate of the power spectrum or
-autocorrelation function is available, then these may be fit with this series to
-initialize the coefficients of the damped sinusoid kernel components.
-Since a Gaussian Process
-must be positive definite, we recommend modeling the logarithm of $a_j$, $b_j$,
-$c_j$, and $d_j$ to enforce this to be the case.  One caveat is
-that to produce certain kernels, a negative value of $a_j$ or $b_j$ is needed (such
-as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
-two coefficients may be sampled from a logarithm, while the individual
-values are allowed to be negative;  this will enforce the positive definite
-constraint, but may allow for other constraints, such as an autocorrelation
-function which has a derivative of zero at $\tau = 0$.
 

### 2017-03-02 paper/ms.tex
a few additions to the docs

@@
 \end{eqnarray}
-A model like this has been used to model the granulation is Solar data
-\citep{Michel:2009}}, which has a kernel of
+This, in turn, is the most commonly used model for the background granulation
+noise in asteoreseismic \citep{Kallinger:2014} and helioseismic
+\citep{Harvey:1985, Michel:2009} analyses.
+The Fourier transform of this PSD corresponds to the kernel
 \begin{equation}
@@
     \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
-\end{equation}
+\end{equation}}
 

### 2017-03-02 paper/ms.tex
fixing #62 [ci skip]

@@
 
-\section{Gaussian processes}
+\section{Gaussian processes}\sectlabel{gps}
 
 Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
-models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
+models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
 covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
@@
 
-To provide some insight for this method, we will start by working through the
-simplest possible example.
+To provide some insight for this method, we will start by working through a
+simple example.
 In this case, we'll assume that we have three data points
@@
 and $d=0$.
-To demonstrate motivate this method, we will write out the full system of
+To demonstrate this method, we will write out the full system of
 equations that we must solve to apply the inverse of $K$ in order to compute
@@
 \begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{x} &=& \bvec{y}, \\
+K\,\bvec{z} &=& \bvec{y}, \\
 \begin{pmatrix}
@@
 \begin{pmatrix}
-    x_1 \\ x_2 \\ x_3
+    z_1 \\ z_2 \\ z_3
 \end{pmatrix} &=&
@@
     y_1 \\ y_2 \\ y_3
-\end{pmatrix} \quad.
-\end{eqnarray}
+\end{pmatrix}
+\end{eqnarray}
+where our goal is to solve for the unknown vector \bvec{z} for a given matrix
+$K$ and vector \bvec{y}.
+In this equation, we have assumed that the mean function is zero but a
+non-zero mean could be included by replacing \bvec{y} by
+$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
 Now, if we introduce the variables
 \begin{eqnarray}\eqlabel{algo-first}
-    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
+    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
 \end{eqnarray}
@@
 \begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
 \end{eqnarray}
@@
 \begin{eqnarray}
-(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
-a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
+(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
+a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
 \end{eqnarray}
@@
 \begin{pmatrix}
-    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
+    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
 \end{pmatrix} &=&
@@
 \begin{eqnarray}
-    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
 \end{eqnarray}
@@
 general~--~and we can take advantage of these algorithms to solve our system
-exactly because the target vector $\bvec{x}$ is a subset of the elements of
-$\bvec{x}_\mathrm{ext}$.
+exactly because the target vector $\bvec{z}$ is a subset of the elements of
+$\bvec{z}_\mathrm{ext}$.
 
@@
 \psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
-g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,x_n + \psi_{n,j}\,h_{n-1,j}\\
-h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,x_n - \psi_{n,j}\,g_{n-1,j}\\
-u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,x_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
-v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,x_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
 \end{eqnarray}
@@
 + \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
-    &=& y_n
+    &=& r_{\bvec{\theta},n}
 \end{eqnarray}
@@
 \begin{eqnarray}
-K\,\bvec{x} &=& \bvec{y} \quad.
-\end{eqnarray}
+K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
+\end{eqnarray}
+if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
+$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
 \eq{full-system-first} through \eq{full-system-last} define a banded matrix
 equation in the ``extended'' space and, as before, this can be used to solve
-for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(N)$ operations.
+for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
+operations.
 \Figure{matrix} shows a pictorial representation of the sparsity pattern of
@@
 Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
-$\bvec{x}_\mathrm{ext}$ and $\bvec{y}_\mathrm{ext}$ are defined schematically
+$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
 as
 \begin{eqnarray}
-{\bvec{x}_\mathrm{ext}} ^\T =
+{\bvec{z}_\mathrm{ext}} ^\T =
 \left(\begin{array}{cccccccccc}
-    x_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & x_2 & u_{2,j} & \cdots &
-    h_{N-1,j} & x_N
+    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
+    h_{N-1,j} & z_N
 \end{array}\right)
@@
 \begin{eqnarray}
-{\bvec{y}_\mathrm{ext}} ^\T =
+{\bvec{r}_\mathrm{ext}} ^\T =
 \left(\begin{array}{cccccccccc}
-    y_1 & 0 & 0 & 0 & 0 & y_2 & 0 & \cdots &
-    0 & y_N
+    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
+    0 & \cdots & 0 & r_{\bvec{\theta},2}
 \end{array}\right) \quad.
@@
 This decomposition can be used to compute the determinant of $K$, solve
-$K^{-1}\,\bvec{y}$, and subsequently calculate the marginalized likelihood in
-\eq{gp-likelihood}.
+$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
+likelihood in \eq{gp-likelihood}.
 

### 2017-03-10 paper/ms.tex
more acks

@@
 Sivaram Ambikasaran,
+Megan Bedell,
 Will Farr,
-and
-David W.\ Hogg
+Sam Grunblatt,
+David W.\ Hogg, and
+Dan Huber
 for helpful contributions to the ideas and code presented here.

### 2017-03-10 celerite/celerite.py

@@ -432,3 +432,22 @@ class GP(ModelSet):
             K[np.diag_indices_from(K)] += diag
         sample = np.random.multivariate_normal(np.zeros_like(x), K, size=size)
         return self.mean.get_value(x) + sample
+
+    def sample_uniform(self, x_min, x_max, nx, size=None):
+        x = np.linspace(x_min, x_max, nx)
+        dx = x - x[0]
+
+        k = self.kernel.get_value(dx)
+        s = np.append(k, k[1:-1][::-1])
+        M = len(s)
+        Fs = np.sqrt(M*np.fft.fft(s))
+        if size is None:
+            nr = 1
+        else:
+            nr = int(np.ceil(0.5 * size))
+        e = (np.random.randn(nr, M) + 1.j * np.random.randn(nr, M)) * Fs
+        y = np.fft.ifft(e)[:, :nx]
+        if size is None:
+            return y[0].real
+        y = np.concatenate((y.real, y.imag), axis=0)
+        return y[:size]

### 2017-03-13 paper/ms.tex
some changes

@@
 For this method to be applicable, the data must be one-dimensional and the
-covariance function must be written as a mixture of damped sinusoid functions.
+covariance function must have a specific form.
 However, there is no further constraint on the data or the model.
@@
 \end{eqnarray}
+For clarity, we will refer to the argument within the sum as a ``\celerite\
+term'' for the remainder of this paper.
 The Fourier transform\footnote{Here and throughout we have defined the Fourier
@@
 
-In the case of $J$ damped sinusoids we can check for negative values of the PSD by solving for the
-roots of the power spectrum, abbreviating with $z = \omega^2$:
+In the case of $J$ \celerite\ terms, we can check for negative values of the
+PSD by solving for the roots of the power spectrum, abbreviating with
+$z = \omega^2$:
 \begin{equation}

### 2017-03-14 paper/ms.tex
moved sturm's theorem to appendix

@@
 \end{figure}
-
-\subsection{Ensuring positive definiteness: Sturm's theorem}
-
-The power spectrum is computed from taking the square of the Fourier transform
-of the data time series; hence, the power-spectrum must be non-negative.  In
-constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
-then it is possible for the power spectrum to go negative, which violates the
-definition of the power spectrum.
-Consequently, if any of these coefficients is negative, it is necessary to
-check that the power spectrum is still non-negative.  Note that a non-negative
-power-spectrum does not require the auto-correlation
-function to be positive for all $\tau$.  The positive power spectrum is
-related to the positive eigenvalues of the covariance matrix which are
-necessary for a positive-definite matrix in limiting cases
-\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
-positive power spectrum results in positive eigenvalues for the covariance
-matrix, and so we describe here how to ensure a positive power spectrum using
-Sturm's theorem.
-%Relation of positive power spectrum to positive definite covariance matrix.
-%Power spectrum of a sum of kernels.
-
-In the case of $J$ \celerite\ terms, we can check for negative values of the
-PSD by solving for the roots of the power spectrum, abbreviating with
-$z = \omega^2$:
-\begin{equation}
-P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
-\end{equation}
-where
-\begin{eqnarray}
-q_j &=& a_jc_j-b_jd_j\\
-r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
-s_j &=& 2(c_j^2-d_j^2)\\
-t_j &=& (c_j^2+d_j^2)^2.
-\end{eqnarray}
-The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
-\begin{equation}
-P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
-\end{equation}
-which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
-which may be solved exactly for the roots.
-
-A procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
-real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
-$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
-$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.
-
-We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
-give us the signs of these polynomials evaluated at $z=0$.
-Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial
-as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where
-$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.
-
-With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each
-of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign
-changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given
-by $N_{+}=\sigma(0)-\sigma(\infty)$.
-
-We have checked that this procedure works for a wide range of parameters, and we find that it robustly
-matches the number of positive real roots which we evaluated numerically.
-
-The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic
-manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
-likelihood may be set to zero.
-%Sturm's theorem applied to numerator of power spectrum.
 
@@
 
+\appendix
+
+\section{Ensuring positive definiteness: Sturm's theorem}
+
+The power spectrum is computed from taking the square of the Fourier transform
+of the data time series; hence, the power-spectrum must be non-negative.  In
+constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
+then it is possible for the power spectrum to go negative, which violates the
+definition of the power spectrum.
+Consequently, if any of these coefficients is negative, it is necessary to
+check that the power spectrum is still non-negative.  Note that a non-negative
+power-spectrum does not require the auto-correlation
+function to be positive for all $\tau$.  The positive power spectrum is
+related to the positive eigenvalues of the covariance matrix which are
+necessary for a positive-definite matrix in limiting cases
+\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
+positive power spectrum results in positive eigenvalues for the covariance
+matrix, and so we describe here how to ensure a positive power spectrum using
+Sturm's theorem.
+%Relation of positive power spectrum to positive definite covariance matrix.
+%Power spectrum of a sum of kernels.
+
+In the case of $J$ \celerite\ terms, we can check for negative values of the
+PSD by solving for the roots of the power spectrum, abbreviating with
+$z = \omega^2$:
+\begin{equation}
+P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
+\end{equation}
+where
+\begin{eqnarray}
+q_j &=& a_jc_j-b_jd_j\\
+r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
+s_j &=& 2(c_j^2-d_j^2)\\
+t_j &=& (c_j^2+d_j^2)^2.
+\end{eqnarray}
+The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
+\begin{equation}
+P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
+\end{equation}
+which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
+which may be solved exactly for the roots.
+
+For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
+real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
+$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
+$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.
+
+We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
+give us the signs of these polynomials evaluated at $z=0$.
+Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial
+as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where
+$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.
+
+With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each
+of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign
+changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given
+by $N_{+}=\sigma(0)-\sigma(\infty)$.
+
+We have checked that this procedure works for a wide range of parameters, and we find that it robustly
+matches the number of positive real roots which we evaluated numerically.
+
+The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic
+manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
+likelihood may be set to zero.
+
+We have implemented this algorithm in subroutines that may be called to check for positive-definiteness
+of a kernel with some values of $a_j$ or $b_j$ negative.
+%Sturm's theorem applied to numerator of power spectrum.
+
 \vspace{1.5em}

### 2017-03-14 paper/ms.tex
Fleshed out abstract

@@
 with a specific emphasis on large astronomical time-series data sets.
-This method can be applied to any Gaussian Process model where the spectral
+This method, which we christen \celerite, can be applied to any Gaussian Process model where the spectral
 density can be expressed as any general mixture of damped sinusoid functions.
+We present a mathematical description of \celerite,
+its relation to the stochastically-driven, damped simple harmonic oscillator, 
+and details of the implementation.  We give examples of \celerite\ applied to both 
+simulated and actual astronomical time-series, using it to estimate the time-series
+power spectrum, while simultaneously accounting for correlated noise when inferring 
+model parameters.  These examples include Bayesian estimation of the strength 
+of stellar granulation, stellar rotation periods, asteroseismic spectra, and 
+transiting planet light curve parameters.  The model is flexible, fast, and 
+general, with potential for application to a wide range of astronomical problems, 
+for which we provide open-source software that implements \celerite\ in 
+\project{C++}, \project{Python}, and \project{Julia}.
 
@@
 Alongside this paper, we have released efficient and well-tested
-implementations of this method written in \project{C++}, \project{Python}, and
-\project{Julia}.
+implementations of this method written in \project{C++}, \project{Python}, and \project{Julia}.
 These implementations are available online at \project{GitHub}

### 2017-03-14 paper/ms.tex
Added references

@@
 inferred with a much larger dataset.
-Looking forward to \tess, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
+Looking forward to \tess \citep{Campante2016}, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
 using only one month of \kepler\ data and compare our results to the results
@@
 One promising early attempt at accounting for correlated noise was the wavelet approach
-of Carter \& Winn (2010).  Their technique allowed for a power-law spectrum for the
+of \citet{Carter2009}.  Their technique allowed for a power-law spectrum for the
 noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
 physically-accurate, description of stellar variability.  Their approach runs in
-${\cal O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
-particular type of noise overpredicts the power spectrum at low frequency compared
-to the granulation power spectrum.  It also cannot describe quasi-periodic noise.
+${\cal O}(N)$ time for $\alpha=1$, which is referred to as `flicker', $1/r$, or pink 
+noise;  however, this particular type of noise overpredicts the power spectrum at low 
+frequency compared to the Solar granulation power spectrum.  It also cannot describe 
+quasi-periodic noise.
 

### 2017-03-14 paper/ms.tex
Added more references

@@
 times.  Some examples of the first are determining the characteristic
-variability timescale of a quasar, measuring the asteroseismic variation of a star,
-or detection of quasi-periodic variability in a high-energy source.  Examples of the
+variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star,
+or detection of quasi-periodic variability in a high-energy source \citep{McAllister2016}.  
+Examples of the
 second are detecting transiting exoplanets and correcting for stellar activity in
-radial velocity measurements. Examples of the third are measuring time delays in
+radial velocity measurements \citep{Haywood2014,Rajpaul2015}. Examples of the third are 
+measuring time delays in
 multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.
@@
 phase functions, and more.
+
+In the future we hope to expand the \celerite\ technique to multi-dimensional
+datasets.  The formalism presented could, for example, be used to model
+the covariance between multiple measured quantities, such as in regression of photometric
+time series versus astrometric precision when correcting for systematic errors
+due to detector sensitivity variations \citep{Aigrain2016}, or in correcting
+radial velocity datasets for stellar activity \citep{Haywood2014,Rajpaul2017}.
 

### 2017-03-14 paper/ms.tex
updating benchmark figures and plots for macos

--- 2017-03-14 paper/ms.tex
@@
 with a specific emphasis on large astronomical time-series data sets.
-This method, which we christen \celerite, can be applied to any Gaussian Process model where the spectral
+This method can be applied to any Gaussian Process model where the spectral
 density can be expressed as any general mixture of damped sinusoid functions.
-We present a mathematical description of \celerite,
-its relation to the stochastically-driven, damped simple harmonic oscillator, 
-and details of the implementation.  We give examples of \celerite\ applied to both 
-simulated and actual astronomical time-series, using it to estimate the time-series
-power spectrum, while simultaneously accounting for correlated noise when inferring 
-model parameters.  These examples include Bayesian estimation of the strength 
-of stellar granulation, stellar rotation periods, asteroseismic spectra, and 
-transiting planet light curve parameters.  The model is flexible, fast, and 
-general, with potential for application to a wide range of astronomical problems, 
-for which we provide open-source software that implements \celerite\ in 
-\project{C++}, \project{Python}, and \project{Julia}.
 
@@
 Alongside this paper, we have released efficient and well-tested
-implementations of this method written in \project{C++}, \project{Python}, and \project{Julia}.
+implementations of this method written in \project{C++}, \project{Python}, and
+\project{Julia}.
 These implementations are available online at \project{GitHub}
@@
 \end{eqnarray}
-
 After constructing the extended matrix (using a compact storage format), the
@@
 Cholesky solver cannot be used for increased efficiency.} routine optimized
-for band matrices.
-This decomposition can be used to compute the determinant of $K$, solve
+for band or sparse matrices.
+This decomposition can then be used to compute the determinant of $K$, solve
 $K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
@@
 
+\item A general sparse LU solver that exploits the sparsity but not the band
+    structure \citep{Guennebaud:2010}.
+
 \item A simple algorithm for computing the LU decomposition for banded
@@
 
-\item A general sparse LU solver that exploits the sparsity but not the band
-    structure \citep{Guennebaud:2010}.
-
 \end{enumerate}
+
+\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$
+with their method that used the sparse LU decomposition implemented in the
+\project{SuperLU} package \citep{Demmel:1999}.
+The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
+because the dimension of the extended matrix scales as $N\,J$ and the
+bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
+We find that, while the vanilla banded solver scales as expected, the LAPACK
+algorithm with tuned BLAS routines scales empirically as $\mathcal{O}(N\,J^2)$
+empirically and offers the fastest solves for $J \gtrsim 8$ on all platforms
+that we tested.
 

### 2017-03-14 paper/ms.tex
Adding yet more references...

--- 2017-03-14 paper/ms.tex
@@
 inferred with a much larger dataset.
-Looking forward to \tess \citep{Campante2016}, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
+Looking forward to \tess \citep{Campante:2016}, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
 using only one month of \kepler\ data and compare our results to the results
@@
 the method is general for one-dimensional GP problems, and may be applied to
-other problems.    Within astrophysics, correlated noise (due to the environment,
-detector, or modeling uncertainty) may be present in
-gravitational wave time series, and so Gaussian processes may be a way to address
-this problem \citep{Moore:2016}.  Accreting black holes show time series which
+other problems.  Accreting black holes show time series which
 may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
@@
 times.  Some examples of the first are determining the characteristic
-variability timescale of a quasar, measuring the asteroseismic variation of a star,
-or detection of quasi-periodic variability in a high-energy source.  Examples of the
+variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star
+\citep{Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
+detection of quasi-periodic variability in a high-energy source \citep{McAllister:2016},
+or classification of variable objects \citep{Zinn:2016}.  Examples of the
 second are detecting transiting exoplanets and correcting for stellar activity in
-radial velocity measurements. Examples of the third are measuring time delays in
-multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.
+radial velocity measurements \citep{Haywood:2014,Rajpaul:2015}. Examples of the third are 
+measuring time delays in
+multiply-imaged gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of AGN.
 
@@
 One promising early attempt at accounting for correlated noise was the wavelet approach
-of \citet{Carter2009}.  Their technique allowed for a power-law spectrum for the
+of \citet{Carter:2009}.  Their technique allowed for a power-law spectrum for the
 noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
@@
 phase functions, and more.
+
+In the future we hope to expand the \celerite\ technique to multi-dimensional
+datasets.  The formalism presented could, for example, be used to model
+the covariance between multiple measured quantities, such as in regression of photometric
+time series versus astrometric precision when correcting for systematic errors
+due to detector sensitivity variations \citep{Aigrain:2016}, in correcting
+radial velocity datasets for stellar activity \citep{Haywood:2014,Rajpaul:2015},
+or in detection of gravitational waves in the presence of systematic errors
+\citep{Moore:2016}.
 

### 2017-03-14 paper/ms.tex
yet more refs

@@
 In the astrophysical literature, GPs have been used to model stochastic
-variability in light curves of stars (CITE), active galactic nuclei (CITE),
+variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
 and X-ray binaries (CITE).
@@
 \citep{MacLeod:2010}, measuring time lags with reverberation mapping
-\citep{Zu:2011}, and modeling time delays in multiply-imaged
+\citep{Zu:2011,Pancoast:2014}, and modeling time delays in multiply-imaged
 gravitationally-lensed systems \citep{Press:1998}.
@@
 variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star
-\citep{Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
+\citep{Brewer:2009,Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
 detection of quasi-periodic variability in a high-energy source \citep{McAllister:2016},

### 2017-03-14 paper/ms.tex

@@
 variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
-and X-ray binaries (CITE).
-They have also been used as models for the cosmic microwave background (CITE),
-correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
+and the logarithmic flux of X-ray binaries \citep{Uttley:2005}.
+They have also been used as models for the cosmic microwave background \citep{Bond:1987,Bond:1999},
+correlated instrumental noise \citep{Gibson:2012}, spectroscopic calibration \cite{Czekala:2017,Evans:2015} and
 residuals caused by model inconsistencies (CITE + better words).

### 2017-03-14 paper/ms.tex
Added some asteroseismology references

@@
 The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ mission (\dfmtodo{cite}) and asteroseismology is a
+curves from the \kepler\ mission \citep{Gilliland:2010,Chaplin:2011,Chaplin:2013,Stello:2013}
+and asteroseismology is a
 key science driver for many of the upcoming large scale photometric surveys
-(\dfmtodo{cite}).
+\citep{Campante:2016,Rauer:2014,Gould:2015}.
 Most asteroseismic analyses have been limited to relatively high

### 2017-03-14 paper/ms.tex

@@
 or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
-\begin{eqnarray}\eqlabel{approx-matern}
+\begin{eqnarray}\eqlabel{approx-matern2}
 k(\tau) &=& \lim_{f \to 0}\,
@@
 This suggests that the Mat\'ern-3/2 covariance can be well approximated using
-the \celerite\ framework with a small value of $f$ in \eq{approx-matern} but we
+the \celerite\ framework with a small value of $f$ in \eq{approx-matern2} but we
 caution that this could also lead to numerical issues with the solver.
@@
 (\dfmtodo{CITE}) and more sophisticated methods are computationally expensive
-and they scale poorly to the current state-of-the-art datasets (\dfmtodo{CITE
-Brewer etc.}).
+and they scale poorly to the current state-of-the-art datasets \citep{Brewer:2009}.
 

### 2017-03-14 paper/ms.tex
tiny typos and formatting

@@
 to astronomical time series
-\vspace{-3\baselineskip}  % OMG AASTEX6 IS SO BROKEN
+\vspace{-3\baselineskip}
 }
@@
 In the astrophysical literature, GPs have been used to model stochastic
-variability in light curves of stars \citep{Brewer:2009}, active galactic nuclei \citep{Kelly:2014},
-and the logarithmic flux of X-ray binaries \citep{Uttley:2005}.
-They have also been used as models for the cosmic microwave background \citep{Bond:1987,Bond:1999},
-correlated instrumental noise \citep{Gibson:2012}, spectroscopic calibration \cite{Czekala:2017,Evans:2015} and
-residuals caused by model inconsistencies (CITE + better words).
+variability in light curves of stars \citep{Brewer:2009}, active galactic
+nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
+\citep{Uttley:2005}.
+They have also been used as models for the cosmic microwave background
+\citep{Bond:1987,Bond:1999}, correlated instrumental noise
+\citep{Gibson:2012}, spectroscopic calibration \citep{Czekala:2017,Evans:2015}
+and residuals caused by model inconsistencies (CITE + better words).
 While these models are widely applicable, their use has been limited, in
@@
 The Fourier transform\footnote{Here and throughout we have defined the Fourier
-transform of the function $f(t)$ as $F(\omega)=(2\,\pi)^{-1/2}\,
+transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
 \int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
@@
 \frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
-{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
+{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
 \end{eqnarray}
@@
 S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
-    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
-    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
+    \frac{1}{1+{\left(\frac{\omega-d_j}{c_j}\right)}^2} +
+    \frac{1}{1+{\left(\frac{\omega+d_j}{c_j}\right)}^2}
 \right] \quad.
@@
 This model can be interpreted intuitively as a quasiperiodic oscillator with
-amplitude $A_j = a_j$, quality factor $Q_j = d_j\,(2\,c_j)^{-1}$, and period
+amplitude $A_j = a_j$, quality factor $Q_j = d_j\,{(2\,c_j)}^{-1}$, and period
 $P_j = 2\,\pi\,{d_j}^{-1}$.
@@
 S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
-    \frac{1}{1+\left(\frac{\omega}{c_j}\right)^2} \quad.
+    \frac{1}{1+{\left(\frac{\omega}{c_j}\right)}^2} \quad.
 \end{eqnarray}

### 2017-03-14 paper/ms.tex
updating sparse benchmark

@@
 \begin{center}
-\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_default.pdf}
-\caption{Benchmark\figurelabel{benchmark}}
+\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin.pdf}
+\caption{A benchmark showing the computational scaling of the \celerite\
+    implementation as a function of the number of data points and the number
+    of terms.
+    (left) The cost of computing \eq{gp-likelihood}
+    \figurelabel{benchmark}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_lapack.pdf}
+\caption{Benchmark\figurelabel{benchmark-lapack}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_sparse.pdf}
+\caption{Benchmark\figurelabel{benchmark-sparse}}
 \end{center}

### 2017-03-14 paper/ms.tex
benchmark plots and captions

@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin.pdf}
-\caption{A benchmark showing the computational scaling of the \celerite\
-    implementation as a function of the number of data points and the number
-    of terms.
-    (left) The cost of computing \eq{gp-likelihood}
+\caption{A benchmark showing the computational scaling of \celerite\ with a
+    vanilla band solver as a function of the number of data points and the
+    number of terms.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different colors show the cost for different numbers of terms $J$ as
+    listed in the legend.
+    To guide the eye, the black line shows linear scaling in $N$.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    The legend shows the value of $N$ for each color and the black line shows
+    quadratic scaling in $J$.
     \figurelabel{benchmark}}
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_lapack.pdf}
-\caption{Benchmark\figurelabel{benchmark-lapack}}
+\caption{The same as \Figure{benchmark} but using the band solver from
+    \project{LAPACK} with the \project{MKL} \project{BLAS} library.
+    \figurelabel{benchmark-lapack}}
 \end{center}
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_sparse.pdf}
-\caption{Benchmark\figurelabel{benchmark-sparse}}
+\caption{The same as \Figure{benchmark} but using the \texttt{SparseLU} solver
+    from \project{Eigen} \citep{Guennebaud:2010}.
+    \figurelabel{benchmark-sparse}}
 \end{center}

### 2017-03-15 paper/ms.tex
benchmark words [ci skip]

@@
 In this \sectionname, we empirically investigate the performance and scaling
-of three different algorithms for solving this extended systems:
+of three different algorithms for solving this extended system:
 
@@
 
-\item A general sparse LU solver that exploits the sparsity but not the band
-    structure \citep{Guennebaud:2010}.
-
-\item A simple algorithm for computing the LU decomposition for banded
-    matrices using Gaussian elimination \citep{Press:1992, Press:2007},
-
-\item The general banded LU decomposition implementation from
+\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
+    for banded matrices using Gaussian elimination \citep{Press:1992,
+    Press:2007},
+
+\item \texttt{lapack}: The general banded LU decomposition implementation from
     LAPACK\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs} methods
-        from LAPACK.} \citep{Anderson:1999} using optimized BLAS routines, and
+    from LAPACK.} \citep{Anderson:1999} using optimized BLAS
+    routines\footnote{The experiments below use the Intel Math Kernel Library
+    (\project{MKL}): \url{https://software.intel.com/en-us/intel-mkl}.}, and
+
+\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
+    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
+    sparsity but not the band structure.
 
@@
 bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
-We find that, while the vanilla banded solver scales as expected, the LAPACK
-algorithm with tuned BLAS routines scales empirically as $\mathcal{O}(N\,J^2)$
-empirically and offers the fastest solves for $J \gtrsim 8$ on all platforms
-that we tested.
+We find that, while the \texttt{vanilla} solver scales as expected, the
+\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
+offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
+
+The benchmark experiments shown here were performed on a MacBook Pro with two
+2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
+CPUs and running Ubuntu.
 

### 2017-03-15 paper/ms.tex
summary of benchmark/scaling results

@@
 CPUs and running Ubuntu.
-
-NNZ: $(10\,J + 1)\,N - 4\,J + 1$
+\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
+with $N$ and $J$ using the \texttt{vanilla} solver.
+As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
+in $J$ for large $J$.
+\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
+the \texttt{lapack} and \texttt{sparse} solvers respectively.
+For each of these optimized solvers, the empirical scaling is
+$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
+Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
+efficient than the other algorithms.
+The real world performance of \celerite\ depends on the specific platform,
+hardware, and \project{LAPACK}/\project{BLAS} implementation but we have found
+qualitatively similar results across popular platforms and state-of-the-art
+libraries.
 
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin.pdf}
-\caption{A benchmark showing the computational scaling of \celerite\ with a
-    vanilla band solver as a function of the number of data points and the
-    number of terms.
+\caption{A benchmark showing the computational scaling of \celerite\ using the
+    \texttt{vanilla} band solver as a function of the number of data points
+    and the number of terms.
     \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_lapack.pdf}
-\caption{The same as \Figure{benchmark} but using the band solver from
-    \project{LAPACK} with the \project{MKL} \project{BLAS} library.
+\caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
+    with the \project{MKL} \project{BLAS} library.
     \figurelabel{benchmark-lapack}}
@@
 \includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_sparse.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{SparseLU} solver
+\caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver
     from \project{Eigen} \citep{Guennebaud:2010}.

### 2017-03-15 paper/ms.tex
moving benchmark figures to paper directory

@@
 \begin{center}
-\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin.pdf}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
 \caption{A benchmark showing the computational scaling of \celerite\ using the
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_lapack.pdf}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin_lapack.pdf}
 \caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
@@
 \begin{center}
-\includegraphics[width=\textwidth]{../examples/benchmark/benchmark_darwin_sparse.pdf}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin_sparse.pdf}
 \caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver

### 2017-03-15 paper/ms.tex
typos and refs

@@
 
-\section{Summary}
-
 % DFM: I'm going to redistribute the following in other sections.
@@
 
-Although we have in mind application of this fast method to stellar variability,
-the method is general for one-dimensional GP problems, and may be applied to
-other problems.  Accreting black holes show time series which
-may be modeled by correlated noise \citep{Kelly:2014};  indeed, this was the
-motivation for the original technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
+Although we have in mind applications of this fast method to model stellar
+variability, the method is general for one-dimensional GP problems, and may be
+applied to other problems.
+Accreting black holes show time series which may be modeled by correlated
+noise \citep{Kelly:2014};  indeed, this was the motivation for the original
+technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
 This approach may be broadly used for characterizing quasar variability
@@
 
-There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
-the variability of an astrophysical system to infer it properties; 2).\
-accounting for astrophysical variability as a source of noise when trying to detect
-additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
-times.  Some examples of the first are determining the characteristic
-variability timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic variation of a star
-\citep{Brewer:2009,Corsaro:2014}, measurement of granulation noise \citep{Bastien:2013,Kallinger:2016},
-detection of quasi-periodic variability in a high-energy source \citep{McAllister:2016},
-or classification of variable objects \citep{Zinn:2016}.  Examples of the
-second are detecting transiting exoplanets and correcting for stellar activity in
-radial velocity measurements \citep{Haywood:2014,Rajpaul:2015}. Examples of the third are
-measuring time delays in
-multiply-imaged gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of AGN.
-
-Our background is in studying transiting exoplanets, a field which has only recently begun
-to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
-One promising early attempt at accounting for correlated noise was the wavelet approach
-of \citet{Carter:2009}.  Their technique allowed for a power-law spectrum for the
-noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
-physically-accurate, description of stellar variability.  Their approach runs in
-${\cal O}(N)$ time for $\alpha=1$, which is referred to as `flicker', $1/r$, or pink
-noise;  however, this particular type of noise overpredicts the power spectrum at low
-frequency compared to the Solar granulation power spectrum.  It also cannot describe
-quasi-periodic noise.
+There are three primary applications we envision for the \celerite\ formalism:
+1).\ modeling the variability of an astrophysical system to infer it
+properties; 2).\ accounting for astrophysical variability as a source of noise
+when trying to detect additional phenomena; 3).\ interpolating or
+extrapolating variability to future or missing times.
+Some examples of the first are determining the characteristic variability
+timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
+variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
+granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
+quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
+classification of variable objects \citep{Zinn:2016}.
+Examples of the second are detecting and characterizing transiting exoplanets
+\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
+Grunblatt:2016, Luger:2016}
+and correcting for stellar activity in radial velocity measurements
+\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
+Examples of the third are measuring time delays in multiply-imaged
+gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
+AGN.
+
+Our background is in studying transiting exoplanets, a field which has only
+recently begun to adopt full covariance matrices in analyzing the noise in
+transiting planet lightcurves.
+One promising early attempt at accounting for correlated noise was the wavelet
+approach of \citet{Carter:2009}.
+Their technique allowed for a power-law spectrum for the noise scaling as
+$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
+physically-accurate, description of stellar variability.
+Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
+as `flicker', $1/r$, or pink noise;  however, this particular type of noise
+overpredicts the power spectrum at low frequency compared to the Solar
+granulation power spectrum.
+It also cannot describe quasi-periodic noise.
 

### 2017-03-15 paper/ms.tex
intro to simulated examples

@@
 \section{Examples with simulated data}
+
+To demonstrate the application of \celerite, we will start by inferring
+posterior constraints on the parameters of a GP model applied to several
+simulated datasets with known properties.
+In the following section, we will expand on these examples by applying
+\celerite\ to real datasets.
+In the first example, we demonstrate that \celerite\ can be used to measure
+the power spectrum of a process when the data are generated from a \celerite\
+model.
+In the second example, we demonstrate that \celerite\ can be used as an
+effective model even if the true process cannot be represented in the space of
+allowed models.
+This is a interesting example because, when analyzing real data, we rarely
+have any fundamental reason to believe that the data were generated by a GP
+model with a specific kernel.
+Even in these cases, GPs can be useful effective models and \celerite\
+provides computational advantages over other GP methods.
 

### 2017-03-15 paper/ms.tex
typos and edits

@@
 
+\vspace{1.5em}
+All of the code used in this project is available from
+\url{https://github.com/dfm/celerite} under the MIT open-source software
+license.
+This code (plus some dependencies) can be run to re-generate all of the
+figures and results in this paper; this version of the paper was generated
+with git commit \texttt{\githash} (\gitdate).
+
+\acknowledgments
+It is a pleasure to thank
+Sivaram Ambikasaran,
+Megan Bedell,
+Will Farr,
+Sam Grunblatt,
+David W.\ Hogg, and
+Dan Huber
+for helpful contributions to the ideas and code presented here.
+
+EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
+NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
+by NASA under cooperative agreement NNH05ZDA001C.
+
+This research made use of the NASA \project{Astrophysics Data System} and the
+NASA Exoplanet Archive.
+The Exoplanet Archive is operated by the California Institute of Technology,
+under contract with NASA under the Exoplanet Exploration Program.
+
+This paper includes data collected by the \kepler\ mission. Funding for the
+\kepler\ mission is provided by the NASA Science Mission directorate.
+We are grateful to the entire \kepler\ team, past and present.
+
+These data were obtained from the Mikulski Archive for Space Telescopes
+(MAST).
+STScI is operated by the Association of Universities for Research in
+Astronomy, Inc., under NASA contract NAS5-26555.
+Support for MAST is provided by the NASA Office of Space Science via grant
+NNX13AC07G and by other grants and contracts.
+
+\facility{Kepler}
+\software{%
+     \project{corner.py} \citep{Foreman-Mackey:2016},
+     \project{Eigen} \citep{Guennebaud:2010},
+     \project{emcee} \citep{Foreman-Mackey:2013},
+     \project{george} \citep{Ambikasaran:2016},
+     \project{LAPACK} \citep{Anderson:1999},
+     \project{matplotlib} \citep{Hunter:2007},
+     \project{numpy} \citep{Van-Der-Walt:2011},
+     \project{transit} \citep{Foreman-Mackey:2016a},
+     \project{scipy} \citep{Jones:2001}.
+}
+
 \appendix
@@
 In the case of $J$ \celerite\ terms, we can check for negative values of the
-PSD by solving for the roots of the power spectrum, abbreviating with
-$z = \omega^2$:
+PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
+abbreviating with $z = \omega^2$:
 \begin{equation}
-P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
+S(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
 \end{equation}
@@
 \end{eqnarray}
-The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
+The denominators of each term are positive, so we can multiply through by
+$\prod_j \left(z^2+s_jz + t_j\right)$ to find
 \begin{equation}
-P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
+Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
+    t_k\right) = 0\quad,
 \end{equation}
-which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
-which may be solved exactly for the roots.
-
-For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
-real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
-$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
-$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.
-
-We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
-give us the signs of these polynomials evaluated at $z=0$.
-Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial
-as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where
-$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.
-
-With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each
-of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign
-changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given
-by $N_{+}=\sigma(0)-\sigma(\infty)$.
-
-We have checked that this procedure works for a wide range of parameters, and we find that it robustly
-matches the number of positive real roots which we evaluated numerically.
-
-The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic
-manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
-likelihood may be set to zero.
-
-We have implemented this algorithm in subroutines that may be called to check for positive-definiteness
-of a kernel with some values of $a_j$ or $b_j$ negative.
-%Sturm's theorem applied to numerator of power spectrum.
-
-\vspace{1.5em}
-All of the code used in this project is available from
-\url{https://github.com/dfm/celerite} under the MIT open-source software
-license.
-This code (plus some dependencies) can be run to re-generate all of the
-figures and results in this paper; this version of the paper was generated
-with git commit \texttt{\githash} (\gitdate).
-
-\acknowledgments
-It is a pleasure to thank
-Sivaram Ambikasaran,
-Megan Bedell,
-Will Farr,
-Sam Grunblatt,
-David W.\ Hogg, and
-Dan Huber
-for helpful contributions to the ideas and code presented here.
-
-EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
-NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
-by NASA under cooperative agreement NNH05ZDA001C.
-
-This research made use of the NASA \project{Astrophysics Data System} and the
-NASA Exoplanet Archive.
-The Exoplanet Archive is operated by the California Institute of Technology,
-under contract with NASA under the Exoplanet Exploration Program.
-
-This paper includes data collected by the \kepler\ mission. Funding for the
-\kepler\ mission is provided by the NASA Science Mission directorate.
-We are grateful to the entire \kepler\ team, past and present.
-
-These data were obtained from the Mikulski Archive for Space Telescopes
-(MAST).
-STScI is operated by the Association of Universities for Research in
-Astronomy, Inc., under NASA contract NAS5-26555.
-Support for MAST is provided by the NASA Office of Space Science via grant
-NNX13AC07G and by other grants and contracts.
-
-\facility{Kepler}
-\software{%
-     \project{corner.py} \citep{Foreman-Mackey:2016},
-     \project{Eigen} \citep{Guennebaud:2010},
-     \project{emcee} \citep{Foreman-Mackey:2013},
-     \project{george} \citep{Ambikasaran:2016},
-     \project{LAPACK} \citep{Anderson:1999},
-     \project{matplotlib} \citep{Hunter:2007},
-     \project{numpy} \citep{Van-Der-Walt:2011},
-     \project{transit} \citep{Foreman-Mackey:2016a},
-     \project{scipy} \citep{Jones:2001}.
-}
+which is a polynomial with order $2\,(J-1)+1$.
+With $J=2$, this yields a cubic equation which may be solved exactly for the
+roots.
+
+For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
+allows one to determine whether there are any real roots within the range
+$(0,\,\infty]$.
+We first construct $Q_0(z)$ and it's derivative $Q_1(z) = {Q_0}^\prime(z)$,
+and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
+\begin{eqnarray}
+Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
+\end{eqnarray}
+where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
+dividing $p(z)$ by $q(z)$.
+
+We evaluate the coefficients of each of the polynomial in the series by
+evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
+of these polynomials evaluated at $z=0$.
+Likewise, we evaluate the coefficients of the largest order term in each
+polynomial which gives the sign of the polynomial as $z \rightarrow \infty$,
+$f_\infty$.
+With the series of coefficients $f_0$ and $f_\infty$, we then determine how
+many times the sign changes in each of these, where $\sigma(0)$ is the number
+of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
+at $z \rightarrow \infty$.
+The total number of real roots in the range
+$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.
+
+We have checked that this procedure works for a wide range of parameters, and
+we find that it robustly matches the number of positive real roots which we
+evaluated numerically.
+The advantage of this procedure is that it does not require computing the
+roots, but only carrying out algebraic manipulation of polynomials to
+determine the number of positive real roots.
+If a non-zero real root is found, the likelihood may be set to zero.
 

### 2017-03-15 paper/ms.tex
updating rotation period figures

@@
 \include{vc}
+\include{figures/rotation}
+
 \usepackage{microtype}
-
 \usepackage{url}
@@
 inferred with a much larger dataset.
-Looking forward to \tess \citep{Campante:2016}, we will infer $\nu_\mathrm{max}$ and $\Delta\nu$
-using only one month of \kepler\ data and compare our results to the results
-inferred from the full 4 year baseline of the mission.
+Looking forward to \tess\ \citep{Campante:2016}, we will infer
+$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
+compare our results to the results inferred from the full 4 year baseline of
+the \kepler\ mission.
 For KIC~11615890, the published asteroseismic parameters based on the full
@@
 as a drop-in replacement for a substantial gain in computational efficiency.
+
+To demonstrate this model, we
+
+\rotationperiod
 

### 2017-03-15 paper/figures/rotation.py

@@ -3,19 +3,20 @@
 
 from __future__ import division, print_function
 
+import kplr
 import emcee3
 import fitsio
 import numpy as np
 import matplotlib.pyplot as plt
 from scipy.optimize import minimize
 
-from plot_setup import setup, get_figsize, COLORS
+from celerite.plot_setup import setup, get_figsize, COLORS
 
 import celerite
 from celerite import terms
 
 np.random.seed(123)
-setup()
+setup(auto=True)
 
 # Define the custom kernel
 class RotationTerm(terms.Term):
@@ -38,29 +39,49 @@ class RotationTerm(terms.Term):
         )
 
 # Load the data
-data = fitsio.read("data/kplr001430163-2013011073258_llc.fits")
-m = data["SAP_QUALITY"] == 0
-m &= np.isfinite(data["TIME"]) & np.isfinite(data["PDCSAP_FLUX"])
-t = np.ascontiguousarray(data["TIME"][m], dtype=np.float64)
-t -= np.min(t)
-y = np.ascontiguousarray(data["PDCSAP_FLUX"][m], dtype=np.float64)
-yerr = np.ascontiguousarray(data["PDCSAP_FLUX_ERR"][m], dtype=np.float64)
-
-# Normalize the data
-mean = np.median(y)
-y = (y / mean - 1.0) * 1e3
-yerr *= 1e3 / mean
+kicid = 1430163
+client = kplr.API()
+star = client.star(kicid)
+x = []
+y = []
+yerr = []
+for lc in star.get_light_curves(short_cadence=False)[4:6]:
+    data = lc.read()
+    x0 = data["TIME"]
+    y0 = data["PDCSAP_FLUX"]
+    m = (data["SAP_QUALITY"] == 0) & np.isfinite(x0) & np.isfinite(y0)
+    x.append(x0[m])
+    mu = np.median(y0[m])
+    y.append((y0[m] / mu - 1.0) * 1e3)
+    yerr.append(1e3 * data["PDCSAP_FLUX_ERR"][m] / mu)
+x = np.concatenate(x)
+y = np.concatenate(y)
+yerr = np.concatenate(yerr)
+inds = np.argsort(x)
+t = np.ascontiguousarray(x[inds], dtype=float)
+y = np.ascontiguousarray(y[inds], dtype=float)
+yerr = np.ascontiguousarray(yerr[inds], dtype=float)
 
 # Set up the GP model
 kernel = RotationTerm(
[…]

### 2017-03-15 paper/figures/transit_sample.py

@@ -5,7 +5,6 @@ from __future__ import division, print_function
 
 import os
 import sys
-import corner
 import emcee3
 import pickle
 import fitsio

### 2017-03-15 paper/ms.tex
words about rotation example

@@
 
-To demonstrate this model, we
-
-\rotationperiod
+To demonstrate this model, we fit a \celerite\ model with a kernel given by
+\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
+This star has a published rotation period of $3.88 \pm 0.58$ measured using
+traditional periodogram and autocorrelation function approaches applied to
+\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
+We used \emcee\ to sample to posterior probability density for the four
+parameters in \eq{rot-kernel} conditioned on two quarters of \kepler\ data and
+find a constraint on the rotation period of $P = \rotationperiod$ in good
+agreement with the literature value.
+\Figure{rotation} shows a subset of the data used in this example and the
+posterior inferences of the PSD and autocorrelation function of the process.
+\Figure{rotation-period} shows the marginalized posterior distribution for the
+rotation period.
 
@@
 \caption{Inferred constraints on a quasiperiodic GP model using the covariance
-    function in \eq{rot-kernel} and one quarter of \kepler\ data.
+    function in \eq{rot-kernel} and two quarters of \kepler\ data.
 (left) The \kepler\ data (black points) and the maximum likelihood model
-    prediction (blue curve).
+    prediction (blue curve) for a 50~day subset of the data used.
     The solid blue line shows the predictive mean and the blue contours show

### 2017-03-15 paper/ms.tex
some words about real examples and captions

@@
 parameters describing the limb-darkening profile of the star \dfmtodo{CITE}.
-
-KIC~1430163 with simulated transit.
-\Figure{transit-ml} and \Figure{transit-corner}.
+As in the previous example, we model the stellar variability using a GP model
+with a kernel given by \eq{rot-kernel}.
+
+To demonstrate this example, we take a month-long segment of the \kepler\
+light curve for KIC~1430163~--~the target from the previous example~--~and
+multiply it by a simulated transit model with known parameters.
+The top panel of \Figure{transit-ml} shows the data including the simulated
+transit.
+The bottom panel shows the same data with the maximum likelihood prediction
+for the stellar variation subtracted.
+To find this model, we maximized the likelihood function in \eq{gp-likelihood}
+with respect to the transit parameters $\bvec{\theta}$ and the variability
+parameters $\bvec{\alpha}$ simultaneously.
+\Figure{transit-corner} shows the marginalized posterior constraints on the
+physical properties of the planet compared to the true values.
+This procedure produces estimates of the planet parameters that are consistent
+with the true values using \celerite\ as an effective model for the stellar
+variability.
 
@@
 \includegraphics[width=\textwidth]{figures/transit-ml.pdf}
-\caption{\figurelabel{transit-ml}}
+    \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
+    KIC~1430163 with a synthetic transit model injected (black points) and the
+    maximum likelihood model for the stellar variability and the transit model
+    (blue line).
+    \emph{(bottom)} The maximum likelihood ``de-trendeding'' of the data in
+    the top panel.
+    In this panel, the maximum likelihood model for the stellar variability
+    has been subtracted to leave only the transits.
+    The de-trended light curve is shown by black error bars and the maximum
+    likelihood transit model is shown as a blue line.
+    \figurelabel{transit-ml}}
 \end{center}
@@
 \includegraphics[width=\textwidth]{figures/transit-corner.pdf}
-\caption{\figurelabel{transit-corner}}
+\caption{The marginalized posterior constraints on the physical parameters of
+    the planet transiting during the light curve shown in the top panel of
+    \Figure{transit-ml}.
+    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
+    regions in the marginalized planes and the histograms along the diagonal
+    show the marginalized posterior for each parameter.
+    The true values used in the simulation are indicated by blue lines.
+    For each parameter, the inference is consistent with the true value.
+    \figurelabel{transit-corner}}
 \end{center}

### 2017-03-16 paper/ms.tex
Adding @sivaramambikasaran as a co-author

@@
 
+\edef \iss {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\iss}{Department of Computational and Data Sciences,
+                    Indian Institute of Science, Bangalore, India}
+
 \author{%
     Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw},
-    Eric~Agol\altaffilmark{\uw}, and
-    Ruth~Angus\altaffilmark{\simons,\columbia}
+    Eric~Agol\altaffilmark{\uw},
+    Ruth~Angus\altaffilmark{\simons,\columbia}, and
+    Sivaram~Ambikasaran\altaffilmark{\iss}
 }

### 2017-03-16 paper/ms.tex
edits to abstract

@@
 
-We present a scalable method for Gaussian Process regression in one dimension
-with a specific emphasis on large astronomical time-series data sets.
-This method, which we christen \celerite, can be applied to any Gaussian Process model where the spectral
-density can be expressed as any general mixture of damped sinusoid functions.
-We present a mathematical description of \celerite,
-its relation to the stochastically-driven, damped simple harmonic oscillator,
-and details of the implementation.  We give examples of \celerite\ applied to both
-simulated and actual astronomical time-series, using it to estimate the time-series
-power spectrum, while simultaneously accounting for correlated noise when inferring
-model parameters.  These examples include Bayesian estimation of the strength
-of stellar granulation, stellar rotation periods, asteroseismic spectra, and
-transiting planet light curve parameters.  The model is flexible, fast, and
-general, with potential for application to a wide range of astronomical problems,
-for which we provide open-source software that implements \celerite\ in
-\project{C++}, \project{Python}, and \project{Julia}.
+\dfmtodo{Make this less douchey:}
+In the current era of large-scale time domain astronomy, scalable methods for
+probabilistic inference are of great interest.
+We present a method for Gaussian Process regression in one dimension with a
+computational cost that scales linearly with the size of the dataset.
+This method exploits structure in the problem when the covariance function is
+expressed as a mixture of complex exponentials but it does not require evenly
+spaced observations or homoscedastic noise.
+This form of the covariance arises naturally when the process is a mixture of
+stochastically-driven damped harmonic oscillators~--~providing a physical
+motivation for and interpretation of this choice~--~but we also demonstrate
+that this function can also be a sufficient effective model in many other
+cases.
+We present a mathematical description of the method, details of the
+implementation, and compare it to existing scalable Gaussian Process methods.
+We demonstrate the application of this method by applying it to simulated and
+real astronomical time series datasets.
+These demonstrations are examples of probabilistic inference of stellar
+rotation periods, asteroseismic oscillation spectra, and transiting planet
+parameters.
+The method is flexible, fast, and most importantly, interpretable.
+There are a wide range of potential applications within astronomical data
+analysis and beyond.
+We provide well tested and documented open-source implementations of this
+method in \project{C++}, \project{Python}, and \project{Julia}.
 

### 2017-03-17 paper/ms.tex
a few edits from my first pass

@@
 the number of data points $\mathcal{O}(N^3)$ and in the current era of large
-time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
-observations each~---~this cost is prohibitive.
-
-In this paper, we present a class of GP models that enable likelihood
-calculations that scale linearly with the number of data points
-$\mathcal{O}(N)$ for one dimensional data sets.
+time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
+$\sim10^{3-5}$ observations each~---~this cost is prohibitive.
+
+In this paper, we present a method for computing a class of GP models that
+scales linearly with the number of data points $\mathcal{O}(N)$ for one
+dimensional data sets.
 This method is a generalization of a method developed by
@@
 This method is especially appealing compared to other similar methods~--~we
-will return to these below~--~because it is exact, flexible, robust, simple,
-and fast.
+will return to these below~--~because it is exact, flexible, simple, and fast.
 
@@
 application on various real and simulated data sets.
-Alongside this paper, we have released efficient and well-tested
-implementations of this method written in \project{C++}, \project{Python}, and \project{Julia}.
-These implementations are available online at \project{GitHub}
+Alongside this paper, we have released well tested and documented open source
+implementations written in \project{C++}, \project{Python}, and
+\project{Julia}.
+These are available online at \project{GitHub}
 \url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
@@
 models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
-covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
-parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
-respectively.
+covariance, autocorrelation, or ``kernel'' function
+$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the parameters
+$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
 Under this model, the log-likelihood of observing a dataset
@@
 maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
-$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
-CITE}.
+$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
 Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
 can be obtained by multiplying the likelihood by a prior
-$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
-\dfmtodo{CITE}) algorithm to sample from the posterior probability density.
-
-The application of GP models is generally limited to small datasets because
-the computational cost of computing the inverse and determinant of the matrix
-$K_\bvec{\alpha}$ scales as the cube of the number of data points $N$,
-$\mathcal{O}(N^3)$.
+$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
+algorithm to sample from the posterior probability density.
+
+GP models have been widely used across the physical sciences but their
+application is generally limited to small datasets because the computational
+cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
+is $\mathcal{O}(N^3)$.
+In other words, this cost is proportional to the cube of the number of data
+points $N$.
 This means that for large datasets, every evaluation of the likelihood will
@@
 \end{eqnarray}
-This PSD is the sum of two Lorentzian or Cauchy distributions with width
-$c_j$ centered on $\omega = \pm d_j$.
+This is the sum of two Lorentzian or Cauchy distributions with width $c_j$
+centered on $\omega = \pm d_j$.
 This model can be interpreted intuitively as a quasiperiodic oscillator with
@@
 
-It's worth noting that the product of two terms of the form found inside the
+Finally, we note that the product of two terms of the form found inside the
 sum in \eq{celerite-kernel} can also be re-written as a sum with updated
@@
 \end{eqnarray}
+Therefore, the method described in the following section can be used to scale
+inferences to large datasets for any model where the kernel function is a
+general sum or product of \celerite\ terms.
 
@@
 \citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
-elements are given by
-\begin{eqnarray}
-K_{nm} = k(\tau_{nm})
-\end{eqnarray}
-for $k(\tau_{nm})$ given by equation \eq{kernel-simple} could be computed
+elements are given by equation \eq{kernel-simple} could be computed
 efficiently by taking advantage of the structure of this covariance function
@@
 
-To provide some insight for this method, we will start by working through a
-simple example.
+To provide some insight for this method, we will follow
+\citet{Ambikasaran:2015} and start by working through a simple example.
 In this case, we'll assume that we have three data points
@@
 \end{eqnarray}
-This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
-and $d=0$.
+This is the result of setting $J=1$, $b=0$, and $d=0$ in
+\eq{celerite-kernel-complex} and this is the model that was studied by
+\citet{Rybicki:1995}.
 To demonstrate this method, we will write out the full system of
@@
 \end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
-started with, it is now has a sparse banded structure that can be exploited to
-solve the system efficiently.
-In particular, band matrix solvers are available that can perform an
-LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
+Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
+with, it is now sparse with banded structure that can be exploited to solve
+the system efficiently.
+In particular, sparse solvers are available that can perform a
+LU-decomposition of matrices like this in $\mathcal{O}(N)$
 operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
@@
 A special case of the \celerite\ model of great physical interest is a
-stochastically-driven simple harmonic oscillator.
+stochastically-driven damped simple harmonic oscillator.
 The differential equation for this system is
@@
 force.
-In the limit of an infinite time series and white random forcing, the PSD of
-this equation is given by \citep{Anderson:1990}
+If $\epsilon(t)$ is white noise, the PSD of this process is given by
+\citep[for example][]{Anderson:1990}
 \begin{equation}\eqlabel{sho-psd}
@@
 of $Q$.
-This figure demonstrates that for $Q \le 1/2$, the model has no oscillatory
-behavior and that for large $Q$, the shape of the PSD near the peak frequency
-approaches a Lorentzian.
+From this figure, it is clear that for $Q \le 1/2$, the model has no
+oscillatory behavior and that for large $Q$, the shape of the PSD near the
+peak frequency approaches a Lorentzian.
 
@@
 The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ mission \citep{Gilliland:2010,Chaplin:2011,Chaplin:2013,Stello:2013}
-and asteroseismology is a
-key science driver for many of the upcoming large scale photometric surveys
-\citep{Campante:2016,Rauer:2014,Gould:2015}.
+curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
+Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
+many of the upcoming large scale photometric surveys \citep{Campante:2016,
+Rauer:2014, Gould:2015}.
 Most asteroseismic analyses have been limited to relatively high
@@
 the measurement uncertainties to the constraints on physical parameters
-(\dfmtodo{CITE}) and more sophisticated methods are computationally expensive
-and they scale poorly to the current state-of-the-art datasets \citep{Brewer:2009}.
+(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
+function in the time domain scale poorly to state-of-the-art datasets
+\citep{Brewer:2009}.
 
@@
 probabilistic model that can be evaluated efficiently even for large datasets.
-In practice, we will model the star as a mixture of stochastically driven
+In practice, one would model the star as a mixture of stochastically-driven
 simple harmonic oscillators where the amplitudes and frequencies of the
 oscillations are computed using a physical model and evaluate the probability
-of the observed dataset using a Gaussian Process with a PSD given by a sum of
-terms given by \eq{sho-psd}.
+of the observed dataset using a GP where the PSD is a sum of terms given by
+\eq{sho-psd}.
 This gives us a method of computing the likelihood function for the parameters
-of the physical model in $\mathcal{O}(N)$ operations and this can be combined
-with standard non-linear optimization or posterior sampling methods to infer
-constraints on the parameters for the given dataset.
+of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$)
+\emph{conditioned on the observed time series} in $\mathcal{O}(N)$ operations.
+In other words, this provides us with a computationally efficient method for
+performing rigorous probabilistic inference of asteroseismic parameters in the
+time domain.
+We expect that this has the potential to push asteroseismic analysis to lower
+signal-to-noise datasets and we hope to revisit this idea in a subsequent
+paper.
 
 To demonstrate this method, we will use a very simple heuristic model based on
-\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
+\dfmtodo{CITE} where the model PSD is given by a mixture of 9 components with
 amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
 several nuisance parameters.
-The first term is used to capture the granulation ``background'' using
-\eq{granulation-psd} with two free parameters $S_g$ and $\omega_g$.
+The first two terms are used to capture the granulation ``background''
+\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
+$S_g$ and $\omega_g$ in each term.
 The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
@@
 the \kepler\ mission.
-For KIC~11615890, the published asteroseismic parameters based on the full
-dataset are \citep{Pinsonneault:2014}
+For KIC~11615890, the published asteroseismic parameters based on several
+years of \kepler\ observations are \citep{Pinsonneault:2014}
 \begin{eqnarray}
@@
 \caption{A comparison between the empirical PSD and the posterior inference of
-the PSD as a mixture of stochastically driven simple harmonic oscillators.
+the PSD as a mixture of stochastically-driven simple harmonic oscillators.
 (top) The periodogram of the \kepler\ light curve for KIC~11615890 computed

### 2017-03-17 paper/ms.tex
some words about alternative methods

@@
 
-\section{celerite as a model of stellar variations}
+\section{celerite as a model of stellar variations}\sectlabel{sho}
 
@@
 To demonstrate this method, we will use a very simple heuristic model based on
-\dfmtodo{CITE} where the model PSD is given by a mixture of 9 components with
+\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
 amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
 several nuisance parameters.
-The first two terms are used to capture the granulation ``background''
+The first term is used to capture the granulation ``background''
 \citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
-$S_g$ and $\omega_g$ in each term.
+$S_g$ and $\omega_g$.
 The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!p]
 \begin{center}
@@
 
-Toeplitz, KISS-GP, CARMA, HODLR.
-
-Limitations of celerite: one-dimension, stationary, etc.
+There are many other methods of scaling GP models to large datasets and in
+this section we will explore some comparisons between \celerite\ and some
+other popular methods and discuss some use cases where \celerite\ shines and
+others where it does not.
+Scalable GP methods tend to fall into two categories: approximate and
+restrictive.
+\celerite\ falls into the latter category because, while the method is exact,
+it requires a specific choice of stationary kernel function and it can only be
+used in one-dimension.
+
+\dfmtodo{Add discussion of original R\&P method here.}
+
+Another commonly used restrictive method uses the fact that, in the limit of
+evenly-spaced data and homoscedastic uncertainties, the covariance matrix is
+``Toeplitz'' \citep[for example][]{Dillon:2013}.
+There are exact methods for solving Toeplitz matrix equations that scale as
+$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
+$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N$
+\citep{Wilson:2014}.
+The Toeplitz method is, in some ways, more flexible than \celerite\ because it
+can be used with any stationary kernel but it requires uniformly spaced data
+and the scaling is worse than \celerite\ so it will, in general, be less
+efficient when applied to large datasets.
+
+The continuous autoregressive moving average (CARMA) models introduced into
+the astrophysics literature by \citet{Kelly:2014} share many features with
+\celerite.
+Like \celerite, the likelihood function for a CARMA model conditioned on a
+sorted one-dimensional dataset can be solved in $\mathcal{O}(N)$ using a
+recursive method.
+The kernel function for CARMA$(J,\,K)$ model is \citep{Kelly:2014}
+\begin{eqnarray}\eqlabel{carma-kernel}
+k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
+\end{eqnarray}
+where
+\begin{eqnarray}\eqlabel{carma-coeff}
+A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
+    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
+    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
+\end{eqnarray}
+and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
+the model.
+Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
+every CARMA model corresponds to a \celerite\ and the parameters $a_j$, $b_j$,
+$c_j$, and $d_j$ can be easily computed analytically.
+The inverse statement is not true, however, for any $J > 1$.
+This means that \celerite\ could be used to compute any CARMA model but using
+CARMA to compute the likelihood of a \celerite\ model would require solving
+\eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
+This is an interesting distinction because, while the computational scaling of
+CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
+faster and the memory requirements are smaller.
+However, as discussed in \sect{sho}, \celerite\ models have physical a
+interpretation as a mixture of stochastically-driven damped harmonic
+oscillators.
+This enabled, for example, the use of \celerite\ as a model for asteroseismic
+oscillations.
+A similar example using CARMA would necessarily be qualitative.
+
+Approximate:
+- inducing point methods \citep{Quinonero-Candela:2007}
+- KISS-GP \citep{Wilson:2015, Wilson:2015a}
+- HODLR \citep{Ambikasaran:2016}
 

### 2017-03-19 paper/ms.tex
a few words for positive definiteness

@@
     KIC~1430163 with a synthetic transit model injected (black points) and the
-    maximum likelihood model for the stellar variability and the transit model
+    maximum likelihood model for the stellar variability
     (blue line).
-    \emph{(bottom)} The maximum likelihood ``de-trendeding'' of the data in
+    \emph{(bottom)} The maximum likelihood ``de-trending'' of the data in
     the top panel.
@@
 It is a pleasure to thank
-Sivaram Ambikasaran,
 Megan Bedell,
@@
 
-\section{Ensuring positive definiteness: Sturm's theorem}
+\section{Ensuring positive definiteness}
+
+For a GP kernel to be valid, it must produce a positive definite covariance
+matrix for all input coordinates.
+For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
+Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
+Fourier transform of a positive finite measure.
+This means that the power spectrum of a positive definite kernel must be
+positive for all frequencies.
+
+Using \eq{celerite-psd}, we find that for a single \celerite\ term, this
+requirement is met when
+\begin{eqnarray}\eqlabel{psd-req}
+\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
+{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2} > 0 \quad.
+\end{eqnarray}
+The denominator is positive for all $c_j \ne 0$ and it can be shown that, when
+$c_j=0$, \eq{psd-req} is satisfied for all $\omega \ne d_j$ where the power is
+identically zero.
+Therefore, when $c_j \ne 0$, we require that the numerator is positive for all
+$\omega$.
 

### 2017-03-20 paper/ms.tex

@@
 
-\dfmtodo{Make this less douchey:}
-In the current era of large-scale time domain astronomy, scalable methods for
-probabilistic inference are of great interest.
+The growing field of large-scale time domain astronomy requires methods for
+probabilistic inference that scale well with amount of data.
 We present a method for Gaussian Process regression in one dimension with a
@@
 This method exploits structure in the problem when the covariance function is
-expressed as a mixture of complex exponentials but it does not require evenly
-spaced observations or homoscedastic noise.
-This form of the covariance arises naturally when the process is a mixture of
+expressed as a mixture of complex exponentials, without requiring evenly
+spaced observations or uniform noise.
+This form of covariance arises naturally when the process is a mixture of
 stochastically-driven damped harmonic oscillators~--~providing a physical
 motivation for and interpretation of this choice~--~but we also demonstrate
-that this function can also be a sufficient effective model in many other
-cases.
-We present a mathematical description of the method, details of the
-implementation, and compare it to existing scalable Gaussian Process methods.
+that it is effective in many other cases.
+We present a mathematical description of the method, the details of the
+implementation, and a comparison to existing scalable Gaussian Process methods.
 We demonstrate the application of this method by applying it to simulated and
@@
 In particular, the measurements don't need to be evenly spaced and the
-uncertainties can be heteroscedastic.
+uncertainties can be heteroscedastic (non-uniform).
 This method is especially appealing compared to other similar methods~--~we
@@
 This means that the power spectrum of a positive definite kernel must be
-positive for all frequencies.
+positive for all frequencies.  This seems true intuitively:  since the
+power spectrum is the square of the Fourier transform of the data time
+series, it must be non-negative.
 
@@
 
-The power spectrum is computed from taking the square of the Fourier transform
-of the data time series; hence, the power-spectrum must be non-negative.  In
-constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
-then it is possible for the power spectrum to go negative, which violates the
-definition of the power spectrum.
-Consequently, if any of these coefficients is negative, it is necessary to
-check that the power spectrum is still non-negative.  Note that a non-negative
-power-spectrum does not require the auto-correlation
-function to be positive for all $\tau$.  The positive power spectrum is
-related to the positive eigenvalues of the covariance matrix which are
-necessary for a positive-definite matrix in limiting cases
-\citep{Messerschmitt:2006}.  We find empirically that requiring an everywhere
-positive power spectrum results in positive eigenvalues for the covariance
-matrix, and so we describe here how to ensure a positive power spectrum using
-Sturm's theorem.
+%The general case with multiple \celerite\ terms can be checked using Sturm's
+%theorem. The power spectrum is computed from taking the square of the Fourier 
+%transform of the data time series; hence, the power-spectrum must be non-negative.  
+%In constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
+%then it is possible for the power spectrum to go negative, which violates the
+%definition of the power spectrum.
+%Consequently, if any of these coefficients is negative, it is necessary to
+%check that the power spectrum is still non-negative.  Note that a non-negative
+%power-spectrum does not require the auto-correlation
+function to be positive for all $\tau$.  %The positive power spectrum is
+%related to the positive eigenvalues of the covariance matrix which are
+%necessary for a positive-definite matrix in limiting cases
+%\citep{Messerschmitt:2006}.  %We find empirically that requiring an everywhere
+%positive power spectrum results in positive eigenvalues for the covariance
+%matrix, and so we describe here how to ensure a positive power spectrum using
+%Sturm's theorem.
 %Relation of positive power spectrum to positive definite covariance matrix.

### 2017-03-20 paper/ms.tex
terminology related to #71

@@
 This means that the power spectrum of a positive definite kernel must be
-positive for all frequencies.  This seems true intuitively:  since the
-power spectrum is the square of the Fourier transform of the data time
-series, it must be non-negative.
+positive for all frequencies.
+This result is also intuitive because, since the power spectrum of a process
+is defined as expected the squared amplitude of the Fourier transform of the
+time series, it must be non-negative.
 

### 2017-03-20 paper/ms.tex
Some intro to section 5; minor edits

@@
 The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
-the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
+the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
 with a small number of operations for each data point.
@@
     r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},2}
+    0 & \cdots & 0 & r_{\bvec{\theta},N}
 \end{array}\right) \quad.
@@
 
-The extended system defined in the previous section is sparse with fewer than
+The extended system defined in the previous section is sparse with typically fewer than
 a few percent non-zero entries and band structure.
@@
 
-\dfmtodo{Add more intro to this section.}
-
-A special case of the \celerite\ model of great physical interest is a
+We now turn to the application of the \celerite\ algorithm to astrophysical
+variability.  Many astrophysical objects have a characteristic timescale
+of variability which is determined by its physical structure. The variability
+is excited by noisy physical processes, but also damped due to dissipation 
+in the system.  As a concrete example, stellar oscillations are strong at 
+resonant frequencies determined by the internal stellar structure,
+which are excited and damped by convective turbulence.
+
+Thus, a special case of the \celerite\ model of great physical interest is a
 stochastically-driven damped simple harmonic oscillator.

### 2017-03-20 paper/ms.tex
minor edits section 5

@@
 We now turn to the application of the \celerite\ algorithm to astrophysical
-variability.  Many astrophysical objects have a characteristic timescale
-of variability which is determined by its physical structure. The variability
-is excited by noisy physical processes, but also damped due to dissipation 
+variability.  Many astronomical objects have a characteristic timescale
+which is determined by the physical structure.  Variability
+is excited by noisy physical processes, which grows most strongly at
+the characteristic timescale, but is also damped due to dissipation 
 in the system.  As a concrete example, stellar oscillations are strong at 
 resonant frequencies determined by the internal stellar structure,
-which are excited and damped by convective turbulence.
+which are both excited and damped by convective turbulence.
 
@@
     For comparison, the dashed line shows the Lorentzian function from
-    \eq{lorentz-psd} with $c_j = \omega_0/2\,Q = 1/20$ and normalized so that
+    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
     $S(d_j)/S(0) = 100$.

### 2017-03-20 paper/ms.tex
Added TESS, SOAP references

@@
 inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Campante:2016}, we will infer
+Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we will infer
 $\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
@@
 quasiperiodic variations in photometric or spectroscopic observations
-(\dfmtodo{CITE}).
+\citep{Dumusque:2014}.
 It has been demonstrated that for light curves with nearly uniform sampling,

### 2017-03-20 paper/ms.tex
Added Aigrain 2015 ref

@@
 the empirical autocorrelation function provides a reliable estimate of the
-rotation period of a star (\dfmtodo{CITE}) and that a GP model with a
+rotation period of a star \citep{Aigrain:2015} and that a GP model with a
 quasiperiodic covariance function can be used to make probabilistic

### 2017-03-21 paper/ms.tex
Added transit refs; described R&P; updated acks

@@
 example, the mean function $\mu_\bvec{\theta}(t)$ is a limb-darkened transit
-light curve \dfmtodo{CITE MA} parameterized by a period $P$, a transit
+light curve \citep{Mandel:2002} parameterized by a period $P$, a transit
 duration $T$, a phase or epoch $t_0$, an impact parameter $b$, the radius
 of the planet in units of the stellar radius $R_P/R_\star$, and several
-parameters describing the limb-darkening profile of the star \dfmtodo{CITE}.
+parameters describing the limb-darkening profile of the star \citep{Claret:2011}.
 As in the previous example, we model the stellar variability using a GP model
@@
 
-\dfmtodo{Add discussion of original R\&P method here.}
+The \celerite\ model is based upon a method developed by Press \& Rybicki twenty 
+years ago \citep{Rybicki:1995}.  They observed that a correlation matrix consisting 
+of a single exponential kernel has an inverse that is tri-diagonal;
+the coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N)$.
+This method does not require an extended matrix, and thus speeds up the solution 
+by a factor of $\approx 2$ for large $N$, but 
+the requirement of an exponential kernel is extremely restrictive,
+equivalent to assuming an Ornstein--Uhlenbeck process, which tends to show sharp 
+variations in the light curve, and cannot describe quasi-periodicity.
 
@@
 Further progress was made by parameterizing the covariance matrix with simple,
-analytic functions that describe the autocorrelation function.  The parameters
-of these functions can then be optimized to best match the observed variability
-pattern of the residuals (after subtracting a transit model).  The disadvantage of
-this approach is that the functions used are chosen somewhat arbitrarily (e.g.\
-the exponential-squared function), and the Cholesky decomposition of the covariance
-matrix for computing the likelihood takes ${\cal O}(N^3)$ operations which prohibits
-application of this technique to large datasets.
+analytic functions that describe the autocorrelation function \citep{Gibson:2012}.  
+The parameters of these functions can then be optimized to best match the observed 
+variability pattern of the residuals (after subtracting a transit model).  The 
+disadvantage of this approach is that the functions used are chosen somewhat 
+arbitrarily (e.g.\ the exponential-squared function), and the Cholesky decomposition 
+of the covariance matrix for computing the likelihood takes ${\cal O}(N^3)$ operations 
+which prohibits application of this technique to large datasets.
 
@@
 
-EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
+This work was performed in part under contract with the Jet Propulsion Laboratory (JPL) 
+funded by NASA through the Sagan Fellowship Program executed by the NASA Exoplanet Science Institute.
+
+EA acknowledges support from NASA grants NNX13AF20G, NNX13A124G, NNX13AF62G, 
+from National Science Foundation (NSF) grant AST-1615315, and from
 NASA Astrobiology Institute's Virtual Planetary Laboratory, supported

### 2017-03-21 paper/ms.tex
Editing summary section

@@
 function in the time domain scale poorly to state-of-the-art datasets
-\citep{Brewer:2009}.
+\citep{Brewer:2009,Corsaro:2014}.
 
@@
 
-Although we have in mind applications of this fast method to model stellar
-variability, the method is general for one-dimensional GP problems, and may be
-applied to other problems.
-Accreting black holes show time series which may be modeled by correlated
-noise \citep{Kelly:2014};  indeed, this was the motivation for the original
-technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
-This approach may be broadly used for characterizing quasar variability
-\citep{MacLeod:2010}, measuring time lags with reverberation mapping
-\citep{Zu:2011,Pancoast:2014}, and modeling time delays in multiply-imaged
-gravitationally-lensed systems \citep{Press:1998}.
-
-Outside of astronomy, this technique may have application to seismology
-\citep{Robinson:1967},
-
-There are three primary applications we envision for the \celerite\ formalism:
-1).\ modeling the variability of an astrophysical system to infer it
-properties; 2).\ accounting for astrophysical variability as a source of noise
-when trying to detect additional phenomena; 3).\ interpolating or
-extrapolating variability to future or missing times.
-Some examples of the first are determining the characteristic variability
-timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
-variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
-granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
-quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
-classification of variable objects \citep{Zinn:2016}.
-Examples of the second are detecting and characterizing transiting exoplanets
-\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
-Grunblatt:2016, Luger:2016}
-and correcting for stellar activity in radial velocity measurements
-\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
-Examples of the third are measuring time delays in multiply-imaged
-gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
-AGN.
+The computational burden of Gaussian process regression has limited its application
+to small astronomical time series.  With the linear scaling of \celerite\, we
+envision the application of Gaussian processes may be expanded to much larger 
+datasets.  Despite the restrictive form of the \celerite\ kernel, with a 
+sufficient number of components it is flexible enough to describe a wide 
+range of astrophysical variability.  In fact, the relation of the \celerite\ kernel 
+to the damped, driven harmonic oscillator matches simple models of astrophysical 
+variability, and makes the parameterization interpretable in terms of resonant 
+frequency, amplitude, and quality factor.  Perhaps the most significant drawback 
+is the $J^2$ scaling of the solver;  however, for many problems a small value of $J$ 
+suffices.
+
+We envision two general varieties of applications:  those interested in the
+the \celerite\ kernel parameters (so-called ``hyper-parameters") which describe the
+astrophysical variability, versus those interested in characterizing the mean 
+model, treating the correlated noise as a nuisance.  Some users may
+be interested in both applications, while other problems may require interpolation
+or extrapolation beyond the dataset that is being modeled.
+
+%Some examples of the first are determining the characteristic variability
+%timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
+%variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
+%granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
+%quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
+%classification of variable objects \citep{Zinn:2016}.
+%Examples of the second are detecting and characterizing transiting exoplanets
+%\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
+%Grunblatt:2016, Luger:2016}
+%and correcting for stellar activity in radial velocity measurements
+%\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
+%Examples of the third are measuring time delays in multiply-imaged
+%gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
+%AGN.
 
@@
 recently begun to adopt full covariance matrices in analyzing the noise in
-transiting planet lightcurves.
+stellar lightcurves when detecting or characterizing transiting planets
+\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016, 
+Grunblatt:2016, Luger:2016}.
 One promising early attempt at accounting for correlated noise was the wavelet
@@
 overpredicts the power spectrum at low frequency compared to the Solar
-granulation power spectrum.
-It also cannot describe quasi-periodic noise.
+granulation power spectrum.  It also cannot describe quasi-periodic noise.
 
@@
 
+Although we have in mind applications of this fast method to model stellar
+variability, the method is general for one-dimensional GP problems, and may be
+applied to other problems.
+Accreting black holes show time series which may be modeled by correlated
+noise \citep{Kelly:2014};  indeed, this was the motivation for the original
+technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
+This approach may be broadly used for characterizing quasar variability
+\citep{MacLeod:2010}, measuring time lags with reverberation mapping
+\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
+gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic 
+variability in a high-energy source \citep{McAllister:2016}, or
+classification of variable objects \citep{Zinn:2016}.
+Outside of astronomy, this technique may have application to terrestrial 
+seismology \citep{Robinson:1967}.
+The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
+non-parametric data modeling \citep{Wilson:2013}.
+
 In the future we hope to expand the \celerite\ technique to multi-dimensional
@@
 or in detection of gravitational waves in the presence of systematic errors
-\citep{Moore:2016}.
+\citep{Moore:2016}.  Other extensions that may be useful are fast simulation
+of \celerite\ processes, prediction and/or interpolation of datasets, and
+extended the \celerite\ kernel to time-varying kernels.
 

### 2017-03-21 paper/ms.tex
more words aboutpostivie definiteness

--- 2017-03-20 paper/ms.tex
@@
 The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
-the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
+the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
 with a small number of operations for each data point.
@@
     r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},N}
+    0 & \cdots & 0 & r_{\bvec{\theta},2}
 \end{array}\right) \quad.
@@
 
-The extended system defined in the previous section is sparse with typically fewer than
+The extended system defined in the previous section is sparse with fewer than
 a few percent non-zero entries and band structure.
@@
 
-We now turn to the application of the \celerite\ algorithm to astrophysical
-variability.  Many astronomical objects have a characteristic timescale
-which is determined by the physical structure.  Variability
-is excited by noisy physical processes, which grows most strongly at
-the characteristic timescale, but is also damped due to dissipation 
-in the system.  As a concrete example, stellar oscillations are strong at 
-resonant frequencies determined by the internal stellar structure,
-which are both excited and damped by convective turbulence.
-
-Thus, a special case of the \celerite\ model of great physical interest is a
+\dfmtodo{Add more intro to this section.}
+
+A special case of the \celerite\ model of great physical interest is a
 stochastically-driven damped simple harmonic oscillator.
@@
     For comparison, the dashed line shows the Lorentzian function from
-    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
+    \eq{lorentz-psd} with $c_j = \omega_0/2\,Q = 1/20$ and normalized so that
     $S(d_j)/S(0) = 100$.
@@
 inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we will infer
+Looking forward to \tess\ \citep{Campante:2016}, we will infer
 $\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
@@
 quasiperiodic variations in photometric or spectroscopic observations
-\citep{Dumusque:2014}.
+(\dfmtodo{CITE}).
 It has been demonstrated that for light curves with nearly uniform sampling,
 the empirical autocorrelation function provides a reliable estimate of the
-rotation period of a star \citep{Aigrain:2015} and that a GP model with a
+rotation period of a star (\dfmtodo{CITE}) and that a GP model with a
 quasiperiodic covariance function can be used to make probabilistic
@@
 $\omega$.
+This requirement can also be written as
+\begin{eqnarray}
+a_j\,c_j > -b_j\,d_j \\
+a_j\,c_j > b_j\,d_j \quad.
+\end{eqnarray}
+Furthermore, we can see that $a_j$ must be positive since $k(0) = a_j$ should
+should be positive and, similarly, by requiring the covariance to be finite at
+infinite lag, we obtain the constraint $c_j \ge 0$.
+Without loss of generality, we can choose $d_j \ge 0$ and find the final
+constraint:
+\begin{eqnarray}
+    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j|
+\end{eqnarray}
 
@@
 %power-spectrum does not require the auto-correlation
-function to be positive for all $\tau$.  %The positive power spectrum is
+%function to be positive for all $\tau$.  %The positive power spectrum is
 %related to the positive eigenvalues of the covariance matrix which are
@@
 \begin{equation}
-S(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
+S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
 \end{equation}
@@
 \begin{eqnarray}
-q_j &=& a_jc_j-b_jd_j\\
-r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
+q_j &=& a_j\,c_j-b_j\,d_j\\
+r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
 s_j &=& 2(c_j^2-d_j^2)\\

### 2017-03-21 paper/ms.tex
adding HODLR benchmark plots

@@
 $\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
-$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N$
+$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
 \citep{Wilson:2014}.
@@
 A similar example using CARMA would necessarily be qualitative.
+
+Another GP method that has been used extensively in astronomy is the
+hierarchical off-diagonal low rank \citep[HODLR,][]{Ambikasaran:2016} solver.
+This method exploits the fact that many commonly used kernel functions produce
+``smooth'' matrices to approximately compute the GP likelihood with the
+scaling $\mathcal{O}(N\,\log^2 N)$.
+This method has the advantage that, unlike \celerite, it can be used with any
+kernel function but, in practice, the cost can still prove to be prohibitively
+high for multi-dimensional inputs.
+
+Many other approximate methods for scaling GP inference exist and we will make
+no attempt to discuss the comparisons exhaustively.
 
@@
 infinite lag, we obtain the constraint $c_j \ge 0$.
-Without loss of generality, we can choose $d_j \ge 0$ and find the final
-constraint:
-\begin{eqnarray}
-    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j|
+Combining these results, we find the constraint
+\begin{eqnarray}
+    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j| \quad.
 \end{eqnarray}

### 2017-03-21 examples/benchmark/plot.py

@@ -14,21 +14,15 @@ setup(auto=True)
 
 parser = argparse.ArgumentParser()
 parser.add_argument("platform")
-parser.add_argument("--default", action="store_true")
-parser.add_argument("--lapack", action="store_true")
-parser.add_argument("--sparse", action="store_true")
+parser.add_argument("--suffix", default=None)
 parser.add_argument("--directory",
                     default=os.path.dirname(os.path.abspath(__file__)))
 args = parser.parse_args()
 
 # Compile into a matrix
 suffix = ""
-if args.default:
-    suffix = "_default"
-elif args.lapack:
-    suffix = "_lapack"
-elif args.sparse:
-    suffix = "_sparse"
+if args.suffix is not None:
+    suffix = "_" + args.suffix
 
 fn = "benchmark_{0}{1}.csv".format(args.platform, suffix)
 fn = os.path.join(args.directory, fn)

### 2017-03-21 examples/benchmark/run.py

@@ -18,6 +18,7 @@ parser = argparse.ArgumentParser()
 parser.add_argument("--lapack", action="store_true")
 parser.add_argument("--default", action="store_true")
 parser.add_argument("--sparse", action="store_true")
+parser.add_argument("--george", action="store_true")
 parser.add_argument("--minnpow", type=int, default=6)
 parser.add_argument("--maxnpow", type=int, default=19)
 parser.add_argument("--minjpow", type=int, default=0)
@@ -25,6 +26,15 @@ parser.add_argument("--maxjpow", type=int, default=8)
 parser.add_argument("--outdir",
                     default=os.path.dirname(os.path.abspath(__file__)))
 args = parser.parse_args()
+
+if args.george:
+    try:
+        import george
+        from george.kernels import CeleriteKernel
+    except ImportError:
+        print("To run the george benchmark, you must install the dev version "
+              "of george with the 'CeleriteKernel' included")
+        raise
 
 # BLAS info
 blas_opt_info = npconf.get_info("blas_opt_info")
@@ -34,8 +44,8 @@ N = 2**np.arange(args.minnpow, args.maxnpow + 1)
 J = 2**np.arange(args.minjpow, args.maxjpow + 1)
 
 header = ""
-for k in ["lapack", "sparse", "default", "minnpow", "maxnpow", "minjpow",
-          "maxjpow"]:
+for k in ["lapack", "sparse", "default", "george",
+          "minnpow", "maxnpow", "minjpow", "maxjpow"]:
     header += "# {0}: {1}\n".format(k, getattr(args, k))
 header += "# platform: {0}\n".format(sys.platform)
 header += "# with_lapack: {0}\n".format(celerite.__with_lapack__)
@@ -51,6 +61,8 @@ elif args.sparse:
     fn += "_sparse"
 elif args.lapack:
     fn += "_lapack"
+elif args.george:
+    fn += "_george"
 fn += ".csv"
 fn = os.path.join(args.outdir, fn)
 print("filename: {0}".format(fn))
@@ -70,24 +82,42 @@ for xi, j in enumerate(J):
         kernel += terms.RealTerm(1.0, 0.1)
     for k in range((2*j - 1) // 2):
         kernel += terms.ComplexTerm(0.1, 2.0, 1.6)
-    c = kernel.coefficients
-    assert 2*j == len(c[0]) + 2*len(c[2]), "Wrong number of terms"
+    coeffs = kernel.coefficients
[…]

### 2017-03-21 paper/figures/rotation.py

@@ -5,7 +5,6 @@ from __future__ import division, print_function
 
 import kplr
 import emcee3
-import fitsio
 import numpy as np
 import matplotlib.pyplot as plt
 from scipy.optimize import minimize
@@ -72,16 +71,6 @@ kernel = RotationTerm(
         [-5.0, 5.0],
     ]
 )
-# kernel += terms.SHOTerm(log_S0=np.log(np.var(y)),
-#                         log_omega0=np.log(2*np.pi/10.0),
-#                         log_Q=-0.5*np.log(2),
-#                         bounds=dict(log_omega0=(np.log(2*np.pi)-np.log(t.max() -
-#                                                                   t.min()),
-#                                                 np.log(2*np.pi)-np.log(10.0)),
-#                                     log_S0=np.log(np.var(y) * np.array([0.01, 1])),
-#                                     ))
-# kernel.terms[1].freeze_parameter("log_Q")
-# kernel.terms[1].freeze_parameter("log_a")
 gp = celerite.GP(kernel, mean=np.median(y))
 gp.compute(t, yerr)
 

### 2017-03-21 paper/ms.tex
started adding discussion of power spectrum estimation

--- 2017-03-21 paper/ms.tex
@@
 Fast and scalable Gaussian process modeling with applications
-to astronomical time series
+to astronomical time series and power spectrum estimation
 \vspace{-3\baselineskip}
@@
 rotation periods, asteroseismic oscillation spectra, and transiting planet
-parameters.
+parameters.  The power spectrum of an irregularly sampled time series may be 
+estimated with Bayesian inference, with a posterior estimate of the associated 
+uncertainties, in principle avoiding the problems associated with window functions 
+and aliasing.
 The method is flexible, fast, and most importantly, interpretable.
@@
 non-parametric data modeling \citep{Wilson:2013}.
+
+The \celerite\ formalism also allows a principled approach to the estimation
+of the power spectrum of a time series, as well as its uncertainties.
+Fourier transforms work for uniformly sampled data;  astronomical data are
+rarely (if ever) uniformly sampled.  Periodograms are affected by the window
+function and aliasing in datasets, and make estimating the significance of
+spectral features ambiguous.  The Bayesian approach to the estimation of a power
+spectrum (CITE) avoids these difficulties:  it allows the computation of
+uncertainties, it accounts for the window function on the uncertainties of
+the power spectrum, and in principle it can treat
 
@@
 $\omega$.
-This requirement can also be written as
-\begin{eqnarray}
-a_j\,c_j > -b_j\,d_j \\
-a_j\,c_j > b_j\,d_j \quad.
-\end{eqnarray}
-Furthermore, we can see that $a_j$ must be positive since $k(0) = a_j$ should
-should be positive and, similarly, by requiring the covariance to be finite at
-infinite lag, we obtain the constraint $c_j \ge 0$.
-Without loss of generality, we can choose $d_j \ge 0$ and find the final
-constraint:
-\begin{eqnarray}
-    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j|
-\end{eqnarray}
 
@@
 %power-spectrum does not require the auto-correlation
-%function to be positive for all $\tau$.  %The positive power spectrum is
+function to be positive for all $\tau$.  %The positive power spectrum is
 %related to the positive eigenvalues of the covariance matrix which are
@@
 \begin{equation}
-S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
+S(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
 \end{equation}
@@
 \begin{eqnarray}
-q_j &=& a_j\,c_j-b_j\,d_j\\
-r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
+q_j &=& a_jc_j-b_jd_j\\
+r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
 s_j &=& 2(c_j^2-d_j^2)\\

### 2017-03-21 paper/ms.tex
HODLR comparison in paper

--- 2017-03-21 paper/ms.tex
@@
 high for multi-dimensional inputs.
-
-Many other approximate methods for scaling GP inference exist and we will make
-no attempt to discuss the comparisons exhaustively.
-
-Approximate:
-- inducing point methods \citep{Quinonero-Candela:2007}
-- KISS-GP \citep{Wilson:2015, Wilson:2015a}
-- HODLR \citep{Ambikasaran:2016}
+The proportionality constant in the $N\,\log^2N$ scaling of the HODLR method
+is a function of the specific kernel and we find~--~using the \project{george}
+software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
+scales approximately linearly with $J$ but with substantial overhead for small
+models when compared to \celerite.
+This means that the HODLR solver can \emph{approximately} evaluate \celerite\
+models more efficiently than \celerite\ for large models $J \gg 10$ with a
+small dataset where $N$ is smaller than a few thousand.
+
+Many other approximate methods for scaling GP inference exist \citep[see, for
+example,][and references therein]{Wilson:2015a} and we will make no attempt to
+discuss the comparisons exhaustively.
 

### 2017-03-21 paper/ms.tex
cleaning up
reverting title [ci skip]
some more discussion of astero example

@@
 Fast and scalable Gaussian process modeling with applications
-to astronomical time series and power spectrum estimation
+to astronomical time series
 \vspace{-3\baselineskip}
@@
 It is also possible to cast some two-dimensional problems in the \celerite\
-framework (\dfmtodo{CITE Ian C}).
+framework \citep[see][for a potential application]{Czekala:2017}.
 
@@
 The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
-Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
-many of the upcoming large scale photometric surveys \citep{Campante:2016,
-Rauer:2014, Gould:2015}.
+curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
+Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
+driver for many of the upcoming large scale photometric surveys
+\citep{Campante:2016, Rauer:2014, Gould:2015}.
 Most asteroseismic analyses have been limited to relatively high
@@
 of the empirical periodogram of the data cannot be used to formally propagate
-the measurement uncertainties to the constraints on physical parameters
-(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
-function in the time domain scale poorly to state-of-the-art datasets
-\citep{Brewer:2009,Corsaro:2014}.
+the measurement uncertainties to the constraints on physical
+parameters~--~instead relying on bootstrapped uncertainty estimators
+\citep{Huber:2009}~--~but more sophisticated methods that compute the
+likelihood function in the time domain scale poorly to state-of-the-art
+datasets \citep{Brewer:2009,Corsaro:2014}.
 
@@
 To demonstrate this method, we will use a very simple heuristic model based on
-\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
+where the model PSD is given by a mixture of 8 components with
 amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
@@
 and $\Delta\nu$ compared to the results from the literature.
-\dfmtodo{Add some more discussion here.}
+
+This model requires about a minute of computation time to find the maximum
+likelihood parameters and then it takes about an hour to run the MCMC sampling
+to convergence on a dual-CPU laptop.
+This is more computationally intensive than traditional methods of measuring
+asteroseismic oscillations but is is also orders of magnitude cheaper than the
+same analysis using another GP solver.
+An in depth discussion of the benefits of rigorous probabilistic inference of
+asteroseismic parameters in the time domain is beyond the scope of this paper
+but we hope to revisit this opportunity in the future.
 
@@
     from the inference shown in \Figure{astero} compared to the published
-    value (error bar) based on the full \kepler\ dataset \dfmtodo{CITE}.
+    value (error bar) based on several years of \kepler\ observations
+    \citep{Pinsonneault:2014}.
     \figurelabel{astero-corner}}
@@
 \includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
-\caption{A comparison between the empirical PSD and the posterior inference of
-the PSD as a mixture of stochastically-driven simple harmonic oscillators.
+\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
+posterior inference of the PSD as a mixture of stochastically-driven simple
+harmonic oscillators.
 (top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
     on the full four year baseline of the mission.
-    The light gray curve shows the raw periodogram and the black curve has
-    been smoothed with a Gaussian filter with \dfmtodo{SOME WIDTH}.
 (middle) The same periodogram computed using about a month of data.
@@
 
-% DFM: I'm going to redistribute the following in other sections.
-
-% Stars are variable, for better or worse:  their variability provides information about the properties
-% of stars, but also results in noise that must be accounted for in detecting and characterizing their
-% planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
-% goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
-% properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
-% flux variability and spectral variations.
-
-% A promising technique for addressing both of these goals is the Gaussian Process formalism
-% \citep{Rasmussen2006,Gibson:2012}.  The basis of this method is to assume that stellar variability
-% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
-% correlations is completely specified by a correlation matrix.  In its simplest form, the
-% correlation matrix is modeled with a time-independent autocorrelation function (although
-% the time-independence can be relaxed), including a white-noise component that may vary
-% with each exposure, and the autocorrelation function is described by a functional form that may be
-% simply parameterized by a relatively small number of parameters, sometimes referred to as
-% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
-% describe the autocorrelation function, thus inferring properties of the stellar variabilty
-% which may be used to characterize properties of the star, such as the rotation period,
-% the asteroseismic frequencies, or granulation noise known as `flicker'
-% \citep{Aerts2010,Noyes:1984,Bastien:2013}.
-% The goals of planet detection include measuring the Doppler shift of the star, measuring
-% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
-% look for dynamical interactions between planets.  To properly account for the impact of
-% stellar variability on the planet parameters requires treating correlations in the data, and
-% to do so also requires evaluating the correlation matrix.
-
-% The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
-% $N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
-% prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
-% its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
-% = N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
-% using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
-% {\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
-% $N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
-% cases.
-
-% One promising solution is to use approximate techiques based upon the fact that correlation
-% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
-% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
-% solver is only approximate \citep{Ambikasaran:2013,Ambikasaran:2016}.  In addition, the
-% solver is complicated and can be challenging to use with general choices of kernel.  Another
-% promising approximate technique is to use a uniform resampling of the dataset, and then
-% rely on the structure of Toeplitz matrices, which require uniformly space time series,
-% to obtain approximate solutions of the likelihood function \citep{Wilson:2015},
-% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
-% this approach has the drawback of being approximate.  A third approach is to use a kernel
-% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
-% model \citep[aka \texttt{CARMA};][]{Kelly:2014}.  This approach can model a stationary power-spectrum that
-% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
-% of the Lorentzians are required to have a specific relation.  In addition, these
-% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}
-
-% The approach we explore in this paper is based upon a method developed by Press \&
-% Rybicki twenty years ago \citep{Rybicki:1995}.  They observed that a correlation
-% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
-% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
-% They also indicated that a combination of two exponential kernels would have a simliar scaling,
-% but they did not demonstrate how to extend this to more than two kernel components.
-
-% A significant advance was made in generalizing the Press-Rybicki method to an
-% arbitrary number of exponential kernels by utilizing the fact that the elements of
-% such matrices can be written in terms of the products of components of two vectors,
-% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
-% Rybicki-Press' (\texttt{celerite}) approach
-% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
-% the coefficients of the kernels need not be stationary, the computation of the
-% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
-% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
-% the coefficients (except that the kernel needs to be positive definite).  Unlike
-% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{celerite} method is exact.  The main drawback
-% of this approach is that the sum of exponential kernels with real coefficients can
-% only represent a limited range of power spectra of auto-correlated time series:
-% those which can be expressed the sum of Lorentzians that peak at zero frequency.
-
-% However, this drawback turns out to not be significant:  the exponential kernel
-% can also have a complex exponential coefficients (plus their complex-conjugates), and
-% thus approximate an almost arbitrary autocorrelation function with enough terms, most
-% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
-% components is a {\it very} good approximation of stellar variability spectra:  the
-% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
-% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
-% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
-% parameterized this way as well.  In this paper
-% we extend the \texttt{celerite} approach to kernels with complex exponentials, which when
-% combined with the complex conjugate leads to damped harmonic auto-correlation functions
-% which are flexible enough to describe a wide range of variability, and may still
-% be solved in $O(N_{time}$ operations.
-
-% We show that the extended matrix that embeds this multi-component kernel can be written
-% in real notation, which speeds up the evaluation of the likelihood function,
-% and allows optimization software to use automatic differential in computing
-% derivatives of the likelihood function.
-
 \section{Summary}
@@
 the power spectrum, and in principle it can treat
+\agoltodo{Finish this\ldots}
 

### 2017-03-21 paper/ms.tex
removing some todos
finished paragraph on power spectrum

@@
     This is consistent with the published rotation period made using the
-    autocorrelation function and the full \kepler\ data (\dfmtodo{CITE}).
+    autocorrelation function and the full \kepler\ baseline
+    \citep{Mathur:2014}.
     \figurelabel{rotation-period}}
@@
 Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
-as `flicker', $1/r$, or pink noise;  however, this particular type of noise
+as `flicker', $1/f$, or pink noise;  however, this particular type of noise
 overpredicts the power spectrum at low frequency compared to the Solar
@@
 uncertainties, it accounts for the window function on the uncertainties of
-the power spectrum, and in principle it can treat
-\agoltodo{Finish this\ldots}
+the power spectrum, and in principle it can treat aliases if the full posterior
+is explored.  One drawback is that the number of $J$ terms to include is
+unknown, but may be inferred with computation of the evidence, or an
+approximation such as the Bayesian information criterion \citep{Kelly:2014}.
+In general, the parameters of the \celerite\ terms are more interesting due
+to their physical meaning rather than the overall shape of the power spectrum,
+as shown in the foregoing examples.
 
@@
 $(0,\,\infty]$.
-We first construct $Q_0(z)$ and it's derivative $Q_1(z) = {Q_0}^\prime(z)$,
+We first construct $Q_0(z)$ and its derivative $Q_1(z) = {Q_0}^\prime(z)$,
 and then loop from $k=2$ to $k=2\,(J-1)+1$, computing

### 2017-03-22 paper/ms.tex
final edits, added more refs

@@
     e^{-\tilde{c}\,\tau}\,[
-        a_+\,\cos(d_+\,\tau) + b_+\,\sin(d_+\,\tau) +
-        a_-\,\cos(d_-\,\tau) + b_-\,\sin(d_-\,\tau)
+        \tilde{a}_+\,\cos(\tilde{d}_+\,\tau) + \tilde{b}_+\,\sin(\tilde{d}_+\,\tau) +
+        \tilde{a}_-\,\cos(\tilde{d}_-\,\tau) + \tilde{b}_-\,\sin(\tilde{d}_-\,\tau)
     ]
@@
 allowed models.
-This is a interesting example because, when analyzing real data, we rarely
+This is an interesting example because, when analyzing real data, we rarely
 have any fundamental reason to believe that the data were generated by a GP
@@
 same analysis using another GP solver.
-An in depth discussion of the benefits of rigorous probabilistic inference of
+An in-depth discussion of the benefits of rigorous probabilistic inference of
 asteroseismic parameters in the time domain is beyond the scope of this paper
@@
 as `flicker', $1/f$, or pink noise;  however, this particular type of noise
-overpredicts the power spectrum at low frequency compared to the Solar
+overpredicts the power spectrum at low and high frequency compared to the Solar
 granulation power spectrum.  It also cannot describe quasi-periodic noise.
@@
 correct inference of planetary transit parameters, and so we expect that \celerite\
-will be important for transit timing, transit spectroscopy, Doppler beaming,
-phase functions, and more.
+will be important for transit detection \citep{ForemanMackey:2015}, transit timing 
+\citep{Agol:2005,Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler 
+beaming \citep{Loeb:2003,Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase 
+functions \citep{Knutson:2007,Zucker:2007}, and more.
 
@@
 seismology \citep{Robinson:1967}.
-The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
-non-parametric data modeling \citep{Wilson:2013}.
+%The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
+%non-parametric data modeling \citep{Wilson:2013}.
 
@@
 spectral features ambiguous.  The Bayesian approach to the estimation of a power
-spectrum (CITE) avoids these difficulties:  it allows the computation of
+spectrum \citep{Wilson:2013} avoids these difficulties:  it allows for the computation of
 uncertainties, it accounts for the window function on the uncertainties of

### 2017-03-22 paper/ms.tex
adding vdp to acks

--- 2017-03-21 paper/ms.tex
@@
 \keywords{%
-% methods: data analysis
-% ---
-% methods: statistical
-% ---
-% catalogs
-% ---
-% planetary systems
-% ---
-% stars: statistics
+ methods: data analysis
+ ---
+ methods: data analysis
+ ---
+ methods: statistical
+ ---
+ asteroseismology
+ ---
+ stars: rotation
+ ---
+ planetary systems
 }
@@
 Sam Grunblatt,
-David W.\ Hogg, and
-Dan Huber
+David W.\ Hogg,
+Dan Huber, and
+Jake VanderPlas
 for helpful contributions to the ideas and code presented here.

### 2017-03-22 paper/ms.tex
OMG GET RID OF ALL THE 'will'S

@@
 The growing field of large-scale time domain astronomy requires methods for
-probabilistic inference that scale well with amount of data.
+probabilistic inference that are computationally tractable even with large
+datasets.
 We present a method for Gaussian Process regression in one dimension with a
@@
 implementation, and a comparison to existing scalable Gaussian Process methods.
-We demonstrate the application of this method by applying it to simulated and
-real astronomical time series datasets.
+We demonstrate the method by applying it to simulated and real astronomical
+time series datasets.
 These demonstrations are examples of probabilistic inference of stellar
 rotation periods, asteroseismic oscillation spectra, and transiting planet
-parameters.  The power spectrum of an irregularly sampled time series may be
-estimated with Bayesian inference, with a posterior estimate of the associated
-uncertainties, in principle avoiding the problems associated with window functions
-and aliasing.
-The method is flexible, fast, and most importantly, interpretable.
-There are a wide range of potential applications within astronomical data
-analysis and beyond.
+parameters.
+In all of these cases, we infer posterior distributions over models for the
+power spectrum of the process that generated these irregularly sampled time
+series without computing a standard algebraic estimator.
+The method is flexible, fast, and most importantly, interpretable, with a wide
+range of potential applications within astronomical data analysis and beyond.
 We provide well tested and documented open-source implementations of this
@@
 \citep{Bond:1987,Bond:1999}, correlated instrumental noise
-\citep{Gibson:2012}, spectroscopic calibration \citep{Czekala:2017,Evans:2015}
-and residuals caused by model inconsistencies (CITE + better words).
+\citep{Gibson:2012}, and spectroscopic calibration
+\citep{Czekala:2017,Evans:2015}.
 While these models are widely applicable, their use has been limited, in
@@
 This method is especially appealing compared to other similar methods~--~we
-will return to these below~--~because it is exact, flexible, simple, and fast.
-
-In the following pages, we will motivate the general problem of GP regression,
+return to these below~--~because it is exact, flexible, simple, and fast.
+
+In the following pages, we motivate the general problem of GP regression,
 describe the previously published scalable method \citep{Rybicki:1995,
@@
 points $N$.
-This means that for large datasets, every evaluation of the likelihood will
-quickly become computationally intractable.
+This means that for large datasets, every evaluation of the likelihood
+quickly becomes computationally intractable.
 In this case, standard non-linear optimization or MCMC will no longer be
@@
 \end{eqnarray}
-In this case, the inverse will be dense but \eq{gp-likelihood} can still be
+In this case, the inverse is dense but \eq{gp-likelihood} can still be
 evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
@@
 The details of this method and a few implementation considerations are
-discussed in the following \sectionname\ but we will first discuss some
+discussed in the following \sectionname\ but we first discuss some
 properties of this covariance function.
@@
 \end{eqnarray}
-For clarity, we will refer to the argument within the sum as a ``\celerite\
+For clarity, we refer to the argument within the sum as a ``\celerite\
 term'' for the remainder of this paper.
@@
 The physical interpretation of this model isn't immediately obvious and we
-will return to a more general discussion of the physical intuition in a moment
-but we can start with a discussion of some useful special cases of this model.
+return to a more general discussion of the physical intuition in a moment
+but we start with a discussion of some useful special cases of this model.
 
@@
 
-To provide some insight for this method, we will follow
+To provide some insight for this method, we follow
 \citet{Ambikasaran:2015} and start by working through a simple example.
@@
 \citet{Rybicki:1995}.
-To demonstrate this method, we will write out the full system of
+To demonstrate this method, we write out the full system of
 equations that we must solve to apply the inverse of $K$ in order to compute
@@
 \end{eqnarray}
-where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
+where we follow \citet{Ambikasaran:2015} and call this the ``extended''
 system and rewrite \eq{impl-matrix} as
@@
 
-In the following section, we will discuss this method more generally but it's
+In the following section, we discuss this method more generally but it's
 worth noting a few important facts that can already be seen in this example.
@@
 points be monotonically sorted in time.
-Neither of these properties will be satisfied in general for multidimensional
-inputs and all of our following discussion will assume a sorted
+Neither of these properties be satisfied in general for multidimensional
+inputs and all of our following discussion assumes a sorted
 one-dimensional dataset.
@@
 First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with a bandwidth that scales linearly
+extended matrix is still be banded with a bandwidth that scales linearly
 with the number of components $J$.
@@
 model for asteroseismic oscillations.
-In practice, we will take a sum over oscillators with different values of $Q$,
+In practice, we take a sum over oscillators with different values of $Q$,
 $S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
@@
 
-To demonstrate the application of \celerite, we will start by inferring
+To demonstrate the application of \celerite, we start by inferring
 posterior constraints on the parameters of a GP model applied to several
 simulated datasets with known properties.
-In the following section, we will expand on these examples by applying
+In the following section, we expand on these examples by applying
 \celerite\ to real datasets.
@@
 
-In this first example, we will simulate a dataset using a known \celerite\
+In this first example, we simulate a dataset using a known \celerite\
 process and fit it with \celerite\ to demonstrate that valid inferences can be
@@
 This means that the true autocorrelation of the process can never be correctly
-represented by the model that we're using to fit but we will use this example
+represented by the model that we're using to fit but we use this example
 to demonstrate that, at least in this case, valid inferences can still be made
@@
 perfectly reproduce the structure of this process.
-That being said, we will demonstrate that rigorous inferences can be made
+That being said, we demonstrate that rigorous inferences can be made
 about $P_\mathrm{true}$ even with an effective model.
@@
 
-In this section we will demonstrate several use cases of \celerite\ when
+In this section we demonstrate several use cases of \celerite\ when
 applied to real datasets.
-Each of these examples touches on an active area of research so we will limit
+Each of these examples touches on an active area of research so we limit
 our examples to be qualitative in nature and make no claim of optimality but
-we hope that these examples will encourage interested readers to investigate
+we hope that these examples encourage interested readers to investigate
 the applicability of \celerite\ to their research.
@@
 
-To demonstrate this method, we will use a very simple heuristic model based on
+To demonstrate this method, we use a very simple heuristic model based on
 where the model PSD is given by a mixture of 8 components with
@@
 inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we will infer
+Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we infer
 $\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
@@
 There are many other methods of scaling GP models to large datasets and in
-this section we will explore some comparisons between \celerite\ and some
+this section we explore some comparisons between \celerite\ and some
 other popular methods and discuss some use cases where \celerite\ shines and
@@
 can be used with any stationary kernel but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it will, in general, be less
+and the scaling is worse than \celerite\ so it, in general, be less
 efficient when applied to large datasets.
@@
 Many other approximate methods for scaling GP inference exist \citep[see, for
-example,][and references therein]{Wilson:2015a} and we will make no attempt to
+example,][and references therein]{Wilson:2015a} and we make no attempt to
 discuss the comparisons exhaustively.
@@
 Jake VanderPlas
-for helpful contributions to the ideas and code presented here.
+for helpful discussions informing the ideas and code presented here.
 
@@
 
-%The general case with multiple \celerite\ terms can be checked using Sturm's
-%theorem. The power spectrum is computed from taking the square of the Fourier
-%transform of the data time series; hence, the power-spectrum must be non-negative.
-%In constructing a kernel, if any of the parameters $a_j$ or $b_j$ are negative,
-%then it is possible for the power spectrum to go negative, which violates the
-%definition of the power spectrum.
-%Consequently, if any of these coefficients is negative, it is necessary to
-%check that the power spectrum is still non-negative.  Note that a non-negative
-%power-spectrum does not require the auto-correlation
-%function to be positive for all $\tau$.  %The positive power spectrum is
-%related to the positive eigenvalues of the covariance matrix which are
-%necessary for a positive-definite matrix in limiting cases
-%\citep{Messerschmitt:2006}.  %We find empirically that requiring an everywhere
-%positive power spectrum results in positive eigenvalues for the covariance
-%matrix, and so we describe here how to ensure a positive power spectrum using
-%Sturm's theorem.
-%Relation of positive power spectrum to positive definite covariance matrix.
-%Power spectrum of a sum of kernels.
-
 In the case of $J$ \celerite\ terms, we can check for negative values of the

### 2017-03-22 paper/ms.tex
paper edits

@@
 practice, by the computational cost and scaling.
-In general, the cost of computing a GP likelihood scales as the third power of
+In general, the cost of computing a GP likelihood scales as the cube of
 the number of data points $\mathcal{O}(N^3)$ and in the current era of large
 time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
-$\sim10^{3-5}$ observations each~---~this cost is prohibitive.
+$\sim10^{3-5}$ observations each~---~this scaling is prohibitive.
 
@@
 $p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
-algorithm to sample from the posterior probability density.
+algorithm to sample the joint posterior probability density.
 
@@
 quickly becomes computationally intractable.
-In this case, standard non-linear optimization or MCMC will no longer be
-practical inference methods.
+In this case, the use of non-linear optimization or MCMC is no longer
+practical.
 
@@
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
-evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
-in the mixture and $N$ is still the number of data points.
-
-It turns out that this kernel function can be made even more general by
+evaluated in $\mathcal{O}(N\,J^2)$ where $J$ is the number of components
+in the mixture.
+
+This kernel function can be made even more general by
 introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
@@
 \end{eqnarray}
-and, for this function, \eq{gp-likelihood} can still be solved with
+and, for this function, \eq{gp-likelihood} can still be evaluated with
 $\mathcal{O}(N)$ operations.
@@
 return to a more general discussion of the physical intuition in a moment
-but we start with a discussion of some useful special cases of this model.
+but we start by investigating some special cases.
 
@@
 \citet{Ambikasaran:2015} and start by working through a simple example.
-In this case, we'll assume that we have three data points
+In this case, we assume that we have three data points
 $\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
@@
 the first term of \eq{gp-likelihood}.
-In matrix notation, this can be written as
+In matrix notation, this is
 \begin{eqnarray}\eqlabel{impl-matrix}
@@
 the extended matrix $K_\mathrm{ext}$.
-Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
+Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
 $\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
@@
 \item \texttt{lapack}: The general banded LU decomposition implementation from
-    LAPACK\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs} methods
-    from LAPACK.} \citep{Anderson:1999} using optimized BLAS
-    routines\footnote{The experiments below use the Intel Math Kernel Library
-    (\project{MKL}): \url{https://software.intel.com/en-us/intel-mkl}.}, and
+    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
+    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
+    \project{BLAS} routines\footnote{The experiments below use the Intel Math
+    Kernel Library (\project{MKL}):
+    \url{https://software.intel.com/en-us/intel-mkl}.}, and
 
@@
 
-We now turn to the application of the \celerite\ algorithm to astrophysical
-variability.  Many astronomical objects have a characteristic timescale
-which is determined by the physical structure.  Variability
-is excited by noisy physical processes, which grows most strongly at
-the characteristic timescale, but is also damped due to dissipation
-in the system.  As a concrete example, stellar oscillations are strong at
-resonant frequencies determined by the internal stellar structure,
-which are both excited and damped by convective turbulence.
-
-Thus, a special case of the \celerite\ model of great physical interest is a
+We now turn to a discussion of \celerite\ as a model of astrophysical
+variability.
+A common concern in the context of GP modeling in astrophysics is the lack of
+physical motivation for the choice of kernel functions.
+Kernels are often chosen simply because they are popular with little
+consideration of the impact of this decision.
+While this isn't necessarily a problem for the inferred results, in this
+section we discuss an exact physical interpretation of the \celerite\ kernel
+that will be applicable to many astrophysical systems but especially the
+time-variability of stars.
+
+Many astronomical objects have characteristic timescales determined by their
+physical structure.
+Variability is excited by noisy physical processes, and grows most strongly
+at the characteristic timescale, but is also damped due to dissipation in the
+system.
+As a concrete example, stellar (asteroseismic) oscillations are strong at
+resonant frequencies determined by the internal stellar structure, which are
+both excited and damped by convective turbulence.
+
+To relate this physical picture to \celerite, we consider the dynamics of a
 stochastically-driven damped simple harmonic oscillator.
@@
 force.
-If $\epsilon(t)$ is white noise, the PSD of this process is given by
-\citep[for example][]{Anderson:1990}
+If $\epsilon(t)$ is white noise, the PSD of this process is
+\citep{Anderson:1990}
 \begin{equation}\eqlabel{sho-psd}
@@
 
-These identities yield a kernel of the form
+For these definitions, the kernel is
 \begin{equation}\eqlabel{sho-kernel}
@@
 where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
-It is interesting to note that, because of the damping, the characteristic
-oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
-is not equal to the frequency of the undamped oscillator $\omega_0$.
+We note that, because of the damping, the characteristic oscillation frequency
+in this model $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
+the frequency of the undamped oscillator $\omega_0$.
 
@@
 \end{eqnarray}
-This, in turn, is the most commonly used model for the background granulation
-noise in asteoreseismic \citep{Kallinger:2014} and helioseismic
-\citep{Harvey:1985, Michel:2009} analyses.
-The Fourier transform of this PSD corresponds to the kernel
+This functional form is commonly used to model for the background granulation
+noise in asteoreseismic and helioseismic \citep{Harvey:1985, Michel:2009,
+Kallinger:2014}  analyses.
+The kernel for this value of $Q$ is
 \begin{equation}
 k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
-    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
+    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)} \quad.
 \end{equation}}
@@
 \citep{Rasmussen:2006}.
-This suggests that the Mat\'ern-3/2 covariance can be well approximated using
+This suggests that the Mat\'ern-3/2 covariance can be approximated using
 the \celerite\ framework with a small value of $f$ in \eq{approx-matern2} but we
-caution that this could also lead to numerical issues with the solver.
+caution that this might lead to numerical issues for the solver.
 }
@@
 \celerite\ to real datasets.
-In the first example, we demonstrate that \celerite\ can be used to measure
+In the first example, we demonstrate that \celerite\ can be used to infer
 the power spectrum of a process when the data are generated from a \celerite\
@@
 where $P_\mathrm{true}$ is the fundamental period of the process.
-This autocorrelation structure yields the power spectrum
+This autocorrelation structure corresponds to the power spectrum
 \begin{eqnarray}
@@
 We note that using \eq{product-rule}, the product of two \celerite\ terms can
-also be expressed as a \celerite\ term.
-We apply log-uniform priors on all the parameters and use \emcee\
-\citep{Foreman-Mackey:2013} to sample the posterior probability for all of the
-parameters.
+also be expressed using \celerite.
+We apply log-uniform priors on all the parameters and use \emcee\ to sample
+the joint posterior probability for all of the parameters.
 The inferred distribution for the parameter $P$ is shown in the right panel of
@@
 be applied to spectroscopy, where wavelength~--~instead of time~--~is the
-independent coordinate and other one-dimensional domains.
-It is also possible to cast some two-dimensional problems in the \celerite\
-framework \citep[see][for a potential application]{Czekala:2017}.
+independent coordinate and other one-dimensional domains \citep[see][for a
+potential application]{Czekala:2017}.
 
@@
 oscillations are computed using a physical model and evaluate the probability
-of the observed dataset using a GP where the PSD is a sum of terms given by
-\eq{sho-psd}.
+of the observed time series using a GP where the PSD is a sum of terms given
+by \eq{sho-psd}.
 This gives us a method of computing the likelihood function for the parameters
-of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$)
-\emph{conditioned on the observed time series} in $\mathcal{O}(N)$ operations.
-In other words, this provides us with a computationally efficient method for
-performing rigorous probabilistic inference of asteroseismic parameters in the
-time domain.
+of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
+other, more fundamental parameters) \emph{conditioned on the observed time
+series} in $\mathcal{O}(N)$ operations.
+In other words, \celerite\ provides a computationally efficient framework that
+can be combined with physically-motivated models of stars and numerical
+inference methods to make rigorous probabilistic measurements of asteroseismic
+parameters in the time domain.
 We expect that this has the potential to push asteroseismic analysis to lower
@@
 
-To demonstrate this method, we use a very simple heuristic model based on
-where the model PSD is given by a mixture of 8 components with
-amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
-several nuisance parameters.
+To demonstrate the method, we use a very simple heuristic model where the PSD
+is given by a mixture of 8 components with amplitudes and frequencies
+specified by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance
+parameters.
 The first term is used to capture the granulation ``background''
@@
 This model could be easily extended to include small frequency splitting and
-$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
-like $\log g$.
+$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
+physical parameters of the star.
 
@@
 inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we infer
+Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
 $\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
@@
 the \kepler\ mission.
-For KIC~11615890, the published asteroseismic parameters based on several
+For KIC~11615890, the published asteroseismic parameters measured using several
 years of \kepler\ observations are \citep{Pinsonneault:2014}
@@
 
-To demonstrate this model, we fit a \celerite\ model with a kernel given by
+To demonstrate this method, we fit a \celerite\ model with a kernel given by
 \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
@@
 \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
-We used \emcee\ to sample to posterior probability density for the four
+We used \emcee\ to sample the joint posterior probability density for the four
 parameters in \eq{rot-kernel} conditioned on two quarters of \kepler\ data and
@@
     credible region.
-(right) Inferred constraints on the model covariance function from
-    \eq{rot-kernel}.
+(right) Inferred constraints on the model covariance function.
     The dashed line shows the maximum likelihood model, the blue solid line
@@
     in \eq{rot-kernel}.
-    This is consistent with the published rotation period made using the
-    autocorrelation function and the full \kepler\ baseline
-    \citep{Mathur:2014}.
+    This is consistent with the published rotation period made using the full
+    \kepler\ baseline \citep{Mathur:2014}.
     \figurelabel{rotation-period}}
@@
 
-To demonstrate this example, we take a month-long segment of the \kepler\
-light curve for KIC~1430163~--~the target from the previous example~--~and
-multiply it by a simulated transit model with known parameters.
+We take a month-long segment of the \kepler\ light curve for
+KIC~1430163~--~the target from the previous example~--~and multiply it by a
+simulated transit model with known parameters.
 The top panel of \Figure{transit-ml} shows the data including the simulated
@@
 \caption{The marginalized posterior constraints on the physical parameters of
-    the planet transiting during the light curve shown in the top panel of
+    the planet transit in the light curve shown in the top panel of
     \Figure{transit-ml}.
@@
 There are many other methods of scaling GP models to large datasets and in
-this section we explore some comparisons between \celerite\ and some
-other popular methods and discuss some use cases where \celerite\ shines and
-others where it does not.
+this section we draw comparisons between \celerite\ and other popular methods.
 Scalable GP methods tend to fall into two categories: approximate and
@@
 
-The \celerite\ model is based upon a method developed by Press \& Rybicki twenty
-years ago \citep{Rybicki:1995}.  They observed that a correlation matrix consisting
-of a single exponential kernel has an inverse that is tri-diagonal;
-the coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N)$.
-This method does not require an extended matrix, and thus speeds up the solution
-by a factor of $\approx 2$ for large $N$, but
-the requirement of an exponential kernel is extremely restrictive,
-equivalent to assuming an Ornstein--Uhlenbeck process, which tends to show sharp
-variations in the light curve, and cannot describe quasi-periodicity.
-
-Another commonly used restrictive method uses the fact that, in the limit of
-evenly-spaced data and homoscedastic uncertainties, the covariance matrix is
-``Toeplitz'' \citep[for example][]{Dillon:2013}.
+As discussed previously, the \celerite\ method is based upon a method
+developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
+In the case where the kernel function has a single term of the form
+\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
+and the Rybicki \& Press method.
+Their method does not require solving the extended matrix because the matrix
+inverse can be calculated analytically but the method is made slightly more
+complicated when white noise is included.
+We implement their full method and find empirically that it is more
+computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$
+but we note that the requirement of a single, real, exponential kernel is
+extremely restrictive and this model cannot describe quasi-periodic
+variability.
+
+Another popular method uses the fact that, in the limit of evenly-spaced data
+and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
+\citep[for example][]{Dillon:2013}.
 There are exact methods for solving Toeplitz matrix equations that scale as
@@
 Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
-every CARMA model corresponds to a \celerite\ and the parameters $a_j$, $b_j$,
-$c_j$, and $d_j$ can be easily computed analytically.
+every CARMA model corresponds to an equivalent \celerite\ model and the
+parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
 The inverse statement is not true, however, for any $J > 1$.
@@
 CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
-faster and the memory requirements are smaller.
+somewhat faster and the memory requirements are smaller.
 However, as discussed in \sect{sho}, \celerite\ models have physical a
@@
 This means that the HODLR solver can \emph{approximately} evaluate \celerite\
-models more efficiently than \celerite\ for large models $J \gg 10$ with a
-small dataset where $N$ is smaller than a few thousand.
+models more efficiently than \celerite\ for large models $J \gg 10$ and
+small datasets where $N$ is smaller than a few thousand.
 
@@
 discuss the comparisons exhaustively.
+
+\dfmtodo{One more sentence in conclusion.}
 

### 2017-03-22 paper/ms.tex
small edits to the intro

@@
 \newcommand{\etc}{\foreign{etc.}}
+\newcommand{\ie}{\foreign{i.e.}}
 
@@
 GP models have been widely used across the physical sciences but their
-application is generally limited to small datasets because the computational
-cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
-is $\mathcal{O}(N^3)$.
+application is generally limited to small datasets because the cost of
+computing the inverse and determinant of the matrix $K_\bvec{\alpha}$ is
+$\mathcal{O}(N^3)$.
 In other words, this cost is proportional to the cube of the number of data
@@
 
-In the following Section, we present a method of substantially improving this
+In the following Section, we present a method for substantially improving this
 scaling in many circumstances.
@@
 of light in a vacuum.} The \celerite\ method requires using a specific model
-for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has
-several limitations but, in subsequent sections, we demonstrate that it can be
-used to increase the computational efficiency of many astronomical data
-analysis problems.
+for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
+has some limitations, we demonstrate in subsequent sections that it can
+increase the computational efficiency of many astronomical data analysis
+problems.
 The main limitation of this method is that it can only be applied to
-one-dimensional datasets.
-When we say ``one-dimensional'' here, it means that the \emph{input
-coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
-using $t$ as the input coordinate because one-dimensional GPs are often
-applied to time series data but this isn't a real restriction and the \celerite\
-method can be applied to \emph{any} one-dimensional dataset.}
-Furthermore, the covariance function for the \celerite\ method is ``stationary''.
-This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
-of $\tau_{nm} \equiv |t_n - t_m|$.
+one-dimensional datasets, where by ``one-dimensional'' we mean that the
+ \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
+t_n$.\footnote{We are using $t$ as the input coordinate because
+one-dimensional GPs are often applied to time series data but this isn't a
+real restriction and the \celerite\ method can be applied to \emph{any}
+one-dimensional dataset.}
+Furthermore, the covariance function for the \celerite\ method is
+``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
+$\tau_{nm} \equiv |t_n - t_m|$.
 

### 2017-03-22 paper/ms.tex
more small edits and comments in the intro

@@
 \newcommand{\agoltodo}[1]{\todo{Agol}{blue}{#1}}
+\newcommand{\racomment}[1]{{\color{magenta}#1}}
 
@@
  \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
-t_n$.\footnote{We are using $t$ as the input coordinate because
+t_n$.\footnote{We use $t$ as the input coordinate because
 one-dimensional GPs are often applied to time series data but this isn't a
@@
 and, for this function, \eq{gp-likelihood} can still be evaluated with
-$\mathcal{O}(N)$ operations.
+$\mathcal{O}(N)$ operations. \racomment{Not $\mathcal{O}(NJ^2)$?}
 The details of this method and a few implementation considerations are
-discussed in the following \sectionname\ but we first discuss some
+outlined in the following \sectionname\ but we first discuss some
 properties of this covariance function.
@@
 \end{eqnarray}
-Therefore, the method described in the following section can be used to scale
-inferences to large datasets for any model where the kernel function is a
-general sum or product of \celerite\ terms.
+Therefore, the method described in the following section can be used to
+perform scalable inference on large datasets for any model where the kernel
+function is a sum or product of \celerite\ terms.
 

### 2017-03-22 paper/ms.tex
a few edits to section 4

@@
 The growing field of large-scale time domain astronomy requires methods for
-probabilistic inference that are computationally tractable even with large
+probabilistic inference that are computationally tractable, even with large
 datasets.
@@
 
-\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$ where the
-elements are given by equation \eq{kernel-simple} could be computed
-efficiently by taking advantage of the structure of this covariance function
-and \citet{Ambikasaran:2015} generalized this computation to apply to the full
+\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
+elements given by equation \eq{kernel-simple}, could be computed
+efficiently by taking advantage of the structure of this covariance function.
+\citet{Ambikasaran:2015} generalized this computation to apply to the full
 mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
@@
 This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and this is the model that was studied by
+\eq{celerite-kernel-complex} and is the model studied by
 \citet{Rybicki:1995}.
-To demonstrate this method, we write out the full system of
-equations that we must solve to apply the inverse of $K$ in order to compute
+To demonstrate that the likelihood of this model can be computed in
+$\mathcal{O}(N)$, we write out the full system of
+equations that must be solved to apply the inverse of $K$ and compute
 the first term of \eq{gp-likelihood}.
@@
 \end{eqnarray}
-The system defined by \eq{algo-first} through \eq{algo-last} can be
-rewritten as a matrix equation to show the benefit that this seemingly trivial
+Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
+as a matrix equation shows the benefit that this seemingly trivial
 reformulation provides:
@@
 \end{eqnarray}
-where we follow \citet{Ambikasaran:2015} and call this the ``extended''
-system and rewrite \eq{impl-matrix} as
+Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
+rewrite \eq{impl-matrix} as
 \begin{eqnarray}
@@
 \end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
+Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
 with, it is now sparse with banded structure that can be exploited to solve
@@
 operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
-general~--~and we can take advantage of these algorithms to solve our system
+general~--~and we can use these algorithms to solve our system
 exactly because the target vector $\bvec{z}$ is a subset of the elements of
@@
 
-In the following section, we discuss this method more generally but it's
+In the following section we discuss this method more generally, but it's
 worth noting a few important facts that can already be seen in this example.
@@
 points be monotonically sorted in time.
-Neither of these properties be satisfied in general for multidimensional
+Neither of these properties will be satisfied in general for multidimensional
 inputs and all of our following discussion assumes a sorted
@@
 First, even if the covariance function is a mixture of exponentials, the
-extended matrix is still be banded with a bandwidth that scales linearly
-with the number of components $J$.
-Second, \citet{Ambikasaran:2015} proved that the determinant of
-$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
-This means that we can use this extended matrix formalism to compute the
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components, $J$.
+Second, \citet{Ambikasaran:2015} proved that the absolute value of the
+determinant of $K_\mathrm{ext}$ is equal to the determinant of $K$.
+This means we can use this extended matrix formalism to compute the
 marginalized likelihood in $\mathcal{O}(N)$ operations.

### 2017-03-22 paper/ms.tex
edits to conclusion

--- 2017-03-22 paper/ms.tex
@@
 \newcommand{\etc}{\foreign{etc.}}
-\newcommand{\ie}{\foreign{i.e.}}
 
@@
 
-Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
-models consisting of a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
-covariance, autocorrelation, or ``kernel'' function
-$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the parameters
-$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
+GP are stochastic models consisting of a mean function
+$\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
+function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the
+parameters $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
 Under this model, the log-likelihood of observing a dataset
@@
 GP models have been widely used across the physical sciences but their
-application is generally limited to small datasets because the cost of
-computing the inverse and determinant of the matrix $K_\bvec{\alpha}$ is
-$\mathcal{O}(N^3)$.
+application is generally limited to small datasets because the computational
+cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
+is $\mathcal{O}(N^3)$.
 In other words, this cost is proportional to the cube of the number of data
@@
 
-In the following Section, we present a method for substantially improving this
+In the following Section, we present a method of substantially improving this
 scaling in many circumstances.
@@
 of light in a vacuum.} The \celerite\ method requires using a specific model
-for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
-has some limitations, we demonstrate in subsequent sections that it can
-increase the computational efficiency of many astronomical data analysis
-problems.
+for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has
+several limitations but, in subsequent sections, we demonstrate that it can be
+used to increase the computational efficiency of many astronomical data
+analysis problems.
 The main limitation of this method is that it can only be applied to
-one-dimensional datasets, where by ``one-dimensional'' we mean that the
- \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
-t_n$.\footnote{We are using $t$ as the input coordinate because
-one-dimensional GPs are often applied to time series data but this isn't a
-real restriction and the \celerite\ method can be applied to \emph{any}
-one-dimensional dataset.}
-Furthermore, the covariance function for the \celerite\ method is
-``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
-$\tau_{nm} \equiv |t_n - t_m|$.
+one-dimensional datasets.
+When we say ``one-dimensional'' here, it means that the \emph{input
+coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
+using $t$ as the input coordinate because one-dimensional GPs are often
+applied to time series data but this isn't a real restriction and the \celerite\
+method can be applied to \emph{any} one-dimensional dataset.}
+Furthermore, the covariance function for the \celerite\ method is ``stationary''.
+This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
+of $\tau_{nm} \equiv |t_n - t_m|$.
 
@@
 
-The computational burden of Gaussian process regression has limited its application
-to small astronomical time series.  With the linear scaling of \celerite\, we
-envision the application of Gaussian processes may be expanded to much larger
-datasets.  Despite the restrictive form of the \celerite\ kernel, with a
-sufficient number of components it is flexible enough to describe a wide
-range of astrophysical variability.  In fact, the relation of the \celerite\ kernel
-to the damped, driven harmonic oscillator matches simple models of astrophysical
-variability, and makes the parameterization interpretable in terms of resonant
-frequency, amplitude, and quality factor.  Perhaps the most significant drawback
-is the $J^2$ scaling of the solver;  however, for many problems a small value of $J$
-suffices.
-
-We envision two general varieties of applications:  those interested in the
-the \celerite\ kernel parameters (so-called ``hyper-parameters") which describe the
-astrophysical variability, versus those interested in characterizing the mean
-model, treating the correlated noise as a nuisance.  Some users may
-be interested in both applications, while other problems may require interpolation
-or extrapolation beyond the dataset that is being modeled.
-
-%Some examples of the first are determining the characteristic variability
-%timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
-%variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
-%granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
-%quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
-%classification of variable objects \citep{Zinn:2016}.
-%Examples of the second are detecting and characterizing transiting exoplanets
-%\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
-%Grunblatt:2016, Luger:2016}
-%and correcting for stellar activity in radial velocity measurements
-%\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
-%Examples of the third are measuring time delays in multiply-imaged
-%gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
-%AGN.
+Gaussian Process models have been fruitfully applied to many problems in
+astronomical data analysis but the fact that the computational cost scales as
+the cube of the number of data points has limited their use to relatively
+small datasets.
+With the linear scaling of \celerite\, we envision the application of Gaussian
+processes may be expanded to much larger datasets.
+Despite the restrictive form of the \celerite\ kernel, with a sufficient
+number of components it is flexible enough to describe a wide range of
+astrophysical variability.
+In fact, the relation of the \celerite\ kernel to the damped, driven harmonic
+oscillator matches simple models of astrophysical variability, and makes the
+parameterization interpretable in terms of resonant frequency, amplitude, and
+quality factor.
+A major drawback of this method is its quadratic scaling with the number of
+terms $J$ but, in many cases, small values of $J$ are sufficient.
 
@@
 recently begun to adopt full covariance matrices in analyzing the noise in
-stellar lightcurves when detecting or characterizing transiting planets
-\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
-Grunblatt:2016, Luger:2016}.
-One promising early attempt at accounting for correlated noise was the wavelet
-approach of \citet{Carter:2009}.
-Their technique allowed for a power-law spectrum for the noise scaling as
-$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
-physically-accurate, description of stellar variability.
-Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
-as `flicker', $1/f$, or pink noise;  however, this particular type of noise
-overpredicts the power spectrum at low and high frequency compared to the Solar
-granulation power spectrum.  It also cannot describe quasi-periodic noise.
-
-Further progress was made by parameterizing the covariance matrix with simple,
-analytic functions that describe the autocorrelation function \citep{Gibson:2012}.
-The parameters of these functions can then be optimized to best match the observed
-variability pattern of the residuals (after subtracting a transit model).  The
-disadvantage of this approach is that the functions used are chosen somewhat
-arbitrarily (e.g.\ the exponential-squared function), and the Cholesky decomposition
-of the covariance matrix for computing the likelihood takes ${\cal O}(N^3)$ operations
-which prohibits application of this technique to large datasets.
-
-Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
-computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
-that accurately describes stellar variability due to the relation to the simple
-harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
-oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
-are obtained, the effects of stellar variability will more dramatically impact the
-correct inference of planetary transit parameters, and so we expect that \celerite\
-will be important for transit detection \citep{ForemanMackey:2015}, transit timing
-\citep{Agol:2005,Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler
-beaming \citep{Loeb:2003,Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
-functions \citep{Knutson:2007,Zucker:2007}, and more.
-
-Although we have in mind applications of this fast method to model stellar
-variability, the method is general for one-dimensional GP problems, and may be
-applied to other problems.
-Accreting black holes show time series which may be modeled by correlated
-noise \citep{Kelly:2014};  indeed, this was the motivation for the original
-technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
+stellar light curves when detecting or characterizing transiting planets
+\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
+Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
+All of these analyses have been limited to small datasets or restrictive
+kernel choices.
+\celerite\ weakens these requirements by providing a scalable method for
+computing the likelihood and a physical motivation for the choice of kernel.
+\celerite\ can be used to accurately model stellar oscillations using the
+relation to the mixture of stochastically driven, damped simple harmonic
+oscillators.
+As higher signal-to-noise observations of transiting exoplanet systems are
+obtained, the effects of stellar variability will more dramatically impact the
+correct inference of planetary transit parameters.
+We expect that \celerite\ will be important for transit detection
+\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
+Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
+\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
+functions \citep{Knutson:2007, Zucker:2007}, and more.
+
+Beyond these applications to model stellar variability, the method is
+generally applicable to other one-dimensional GP models.
+Accreting black holes show time series which may be modeled using a GP
+\citep{Kelly:2014}; indeed, this was the motivation for the original technique
+developed by Rybicki \& Press \citep{Rybicki:1992, Rybicki:1995}.
 This approach may be broadly used for characterizing quasar variability
 \citep{MacLeod:2010}, measuring time lags with reverberation mapping
-\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
-gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic
-variability in a high-energy source \citep{McAllister:2016}, or
+\citep{Zu:2011, Pancoast:2014}, modeling time delays in multiply-imaged
+gravitationally-lensed systems \citep{Press:1998}, characterizing
+quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
 classification of variable objects \citep{Zinn:2016}.
-Outside of astronomy, this technique may have application to terrestrial
-seismology \citep{Robinson:1967}.
-%The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
-%non-parametric data modeling \citep{Wilson:2013}.
-
-The \celerite\ formalism also allows a principled approach to the estimation
-of the power spectrum of a time series, as well as its uncertainties.
-Fourier transforms work for uniformly sampled data;  astronomical data are
-rarely (if ever) uniformly sampled.  Periodograms are affected by the window
-function and aliasing in datasets, and make estimating the significance of
-spectral features ambiguous.  The Bayesian approach to the estimation of a power
-spectrum \citep{Wilson:2013} avoids these difficulties:  it allows for the computation of
-uncertainties, it accounts for the window function on the uncertainties of
-the power spectrum, and in principle it can treat aliases if the full posterior
-is explored.  One drawback is that the number of $J$ terms to include is
-unknown, but may be inferred with computation of the evidence, or an
-approximation such as the Bayesian information criterion \citep{Kelly:2014}.
-In general, the parameters of the \celerite\ terms are more interesting due
-to their physical meaning rather than the overall shape of the power spectrum,
-as shown in the foregoing examples.
+We expect that there are also applications beyond astronomy.
+
+The \celerite\ formalism can also be used for power spectrum estimation and
+quantification of its uncertainties.
+In principle, a large number of \celerite\ terms could be used to perform
+non-parametric probabilistic inference of the power spectrum despite
+unevenly-spaced data with heteroscedastic noise \citep[for
+example,][]{Wilson:2013, Kelly:2014}.
+This type of analysis will be limited by the quadratic scaling of \celerite\
+with the number of terms $J$ but this limits existing methods as well
+\citep[CARMA models,][]{Kelly:2014}.
+In general, we encourage the use of physically motivated models for parameter
+estimation instead of qualitative modeling of the power spectrum itself.
+
+There are many data analysis problems where \celerite\ will not be applicable.
+In particular, any multidimensional
 

### 2017-03-22 paper/ms.tex

@@
 of light in a vacuum.} The \celerite\ method requires using a specific model
-for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has
-several limitations but, in subsequent sections, we demonstrate that it can be
-used to increase the computational efficiency of many astronomical data
-analysis problems.
+for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it
+has some limitations, we demonstrate in subsequent sections that it can
+increase the computational efficiency of many astronomical data analysis
+problems.
 The main limitation of this method is that it can only be applied to
-one-dimensional datasets.
-When we say ``one-dimensional'' here, it means that the \emph{input
-coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
-using $t$ as the input coordinate because one-dimensional GPs are often
-applied to time series data but this isn't a real restriction and the \celerite\
-method can be applied to \emph{any} one-dimensional dataset.}
-Furthermore, the covariance function for the \celerite\ method is ``stationary''.
-This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
-of $\tau_{nm} \equiv |t_n - t_m|$.
+one-dimensional datasets, where by ``one-dimensional'' we mean that the
+ \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
+t_n$.\footnote{We use $t$ as the input coordinate because
+one-dimensional GPs are often applied to time series data but this isn't a
+real restriction and the \celerite\ method can be applied to \emph{any}
+one-dimensional dataset.}
+Furthermore, the covariance function for the \celerite\ method is
+``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
+$\tau_{nm} \equiv |t_n - t_m|$.
 

### 2017-03-22 paper/ms.tex
I don't approve of the use of 'i.e.'

@@
 Furthermore, the covariance function for the \celerite\ method is
-``stationary'', \ie\ $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of
-$\tau_{nm} \equiv |t_n - t_m|$.
+``stationary''.
+In other words, $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function of $\tau_{nm}
+\equiv |t_n - t_m|$.
 

### 2017-03-22 paper/ms.tex
adding comparison to Carter & Winn

@@
 They have also been used as models for the cosmic microwave background
-\citep{Bond:1987,Bond:1999}, correlated instrumental noise
-\citep{Gibson:2012}, and spectroscopic calibration
-\citep{Czekala:2017,Evans:2015}.
+\citep{Bond:1987, Bond:1999, Wandelt:2003}, correlated instrumental noise
+\citep{Gibson:2012}, and spectroscopic calibration \citep{Czekala:2017,
+Evans:2015}.
 While these models are widely applicable, their use has been limited, in
@@
 and, for this function, \eq{gp-likelihood} can still be evaluated with
-$\mathcal{O}(N)$ operations. \racomment{Not $\mathcal{O}(NJ^2)$?}
+$\mathcal{O}(N\,J^2)$ operations.
 The details of this method and a few implementation considerations are
@@
 
+\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
+wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
+scaling.
+This method has been widely applied in the context of exoplanet transit
+characterization but it requires evenly spaced observations and the power
+spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
+gain the computational advantage.
+This wavelet method has been demonstrated to improve parameter estimation for
+transiting exoplanets \citep{Carter:2009} but these strict requirements make
+this method applicable for only a limited set of use cases.
+
 The continuous autoregressive moving average (CARMA) models introduced into
@@
 There are many data analysis problems where \celerite\ will not be applicable.
-In particular, any multidimensional
-
-In the future we hope to expand the \celerite\ technique to multi-dimensional
-datasets.  The formalism presented could, for example, be used to model
-the covariance between multiple measured quantities, such as in regression of photometric
-time series versus astrometric precision when correcting for systematic errors
-due to detector sensitivity variations \citep{Aigrain:2016}, in correcting
-radial velocity datasets for stellar activity \citep{Haywood:2014,Rajpaul:2015},
-or in detection of gravitational waves in the presence of systematic errors
-\citep{Moore:2016}.  Other extensions that may be useful are fast simulation
-of \celerite\ processes, prediction and/or interpolation of datasets, and
-extended the \celerite\ kernel to time-varying kernels.
-
-%\section{Prediction/interpolation}
-%
-%Precition/interpolation involves constructing a matrix of covariances
-%between the (noisy) training set and the times at which we wish to
-%predict or interpolate data points.   If our training times are given by
-%${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
-%mean values at the predicted times are given by (RW 2.23):
-%\begin{eqnarray}
-%\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
-%&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
-%&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
-%y_k &=& \sum_j k(t_j,t_k) h_j,
-%\end{eqnarray}
-%where the latter equation may be solved for ${\bf h}$ using the extended
-%matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
-%matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
-%of $MN$ multiplications is required to obtain the predicted values.
-%If $M$ is large, this can become quickly prohibitive.  It turns out
-%that the structure of the sum of exponential kernels may be exploited
-%to obtain the predicted mean in of order $J(M+N)+J^2N$ multiplications,
-%i.e.\ a linear scaling with the total number of data points, which can
-%be much fewer if $M$ is large.
-%
-%The procedure works as follows:
-%\begin{itemize}
-%\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
-%\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
-%compute:
-%\begin{eqnarray} \label{prediction}
-%\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
-%\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} h_j\\
-%\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} h_j,
-%\end{eqnarray}
-%where we have separated out the training times before and after the predicted time $t_{*,k}$.
-%
-%\item Next, we update the coefficients recursively:
-%\begin{eqnarray}
-%\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} h_j\right]\\
-%\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} h_j\right].
-%\end{eqnarray}
-%With these updates, we can then use equation \ref{prediction} to compute the expected
-%mean at $t_{*,k}$ with an additional $J$ operations.  Since there are $N$ values of
-%$t_k$, the number of operations needed to update $\epsilon$ is of order $J(N+M)$, and
-%so this procedure avoids a heavy computational burden.
-%
-%The variance may be handled in a similar manner, but with $J^2$ coefficients
-%to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
-%\begin{eqnarray}
-%cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
-%&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
-%\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
-%&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
-%\end{eqnarray}
-%The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
-%\end{itemize}
-%
-%Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
-%then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
-%just solve for the difference in fewer operations? {\bf TBD}
-
-\vspace{1.5em}
-All of the code used in this project is available from
-\url{https://github.com/dfm/celerite} under the MIT open-source software
-license.
-This code (plus some dependencies) can be run to re-generate all of the
-figures and results in this paper; this version of the paper was generated
-with git commit \texttt{\githash} (\gitdate).
+In particular, the restriction to one-dimensional problems is significant.
+There are many examples of multidimensional GP modeling the astrophysics
+literature \citep[recent examples from the field of exoplanet characterization
+include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016} but \celerite\ could not
+be used to accelerate any of these analyses.
+It is plausible that an extension could be derived to tackle some
+multidimensional problems with the right structure~--~simultaneous parallel
+time series, for example~--~and we hope to revisit this possibility in future
+work.
+
+Alongside this paper, we have released a well tested and documented open
+source software package that implements the method and all of the examples
+discussed in these pages.
+This software is available at \url{https://github.com/dfm/celerite}\footnote{%
+This version of the paper was generated with git commit \texttt{\githash}
+(\gitdate).} and it is made available under the MIT license.
 
@@
 
-This work was performed in part under contract with the Jet Propulsion Laboratory (JPL)
-funded by NASA through the Sagan Fellowship Program executed by the NASA Exoplanet Science Institute.
+This work was performed in part under contract with the Jet Propulsion
+Laboratory (JPL) funded by NASA through the Sagan Fellowship Program executed
+by the NASA Exoplanet Science Institute.
 

### 2017-03-22 paper/ms.tex
summary paragraph in comparison section

@@
 
+In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
+case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
+will also be identically zero for all $n$.
+Removing these trivial rows from the extended matrix results in factor of two
+increase in the computational efficiency for these terms.
+In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
+$d_j=0$, and the general terms are called ``complex''.
+
 \begin{figure}[!htbp]
@@
 Many other approximate methods for scaling GP inference exist \citep[see, for
-example,][and references therein]{Wilson:2015a} and we make no attempt to
-discuss the comparisons exhaustively.
-
-\dfmtodo{One more sentence in conclusion.}
+example,][and references therein]{Wilson:2015a} and we make no attempt to make
+our discussion exhaustive.
+The key take away here is that \celerite\ provides and \emph{exact} method for
+GP inference for a specific but flexible class of one-dimensional kernel
+functions.
+The closest cousin is the CARMA method with similar strengths and limitations
+but, since there is a simpler relationship between the parameters of a
+\celerite\ model and its behavior, \celerite\ can be more easily integrated
+with physical or physically interpretable models.
 

### 2017-03-23 paper/ms.tex
minor edits

@@
 can be used with any stationary kernel but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it, in general, be less
+and the scaling is worse than \celerite\ so it, in general, is less
 efficient when applied to large datasets.
@@
 our discussion exhaustive.
-The key take away here is that \celerite\ provides and \emph{exact} method for
+The key take away here is that \celerite\ provides an \emph{exact} method for
 GP inference for a specific but flexible class of one-dimensional kernel
@@
 astrophysical variability.
-In fact, the relation of the \celerite\ kernel to the damped, driven harmonic
+In fact, the relation of the \celerite\ kernel to the damped, stochastically-driven harmonic
 oscillator matches simple models of astrophysical variability, and makes the
@@
 \celerite\ can be used to accurately model stellar oscillations using the
-relation to the mixture of stochastically driven, damped simple harmonic
+relation to the mixture of stochastically-driven, damped simple harmonic
 oscillators.
@@
 In the case of $J$ \celerite\ terms, we can check for negative values of the
-PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
+PSD by solving for the roots of the power spectrum; if there are any real,
+positive roots, then the power-spectrum goes negative (or zero), and thus does not
+represent a valid kernel. We rewrite the power spectrum, \eqalt{celerite-psd}),
 abbreviating with $z = \omega^2$:

### 2017-03-23 paper/ms.tex
a few more edits

--- 2017-03-22 paper/ms.tex
@@
 
-Many astronomical objects have characteristic timescales determined by their
+Many astronomical objects are variable on timescales determined by their
 physical structure.
-Variability is excited by noisy physical processes, and grows most strongly
-at the characteristic timescale, but is also damped due to dissipation in the
-system.
-As a concrete example, stellar (asteroseismic) oscillations are strong at
-resonant frequencies determined by the internal stellar structure, which are
-both excited and damped by convective turbulence.
+For phenomena such as stellar (asteroseismic) oscillations, variability is
+excited by noisy physical processes and grows most strongly at the
+characteristic timescale but is also damped due to dissipation in the system.
+These oscillations are strong at resonant frequencies determined by the
+internal stellar structure, which are both excited and damped by convective
+turbulence.
 
@@
 We note that, because of the damping, the characteristic oscillation frequency
-in this model $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
-the frequency of the undamped oscillator $\omega_0$.
+in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
+the frequency of the undamped oscillator, $\omega_0$.
 
@@
 
-For this example, we simulate a dataset using a known GP model with a kernel
+For this examplewe simulate a dataset using a known GP model with a kernel
 outside of the support of a \celerite\ process.
-This means that the true autocorrelation of the process can never be correctly
+This means the true autocorrelation of the process can never be correctly
 represented by the model that we're using to fit but we use this example
@@
 
-For this example, the data are simulated from a quasiperiodic GP with the
+For this example the data are simulated from a quasiperiodic GP with the
 kernel
@@
 \end{eqnarray}
-that for large $\omega$ falls off exponentially.
-When compared to \eq{celerite-psd}~--~that for large $\omega$ goes as
-$\omega^{-4}$, at most~--~it is clear that a \celerite\ model can never
+which, for large values of $\omega$, falls off exponentially.
+When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
+$\omega^{-4}$ at most~--~it is clear that a \celerite\ model can never
 perfectly reproduce the structure of this process.
@@
 applied to real datasets.
-Each of these examples touches on an active area of research so we limit
-our examples to be qualitative in nature and make no claim of optimality but
-we hope that these examples encourage interested readers to investigate
-the applicability of \celerite\ to their research.
+Each of these examples touches on an active area of research so we limit our
+examples to be qualitative in nature and do not claim that \celerite\ is the
+optimum method but we hope these examples encourage interested readers to
+investigate the applicability of \celerite\ to their research.
 
@@
 To demonstrate the applicability of this model, we apply it to infer the
-asteroseismic parameters of the giant star KIC~11615890 observed by the
+asteroseismic parameters of the giant star KIC~11615890, observed by the
 \kepler\ Mission.
@@
 The key difference between this function and other quasiperiodic kernels is
-that it is everywhere positive.
+that it is positive everywhere.
 We can construct a simple \celerite\ covariance function with similar properties
@@
 \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
-This star has a published rotation period of $3.88 \pm 0.58$ measured using
+This star has a published rotation period of $3.88 \pm 0.58$, measured using
 traditional periodogram and autocorrelation function approaches applied to
@@
 We used \emcee\ to sample the joint posterior probability density for the four
-parameters in \eq{rot-kernel} conditioned on two quarters of \kepler\ data and
-find a constraint on the rotation period of $P = \rotationperiod$ in good
-agreement with the literature value.
+parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
+and find a constraint on the rotation period of $P = \rotationperiod$.
+This is in good agreement with the literature value.
 \Figure{rotation} shows a subset of the data used in this example and the
@@
 parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
-The inverse statement is not true, however, for any $J > 1$.
-This means that \celerite\ could be used to compute any CARMA model but using
+The inverse statement is not true however, for any $J > 1$.
+This means \celerite\ could be used to compute any CARMA model but using
 CARMA to compute the likelihood of a \celerite\ model would require solving
@@
 somewhat faster and the memory requirements are smaller.
-However, as discussed in \sect{sho}, \celerite\ models have physical a
+However, as discussed in \sect{sho}, \celerite\ models have a physical
 interpretation as a mixture of stochastically-driven damped harmonic
 oscillators.
-This enabled, for example, the use of \celerite\ as a model for asteroseismic
+This enables, for example, the use of \celerite\ as a model for asteroseismic
 oscillations.

### 2017-03-23 paper/ms.tex
formatting

--- 2017-03-23 paper/ms.tex
@@
 
-\widowpenalty=10000
-\clubpenalty=10000
-
 \sloppy\sloppypar
+
+%\widowpenalty=10000
+\clubpenalty=1000000000
 
@@
 \title{%
+\vspace{-\baselineskip}
 Fast and scalable Gaussian process modeling with applications
 to astronomical time series
-\vspace{-3\baselineskip}
+\vspace{-3.2\baselineskip}
 }
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[p]
 \begin{center}
@@
 can be used with any stationary kernel but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it, in general, is less
+and the scaling is worse than \celerite\ so it, in general, be less
 efficient when applied to large datasets.
@@
 our discussion exhaustive.
-The key take away here is that \celerite\ provides an \emph{exact} method for
+The key take away here is that \celerite\ provides and \emph{exact} method for
 GP inference for a specific but flexible class of one-dimensional kernel
@@
 astrophysical variability.
-In fact, the relation of the \celerite\ kernel to the damped, stochastically-driven harmonic
+In fact, the relation of the \celerite\ kernel to the damped, driven harmonic
 oscillator matches simple models of astrophysical variability, and makes the
@@
 \celerite\ can be used to accurately model stellar oscillations using the
-relation to the mixture of stochastically-driven, damped simple harmonic
+relation to the mixture of stochastically driven, damped simple harmonic
 oscillators.
@@
 In the case of $J$ \celerite\ terms, we can check for negative values of the
-PSD by solving for the roots of the power spectrum; if there are any real,
-positive roots, then the power-spectrum goes negative (or zero), and thus does not
-represent a valid kernel. We rewrite the power spectrum, \eqalt{celerite-psd}),
+PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
 abbreviating with $z = \omega^2$:

### 2017-03-23 paper/ms.tex
fixing #78 - Equation A4
adding references to Mcquillan 2013, 2014

--- 2017-03-23 paper/ms.tex
@@
 the empirical autocorrelation function provides a reliable estimate of the
-rotation period of a star \citep{Aigrain:2015} and that a GP model with a
+rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
+and that a GP model with a
 quasiperiodic covariance function can be used to make probabilistic
@@
 \begin{eqnarray}
-    -|b_j\,d_j| < a_j\,c_j < |b_j\,d_j| \quad.
+    |b_j\,d_j| < a_j\,c_j \quad.
 \end{eqnarray}

### 2017-03-23 paper/ms.tex
tiny edit

--- 2017-03-23 paper/ms.tex
@@
 Second, \citet{Ambikasaran:2015} proved that the absolute value of the
-determinant of $K_\mathrm{ext}$ is equal to the determinant of $K$.
+determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
+determinant of $K$.
 This means we can use this extended matrix formalism to compute the
@@
 
-\begin{figure}[!htbp]
+\begin{figure}[tp]
 \begin{center}
@@
 
+\newpage
 \section{Comparisons to other methods}

### 2017-03-24 paper/ms.tex
Typo :0

--- 2017-03-23 paper/ms.tex
@@
 
-\begin{figure}[p]
+\begin{figure}[tp]
 \begin{center}
@@
 
-For this examplewe simulate a dataset using a known GP model with a kernel
+For this example we simulate a dataset using a known GP model with a kernel
 outside of the support of a \celerite\ process.

### 2017-03-24 paper/ms.tex
Pedantic footnote placements.

@@
 range of potential applications within astronomical data analysis and beyond.
-We provide well tested and documented open-source implementations of this
+We provide well-tested and documented open-source implementations of this
 method in \project{C++}, \project{Python}, and \project{Julia}.
@@
 application on various real and simulated data sets.
-Alongside this paper, we have released well tested and documented open source
+Alongside this paper, we have released well-tested and documented open source
 implementations written in \project{C++}, \project{Python}, and
@@
 
-Similarly, setting both $b_j$ and $d_j$ to zero, we get a Ornstein--Uhlenbeck
+Similarly, setting both $b_j$ and $d_j$ to zero, we get an Ornstein--Uhlenbeck
 process
@@
     methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines\footnote{The experiments below use the Intel Math
+    \project{BLAS} routines,\footnote{The experiments below use the Intel Math
     Kernel Library (\project{MKL}):
-    \url{https://software.intel.com/en-us/intel-mkl}.}, and
+    \url{https://software.intel.com/en-us/intel-mkl}.} and
 
@@
 our discussion exhaustive.
-The key take away here is that \celerite\ provides an \emph{exact} method for
+The key takeaway here is that \celerite\ provides an \emph{exact} method for
 GP inference for a specific but flexible class of one-dimensional kernel
@@
 
-Alongside this paper, we have released a well tested and documented open
+Alongside this paper, we have released a well-tested and documented open
 source software package that implements the method and all of the examples

### 2017-03-24 paper/ms.tex
small abstract edits

@@
 datasets.
-We present a method for Gaussian Process regression in one dimension with a
-computational cost that scales linearly with the size of the dataset.
+Gaussian Processes are a popular class of models used for this purpose but,
+since the computational cost scales as the cube of the number of data points,
+their application has been limited to relatively small datasets.
+In this paper, we present a method for Gaussian Process regression in
+one-dimension where the computational requirements scale linearly with the
+size of the dataset.
 This method exploits structure in the problem when the covariance function is

### 2017-03-24 paper/ms.tex
edits from @sivaramambikasaran

@@
 
-\edef \iss {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\iss}{Department of Computational and Data Sciences,
+\edef \iis {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\iis}{Department of Computational and Data Sciences,
                     Indian Institute of Science, Bangalore, India}
@@
     Ruth~Angus\altaffilmark{\simons,\columbia}, and
-    Sivaram~Ambikasaran\altaffilmark{\iss}
+    Sivaram~Ambikasaran\altaffilmark{\iis}
 }
@@
 
-In the following Section, we present a method for substantially improving this
+In the following section, we present a method for substantially improving this
 scaling in many circumstances.
@@
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
-evaluated in $\mathcal{O}(N\,J^2)$ where $J$ is the number of components
+evaluated in $\mathcal{O}(N\,J^2)$, where $J$ is the number of components
 in the mixture.
@@
 The details of this method and a few implementation considerations are
-outlined in the following \sectionname\ but we first discuss some
+outlined in the following section but we first discuss some
 properties of this covariance function.
@@
 Therefore, the method described in the following section can be used to
-perform scalable inference on large datasets for any model where the kernel
+perform scalable inference on large datasets for any model, where the kernel
 function is a sum or product of \celerite\ terms.
@@
 \citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
-elements given by equation \eq{kernel-simple}, could be computed
+elements given by \eq{kernel-simple}, could be computed
 efficiently by taking advantage of the structure of this covariance function.
@@
 $K$ and vector \bvec{y}.
-In this equation, we have assumed that the mean function is zero but a
+In \eq{impl-matrix}, we have assumed that the mean function is zero but a
 non-zero mean could be included by replacing \bvec{y} by
@@
 a few percent non-zero entries and band structure.
-In this \sectionname, we empirically investigate the performance and scaling
+In this section, we empirically investigate the performance and scaling
 of three different algorithms for solving this extended system:
@@
 
-\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$
-with their method that used the sparse LU decomposition implemented in the
-\project{SuperLU} package \citep{Demmel:1999}.
 The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
@@
 bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
+\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
+\project{SuperLU} package \citep{Demmel:1999}.
 We find that, while the \texttt{vanilla} solver scales as expected, the
@@
 
-For this example we simulate a dataset using a known GP model with a kernel
+For this example, we simulate a dataset using a known GP model with a kernel
 outside of the support of a \celerite\ process.
@@
 
-In this section we demonstrate several use cases of \celerite\ when
+In this section, we demonstrate several use cases of \celerite\ when
 applied to real datasets.
@@
 quality factor.
-A major drawback of this method is its quadratic scaling with the number of
+% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
+A drawback of this method is its quadratic scaling with the number of
 terms $J$ but, in many cases, small values of $J$ are sufficient.
@@
 literature \citep[recent examples from the field of exoplanet characterization
-include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016} but \celerite\ could not
+include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
 be used to accelerate any of these analyses.
@@
 which is a polynomial with order $2\,(J-1)+1$.
-With $J=2$, this yields a cubic equation which may be solved exactly for the
-roots.
+With $J=2$, this yields a cubic equation whose roots can be obtained exactly.
 
@@
 Likewise, we evaluate the coefficients of the largest order term in each
-polynomial which gives the sign of the polynomial as $z \rightarrow \infty$,
+polynomial that gives the sign of the polynomial as $z \rightarrow \infty$,
 $f_\infty$.
-With the series of coefficients $f_0$ and $f_\infty$, we then determine how
+With the sequence of coefficients $f_0$ and $f_\infty$, we then determine how
 many times the sign changes in each of these, where $\sigma(0)$ is the number

### 2017-03-25 paper/ms.tex
addressing comments from @jakevdp

@@
 \newcommand{\celerite}{\project{celerite}}
+\newcommand{\celeriteterm}{\emph{celerite}}
 \newcommand{\emcee}{\project{emcee}}
@@
 For this method to be applicable, the data must be one-dimensional and the
-covariance function must have a specific form.
+covariance function must be represented by a mixture of exponentials.
 However, there is no further constraint on the data or the model.
@@
 uncertainties can be heteroscedastic (non-uniform).
+
+This method achieves linear scaling by exploiting structure in the covariance
+matrix when it is generated by a mixture of exponentials.
+More precisely, the matrix can be embedded in an equivalent higher-dimensional
+system that can be solved efficiently using a sparse or banded solver.
+As we will discuss in the following pages, this choice of kernel function is
+not as restrictive as it first seems because it arises naturally in some
+physical systems and it can be used as an effective model in many other cases.
 This method is especially appealing compared to other similar methods~--~we
@@
 \end{eqnarray}
-For clarity, we refer to the argument within the sum as a ``\celerite\
+For clarity, we refer to the argument within the sum as a ``\celeriteterm\
 term'' for the remainder of this paper.
@@
 perform scalable inference on large datasets for any model, where the kernel
-function is a sum or product of \celerite\ terms.
+function is a sum or product of \celeriteterm\ terms.
 
@@
 
-In the case of the full \celerite\ covariance function
+In the case of the full \celeriteterm\ covariance function
 (\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
@@
 the extended matrix $K_\mathrm{ext}$.
+The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
+with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
+non-zero entries.
 Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
@@
 
-We now turn to a discussion of \celerite\ as a model of astrophysical
+We now turn to a discussion of \celeriteterm\ as a model of astrophysical
 variability.
@@
 While this isn't necessarily a problem for the inferred results, in this
-section we discuss an exact physical interpretation of the \celerite\ kernel
+section we discuss an exact physical interpretation of the \celeriteterm\ kernel
 that will be applicable to many astrophysical systems but especially the
@@
 
-To relate this physical picture to \celerite, we consider the dynamics of a
+To relate this physical picture to \celeriteterm, we consider the dynamics of a
 stochastically-driven damped simple harmonic oscillator.
@@
 for $Q \ge \frac{1}{2}$.
-For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
+For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of
+\celeriteterm\
 terms with parameters
@@
 This suggests that the Mat\'ern-3/2 covariance can be approximated using
-the \celerite\ framework with a small value of $f$ in \eq{approx-matern2} but we
+the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
 caution that this might lead to numerical issues for the solver.
@@
 In the first example, we demonstrate that \celerite\ can be used to infer
-the power spectrum of a process when the data are generated from a \celerite\
+the power spectrum of a process when the data are generated from a
+\celeriteterm\
 model.
@@
 
-In this first example, we simulate a dataset using a known \celerite\
+In this first example, we simulate a dataset using a known \celeriteterm\
 process and fit it with \celerite\ to demonstrate that valid inferences can be
@@
 For this example, we simulate a dataset using a known GP model with a kernel
-outside of the support of a \celerite\ process.
+outside of the support of a \celeriteterm\ process.
 This means the true autocorrelation of the process can never be correctly
@@
 When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
-$\omega^{-4}$ at most~--~it is clear that a \celerite\ model can never
+$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
 perfectly reproduce the structure of this process.
@@
 1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
-We note that using \eq{product-rule}, the product of two \celerite\ terms can
+We note that using \eq{product-rule}, the product of two \celeriteterm\ terms can
 also be expressed using \celerite.
@@
 the inferences made using the correct model (\eqalt{sim-wrong-true}).
-The inference made using this effective \celerite\ model are indistinguishable
+The inference made using this effective \celeriteterm\ model are indistinguishable
 from the inferences made using the correct model but substantially less
@@
 We randomly select a month-long segment of \kepler\ data, initialize our
-\celerite\ model using a grid search in the parameter space, and then use
+\celeriteterm\ model using a grid search in the parameter space, and then use
 \emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
@@
 \end{eqnarray}
-where $P$ is the period of the oscillation.
-The key difference between this function and other quasiperiodic kernels is
-that it is positive everywhere.
-We can construct a simple \celerite\ covariance function with similar properties
+where $P$ is the rotation period of the star.
+GP modeling with the same kernel function has been proposed as a method of
+measuring the mean periodicity in quasiperiodic photometric time series in
+general \citep{Wang:2012}.
+The key difference between \eq{sine2} and other quasiperiodic kernels is that
+it is positive everywhere.
+We can construct a simple \celeriteterm\ covariance function with similar properties
 as follows
@@
 
-To demonstrate this method, we fit a \celerite\ model with a kernel given by
+To demonstrate this method, we fit a \celeriteterm\ model with a kernel given by
 \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
@@
 Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
-every CARMA model corresponds to an equivalent \celerite\ model and the
+every CARMA model corresponds to an equivalent \celeriteterm\ model and the
 parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
@@
 This means \celerite\ could be used to compute any CARMA model but using
-CARMA to compute the likelihood of a \celerite\ model would require solving
+CARMA to compute the likelihood of a \celeriteterm\ model would require solving
 \eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
@@
 somewhat faster and the memory requirements are smaller.
-However, as discussed in \sect{sho}, \celerite\ models have a physical
+However, as discussed in \sect{sho}, \celeriteterm\ models have a physical
 interpretation as a mixture of stochastically-driven damped harmonic
 oscillators.
-This enables, for example, the use of \celerite\ as a model for asteroseismic
+This enables, for example, the use of \celeriteterm\ as a model for asteroseismic
 oscillations.
@@
 models when compared to \celerite.
-This means that the HODLR solver can \emph{approximately} evaluate \celerite\
+This means that the HODLR solver can \emph{approximately} evaluate
+\celeriteterm\
 models more efficiently than \celerite\ for large models $J \gg 10$ and
@@
 but, since there is a simpler relationship between the parameters of a
-\celerite\ model and its behavior, \celerite\ can be more easily integrated
+\celeriteterm\ model and its behavior, \celerite\ can be more easily integrated
 with physical or physically interpretable models.
@@
 processes may be expanded to much larger datasets.
-Despite the restrictive form of the \celerite\ kernel, with a sufficient
+Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
 number of components it is flexible enough to describe a wide range of
 astrophysical variability.
-In fact, the relation of the \celerite\ kernel to the damped, stochastically-driven harmonic
+In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
 oscillator matches simple models of astrophysical variability, and makes the
@@
 computing the likelihood and a physical motivation for the choice of kernel.
-\celerite\ can be used to accurately model stellar oscillations using the
+\celeriteterm\ can be used to accurately model stellar oscillations using the
 relation to the mixture of stochastically-driven, damped simple harmonic
@@
 
-The \celerite\ formalism can also be used for power spectrum estimation and
+The \celeriteterm\ formalism can also be used for power spectrum estimation and
 quantification of its uncertainties.
-In principle, a large number of \celerite\ terms could be used to perform
+In principle, a large number of \celeriteterm\ terms could be used to perform
 non-parametric probabilistic inference of the power spectrum despite
@@
 
-Using \eq{celerite-psd}, we find that for a single \celerite\ term, this
+Using \eq{celerite-psd}, we find that for a single \celeriteterm\ term, this
 requirement is met when
@@
 
-In the case of $J$ \celerite\ terms, we can check for negative values of the
+In the case of $J$ \celeriteterm\ terms, we can check for negative values of the
 PSD by solving for the roots of the power spectrum; if there are any real,

### 2017-03-25 paper/figures/matrix.py

@@ -70,7 +70,8 @@ for n in range(1, N):
 
 nband = len(full) * 2*(2*J+2)
 nnz = np.sum(full > 0)
-print("Sparsity: {0}".format(nnz / nband))
+print("nnz: {0}, {1}".format(nnz, (20*J+1)*N - 28*J))
+# print("nnz: {0}, {1}".format(nnz, (20*J+1)*(N-2)+12*J+2))
 
 # Legend
 d = 1

### 2017-03-25 paper/ms.tex
addressing comments from AGW

@@
 
+The structured kernel interpolation \citep[SKI/KISS-GP][]{Wilson:2015}
+framework is another approximate method that can be used to scale GPs to large
+datasets.
+This method uses fast interpolation of the kernel into a space where Toeplitz
+or Kronecker structure can be used to scale inference and prediction.
+The SKI/KISS-GP framework can be applied to scale GP inference with a wide
+range of kernels, but the computational cost will depend on the specific
+dataset, model, and precision requirements for the approximation.
+The SKI method has an impressive $\mathcal{O}(1)$ cost for test-time
+predictions and it is interesting to consider how this could be applied to
+\celeriteterm\ models.
+
 Many other approximate methods for scaling GP inference exist \citep[see, for
@@
 Megan Bedell,
+Ian Czekala,
 Will Farr,
@@
 David W.\ Hogg,
-Dan Huber, and
-Jake VanderPlas
+Dan Huber,
+Dennis Stello,
+Jake VanderPlas, and
+Andrew Gordon Wilson
 for helpful discussions informing the ideas and code presented here.

### 2017-03-25 paper/ms.tex
more comments from AGW

@@
 term'' for the remainder of this paper.
+The kernel function in \eq{celerite-kernel} is qualitatively similar to the
 The Fourier transform\footnote{Here and throughout we have defined the Fourier

### 2017-03-25 paper/ms.tex
words about rotation period uncertainties

@@
 $\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
-function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by the
-parameters $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
+function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by
+$\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
 Under this model, the log-likelihood of observing a dataset
@@
 The physical interpretation of this model isn't immediately obvious and we
-return to a more general discussion of the physical intuition in a moment
-but we start by investigating some special cases.
+return to a more general discussion shortly but we start by considering some
+special cases.
 
@@
 quasiperiodic covariance function can be used to make probabilistic
-measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
+measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
 The covariance function used for this type of analysis has the form
@@
 and find a constraint on the rotation period of $P = \rotationperiod$.
-This is in good agreement with the literature value.
+This is in good agreement with the literature value with somewhat smaller
+uncertainties.
+A detailed comparison of GP rotation period measurements and the traditional
+methods is beyond the scope of this paper but forthcoming work (R.~Angus,
+\etal, submitted) demonstrates that GP inferences are, at a population level,
+more reliable than other methods.
 \Figure{rotation} shows a subset of the data used in this example and the
@@
 
-\newpage
 \section{Comparisons to other methods}
@@
 positive for all frequencies.
-This result is also intuitive because, since the power spectrum of a process
-is defined as expected the squared amplitude of the Fourier transform of the
-time series, it must be non-negative.
+This result is intuitive because, since the power spectrum of a process is
+defined as the expected squared amplitude of the Fourier transform of the time
+series, it must be non-negative.
 

### 2017-03-25 paper/figures/rotation.py

@@ -215,10 +215,10 @@ ax.set_xlabel("rotation period [days]")
 fig.savefig("rotation-period.pdf", bbox_inches="tight", dpi=300)
 
 q = np.percentile(period_samps, [16, 50, 84])
-print(q, np.diff(q))
+print(q, np.diff(q), np.mean(period_samps), np.std(period_samps))
 
 with open("rotation.tex", "w") as f:
     f.write("% Automatically generated\n")
     f.write(("\\newcommand{{\\rotationperiod}}{{\\ensuremath{{{{"
-             "{0:.2f}_{{-{1:.2f}}}^{{+{2:.2f}}} }}}}}}\n")
-            .format(q[1], q[1]-q[0], q[2]-q[1]))
+             "{0:.2f} \pm {1:.2f}}}}}}}\n")
+            .format(np.mean(period_samps), np.std(period_samps)))

### 2017-03-26 paper/ms.tex
addressing some comments from Stello

@@
 the \kepler\ mission.
+Unlike typical giants, KIC~11615890 is a member of a class of stars where the
+dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
+in the core \citep{Stello:2016}.
+This makes this target simpler for the purposes of this demonstration because
+we can neglect the $\ell = 1$ modes and the model proposed above will be an
+effective model for the combined signal from the $\ell = 0$ and 2 modes.
+
 For KIC~11615890, the published asteroseismic parameters measured using several
@@
 
+This research made use of Astropy, a community-developed core Python package
+for Astronomy \citep{Astropy-Collaboration:2013}.
+
 \facility{Kepler}
 \software{%
+     \project{AtroPy} \citep{Astropy-Collaboration:2013},
      \project{corner.py} \citep{Foreman-Mackey:2016},

### 2017-03-27 paper/ms.tex

@@
 time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
-$\sim10^{3-5}$ observations each~---~this scaling is prohibitive.
+$\sim10^{3-5}$ observations each~--~this scaling is prohibitive.
 
@@
 
-GP are stochastic models consisting of a mean function
+GPs are stochastic models consisting of a mean function
 $\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
@@
 term'' for the remainder of this paper.
-The kernel function in \eq{celerite-kernel} is qualitatively similar to the
 The Fourier transform\footnote{Here and throughout we have defined the Fourier
@@
 This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and is the model studied by
+\eq{celerite-kernel-complex} and it is the model studied by
 \citet{Rybicki:1995}.
@@
 
-The extended system defined in the previous section is sparse with typically fewer than
-a few percent non-zero entries and band structure.
+The extended system defined in the previous section is sparse, with typically
+fewer than a few percent non-zero entries and band structure.
 In this section, we empirically investigate the performance and scaling
@@
 The real world performance of \celerite\ depends on the specific platform,
-hardware, and \project{LAPACK}/\project{BLAS} implementation but we have found
-qualitatively similar results across popular platforms and state-of-the-art
-libraries.
+hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
+found qualitatively similar results across popular platforms and
+state-of-the-art libraries.
 
@@
 likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
-measurements in $\mathcal{O}(N)$ operations using the \celerite\ method
+measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
 described in the previous section.
@@
 This means the true autocorrelation of the process can never be correctly
-represented by the model that we're using to fit but we use this example
+represented by the model that we are using to fit, but we use this example
 to demonstrate that, at least in this case, valid inferences can still be made
@@
 
-For this example the data are simulated from a quasiperiodic GP with the
+For this example, the data are simulated from a quasiperiodic GP with the
 kernel
@@
 The inference made using this effective \celeriteterm\ model are indistinguishable
-from the inferences made using the correct model but substantially less
+from the inferences made using the correct model and substantially less
 computation time is required for the \celerite\ inference.
@@
 of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
-other, more fundamental parameters) \emph{conditioned on the observed time
+other more fundamental parameters) \emph{conditioned on the observed time
 series} in $\mathcal{O}(N)$ operations.
@@
 are shared nuisance parameters.
-This model could be easily extended to include small frequency splitting and
+This model could be extended to include more angular modes and
 $\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
@@
 The key difference between \eq{sine2} and other quasiperiodic kernels is that
-it is positive everywhere.
+it is positive for all values of $\tau$.
 We can construct a simple \celeriteterm\ covariance function with similar properties
@@
 A detailed comparison of GP rotation period measurements and the traditional
-methods is beyond the scope of this paper but forthcoming work (R.~Angus,
+methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
 \etal, submitted) demonstrates that GP inferences are, at a population level,
@@
 developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
-In the case where the kernel function has a single term of the form
+In the case where the kernel function has a single term of the form given in
 \eq{kernel-simple}, it is interesting to compare the performance of \celerite\
@@
 The Toeplitz method is, in some ways, more flexible than \celerite\ because it
-can be used with any stationary kernel but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it, in general, is less
-efficient when applied to large datasets.
+can be used with any stationary kernel, but it requires uniformly spaced data
+and the scaling is worse than \celerite\ so it is less efficient when applied
+to large datasets.
 
@@
 \celerite.
-Like \celerite, the likelihood function for a CARMA model conditioned on a
-sorted one-dimensional dataset can be solved in $\mathcal{O}(N)$ using a
-recursive method.
-The kernel function for CARMA$(J,\,K)$ model is \citep{Kelly:2014}
+CARMA models are derived by solving a stochastic differential equation and,
+like \celerite, the likelihood function for a CARMA model conditioned on a
+one-dimensional dataset can be solved in $\mathcal{O}(N)$.
+The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
 \begin{eqnarray}\eqlabel{carma-kernel}
@@
 parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
-The inverse statement is not true however, for any $J > 1$.
-This means \celerite\ could be used to compute any CARMA model but using
-CARMA to compute the likelihood of a \celeriteterm\ model would require solving
-\eq{carma-coeff} for a given set of $\{A_j\}_{j=1}^J$ numerically.
-This is an interesting distinction because, while the computational scaling of
-CARMA models is also $\mathcal{O}(N\,J^2)$, CARMA solvers are, in practice,
-somewhat faster and the memory requirements are smaller.
-However, as discussed in \sect{sho}, \celeriteterm\ models have a physical
-interpretation as a mixture of stochastically-driven damped harmonic
-oscillators.
-This enables, for example, the use of \celeriteterm\ as a model for asteroseismic
-oscillations.
-A similar example using CARMA would necessarily be qualitative.
+The inverse statement is not as simple.
+In practice, this means that \celerite\ could be trivially used to compute any
+CARMA model.
+Using the CARMA solver to compute a \celeriteterm\ model, however, requires
+solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
+While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
+the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
+is, in practice, somewhat faster and the memory requirements are smaller.
+One benefit of the \celeriteterm\ model is that it is parameterized in terms
+of the amplitudes directly.
+This means that the physical interpretation of these parameters is, in some
+cases, simpler.
+On the other hand, CARMA models are parameterized by modeling the power
+spectrum of the stochastic driving force and this might be the quantity of
+interest in other situations.
+Another benefit of the \celeriteterm\ framework is that, unlike
+filtering-based CARMA solvers, it directly solves matrix equations of the form
+${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
+This is a crucial feature when GP modeling is combined with linear models
+\citep[for example][]{Luger:2017}.
 
@@
 software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
-scales approximately linearly with $J$ but with substantial overhead for small
-models when compared to \celerite.
+scales approximately linearly with $J$, but it requires substantial overhead
+for small models.
 This means that the HODLR solver can \emph{approximately} evaluate
@@
 functions.
-The closest cousin is the CARMA method with similar strengths and limitations
-but, since there is a simpler relationship between the parameters of a
-\celeriteterm\ model and its behavior, \celerite\ can be more easily integrated
-with physical or physically interpretable models.
+Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
+stochastically-driven, damped simple harmonic oscillators, they are a
+physically motivated choice of covariance function in many applications.
 
@@
 computing the likelihood and a physical motivation for the choice of kernel.
-\celeriteterm\ can be used to accurately model stellar oscillations using the
-relation to the mixture of stochastically-driven, damped simple harmonic
-oscillators.
+\celerite\ can be used to model stellar oscillations using the relation to the
+mixture of stochastically-driven, damped simple harmonic oscillators.
 As higher signal-to-noise observations of transiting exoplanet systems are

### 2017-03-27 paper/ms.tex
A few minor edits, commas

@@
 $S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
-stellar time series.
+of stellar time series.
 Since this kernel is exactly described by the exponential kernel, the
@@
 Their method does not require solving the extended matrix because the matrix
-inverse can be calculated analytically but the method is made slightly more
+inverse can be calculated analytically, but the method is made slightly more
 complicated when white noise is included.
 We implement their full method and find empirically that it is more
-computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$
+computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
 but we note that the requirement of a single, real, exponential kernel is
@@
 This method has been widely applied in the context of exoplanet transit
-characterization but it requires evenly spaced observations and the power
+characterization, but it requires evenly spaced observations and the power
 spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
@@
 This wavelet method has been demonstrated to improve parameter estimation for
-transiting exoplanets \citep{Carter:2009} but these strict requirements make
+transiting exoplanets \citep{Carter:2009}, but these strict requirements make
 this method applicable for only a limited set of use cases.
@@
 The key takeaway here is that \celerite\ provides an \emph{exact} method for
-GP inference for a specific but flexible class of one-dimensional kernel
+GP inference for a specific, but flexible, class of one-dimensional kernel
 functions.
@@
 Gaussian Process models have been fruitfully applied to many problems in
-astronomical data analysis but the fact that the computational cost scales as
+astronomical data analysis, but the fact that the computational cost scales as
 the cube of the number of data points has limited their use to relatively
@@
 A drawback of this method is its quadratic scaling with the number of
-terms $J$ but, in many cases, small values of $J$ are sufficient.
+terms $J$, but, in many cases, small values of $J$ are sufficient.
 
@@
 This type of analysis will be limited by the quadratic scaling of \celerite\
-with the number of terms $J$ but this limits existing methods as well
+with the number of terms $J$, but this limits existing methods as well
 \citep[CARMA models,][]{Kelly:2014}.

### 2017-03-27 celerite/__init__.py
bumping version number, figures for julia tutorial, publish script

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-__version__ = "0.1.3.dev0"
+__version__ = "0.1.3"
 
 try:
     __CELERITE_SETUP__

### 2017-03-27 paper/ms.tex
bumping version number, figures for julia tutorial, publish script
adding Zenodo DOI to ms

--- 2017-03-27 paper/ms.tex
@@
 These are available online at \project{GitHub}
-\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
-archive}.
+\url{https://github.com/dfm/celerite} and \project{Zenodo}
+\citep{Foreman-Mackey:2017}.
 
@@
 (top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
-    on the full four year baseline of the mission.
+    on the full four year baseline of the mission. The orange line shows a
+    smoothed periodogram and the blue line indicates the level of the
+    measurement uncertainties.
 (middle) The same periodogram computed using about a month of data.
@@
 discussed in these pages.
-This software is available at \url{https://github.com/dfm/celerite}\footnote{%
-This version of the paper was generated with git commit \texttt{\githash}
-(\gitdate).} and it is made available under the MIT license.
+This software is available on GitHub
+\url{https://github.com/dfm/celerite}\footnote{% This version of the paper was
+generated with git commit \texttt{\githash} (\gitdate).} and Zenodo
+\citep{Foreman-Mackey:2017}, and it is made available under the MIT license.
 

### 2017-03-28 paper/ms.tex
page breaks for formatting
re-working text based on comments from M Rawls

@@
 The growing field of large-scale time domain astronomy requires methods for
-probabilistic inference that are computationally tractable, even with large
-datasets.
+probabilistic data analysis that are computationally tractable, even with
+large datasets.
 Gaussian Processes are a popular class of models used for this purpose but,
@@
 their application has been limited to relatively small datasets.
-In this paper, we present a method for Gaussian Process regression in
+In this paper, we present a method for Gaussian Process modeling in
 one-dimension where the computational requirements scale linearly with the
 size of the dataset.
-This method exploits structure in the problem when the covariance function is
+We demonstrate the method by applying it to simulated and real astronomical
+time series datasets.
+These demonstrations are examples of probabilistic inference of stellar
+rotation periods, asteroseismic oscillation spectra, and transiting planet
+parameters.
+The method exploits structure in the problem when the covariance function is
 expressed as a mixture of complex exponentials, without requiring evenly
@@
 implementation, and a comparison to existing scalable Gaussian Process methods.
-We demonstrate the method by applying it to simulated and real astronomical
-time series datasets.
-These demonstrations are examples of probabilistic inference of stellar
-rotation periods, asteroseismic oscillation spectra, and transiting planet
-parameters.
-In all of these cases, we infer posterior distributions over models for the
-power spectrum of the process that generated these irregularly sampled time
-series without computing a standard algebraic estimator.
 The method is flexible, fast, and most importantly, interpretable, with a wide
@@
 \keywords{%
- methods: data analysis
- ---
  methods: data analysis
@@
 
-Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
-models for time-series analysis.
-For GP modeling, a functional form is chosen to describe the autocovariance
-of the data and the parameters of this function are fit for or marginalized.
-In the astrophysical literature, GPs have been used to model stochastic
-variability in light curves of stars \citep{Brewer:2009}, active galactic
-nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
+In the astrophysical literature, Gaussian Processes
+\citep[GPs;][]{Rasmussen:2006} have been used to model stochastic variability
+in light curves of stars \citep{Brewer:2009}, active galactic nuclei
+\citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
 \citep{Uttley:2005}.
@@
 year old paper \citep{Rybicki:1995}.
-For this method to be applicable, the data must be one-dimensional and the
-covariance function must be represented by a mixture of exponentials.
+This method can only be used with one-dimensional datasets and the covariance
+function must be represented by a mixture of exponentials, we will return to a
+discussion of what this means in detail in the following sections.
 However, there is no further constraint on the data or the model.
@@
 
-In the following pages, we motivate the general problem of GP regression,
+Our main expertise lies in the field of exoplanet discovery and
+characterization where GPs have become a model of choice.
+We are confident that the method described in the following pages will benefit
+this field but we also expect that there will be applications in other
+branches of astrophysics and beyond.
+In \sect{examples-real}, we present applications of the method to research
+problems in stellar rotation, asteroseismic analysis, and exoplanet transit
+fitting.
+Some readers might consider starting with that section for some motivating
+examples before delving into the detailed derivations of the earlier sections.
+
+In the following pages, we motivate the general problem of GP modeling,
 describe the previously published scalable method \citep{Rybicki:1995,
@@
 $\mu_\bvec{\theta}(\bvec{x})$ and a covariance, autocorrelation, or ``kernel''
-function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ parameterized by
+function $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ parameterized by
 $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
@@
 given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
+\eq{gp-likelihood} is the equation of an $N$-dimensional Gaussian and it can
+be derived as the generalization of what we astronomers call ``$\chi^2$''
+to include the effects of correlated noise.
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
@@
  \emph{input coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv
-t_n$.\footnote{We use $t$ as the input coordinate because
-one-dimensional GPs are often applied to time series data but this isn't a
-real restriction and the \celerite\ method can be applied to \emph{any}
-one-dimensional dataset.}
+t_n$.
+We use $t$ as the input coordinate because one-dimensional GPs are often
+applied to time series data but this isn't a real restriction and the
+\celerite\ method can be applied to \emph{any} one-dimensional dataset.
 Furthermore, the covariance function for the \celerite\ method is
@@
 
-Similarly, setting both $b_j$ and $d_j$ to zero, we get an Ornstein--Uhlenbeck
-process
+Similarly, setting both $b_j$ and $d_j$ to zero, we get
 \begin{eqnarray}
@@
 \end{eqnarray}
+This model is often referred to as an Ornstein--Uhlenbeck process and this is
+the kernel that was studied by Rybicki \& Press
+\citep{Rybicki:1992,Rybicki:1995}.
 
@@
 Now, if we introduce the variables
+\begin{eqnarray}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
+\end{eqnarray}
+where $g_{0} = 0$, and
 \begin{eqnarray}\eqlabel{algo-first}
@@
 \end{eqnarray}
-where $u_{N} = 0$, and
-\begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
-\end{eqnarray}
-where $g_{0} = 0$,
+where $u_{N} = 0$,
 the system of equations can be rewritten as
@@
 \end{eqnarray}
-Furthermore, our derivation of the extended matrix requires that the data
-points be monotonically sorted in time.
-Neither of these properties will be satisfied in general for multidimensional
-inputs and all of our following discussion assumes a sorted
-one-dimensional dataset.
+Second, our derivation of the extended matrix requires that the data
+points be monotonically sorted.
+This is the motivation for our previous statement that this method can only be
+applied in one dimension since neither of these properties will be satisfied
+by multidimensional inputs.
 
@@
 \eq{full-system-first} through \eq{full-system-last} define a banded matrix
-equation in the ``extended'' space and, as before, this can be used to solve
+equation in the ``extended'' space
+\begin{eqnarray}\eqlabel{kext-full}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
+    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
+\end{eqnarray}
+and, as before, this can be used to solve
 for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
@@
 \caption{A pictorial representation of the sparse extended matrix
-$K_\mathrm{ext}$ with $N=5$ and $J=2$.
+$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
 Each colored block corresponds to a non-zero entry in the matrix as described
@@
 \item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
-    for banded matrices using Gaussian elimination \citep{Press:1992,
-    Press:2007},
+    for banded matrices using Gaussian elimination \citep[we implemented this
+    method following][]{Press:1992, Press:2007},
 
@@
     methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines,\footnote{The experiments below use the Intel Math
-    Kernel Library (\project{MKL}):
+    \project{BLAS} routines,\footnote{The experiments in this paper use the
+    Intel Math Kernel Library (\project{MKL}):
     \url{https://software.intel.com/en-us/intel-mkl}.} and
@@
     points $N$.
-    The different colors show the cost for different numbers of terms $J$ as
-    listed in the legend.
-    To guide the eye, the black line shows linear scaling in $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
     \emph{(right)} The same information plotted as a function of $J$ for
     different values of $N$.
-    The legend shows the value of $N$ for each color and the black line shows
-    quadratic scaling in $J$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
     \figurelabel{benchmark}}
@@
 oscillator provides a physically motivated model that is flexible enough to
-describe a wide range of stellar variations.
+describe a wide range of stellar variations and we return to give some
+specific examples in \sect{examples-real}.
 Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
@@
 
-\section{Examples with real data}
+\section{Examples with real data} \sectlabel{examples-real}
 
@@
 examples to be qualitative in nature and do not claim that \celerite\ is the
-optimum method but we hope these examples encourage interested readers to
+optimum method, but we hope these examples encourage interested readers to
 investigate the applicability of \celerite\ to their research.
@@
 All of the following examples show time domain datasets with a clear bias in
-favor of large homogeneous photometric surveys but these methods can similarly
-be applied to spectroscopy, where wavelength~--~instead of time~--~is the
-independent coordinate and other one-dimensional domains \citep[see][for a
-potential application]{Czekala:2017}.
-
-\subsection{Asteroseismic oscillations}
-
-The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
-Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
-driver for many of the upcoming large scale photometric surveys
-\citep{Campante:2016, Rauer:2014, Gould:2015}.
-Most asteroseismic analyses have been limited to relatively high
-signal-to-noise oscillations because the standard methods based on statistics
-of the empirical periodogram of the data cannot be used to formally propagate
-the measurement uncertainties to the constraints on physical
-parameters~--~instead relying on bootstrapped uncertainty estimators
-\citep{Huber:2009}~--~but more sophisticated methods that compute the
-likelihood function in the time domain scale poorly to state-of-the-art
-datasets \citep{Brewer:2009,Corsaro:2014}.
-
-\celerite\ alleviates these problems by providing a physically motivated
-probabilistic model that can be evaluated efficiently even for large datasets.
-In practice, one would model the star as a mixture of stochastically-driven
-simple harmonic oscillators where the amplitudes and frequencies of the
-oscillations are computed using a physical model and evaluate the probability
-of the observed time series using a GP where the PSD is a sum of terms given
-by \eq{sho-psd}.
-This gives us a method of computing the likelihood function for the parameters
-of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
-other more fundamental parameters) \emph{conditioned on the observed time
-series} in $\mathcal{O}(N)$ operations.
-In other words, \celerite\ provides a computationally efficient framework that
-can be combined with physically-motivated models of stars and numerical
-inference methods to make rigorous probabilistic measurements of asteroseismic
-parameters in the time domain.
-We expect that this has the potential to push asteroseismic analysis to lower
-signal-to-noise datasets and we hope to revisit this idea in a subsequent
-paper.
-
-To demonstrate the method, we use a very simple heuristic model where the PSD
-is given by a mixture of 8 components with amplitudes and frequencies
-specified by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance
-parameters.
-The first term is used to capture the granulation ``background''
-\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
-$S_g$ and $\omega_g$.
-The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
-parameter shared between terms and the frequencies are given by
-\begin{eqnarray}
-\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
-\end{eqnarray}
-and the amplitudes are given by
-\begin{eqnarray}
-S_{0,\,j} =
-    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
-\end{eqnarray}
-where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters.
-This model could be extended to include more angular modes and
-$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
-physical parameters of the star.
-
-To demonstrate the applicability of this model, we apply it to infer the
-asteroseismic parameters of the giant star KIC~11615890, observed by the
-\kepler\ Mission.
-The goal of this example is to show that, even for a low signal-to-noise
-dataset with a short baseline, it is possible to infer asteroseismic
-parameters with formal uncertainties that are consistent with the parameters
-inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
-$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
-compare our results to the results inferred from the full 4 year baseline of
-the \kepler\ mission.
-Unlike typical giants, KIC~11615890 is a member of a class of stars where the
-dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
-in the core \citep{Stello:2016}.
-This makes this target simpler for the purposes of this demonstration because
-we can neglect the $\ell = 1$ modes and the model proposed above will be an
-effective model for the combined signal from the $\ell = 0$ and 2 modes.
-
-For KIC~11615890, the published asteroseismic parameters measured using several
-years of \kepler\ observations are \citep{Pinsonneault:2014}
-\begin{eqnarray}
-    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
-    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
-\end{eqnarray}
-We randomly select a month-long segment of \kepler\ data, initialize our
-\celeriteterm\ model using a grid search in the parameter space, and then use
-\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
-the full set of parameters.
-\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
-and $\Delta\nu$ compared to the results from the literature.
-
-This model requires about a minute of computation time to find the maximum
-likelihood parameters and then it takes about an hour to run the MCMC sampling
-to convergence on a dual-CPU laptop.
-This is more computationally intensive than traditional methods of measuring
-asteroseismic oscillations but is is also orders of magnitude cheaper than the
-same analysis using another GP solver.
-An in-depth discussion of the benefits of rigorous probabilistic inference of
-asteroseismic parameters in the time domain is beyond the scope of this paper
-but we hope to revisit this opportunity in the future.
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
-\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
-    from the inference shown in \Figure{astero} compared to the published
-    value (error bar) based on several years of \kepler\ observations
-    \citep{Pinsonneault:2014}.
-    \figurelabel{astero-corner}}
-\end{center}
-\end{figure}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
-\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
-posterior inference of the PSD as a mixture of stochastically-driven simple
-harmonic oscillators.
-(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
-    on the full four year baseline of the mission. The orange line shows a
-    smoothed periodogram and the blue line indicates the level of the
-    measurement uncertainties.
-(middle) The same periodogram computed using about a month of data.
-(bottom) The power spectrum inferred using the mixture of SHOs model described
-    in the text and only one month of \kepler\ data.
-    The black line shows the median of posterior PSD and the gray contours
-    show the 68\% credible region.
-    \figurelabel{astero}}
-\end{center}
-\end{figure}
+favor of large homogeneous photometric surveys, but these methods can
+similarly be applied to spectroscopy, where wavelength~--~instead of
+time~--~is the independent coordinate and other one-dimensional domains
+\citep[see][for a potential application]{Czekala:2017}.
 
@@
 
-Another source of variability that can be measured from time series
-measurements of stars is rotation.
+A source of variability that can be measured from time series measurements of
+stars is rotation.
 The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
@@
 
-To demonstrate this method, we fit a \celeriteterm\ model with a kernel given by
-\eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
-This star has a published rotation period of $3.88 \pm 0.58$, measured using
-traditional periodogram and autocorrelation function approaches applied to
-\kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
+To demonstrate this method, we fit a \celeriteterm\ model with a kernel given
+by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
+This star has a published rotation period of $3.88 \pm 0.58\,\mathrm{days}$,
+measured using traditional periodogram and autocorrelation function approaches
+applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
 We used \emcee\ to sample the joint posterior probability density for the four
@@
     \kepler\ baseline \citep{Mathur:2014}.
+    The error bars from the literature are indicated by the two vertical
+    dashed lines.
     \figurelabel{rotation-period}}
+\end{center}
+\end{figure}
+
+\subsection{Asteroseismic oscillations}
+
+The asteroseismic oscillations of thousands of stars were measured using light
+curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
+Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
+driver for many of the upcoming large scale photometric surveys
+\citep{Campante:2016, Rauer:2014, Gould:2015}.
+Most asteroseismic analyses have been limited to relatively high
+signal-to-noise oscillations.
+The standard methods use statistics of the empirical periodogram and cannot
+formally propagate the measurement uncertainties to the constraints on
+physical parameters, instead relying on bootstrapped uncertainty estimators
+\citep{Huber:2009}.
+More sophisticated methods that compute the likelihood function in the time
+domain scale poorly to state-of-the-art datasets \citep{Brewer:2009,
+Corsaro:2014}.
+
+\celerite\ alleviates these problems by providing a physically motivated
+probabilistic model that can be evaluated efficiently even for large datasets.
+In practice, one would model the star as a mixture of stochastically-driven
+simple harmonic oscillators where the amplitudes and frequencies of the
+oscillations are computed using a physical model, and evaluate the probability
+of the observed time series using a GP where the PSD is a sum of terms given
+by \eq{sho-psd}.
+This gives us a method of computing the likelihood function for the parameters
+of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
+other more fundamental parameters) \emph{conditioned on the observed time
+series} in $\mathcal{O}(N)$ operations.
+In other words, \celerite\ provides a computationally efficient framework that
+can be combined with physically-motivated models of stars and numerical
+inference methods to make rigorous probabilistic measurements of asteroseismic
+parameters in the time domain.
+We expect that this has the potential to push asteroseismic analysis to lower
+signal-to-noise datasets and we hope to revisit this idea in a subsequent
+paper.
+
+To demonstrate the method, we use a simple heuristic model where the PSD is
+given by a mixture of 8 components with amplitudes and frequencies specified
+by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
+The first term is used to capture the granulation ``background''
+\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
+$S_g$ and $\omega_g$.
+The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
+parameter shared between terms and the frequencies are given by
+\begin{eqnarray}
+\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
+\end{eqnarray}
+and the amplitudes are given by
+\begin{eqnarray}
+S_{0,\,j} =
+    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
+\end{eqnarray}
+where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
+are shared nuisance parameters.
+This model could be extended to include more angular modes and
+$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
+physical parameters of the star.
+
+To demonstrate the applicability of this model, we apply it to infer the
+asteroseismic parameters of the giant star KIC~11615890, observed by the
+\kepler\ Mission.
+The goal of this example is to show that, even for a low signal-to-noise
+dataset with a short baseline, it is possible to infer asteroseismic
+parameters with formal uncertainties that are consistent with the parameters
+inferred with a much larger dataset.
+Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
+$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
+compare our results to the results inferred from the full 4 year baseline of
+the \kepler\ mission.
+Unlike typical giants, KIC~11615890 is a member of a class of stars where the
+dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
+in the core \citep{Stello:2016}.
+This makes this target simpler for the purposes of this demonstration because
+we can neglect the $\ell = 1$ modes and the model proposed above will be an
+effective model for the combined signal from the $\ell = 0$ and 2 modes.
+
+For KIC~11615890, the published asteroseismic parameters measured using several
+years of \kepler\ observations are \citep{Pinsonneault:2014}
+\begin{eqnarray}
+    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
+    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
+\end{eqnarray}
+We randomly select a month-long segment of PDC \citep{Stumpe:2012, Smith:2012}
+\kepler\ data, initialize our \celeriteterm\ model using a grid search in the
+parameter space, and then use \emcee\ \citep{Foreman-Mackey:2013} to sample
+the joint posterior density for the full set of parameters.
+\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
+and $\Delta\nu$ compared to the results from the literature.
+This result is consistent within the published error bars and the posterior
+constraints are tighter than the published results.
+All asteroseismic analyses are known to have substantial systematic and
+method-dependent uncertainties \citep{Verner:2011} so further experiments
+would be needed to assess the reliability of this method.
+
+This model requires about a minute of computation time to find the maximum
+likelihood parameters and then it takes about an hour to run the MCMC sampling
+to convergence on a dual-CPU laptop.
+This is more computationally intensive than traditional methods of measuring
+asteroseismic oscillations, but is is also orders of magnitude cheaper than the
+same analysis using another GP solver.
+An in-depth discussion of the benefits of rigorous probabilistic inference of
+asteroseismic parameters in the time domain is beyond the scope of this paper,
+but we hope to revisit this opportunity in the future.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
+\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
+    from the inference shown in \Figure{astero} compared to the published
+    value (error bar) based on several years of \kepler\ observations
+    \citep{Pinsonneault:2014}.
+    \figurelabel{astero-corner}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
+\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
+posterior inference of the PSD as a mixture of stochastically-driven simple
+harmonic oscillators.
+(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
+    on the full four year baseline of the mission. The orange line shows a
+    smoothed periodogram and the blue line indicates the level of the
+    measurement uncertainties.
+(middle) The same periodogram computed using about a month of data.
+(bottom) The power spectrum inferred using the mixture of SHOs model described
+    in the text and only one month of \kepler\ data.
+    The black line shows the median of posterior PSD and the gray contours
+    show the 68\% credible region.
+    \figurelabel{astero}}
 \end{center}
@@
 using \celerite.
-This example is different from all the previous examples because in this case,
-we are uninterested in the inferred parameters of the covariance model.
+This example is different from all the previous examples~--~both simulated and
+real~--~because in this case, we are uninterested in the inferred parameters
+of the covariance model.
 Instead, we're interested in inferring constraints on the parameters of the
-mean model.
-In \eq{gp-likelihood} these parameters are called $\bvec{\theta}$ and in this
-example, the mean function $\mu_\bvec{\theta}(t)$ is a limb-darkened transit
-light curve \citep{Mandel:2002} parameterized by a period $P$, a transit
-duration $T$, a phase or epoch $t_0$, an impact parameter $b$, the radius
-of the planet in units of the stellar radius $R_P/R_\star$, and several
-parameters describing the limb-darkening profile of the star \citep{Claret:2011}.
+mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}).
+When we say ``mean model'' here, we're referring to what is normally just
+called the ``model''.
+In this case, we have a light curve of a rotating star with a planet
+transiting and we would like to fit for the physical parameters of the planet
+while taking our uncertainty about the star's variability into account.
+GP modeling has been used for this purpose throughout the exoplanet literature
+\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
+Foreman-Mackey:2016b, Grunblatt:2016}.
+
+In \eq{gp-likelihood} the physical parameters of the planet are called
+$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
+is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
+$\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
+$t_0$, an impact parameter $b$, the radius of the planet in units of the
+stellar radius $R_P/R_\star$, and several parameters describing the
+limb-darkening profile of the star \citep{Claret:2011}.
 As in the previous example, we model the stellar variability using a GP model
-with a kernel given by \eq{rot-kernel}.
+with a kernel given by \eq{rot-kernel} and fit for the parameters of the
+planet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
+simultaneously.
 
@@
 simulated transit model with known parameters.
+The fact that we are using a simulated transit enables us to confirm that we
+correctly measure the planet parameters but we do not use this knowledge to
+inform our fit.
 The top panel of \Figure{transit-ml} shows the data including the simulated
@@
 Dan Huber,
+Meredith Rawls,
 Dennis Stello,

### 2017-03-28 paper/figures/rotation.py

@@ -212,6 +212,8 @@ ax.yaxis.set_major_locator(plt.NullLocator())
 mu, std = np.mean(period_samps), np.std(period_samps)
 ax.set_xlim(mu - 3.5*std, mu + 3.5*std)
 ax.set_xlabel("rotation period [days]")
+ax.axvline(3.80 + 0.15, color="k", ls="dashed")
+ax.axvline(3.80 - 0.15, color="k", ls="dashed")
 fig.savefig("rotation-period.pdf", bbox_inches="tight", dpi=300)
 
 q = np.percentile(period_samps, [16, 50, 84])

### 2017-03-28 paper/ms.tex
typos and formatting

@@
 to astronomical time series
-\vspace{-3.2\baselineskip}
+\vspace{-2\baselineskip}
 }
@@
 \end{eqnarray}
-and the PSD for the this component is
+and the PSD for this component is
 \begin{eqnarray}\eqlabel{lorentz-psd}
@@
 
+\newpage
 \subsection{Inferences with the ``wrong'' model}
@@
 
-\subsection{Stellar rotation}
+\subsection{Stellar rotation} \sectlabel{rotation}
 
@@
 This example is different from all the previous examples~--~both simulated and
-real~--~because in this case, we are uninterested in the inferred parameters
-of the covariance model.
+real~--~because, in this case, we are uninterested in the parameters of the
+covariance model.
 Instead, we're interested in inferring constraints on the parameters of the
-mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}).
-When we say ``mean model'' here, we're referring to what is normally just
-called the ``model''.
+mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
+signal caused by the transit of the planet.
+We say ``mean model'' here to unambiguously distinguish it from the
+``covariance model'', but this is what would generally be called the ``model''
+in other analyses.
 In this case, we have a light curve of a rotating star with a planet
@@
 We take a month-long segment of the \kepler\ light curve for
-KIC~1430163~--~the target from the previous example~--~and multiply it by a
+KIC~1430163~--~the target from \sect{rotation}~--~and multiply it by a
 simulated transit model with known parameters.
@@
 This software is available on GitHub
-\url{https://github.com/dfm/celerite}\footnote{% This version of the paper was
+\url{https://github.com/dfm/celerite}\footnote{This version of the paper was
 generated with git commit \texttt{\githash} (\gitdate).} and Zenodo
@@
 by the NASA Exoplanet Science Institute.
-
 EA acknowledges support from NASA grants NNX13AF20G, NNX13A124G, NNX13AF62G,
@@
 We are grateful to the entire \kepler\ team, past and present.
-
 These data were obtained from the Mikulski Archive for Space Telescopes
@@
 
+\newpage
 \appendix
@@
 
+\newpage
 \bibliography{celerite}

### 2017-03-29 .ci/build-paper.sh
skeleton of second paper

@@ -13,7 +13,8 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan $TRAVIS_BRANCH-pdf
   git rm -rf .
-  git add -f paper/ms.pdf
-  git -c user.name='travis' -c user.email='travis' commit -m "making a paper"
+  git add -f paper/paper1/ms.pdf
+  git add -f paper/paper2/celerite2.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "building the papers"
   git push -q -f https://$GITHUB_USER:$GITHUB_API_KEY@github.com/$TRAVIS_REPO_SLUG $TRAVIS_BRANCH-pdf
 fi

### 2017-03-29 .ci/travis.sh
another paper change

@@ -3,7 +3,7 @@
 # If building the paper, do that here
 if [[ $TEST_LANG == paper ]]
 then
-  if git diff --name-only $TRAVIS_COMMIT_RANGE | grep 'paper/'
+  if git diff --name-only $TRAVIS_COMMIT_RANGE | grep 'papers/'
   then
     echo "Building the paper..."
     export CELERITE_BUILDING_PAPER=true

### 2017-04-28 papers/paper1/ms.tex
adding OU process ref

--- 2017-03-28 paper/ms.tex
@@
 \newcommand{\racomment}[1]{{\color{magenta}#1}}
+
+\newcommand{\response}[1]{{\color{green}#1}}
 
@@
 \end{eqnarray}
-This model is often referred to as an Ornstein--Uhlenbeck process and this is
-the kernel that was studied by Rybicki \& Press
-\citep{Rybicki:1992,Rybicki:1995}.
+This model is often called an Ornstein--Uhlenbeck process \response{\citep[in
+reference to the classic paper,][]{Uhlenbeck:1930}} and this is the kernel
+that was studied by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
 
@@
 \software{%
-     \project{AtroPy} \citep{Astropy-Collaboration:2013},
+     \project{AstroPy} \citep{Astropy-Collaboration:2013},
      \project{corner.py} \citep{Foreman-Mackey:2016},

### 2017-05-02 papers/paper1/ms.tex
some words about Cholesky factorization

@@
 
-\section{Implementation \& performance}
-
-\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
-elements given by \eq{kernel-simple}, could be computed
-efficiently by taking advantage of the structure of this covariance function.
-\citet{Ambikasaran:2015} generalized this computation to apply to the full
-mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
-efficient method for computing the determinant of $K$.
-
-\subsection{An example}
-
-To provide some insight for this method, we follow
-\citet{Ambikasaran:2015} and start by working through a simple example.
-In this case, we assume that we have three data points
-$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
-variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
-like to compute the likelihood of these data under a GP model with the
-covariance function
-\begin{eqnarray}
-k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
-\end{eqnarray}
-This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and it is the model studied by
-\citet{Rybicki:1995}.
-To demonstrate that the likelihood of this model can be computed in
-$\mathcal{O}(N)$, we write out the full system of
-equations that must be solved to apply the inverse of $K$ and compute
-the first term of \eq{gp-likelihood}.
-In matrix notation, this is
-\begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{z} &=& \bvec{y}, \\
-\begin{pmatrix}
-    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
-    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
-    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ z_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ y_2 \\ y_3
-\end{pmatrix}
-\end{eqnarray}
-where our goal is to solve for the unknown vector \bvec{z} for a given matrix
-$K$ and vector \bvec{y}.
-In \eq{impl-matrix}, we have assumed that the mean function is zero but a
-non-zero mean could be included by replacing \bvec{y} by
-$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
-Now, if we introduce the variables
-\begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
-\end{eqnarray}
-where $g_{0} = 0$, and
-\begin{eqnarray}\eqlabel{algo-first}
-    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
-\end{eqnarray}
-where $u_{N} = 0$,
-the system of equations can be rewritten as
-\begin{eqnarray}
-(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
-a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
-\end{eqnarray}
-Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
-as a matrix equation shows the benefit that this seemingly trivial
-reformulation provides:
-\begin{eqnarray}
-\begin{pmatrix}
-    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
-    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
-    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
-    0 & 0 & 0 & 0 & -1 & 0 & a \\
-    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
-\end{pmatrix}\nonumber
-\end{eqnarray}
-Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
-rewrite \eq{impl-matrix} as
-\begin{eqnarray}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
-\end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
-with, it is now sparse with banded structure that can be exploited to solve
-the system efficiently.
-In particular, sparse solvers are available that can perform a
-LU-decomposition of matrices like this in $\mathcal{O}(N)$
-operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
-general~--~and we can use these algorithms to solve our system
-exactly because the target vector $\bvec{z}$ is a subset of the elements of
-$\bvec{z}_\mathrm{ext}$.
-
-In the following section we discuss this method more generally, but it's
-worth noting a few important facts that can already be seen in this example.
-First, the fundamental reason why this matrix $K$ can be solved efficiently is
-the following property of exponentials
-\begin{eqnarray}
-    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
-    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
-    e^{-c\,(t_3 - t_1)}
-\end{eqnarray}
-and it is important to note that this property does not extend to other common
-covariance functions like the ``exponential-squared'' function
-\begin{eqnarray}
-    k(\tau) \propto e^{-c\,\tau^2} \quad.
-\end{eqnarray}
-Second, our derivation of the extended matrix requires that the data
-points be monotonically sorted.
-This is the motivation for our previous statement that this method can only be
-applied in one dimension since neither of these properties will be satisfied
-by multidimensional inputs.
-
-\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
-extended matrix formalism in practice.
-First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with a bandwidth that scales linearly
-with the number of components, $J$.
-Second, \citet{Ambikasaran:2015} proved that the absolute value of the
-determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
-determinant of $K$.
-This means we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(N)$ operations.
-
-\subsection{The algorithm}
-
-In this section, we generalize the method from the previous section to the
-covariance function given by \eq{celerite-kernel}.
-This derivation follows \citet{Ambikasaran:2015} but it includes explicit
-treatment of complex parameters, and their complex conjugates.
-
-In the case of the full \celeriteterm\ covariance function
-(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
-analogy to the $u_n$ and $g_n$ that we introduced in the previous section
-\begin{eqnarray}\eqlabel{full-system-first}
-\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
-\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
-g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
-h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
-u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
-v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
-\end{eqnarray}
-with the boundary conditions
-\begin{eqnarray}
-    g_{0,j} = 0 \quad, \quad
-    h_{0,j} = 0 \quad, \quad
-    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
-    v_{N,j} = 0
-\end{eqnarray}
-for all $j$.
-Using these variables and some algebra, we find that the following expression
-\begin{eqnarray}\eqlabel{full-system-last}
-\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
-+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
-+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
-    &=& r_{\bvec{\theta},n}
-\end{eqnarray}
-is equivalent to the target matrix equation
-\begin{eqnarray}
-K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
-\end{eqnarray}
-if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
-$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
-\eq{full-system-first} through \eq{full-system-last} define a banded matrix
-equation in the ``extended'' space
-\begin{eqnarray}\eqlabel{kext-full}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
-    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
-\end{eqnarray}
-and, as before, this can be used to solve
-for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
-operations.
-\Figure{matrix} shows a pictorial representation of the sparsity pattern of
-the extended matrix $K_\mathrm{ext}$.
-The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
-with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
-non-zero entries.
-Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
-$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
-as
-\begin{eqnarray}
-{\bvec{z}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
-    h_{N-1,j} & z_N
-\end{array}\right)
-\end{eqnarray}
-and
-\begin{eqnarray}
-{\bvec{r}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},N}
-\end{array}\right) \quad.
-\end{eqnarray}
-After constructing the extended matrix (using a compact storage format), the
-extended matrix can be factorized using a LU-decomposition\footnote{Even
-though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
-Cholesky solver cannot be used for increased efficiency.} routine optimized
-for band or sparse matrices.
-This decomposition can then be used to compute the determinant of $K$, solve
-$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
-likelihood in \eq{gp-likelihood}.
-
-In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
-case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
-will also be identically zero for all $n$.
-Removing these trivial rows from the extended matrix results in factor of two
-increase in the computational efficiency for these terms.
-In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
-$d_j=0$, and the general terms are called ``complex''.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
-\caption{A pictorial representation of the sparse extended matrix
-$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
-Each colored block corresponds to a non-zero entry in the matrix as described
-in the legend.
-    \figurelabel{matrix}}
-\end{center}
-\end{figure}
-
-\subsection{Implementation considerations \& scaling}
-
-The extended system defined in the previous section is sparse, with typically
-fewer than a few percent non-zero entries and band structure.
-In this section, we empirically investigate the performance and scaling
-of three different algorithms for solving this extended system:
-
-\begin{enumerate}
-
-\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
-    for banded matrices using Gaussian elimination \citep[we implemented this
-    method following][]{Press:1992, Press:2007},
-
-\item \texttt{lapack}: The general banded LU decomposition implementation from
-    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
-    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines,\footnote{The experiments in this paper use the
-    Intel Math Kernel Library (\project{MKL}):
-    \url{https://software.intel.com/en-us/intel-mkl}.} and
-
-\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
-    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
-    sparsity but not the band structure.
-
-\end{enumerate}
-
-The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
-because the dimension of the extended matrix scales as $N\,J$ and the
-bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
-\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
-\project{SuperLU} package \citep{Demmel:1999}.
-We find that, while the \texttt{vanilla} solver scales as expected, the
-\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
-offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
-
-The benchmark experiments shown here were performed on a MacBook Pro with two
-2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
-CPUs and running Ubuntu.
-\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
-with $N$ and $J$ using the \texttt{vanilla} solver.
-As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
-in $J$ for large $J$.
-\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
-the \texttt{lapack} and \texttt{sparse} solvers respectively.
-For each of these optimized solvers, the empirical scaling is
-$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
-Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
-efficient than the other algorithms.
-The real world performance of \celerite\ depends on the specific platform,
-hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
-found qualitatively similar results across popular platforms and
-state-of-the-art libraries.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
-\caption{A benchmark showing the computational scaling of \celerite\ using the
-    \texttt{vanilla} band solver as a function of the number of data points
-    and the number of terms.
-    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
-    matrix given by \eq{celerite-kernel} as a function of the number of data
-    points $N$.
-    The different lines show the cost for different numbers of terms $J$
-    increasing from bottom to top.
-    To guide the eye, the straight black line without points
-    shows linear scaling in $N$.
-    \emph{(right)} The same information plotted as a function of $J$ for
-    different values of $N$.
-    Each line shows the scaling for a specific value of $N$ increasing from
-    bottom to top.
-    The black line shows quadratic scaling in $J$.
-    \figurelabel{benchmark}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_lapack.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
-    with the \project{MKL} \project{BLAS} library.
-    \figurelabel{benchmark-lapack}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_sparse.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver
-    from \project{Eigen} \citep{Guennebaud:2010}.
-    \figurelabel{benchmark-sparse}}
-\end{center}
-\end{figure}
+\section{Semi-separable matrix factorization}
+
+\subsection{Cholesky factorization}
+
+The relationship between the \celeriteterm\ model and semi-separable linear
+algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
+A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
+written as
+\begin{eqnarray}
+K_{n,m} = \left\{\begin{array}{ll}
+    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
+    A_{n,n} & \mathrm{if}\,m=n \\
+    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
+\end{array}\right.
+\end{eqnarray}
+for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
+N$-dimensional diagonal matrix $A$.
+This definition can be equivalently written as
+\begin{eqnarray}\eqlabel{semi-sep}
+K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+\end{eqnarray}
+where the $\mathrm{tril}$ function extracts the strict lower triangular
+component of its argument and $\mathrm{triu}$ is the equivalent strict upper
+triangular operation.
+For the \celeriteterm\ model described in the previous section, the rank of
+the semi-separable matrix is $R = 2\,J$ and the components of the relevant
+matrices are
+\begin{eqnarray}
+U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
+    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
+U_{n,2\,j+1} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
+    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
+V_{m,2\,j+1} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
+\end{eqnarray}
+
+To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
+compute the log-determinant of $K$ and solve linear systems of the form
+$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
+semi-separable matrix.
+If $K$ is positive semi-definite, these can be computed using the Cholesky
+factorization of $K$:
+\begin{eqnarray}
+K = L\,D\,L^\T
+\end{eqnarray}
+where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
+diagonal matrix.
+We use the ansatz that $L$ has the form
+\begin{eqnarray}
+    L = I + \mathrm{tril} (U\,X^T)
+\end{eqnarray}
+where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
+$N \times 2\,J$ matrix.
+Combining this with \eq{semi-sep}, we find the equation
+\begin{eqnarray}\eqlabel{cholesky}
+A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,X^\T)\,D
+    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $X$.
+Setting each element on the left side of \eq{cholesky} equal to the
+corresponding element on the right side, we derive the following recursion
+relations
+\begin{eqnarray}
+S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
+    \nonumber\\
+X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
+    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
+    \eqlabel{algorithm}
+\end{eqnarray}
+where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
+$S_{0,j,k}=0$.
+The computational cost of one iteration of the update in \eq{algorithm} is
+$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
+observation~--~so the total cost of this factorization scales as
+$\mathcal{O}(N\,J^2)$.
+
+\subsection{Solving}
+
+After computing the Cholesky factorization of the matrix $K$, we can apply
+the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
+$\mathcal{O}(N)$ operations respectively.
+The log-determinant of $K$ is given by
+\begin{eqnarray}
+    \log \det K = \sum_{n=1}^N \log D_{n,n}
+\end{eqnarray}
+where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
+
+The inverse of $K$ can be applied using the relationship
+\begin{eqnarray}
+\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
+substitution algorithm
+\begin{eqnarray}
+    f_{n,j} &=& f_{n-1,j} + X_{n-1,j}\,z_{n-1} \nonumber\\
+    z_n &=& y_n - \sum_{j=1}^{2\,J} U_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{0,j} = 0$ for all $j$.
+Using this result and the following backward substitution algorithm, we can
+compute $\bvec{b}$
+\begin{eqnarray}
+    f_{n,j} &=& f_{n+1,j} + U_{n+1,j}\,b_{n+1} \nonumber\\
+    b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} X_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{N+1,j} = 0$ for all $j$.
+The total cost of the forward and backward substitution passes scales as
+$\mathcal{O}(N\,J)$.
+
+\subsection{Pre-conditioning}
 
@@
 
+\section{An alternative solver using the extended matrix formalism}
+
+\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
+elements given by \eq{kernel-simple}, could be computed
+efficiently by taking advantage of the structure of this covariance function.
+\citet{Ambikasaran:2015} generalized this computation to apply to the full
+mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
+efficient method for computing the determinant of $K$.
+
+\subsection{An example}
+
+To provide some insight for this method, we follow
+\citet{Ambikasaran:2015} and start by working through a simple example.
+In this case, we assume that we have three data points
+$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
+variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
+like to compute the likelihood of these data under a GP model with the
+covariance function
+\begin{eqnarray}
+k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
+\end{eqnarray}
+This is the result of setting $J=1$, $b=0$, and $d=0$ in
+\eq{celerite-kernel-complex} and it is the model studied by
+\citet{Rybicki:1995}.
+To demonstrate that the likelihood of this model can be computed in
+$\mathcal{O}(N)$, we write out the full system of
+equations that must be solved to apply the inverse of $K$ and compute
+the first term of \eq{gp-likelihood}.
+In matrix notation, this is
+\begin{eqnarray}\eqlabel{impl-matrix}
+K\,\bvec{z} &=& \bvec{y}, \\
+\begin{pmatrix}
+    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
+    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
+    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ z_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ y_2 \\ y_3
+\end{pmatrix}
+\end{eqnarray}
+where our goal is to solve for the unknown vector \bvec{z} for a given matrix
+$K$ and vector \bvec{y}.
+In \eq{impl-matrix}, we have assumed that the mean function is zero but a
+non-zero mean could be included by replacing \bvec{y} by
+$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
+Now, if we introduce the variables
+\begin{eqnarray}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
+\end{eqnarray}
+where $g_{0} = 0$, and
+\begin{eqnarray}\eqlabel{algo-first}
+    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
+\end{eqnarray}
+where $u_{N} = 0$,
+the system of equations can be rewritten as
+\begin{eqnarray}
+(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
+a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
+\end{eqnarray}
+Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
+as a matrix equation shows the benefit that this seemingly trivial
+reformulation provides:
+\begin{eqnarray}
+\begin{pmatrix}
+    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
+    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
+    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
+    0 & 0 & 0 & 0 & -1 & 0 & a \\
+    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
+\end{pmatrix}\nonumber
+\end{eqnarray}
+Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
+rewrite \eq{impl-matrix} as
+\begin{eqnarray}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
+\end{eqnarray}
+Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
+with, it is now sparse with banded structure that can be exploited to solve
+the system efficiently.
+In particular, sparse solvers are available that can perform a
+LU-decomposition of matrices like this in $\mathcal{O}(N)$
+operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
+general~--~and we can use these algorithms to solve our system
+exactly because the target vector $\bvec{z}$ is a subset of the elements of
+$\bvec{z}_\mathrm{ext}$.
+
+In the following section we discuss this method more generally, but it's
+worth noting a few important facts that can already be seen in this example.
+First, the fundamental reason why this matrix $K$ can be solved efficiently is
+the following property of exponentials
+\begin{eqnarray}
+    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
+    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
+    e^{-c\,(t_3 - t_1)}
+\end{eqnarray}
+and it is important to note that this property does not extend to other common
+covariance functions like the ``exponential-squared'' function
+\begin{eqnarray}
+    k(\tau) \propto e^{-c\,\tau^2} \quad.
+\end{eqnarray}
+Second, our derivation of the extended matrix requires that the data
+points be monotonically sorted.
+This is the motivation for our previous statement that this method can only be
+applied in one dimension since neither of these properties will be satisfied
+by multidimensional inputs.
+
+\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
+extended matrix formalism in practice.
+First, even if the covariance function is a mixture of exponentials, the
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components, $J$.
+Second, \citet{Ambikasaran:2015} proved that the absolute value of the
+determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
+determinant of $K$.
+This means we can use this extended matrix formalism to compute the
+marginalized likelihood in $\mathcal{O}(N)$ operations.
+
+\subsection{The algorithm}
+
+In this section, we generalize the method from the previous section to the
+covariance function given by \eq{celerite-kernel}.
+This derivation follows \citet{Ambikasaran:2015} but it includes explicit
+treatment of complex parameters, and their complex conjugates.
+
+In the case of the full \celeriteterm\ covariance function
+(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
+analogy to the $u_n$ and $g_n$ that we introduced in the previous section
+\begin{eqnarray}\eqlabel{full-system-first}
+\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
+\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
+\end{eqnarray}
+with the boundary conditions
+\begin{eqnarray}
+    g_{0,j} = 0 \quad, \quad
+    h_{0,j} = 0 \quad, \quad
+    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
+    v_{N,j} = 0
+\end{eqnarray}
+for all $j$.
+Using these variables and some algebra, we find that the following expression
+\begin{eqnarray}\eqlabel{full-system-last}
+\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
++ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
++ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
+    &=& r_{\bvec{\theta},n}
+\end{eqnarray}
+is equivalent to the target matrix equation
+\begin{eqnarray}
+K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
+\end{eqnarray}
+if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
+$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
+\eq{full-system-first} through \eq{full-system-last} define a banded matrix
+equation in the ``extended'' space
+\begin{eqnarray}\eqlabel{kext-full}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
+    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
+\end{eqnarray}
+and, as before, this can be used to solve
+for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
+operations.
+\Figure{matrix} shows a pictorial representation of the sparsity pattern of
+the extended matrix $K_\mathrm{ext}$.
+The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
+with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
+non-zero entries.
+Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
+$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
+as
+\begin{eqnarray}
+{\bvec{z}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
+    h_{N-1,j} & z_N
+\end{array}\right)
+\end{eqnarray}
+and
+\begin{eqnarray}
+{\bvec{r}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
+    0 & \cdots & 0 & r_{\bvec{\theta},N}
+\end{array}\right) \quad.
+\end{eqnarray}
+After constructing the extended matrix (using a compact storage format), the
+extended matrix can be factorized using a LU-decomposition\footnote{Even
+though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
+Cholesky solver cannot be used for increased efficiency.} routine optimized
+for band or sparse matrices.
+This decomposition can then be used to compute the determinant of $K$, solve
+$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
+likelihood in \eq{gp-likelihood}.
+
+In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
+case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
+will also be identically zero for all $n$.
+Removing these trivial rows from the extended matrix results in factor of two
+increase in the computational efficiency for these terms.
+In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
+$d_j=0$, and the general terms are called ``complex''.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
+\caption{A pictorial representation of the sparse extended matrix
+$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
+Each colored block corresponds to a non-zero entry in the matrix as described
+in the legend.
+    \figurelabel{matrix}}
+\end{center}
+\end{figure}
+
+\subsection{Implementation considerations \& scaling}
+
+The extended system defined in the previous section is sparse, with typically
+fewer than a few percent non-zero entries and band structure.
+In this section, we empirically investigate the performance and scaling
+of three different algorithms for solving this extended system:
+
+\begin{enumerate}
+
+\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
+    for banded matrices using Gaussian elimination \citep[we implemented this
+    method following][]{Press:1992, Press:2007},
+
+\item \texttt{lapack}: The general banded LU decomposition implementation from
+    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
+    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
+    \project{BLAS} routines,\footnote{The experiments in this paper use the
+    Intel Math Kernel Library (\project{MKL}):
+    \url{https://software.intel.com/en-us/intel-mkl}.} and
+
+\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
+    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
+    sparsity but not the band structure.
+
+\end{enumerate}
+
+The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
+because the dimension of the extended matrix scales as $N\,J$ and the
+bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
+\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
+\project{SuperLU} package \citep{Demmel:1999}.
+We find that, while the \texttt{vanilla} solver scales as expected, the
+\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
+offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
+
+The benchmark experiments shown here were performed on a MacBook Pro with two
+2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
+CPUs and running Ubuntu.
+\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
+with $N$ and $J$ using the \texttt{vanilla} solver.
+As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
+in $J$ for large $J$.
+\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
+the \texttt{lapack} and \texttt{sparse} solvers respectively.
+For each of these optimized solvers, the empirical scaling is
+$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
+Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
+efficient than the other algorithms.
+The real world performance of \celerite\ depends on the specific platform,
+hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
+found qualitatively similar results across popular platforms and
+state-of-the-art libraries.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
+\caption{A benchmark showing the computational scaling of \celerite\ using the
+    \texttt{vanilla} band solver as a function of the number of data points
+    and the number of terms.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
+    \figurelabel{benchmark}}
+\end{center}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin_lapack.pdf}
+\caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
+    with the \project{MKL} \project{BLAS} library.
+    \figurelabel{benchmark-lapack}}
+\end{center}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin_sparse.pdf}
+\caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver
+    from \project{Eigen} \citep{Guennebaud:2010}.
+    \figurelabel{benchmark-sparse}}
+\end{center}
+\end{figure}
+
 \section{Ensuring positive definiteness}

### 2017-05-08 papers/paper1/ms.tex
Corrected indexing in equations (21)-(24)

@@
 \begin{eqnarray}
-U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
+U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
     b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
-U_{n,2\,j+1} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
+U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
     b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
-V_{m,2\,j} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
-V_{m,2\,j+1} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
 A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.

### 2017-05-10 papers/paper1/ms.tex
adding some discussion of pre-conditioned algorithm

@@
 where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
-$S_{0,j,k}=0$.
+$S_{1,j,k}=0$.
 The computational cost of one iteration of the update in \eq{algorithm} is
@@
 
+\subsection{Pre-conditioning}
+
+Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
+overflow and underflow issues in many realistic cases because both $U$ and $V$
+have factors of the form $e^{\pm t}$, where $t$ can be any arbitrarily large
+real number.
+This can be avoided with the appropriate reparameterization.
+We define the following pre-conditioned variables
+\begin{eqnarray}
+\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
+\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
+\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
+\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
+\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
+\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
+\end{eqnarray}
+and the set of variables
+\begin{eqnarray}
+\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
+\end{eqnarray}
+with the constraint
+\begin{eqnarray}
+\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
+\end{eqnarray}
+Using this parameterization, we find the following, numerically stable,
+algorithm
+\begin{eqnarray}
+    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
+    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
+    \nonumber\\
+\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
+    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
+\end{eqnarray}
+
 \subsection{Solving}
@@
 \begin{eqnarray}
-    f_{n,j} &=& f_{n-1,j} + X_{n-1,j}\,z_{n-1} \nonumber\\
-    z_n &=& y_n - \sum_{j=1}^{2\,J} U_{n,j}\,f_{n,j}
+    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
+    \tilde{X}_{n-1,j}\,z_{n-1}\right] \nonumber\\
+    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
 \end{eqnarray}
@@
 \begin{eqnarray}
-    f_{n,j} &=& f_{n+1,j} + U_{n+1,j}\,b_{n+1} \nonumber\\
-    b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} X_{n,j}\,f_{n,j}
-\end{eqnarray}
-where $f_{N+1,j} = 0$ for all $j$.
+g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
+    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
+b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{X}_{n,j}\,g_{n,j}
+\end{eqnarray}
+where $g_{N+1,j} = 0$ for all $j$.
 The total cost of the forward and backward substitution passes scales as
@@
 
-\subsection{Pre-conditioning}
+\subsection{Simulation}
 

### 2017-05-11 papers/paper1/ms.tex

@@
 dimensional data sets.
-This method is a generalization of a method developed by
-\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
+\response{This method is a generalization of a method developed by
+\citet{Ambikasaran:2015} 
+ ({\bf This is no longer correct -different Ambikasaran result!})}
+that was, in turn, built on intuition from a twenty
 year old paper \citep{Rybicki:1995}.
 This method can only be used with one-dimensional datasets and the covariance
-function must be represented by a mixture of exponentials, we will return to a
+function must be represented by a mixture of exponentials; we will return to a
 discussion of what this means in detail in the following sections.
@@
 matrix when it is generated by a mixture of exponentials.
-More precisely, the matrix can be embedded in an equivalent higher-dimensional
-system that can be solved efficiently using a sparse or banded solver.
+\response{More precisely, the matrix can be embedded in an equivalent higher-dimensional
+system that can be solved efficiently using a sparse or banded solver. Replace
+with(?): The matrix can be expressed as the sum of semi-separable matrices, each
+of which is constructed from the outer product of one-dimensional vectors.  These
+are a type of low-rank matrix, which may be solved by Cholesky decomposition
+using a novel technique developed by one of us (SA).}
 As we will discuss in the following pages, this choice of kernel function is
@@
 In the following pages, we motivate the general problem of GP modeling,
-describe the previously published scalable method \citep{Rybicki:1995,
-Ambikasaran:2015} and our generalization, and demonstrate the model's
-application on various real and simulated data sets.
+describe the \response{(delete) previously published} scalable method 
+\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate 
+the model's application on various real and simulated data sets.
 Alongside this paper, we have released well-tested and documented open source

### 2017-05-24 papers/paper1/figures/simulated/correct.py
updating simulated data examples using notebooks

@@ -1,97 +1,0 @@
-#!/usr/bin/env python
-# -*- coding: utf-8 -*-
-
-from __future__ import division, print_function
-
-import emcee
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.optimize import minimize
-
-import celerite
-from celerite import terms
-from celerite import plot_setup
-
-np.random.seed(42)
-plot_setup.setup(auto=True)
-
-# Simulate some data
-kernel = terms.SHOTerm(log_S0=0.0, log_omega0=2.0, log_Q=2.0,
-                       bounds=[(-10, 10), (-10, 10), (-10, 10)])
-gp = celerite.GP(kernel)
-true_params = np.array(gp.get_parameter_vector())
-omega = 2*np.pi*np.exp(np.linspace(-np.log(10.0), -np.log(0.1), 5000))
-true_psd = gp.kernel.get_psd(omega)
-N = 200
-t = np.sort(np.random.uniform(0, 10, N))
-yerr = 2.5
-y = gp.sample(t, diag=yerr**2)
-
-# Find the maximum likelihood model
-gp.compute(t, yerr)
-
-def nll(params, gp, y):
-    gp.set_parameter_vector(params)
-    if not np.isfinite(gp.log_prior()):
-        return 1e10
-    ll = gp.log_likelihood(y)
-    return -ll if np.isfinite(ll) else 1e10
-
-# Run the MCMC
-p0 = true_params + 1e-4*np.random.randn(len(true_params))
-soln = minimize(nll, p0, method="L-BFGS-B", args=(gp, y))
-gp.set_parameter_vector(soln.x)
-ml_psd = gp.kernel.get_psd(omega)
-
-def log_probability(params):
-    gp.set_parameter_vector(params)
-    lp = gp.log_prior()
-    if not np.isfinite(lp):
-        return -np.inf
-    ll = gp.log_likelihood(y)
-    return ll + lp if np.isfinite(ll) else -np.inf
-
-ndim = len(soln.x)
-nwalkers = 32
-coords = soln.x + 1e-4 * np.random.randn(nwalkers, ndim)
-
-sampler = emcee.EnsembleSampler(nwalkers, ndim, log_probability)
-coords, _, _ = sampler.run_mcmc(coords, 500)
-sampler.reset()
-coords, _, _ = sampler.run_mcmc(coords, 2000)
-
[…]

### 2017-05-24 papers/paper1/figures/simulated/wrong-qpo.py

@@ -1,135 +1,0 @@
-#!/usr/bin/env python
-# -*- coding: utf-8 -*-
-
-from __future__ import division, print_function
-
-import emcee
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.optimize import minimize
-from scipy.linalg import cho_factor, cho_solve
-
-import celerite
-from celerite import terms
-from celerite.modeling import Model
-
-from celerite import plot_setup
-
-plot_setup.setup(auto=True)
-
-class TrueModel(Model):
-    parameter_names = ("log_amp", "log_ell", "log_period")
-
-    def get_K(self, x):
-        tau = x[:, None] - x[None, :]
-        return (
-            np.exp(self.log_amp - 0.5 * tau**2 * np.exp(-2.0*self.log_ell)) *
-            np.cos(2*np.pi*tau*np.exp(-self.log_period))
-        )
-
-    def __call__(self, params, x, y, yerr):
-        self.set_parameter_vector(params)
-        lp = self.log_prior()
-        if not np.isfinite(lp):
-            return -np.inf
-
-        K = self.get_K(x)
-        K[np.diag_indices_from(K)] += yerr**2
-        try:
-            factor = cho_factor(K, overwrite_a=True)
-        except (np.linalg.LinAlgError, ValueError):
-            return -np.inf
-        ld = 2.0 * np.sum(np.log(np.diag(factor[0])))
-        return -0.5*(np.dot(y, cho_solve(factor, y))+ld) + lp
-
-true_model = TrueModel(log_amp=0.0, log_ell=np.log(5.0), log_period=0.0,
-                       bounds=[(-10, 10), (-10, 10), (-10, 10)])
-
-# Simulate a dataset from the true model
-np.random.seed(42)
-N = 100
-t = np.sort(np.random.uniform(0, 20, N))
-yerr = 0.5
-K = true_model.get_K(t)
-K[np.diag_indices_from(K)] += yerr**2
-y = np.random.multivariate_normal(np.zeros(N), K)
-
-# Set up the celerite model that we will use to fit - product of two SHOs
-log_Q = 1.0
-kernel = terms.SHOTerm(log_S0=np.log(np.var(y))-2*log_Q, log_Q=log_Q,
-                       log_omega0=np.log(2*np.pi))
[…]

### 2017-05-26 papers/paper1/ms.tex
updating simulated data examples using notebooks
rasterizing points on rotation plot
updating tex for new figures etc.

@@
 \include{vc}
-\include{figures/rotation}
+\include{figures/rotation/rotation}
 
@@
 \includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
-\caption{(left) A simulated dataset.
+\caption{(top left) A simulated dataset (black error bars), and maximum
+    likelihood model (blue contours).
+    (bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
     (right) The inferred PSD~--~the blue contours encompass 68\% of the
@@
 \includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
-\caption{(left) A simulated dataset.
+\caption{(top left) A simulated dataset (black error bars), and maximum
+    likelihood model (blue contours).
+    (bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
     (right) The inferred period of the process. The true period is indicated
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/rotation.pdf}
+\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
 \caption{Inferred constraints on a quasiperiodic GP model using the covariance
     function in \eq{rot-kernel} and two quarters of \kepler\ data.
-(left) The \kepler\ data (black points) and the maximum likelihood model
-    prediction (blue curve) for a 50~day subset of the data used.
+(top left) The \kepler\ data (black points) and the maximum likelihood model
+    prediction (blue curve) for a 60~day subset of the data used.
     The solid blue line shows the predictive mean and the blue contours show
     the predictive standard deviation.
-(center) Inferred constraints on the model PSD.
-    The dashed line shows the maximum likelihood PSD, the blue solid line
-    shows the median of posterior PSD, and the blue contours show the 68\%
-    credible region.
-(right) Inferred constraints on the model covariance function.
-    The dashed line shows the maximum likelihood model, the blue solid line
-    shows the median of posterior, and the blue contours show the 68\%
-    credible region.
-    \figurelabel{rotation}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{figures/rotation-period.pdf}
-\caption{The posterior constraint on the rotation period of KIC~1430163 using
+(bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
+(right) The posterior constraint on the rotation period of KIC~1430163 using
     the dataset and model from \Figure{rotation}.
@@
     dashed lines.
-    \figurelabel{rotation-period}}
+    \figurelabel{rotation}}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
+\includegraphics[width=0.8\textwidth]{figures/astero/astero.pdf}
+\caption{
+\response{
+(top) The \kepler\ data (black points) and the maximum likelihood model
+    prediction (blue curve) for a 10~day subset of the month-long dataset that
+    was used for the fit.
+    The solid blue line shows the predictive mean and the blue contours show
+    the predictive standard deviation.
+(bottom) The residuals between the mean predictive model and the data
+    shown in the top figure.
+}
+    \figurelabel{astero}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
 \caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
+\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
 \caption{A comparison between the Lomb-Scargle estimator of the PSD and the
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/transit-ml.pdf}
+\includegraphics[width=\textwidth]{figures/transit/transit-ml.pdf}
     \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
@@
 \begin{center}
-\includegraphics[width=\textwidth]{figures/transit-corner.pdf}
+\includegraphics[width=\textwidth]{figures/transit/transit-corner.pdf}
 \caption{The marginalized posterior constraints on the physical parameters of
@@
 \includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
-\caption{A benchmark showing the computational scaling of \celerite\ using the
+\caption{\dfmtodo{Cholesky} A benchmark showing the computational scaling of \celerite\ using the
     \texttt{vanilla} band solver as a function of the number of data points
@@
 
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_lapack.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
-    with the \project{MKL} \project{BLAS} library.
-    \figurelabel{benchmark-lapack}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_sparse.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver
-    from \project{Eigen} \citep{Guennebaud:2010}.
-    \figurelabel{benchmark-sparse}}
-\end{center}
-\end{figure}
-
 \section{Ensuring positive definiteness}

### 2017-05-26 paper/ms.tex
check for paper changes

@@
 dimensional data sets.
-This method is a generalization of a method developed by
-\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
+\response{This method is a generalization of a method developed by
+\citet{Ambikasaran:2015}
+ ({\bf This is no longer correct -different Ambikasaran result!})}
+that was, in turn, built on intuition from a twenty
 year old paper \citep{Rybicki:1995}.
 This method can only be used with one-dimensional datasets and the covariance
-function must be represented by a mixture of exponentials, we will return to a
+function must be represented by a mixture of exponentials; we will return to a
 discussion of what this means in detail in the following sections.
@@
 In the following pages, we motivate the general problem of GP modeling,
-describe the previously published scalable method \citep{Rybicki:1995,
-Ambikasaran:2015} and our generalization, and demonstrate the model's
-application on various real and simulated data sets.
+describe the \response{(delete) previously published} scalable method
+\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
+the model's application on various real and simulated data sets.
 Alongside this paper, we have released well-tested and documented open source
@@
 
-\section{Implementation \& performance}
-
-\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
-elements given by \eq{kernel-simple}, could be computed
-efficiently by taking advantage of the structure of this covariance function.
-\citet{Ambikasaran:2015} generalized this computation to apply to the full
-mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
-efficient method for computing the determinant of $K$.
-
-\subsection{An example}
-
-To provide some insight for this method, we follow
-\citet{Ambikasaran:2015} and start by working through a simple example.
-In this case, we assume that we have three data points
-$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
-variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
-like to compute the likelihood of these data under a GP model with the
-covariance function
-\begin{eqnarray}
-k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
-\end{eqnarray}
-This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and it is the model studied by
-\citet{Rybicki:1995}.
-To demonstrate that the likelihood of this model can be computed in
-$\mathcal{O}(N)$, we write out the full system of
-equations that must be solved to apply the inverse of $K$ and compute
-the first term of \eq{gp-likelihood}.
-In matrix notation, this is
-\begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{z} &=& \bvec{y}, \\
-\begin{pmatrix}
-    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
-    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
-    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ z_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ y_2 \\ y_3
-\end{pmatrix}
-\end{eqnarray}
-where our goal is to solve for the unknown vector \bvec{z} for a given matrix
-$K$ and vector \bvec{y}.
-In \eq{impl-matrix}, we have assumed that the mean function is zero but a
-non-zero mean could be included by replacing \bvec{y} by
-$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
-Now, if we introduce the variables
-\begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
-\end{eqnarray}
-where $g_{0} = 0$, and
-\begin{eqnarray}\eqlabel{algo-first}
-    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
-\end{eqnarray}
-where $u_{N} = 0$,
-the system of equations can be rewritten as
-\begin{eqnarray}
-(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
-a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
-\end{eqnarray}
-Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
-as a matrix equation shows the benefit that this seemingly trivial
-reformulation provides:
-\begin{eqnarray}
-\begin{pmatrix}
-    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
-    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
-    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
-    0 & 0 & 0 & 0 & -1 & 0 & a \\
-    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
-\end{pmatrix}\nonumber
-\end{eqnarray}
-Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
-rewrite \eq{impl-matrix} as
-\begin{eqnarray}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
-\end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
-with, it is now sparse with banded structure that can be exploited to solve
-the system efficiently.
-In particular, sparse solvers are available that can perform a
-LU-decomposition of matrices like this in $\mathcal{O}(N)$
-operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
-general~--~and we can use these algorithms to solve our system
-exactly because the target vector $\bvec{z}$ is a subset of the elements of
-$\bvec{z}_\mathrm{ext}$.
-
-In the following section we discuss this method more generally, but it's
-worth noting a few important facts that can already be seen in this example.
-First, the fundamental reason why this matrix $K$ can be solved efficiently is
-the following property of exponentials
-\begin{eqnarray}
-    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
-    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
-    e^{-c\,(t_3 - t_1)}
-\end{eqnarray}
-and it is important to note that this property does not extend to other common
-covariance functions like the ``exponential-squared'' function
-\begin{eqnarray}
-    k(\tau) \propto e^{-c\,\tau^2} \quad.
-\end{eqnarray}
-Second, our derivation of the extended matrix requires that the data
-points be monotonically sorted.
-This is the motivation for our previous statement that this method can only be
-applied in one dimension since neither of these properties will be satisfied
-by multidimensional inputs.
-
-\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
-extended matrix formalism in practice.
-First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with a bandwidth that scales linearly
-with the number of components, $J$.
-Second, \citet{Ambikasaran:2015} proved that the absolute value of the
-determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
-determinant of $K$.
-This means we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(N)$ operations.
-
-\subsection{The algorithm}
-
-In this section, we generalize the method from the previous section to the
-covariance function given by \eq{celerite-kernel}.
-This derivation follows \citet{Ambikasaran:2015} but it includes explicit
-treatment of complex parameters, and their complex conjugates.
-
-In the case of the full \celeriteterm\ covariance function
-(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
-analogy to the $u_n$ and $g_n$ that we introduced in the previous section
-\begin{eqnarray}\eqlabel{full-system-first}
-\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
-\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
-g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
-h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
-u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
-v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
-\end{eqnarray}
-with the boundary conditions
-\begin{eqnarray}
-    g_{0,j} = 0 \quad, \quad
-    h_{0,j} = 0 \quad, \quad
-    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
-    v_{N,j} = 0
-\end{eqnarray}
-for all $j$.
-Using these variables and some algebra, we find that the following expression
-\begin{eqnarray}\eqlabel{full-system-last}
-\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
-+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
-+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
-    &=& r_{\bvec{\theta},n}
-\end{eqnarray}
-is equivalent to the target matrix equation
-\begin{eqnarray}
-K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
-\end{eqnarray}
-if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
-$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
-\eq{full-system-first} through \eq{full-system-last} define a banded matrix
-equation in the ``extended'' space
-\begin{eqnarray}\eqlabel{kext-full}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
-    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
-\end{eqnarray}
-and, as before, this can be used to solve
-for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
-operations.
-\Figure{matrix} shows a pictorial representation of the sparsity pattern of
-the extended matrix $K_\mathrm{ext}$.
-The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
-with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
-non-zero entries.
-Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
-$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
-as
-\begin{eqnarray}
-{\bvec{z}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
-    h_{N-1,j} & z_N
-\end{array}\right)
-\end{eqnarray}
-and
-\begin{eqnarray}
-{\bvec{r}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},N}
-\end{array}\right) \quad.
-\end{eqnarray}
-After constructing the extended matrix (using a compact storage format), the
-extended matrix can be factorized using a LU-decomposition\footnote{Even
-though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
-Cholesky solver cannot be used for increased efficiency.} routine optimized
-for band or sparse matrices.
-This decomposition can then be used to compute the determinant of $K$, solve
-$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
-likelihood in \eq{gp-likelihood}.
-
-In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
-case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
-will also be identically zero for all $n$.
-Removing these trivial rows from the extended matrix results in factor of two
-increase in the computational efficiency for these terms.
-In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
-$d_j=0$, and the general terms are called ``complex''.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
-\caption{A pictorial representation of the sparse extended matrix
-$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
-Each colored block corresponds to a non-zero entry in the matrix as described
-in the legend.
-    \figurelabel{matrix}}
-\end{center}
-\end{figure}
-
-\subsection{Implementation considerations \& scaling}
-
-The extended system defined in the previous section is sparse, with typically
-fewer than a few percent non-zero entries and band structure.
-In this section, we empirically investigate the performance and scaling
-of three different algorithms for solving this extended system:
-
-\begin{enumerate}
-
-\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
-    for banded matrices using Gaussian elimination \citep[we implemented this
-    method following][]{Press:1992, Press:2007},
-
-\item \texttt{lapack}: The general banded LU decomposition implementation from
-    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
-    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines,\footnote{The experiments in this paper use the
-    Intel Math Kernel Library (\project{MKL}):
-    \url{https://software.intel.com/en-us/intel-mkl}.} and
-
-\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
-    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
-    sparsity but not the band structure.
-
-\end{enumerate}
-
-The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
-because the dimension of the extended matrix scales as $N\,J$ and the
-bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
-\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
-\project{SuperLU} package \citep{Demmel:1999}.
-We find that, while the \texttt{vanilla} solver scales as expected, the
-\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
-offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
-
-The benchmark experiments shown here were performed on a MacBook Pro with two
-2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
-CPUs and running Ubuntu.
-\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
-with $N$ and $J$ using the \texttt{vanilla} solver.
-As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
-in $J$ for large $J$.
-\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
-the \texttt{lapack} and \texttt{sparse} solvers respectively.
-For each of these optimized solvers, the empirical scaling is
-$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
-Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
-efficient than the other algorithms.
-The real world performance of \celerite\ depends on the specific platform,
-hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
-found qualitatively similar results across popular platforms and
-state-of-the-art libraries.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
-\caption{A benchmark showing the computational scaling of \celerite\ using the
-    \texttt{vanilla} band solver as a function of the number of data points
-    and the number of terms.
-    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
-    matrix given by \eq{celerite-kernel} as a function of the number of data
-    points $N$.
-    The different lines show the cost for different numbers of terms $J$
-    increasing from bottom to top.
-    To guide the eye, the straight black line without points
-    shows linear scaling in $N$.
-    \emph{(right)} The same information plotted as a function of $J$ for
-    different values of $N$.
-    Each line shows the scaling for a specific value of $N$ increasing from
-    bottom to top.
-    The black line shows quadratic scaling in $J$.
-    \figurelabel{benchmark}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_lapack.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{lapack} solver
-    with the \project{MKL} \project{BLAS} library.
-    \figurelabel{benchmark-lapack}}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin_sparse.pdf}
-\caption{The same as \Figure{benchmark} but using the \texttt{sparse} solver
-    from \project{Eigen} \citep{Guennebaud:2010}.
-    \figurelabel{benchmark-sparse}}
-\end{center}
-\end{figure}
+\section{Semi-separable matrix factorization}
+
+\subsection{Cholesky factorization}
+
+The relationship between the \celeriteterm\ model and semi-separable linear
+algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
+A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
+written as
+\begin{eqnarray}
+K_{n,m} = \left\{\begin{array}{ll}
+    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
+    A_{n,n} & \mathrm{if}\,m=n \\
+    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
+\end{array}\right.
+\end{eqnarray}
+for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
+N$-dimensional diagonal matrix $A$.
+This definition can be equivalently written as
+\begin{eqnarray}\eqlabel{semi-sep}
+K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+\end{eqnarray}
+where the $\mathrm{tril}$ function extracts the strict lower triangular
+component of its argument and $\mathrm{triu}$ is the equivalent strict upper
+triangular operation.
+For the \celeriteterm\ model described in the previous section, the rank of
+the semi-separable matrix is $R = 2\,J$ and the components of the relevant
+matrices are
+\begin{eqnarray}
+U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
+    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
+U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
+    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
+V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
+\end{eqnarray}
+
+To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
+compute the log-determinant of $K$ and solve linear systems of the form
+$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
+semi-separable matrix.
+If $K$ is positive semi-definite, these can be computed using the Cholesky
+factorization of $K$:
+\begin{eqnarray}
+K = L\,D\,L^\T
+\end{eqnarray}
+where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
+diagonal matrix.
+We use the ansatz that $L$ has the form
+\begin{eqnarray}
+    L = I + \mathrm{tril} (U\,X^T)
+\end{eqnarray}
+where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
+$N \times 2\,J$ matrix.
+Combining this with \eq{semi-sep}, we find the equation
+\begin{eqnarray}\eqlabel{cholesky}
+A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,X^\T)\,D
+    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $X$.
+Setting each element on the left side of \eq{cholesky} equal to the
+corresponding element on the right side, we derive the following recursion
+relations
+\begin{eqnarray}
+S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
+    \nonumber\\
+X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
+    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
+    \eqlabel{algorithm}
+\end{eqnarray}
+where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
+$S_{1,j,k}=0$.
+The computational cost of one iteration of the update in \eq{algorithm} is
+$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
+observation~--~so the total cost of this factorization scales as
+$\mathcal{O}(N\,J^2)$.
+
+\subsection{Pre-conditioning}
+
+Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
+overflow and underflow issues in many realistic cases because both $U$ and $V$
+have factors of the form $e^{\pm t}$, where $t$ can be any arbitrarily large
+real number.
+This can be avoided with the appropriate reparameterization.
+We define the following pre-conditioned variables
+\begin{eqnarray}
+\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
+\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
+\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
+\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
+\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
+\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
+\end{eqnarray}
+and the set of variables
+\begin{eqnarray}
+\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
+\end{eqnarray}
+with the constraint
+\begin{eqnarray}
+\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
+\end{eqnarray}
+Using this parameterization, we find the following, numerically stable,
+algorithm
+\begin{eqnarray}
+    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
+    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
+    \nonumber\\
+\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
+    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
+\end{eqnarray}
+
+\subsection{Solving}
+
+After computing the Cholesky factorization of the matrix $K$, we can apply
+the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
+$\mathcal{O}(N)$ operations respectively.
+The log-determinant of $K$ is given by
+\begin{eqnarray}
+    \log \det K = \sum_{n=1}^N \log D_{n,n}
+\end{eqnarray}
+where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
+
+The inverse of $K$ can be applied using the relationship
+\begin{eqnarray}
+\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
+substitution algorithm
+\begin{eqnarray}
+    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
+    \tilde{X}_{n-1,j}\,z_{n-1}\right] \nonumber\\
+    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{0,j} = 0$ for all $j$.
+Using this result and the following backward substitution algorithm, we can
+compute $\bvec{b}$
+\begin{eqnarray}
+g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
+    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
+b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{X}_{n,j}\,g_{n,j}
+\end{eqnarray}
+where $g_{N+1,j} = 0$ for all $j$.
+The total cost of the forward and backward substitution passes scales as
+$\mathcal{O}(N\,J)$.
+
+\subsection{Simulation}
 
@@
 more reliable than other methods.
-\Figure{rotation} shows a subset of the data used in this example and the
-posterior inferences of the PSD and autocorrelation function of the process.
-\Figure{rotation-period} shows the marginalized posterior distribution for the
-rotation period.
+The left panels of \Figure{rotation} show a subset of the data used in this
+example and the residuals away from the maximum likelihood predictive mean.
+The right panel of \Figure{rotation} shows the marginalized posterior
+distribution for the rotation period.
 
@@
 \appendix
+
+\section{An alternative solver using the extended matrix formalism}
+
+\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
+elements given by \eq{kernel-simple}, could be computed
+efficiently by taking advantage of the structure of this covariance function.
+\citet{Ambikasaran:2015} generalized this computation to apply to the full
+mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
+efficient method for computing the determinant of $K$.
+
+\subsection{An example}
+
+To provide some insight for this method, we follow
+\citet{Ambikasaran:2015} and start by working through a simple example.
+In this case, we assume that we have three data points
+$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
+variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
+like to compute the likelihood of these data under a GP model with the
+covariance function
+\begin{eqnarray}
+k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
+\end{eqnarray}
+This is the result of setting $J=1$, $b=0$, and $d=0$ in
+\eq{celerite-kernel-complex} and it is the model studied by
+\citet{Rybicki:1995}.
+To demonstrate that the likelihood of this model can be computed in
+$\mathcal{O}(N)$, we write out the full system of
+equations that must be solved to apply the inverse of $K$ and compute
+the first term of \eq{gp-likelihood}.
+In matrix notation, this is
+\begin{eqnarray}\eqlabel{impl-matrix}
+K\,\bvec{z} &=& \bvec{y}, \\
+\begin{pmatrix}
+    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
+    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
+    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ z_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ y_2 \\ y_3
+\end{pmatrix}
+\end{eqnarray}
+where our goal is to solve for the unknown vector \bvec{z} for a given matrix
+$K$ and vector \bvec{y}.
+In \eq{impl-matrix}, we have assumed that the mean function is zero but a
+non-zero mean could be included by replacing \bvec{y} by
+$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
+Now, if we introduce the variables
+\begin{eqnarray}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
+\end{eqnarray}
+where $g_{0} = 0$, and
+\begin{eqnarray}\eqlabel{algo-first}
+    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
+\end{eqnarray}
+where $u_{N} = 0$,
+the system of equations can be rewritten as
+\begin{eqnarray}
+(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
+a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
+\end{eqnarray}
+Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
+as a matrix equation shows the benefit that this seemingly trivial
+reformulation provides:
+\begin{eqnarray}
+\begin{pmatrix}
+    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
+    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
+    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
+    0 & 0 & 0 & 0 & -1 & 0 & a \\
+    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
+\end{pmatrix}\nonumber
+\end{eqnarray}
+Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
+rewrite \eq{impl-matrix} as
+\begin{eqnarray}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
+\end{eqnarray}
+Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
+with, it is now sparse with banded structure that can be exploited to solve
+the system efficiently.
+In particular, sparse solvers are available that can perform a
+LU-decomposition of matrices like this in $\mathcal{O}(N)$
+operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
+general~--~and we can use these algorithms to solve our system
+exactly because the target vector $\bvec{z}$ is a subset of the elements of
+$\bvec{z}_\mathrm{ext}$.
+
+In the following section we discuss this method more generally, but it's
+worth noting a few important facts that can already be seen in this example.
+First, the fundamental reason why this matrix $K$ can be solved efficiently is
+the following property of exponentials
+\begin{eqnarray}
+    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
+    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
+    e^{-c\,(t_3 - t_1)}
+\end{eqnarray}
+and it is important to note that this property does not extend to other common
+covariance functions like the ``exponential-squared'' function
+\begin{eqnarray}
+    k(\tau) \propto e^{-c\,\tau^2} \quad.
+\end{eqnarray}
+Second, our derivation of the extended matrix requires that the data
+points be monotonically sorted.
+This is the motivation for our previous statement that this method can only be
+applied in one dimension since neither of these properties will be satisfied
+by multidimensional inputs.
+
+\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
+extended matrix formalism in practice.
+First, even if the covariance function is a mixture of exponentials, the
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components, $J$.
+Second, \citet{Ambikasaran:2015} proved that the absolute value of the
+determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
+determinant of $K$.
+This means we can use this extended matrix formalism to compute the
+marginalized likelihood in $\mathcal{O}(N)$ operations.
+
+\subsection{The algorithm}
+
+In this section, we generalize the method from the previous section to the
+covariance function given by \eq{celerite-kernel}.
+This derivation follows \citet{Ambikasaran:2015} but it includes explicit
+treatment of complex parameters, and their complex conjugates.
+
+In the case of the full \celeriteterm\ covariance function
+(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
+analogy to the $u_n$ and $g_n$ that we introduced in the previous section
+\begin{eqnarray}\eqlabel{full-system-first}
+\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
+\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
+\end{eqnarray}
+with the boundary conditions
+\begin{eqnarray}
+    g_{0,j} = 0 \quad, \quad
+    h_{0,j} = 0 \quad, \quad
+    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
+    v_{N,j} = 0
+\end{eqnarray}
+for all $j$.
+Using these variables and some algebra, we find that the following expression
+\begin{eqnarray}\eqlabel{full-system-last}
+\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
++ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
++ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
+    &=& r_{\bvec{\theta},n}
+\end{eqnarray}
+is equivalent to the target matrix equation
+\begin{eqnarray}
+K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
+\end{eqnarray}
+if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
+$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
+\eq{full-system-first} through \eq{full-system-last} define a banded matrix
+equation in the ``extended'' space
+\begin{eqnarray}\eqlabel{kext-full}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
+    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
+\end{eqnarray}
+and, as before, this can be used to solve
+for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
+operations.
+\Figure{matrix} shows a pictorial representation of the sparsity pattern of
+the extended matrix $K_\mathrm{ext}$.
+The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
+with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
+non-zero entries.
+Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
+$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
+as
+\begin{eqnarray}
+{\bvec{z}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
+    h_{N-1,j} & z_N
+\end{array}\right)
+\end{eqnarray}
+and
+\begin{eqnarray}
+{\bvec{r}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
+    0 & \cdots & 0 & r_{\bvec{\theta},N}
+\end{array}\right) \quad.
+\end{eqnarray}
+After constructing the extended matrix (using a compact storage format), the
+extended matrix can be factorized using a LU-decomposition\footnote{Even
+though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
+Cholesky solver cannot be used for increased efficiency.} routine optimized
+for band or sparse matrices.
+This decomposition can then be used to compute the determinant of $K$, solve
+$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
+likelihood in \eq{gp-likelihood}.
+
+In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
+case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
+will also be identically zero for all $n$.
+Removing these trivial rows from the extended matrix results in factor of two
+increase in the computational efficiency for these terms.
+In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
+$d_j=0$, and the general terms are called ``complex''.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
+\caption{A pictorial representation of the sparse extended matrix
+$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
+Each colored block corresponds to a non-zero entry in the matrix as described
+in the legend.
+    \figurelabel{matrix}}
+\end{center}
+\end{figure}
+
+\subsection{Implementation considerations \& scaling}
+
+The extended system defined in the previous section is sparse, with typically
+fewer than a few percent non-zero entries and band structure.
+In this section, we empirically investigate the performance and scaling
+of three different algorithms for solving this extended system:
+
+\begin{enumerate}
+
+\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
+    for banded matrices using Gaussian elimination \citep[we implemented this
+    method following][]{Press:1992, Press:2007},
+
+\item \texttt{lapack}: The general banded LU decomposition implementation from
+    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
+    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
+    \project{BLAS} routines,\footnote{The experiments in this paper use the
+    Intel Math Kernel Library (\project{MKL}):
+    \url{https://software.intel.com/en-us/intel-mkl}.} and
+
+\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
+    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
+    sparsity but not the band structure.
+
+\end{enumerate}
+
+The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
+because the dimension of the extended matrix scales as $N\,J$ and the
+bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
+\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
+\project{SuperLU} package \citep{Demmel:1999}.
+We find that, while the \texttt{vanilla} solver scales as expected, the
+\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
+offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
+
+The benchmark experiments shown here were performed on a MacBook Pro with two
+2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
+CPUs and running Ubuntu.
+\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
+with $N$ and $J$ using the \texttt{vanilla} solver.
+As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
+in $J$ for large $J$.
+\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
+the \texttt{lapack} and \texttt{sparse} solvers respectively.
+For each of these optimized solvers, the empirical scaling is
+$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
+Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
+efficient than the other algorithms.
+The real world performance of \celerite\ depends on the specific platform,
+hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
+found qualitatively similar results across popular platforms and
+state-of-the-art libraries.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
+\caption{\dfmtodo{Cholesky} A benchmark showing the computational scaling of \celerite\ using the
+    \texttt{vanilla} band solver as a function of the number of data points
+    and the number of terms.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
+    \figurelabel{benchmark}}
+\end{center}
+\end{figure}
 

### 2017-05-26 paper/ms.tex
response color [ci skip]

@@
 
-\newcommand{\response}[1]{{\color{green}#1}}
+\newcommand{\response}[1]{{\color{blue}#1}}
 
@@
 \celerite\ comes from the French word \foreign{c\'elerit\'e} meaning the speed
-of light in a vacuum.} The \celerite\ method requires using a specific model
+of light in a vacuum.
+\response{Throughout the paper, when referring to the \celeriteterm\ model, we
+typeset \celeriteterm\ in italics and, in reference to the implementation of
+this model, we typeset \celerite\ in sans-serif.}}
+The \celerite\ method requires using a specific model
 for the covariance $k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and, although it

### 2017-05-27 paper/figures/example-stats.tex
adding full specifications for the models

\begin{floattable}
\begin{deluxetable}{cccccccc}
\caption{The scaling and convergence stats for each example. \label{tab:example-stats}}
\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & \colhead{direct cost} & \colhead{\celerite\ cost} & \colhead{$D$} & \colhead{steps} & \colhead{effective samples}\\
&&& \colhead{ms} & \colhead{ms} &&&}
\startdata
1 & 200 & 1 & 2.85 & 0.26 & 3 & 80000 & 1737 \\
2 & 100 & 2 & 0.67 & 0.36 & 4 & 80000 & 1134 \\
3 & 6950 & 2 & 7351.71 & 1.46 & 4 & 176000 & 3606 \\
4 & 1440 & 8 & 828.93 & 2.74 & 9 & 640000 & 1443 \\
5 & 1000 & 2 & 82.73 & 0.94 & 13 & 640000 & 1141 \\
\enddata
\end{deluxetable}
\end{floattable}

### 2017-05-27 paper/ms.tex
adding full specifications for the models

@@
 \begin{eqnarray}
-    \log \det K = \sum_{n=1}^N \log D_{n,n}
+    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
 \end{eqnarray}
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
+\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
 \caption{(left) The power spectrum of a stochastically-driven simple harmonic
@@
     $S(d_j)/S(0) = 100$.
-    (right) The corresponding autocorrelation functions with the same colors.
+    (middle) The corresponding autocorrelation functions with the same colors.
+    \response{(right) Three realizations from each model in the same colors.}
     \figurelabel{sho}}
@@
 reproduces the true PSD.
+
+\input{figures/simulated/correct-params}
 
@@
 
+\input{figures/simulated/wrong-qpo-params}
+
 \begin{figure}[!htbp]
@@
 \begin{eqnarray}\eqlabel{rot-kernel}
-k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
+k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
     \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
@@
 distribution for the rotation period.
+
+\input{figures/rotation/rotation-params}
 
@@
 
+\input{figures/astero/astero-params}
+
 \begin{figure}[!htbp]
@@
 
+\newpage
+\input{figures/transit/transit-params}
+
 \begin{figure}[!htbp]
@@
 \end{figure}
+
+\subsection{Summary}
+
+\input{figures/example-stats}
 
@@
 for helpful discussions informing the ideas and code presented here.
+\response{We would also like to thank the anonymous referee for thorough and
+constructive feedback that greatly improved the paper.}
 

### 2017-05-27 paper/figures/collect_results.py

@@ -1,0 +1,65 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import json
+
+stats_rows = []
+for i, example in enumerate(["simulated/correct", "simulated/wrong-qpo",
+                             "rotation/rotation", "astero/astero",
+                             "transit/transit"]):
+    with open(example + ".json") as f:
+        data = json.load(f)
+
+    data["ntot"] = data["nwalkers"] * (data["nburn"] + data["nsteps"])
+    data["direct_time"] *= 1000
+    data["time"] *= 1000
+    stats_rows.append(
+        (r"{0} & {N} & {J} & {direct_time:.2f} & {time:.2f} & {ndim} & {ntot} "
+         r"& {neff:.0f} \\")
+        .format(i+1, **data)
+    )
+
+    with open(example + "-params.json") as f:
+        params = json.load(f)
+
+    rows = []
+    for p in params:
+        if len(p[1]) == 1:
+            rows.append(r"{0} & {1} \\"
+                        .format(p[0], p[1][0]))
+        elif len(p[1]) == 2:
+            rows.append(r"{0} & $\mathcal{{U}}({1[0]},\,{1[1]})$ \\"
+                        .format(*p))
+        else:
+            assert False
+
+    with open(example + "-params.tex", "w") as f:
+        f.write("\\begin{floattable}\n")
+        f.write("\\begin{deluxetable}{cc}\n")
+        f.write("\\tablecaption{{The parameters for Example {0}. "
+                "\label{{tab:example-{0}-params}}}}\n".format(i+1))
+        f.write("\\tablehead{\colhead{parameter} & \colhead{prior}}\n")
+        f.write("\\startdata\n")
+        f.write("\n".join(rows))
+        f.write("\n\\enddata\n")
+        f.write("\end{deluxetable}\n")
+        f.write("\end{floattable}\n")
+
+
+with open("example-stats.tex", "w") as f:
+    f.write("\\begin{floattable}\n")
+    f.write("\\begin{deluxetable}{cccccccc}\n")
+    f.write("\caption{The scaling and convergence stats for each example. "
+            "\label{tab:example-stats}}\n")
+    f.write("\\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & "
+            "\colhead{direct cost} & \colhead{\celerite\ cost} & "
[…]

### 2017-05-27 paper/figures/sho.py

@@ -7,6 +7,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 from celerite.plot_setup import setup, get_figsize
 
+np.random.seed(42)
 setup(auto=True)
 
 def sho_psd(Q, x):
@@ -30,14 +31,23 @@ def lorentz_acf(Q, tau):
     t = np.abs(tau)
     return np.exp(-0.5*t/Q) * np.cos(t)
 
-fig, (ax1, ax2) = plt.subplots(1, 2, figsize=get_figsize(1, 2))
+fig, (ax1, ax2, ax3) = plt.subplots(1, 3, figsize=get_figsize(1, 3))
 x = 10**np.linspace(-1.1, 1.1, 5000)
-tau = np.linspace(0, 20, 5000)
+tau = np.linspace(0, 20, 1000)
 
-for Q_name, Q in [("1/2", 0.5), ("1/\\sqrt{2}", 1./np.sqrt(2)),
-                  ("2", 2.0), ("10", 10.0)]:
-    ax1.plot(x, sho_psd(Q, x), label="$Q = {0}$".format(Q_name), lw=1.5)
-    ax2.plot(tau, sho_acf(Q, tau), label="$Q = {0}$".format(Q_name), lw=1.5)
+for i, (Q_name, Q) in enumerate(
+        [("1/2", 0.5), ("1/\\sqrt{2}", 1./np.sqrt(2)), ("2", 2.0),
+         ("10", 10.0)]):
+    l, = ax1.plot(x, sho_psd(Q, x), label="$Q = {0}$".format(Q_name), lw=1.5)
+    c = l.get_color()
+    ax2.plot(tau, sho_acf(Q, tau), label="$Q = {0}$".format(Q_name), lw=1.5,
+             color=c)
+
+    K = sho_acf(Q, tau[:, None] - tau[None, :])
+    y = np.random.multivariate_normal(np.zeros(len(tau)), K, size=3)
+    ax3.axhline(-5*i, color="k", lw=0.75)
+    ax3.plot(tau, -5*i + (y - np.mean(y, axis=1)[:, None]).T, color=c,
+             lw=1)
 
 ax1.plot(x, lorentz_psd(10.0, x), "--k")
 ax2.plot(tau, lorentz_acf(10.0, tau), "--k")
@@ -54,5 +64,9 @@ ax2.set_xlim(tau.min(), tau.max())
[…]

### 2017-05-30 paper/figures/example-stats.tex
words about simulated examples

@@
 \caption{The scaling and convergence stats for each example. \label{tab:example-stats}}
-\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & \colhead{direct cost} & \colhead{\celerite\ cost} & \colhead{$D$} & \colhead{steps} & \colhead{effective samples}\\
+\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & \colhead{direct cost} & \colhead{\celerite\ cost} & \colhead{$D$} & \colhead{model evaluations} & \colhead{effective samples}\\
 &&& \colhead{ms} & \colhead{ms} &&&}

### 2017-05-30 paper/ms.tex
words about simulated examples

@@
 \include{vc}
+\include{figures/examples-macros}
 \include{figures/rotation/rotation}
@@
 $\bvec{\theta}$ and $\bvec{\alpha}$ respectively.
-Under this model, the log-likelihood of observing a dataset
+Under this model, \response{the log-likelihood function $\mathcal{L}
+(\bvec{\theta},\,\bvec{\alpha})$} with a dataset
 \begin{eqnarray}
@@
 \begin{eqnarray}\eqlabel{gp-likelihood}
+\ln \mathcal{L} (\bvec{\theta},\,\bvec{\alpha}) =
 \ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
@@
 While this isn't necessarily a problem for the inferred results, in this
-section we discuss an exact physical interpretation of the \celeriteterm\ kernel
-that will be applicable to many astrophysical systems but especially the
-time-variability of stars.
+section we discuss an exact physical interpretation of the \celeriteterm\
+kernel that will be applicable to many astrophysical systems but especially
+the time-variability of stars.
 
@@
 To relate this physical picture to \celeriteterm, we consider the dynamics of a
-stochastically-driven damped simple harmonic oscillator.
+stochastically-driven damped simple harmonic oscillator (SHO).
 The differential equation for this system is
@@
 \begin{equation}\eqlabel{sho-kernel}
-k(\tau) = S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
+k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_0) =
+S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
 \begin{cases}
@@
 the power spectrum of a process when the data are generated from a
-\celeriteterm\
-model.
+\celeriteterm\ model.
 In the second example, we demonstrate that \celerite\ can be used as an
@@
 
+\response{In each example, the joint posterior probability density is given by
+\begin{eqnarray}\eqlabel{joint-post}
+p(\bvec{\theta},\,\bvec{\alpha}\,|\,\bvec{y},\,X) \propto
+p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})\,
+p(\bvec{\theta},\,\bvec{\alpha})
+\end{eqnarray}
+where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
+likelihood defined in \eq{gp-likelihood} and
+$p(\bvec{\theta},\,\bvec{\alpha})$ is the joint prior probability density for
+the parameters.
+The specific parameters and priors are discussed in detail in each section
+but, in most cases, we assume separable uniform priors with plausibly broad
+support.
+}
+
 \subsection{Recovery of a celerite process}
@@
 made in this idealized case.
-The simulated dataset is shown in the left panel of \Figure{simulated-correct}
-and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
-$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
-The true PSD is shown as a dashed line in the right panel of
+\response{We simulated a small ($N = \exampleindata$) dataset using a GP model
+with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
+1\,\mathrm{ppm}^2$, $\omega_0 = e^2\,\mathrm{day}^{-1}$, and $Q = e^2$, where
+the units are arbitrarily chosen to simplify the discussion.
+We add further white noise with the amplitude $\sigma_n = 2.5\, \mathrm{ppm}$
+to each data point.
+The simulated data are plotted in the top left panel of
 \Figure{simulated-correct}.
-We applied log-uniform priors to all of the parameters and used \emcee\
-\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
-the marginalized posterior inference of the PSD.
-This inference is shown in the right panel of \Figure{simulated-correct} as a
-blue contour indicating 68\% of the posterior mass.
+
+In this case, when simulating and fitting, we set the mean function
+$\mu_\bvec{\theta}$ to zero~--~this means that the parameter vector
+$\bvec{\theta}$ is empty~--~and the elements of the covariance matrix are
+given by \eq{sho-kernel} with three parameters $\bvec{\alpha} = (S_0,\,
+\omega_0,\,Q)$.
+We choose a proper separable prior for $\bvec{\alpha}$ with log-uniform
+densities for each parameter as listed in Table~\ref{tab:example-1-params}.
+
+To start, we estimate the maximum \emph{a posteriori} (MAP) parameters using
+the \project{L-BFGS-B} non-linear optimization routine \citep{Byrd:1995,
+Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
+The top left panel of \Figure{simulated-correct} shows the conditional mean
+and standard deviation of the MAP model over-plotted on the simulated data and
+the bottom panel shows the residuals away from this model.
+Since we are using the correct model to fit the data, it is reassuring that
+the residuals appear qualitatively uncorrelated in this figure.
+
+We then sample the joint posterior probability (\eqalt{joint-post}) using
+\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
+We initialize \exampleinwalkers~walkers in a Gaussian ball with width
+$10^{-4}$ in each parameter around the MAP parameter vector, run
+\exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
+To asses convergence, we estimate the mean integrated autocorrelation time of
+the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
+chain results in \exampleineff~effective samples.
+
+Each sample in the chain corresponds to a model PSD and we compare this
+posterior constraint on the PSD to the true spectral density in the
+right-hand panel of \Figure{simulated-correct}.
+In this figure, the true PSD is plotted as a dashed black line and the
+numerical estimate of the posterior inference is shown as a blue contour
+indicating 68\% of the MCMC samples.
 It is clear from this figure that, as expected, the inference correctly
 reproduces the true PSD.
+}
 
@@
 
-For this example, we simulate a dataset using a known GP model with a kernel
+In this example, we simulate a dataset using a known GP model with a kernel
 outside of the support of a \celeriteterm\ process.
@@
 
-For this example, the data are simulated from a quasiperiodic GP with the
-kernel
+The data are simulated from a quasiperiodic GP with the kernel
 \begin{eqnarray}\eqlabel{sim-wrong-true}
@@
 This autocorrelation structure corresponds to the power spectrum
+\citep{Wilson:2013}
 \begin{eqnarray}
@@
 $\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
-perfectly reproduce the structure of this process.
-That being said, we demonstrate that rigorous inferences can be made
-about $P_\mathrm{true}$ even with an effective model.
-The left panel of \Figure{simulated-wrong} shows the simulated dataset.
+exactly reproduce the structure of this process.
+That being said, we demonstrate that robust inferences can be made about
+$P_\mathrm{true}$ even with this effective model.
+
+\response{We generate a realization of a GP model with the kernel given in
+\eq{sim-wrong-true} with $N = \exampleiindata$, $\alpha = 1\,\mathrm{ppm}^2$,
+$\lambda = 5\,\mathrm{day}$, and $P_\mathrm{true} = 1\,\mathrm{day}$.
+We also add white noise with amplitude $\sigma = 0.5\,\mathrm{ppm}$ to each
+data point.
+The top left panel of \Figure{simulated-wrong} shows this simulated dataset.
+
 We then fit this simulated data using the product of two SHO terms
 (\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
-1/\sqrt{2}$ and the other has $\omega_0 = 2\,\pi/P$.
-We note that using \eq{product-rule}, the product of two \celeriteterm\ terms can
-also be expressed using \celerite.
-We apply log-uniform priors on all the parameters and use \emcee\ to sample
-the joint posterior probability for all of the parameters.
-The inferred distribution for the parameter $P$ is shown in the right panel of
-\Figure{simulated-wrong} and compared to the true period $P_\mathrm{true}$ and
-the inferences made using the correct model (\eqalt{sim-wrong-true}).
-The inference made using this effective \celeriteterm\ model are indistinguishable
-from the inferences made using the correct model and substantially less
-computation time is required for the \celerite\ inference.
+1/\sqrt{2}$ fixed.
+The kernel for this model is
+\begin{eqnarray}
+k(\tau) = k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_1) \,
+    k_\mathrm{SHO}(\tau;\,S_0 = 1,\,Q = 1/\sqrt{2},\,\omega_2)
+\end{eqnarray}
+where $k_\mathrm{SHO}$ is defined in \eq{sho-kernel} and the inferred period
+is related to the parameters by $P=2\,\pi/\omega_1$.
+We note that using \eq{product-rule}, the product of two \celeriteterm\ terms
+can also be expressed using \celerite.
+
+As in the previous example, we set the mean function to zero and can,
+therefore, omit the parameters $\bvec{\theta}$.
+Table~\ref{tab:example-2-params} lists the proper log-uniform priors that we
+choose for each parameter in $\bvec{\alpha} =
+(S_0,\,Q,\,\omega_1,\,\omega_2)$.
+These priors, together with the GP likelihood (\eqalt{gp-likelihood}) fully
+specify the posterior probability density.
+
+As above, we estimate the MAP parameters using \project{L-BFGS-B} and sample
+the posterior probability density using \emcee.
+The top left panel of \Figure{simulated-wrong} shows the conditional mean and
+standard deviation of the MAP model.
+The bottom left panel shows the residuals between the data and this MAP model
+and, even though this GP model is formally ``wrong'', there are no obvious
+correlations in these residuals.
+
+To perform posterior inference, we initialize \exampleiinwalkers~walkers by
+sampling from the 4-dimensional Gaussian centered on the MAP parameters with
+an isotropic standard deviation of $10^{-4}$.
+We then run \exampleiinburn~steps of burn-in and \exampleiinsteps~steps of
+MCMC.
+To estimate the number of effective samples, we estimate the integrated
+autocorrelation time of the chain for the parameter $\omega_1$~--~the
+parameter of primary interest~--~and find \exampleiineff~effective samples.
+
+For comparison, we run the same number of steps of MCMC to sample the
+``correct'' joint posterior density.
+For this reference inference, we use a GP likelihood with the kernel given by
+\eq{sim-wrong-true} and choose log-uniform priors on each of the
+three parameters $\ln \alpha/\mathrm{ppm}^2 \sim \mathcal{U}(-10,\,10)$,
+$\ln \lambda/\mathrm{day} \sim \mathcal{U}(-10,\,10)$, and
+$\ln P_\mathrm{true}/\mathrm{day} \sim \mathcal{U}(-10,\,10)$.
+
+The marginalized posterior inferences of the characteristic period of the
+process are shown in the right panel of \Figure{simulated-wrong}.
+The inference using the correct model is shown as a dashed blue histogram and
+the inference made using the effective model is shown as a solid black
+histogram.
+These inferences are consistent with each other and with the true period used
+for the simulation (shown as a vertical gray line).
+This demonstrates that, in this case, \celerite\ can be used as a
+computationally efficient effective model and hints that this may be true in
+other problems as well.
+}
 

### 2017-05-31 paper/ms.tex
added Guggenheim; boldface X?

--- 2017-05-27 paper/ms.tex
@@
 
+\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\guggenheim}{Guggenheim Fellow}
+
 \edef \columbia {\arabic{affilcounter}}\stepcounter{affilcounter}
@@
     Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw},
-    Eric~Agol\altaffilmark{\uw},
+    Eric~Agol\altaffilmark{\uw,\guggenheim},
     Ruth~Angus\altaffilmark{\simons,\columbia}, and
@@
 motivation for and interpretation of this choice~--~but we also demonstrate
-that it is effective in many other cases.
+that it is effective in other cases.
 We present a mathematical description of the method, the details of the
@@
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
-$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
+$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \response{(Why
+isn't X boldface?)} can be found by
 maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and

### 2017-05-31 paper/ms.tex
updating to latest version
Flagged variables that are reused (which can lead to confusion of reader).  Discussed real terms.

@@
 
-\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
+\edef \guggenheim {\arabic{affilcounter}}\stepcounter{affilcounter}
 \altaffiltext{\guggenheim}{Guggenheim Fellow}
@@
 \end{eqnarray}
-that we can solve for $X$.
+that we can solve for $X$. \response{X was used to represent independent
+parameter matrix in definition of likelihood.}
 Setting each element on the left side of \eq{cholesky} equal to the
@@
 $\mathcal{O}(N\,J^2)$.
+
+Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
+then these terms only add one to the rank of $U$ and $V$.  We use this fact
+to simplify the algebra when real components are included:  if $J_{real}$
+and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then 
+the $U$, $V$, and $X$ matrices have size $N \times (J_{real}+2J_{complex})$.
 
@@
 \end{eqnarray}
-which, for large values of $\omega$, falls off exponentially.
+which, for large values of $\omega$, falls off exponentially. \response{$\alpha$
+means the parameter vector in definition of likelihood.}
 When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
@@
 \end{eqnarray}
-for $a>0$, $b>0$, and $c>0$.
+for $a>0$, $b>0$, and $c>0$. \response{a,b,c already have different meaning
+in celerite kernel.}
 The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
@@
 where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters.
+are shared nuisance parameters. \response{A was used for diagonal of
+covariance matrix.}
 This model could be extended to include more angular modes and
@@
 \newpage
+
+\response{Table 5 has two parameter with name "b":  the celerite coefficient,
+and the impact parameter.}
+
 \input{figures/transit/transit-params}
@@
 \subsection{Summary}
+
+\response{The D in Table 6 may be confused with the D used in the Cholesky solver.}
 

### 2017-05-31 paper/ms.tex

--- 2017-05-31 paper/ms.tex
@@
 dimensional data sets.
-\response{This method is a generalization of a method developed by
+\todo{This method is a generalization of a method developed by
 \citet{Ambikasaran:2015}
@@
 matrix when it is generated by a mixture of exponentials.
-\response{More precisely, the matrix can be embedded in an equivalent higher-dimensional
+\todo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
 system that can be solved efficiently using a sparse or banded solver. Replace
@@
 In the following pages, we motivate the general problem of GP modeling,
-describe the \response{(delete) previously published} scalable method
+describe the \todo{(delete) previously published} scalable method
 \citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
@@
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
-$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \response{(Why
+$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \todo{(Why
 isn't X boldface?)} can be found by
@@
 \end{eqnarray}
-that we can solve for $X$.
+that we can solve for $X$. \todo{X was used to represent independent
+parameter matrix in definition of likelihood.}
 Setting each element on the left side of \eq{cholesky} equal to the
@@
 \end{eqnarray}
-which, for large values of $\omega$, falls off exponentially.
+which, for large values of $\omega$, falls off exponentially. \todo{$\alpha$
+means the parameter vector in definition of likelihood.}
 When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
@@
 \end{eqnarray}
-for $a>0$, $b>0$, and $c>0$.
+for $a>0$, $b>0$, and $c>0$. \todo{a,b,c already have different meaning
+in celerite kernel.}
 The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
@@
 where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters.
+are shared nuisance parameters. \todo{A was used for diagonal of
+covariance matrix.}
 This model could be extended to include more angular modes and
@@
 \newpage
+
+\todo{Table 5 has two parameter with name "b":  the celerite coefficient,
+and the impact parameter.}
+
 \input{figures/transit/transit-params}
@@
 \subsection{Summary}
+
+\todo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
 

### 2017-05-31 paper/ms.tex
one examples section

--- 2017-05-31 paper/ms.tex
@@
 
-\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\guggenheim}{Guggenheim Fellow}
-
 \edef \columbia {\arabic{affilcounter}}\stepcounter{affilcounter}
@@
     Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw},
-    Eric~Agol\altaffilmark{\uw,\guggenheim},
+    Eric~Agol\altaffilmark{\uw},
     Ruth~Angus\altaffilmark{\simons,\columbia}, and
@@
 motivation for and interpretation of this choice~--~but we also demonstrate
-that it is effective in other cases.
+that it is effective in many other cases.
 We present a mathematical description of the method, the details of the
@@
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
-$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \response{(Why
-isn't X boldface?)} can be found by
+$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
 maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
@@
 
-\section{Examples with simulated data}
-
-To demonstrate the application of \celerite, we start by inferring
+\section{Examples}
+
+To demonstrate the application of \celerite, we use it to perform posterior
+inference in two examples with realistic simulated data and three examples of
+real-world research problems.
+
+start by inferring
 posterior constraints on the parameters of a GP model applied to several
 simulated datasets with known properties.
+
 In the following section, we expand on these examples by applying
@@
 provides computational advantages over other GP methods.
+
+In this section, we demonstrate several use cases of \celerite\ when
+applied to real datasets.
+Each of these examples touches on an active area of research so we limit our
+examples to be qualitative in nature and do not claim that \celerite\ is the
+optimum method, but we hope these examples encourage interested readers to
+investigate the applicability of \celerite\ to their research.
+
+All of the following examples show time domain datasets with a clear bias in
+favor of large homogeneous photometric surveys, but these methods can
+similarly be applied to spectroscopy, where wavelength~--~instead of
+time~--~is the independent coordinate and other one-dimensional domains
+\citep[see][for a potential application]{Czekala:2017}.
 
@@
 \end{figure}
-
-\section{Examples with real data} \sectlabel{examples-real}
-
-In this section, we demonstrate several use cases of \celerite\ when
-applied to real datasets.
-Each of these examples touches on an active area of research so we limit our
-examples to be qualitative in nature and do not claim that \celerite\ is the
-optimum method, but we hope these examples encourage interested readers to
-investigate the applicability of \celerite\ to their research.
-
-All of the following examples show time domain datasets with a clear bias in
-favor of large homogeneous photometric surveys, but these methods can
-similarly be applied to spectroscopy, where wavelength~--~instead of
-time~--~is the independent coordinate and other one-dimensional domains
-\citep[see][for a potential application]{Czekala:2017}.
 

### 2017-05-31 paper/ms.tex
fixing todos macros

@@
 dimensional data sets.
-\todo{This method is a generalization of a method developed by
+\dfmtodo{This method is a generalization of a method developed by
 \citet{Ambikasaran:2015}
@@
 matrix when it is generated by a mixture of exponentials.
-\todo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
+\dfmtodo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
 system that can be solved efficiently using a sparse or banded solver. Replace
@@
 In the following pages, we motivate the general problem of GP modeling,
-describe the \todo{(delete) previously published} scalable method
+describe the \dfmtodo{(delete) previously published} scalable method
 \citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
@@
 The maximum likelihood values for the parameters $\bvec{\theta}$ and
-$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ \todo{(Why
-isn't X boldface?)} can be found by
+$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
 maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
@@
 \end{eqnarray}
-that we can solve for $X$. \todo{X was used to represent independent
+that we can solve for $X$. \dfmtodo{X was used to represent independent
 parameter matrix in definition of likelihood.}
@@
 \end{eqnarray}
-which, for large values of $\omega$, falls off exponentially. \todo{$\alpha$
+which, for large values of $\omega$, falls off exponentially. \dfmtodo{$\alpha$
 means the parameter vector in definition of likelihood.}
@@
 \end{eqnarray}
-for $a>0$, $b>0$, and $c>0$. \todo{a,b,c already have different meaning
+for $a>0$, $b>0$, and $c>0$. \dfmtodo{a,b,c already have different meaning
 in celerite kernel.}
@@
 where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters. \todo{A was used for diagonal of
+are shared nuisance parameters. \dfmtodo{A was used for diagonal of
 covariance matrix.}
@@
 
-\todo{Table 5 has two parameter with name "b":  the celerite coefficient,
+\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
 and the impact parameter.}
@@
 
-\todo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
+\dfmtodo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
 

### 2017-05-31 paper/ms.tex
Added an example of the semi-separable matrix decomposition

@@
 component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.
+triangular operation.  Note that the last two terms in this equation are
+transposes of one another.
 For the \celeriteterm\ model described in the previous section, the rank of
@@
 \end{eqnarray}
+
+To help clarify these equations, we write out here an example of a semi-separable matrix with a 
+single \celeriteterm\ term with $N=3$ and $c=d=1$:
+\begin{eqnarray}\eqlabel{semi-separable-structure}
+K &=& \\
+&&\begin{pmatrix}
+%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
+%    {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}\\
+%    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
+    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
+    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
+    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
+\end{pmatrix}\\
+&=&
+\begin{pmatrix}
+    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
+    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
+    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
+\end{pmatrix}\\
+&+&\mathrm{tril}\left[\begin{pmatrix}
+%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}\\
+%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}\\
+%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
+    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
+    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
+    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
+    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
+    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
+    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
+\end{pmatrix} %\right.\\
+%&\times &
+\times
+%\left.
+\begin{pmatrix}
+% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} &{\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}}\\
+% {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
+% {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_3} \cos{t_3}}\\
+% {\scriptscriptstyle e^{t_1} \sin{t_1}} & {\scriptscriptstyle e^{t_2} \sin{t_2}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
+ {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
+ {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
+ {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
+\end{pmatrix}^T\right]\\
+&+&
+\mathrm{triu}\left[\begin{pmatrix}
+% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} & {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} \\
+% {\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}}\\
+% {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
+ {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
+ {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
+ {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
+\end{pmatrix} % \right.\\
+\times
+%\left.
+\begin{pmatrix}
+%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}\\
+%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}&
+%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
+%    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
+%    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
+%    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}\\
+%    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}&
+%    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}&
+%    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
+    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
+    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
+    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
+    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
+    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
+    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
+\end{pmatrix}^T\right]\\
+&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
+\end{eqnarray}
+Note that rather than storing the full $N \times N$ covariance matrix, it
+is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
+as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
+point numbers.
 

### 2017-05-31 paper/ms.tex
Added text in simulation section.

@@
 \begin{eqnarray}\eqlabel{semi-separable-structure}
-K &=& \\
-&&\begin{pmatrix}
+K &=&
+\begin{pmatrix}
 %    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
@@
 
+Simulation of a Gaussian process for large $N$ requires multiplying an
+$N$-dimensional vector of Gaussian normal deviates, $\bf{g}$, by the square root
+of the covariance matrix:
+\begin{equation}
+{\bf y} = K_\bvec{\alpha}^{1/2} {\bf g},
+\end{equation}
+where $g_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
+comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
+the likelihood with $\mu_\theta =0$ is given by
+\begin{eqnarray}
+Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
+&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
+&=& -\frac{1}{2}{\bf g}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf g}\\
+&=& -\frac{1}{2}{\bf g}^T {\bf g}.
+\end{eqnarray}
+The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
+
+To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
+low-rank Cholesky decomposition first, generating the vector ${\bf g}$ of random normal deviates, and then multiplying
+\begin{equation}
+{\bf y} = L D^{1/2} {\bf g}.
+\end{equation}
+This may be carried out in $O(JN)$ operations with
+\begin{eqnarray}
+f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} g_{n-1}\right)\\
+y_n &=& D_{n,n}^{1/2} g_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
+\end{eqnarray}
+for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $g_0 = 0$.
+We expect that this simulation algorithm will be useful for testing Gaussian
+process inference algorithms with large datasets, which becomes prohibitive
+for large $N$ if standard Cholesky decomposition and multiplication are used.
+
+\subsection{Prediction}
+
 \section{celerite as a model of stellar variations}\sectlabel{sho}

### 2017-05-31 paper/ms.tex

@@
 for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $g_0 = 0$.
+\agoltodo{Check that these equations are in fact correct.}
 We expect that this simulation algorithm will be useful for testing Gaussian
@@
 
+Multiplication of the full covariance matrix times a vector can be accomplished
+by yet another multiplication by the transpose of the Cholesky decomposition.
+\dfmtodo{Should we spell this out as well?}
+
 \subsection{Prediction}
+
+\agoltodo{Write up the prediction algorithm.}
 

### 2017-06-01 paper/ms.tex

--- 2017-05-31 paper/ms.tex
@@
 
-\subsection{Cholesky factorization}
-
 The relationship between the \celeriteterm\ model and semi-separable linear
@@
 To help clarify these equations, we write out here an example of a semi-separable matrix with a 
-single \celeriteterm\ term with $N=3$ and $c=d=1$:
+single \celeriteterm\ term with $c=d=1$: %and $N=3$ 
 \begin{eqnarray}\eqlabel{semi-separable-structure}
-K &=& \\
-&&\begin{pmatrix}
+K&=&%
+\begin{pmatrix}
 %    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
@@
 %    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
-    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
-    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
-    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
+%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
+%    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
+%    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
+    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
+    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
+    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
+    \ldots & \ldots & \ldots & \ldots
 \end{pmatrix}\\
@@
 \begin{pmatrix}
-    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
+%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
+%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
+%    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
+    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
+    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
+    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
+%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & \ldots \\
+%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& \ldots \\
+    \ldots & \ldots & \ldots & \ldots
 \end{pmatrix}\\
@@
     {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
+    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
+   \ldots & \ldots
 \end{pmatrix} %\right.\\
@@
  {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
+ {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
+ \ldots & \ldots
 \end{pmatrix}^T\right]\\
@@
  {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
+ {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
+ \ldots & \ldots
 \end{pmatrix} % \right.\\
@@
     {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
+    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
+   \ldots & \ldots
 \end{pmatrix}^T\right]\\
@@
 point numbers.
+
+\subsection{Cholesky factorization}
 
@@
 overflow and underflow issues in many realistic cases because both $U$ and $V$
-have factors of the form $e^{\pm t}$, where $t$ can be any arbitrarily large
+have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
 real number.
@@
 \subsection{Simulation}
+
+Simulation of a Gaussian process for large $N$ requires multiplying an
+$N$-dimensional vector of Gaussian normal deviates, $\bf{q}$, by the square root
+of the covariance matrix:
+\begin{equation}
+{\bf y} = K_\bvec{\alpha}^{1/2} {\bf q},
+\end{equation}
+where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
+comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
+the likelihood with $\mu_\bvec{\theta} =0$ is given by
+\begin{eqnarray}
+Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
+&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
+&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
+&=& -\frac{1}{2}{\bf q}^T {\bf q}.
+\end{eqnarray}
+The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
+
+To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
+low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
+\begin{equation}
+{\bf y} = L D^{1/2} {\bf q}.
+\end{equation}
+This may be carried out in $O(JN)$ operations with
+\begin{eqnarray}
+f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} q_{n-1}\right)\\
+y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
+\end{eqnarray}
+for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
+\agoltodo{Check that these equations are in fact correct.}
+We expect that this simulation algorithm will be useful for test-driving Gaussian
+process inference algorithms applied to large datasets; even simulation of
+Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition 
+and multiplication are used.
+
+Multiplication of the full covariance matrix times a vector can be accomplished
+by yet another multiplication by the transpose of the Cholesky decomposition.
+\dfmtodo{Should we spell this out as well?}
+
+\subsection{Prediction}
+
+\agoltodo{Write up the prediction algorithm.}
 

### 2017-06-02 paper/ms.tex
Adding prediction section.  Needs work.

@@
 by yet another multiplication by the transpose of the Cholesky decomposition.
-\dfmtodo{Should we spell this out as well?}
-
-\subsection{Prediction}
-
-\agoltodo{Write up the prediction algorithm.}
+\dfmtodo{Should we spell this out as well?  This is useful, for example, for
+filter out a particular component of variability.}
+
+\subsection{Prediction and interpolation}
+
+%Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.
+With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $O(N+M)$ operations.
+
+For predicting or interpolating missing or future data points (and their variance) at times ${\bf t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.
+The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
+\begin{eqnarray}
+\bar f_* &=& K({\bf t_*},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
+&=& K({\bf t_*},{\bf t}) {\bf b}
+\end{eqnarray}
+where ${\bf y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
+${\bf b} = \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y}$ can be computed from
+the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.
+
+Since the covariance matrix may be written as the sum of exponentials, we can write:
+\begin{eqnarray}
+\bar f_{*,i} &=& K(t_{*,i},{\bf t}) {\bf b} \\
+&=& \sum_n \sum_p \alpha_p e^{-\beta_p \vert t_{*,i}-t_n\vert} b_j.
+\end{eqnarray}
+\agoltodo{Convert this to real, \celerite\ notation .}
+Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
+$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
+\begin{equation}
+\bar f_{*,i} = \sum_{n=1}^{n_0} \sum_p \alpha_p e^{-\beta_p( t_{*,i}-t_n)} b_n + \sum_{n=n_0+1}^N \sum_p \alpha_p e^{-\beta_p(t_n-t_{*,i})} b_n.
+\end{equation}
+Each of these double-sums may have the sums interchanged, giving:
+\begin{eqnarray}
+\bar f_{*,i} &=& \sum_p e^{-\beta_p ( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} \alpha_p e^{-\beta_p ( t_{*,i-1}-t_n) } b_n \\
+&+&
+\sum_p e^{-\beta_p ( t_{*,i+1}-t_{*,i})} \sum_{n=n+0+1}^{N} \alpha_p e^{-\beta_p ( t_n-t_{*,i+1}) } b_n
+\end{eqnarray}
+We can compute this recursively.
+Defining $Q^-_{i,n_0,p} = \sum_{j=1}^{n_0} \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ and defining $Q^+_{i,n_0,p} = \sum_{j=n_0}^N \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ gives:
+\begin{equation}
+\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
+\end{equation}
+Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $O(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx O(JM)$ operations
+\begin{eqnarray}
+Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
+Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
+\end{eqnarray} 
+for a total scaling of $\approx O(J^2N+J(N+M))$.
+
+The variance evaluated at future predicted times ${\bf t_*} > {\bf t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
+\begin{eqnarray}
+\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},{\bf t}^T)K({\bf t},{\bf t})^{-1} K({\bf t},t_{*,i+1})\\
+&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
+&\times& e^{-\beta_p(t_{*,i}-{\bf t}^T)}K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}\\
+&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
+\end{eqnarray}
+for $1 \le p,q \le J$,
+where
+\begin{equation}
+P_{i,p,q} = e^{-\beta_p(t_{*,i}-{\bf t}^T)}
+K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}.
+\end{equation}
+For $i=1$, $P_{1,p,q}$ can be obtained in $O(J^2N)$ operations
+with the Cholesky decomposition.
+Subsequent $P_{i+1,p,q}$ may be obtained recursively as
+\begin{equation}
+P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
+\end{equation}
+in $O(J^2M)$ operations, for a total of $\approx O(J^2(N+M))$ operations.
+Note that this variance formula {\it only} works for prediction; we have
+yet to derive a formula for which ${\bf t_*}$ and ${\bf t}$ are
+interspersed.
 

### 2017-06-02 paper/ms.tex
renaming the impact parameter

--- 2017-05-31 paper/ms.tex
@@
 component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.  Note that the last two terms in this equation are
-transposes of one another.
+triangular operation.
 For the \celeriteterm\ model described in the previous section, the rank of
@@
 \end{eqnarray}
-
-To help clarify these equations, we write out here an example of a semi-separable matrix with a 
-single \celeriteterm\ term with $N=3$ and $c=d=1$:
-\begin{eqnarray}\eqlabel{semi-separable-structure}
-K &=&
-\begin{pmatrix}
-%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
-%    {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}\\
-%    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
-    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
-    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
-    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
-\end{pmatrix}\\
-&=&
-\begin{pmatrix}
-    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
-\end{pmatrix}\\
-&+&\mathrm{tril}\left[\begin{pmatrix}
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
-    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
-    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
-    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
-\end{pmatrix} %\right.\\
-%&\times &
-\times
-%\left.
-\begin{pmatrix}
-% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} &{\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}}\\
-% {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
-% {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_3} \cos{t_3}}\\
-% {\scriptscriptstyle e^{t_1} \sin{t_1}} & {\scriptscriptstyle e^{t_2} \sin{t_2}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
- {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
- {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
-\end{pmatrix}^T\right]\\
-&+&
-\mathrm{triu}\left[\begin{pmatrix}
-% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} & {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} \\
-% {\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}}\\
-% {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
- {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
- {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
-\end{pmatrix} % \right.\\
-\times
-%\left.
-\begin{pmatrix}
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
-%    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-%    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-%    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}\\
-%    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}&
-%    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}&
-%    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
-    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
-    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
-    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
-\end{pmatrix}^T\right]\\
-&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
-\end{eqnarray}
-Note that rather than storing the full $N \times N$ covariance matrix, it
-is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
-as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
-point numbers.
 
@@
 \subsection{Simulation}
-
-Simulation of a Gaussian process for large $N$ requires multiplying an
-$N$-dimensional vector of Gaussian normal deviates, $\bf{g}$, by the square root
-of the covariance matrix:
-\begin{equation}
-{\bf y} = K_\bvec{\alpha}^{1/2} {\bf g},
-\end{equation}
-where $g_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
-comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
-the likelihood with $\mu_\theta =0$ is given by
-\begin{eqnarray}
-Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
-&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
-&=& -\frac{1}{2}{\bf g}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf g}\\
-&=& -\frac{1}{2}{\bf g}^T {\bf g}.
-\end{eqnarray}
-The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
-
-To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
-low-rank Cholesky decomposition first, generating the vector ${\bf g}$ of random normal deviates, and then multiplying
-\begin{equation}
-{\bf y} = L D^{1/2} {\bf g}.
-\end{equation}
-This may be carried out in $O(JN)$ operations with
-\begin{eqnarray}
-f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} g_{n-1}\right)\\
-y_n &=& D_{n,n}^{1/2} g_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
-\end{eqnarray}
-for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $g_0 = 0$.
-\agoltodo{Check that these equations are in fact correct.}
-We expect that this simulation algorithm will be useful for testing Gaussian
-process inference algorithms with large datasets, which becomes prohibitive
-for large $N$ if standard Cholesky decomposition and multiplication are used.
-
-Multiplication of the full covariance matrix times a vector can be accomplished
-by yet another multiplication by the transpose of the Cholesky decomposition.
-\dfmtodo{Should we spell this out as well?}
-
-\subsection{Prediction}
-
-\agoltodo{Write up the prediction algorithm.}
 
@@
 $\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
-$t_0$, an impact parameter $b$, the radius of the planet in units of the
+$t_0$, an impact parameter $B$, the radius of the planet in units of the
 stellar radius $R_P/R_\star$, and several parameters describing the

### 2017-06-02 paper/ms.tex
Minor edit

--- 2017-06-02 paper/ms.tex
@@
 that we can solve for $X$. \dfmtodo{X was used to represent independent
-parameter matrix in definition of likelihood.}
+parameter matrix in definition of likelihood. Perhaps we should use W?}
 Setting each element on the left side of \eq{cholesky} equal to the

### 2017-06-02 paper/ms.tex
words about rotation and turning off omp for mkl

--- 2017-06-02 paper/ms.tex
@@
 \makeatletter
-\def\env@matrix{\hskip -\arraycolsep % taken from amsmath.sty lines 895ff
-  \let\@ifnextchar\new@ifnextchar
-  \array{*{\c@MaxMatrixCols}c}}
+\def\env@matrix{\hskip -\arraycolsep\let\@ifnextchar\new@ifnextchar\array{*{\c@MaxMatrixCols}c}}
 \makeatother
@@
 \newcommand{\kepler}{\project{Kepler}}
+\newcommand{\lsst}{\project{LSST}}
 \newcommand{\tess}{\project{TESS}}
@@
 
+\dfmtodo{Talk about LSST here too!}
+
 In the astrophysical literature, Gaussian Processes
@@
 In particular, the measurements don't need to be evenly spaced and the
-uncertainties can be heteroscedastic (non-uniform).
+uncertainties can be heteroscedastic.
 
@@
 
-To demonstrate the application of \celerite, we use it to perform posterior
+\response{
+To show the application of \celerite, we use it to perform posterior
 inference in two examples with realistic simulated data and three examples of
 real-world research problems.
-
-start by inferring
-posterior constraints on the parameters of a GP model applied to several
-simulated datasets with known properties.
-
-In the following section, we expand on these examples by applying
-\celerite\ to real datasets.
+These examples have been chosen to justify the application of \celerite\ for a
+range of use cases, but they are by no means exhaustive.
+These examples are framed in the time domain with a clear bias in favor of
+large homogeneous photometric surveys, but \celerite\ can also be used for
+other one-dimensional problems like, for example, spectroscopy, where
+wavelength~--~instead of time~--~is the independent coordinate \citep[see][for
+a potential application]{Czekala:2017}.
+
 In the first example, we demonstrate that \celerite\ can be used to infer
@@
 
-In this section, we demonstrate several use cases of \celerite\ when
-applied to real datasets.
+The next three examples show \celerite\ used to infer the properties of stars
+and exoplanets observed by NASA's \kepler\ Mission.
 Each of these examples touches on an active area of research so we limit our
 examples to be qualitative in nature and do not claim that \celerite\ is the
-optimum method, but we hope these examples encourage interested readers to
+optimal method, but we hope these examples encourage interested readers to
 investigate the applicability of \celerite\ to their research.
 
-All of the following examples show time domain datasets with a clear bias in
-favor of large homogeneous photometric surveys, but these methods can
-similarly be applied to spectroscopy, where wavelength~--~instead of
-time~--~is the independent coordinate and other one-dimensional domains
-\citep[see][for a potential application]{Czekala:2017}.
-
-\response{In each example, the joint posterior probability density is given by
+In each example, the joint posterior probability density is given by
 \begin{eqnarray}\eqlabel{joint-post}
@@
 where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
-likelihood defined in \eq{gp-likelihood} and
-$p(\bvec{\theta},\,\bvec{\alpha})$ is the joint prior probability density for
-the parameters.
-The specific parameters and priors are discussed in detail in each section
-but, in most cases, we assume separable uniform priors with plausibly broad
-support.
+likelihood defined in \eq{gp-likelihood} and $p(\bvec{\theta},\,\bvec{\alpha})$
+is the joint prior probability density for the parameters.
+We sample each posterior density using MCMC and investigate the performance of
+the model when used to interpret the physical processes that generated the
+data.
+The specific parameters and priors are discussed in detail in each section,
+but we generally assume separable uniform priors with plausibly broad support.
 }
 
-\subsection{Recovery of a celerite process}
+\subsection{Example 1: Recovery of a celerite process}
 
 In this first example, we simulate a dataset using a known \celeriteterm\
-process and fit it with \celerite\ to demonstrate that valid inferences can be
-made in this idealized case.
+process and fit it with \celerite\ to show that valid inferences can be made
+in this idealized case.
 \response{We simulated a small ($N = \exampleindata$) dataset using a GP model
@@
 \exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
-To asses convergence, we estimate the mean integrated autocorrelation time of
+To assess convergence, we estimate the mean integrated autocorrelation time of
 the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
@@
 
-\newpage
-\subsection{Inferences with the ``wrong'' model}
+\subsection{Example 2: Inferences with the ``wrong'' model}
 
@@
 \begin{eqnarray}\eqlabel{sim-wrong-true}
-k_\mathrm{true} (\tau) = \alpha\,
+k_\mathrm{true} (\tau) = A\,
     \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
@@
 \begin{eqnarray}
-S_\mathrm{true} (\omega) = \frac{\lambda\,\alpha}{2}\,\left[
+S_\mathrm{true} (\omega) = \frac{\lambda\,A}{2}\,\left[
     \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
@@
 \end{eqnarray}
-which, for large values of $\omega$, falls off exponentially. \dfmtodo{$\alpha$
-means the parameter vector in definition of likelihood.}
+which, for large values of $\omega$, falls off exponentially.
 When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
@@
 \response{We generate a realization of a GP model with the kernel given in
-\eq{sim-wrong-true} with $N = \exampleiindata$, $\alpha = 1\,\mathrm{ppm}^2$,
+\eq{sim-wrong-true} with $N = \exampleiindata$, $A = 1\,\mathrm{ppm}^2$,
 $\lambda = 5\,\mathrm{day}$, and $P_\mathrm{true} = 1\,\mathrm{day}$.
@@
 \end{eqnarray}
-where $k_\mathrm{SHO}$ is defined in \eq{sho-kernel} and the inferred period
-is related to the parameters by $P=2\,\pi/\omega_1$.
+where $k_\mathrm{SHO}$ is defined in \eq{sho-kernel} and the period of the
+process is $P=2\,\pi/\omega_1$.
 We note that using \eq{product-rule}, the product of two \celeriteterm\ terms
@@
 \eq{sim-wrong-true} and choose log-uniform priors on each of the
-three parameters $\ln \alpha/\mathrm{ppm}^2 \sim \mathcal{U}(-10,\,10)$,
+three parameters $\ln A/\mathrm{ppm}^2 \sim \mathcal{U}(-10,\,10)$,
 $\ln \lambda/\mathrm{day} \sim \mathcal{U}(-10,\,10)$, and
@@
 
-\subsection{Stellar rotation} \sectlabel{rotation}
+\subsection{Example 3: Stellar rotation} \sectlabel{rotation}
 
@@
 it is positive for all values of $\tau$.
-We can construct a simple \celeriteterm\ covariance function with similar properties
-as follows
+We construct a simple \celeriteterm\ covariance function with similar
+properties as follows
 \begin{eqnarray}\eqlabel{rot-kernel}
-k(\tau) = \frac{a}{2+b}\,e^{-\tau/c}\,\left[
-    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
+k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
+    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + C)
 \right]
 \end{eqnarray}
-for $a>0$, $b>0$, and $c>0$. \dfmtodo{a,b,c already have different meaning
-in celerite kernel.}
+for $B>0$, $C>0$, and $L>0$.
 The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
 but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
-as a drop-in replacement for a substantial gain in computational efficiency.
-
-To demonstrate this method, we fit a \celeriteterm\ model with a kernel given
+as a drop-in replacement for a significant gain in computational efficiency.
+
+\response{GPs have been used to measure stellar rotation periods for
+individual datasets \citep[for example][]{Littlefair:2017}, but the
+computational cost of traditional GP methods has hindered the industrial
+application to existing surveys like \kepler\ with hundreds of thousands of
+targets.
+The increase in computational efficiency and scalability provided by
+\celerite\ opens the possibility of inferring rotation periods using GPs at
+scale of existing and forthcoming surveys like \kepler, \tess, and \lsst.
+}
+
+As a demonstration, we fit a \celeriteterm\ model with a kernel given
 by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
@@
 applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
+
+\response{We select about 180~days of contiguous observations of
+KIC~1430163 from \kepler.
+This dataset has \exampleiiindata\ measurements and, using a tuned linear
+algebra
+implementation\footnote{We use the Intel Math Kernel Library
+\url{https://software.intel.com/en-us/intel-mkl}}, a
+single evaluation of the likelihood requires over 8~seconds on a modern Intel
+CPU.
+This calculation, using \celerite\ with the model in \eq{rot-kernel}, only
+takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
+magnitude per model evaluation.
+
+We set the mean function $\mu_\bvec{\theta}$ to zero and the remaining
+parameters $\bvec{\alpha}$ and their priors are listed in
+Table~\ref{tab:example-3-params}.
+As with the earlier examples, we start by estimating the MAP parameters using
+\project{L-BFGS-B} and initialize \exampleiiinwalkers~walkers by sampling from
+an isotropic Gaussian with a standard deviation of $10^{-5}$ centered on the
+MAP parameters.
+}
+
 We used \emcee\ to sample the joint posterior probability density for the four
@@
 and find a constraint on the rotation period of $P = \rotationperiod$.
-This is in good agreement with the literature value with somewhat smaller
+This is in good agreement with the literature value with smaller
 uncertainties.
@@
     This is consistent with the published rotation period made using the full
-    \kepler\ baseline \citep{Mathur:2014}.
-    The error bars from the literature are indicated by the two vertical
-    dashed lines.
+    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
     \figurelabel{rotation}}

### 2017-06-02 paper/ms.tex
Added section on multiplication and filtering

--- 2017-06-02 paper/ms.tex
@@
 \begin{eqnarray}
-    L = I + \mathrm{tril} (U\,X^T)
-\end{eqnarray}
-where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
+    L = I + \mathrm{tril} (U\,W^T)
+\end{eqnarray}
+where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
 $N \times 2\,J$ matrix.
@@
 A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
-&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
-    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
-&=& D + \mathrm{tril}(U\,X^\T)\,D
-    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
-&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
-\end{eqnarray}
-that we can solve for $X$. \dfmtodo{X was used to represent independent
-parameter matrix in definition of likelihood. Perhaps we should use W?}
+&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,W^\T)\,D
+    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $W$.
 Setting each element on the left side of \eq{cholesky} equal to the
@@
 \begin{eqnarray}
-S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
+S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
 D_{n,n} &=& A_{n,n} -
@@
     \nonumber\\
-X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
+W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
     \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
@@
 and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
-the $U$, $V$, and $X$ matrices have size $N \times (J_{real}+2J_{complex})$.
+the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
 
@@
 \tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
-\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
-\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
+\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
+\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
 \end{eqnarray}
@@
     S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
-    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
+    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
 D_{n,n} &=& A_{n,n} -
@@
     \nonumber\\
-\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
+\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
     \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
@@
     f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
-    \tilde{X}_{n-1,j}\,z_{n-1}\right] \nonumber\\
+    \tilde{W}_{n-1,j}\,z_{n-1}\right] \nonumber\\
     z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
@@
     \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
-b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{X}_{n,j}\,g_{n,j}
+b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
 \end{eqnarray}
@@
 \end{equation}
-This may be carried out in $O(JN)$ operations with
-\begin{eqnarray}
-f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} q_{n-1}\right)\\
+This may be carried out in $\mathcal{O}(JN)$ operations with
+\begin{eqnarray}
+f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
 y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
@@
 
-Multiplication of the full covariance matrix times a vector can be accomplished
-by yet another multiplication by the transpose of the Cholesky decomposition.
-\dfmtodo{Should we spell this out as well?  This is useful, for example, for
-filter out a particular component of variability.}
+\subsection{Multiplication and filtering}
+
+Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
+operations with a stable recursive algorithm using the semi-separable form:
+\begin{equation}
+{\bf y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) {\bf z}.
+\end{equation}
+The algorithm for multiplication is given by
+\begin{eqnarray}
+f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
+f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
+y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
+\end{eqnarray}
+where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
+two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards 
+from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
+indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}
+
+Multiplication of the full covariance matrix is useful, for example, for isolating
+a particular component of a kernel, which amounts to a customized filter of a dataset.
+For example, a white-noise filter is given by:
+\begin{eqnarray}
+\bar f &=& K({\bf t},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
+&=& K({\bf t},{\bf t}) {\bf b}
+\end{eqnarray}
+where ${\bf y}$ is the vector of measurements and ${\bf b}$ may be computed from the Cholesky
+decomposition as described above.  The second step multiplication of the white-noise-free
+kernel times ${\bf b}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
+Note that for this particular example it is more efficient to compute $\bar f = {\bf y} -\sigma^2 {\bf I} {\bf b}$,
+but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.
 
@@
 %Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.
-With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $O(N+M)$ operations.
+With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $\mathcal{O}(N+M)$ operations.
 
@@
 \end{equation}
-Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $O(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx O(JM)$ operations
+Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
 \begin{eqnarray}
@@
 \end{eqnarray} 
-for a total scaling of $\approx O(J^2N+J(N+M))$.
+for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
 
@@
 \end{equation}
-For $i=1$, $P_{1,p,q}$ can be obtained in $O(J^2N)$ operations
+For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
 with the Cholesky decomposition.
@@
 \end{equation}
-in $O(J^2M)$ operations, for a total of $\approx O(J^2(N+M))$ operations.
+in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
 Note that this variance formula {\it only} works for prediction; we have

### 2017-06-02 paper/ms.tex

--- 2017-06-02 paper/ms.tex
@@
 A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
-&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
-    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
-&=& D + \mathrm{tril}(U\,X^\T)\,D
-    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
-&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
-\end{eqnarray}
-that we can solve for $X$. \dfmtodo{X was used to represent independent
-parameter matrix in definition of likelihood.}
+&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,W^\T)\,D
+    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $W$.
 Setting each element on the left side of \eq{cholesky} equal to the

### 2017-06-02 paper/ms.tex
words about astero example

@@
 
+\edef \guggenheim {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\guggenheim}{Guggenheim Fellow}
+
 \edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
 \altaffiltext{\simons}{Simons Fellow}
-
-\edef \guggenheim {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\guggenheim}{Guggenheim Fellow}
 
@@
 MAP parameters.
-}
-
-We used \emcee\ to sample the joint posterior probability density for the four
-parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
-and find a constraint on the rotation period of $P = \rotationperiod$.
-This is in good agreement with the literature value with smaller
+The left panels of \Figure{rotation} show a subset of the data used in this
+example and the residuals away from the MAP predictive mean.
+
+We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
+of MCMC using \emcee.
+We estimate the integrated autocorrelation time of the chain for $\ln P$ and
+estimate that we have \exampleiiineff\ independent samples.
+These samples give a posterior constraint on the period of $P =
+\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
+in the right panel of \Figure{rotation}.
+This result is in good agreement with the literature value with smaller
 uncertainties.
@@
 more reliable than other methods.
-The left panels of \Figure{rotation} show a subset of the data used in this
-example and the residuals away from the maximum likelihood predictive mean.
-The right panel of \Figure{rotation} shows the marginalized posterior
-distribution for the rotation period.
+}
 
@@
 
-\subsection{Asteroseismic oscillations}
+\subsection{Example 4: Asteroseismic oscillations}
 
 The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
+curves from the \kepler\ Mission \citep{Gilliland:2010, Huber:2011,
 Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
@@
 \citep{Campante:2016, Rauer:2014, Gould:2015}.
-Most asteroseismic analyses have been limited to relatively high
-signal-to-noise oscillations.
-The standard methods use statistics of the empirical periodogram and cannot
-formally propagate the measurement uncertainties to the constraints on
-physical parameters, instead relying on bootstrapped uncertainty estimators
-\citep{Huber:2009}.
+Most asteroseismic analyses have been limited to high signal-to-noise
+oscillations because the standard methods use statistics of the empirical
+periodogram and cannot formally propagate the measurement uncertainties to the
+constraints on physical parameters, instead relying on bootstrapped
+uncertainty estimators \citep{Huber:2009}.
 More sophisticated methods that compute the likelihood function in the time
-domain scale poorly to state-of-the-art datasets \citep{Brewer:2009,
+domain scale poorly to large survey datasets \citep{Brewer:2009,
 Corsaro:2014}.
@@
 probabilistic model that can be evaluated efficiently even for large datasets.
-In practice, one would model the star as a mixture of stochastically-driven
-simple harmonic oscillators where the amplitudes and frequencies of the
-oscillations are computed using a physical model, and evaluate the probability
-of the observed time series using a GP where the PSD is a sum of terms given
-by \eq{sho-psd}.
-This gives us a method of computing the likelihood function for the parameters
-of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
-other more fundamental parameters) \emph{conditioned on the observed time
-series} in $\mathcal{O}(N)$ operations.
+In practice, we model the star as a mixture of stochastically-driven simple
+harmonic oscillators where the amplitudes and frequencies of the oscillations
+are computed using a physical model, and evaluate the probability of the
+observed time series using a GP where the PSD is a sum of terms given by
+\eq{sho-psd}.
+This gives us a method for computing the likelihood function for the
+parameters of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta
+\nu$, or other more fundamental parameters) \emph{conditioned on the observed
+time series} in $\mathcal{O}(N)$ operations.
 In other words, \celerite\ provides a computationally efficient framework that
@@
 parameters in the time domain.
-We expect that this has the potential to push asteroseismic analysis to lower
-signal-to-noise datasets and we hope to revisit this idea in a subsequent
-paper.
+This has the potential to push asteroseismic analysis to lower signal-to-noise
+datasets and we hope to revisit this idea in a future paper.
 
@@
 given by a mixture of 8 components with amplitudes and frequencies specified
-by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
+by $\nu_\mathrm{max}$, $\Delta \nu$, and some nuisance parameters.
 The first term is used to capture the granulation ``background''
@@
 where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters. \dfmtodo{A was used for diagonal of
-covariance matrix.}
-This model could be extended to include more angular modes and
-$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
-physical parameters of the star.
+are shared nuisance parameters.
+\response{Finally, we also fit for the amplitude of the white noise by adding
+a parameter $\sigma$ in quadrature with the uncertainties given for the
+\kepler\ observations.
+All of these parameters and their chosen priors are listed in
+Table~\ref{tab:example-4-params}.
+As before, these priors are all log-uniform except for $\epsilon$ where we use
+a zero-mean normal prior with a broad variance of $1~\mathrm{day}^2$ to break
+the degeneracy between $\nu_\mathrm{max}$ and $\epsilon$.
+}
+To build a more realistic model, this prescription could be extended to
+include more angular modes, or $\nu_\mathrm{max}$ and $\Delta \nu$ could be
+replaced by the fundamental physical parameters of the star.
 
@@
 compare our results to the results inferred from the full 4 year baseline of
-the \kepler\ mission.
+the \kepler\ Mission.
+For KIC~11615890, the published asteroseismic parameters measured using several
+years of \kepler\ observations are \citep{Pinsonneault:2014}
+\begin{eqnarray}
+    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
+    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
+\end{eqnarray}
+
 Unlike typical giants, KIC~11615890 is a member of a class of stars where the
@@
 
-For KIC~11615890, the published asteroseismic parameters measured using several
-years of \kepler\ observations are \citep{Pinsonneault:2014}
-\begin{eqnarray}
-    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
-    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
-\end{eqnarray}
-We randomly select a month-long segment of PDC \citep{Stumpe:2012, Smith:2012}
-\kepler\ data, initialize our \celeriteterm\ model using a grid search in the
-parameter space, and then use \emcee\ \citep{Foreman-Mackey:2013} to sample
-the joint posterior density for the full set of parameters.
+\response{
+For this demonstration, we randomly select a month-long segment of PDC
+\citep{Stumpe:2012, Smith:2012} \kepler\ data.
+Unlike the previous examples, the posterior distribution is sharply
+multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
+not practical.
+Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
+$\Delta\nu$ using only the month-long subset of data and following the
+standard procedure described by \citet{Huber:2009}.
+We then run set of \project{L-BFGS-B} optimizations with values of
+$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
+estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
+using the empirical relationship between these two quantities
+\citep{Stello:2009}.
+This initialization procedure is sufficient for this example, but the general
+application of this method will require a more sophisticated prescription.
+
+\Figure{astero} shows a 10~day subset of the dataset used for this example.
+The MAP model is overplotted on these data and the residuals away from the
+mean prediction of this model are shown in the bottom panel of
+\Figure{astero}.
+There is no obvious structure in these residuals, lending some credibility to
+the model specification.
+
+We initialize \exampleivnwalkers~walkers by sampling from an isotropic
+Gaussian centered on the MAP parameters (the full set of parameters and their
+priors are listed in Table~\ref{tab:example-4-params}), run
+\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
+\emcee.
+We estimate the mean autocorrelation time for the chains of
+$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
+samples from the marginalized posterior density.
 \Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
@@
 method-dependent uncertainties \citep{Verner:2011} so further experiments
-would be needed to assess the reliability of this method.
-
-This model requires about a minute of computation time to find the maximum
-likelihood parameters and then it takes about an hour to run the MCMC sampling
-to convergence on a dual-CPU laptop.
+would be needed to fully assess the reliability of this method.
+
+This model requires about 10~CPU minutes to run the MCMC to convergence.
 This is more computationally intensive than traditional methods of measuring
-asteroseismic oscillations, but is is also orders of magnitude cheaper than the
-same analysis using another GP solver.
+asteroseismic oscillations, but it is much cheaper than the same analysis
+using a general direct GP solver.
+For comparison, we estimate that repeating this analysis using a general
+Cholesky factorization implemented as part of a tuned linear algebra
+library\footnote{We use the Intel Math Kernel Library
+\url{https://software.intel.com/en-us/intel-mkl}} would require about 15~CPU
+hours.
 An in-depth discussion of the benefits of rigorous probabilistic inference of
@@
 but we hope to revisit this opportunity in the future.
+}
 
@@
 
-\subsection{Exoplanet transit fitting}
+\subsection{Example 5: Exoplanet transit fitting}
 
@@
 
-This paper includes data collected by the \kepler\ mission. Funding for the
-\kepler\ mission is provided by the NASA Science Mission directorate.
+This paper includes data collected by the \kepler\ Mission. Funding for the
+\kepler\ Mission is provided by the NASA Science Mission directorate.
 We are grateful to the entire \kepler\ team, past and present.

### 2017-06-02 paper/ms.tex
Minor edits of simulation section.

@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
-the likelihood with $\mu_\bvec{\theta} =0$ is given by
-\begin{eqnarray}
-Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
-&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
-&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
-&=& -\frac{1}{2}{\bf q}^T {\bf q}.
-\end{eqnarray}
-The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term in $-2\ln{\mathcal{L}}$
+with $\mu_\bvec{\theta} =0$ is given by
+\begin{eqnarray}
+\mathcal{Q} &=&{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
+&=& {\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
+&=& {\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
+&=& {\bf q}^T {\bf q}.
+\end{eqnarray}
+This last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.
 

### 2017-06-03 paper/ms.tex
Working on prediction section

--- 2017-06-02 paper/ms.tex
@@
 compute the log-determinant of $K$ and solve linear systems of the form
-$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
+$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
 semi-separable matrix.
@@
 \begin{eqnarray}
-\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
-\end{eqnarray}
-First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
+\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
 substitution algorithm
@@
     f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
-    \tilde{W}_{n-1,j}\,z_{n-1}\right] \nonumber\\
-    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
+    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
+    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
 \end{eqnarray}
@@
 Using this result and the following backward substitution algorithm, we can
-compute $\bvec{b}$
+compute $\bvec{z}$
 \begin{eqnarray}
 g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
-    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
-b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
+    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
+z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
 \end{eqnarray}
@@
 Simulation of a Gaussian process for large $N$ requires multiplying an
-$N$-dimensional vector of Gaussian normal deviates, $\bf{q}$, by the square root
+$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
 of the covariance matrix:
 \begin{equation}
-{\bf y} = K_\bvec{\alpha}^{1/2} {\bf q},
+\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
 \end{equation}
@@
 \begin{eqnarray}
-Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
-&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
-&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
-&=& -\frac{1}{2}{\bf q}^T {\bf q}.
+\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
+&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
+&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
+&=& \bvec{q}^T \bvec{q}.
 \end{eqnarray}
@@
 To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
-low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
+low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
 \begin{equation}
-{\bf y} = L D^{1/2} {\bf q}.
+\bvec{y} = L D^{1/2} \bvec{q}.
 \end{equation}
@@
 \begin{equation}
-{\bf y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) {\bf z}.
+\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
 \end{equation}
@@
 \begin{eqnarray}
-\bar f &=& K({\bf t},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
-&=& K({\bf t},{\bf t}) {\bf b}
-\end{eqnarray}
-where ${\bf y}$ is the vector of measurements and ${\bf b}$ may be computed from the Cholesky
+\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
+&=& K(\bvec{t},\bvec{t}) \bvec{z}
+\end{eqnarray}
+where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
 decomposition as described above.  The second step multiplication of the white-noise-free
-kernel times ${\bf b}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
-Note that for this particular example it is more efficient to compute $\bar f = {\bf y} -\sigma^2 {\bf I} {\bf b}$,
+kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
+Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
 but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.
@@
 
-For predicting or interpolating missing or future data points (and their variance) at times ${\bf t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.
+For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.
 The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
 \begin{eqnarray}
-\bar f_* &=& K({\bf t_*},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
-&=& K({\bf t_*},{\bf t}) {\bf b}
-\end{eqnarray}
-where ${\bf y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
-${\bf b} = \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y}$ can be computed from
+\bar f_* &=& K(\bvec{t_*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
+&=& K(\bvec{t_*},\bvec{t}) \bvec{z}
+\end{eqnarray}
+where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
+$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
 the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.
@@
 \begin{eqnarray}
-\bar f_{*,i} &=& K(t_{*,i},{\bf t}) {\bf b} \\
-&=& \sum_n \sum_p \alpha_p e^{-\beta_p \vert t_{*,i}-t_n\vert} b_j.
-\end{eqnarray}
-\agoltodo{Convert this to real, \celerite\ notation .}
+\bar f_{*,i} &=& K(t_{*,i},\bvec{t}) \bvec{z} \\
+&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,i}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,i}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,i}-t_n\vert)}\right) z_j.
+\end{eqnarray}
 Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
 $\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
-\begin{equation}
-\bar f_{*,i} = \sum_{n=1}^{n_0} \sum_p \alpha_p e^{-\beta_p( t_{*,i}-t_n)} b_n + \sum_{n=n_0+1}^N \sum_p \alpha_p e^{-\beta_p(t_n-t_{*,i})} b_n.
-\end{equation}
-Each of these double-sums may have the sums interchanged, giving:
-\begin{eqnarray}
-\bar f_{*,i} &=& \sum_p e^{-\beta_p ( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} \alpha_p e^{-\beta_p ( t_{*,i-1}-t_n) } b_n \\
-&+&
-\sum_p e^{-\beta_p ( t_{*,i+1}-t_{*,i})} \sum_{n=n+0+1}^{N} \alpha_p e^{-\beta_p ( t_n-t_{*,i+1}) } b_n
-\end{eqnarray}
-We can compute this recursively.
-Defining $Q^-_{i,n_0,p} = \sum_{j=1}^{n_0} \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ and defining $Q^+_{i,n_0,p} = \sum_{j=n_0}^N \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ gives:
+\begin{eqnarray}
+\bar f_{*,i} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,i}-t_n)} \left[a_j \cos{(d_j(t_{*,i}-t_n))}+b_j\sin{(d_j(t_{*,i}-t_n))}\right] z_n \\
+&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,i})} \left[a_j \cos{(d_j(t_n-t_{*,i}))}+b_j\sin{(d_j(t_n-t_{*,i}))}\right] z_n.
+\end{eqnarray}
+%Each of these double-sums may have the sums interchanged, giving:
+%\begin{eqnarray}
+%\bar f_{*,i} &=& Re\left[\sum_j e^{-(c_j+id_j)( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} (a_j+ib_j) e^{-(c_j+id_j)( t_{*,i-1}-t_n) } z_n\right. \\
+%&+&\left.
+%\sum_j e^{-(c_j+id_j)( t_{*,i+1}-t_{*,i})} \sum_{n=n_0+1}^{N} (a_j+ib_j) e^{-(c_j+id_j)( t_n-t_{*,i+1}) } z_n\right]
+%\end{eqnarray}
+We can compute this recursively.  Defining
+\begin{eqnarray}
+Q^-_{i,n_0,2j-1} &=& \sum_{n=1}^{n_0} a_j e^{-c_j(t_{*,i}-t_n)} \cos{(d_j(t_{*,i}-t_n))} z_n,\\
+Q^-_{i,n_0,2j} &=& \sum_{n=1}^{n_0} b_j e^{-c_j(t_{*,i}-t_n)} \sin{(d_j(t_{*,i}-t_n))} z_n,\\
+Q^+_{i,n_0,2j-1} &=& \sum_{n=n_0+1}^N a_j e^{-c_j(t_n-t_{*,i})}\cos{(d_j(t_n-t_{*,i}))} z_n,\\
+Q^+_{i,n_0,2j} &=& \sum_{n=n_0+1}^N b_j e^{-c_j(t_n-t_{*,i})} \sin{(d_j(t_n-t_{*,i}))} z_n,
+\end{eqnarray} gives:
+\agoltodo{Convert the following to real, \celerite\ notation .}
 \begin{equation}
@@
 
-The variance evaluated at future predicted times ${\bf t_*} > {\bf t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
-\begin{eqnarray}
-\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},{\bf t}^T)K({\bf t},{\bf t})^{-1} K({\bf t},t_{*,i+1})\\
+The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
+\begin{eqnarray}
+\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
 &=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
-&\times& e^{-\beta_p(t_{*,i}-{\bf t}^T)}K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}\\
+&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
 &=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
@@
 \begin{equation}
-P_{i,p,q} = e^{-\beta_p(t_{*,i}-{\bf t}^T)}
-K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}.
+P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
+K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
 \end{equation}
@@
 Note that this variance formula {\it only} works for prediction; we have
-yet to derive a formula for which ${\bf t_*}$ and ${\bf t}$ are
+yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
 interspersed.
@@
     \citep{Pinsonneault:2014}.
+    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
+    regions in the marginalized planes and the histograms along the diagonal
+    show the marginalized posterior for each parameter.
     \figurelabel{astero-corner}}
@@
 
-In this example, we inject the signal of a simulated exoplanet transit into a
-real \kepler\ light curve and then demonstrate that we can recover the true
-physical parameters of the exoplanet while modeling the stellar variability
-using \celerite.
+\response{
 This example is different from all the previous examples~--~both simulated and
-real~--~because, in this case, we are uninterested in the parameters of the
-covariance model.
-Instead, we're interested in inferring constraints on the parameters of the
-mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
-signal caused by the transit of the planet.
-We say ``mean model'' here to unambiguously distinguish it from the
-``covariance model'', but this is what would generally be called the ``model''
-in other analyses.
-In this case, we have a light curve of a rotating star with a planet
-transiting and we would like to fit for the physical parameters of the planet
-while taking our uncertainty about the star's variability into account.
-GP modeling has been used for this purpose throughout the exoplanet literature
-\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
-Foreman-Mackey:2016b, Grunblatt:2016}.
-
-In \eq{gp-likelihood} the physical parameters of the planet are called
+real~--~because, in this case, do not set the deterministic
+mean function $\mu_\bvec{\theta}$ to zero.
+Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
+physically significant function and the parameters are fundamental properties
+of the system.
+This is an example using real data~--~because we use the light curve from
+\sect{rotation}~--~but we multiply these data by a simulated transiting
+exoplanet model with known parameters.
+This allows us to show that we can recover the true parameters of the planet
+even when the transit signal is superimposed on the real variability of a
+star.
+}
+GP modeling has been used extensively for this purpose throughout the
+exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
+Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.
+
+In \eq{gp-likelihood} the physical parameters of the exoplanet are called
 $\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
 is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
-$\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
-$t_0$, an impact parameter $B$, the radius of the planet in units of the
-stellar radius $R_P/R_\star$, and several parameters describing the
-limb-darkening profile of the star \citep{Claret:2011}.
-As in the previous example, we model the stellar variability using a GP model
+$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
+$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
+planet in units of the stellar radius $R_P/R_\star$, and two parameters
+describing the limb-darkening profile of the star
+\citep{Claret:2011,Kipping:2013}.
+As in \sect{rotation}, we model the stellar variability using a GP model
 with a kernel given by \eq{rot-kernel} and fit for the parameters of the
-planet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
+exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
 simultaneously.
-
-We take a month-long segment of the \kepler\ light curve for
-KIC~1430163~--~the target from \sect{rotation}~--~and multiply it by a
-simulated transit model with known parameters.
-The fact that we are using a simulated transit enables us to confirm that we
-correctly measure the planet parameters but we do not use this knowledge to
-inform our fit.
+\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
+listed in Table~\ref{tab:example-5-params} along with their priors and the
+true values for $\bvec{\theta}$.}
+
+\response{ We take a 20~day segment of the \kepler\ light curve for
+KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
+model with the parameters listed in Table~\ref{tab:example-5-params}.
+Using these data, we maximize the joint posterior defined by the likelihood in
+\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
+\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
+simultaneously.
 The top panel of \Figure{transit-ml} shows the data including the simulated
-transit.
-The bottom panel shows the same data with the maximum likelihood prediction
-for the stellar variation subtracted.
-To find this model, we maximized the likelihood function in \eq{gp-likelihood}
-with respect to the transit parameters $\bvec{\theta}$ and the variability
-parameters $\bvec{\alpha}$ simultaneously.
+transit as black points with the MAP model prediction over-plotted in blue.
+The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
+where the MAP model has been subtracted and the MAP mean model
+$\mu_\bvec{\theta}$ has been added back in.
+For comparison, the transit model is over-plotted in the bottom panel of
+\Figure{transit-ml} and we see no obvious correlations in the residuals.
+
+Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
+MAP parameters, we run \examplevnburn~steps of burn-in and
+\examplevnsteps~steps of production MCMC using \emcee.
+We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
+chain and find \examplevneff~effective samples across the full chain.
 \Figure{transit-corner} shows the marginalized posterior constraints on the
 physical properties of the planet compared to the true values.
-This procedure produces estimates of the planet parameters that are consistent
-with the true values using \celerite\ as an effective model for the stellar
-variability.
-
-\newpage
-
-\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
-and the impact parameter.}
-
-\input{figures/transit/transit-params}
+Even though the \celeriteterm\ representation of the stellar variability is
+only an effective model, the inferred distributions for the physical
+properties of the planet $\bvec{\theta}$ are consistent with the true values.
+This promising result suggests that \celerite\ can be used as an effective
+model for transit inference for a significant gain in computational
+tractability.
+Even for this example with small $N$, the cost of computing the likelihood
+using \celerite\ is nearly two orders of magnitude faster than the same
+computation using a general direct solver.
+}
+
+\input{figures/transit/transit-params-edited}
 
@@
 
-\dfmtodo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
+\dfmtodo{Do this.}
 
@@
 
-There are many other methods of scaling GP models to large datasets and in
-this section we draw comparisons between \celerite\ and other popular methods.
+There are other methods of scaling GP models to large datasets and in this
+section we draw comparisons between \celerite\ and other popular methods.
 Scalable GP methods tend to fall into two categories: approximate and
@@
 
-As discussed previously, the \celerite\ method is based upon a method
-developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
-In the case where the kernel function has a single term of the form given in
-\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
-and the Rybicki \& Press method.
-Their method does not require solving the extended matrix because the matrix
-inverse can be calculated analytically, but the method is made slightly more
-complicated when white noise is included.
-We implement their full method and find empirically that it is more
-computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
-but we note that the requirement of a single, real, exponential kernel is
-extremely restrictive and this model cannot describe quasi-periodic
-variability.
-
-Another popular method uses the fact that, in the limit of evenly-spaced data
-and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
-\citep[for example][]{Dillon:2013}.
-There are exact methods for solving Toeplitz matrix equations that scale as
-$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
-$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
-\citep{Wilson:2014}.
-The Toeplitz method is, in some ways, more flexible than \celerite\ because it
-can be used with any stationary kernel, but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it is less efficient when applied
-to large datasets.
-
-\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
-wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
-scaling.
-This method has been widely applied in the context of exoplanet transit
-characterization, but it requires evenly spaced observations and the power
-spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
-gain the computational advantage.
-This wavelet method has been demonstrated to improve parameter estimation for
-transiting exoplanets \citep{Carter:2009}, but these strict requirements make
-this method applicable for only a limited set of use cases.
-
+\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}
+
+\response{
 The continuous autoregressive moving average (CARMA) models introduced into
@@
 \citep[for example][]{Luger:2017}.
+}
+
+\subsection{Other scalable methods}
+
+%As discussed previously, the \celerite\ method is based upon a method
+%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
+%In the case where the kernel function has a single term of the form given in
+%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
+%and the Rybicki \& Press method.
+%Their method does not require solving the extended matrix because the matrix
+%inverse can be calculated analytically, but the method is made slightly more
+%complicated when white noise is included.
+%We implement their full method and find empirically that it is more
+%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
+%but we note that the requirement of a single, real, exponential kernel is
+%extremely restrictive and this model cannot describe quasi-periodic
+%variability.
+
+Another popular method uses the fact that, in the limit of evenly-spaced data
+and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
+\citep[for example][]{Dillon:2013}.
+There are exact methods for solving Toeplitz matrix equations that scale as
+$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
+$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
+\citep{Wilson:2014}.
+The Toeplitz method is, in some ways, more flexible than \celerite\ because it
+can be used with any stationary kernel, but it requires uniformly spaced data
+and the scaling is worse than \celerite\ so it is less efficient when applied
+to large datasets.
+
+\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
+wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
+scaling.
+This method has been widely applied in the context of exoplanet transit
+characterization, but it requires evenly spaced observations and the power
+spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
+gain the computational advantage.
+This wavelet method has been demonstrated to improve parameter estimation for
+transiting exoplanets \citep{Carter:2009}, but these strict requirements make
+this method applicable for only a limited set of use cases.
 

### 2017-06-03 paper/ms.tex
Working on prediction section

@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
-the likelihood with $\mu_\bvec{\theta} =0$ is given by
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term in $-2\ln{\mathcal{L}}$
+with $\mu_\bvec{\theta} =0$ is given by
 \begin{eqnarray}
@@
 \end{eqnarray}
-The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
+This last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.
 
@@
 
-\begin{figure}[!p]
+\begin{figure}[!htbp]
 \begin{center}
@@
     \citep{Pinsonneault:2014}.
-    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
-    regions in the marginalized planes and the histograms along the diagonal
-    show the marginalized posterior for each parameter.
     \figurelabel{astero-corner}}
@@
 
-\response{
+In this example, we inject the signal of a simulated exoplanet transit into a
+real \kepler\ light curve and then demonstrate that we can recover the true
+physical parameters of the exoplanet while modeling the stellar variability
+using \celerite.
 This example is different from all the previous examples~--~both simulated and
-real~--~because, in this case, do not set the deterministic
-mean function $\mu_\bvec{\theta}$ to zero.
-Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
-physically significant function and the parameters are fundamental properties
-of the system.
-This is an example using real data~--~because we use the light curve from
-\sect{rotation}~--~but we multiply these data by a simulated transiting
-exoplanet model with known parameters.
-This allows us to show that we can recover the true parameters of the planet
-even when the transit signal is superimposed on the real variability of a
-star.
-}
-GP modeling has been used extensively for this purpose throughout the
-exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
-Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.
-
-In \eq{gp-likelihood} the physical parameters of the exoplanet are called
+real~--~because, in this case, we are uninterested in the parameters of the
+covariance model.
+Instead, we're interested in inferring constraints on the parameters of the
+mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
+signal caused by the transit of the planet.
+We say ``mean model'' here to unambiguously distinguish it from the
+``covariance model'', but this is what would generally be called the ``model''
+in other analyses.
+In this case, we have a light curve of a rotating star with a planet
+transiting and we would like to fit for the physical parameters of the planet
+while taking our uncertainty about the star's variability into account.
+GP modeling has been used for this purpose throughout the exoplanet literature
+\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
+Foreman-Mackey:2016b, Grunblatt:2016}.
+
+In \eq{gp-likelihood} the physical parameters of the planet are called
 $\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
 is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
-$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
-$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
-planet in units of the stellar radius $R_P/R_\star$, and two parameters
-describing the limb-darkening profile of the star
-\citep{Claret:2011,Kipping:2013}.
-As in \sect{rotation}, we model the stellar variability using a GP model
+$\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
+$t_0$, an impact parameter $B$, the radius of the planet in units of the
+stellar radius $R_P/R_\star$, and several parameters describing the
+limb-darkening profile of the star \citep{Claret:2011}.
+As in the previous example, we model the stellar variability using a GP model
 with a kernel given by \eq{rot-kernel} and fit for the parameters of the
-exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
+planet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
 simultaneously.
-\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
-listed in Table~\ref{tab:example-5-params} along with their priors and the
-true values for $\bvec{\theta}$.}
-
-\response{ We take a 20~day segment of the \kepler\ light curve for
-KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
-model with the parameters listed in Table~\ref{tab:example-5-params}.
-Using these data, we maximize the joint posterior defined by the likelihood in
-\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
-\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
-simultaneously.
+
+We take a month-long segment of the \kepler\ light curve for
+KIC~1430163~--~the target from \sect{rotation}~--~and multiply it by a
+simulated transit model with known parameters.
+The fact that we are using a simulated transit enables us to confirm that we
+correctly measure the planet parameters but we do not use this knowledge to
+inform our fit.
 The top panel of \Figure{transit-ml} shows the data including the simulated
-transit as black points with the MAP model prediction over-plotted in blue.
-The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
-where the MAP model has been subtracted and the MAP mean model
-$\mu_\bvec{\theta}$ has been added back in.
-For comparison, the transit model is over-plotted in the bottom panel of
-\Figure{transit-ml} and we see no obvious correlations in the residuals.
-
-Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
-MAP parameters, we run \examplevnburn~steps of burn-in and
-\examplevnsteps~steps of production MCMC using \emcee.
-We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
-chain and find \examplevneff~effective samples across the full chain.
+transit.
+The bottom panel shows the same data with the maximum likelihood prediction
+for the stellar variation subtracted.
+To find this model, we maximized the likelihood function in \eq{gp-likelihood}
+with respect to the transit parameters $\bvec{\theta}$ and the variability
+parameters $\bvec{\alpha}$ simultaneously.
 \Figure{transit-corner} shows the marginalized posterior constraints on the
 physical properties of the planet compared to the true values.
-Even though the \celeriteterm\ representation of the stellar variability is
-only an effective model, the inferred distributions for the physical
-properties of the planet $\bvec{\theta}$ are consistent with the true values.
-This promising result suggests that \celerite\ can be used as an effective
-model for transit inference for a significant gain in computational
-tractability.
-Even for this example with small $N$, the cost of computing the likelihood
-using \celerite\ is nearly two orders of magnitude faster than the same
-computation using a general direct solver.
-}
-
-\input{figures/transit/transit-params-edited}
+This procedure produces estimates of the planet parameters that are consistent
+with the true values using \celerite\ as an effective model for the stellar
+variability.
+
+\newpage
+
+\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
+and the impact parameter.}
+
+\input{figures/transit/transit-params}
 
@@
 
-\dfmtodo{Do this.}
+\dfmtodo{The D in Table 6 may be confused with the D used in the Cholesky solver.}
 
@@
 
-There are other methods of scaling GP models to large datasets and in this
-section we draw comparisons between \celerite\ and other popular methods.
+There are many other methods of scaling GP models to large datasets and in
+this section we draw comparisons between \celerite\ and other popular methods.
 Scalable GP methods tend to fall into two categories: approximate and
@@
 
-\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}
-
-\response{
+As discussed previously, the \celerite\ method is based upon a method
+developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
+In the case where the kernel function has a single term of the form given in
+\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
+and the Rybicki \& Press method.
+Their method does not require solving the extended matrix because the matrix
+inverse can be calculated analytically, but the method is made slightly more
+complicated when white noise is included.
+We implement their full method and find empirically that it is more
+computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
+but we note that the requirement of a single, real, exponential kernel is
+extremely restrictive and this model cannot describe quasi-periodic
+variability.
+
+Another popular method uses the fact that, in the limit of evenly-spaced data
+and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
+\citep[for example][]{Dillon:2013}.
+There are exact methods for solving Toeplitz matrix equations that scale as
+$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
+$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
+\citep{Wilson:2014}.
+The Toeplitz method is, in some ways, more flexible than \celerite\ because it
+can be used with any stationary kernel, but it requires uniformly spaced data
+and the scaling is worse than \celerite\ so it is less efficient when applied
+to large datasets.
+
+\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
+wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
+scaling.
+This method has been widely applied in the context of exoplanet transit
+characterization, but it requires evenly spaced observations and the power
+spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
+gain the computational advantage.
+This wavelet method has been demonstrated to improve parameter estimation for
+transiting exoplanets \citep{Carter:2009}, but these strict requirements make
+this method applicable for only a limited set of use cases.
+
 The continuous autoregressive moving average (CARMA) models introduced into
@@
 \citep[for example][]{Luger:2017}.
-}
-
-\subsection{Other scalable methods}
-
-%As discussed previously, the \celerite\ method is based upon a method
-%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
-%In the case where the kernel function has a single term of the form given in
-%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
-%and the Rybicki \& Press method.
-%Their method does not require solving the extended matrix because the matrix
-%inverse can be calculated analytically, but the method is made slightly more
-%complicated when white noise is included.
-%We implement their full method and find empirically that it is more
-%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
-%but we note that the requirement of a single, real, exponential kernel is
-%extremely restrictive and this model cannot describe quasi-periodic
-%variability.
-
-Another popular method uses the fact that, in the limit of evenly-spaced data
-and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
-\citep[for example][]{Dillon:2013}.
-There are exact methods for solving Toeplitz matrix equations that scale as
-$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
-$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
-\citep{Wilson:2014}.
-The Toeplitz method is, in some ways, more flexible than \celerite\ because it
-can be used with any stationary kernel, but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it is less efficient when applied
-to large datasets.
-
-\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
-wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
-scaling.
-This method has been widely applied in the context of exoplanet transit
-characterization, but it requires evenly spaced observations and the power
-spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
-gain the computational advantage.
-This wavelet method has been demonstrated to improve parameter estimation for
-transiting exoplanets \citep{Carter:2009}, but these strict requirements make
-this method applicable for only a limited set of use cases.
 

### 2017-06-04 paper/ms.tex
adding some words and the true values for the transit example

--- 2017-06-02 paper/ms.tex
@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term in $-2\ln{\mathcal{L}}$
-with $\mu_\bvec{\theta} =0$ is given by
-\begin{eqnarray}
-\mathcal{Q} &=&{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
-&=& {\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
-&=& {\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
-&=& {\bf q}^T {\bf q}.
-\end{eqnarray}
-This last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
+the likelihood with $\mu_\bvec{\theta} =0$ is given by
+\begin{eqnarray}
+Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
+&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
+&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
+&=& -\frac{1}{2}{\bf q}^T {\bf q}.
+\end{eqnarray}
+The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
 
@@
 
-In this example, we inject the signal of a simulated exoplanet transit into a
-real \kepler\ light curve and then demonstrate that we can recover the true
-physical parameters of the exoplanet while modeling the stellar variability
-using \celerite.
+\response{
 This example is different from all the previous examples~--~both simulated and
-real~--~because, in this case, we are uninterested in the parameters of the
-covariance model.
-Instead, we're interested in inferring constraints on the parameters of the
-mean model ($\mu_\bvec{\theta}(\bvec{x})$ from \sect{gps}), the model of the
-signal caused by the transit of the planet.
-We say ``mean model'' here to unambiguously distinguish it from the
-``covariance model'', but this is what would generally be called the ``model''
-in other analyses.
-In this case, we have a light curve of a rotating star with a planet
-transiting and we would like to fit for the physical parameters of the planet
-while taking our uncertainty about the star's variability into account.
-GP modeling has been used for this purpose throughout the exoplanet literature
-\citep[for example][]{Dawson:2014, Barclay:2015, Evans:2015,
-Foreman-Mackey:2016b, Grunblatt:2016}.
-
-In \eq{gp-likelihood} the physical parameters of the planet are called
+real~--~because, in this case, do not set the deterministic
+mean function $\mu_\bvec{\theta}$ to zero.
+Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
+physically significant function and the parameters are fundamental properties
+of the system.
+This is an example using real data~--~because we use the light curve from
+\sect{rotation}~--~but we multiply these data by a simulated transiting
+exoplanet model with known parameters.
+This allows us to show that we can recover the true parameters of the planet
+even when the transit signal is superimposed on the real variability of a
+star.
+}
+GP modeling has been used extensively for this purpose throughout the
+exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
+Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.
+
+In \eq{gp-likelihood} the physical parameters of the exoplanet are called
 $\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
 is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
-$\bvec{\theta}$ are a period $P$, a transit duration $T$, a phase or epoch
-$t_0$, an impact parameter $B$, the radius of the planet in units of the
-stellar radius $R_P/R_\star$, and several parameters describing the
-limb-darkening profile of the star \citep{Claret:2011}.
-As in the previous example, we model the stellar variability using a GP model
+$\bvec{\theta}$ are the orbital period $P$, the transit duration $T$, the
+phase or epoch $t_0$, the impact parameter $B$, the radius of the planet in
+units of the stellar radius $R_P/R_\star$, and two parameters describing
+the limb-darkening profile of the star \citep{Claret:2011,Kipping:2013}.
+As in \sect{rotation}, we model the stellar variability using a GP model
 with a kernel given by \eq{rot-kernel} and fit for the parameters of the
-planet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
+exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
 simultaneously.
 
-We take a month-long segment of the \kepler\ light curve for
-KIC~1430163~--~the target from \sect{rotation}~--~and multiply it by a
-simulated transit model with known parameters.
+We take a month-long segment of the \kepler\ light curve for KIC~1430163 and
+multiply it by a simulated transit model with known parameters.
 The fact that we are using a simulated transit enables us to confirm that we
@@
 
-\dfmtodo{Table 5 has two parameter with name "b":  the celerite coefficient,
-and the impact parameter.}
-
-\input{figures/transit/transit-params}
+\input{figures/transit/transit-params-edited}
 

### 2017-06-04 paper/figures/example-stats.tex
words about transit and more steps for transit

--- 2017-05-27 paper/figures/example-stats.tex
@@
 \begin{deluxetable}{cccccccc}
-\caption{The scaling and convergence stats for each example. \label{tab:example-stats}}
-\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & \colhead{direct cost} & \colhead{\celerite\ cost} & \colhead{$D$} & \colhead{steps} & \colhead{effective samples}\\
+\caption{The computational cost and convergence stats for each example. \label{tab:example-stats}}
+\tablehead{& \colhead{$N$} & \colhead{$J$} & \colhead{direct\tablenotemark{a}} & \colhead{\celerite\tablenotemark{b}} & \colhead{dimension\tablenotemark{c}} & \colhead{evaluations\tablenotemark{d}} & \colhead{$N_\mathrm{eff}$\tablenotemark{e}}\\
 &&& \colhead{ms} & \colhead{ms} &&&}
@@
 2 & 100 & 2 & 0.67 & 0.36 & 4 & 80000 & 1134 \\
-3 & 6950 & 2 & 7351.71 & 1.46 & 4 & 176000 & 3606 \\
+3 & 6950 & 2 & 8119.11 & 1.47 & 4 & 176000 & 2900 \\
 4 & 1440 & 8 & 828.93 & 2.74 & 9 & 640000 & 1443 \\
-5 & 1000 & 2 & 82.73 & 0.94 & 13 & 640000 & 1141 \\
+5 & 1000 & 2 & 88.30 & 0.96 & 13 & 1280000 & 1490 \\
 \enddata
+
+\tablenotetext{a}{The computational cost of computing the GP model using the
+general Cholesky factorization routine implemented in the Intel MKL.}
+\tablenotetext{b}{The computational cost of computing the GP model using
+\celerite.}
+\tablenotetext{c}{The total number of parameters in the model.}
+\tablenotetext{d}{The total number of evaluations of the model to run MCMC to
+convergence.}
+\tablenotetext{e}{The effective number of independent samples estimated by
+computing the integrated autocorrelation time of the chain.}
 \end{deluxetable}

### 2017-06-04 paper/ms.tex
words about transit and more steps for transit

--- 2017-06-04 paper/ms.tex
@@
 is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
-$\bvec{\theta}$ are the orbital period $P$, the transit duration $T$, the
-phase or epoch $t_0$, the impact parameter $B$, the radius of the planet in
-units of the stellar radius $R_P/R_\star$, and two parameters describing
-the limb-darkening profile of the star \citep{Claret:2011,Kipping:2013}.
+$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
+$T$, the phase or epoch $t_0$, the impact parameter $B$, the radius of the
+planet in units of the stellar radius $R_P/R_\star$, and two parameters
+describing the limb-darkening profile of the star
+\citep{Claret:2011,Kipping:2013}.
 As in \sect{rotation}, we model the stellar variability using a GP model
@@
 simultaneously.
-
-We take a month-long segment of the \kepler\ light curve for KIC~1430163 and
-multiply it by a simulated transit model with known parameters.
-The fact that we are using a simulated transit enables us to confirm that we
-correctly measure the planet parameters but we do not use this knowledge to
-inform our fit.
+\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
+listed in Table~\ref{tab:example-5-params} along with their priors and the
+true values for $\bvec{\theta}$.}
+
+\response{ We take a 20~day segment of the \kepler\ light curve for
+KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
+model with the parameters listed in Table~\ref{tab:example-5-params}.
+Using these data, we maximize the joint posterior defined by the likelihood in
+\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
+\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
+simultaneously.
 The top panel of \Figure{transit-ml} shows the data including the simulated
-transit.
-The bottom panel shows the same data with the maximum likelihood prediction
-for the stellar variation subtracted.
-To find this model, we maximized the likelihood function in \eq{gp-likelihood}
-with respect to the transit parameters $\bvec{\theta}$ and the variability
-parameters $\bvec{\alpha}$ simultaneously.
+transit as black points with the MAP model prediction over-plotted in blue.
+The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
+where the MAP model has been subtracted and the MAP mean model
+$\mu_\bvec{\theta}$ has been added back in.
+For comparison, the transit model is over-plotted in the bottom panel of
+\Figure{transit-ml} and we see no obvious correlations in the residuals.
+
+Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
+MAP parameters, we run \examplevnburn~steps of burn-in and
+\examplevnsteps~steps of production MCMC using \emcee.
+We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
+chain and find \examplevneff~effective samples across the full chain.
 \Figure{transit-corner} shows the marginalized posterior constraints on the
 physical properties of the planet compared to the true values.
-This procedure produces estimates of the planet parameters that are consistent
-with the true values using \celerite\ as an effective model for the stellar
-variability.
-
-\newpage
+Even though the \celeriteterm\ representation of the stellar variability is
+only an effective model, the inferred distributions for the physical
+properties of the planet $\bvec{\theta}$ are consistent with the true values.
+This promising result suggests that \celerite\ can be used as an effective
+model for transit inference for a significant gain in computational
+tractability.
+Even for this example with small $N$, the cost of computing the likelihood
+using \celerite\ is nearly two orders of magnitude faster than the same
+computation using a general direct solver.
+}
 

### 2017-06-04 paper/figures/collect_results.py

@@ -68,15 +68,28 @@ for i, example in enumerate(["simulated/correct", "simulated/wrong-qpo",
 with open("example-stats.tex", "w") as f:
     f.write("\\begin{floattable}\n")
     f.write("\\begin{deluxetable}{cccccccc}\n")
-    f.write("\caption{The scaling and convergence stats for each example. "
-            "\label{tab:example-stats}}\n")
-    f.write("\\tablehead{\colhead{example} & \colhead{$N$} & \colhead{$J$} & "
-            "\colhead{direct cost} & \colhead{\celerite\ cost} & "
-            "\colhead{$D$} & "
-            "\colhead{model evaluations} & \colhead{effective samples}\\\\\n")
+    f.write("\caption{The computational cost and convergence stats for each "
+            "example. \label{tab:example-stats}}\n")
+    f.write("\\tablehead{& \colhead{$N$} & \colhead{$J$} & "
+            "\colhead{direct\\tablenotemark{a}} & "
+            "\colhead{\celerite\\tablenotemark{b}} & "
+            "\colhead{dimension\\tablenotemark{c}} & "
+            "\colhead{evaluations\\tablenotemark{d}} "
+            "& \colhead{$N_\mathrm{eff}$\\tablenotemark{e}}\\\\\n")
     f.write("&&& \colhead{ms} & \colhead{ms} &&&}\n")
     f.write("\\startdata\n")
     f.write("\n".join(stats_rows))
     f.write("\n\\enddata\n")
+    f.write(r"""
+\tablenotetext{a}{The computational cost of computing the GP model using the
+general Cholesky factorization routine implemented in the Intel MKL.}
+\tablenotetext{b}{The computational cost of computing the GP model using
+\celerite.}
+\tablenotetext{c}{The total number of parameters in the model.}
+\tablenotetext{d}{The total number of evaluations of the model to run MCMC to
+convergence.}
+\tablenotetext{e}{The effective number of independent samples estimated by
+computing the integrated autocorrelation time of the chain.}
+""")
     f.write("\end{deluxetable}\n")
     f.write("\end{floattable}\n")

### 2017-06-04 paper/ms.tex
transit terminology

@@
 $\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
-$T$, the phase or epoch $t_0$, the impact parameter $B$, the radius of the
+$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
 planet in units of the stellar radius $R_P/R_\star$, and two parameters

### 2017-06-05 paper/ms.tex

@@
 $\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
-$T$, the phase or epoch $t_0$, the impact parameter $b$, the radius of the
+$T$, the phase or epoch $t_0$, the impact parameter $B$, the radius of the
 planet in units of the stellar radius $R_P/R_\star$, and two parameters

### 2017-06-05 paper/ms.tex
some words

@@
 k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
-    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
-\end{eqnarray}
-where $P$ is the rotation period of the star.
+    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
+\end{eqnarray}
+where $P_\mathrm{rot}$ is the rotation period of the star.
 GP modeling with the same kernel function has been proposed as a method of
@@
 k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
-    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + C)
+    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
 \right]
@@
 of MCMC using \emcee.
-We estimate the integrated autocorrelation time of the chain for $\ln P$ and
-estimate that we have \exampleiiineff\ independent samples.
-These samples give a posterior constraint on the period of $P =
+We estimate the integrated autocorrelation time of the chain for $\ln
+P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
+samples.
+These samples give a posterior constraint on the period of $P_\mathrm{rot} =
 \rotationperiod$ and the marginalized posterior distribution for $P$ is shown
@@
     the dataset and model from \Figure{rotation}.
-    The period is the parameter $P$ in \eq{rot-kernel} and this figure shows
-    the posterior distribution marginalized over all other nuisance parameters
-    in \eq{rot-kernel}.
-    This is consistent with the published rotation period made using the full
+    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
+    figure shows the posterior distribution marginalized over all other
+    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
+    published rotation period made using the full
     \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
@@
 $\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
-is a limb-darkened transit light curve \citep{Mandel:2002} and the parameters
-$\bvec{\theta}$ are the orbital period $P_\mathrm{orb}$, the transit duration
-$T$, the phase or epoch $t_0$, the impact parameter $B$, the radius of the
-planet in units of the stellar radius $R_P/R_\star$, and two parameters
-describing the limb-darkening profile of the star
-\citep{Claret:2011,Kipping:2013}.
+is a limb-darkened transit light curve \citep{Mandel:2002,
+Foreman-Mackey:2016a} and the parameters $\bvec{\theta}$ are the orbital
+period $P_\mathrm{orb}$, the transit duration $T$, the phase or epoch $t_0$,
+the impact parameter $b$, the radius of the planet in units of the stellar
+radius $R_P/R_\star$, the baseline relative flux of the light curve $f_0$, and
+two parameters describing the limb-darkening profile of the star
+\citep{Claret:2011, Kipping:2013}.
 As in \sect{rotation}, we model the stellar variability using a GP model
@@
 
-\dfmtodo{Do this.}
+\response{
+In this section, we demonstrated five potential use cases for \celerite.
+These examples span a range of data sizes and model complexities, and the
+last three are all based on active areas of research with immediate real-world
+applicability.
+The sizes of these datasets are modest on the scale of some existing and
+forthcoming survey datasets because we designed the experiments to be easily
+reproducible but, even in these cases, the inferences made here would be
+intractable without substantial computational investment.
+
+Table~\ref{tab:example-stats} lists the specifications of each example.
+
+In Example~3, for example, we inferred the rotation period of a star observed
+by \kepler\ with a cost of about 4~CPU-minutes.
+By contrast, the same inference using a general but optimized Cholesky
+factorization routine would require nearly 400~CPU-hours.
+The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
+}
 

### 2017-06-05 paper/figures/collect_results.py

@@ -24,6 +24,8 @@ for i, example in enumerate(["simulated/correct", "simulated/wrong-qpo",
          r"& {neff:.0f} \\")
         .format(i+1, **data)
     )
+    print(data["direct_time"] * data["ntot"] / 1000.0 / 60.0 / 60.0,
+          data["time"] * data["ntot"] / 1000.0 / 60.0)
 
     with open(macros_fn, "a") as f:
         f.write("\\newcommand{{\\example{0}ndata}}{{{1}}}\n"

### 2017-06-05 paper/ms.tex
Completed prediction section.

--- 2017-06-05 paper/ms.tex
@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
-the likelihood with $\mu_\bvec{\theta} =0$ is given by
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that $\mathcal{Q} = -2 \mathcal{L}$
+with $\mu_\bvec{\theta} =0$ is given by
 \begin{eqnarray}
@@
 The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.
 
@@
 
-Since the covariance matrix may be written as the sum of exponentials, we can write:
-\begin{eqnarray}
-\bar f_{*,i} &=& K(t_{*,i},\bvec{t}) \bvec{z} \\
-&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,i}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,i}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,i}-t_n\vert)}\right) z_j.
-\end{eqnarray}
-Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
-$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
-\begin{eqnarray}
-\bar f_{*,i} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,i}-t_n)} \left[a_j \cos{(d_j(t_{*,i}-t_n))}+b_j\sin{(d_j(t_{*,i}-t_n))}\right] z_n \\
-&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,i})} \left[a_j \cos{(d_j(t_n-t_{*,i}))}+b_j\sin{(d_j(t_n-t_{*,i}))}\right] z_n.
-\end{eqnarray}
-%Each of these double-sums may have the sums interchanged, giving:
+Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
+\begin{eqnarray}
+\bar f_{*,m} &=& K(t_{*,m},\bvec{t}) \bvec{z} \\
+&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,m}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,m}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,m}-t_n\vert)}\right) z_j,
+\end{eqnarray}
+for $m =1$ to $M$.
+Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,m}$ and
+$\{t_{n0+1},...,t_N\} \ge t_{*,m}$, gives
+\begin{eqnarray}
+\bar f_{*,m} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,m}-t_n)} \left[a_j \cos{(d_j(t_{*,m}-t_n))}+b_j\sin{(d_j(t_{*,m}-t_n))}\right] z_n \\
+&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,m})} \left[a_j \cos{(d_j(t_n-t_{*,m}))}+b_j\sin{(d_j(t_n-t_{*,m}))}\right] z_n.
+\end{eqnarray}
+We can compute this recursively.  Defining
+\begin{eqnarray}
+Q^-_{n_0,i} &=& \left(Q^-_{n_0-1,i}+z_n\tilde V_{n_0,i}\right)e^{-c_j(t_{n+1}-t_n)},\\
+X^-_{m,n_0,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
+Q^+_{n_0,i} &=& \left(Q^+_{n_0+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_n-t_{n-1})},\\
+X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
+\end{eqnarray}
+where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
+for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_{m}^*$.
+Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
+$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
+For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
+point the predicted mean can be computed from
+\begin{equation}
+\bar f_{*,i} = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
+\end{equation}
+
+%The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
 %\begin{eqnarray}
-%\bar f_{*,i} &=& Re\left[\sum_j e^{-(c_j+id_j)( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} (a_j+ib_j) e^{-(c_j+id_j)( t_{*,i-1}-t_n) } z_n\right. \\
-%&+&\left.
-%\sum_j e^{-(c_j+id_j)( t_{*,i+1}-t_{*,i})} \sum_{n=n_0+1}^{N} (a_j+ib_j) e^{-(c_j+id_j)( t_n-t_{*,i+1}) } z_n\right]
+%\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
+%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
+%&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
+%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
 %\end{eqnarray}
-We can compute this recursively.  Defining
-\begin{eqnarray}
-Q^-_{i,n_0,2j-1} &=& \sum_{n=1}^{n_0} a_j e^{-c_j(t_{*,i}-t_n)} \cos{(d_j(t_{*,i}-t_n))} z_n,\\
-Q^-_{i,n_0,2j} &=& \sum_{n=1}^{n_0} b_j e^{-c_j(t_{*,i}-t_n)} \sin{(d_j(t_{*,i}-t_n))} z_n,\\
-Q^+_{i,n_0,2j-1} &=& \sum_{n=n_0+1}^N a_j e^{-c_j(t_n-t_{*,i})}\cos{(d_j(t_n-t_{*,i}))} z_n,\\
-Q^+_{i,n_0,2j} &=& \sum_{n=n_0+1}^N b_j e^{-c_j(t_n-t_{*,i})} \sin{(d_j(t_n-t_{*,i}))} z_n,
-\end{eqnarray} gives:
-\agoltodo{Convert the following to real, \celerite\ notation .}
-\begin{equation}
-\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
-\end{equation}
-Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
-\begin{eqnarray}
-Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
-Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
-\end{eqnarray}
-for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
-
-The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
-\begin{eqnarray}
-\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
-&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
-&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
-&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
-\end{eqnarray}
-for $1 \le p,q \le J$,
-where
-\begin{equation}
-P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
-K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
-\end{equation}
-For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
-with the Cholesky decomposition.
-Subsequent $P_{i+1,p,q}$ may be obtained recursively as
-\begin{equation}
-P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
-\end{equation}
-in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
-Note that this variance formula {\it only} works for prediction; we have
-yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
-interspersed.
+%for $1 \le p,q \le J$,
+%where
+%\begin{equation}
+%P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
+%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
+%\end{equation}
+%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
+%with the Cholesky decomposition.
+%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
+%\begin{equation}
+%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
+%\end{equation}
+%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
+%Note that this variance formula {\it only} works for prediction; we have
+%yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
+%interspersed.
 

### 2017-06-05 paper/ms.tex
Made predicted time notation consistent

@@
 
-For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.
+For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t^*} = \left\{t_{1}^*,t_{2}^*,...,t_{M}^*\right\}$, we need to apply the covariance between the measured and predicted times.
 The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
 \begin{eqnarray}
-\bar f_* &=& K(\bvec{t_*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
-&=& K(\bvec{t_*},\bvec{t}) \bvec{z}
+\bar f_* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
+&=& K(\bvec{t^*},\bvec{t}) \bvec{z}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\bar f_{*,m} &=& K(t_{*,m},\bvec{t}) \bvec{z} \\
-&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,m}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,m}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,m}-t_n\vert)}\right) z_j,
+\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
+&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
 \end{eqnarray}
 for $m =1$ to $M$.
-Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,m}$ and
-$\{t_{n0+1},...,t_N\} \ge t_{*,m}$, gives
-\begin{eqnarray}
-\bar f_{*,m} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,m}-t_n)} \left[a_j \cos{(d_j(t_{*,m}-t_n))}+b_j\sin{(d_j(t_{*,m}-t_n))}\right] z_n \\
-&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,m})} \left[a_j \cos{(d_j(t_n-t_{*,m}))}+b_j\sin{(d_j(t_n-t_{*,m}))}\right] z_n.
+Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
+$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
+\begin{eqnarray}
+\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
+&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
 \end{eqnarray}
@@
 where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
-for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_{m}^*$.
+for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_m^*$.
 Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
@@
 \begin{equation}
-\bar f_{*,i} = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
+\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
 \end{equation}
 
-%The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
+%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
 %\begin{eqnarray}
-%\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
-%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
-%&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
-%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
+%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
+%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
+%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
+%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
 %\end{eqnarray}
@@
 %\begin{equation}
-%P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
-%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
+%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
+%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
 %\end{equation}
@@
 %\begin{equation}
-%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
+%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
 %\end{equation}
@@
 %Note that this variance formula {\it only} works for prediction; we have
-%yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
+%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
 %interspersed.

### 2017-06-05 paper/ms.tex

@@
 \begin{eqnarray}
-Q^-_{n_0,i} &=& \left(Q^-_{n_0-1,i}+z_n\tilde V_{n_0,i}\right)e^{-c_j(t_{n+1}-t_n)},\\
-X^-_{m,n_0,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
-Q^+_{n_0,i} &=& \left(Q^+_{n_0+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_n-t_{n-1})},\\
+Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
+X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
+Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
 X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
 \end{eqnarray}
-where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
+where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
 for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_m^*$.
@@
 \begin{equation}
-\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
+\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
 \end{equation}

### 2017-06-06 paper/ms.tex
a change
Further edits to manuscript.
intro edits

@@
 
-\dfmtodo{Talk about LSST here too!}
-
 In the astrophysical literature, Gaussian Processes
@@
 practice, by the computational cost and scaling.
-In general, the cost of computing a GP likelihood scales as the cube of
-the number of data points $\mathcal{O}(N^3)$ and in the current era of large
-time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
-$\sim10^{3-5}$ observations each~--~this scaling is prohibitive.
-
-In this paper, we present a method for computing a class of GP models that
-scales linearly with the number of data points $\mathcal{O}(N)$ for one
-dimensional data sets.
-\dfmtodo{This method is a generalization of a method developed by
-\citet{Ambikasaran:2015}
- ({\bf This is no longer correct -different Ambikasaran result!})}
-that was, in turn, built on intuition from a twenty
-year old paper \citep{Rybicki:1995}.
+The cost of computing a general GP likelihood scales as the cube of the number
+of data points $\mathcal{O}(N^3)$ and in the current era of large time domain
+surveys~--~with as many as $\sim10^{4-9}$ targets with $\sim10^{3-5}$
+observations each~--~this scaling is prohibitive.
+
+\response{Existing astronomical time series datasets have already reached the
+limit where na\"ive application GP models is no longer tractable.
+NASA's \kepler\ Mission \citep{Borucki:2010}, for example, measured light
+curves with more than 60,000 observations each for about 190,000 stars.
+Current and forthcoming surveys such as \project{K2} \citep{Howell:2014},
+\tess\ \citep{Ricker:2014}, \lsst\ \citep{Ivezic:2008}, \project{WFIRST}
+\citep{Spergel:2015}, and PLATO \citep{Rauer:2014} will continue to produce
+similar or larger data volumes.}
+
+In this paper, we present a method for directly and exactly computing a class
+of GP models that scales linearly with the number of data points
+$\mathcal{O}(N)$ for one dimensional data sets.
+\response{Unlike earlier linear methods using Kalman filters \citep[for
+example,][]{Kelly:2014}, this novel algorithm exploits the semi-separable
+structure of a specific class of covariance matrices to directly factorize and
+solve the system.
+}
 This method can only be used with one-dimensional datasets and the covariance
@@
 discussion of what this means in detail in the following sections.
-However, there is no further constraint on the data or the model.
-In particular, the measurements don't need to be evenly spaced and the
-uncertainties can be heteroscedastic.
+However, the measurements don't need to be evenly spaced and the uncertainties
+can be heteroscedastic.
 
@@
 matrix when it is generated by a mixture of exponentials.
-\dfmtodo{More precisely, the matrix can be embedded in an equivalent higher-dimensional
-system that can be solved efficiently using a sparse or banded solver. Replace
-with(?): The matrix can be expressed as the sum of semi-separable matrices, each
-of which is constructed from the outer product of one-dimensional vectors.  These
-are a type of low-rank matrix, which may be solved by Cholesky decomposition
-using a novel technique developed by one of us (SA).}
-As we will discuss in the following pages, this choice of kernel function is
-not as restrictive as it first seems because it arises naturally in some
-physical systems and it can be used as an effective model in many other cases.
+\response{The semi-separable nature of these matrices was first recognized by
+\citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
+\citep{Rybicki:1995}.}
+As we will discuss in the following pages, this choice of kernel function
+arises naturally in some physical systems and we demonstrate it can be used as
+an effective model in some other cases.
 This method is especially appealing compared to other similar methods~--~we
-return to these below~--~because it is exact, flexible, simple, and fast.
+discuss these comparisons in \sect{compare}~--~because it is exact, simple,
+and fast.
 
@@
 We are confident that the method described in the following pages will benefit
-this field but we also expect that there will be applications in other
+this field, but we also expect that there will be applications in other
 branches of astrophysics and beyond.
-In \sect{examples-real}, we present applications of the method to research
-problems in stellar rotation, asteroseismic analysis, and exoplanet transit
-fitting.
-Some readers might consider starting with that section for some motivating
+In \sect{examples}, we present applications of the method to research problems
+in stellar rotation (\sect{rotation}), asteroseismic analysis (\sect{astero}),
+and exoplanet transit fitting (\sect{transit}).
+Some readers might consider starting with these section for motivating
 examples before delving into the detailed derivations of the earlier sections.
 
-In the following pages, we motivate the general problem of GP modeling,
-describe the \dfmtodo{(delete) previously published} scalable method
-\citep{Rybicki:1995, Ambikasaran:2015} and our generalization, and demonstrate
-the model's application on various real and simulated data sets.
+In the following pages, we motivate the general problem of GP modeling, derive
+our novel direct solver, and demonstrate the model's application on real and
+simulated data sets.
 Alongside this paper, we have released well-tested and documented open source
@@
 to include the effects of correlated noise.
-The maximum likelihood values for the parameters $\bvec{\theta}$ and
+A point estimate for the values of the parameters $\bvec{\theta}$ and
 $\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
@@
 $\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
-Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
-can be obtained by multiplying the likelihood by a prior
+\response{Furthermore}, probabilistic constraints on $\bvec{\theta}$ and
+$\bvec{\alpha}$ can be obtained by multiplying the likelihood by a prior
 $p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
@@
 application is generally limited to small datasets because the cost of
-computing the inverse and determinant of the matrix $K_\bvec{\alpha}$ is
+computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
 $\mathcal{O}(N^3)$.
@@
 points $N$.
-This means that for large datasets, every evaluation of the likelihood
-quickly becomes computationally intractable.
+Furthermore, the storage requirements also scale as $\mathcal{O}(N^2)$.
+This means that for large datasets, every evaluation of the likelihood quickly
+becomes computationally intractable, and representing the matrix in memory can
+become expensive.
 In this case, the use of non-linear optimization or MCMC is no longer
@@
 
-In the following section, we present a method for substantially improving this
-scaling in many circumstances.
+\response{This is not a purely academic concern.
+While the cost of directly evaluating the GP likelihood using a tuned linear
+algebra library\footnote{For comparisons throughout this paper, we use the
+Intel Math Kernel Library (\url{https://software.intel.com/en-us/intel-mkl})
+running on a recent Apple MacBook Pro.} with a dataset of $N=1024$
+measurements is less than a tenth of a second, the same calculation with
+$N=8192$ takes over 8~seconds.
+Furthermore, most optimization and sampling routines are iterative, requiring
+many evaluations of this model to perform parameter estimation or inference.
+Existing datasets from astronomical surveys like \kepler\ and \project{K2}
+include tens of thousands of observations for hundreds of thousands of
+targets, and upcoming surveys like \lsst\ are expected to produce thousands or
+tens of thousands of measurements each for billions of astronomical sources
+\citep{Ivezic:2008}.
+Scalable methods will be required if we ever hope to use GP models with these
+datasets.
+}
+
+In the following section, we present a method for improving the cubic scaling
+for some use cases.
 We call our method and its implementations \celerite.\footnote{The name
@@
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
-    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
-\end{eqnarray}
+    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
+\end{eqnarray}
+\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
+kernel is semi-separable.}
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
-evaluated in $\mathcal{O}(N\,J^2)$, where $J$ is the number of components
-in the mixture.
+evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
+components in the mixture.
+\response{In the following section, we derive a novel method with the same
+$\mathcal{O}(N\,J^2)$ scaling that can be used when
+$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}
 
@@
 
-\section{Examples}
+\section{Examples}\sectlabel{examples}
 
@@
 
-\subsection{Example 1: Recovery of a celerite process}
+\subsection{Example 1: Recovery of a celerite process}\sectlabel{example1}
 
@@
 
-\subsection{Example 2: Inferences with the ``wrong'' model}
+\subsection{Example 2: Inferences with the ``wrong'' model}\sectlabel{example2}
 
@@
 
-\subsection{Example 4: Asteroseismic oscillations}
+\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}
 
@@
 
-\subsection{Example 5: Exoplanet transit fitting}
+\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}
 
@@
 
-\section{Comparisons to other methods}
+\section{Comparisons to other methods}\sectlabel{compare}
 

### 2017-06-07 paper/ms.tex
a tiny change
moving prediction, etc. to appendix

@@
 method in \project{C++}, \project{Python}, and \project{Julia}.
+A change.
 
@@
 \end{eqnarray}
+
+Note that rather than storing the full $N \times N$ covariance matrix, it
+is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
+as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
+point numbers.
+
+\subsection{Cholesky factorization}
+
+To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
+compute the log-determinant of $K$ and solve linear systems of the form
+$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
+semi-separable matrix.
+If $K$ is positive semi-definite, these can be computed using the Cholesky
+factorization of $K$:
+\begin{eqnarray}
+K = L\,D\,L^\T
+\end{eqnarray}
+where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
+diagonal matrix.
+We use the ansatz that $L$ has the form
+\begin{eqnarray}
+    L = I + \mathrm{tril} (U\,W^T)
+\end{eqnarray}
+where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
+$N \times 2\,J$ matrix.
+Combining this with \eq{semi-sep}, we find the equation
+\begin{eqnarray}\eqlabel{cholesky}
+A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,W^\T)\,D
+    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $W$.
+Setting each element on the left side of \eq{cholesky} equal to the
+corresponding element on the right side, we derive the following recursion
+relations
+\begin{eqnarray}
+S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
+    \nonumber\\
+W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
+    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
+    \eqlabel{algorithm}
+\end{eqnarray}
+where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
+$S_{1,j,k}=0$.
+The computational cost of one iteration of the update in \eq{algorithm} is
+$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
+observation~--~so the total cost of this factorization scales as
+$\mathcal{O}(N\,J^2)$.
+
+Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
+then these terms only add one to the rank of $U$ and $V$.  We use this fact
+to simplify the algebra when real components are included:  if $J_{real}$
+and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
+the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
+
+\subsection{Pre-conditioning}
+
+Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
+overflow and underflow issues in many realistic cases because both $U$ and $V$
+have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
+real number.
+This can be avoided with the appropriate reparameterization.
+We define the following pre-conditioned variables
+\begin{eqnarray}
+\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
+\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
+\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
+\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
+\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
+\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
+\end{eqnarray}
+and the set of variables
+\begin{eqnarray}
+\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
+\end{eqnarray}
+with the constraint
+\begin{eqnarray}
+\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
+\end{eqnarray}
+Using this parameterization, we find the following, numerically stable,
+algorithm
+\begin{eqnarray}
+    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
+    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
+    \nonumber\\
+\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
+    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
+\end{eqnarray}
+
+\subsection{Solving}
+
+After computing the Cholesky factorization of the matrix $K$, we can apply
+the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
+$\mathcal{O}(N)$ operations respectively.
+The log-determinant of $K$ is given by
+\begin{eqnarray}
+    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
+\end{eqnarray}
+where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
+
+The inverse of $K$ can be applied using the relationship
+\begin{eqnarray}
+\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
+substitution algorithm
+\begin{eqnarray}
+    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
+    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
+    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{0,j} = 0$ for all $j$.
+Using this result and the following backward substitution algorithm, we can
+compute $\bvec{z}$
+\begin{eqnarray}
+g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
+    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
+z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
+\end{eqnarray}
+where $g_{N+1,j} = 0$ for all $j$.
+The total cost of the forward and backward substitution passes scales as
+$\mathcal{O}(N\,J)$.
+
+\section{celerite as a model of stellar variations}\sectlabel{sho}
+
+We now turn to a discussion of \celeriteterm\ as a model of astrophysical
+variability.
+A common concern in the context of GP modeling in astrophysics is the lack of
+physical motivation for the choice of kernel functions.
+Kernels are often chosen simply because they are popular with little
+consideration of the impact of this decision.
+While this isn't necessarily a problem for the inferred results, in this
+section we discuss an exact physical interpretation of the \celeriteterm\
+kernel that will be applicable to many astrophysical systems but especially
+the time-variability of stars.
+
+Many astronomical objects are variable on timescales determined by their
+physical structure.
+For phenomena such as stellar (asteroseismic) oscillations, variability is
+excited by noisy physical processes and grows most strongly at the
+characteristic timescale but is also damped due to dissipation in the system.
+These oscillations are strong at resonant frequencies determined by the
+internal stellar structure, which are both excited and damped by convective
+turbulence.
+
+To relate this physical picture to \celeriteterm, we consider the dynamics of a
+stochastically-driven damped simple harmonic oscillator (SHO).
+The differential equation for this system is
+\begin{equation}
+    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
+    + \omega_0^2\right]\, y(t) = \epsilon(t)
+\end{equation}
+where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
+quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
+force.
+If $\epsilon(t)$ is white noise, the PSD of this process is
+\citep{Anderson:1990}
+\begin{equation}\eqlabel{sho-psd}
+S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
+    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
+\end{equation}
+where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
+= \sqrt{2/\pi}\,S_0\,Q^2$.
+The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
+\begin{eqnarray}\eqlabel{sho-complex}
+a_j &=& S_0\,\omega_0\,Q \\
+b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
+c_j &=& \frac{\omega_0}{2\,Q}\\
+d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
+\end{eqnarray}
+for $Q \ge \frac{1}{2}$.
+For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of
+\celeriteterm\
+terms with parameters
+\begin{eqnarray}\eqlabel{sho-real}
+a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
+        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
+b_{j\pm} &=& 0 \nonumber\\
+    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
+    \nonumber\\
+d_{j\pm} &=& 0 \quad. \nonumber
+\end{eqnarray}
+
+For these definitions, the kernel is
+\begin{equation}\eqlabel{sho-kernel}
+k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_0) =
+S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
+\begin{cases}
+    \cosh{(\eta\,\omega_0\,\tau)} +
+        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
+    2\,(1+\omega_0\,\tau), & Q = 1/2\\
+    \cos{(\eta\,\omega_0\,\tau)} +
+        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
+\end{cases}
+\end{equation}
+where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
+We note that, because of the damping, the characteristic oscillation frequency
+in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
+the frequency of the undamped oscillator, $\omega_0$.
+
+The power spectrum in \eq{sho-psd} has several limits of physical interest:
+\begin{itemize}
+
+{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
+\begin{eqnarray}\eqlabel{granulation-psd}
+S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
+\end{eqnarray}
+This functional form is commonly used to model for the background granulation
+noise in asteoreseismic and helioseismic \citep{Harvey:1985, Michel:2009,
+Kallinger:2014}  analyses.
+The kernel for this value of $Q$ is
+\begin{equation}
+k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
+    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)} \quad.
+\end{equation}}
+
+{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
+\begin{eqnarray}
+S(\omega) =
+    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
+\end{eqnarray}
+with the corresponding covariance function (using
+\eqalt{celerite-kernel} and \eqalt{sho-real})
+\begin{eqnarray}\eqlabel{approx-matern}
+k(\tau) &=& \lim_{f \to 0}\,
+    \frac{1}{2}\,S_0\,\omega_0\,
+    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
+          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
+    \right] \\
+&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
+\end{eqnarray}
+or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
+\begin{eqnarray}\eqlabel{approx-matern2}
+k(\tau) &=& \lim_{f \to 0}\,
+    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
+    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
+&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
+\end{eqnarray}
+This covariance function is also known as the Mat\'ern-3/2 function
+\citep{Rasmussen:2006}.
+This suggests that the Mat\'ern-3/2 covariance can be approximated using
+the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
+caution that this might lead to numerical issues for the solver.
+}
+
+{\item Finally, in the limit of large $Q$, the model approaches a high
+    quality oscillation with frequency $\omega_0$ and covariance function
+\begin{eqnarray}
+k(\tau) \approx
+    S_0\,\omega_0\,Q\,
+    \exp\left(-\frac{\omega_0\,\tau}{2\,Q}\right)\,
+    \cos\left(\omega_0\,\tau\right) \quad.
+\end{eqnarray}}
+
+\end{itemize}
+\Figure{sho} shows a plot of the PSD for these limits and several other values
+of $Q$.
+From this figure, it is clear that for $Q \le 1/2$, the model has no
+oscillatory behavior and that for large $Q$, the shape of the PSD near the
+peak frequency approaches a Lorentzian.
+
+These special cases demonstrate that the stochastically-driven simple harmonic
+oscillator provides a physically motivated model that is flexible enough to
+describe a wide range of stellar variations and we return to give some
+specific examples in \sect{examples-real}.
+Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
+model for asteroseismic oscillations.
+In practice, we take a sum over oscillators with different values of $Q$,
+$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
+of stellar time series.
+Since this kernel is exactly described by the exponential kernel, the
+likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
+measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
+described in the previous section.
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
+\caption{(left) The power spectrum of a stochastically-driven simple harmonic
+    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
+    factor $Q$.
+    For comparison, the dashed line shows the Lorentzian function from
+    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
+    $S(d_j)/S(0) = 100$.
+    (middle) The corresponding autocorrelation functions with the same colors.
+    \response{(right) Three realizations from each model in the same colors.}
+    \figurelabel{sho}}
+\end{center}
+\end{figure}
+
+\section{Examples}\sectlabel{examples}
+
+\response{
+To show the application of \celerite, we use it to perform posterior
+inference in two examples with realistic simulated data and three examples of
+real-world research problems.
+These examples have been chosen to justify the application of \celerite\ for a
+range of use cases, but they are by no means exhaustive.
+These examples are framed in the time domain with a clear bias in favor of
+large homogeneous photometric surveys, but \celerite\ can also be used for
+other one-dimensional problems like, for example, spectroscopy, where
+wavelength~--~instead of time~--~is the independent coordinate \citep[see][for
+a potential application]{Czekala:2017}.
+
+In the first example, we demonstrate that \celerite\ can be used to infer
+the power spectrum of a process when the data are generated from a
+\celeriteterm\ model.
+In the second example, we demonstrate that \celerite\ can be used as an
+effective model even if the true process cannot be represented in the space of
+allowed models.
+This is an interesting example because, when analyzing real data, we rarely
+have any fundamental reason to believe that the data were generated by a GP
+model with a specific kernel.
+Even in these cases, GPs can be useful effective models and \celerite\
+provides computational advantages over other GP methods.
+
+The next three examples show \celerite\ used to infer the properties of stars
+and exoplanets observed by NASA's \kepler\ Mission.
+Each of these examples touches on an active area of research so we limit our
+examples to be qualitative in nature and do not claim that \celerite\ is the
+optimal method, but we hope these examples encourage interested readers to
+investigate the applicability of \celerite\ to their research.
+
+In each example, the joint posterior probability density is given by
+\begin{eqnarray}\eqlabel{joint-post}
+p(\bvec{\theta},\,\bvec{\alpha}\,|\,\bvec{y},\,X) \propto
+p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})\,
+p(\bvec{\theta},\,\bvec{\alpha})
+\end{eqnarray}
+where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
+likelihood defined in \eq{gp-likelihood} and $p(\bvec{\theta},\,\bvec{\alpha})$
+is the joint prior probability density for the parameters.
+We sample each posterior density using MCMC and investigate the performance of
+the model when used to interpret the physical processes that generated the
+data.
+The specific parameters and priors are discussed in detail in each section,
+but we generally assume separable uniform priors with plausibly broad support.
+}
+
+\subsection{Example 1: Recovery of a celerite process}\sectlabel{example1}
+
+In this first example, we simulate a dataset using a known \celeriteterm\
+process and fit it with \celerite\ to show that valid inferences can be made
+in this idealized case.
+\response{We simulated a small ($N = \exampleindata$) dataset using a GP model
+with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
+1\,\mathrm{ppm}^2$, $\omega_0 = e^2\,\mathrm{day}^{-1}$, and $Q = e^2$, where
+the units are arbitrarily chosen to simplify the discussion.
+We add further white noise with the amplitude $\sigma_n = 2.5\, \mathrm{ppm}$
+to each data point.
+The simulated data are plotted in the top left panel of
+\Figure{simulated-correct}.
+
+In this case, when simulating and fitting, we set the mean function
+$\mu_\bvec{\theta}$ to zero~--~this means that the parameter vector
+$\bvec{\theta}$ is empty~--~and the elements of the covariance matrix are
+given by \eq{sho-kernel} with three parameters $\bvec{\alpha} = (S_0,\,
+\omega_0,\,Q)$.
+We choose a proper separable prior for $\bvec{\alpha}$ with log-uniform
+densities for each parameter as listed in Table~\ref{tab:example-1-params}.
+
+To start, we estimate the maximum \emph{a posteriori} (MAP) parameters using
+the \project{L-BFGS-B} non-linear optimization routine \citep{Byrd:1995,
+Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
+The top left panel of \Figure{simulated-correct} shows the conditional mean
+and standard deviation of the MAP model over-plotted on the simulated data and
+the bottom panel shows the residuals away from this model.
+Since we are using the correct model to fit the data, it is reassuring that
+the residuals appear qualitatively uncorrelated in this figure.
+
+We then sample the joint posterior probability (\eqalt{joint-post}) using
+\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
+We initialize \exampleinwalkers~walkers in a Gaussian ball with width
+$10^{-4}$ in each parameter around the MAP parameter vector, run
+\exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
+To assess convergence, we estimate the mean integrated autocorrelation time of
+the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
+chain results in \exampleineff~effective samples.
+
+Each sample in the chain corresponds to a model PSD and we compare this
+posterior constraint on the PSD to the true spectral density in the
+right-hand panel of \Figure{simulated-correct}.
+In this figure, the true PSD is plotted as a dashed black line and the
+numerical estimate of the posterior inference is shown as a blue contour
+indicating 68\% of the MCMC samples.
+It is clear from this figure that, as expected, the inference correctly
+reproduces the true PSD.
+}
+
+\input{figures/simulated/correct-params}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
+\caption{(top left) A simulated dataset (black error bars), and maximum
+    likelihood model (blue contours).
+    (bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
+    (right) The inferred PSD~--~the blue contours encompass 68\% of the
+    posterior mass~--~compared to the true PSD (dashed black line).
+    \figurelabel{simulated-correct}}
+\end{center}
+\end{figure}
+
+\subsection{Example 2: Inferences with the ``wrong'' model}\sectlabel{example2}
+
+In this example, we simulate a dataset using a known GP model with a kernel
+outside of the support of a \celeriteterm\ process.
+This means the true autocorrelation of the process can never be correctly
+represented by the model that we are using to fit, but we use this example
+to demonstrate that, at least in this case, valid inferences can still be made
+about the physical parameters of the model.
+
+The data are simulated from a quasiperiodic GP with the kernel
+\begin{eqnarray}\eqlabel{sim-wrong-true}
+k_\mathrm{true} (\tau) = A\,
+    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
+    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
+\end{eqnarray}
+where $P_\mathrm{true}$ is the fundamental period of the process.
+This autocorrelation structure corresponds to the power spectrum
+\citep{Wilson:2013}
+\begin{eqnarray}
+S_\mathrm{true} (\omega) = \frac{\lambda\,A}{2}\,\left[
+    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
+        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
+    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
+        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
+\right]
+\end{eqnarray}
+which, for large values of $\omega$, falls off exponentially.
+When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
+$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
+exactly reproduce the structure of this process.
+That being said, we demonstrate that robust inferences can be made about
+$P_\mathrm{true}$ even with this effective model.
+
+\response{We generate a realization of a GP model with the kernel given in
+\eq{sim-wrong-true} with $N = \exampleiindata$, $A = 1\,\mathrm{ppm}^2$,
+$\lambda = 5\,\mathrm{day}$, and $P_\mathrm{true} = 1\,\mathrm{day}$.
+We also add white noise with amplitude $\sigma = 0.5\,\mathrm{ppm}$ to each
+data point.
+The top left panel of \Figure{simulated-wrong} shows this simulated dataset.
+
+We then fit this simulated data using the product of two SHO terms
+(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
+1/\sqrt{2}$ fixed.
+The kernel for this model is
+\begin{eqnarray}
+k(\tau) = k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_1) \,
+    k_\mathrm{SHO}(\tau;\,S_0 = 1,\,Q = 1/\sqrt{2},\,\omega_2)
+\end{eqnarray}
+where $k_\mathrm{SHO}$ is defined in \eq{sho-kernel} and the period of the
+process is $P=2\,\pi/\omega_1$.
+We note that using \eq{product-rule}, the product of two \celeriteterm\ terms
+can also be expressed using \celerite.
+
+As in the previous example, we set the mean function to zero and can,
+therefore, omit the parameters $\bvec{\theta}$.
+Table~\ref{tab:example-2-params} lists the proper log-uniform priors that we
+choose for each parameter in $\bvec{\alpha} =
+(S_0,\,Q,\,\omega_1,\,\omega_2)$.
+These priors, together with the GP likelihood (\eqalt{gp-likelihood}) fully
+specify the posterior probability density.
+
+As above, we estimate the MAP parameters using \project{L-BFGS-B} and sample
+the posterior probability density using \emcee.
+The top left panel of \Figure{simulated-wrong} shows the conditional mean and
+standard deviation of the MAP model.
+The bottom left panel shows the residuals between the data and this MAP model
+and, even though this GP model is formally ``wrong'', there are no obvious
+correlations in these residuals.
+
+To perform posterior inference, we initialize \exampleiinwalkers~walkers by
+sampling from the 4-dimensional Gaussian centered on the MAP parameters with
+an isotropic standard deviation of $10^{-4}$.
+We then run \exampleiinburn~steps of burn-in and \exampleiinsteps~steps of
+MCMC.
+To estimate the number of effective samples, we estimate the integrated
+autocorrelation time of the chain for the parameter $\omega_1$~--~the
+parameter of primary interest~--~and find \exampleiineff~effective samples.
+
+For comparison, we run the same number of steps of MCMC to sample the
+``correct'' joint posterior density.
+For this reference inference, we use a GP likelihood with the kernel given by
+\eq{sim-wrong-true} and choose log-uniform priors on each of the
+three parameters $\ln A/\mathrm{ppm}^2 \sim \mathcal{U}(-10,\,10)$,
+$\ln \lambda/\mathrm{day} \sim \mathcal{U}(-10,\,10)$, and
+$\ln P_\mathrm{true}/\mathrm{day} \sim \mathcal{U}(-10,\,10)$.
+
+The marginalized posterior inferences of the characteristic period of the
+process are shown in the right panel of \Figure{simulated-wrong}.
+The inference using the correct model is shown as a dashed blue histogram and
+the inference made using the effective model is shown as a solid black
+histogram.
+These inferences are consistent with each other and with the true period used
+for the simulation (shown as a vertical gray line).
+This demonstrates that, in this case, \celerite\ can be used as a
+computationally efficient effective model and hints that this may be true in
+other problems as well.
+}
+
+\input{figures/simulated/wrong-qpo-params}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
+\caption{(top left) A simulated dataset (black error bars), and maximum
+    likelihood model (blue contours).
+    (bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
+    (right) The inferred period of the process. The true period is indicated
+    by the vertical orange line, the posterior inference using the correct
+    model is shown as the blue dashed histogram, and the inference made using
+    the ``wrong'' effective model is shown as the black histogram.
+    \figurelabel{simulated-wrong}}
+\end{center}
+\end{figure}
+
+\subsection{Example 3: Stellar rotation} \sectlabel{rotation}
+
+A source of variability that can be measured from time series measurements of
+stars is rotation.
+The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
+quasiperiodic variations in photometric or spectroscopic observations
+\citep{Dumusque:2014}.
+It has been demonstrated that for light curves with nearly uniform sampling,
+the empirical autocorrelation function provides a reliable estimate of the
+rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
+and that a GP model with a
+quasiperiodic covariance function can be used to make probabilistic
+measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
+The covariance function used for this type of analysis has the form
+\begin{eqnarray}\eqlabel{sine2}
+k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
+    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
+\end{eqnarray}
+where $P_\mathrm{rot}$ is the rotation period of the star.
+GP modeling with the same kernel function has been proposed as a method of
+measuring the mean periodicity in quasiperiodic photometric time series in
+general \citep{Wang:2012}.
+The key difference between \eq{sine2} and other quasiperiodic kernels is that
+it is positive for all values of $\tau$.
+We construct a simple \celeriteterm\ covariance function with similar
+properties as follows
+\begin{eqnarray}\eqlabel{rot-kernel}
+k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
+    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
+\right]
+\end{eqnarray}
+for $B>0$, $C>0$, and $L>0$.
+The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
+but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
+as a drop-in replacement for a significant gain in computational efficiency.
+
+\response{GPs have been used to measure stellar rotation periods for
+individual datasets \citep[for example][]{Littlefair:2017}, but the
+computational cost of traditional GP methods has hindered the industrial
+application to existing surveys like \kepler\ with hundreds of thousands of
+targets.
+The increase in computational efficiency and scalability provided by
+\celerite\ opens the possibility of inferring rotation periods using GPs at
+scale of existing and forthcoming surveys like \kepler, \tess, and \lsst.
+}
+
+As a demonstration, we fit a \celeriteterm\ model with a kernel given
+by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
+This star has a published rotation period of $3.88 \pm 0.58\,\mathrm{days}$,
+measured using traditional periodogram and autocorrelation function approaches
+applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
+
+\response{We select about 180~days of contiguous observations of
+KIC~1430163 from \kepler.
+This dataset has \exampleiiindata\ measurements and, using a tuned linear
+algebra
+implementation\footnote{We use the Intel Math Kernel Library
+\url{https://software.intel.com/en-us/intel-mkl}}, a
+single evaluation of the likelihood requires over 8~seconds on a modern Intel
+CPU.
+This calculation, using \celerite\ with the model in \eq{rot-kernel}, only
+takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
+magnitude per model evaluation.
+
+We set the mean function $\mu_\bvec{\theta}$ to zero and the remaining
+parameters $\bvec{\alpha}$ and their priors are listed in
+Table~\ref{tab:example-3-params}.
+As with the earlier examples, we start by estimating the MAP parameters using
+\project{L-BFGS-B} and initialize \exampleiiinwalkers~walkers by sampling from
+an isotropic Gaussian with a standard deviation of $10^{-5}$ centered on the
+MAP parameters.
+The left panels of \Figure{rotation} show a subset of the data used in this
+example and the residuals away from the MAP predictive mean.
+
+We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
+of MCMC using \emcee.
+We estimate the integrated autocorrelation time of the chain for $\ln
+P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
+samples.
+These samples give a posterior constraint on the period of $P_\mathrm{rot} =
+\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
+in the right panel of \Figure{rotation}.
+This result is in good agreement with the literature value with smaller
+uncertainties.
+A detailed comparison of GP rotation period measurements and the traditional
+methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
+\etal, submitted) demonstrates that GP inferences are, at a population level,
+more reliable than other methods.
+}
+
+\input{figures/rotation/rotation-params}
+
+\begin{figure}[p]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
+\caption{Inferred constraints on a quasiperiodic GP model using the covariance
+    function in \eq{rot-kernel} and two quarters of \kepler\ data.
+(top left) The \kepler\ data (black points) and the maximum likelihood model
+    prediction (blue curve) for a 60~day subset of the data used.
+    The solid blue line shows the predictive mean and the blue contours show
+    the predictive standard deviation.
+(bottom left) The residuals between the mean predictive model and the data
+    shown in the top left figure.
+(right) The posterior constraint on the rotation period of KIC~1430163 using
+    the dataset and model from \Figure{rotation}.
+    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
+    figure shows the posterior distribution marginalized over all other
+    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
+    published rotation period made using the full
+    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
+    \figurelabel{rotation}}
+\end{center}
+\end{figure}
+
+\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}
+
+The asteroseismic oscillations of thousands of stars were measured using light
+curves from the \kepler\ Mission \citep{Gilliland:2010, Huber:2011,
+Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
+driver for many of the upcoming large scale photometric surveys
+\citep{Campante:2016, Rauer:2014, Gould:2015}.
+Most asteroseismic analyses have been limited to high signal-to-noise
+oscillations because the standard methods use statistics of the empirical
+periodogram and cannot formally propagate the measurement uncertainties to the
+constraints on physical parameters, instead relying on bootstrapped
+uncertainty estimators \citep{Huber:2009}.
+More sophisticated methods that compute the likelihood function in the time
+domain scale poorly to large survey datasets \citep{Brewer:2009,
+Corsaro:2014}.
+
+\celerite\ alleviates these problems by providing a physically motivated
+probabilistic model that can be evaluated efficiently even for large datasets.
+In practice, we model the star as a mixture of stochastically-driven simple
+harmonic oscillators where the amplitudes and frequencies of the oscillations
+are computed using a physical model, and evaluate the probability of the
+observed time series using a GP where the PSD is a sum of terms given by
+\eq{sho-psd}.
+This gives us a method for computing the likelihood function for the
+parameters of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta
+\nu$, or other more fundamental parameters) \emph{conditioned on the observed
+time series} in $\mathcal{O}(N)$ operations.
+In other words, \celerite\ provides a computationally efficient framework that
+can be combined with physically-motivated models of stars and numerical
+inference methods to make rigorous probabilistic measurements of asteroseismic
+parameters in the time domain.
+This has the potential to push asteroseismic analysis to lower signal-to-noise
+datasets and we hope to revisit this idea in a future paper.
+
+To demonstrate the method, we use a simple heuristic model where the PSD is
+given by a mixture of 8 components with amplitudes and frequencies specified
+by $\nu_\mathrm{max}$, $\Delta \nu$, and some nuisance parameters.
+The first term is used to capture the granulation ``background''
+\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
+$S_g$ and $\omega_g$.
+The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
+parameter shared between terms and the frequencies are given by
+\begin{eqnarray}
+\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
+\end{eqnarray}
+and the amplitudes are given by
+\begin{eqnarray}
+S_{0,\,j} =
+    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
+\end{eqnarray}
+where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
+are shared nuisance parameters.
+\response{Finally, we also fit for the amplitude of the white noise by adding
+a parameter $\sigma$ in quadrature with the uncertainties given for the
+\kepler\ observations.
+All of these parameters and their chosen priors are listed in
+Table~\ref{tab:example-4-params}.
+As before, these priors are all log-uniform except for $\epsilon$ where we use
+a zero-mean normal prior with a broad variance of $1~\mathrm{day}^2$ to break
+the degeneracy between $\nu_\mathrm{max}$ and $\epsilon$.
+}
+To build a more realistic model, this prescription could be extended to
+include more angular modes, or $\nu_\mathrm{max}$ and $\Delta \nu$ could be
+replaced by the fundamental physical parameters of the star.
+
+To demonstrate the applicability of this model, we apply it to infer the
+asteroseismic parameters of the giant star KIC~11615890, observed by the
+\kepler\ Mission.
+The goal of this example is to show that, even for a low signal-to-noise
+dataset with a short baseline, it is possible to infer asteroseismic
+parameters with formal uncertainties that are consistent with the parameters
+inferred with a much larger dataset.
+Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
+$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
+compare our results to the results inferred from the full 4 year baseline of
+the \kepler\ Mission.
+For KIC~11615890, the published asteroseismic parameters measured using several
+years of \kepler\ observations are \citep{Pinsonneault:2014}
+\begin{eqnarray}
+    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
+    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
+\end{eqnarray}
+
+Unlike typical giants, KIC~11615890 is a member of a class of stars where the
+dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
+in the core \citep{Stello:2016}.
+This makes this target simpler for the purposes of this demonstration because
+we can neglect the $\ell = 1$ modes and the model proposed above will be an
+effective model for the combined signal from the $\ell = 0$ and 2 modes.
+
+\response{
+For this demonstration, we randomly select a month-long segment of PDC
+\citep{Stumpe:2012, Smith:2012} \kepler\ data.
+Unlike the previous examples, the posterior distribution is sharply
+multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
+not practical.
+Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
+$\Delta\nu$ using only the month-long subset of data and following the
+standard procedure described by \citet{Huber:2009}.
+We then run set of \project{L-BFGS-B} optimizations with values of
+$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
+estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
+using the empirical relationship between these two quantities
+\citep{Stello:2009}.
+This initialization procedure is sufficient for this example, but the general
+application of this method will require a more sophisticated prescription.
+
+\Figure{astero} shows a 10~day subset of the dataset used for this example.
+The MAP model is overplotted on these data and the residuals away from the
+mean prediction of this model are shown in the bottom panel of
+\Figure{astero}.
+There is no obvious structure in these residuals, lending some credibility to
+the model specification.
+
+We initialize \exampleivnwalkers~walkers by sampling from an isotropic
+Gaussian centered on the MAP parameters (the full set of parameters and their
+priors are listed in Table~\ref{tab:example-4-params}), run
+\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
+\emcee.
+We estimate the mean autocorrelation time for the chains of
+$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
+samples from the marginalized posterior density.
+\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
+and $\Delta\nu$ compared to the results from the literature.
+This result is consistent within the published error bars and the posterior
+constraints are tighter than the published results.
+All asteroseismic analyses are known to have substantial systematic and
+method-dependent uncertainties \citep{Verner:2011} so further experiments
+would be needed to fully assess the reliability of this method.
+
+This model requires about 10~CPU minutes to run the MCMC to convergence.
+This is more computationally intensive than traditional methods of measuring
+asteroseismic oscillations, but it is much cheaper than the same analysis
+using a general direct GP solver.
+For comparison, we estimate that repeating this analysis using a general
+Cholesky factorization implemented as part of a tuned linear algebra
+library\footnote{We use the Intel Math Kernel Library
+\url{https://software.intel.com/en-us/intel-mkl}} would require about 15~CPU
+hours.
+An in-depth discussion of the benefits of rigorous probabilistic inference of
+asteroseismic parameters in the time domain is beyond the scope of this paper,
+but we hope to revisit this opportunity in the future.
+}
+
+\input{figures/astero/astero-params}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero/astero.pdf}
+\caption{
+\response{
+(top) The \kepler\ data (black points) and the maximum likelihood model
+    prediction (blue curve) for a 10~day subset of the month-long dataset that
+    was used for the fit.
+    The solid blue line shows the predictive mean and the blue contours show
+    the predictive standard deviation.
+(bottom) The residuals between the mean predictive model and the data
+    shown in the top figure.
+}
+    \figurelabel{astero}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!p]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
+\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
+    from the inference shown in \Figure{astero} compared to the published
+    value (error bar) based on several years of \kepler\ observations
+    \citep{Pinsonneault:2014}.
+    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
+    regions in the marginalized planes and the histograms along the diagonal
+    show the marginalized posterior for each parameter.
+    \figurelabel{astero-corner}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!p]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
+\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
+posterior inference of the PSD as a mixture of stochastically-driven simple
+harmonic oscillators.
+(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
+    on the full four year baseline of the mission. The orange line shows a
+    smoothed periodogram and the blue line indicates the level of the
+    measurement uncertainties.
+(middle) The same periodogram computed using about a month of data.
+(bottom) The power spectrum inferred using the mixture of SHOs model described
+    in the text and only one month of \kepler\ data.
+    The black line shows the median of posterior PSD and the gray contours
+    show the 68\% credible region.
+    \figurelabel{astero}}
+\end{center}
+\end{figure}
+
+\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}
+
+\response{
+This example is different from all the previous examples~--~both simulated and
+real~--~because, in this case, do not set the deterministic
+mean function $\mu_\bvec{\theta}$ to zero.
+Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
+physically significant function and the parameters are fundamental properties
+of the system.
+This is an example using real data~--~because we use the light curve from
+\sect{rotation}~--~but we multiply these data by a simulated transiting
+exoplanet model with known parameters.
+This allows us to show that we can recover the true parameters of the planet
+even when the transit signal is superimposed on the real variability of a
+star.
+}
+GP modeling has been used extensively for this purpose throughout the
+exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
+Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.
+
+In \eq{gp-likelihood} the physical parameters of the exoplanet are called
+$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
+is a limb-darkened transit light curve \citep{Mandel:2002,
+Foreman-Mackey:2016a} and the parameters $\bvec{\theta}$ are the orbital
+period $P_\mathrm{orb}$, the transit duration $T$, the phase or epoch $t_0$,
+the impact parameter $b$, the radius of the planet in units of the stellar
+radius $R_P/R_\star$, the baseline relative flux of the light curve $f_0$, and
+two parameters describing the limb-darkening profile of the star
+\citep{Claret:2011, Kipping:2013}.
+As in \sect{rotation}, we model the stellar variability using a GP model
+with a kernel given by \eq{rot-kernel} and fit for the parameters of the
+exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
+simultaneously.
+\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
+listed in Table~\ref{tab:example-5-params} along with their priors and the
+true values for $\bvec{\theta}$.}
+
+\response{ We take a 20~day segment of the \kepler\ light curve for
+KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
+model with the parameters listed in Table~\ref{tab:example-5-params}.
+Using these data, we maximize the joint posterior defined by the likelihood in
+\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
+\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
+simultaneously.
+The top panel of \Figure{transit-ml} shows the data including the simulated
+transit as black points with the MAP model prediction over-plotted in blue.
+The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
+where the MAP model has been subtracted and the MAP mean model
+$\mu_\bvec{\theta}$ has been added back in.
+For comparison, the transit model is over-plotted in the bottom panel of
+\Figure{transit-ml} and we see no obvious correlations in the residuals.
+
+Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
+MAP parameters, we run \examplevnburn~steps of burn-in and
+\examplevnsteps~steps of production MCMC using \emcee.
+We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
+chain and find \examplevneff~effective samples across the full chain.
+\Figure{transit-corner} shows the marginalized posterior constraints on the
+physical properties of the planet compared to the true values.
+Even though the \celeriteterm\ representation of the stellar variability is
+only an effective model, the inferred distributions for the physical
+properties of the planet $\bvec{\theta}$ are consistent with the true values.
+This promising result suggests that \celerite\ can be used as an effective
+model for transit inference for a significant gain in computational
+tractability.
+Even for this example with small $N$, the cost of computing the likelihood
+using \celerite\ is nearly two orders of magnitude faster than the same
+computation using a general direct solver.
+}
+
+\input{figures/transit/transit-params-edited}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/transit/transit-ml.pdf}
+    \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
+    KIC~1430163 with a synthetic transit model injected (black points) and the
+    maximum likelihood model for the stellar variability
+    (blue line).
+    \emph{(bottom)} The maximum likelihood ``de-trending'' of the data in
+    the top panel.
+    In this panel, the maximum likelihood model for the stellar variability
+    has been subtracted to leave only the transits.
+    The de-trended light curve is shown by black error bars and the maximum
+    likelihood transit model is shown as a blue line.
+    \figurelabel{transit-ml}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!p]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/transit/transit-corner.pdf}
+\caption{The marginalized posterior constraints on the physical parameters of
+    the planet transit in the light curve shown in the top panel of
+    \Figure{transit-ml}.
+    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
+    regions in the marginalized planes and the histograms along the diagonal
+    show the marginalized posterior for each parameter.
+    The true values used in the simulation are indicated by blue lines.
+    For each parameter, the inference is consistent with the true value.
+    \figurelabel{transit-corner}}
+\end{center}
+\end{figure}
+
+\subsection{Summary}
+
+\response{
+In this section, we demonstrated five potential use cases for \celerite.
+These examples span a range of data sizes and model complexities, and the
+last three are all based on active areas of research with immediate real-world
+applicability.
+The sizes of these datasets are modest on the scale of some existing and
+forthcoming survey datasets because we designed the experiments to be easily
+reproducible but, even in these cases, the inferences made here would be
+intractable without substantial computational investment.
+
+Table~\ref{tab:example-stats} lists the specifications of each example.
+
+In Example~3, for example, we inferred the rotation period of a star observed
+by \kepler\ with a cost of about 4~CPU-minutes.
+By contrast, the same inference using a general but optimized Cholesky
+factorization routine would require nearly 400~CPU-hours.
+The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
+}
+
+\input{figures/example-stats}
+
+\section{Comparisons to other methods}\sectlabel{compare}
+
+There are other methods of scaling GP models to large datasets and in this
+section we draw comparisons between \celerite\ and other popular methods.
+Scalable GP methods tend to fall into two categories: approximate and
+restrictive.
+\celerite\ falls into the latter category because, while the method is exact,
+it requires a specific choice of stationary kernel function and it can only be
+used in one-dimension.
+
+\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}
+
+\response{
+The continuous autoregressive moving average (CARMA) models introduced into
+the astrophysics literature by \citet{Kelly:2014} share many features with
+\celerite.
+CARMA models are derived by solving a stochastic differential equation and,
+like \celerite, the likelihood function for a CARMA model conditioned on a
+one-dimensional dataset can be solved in $\mathcal{O}(N)$.
+The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
+\begin{eqnarray}\eqlabel{carma-kernel}
+k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
+\end{eqnarray}
+where
+\begin{eqnarray}\eqlabel{carma-coeff}
+A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
+    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
+    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
+\end{eqnarray}
+and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
+the model.
+Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
+every CARMA model corresponds to an equivalent \celeriteterm\ model and the
+parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
+The inverse statement is not as simple.
+In practice, this means that \celerite\ could be trivially used to compute any
+CARMA model.
+Using the CARMA solver to compute a \celeriteterm\ model, however, requires
+solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
+While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
+the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
+is, in practice, somewhat faster and the memory requirements are smaller.
+One benefit of the \celeriteterm\ model is that it is parameterized in terms
+of the amplitudes directly.
+This means that the physical interpretation of these parameters is, in some
+cases, simpler.
+On the other hand, CARMA models are parameterized by modeling the power
+spectrum of the stochastic driving force and this might be the quantity of
+interest in other situations.
+Another benefit of the \celeriteterm\ framework is that, unlike
+filtering-based CARMA solvers, it directly solves matrix equations of the form
+${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
+This is a crucial feature when GP modeling is combined with linear models
+\citep[for example][]{Luger:2017}.
+}
+
+\subsection{Other scalable methods}
+
+%As discussed previously, the \celerite\ method is based upon a method
+%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
+%In the case where the kernel function has a single term of the form given in
+%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
+%and the Rybicki \& Press method.
+%Their method does not require solving the extended matrix because the matrix
+%inverse can be calculated analytically, but the method is made slightly more
+%complicated when white noise is included.
+%We implement their full method and find empirically that it is more
+%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
+%but we note that the requirement of a single, real, exponential kernel is
+%extremely restrictive and this model cannot describe quasi-periodic
+%variability.
+
+Another popular method uses the fact that, in the limit of evenly-spaced data
+and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
+\citep[for example][]{Dillon:2013}.
+There are exact methods for solving Toeplitz matrix equations that scale as
+$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
+$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
+\citep{Wilson:2014}.
+The Toeplitz method is, in some ways, more flexible than \celerite\ because it
+can be used with any stationary kernel, but it requires uniformly spaced data
+and the scaling is worse than \celerite\ so it is less efficient when applied
+to large datasets.
+
+\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
+wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
+scaling.
+This method has been widely applied in the context of exoplanet transit
+characterization, but it requires evenly spaced observations and the power
+spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
+gain the computational advantage.
+This wavelet method has been demonstrated to improve parameter estimation for
+transiting exoplanets \citep{Carter:2009}, but these strict requirements make
+this method applicable for only a limited set of use cases.
+
+Another GP method that has been used extensively in astronomy is the
+hierarchical off-diagonal low rank \citep[HODLR,][]{Ambikasaran:2016} solver.
+This method exploits the fact that many commonly used kernel functions produce
+``smooth'' matrices to approximately compute the GP likelihood with the
+scaling $\mathcal{O}(N\,\log^2 N)$.
+This method has the advantage that, unlike \celerite, it can be used with any
+kernel function but, in practice, the cost can still prove to be prohibitively
+high for multi-dimensional inputs.
+The proportionality constant in the $N\,\log^2N$ scaling of the HODLR method
+is a function of the specific kernel and we find~--~using the \project{george}
+software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
+scales approximately linearly with $J$, but it requires substantial overhead
+for small models.
+This means that the HODLR solver can \emph{approximately} evaluate
+\celeriteterm\
+models more efficiently than \celerite\ for large models $J \gg 10$ and
+small datasets where $N$ is smaller than a few thousand.
+
+The structured kernel interpolation \citep[SKI/KISS-GP][]{Wilson:2015}
+framework is another approximate method that can be used to scale GPs to large
+datasets.
+This method uses fast interpolation of the kernel into a space where Toeplitz
+or Kronecker structure can be used to scale inference and prediction.
+The SKI/KISS-GP framework can be applied to scale GP inference with a wide
+range of kernels, but the computational cost will depend on the specific
+dataset, model, and precision requirements for the approximation.
+The SKI method has an impressive $\mathcal{O}(1)$ cost for test-time
+predictions and it is interesting to consider how this could be applied to
+\celeriteterm\ models.
+
+Many other approximate methods for scaling GP inference exist \citep[see, for
+example,][and references therein]{Wilson:2015a} and we make no attempt to make
+our discussion exhaustive.
+The key takeaway here is that \celerite\ provides an \emph{exact} method for
+GP inference for a specific, but flexible, class of one-dimensional kernel
+functions.
+Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
+stochastically-driven, damped simple harmonic oscillators, they are a
+physically motivated choice of covariance function in many applications.
+
+\section{Summary}
+
+Gaussian Process models have been fruitfully applied to many problems in
+astronomical data analysis, but the fact that the computational cost scales as
+the cube of the number of data points has limited their use to relatively
+small datasets.
+With the linear scaling of \celerite\, we envision the application of Gaussian
+processes may be expanded to much larger datasets.
+Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
+number of components it is flexible enough to describe a wide range of
+astrophysical variability.
+In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
+oscillator matches simple models of astrophysical variability, and makes the
+parameterization interpretable in terms of resonant frequency, amplitude, and
+quality factor.
+% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
+A drawback of this method is its quadratic scaling with the number of
+terms $J$, but, in many cases, small values of $J$ are sufficient.
+
+Our background is in studying transiting exoplanets, a field which has only
+recently begun to adopt full covariance matrices in analyzing the noise in
+stellar light curves when detecting or characterizing transiting planets
+\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
+Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
+All of these analyses have been limited to small datasets or restrictive
+kernel choices.
+\celerite\ weakens these requirements by providing a scalable method for
+computing the likelihood and a physical motivation for the choice of kernel.
+\celerite\ can be used to model stellar oscillations using the relation to the
+mixture of stochastically-driven, damped simple harmonic oscillators.
+As higher signal-to-noise observations of transiting exoplanet systems are
+obtained, the effects of stellar variability will more dramatically impact the
+correct inference of planetary transit parameters.
+We expect that \celerite\ will be important for transit detection
+\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
+Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
+\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
+functions \citep{Knutson:2007, Zucker:2007}, and more.
+
+Beyond these applications to model stellar variability, the method is
+generally applicable to other one-dimensional GP models.
+Accreting black holes show time series which may be modeled using a GP
+\citep{Kelly:2014}; indeed, this was the motivation for the original technique
+developed by Rybicki \& Press \citep{Rybicki:1992, Rybicki:1995}.
+This approach may be broadly used for characterizing quasar variability
+\citep{MacLeod:2010}, measuring time lags with reverberation mapping
+\citep{Zu:2011, Pancoast:2014}, modeling time delays in multiply-imaged
+gravitationally-lensed systems \citep{Press:1998}, characterizing
+quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
+classification of variable objects \citep{Zinn:2016}.
+We expect that there are also applications beyond astronomy.
+
+The \celeriteterm\ formalism can also be used for power spectrum estimation and
+quantification of its uncertainties.
+In principle, a large number of \celeriteterm\ terms could be used to perform
+non-parametric probabilistic inference of the power spectrum despite
+unevenly-spaced data with heteroscedastic noise \citep[for
+example,][]{Wilson:2013, Kelly:2014}.
+This type of analysis will be limited by the quadratic scaling of \celerite\
+with the number of terms $J$, but this limits existing methods as well
+\citep[CARMA models,][]{Kelly:2014}.
+In general, we encourage the use of physically motivated models for parameter
+estimation instead of qualitative modeling of the power spectrum itself.
+
+There are many data analysis problems where \celerite\ will not be applicable.
+In particular, the restriction to one-dimensional problems is significant.
+There are many examples of multidimensional GP modeling the astrophysics
+literature \citep[recent examples from the field of exoplanet characterization
+include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
+be used to accelerate any of these analyses.
+It is plausible that an extension could be derived to tackle some
+multidimensional problems with the right structure~--~simultaneous parallel
+time series, for example~--~and we hope to revisit this possibility in future
+work.
+
+Alongside this paper, we have released a well-tested and documented open
+source software package that implements the method and all of the examples
+discussed in these pages.
+This software is available on GitHub
+\url{https://github.com/dfm/celerite}\footnote{This version of the paper was
+generated with git commit \texttt{\githash} (\gitdate).} and Zenodo
+\citep{Foreman-Mackey:2017}, and it is made available under the MIT license.
+
+\acknowledgments
+It is a pleasure to thank
+Megan Bedell,
+Ian Czekala,
+Will Farr,
+Sam Grunblatt,
+David W.\ Hogg,
+Dan Huber,
+Meredith Rawls,
+Dennis Stello,
+Jake VanderPlas, and
+Andrew Gordon Wilson
+for helpful discussions informing the ideas and code presented here.
+\response{We would also like to thank the anonymous referee for thorough and
+constructive feedback that greatly improved the paper.}
+
+This work was performed in part under contract with the Jet Propulsion
+Laboratory (JPL) funded by NASA through the Sagan Fellowship Program executed
+by the NASA Exoplanet Science Institute.
+EA acknowledges support from NASA grants NNX13AF20G, NNX13A124G, NNX13AF62G,
+from National Science Foundation (NSF) grant AST-1615315, and from
+NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
+by NASA under cooperative agreement NNH05ZDA001C.
+
+This research made use of the NASA \project{Astrophysics Data System} and the
+NASA Exoplanet Archive.
+The Exoplanet Archive is operated by the California Institute of Technology,
+under contract with NASA under the Exoplanet Exploration Program.
+
+This paper includes data collected by the \kepler\ Mission. Funding for the
+\kepler\ Mission is provided by the NASA Science Mission directorate.
+We are grateful to the entire \kepler\ team, past and present.
+These data were obtained from the Mikulski Archive for Space Telescopes
+(MAST).
+STScI is operated by the Association of Universities for Research in
+Astronomy, Inc., under NASA contract NAS5-26555.
+Support for MAST is provided by the NASA Office of Space Science via grant
+NNX13AC07G and by other grants and contracts.
+
+This research made use of Astropy, a community-developed core Python package
+for Astronomy \citep{Astropy-Collaboration:2013}.
+
+\facility{Kepler}
+\software{%
+     \project{AstroPy} \citep{Astropy-Collaboration:2013},
+     \project{corner.py} \citep{Foreman-Mackey:2016},
+     \project{Eigen} \citep{Guennebaud:2010},
+     \project{emcee} \citep{Foreman-Mackey:2013},
+     \project{george} \citep{Ambikasaran:2016},
+     \project{LAPACK} \citep{Anderson:1999},
+     \project{matplotlib} \citep{Hunter:2007},
+     \project{numpy} \citep{Van-Der-Walt:2011},
+     \project{transit} \citep{Foreman-Mackey:2016a},
+     \project{scipy} \citep{Jones:2001}.
+}
+
+\newpage
+\appendix
+
+\section{An alternative solver using the extended matrix formalism}
+
+\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
+elements given by \eq{kernel-simple}, could be computed
+efficiently by taking advantage of the structure of this covariance function.
+\citet{Ambikasaran:2015} generalized this computation to apply to the full
+mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
+efficient method for computing the determinant of $K$.
+
+\subsection{An example}
+
+To provide some insight for this method, we follow
+\citet{Ambikasaran:2015} and start by working through a simple example.
+In this case, we assume that we have three data points
+$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
+variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
+like to compute the likelihood of these data under a GP model with the
+covariance function
+\begin{eqnarray}
+k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
+\end{eqnarray}
+This is the result of setting $J=1$, $b=0$, and $d=0$ in
+\eq{celerite-kernel-complex} and it is the model studied by
+\citet{Rybicki:1995}.
+To demonstrate that the likelihood of this model can be computed in
+$\mathcal{O}(N)$, we write out the full system of
+equations that must be solved to apply the inverse of $K$ and compute
+the first term of \eq{gp-likelihood}.
+In matrix notation, this is
+\begin{eqnarray}\eqlabel{impl-matrix}
+K\,\bvec{z} &=& \bvec{y}, \\
+\begin{pmatrix}
+    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
+    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
+    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ z_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ y_2 \\ y_3
+\end{pmatrix}
+\end{eqnarray}
+where our goal is to solve for the unknown vector \bvec{z} for a given matrix
+$K$ and vector \bvec{y}.
+In \eq{impl-matrix}, we have assumed that the mean function is zero but a
+non-zero mean could be included by replacing \bvec{y} by
+$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
+Now, if we introduce the variables
+\begin{eqnarray}
+    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
+\end{eqnarray}
+where $g_{0} = 0$, and
+\begin{eqnarray}\eqlabel{algo-first}
+    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
+\end{eqnarray}
+where $u_{N} = 0$,
+the system of equations can be rewritten as
+\begin{eqnarray}
+(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
+a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
+a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
+\end{eqnarray}
+Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
+as a matrix equation shows the benefit that this seemingly trivial
+reformulation provides:
+\begin{eqnarray}
+\begin{pmatrix}
+    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
+    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
+    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
+    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
+    0 & 0 & 0 & 0 & -1 & 0 & a \\
+    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
+\end{pmatrix}\,
+\begin{pmatrix}
+    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
+\end{pmatrix} &=&
+\begin{pmatrix}
+    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
+\end{pmatrix}\nonumber
+\end{eqnarray}
+Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
+rewrite \eq{impl-matrix} as
+\begin{eqnarray}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
+\end{eqnarray}
+Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
+with, it is now sparse with banded structure that can be exploited to solve
+the system efficiently.
+In particular, sparse solvers are available that can perform a
+LU-decomposition of matrices like this in $\mathcal{O}(N)$
+operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
+general~--~and we can use these algorithms to solve our system
+exactly because the target vector $\bvec{z}$ is a subset of the elements of
+$\bvec{z}_\mathrm{ext}$.
+
+In the following section we discuss this method more generally, but it's
+worth noting a few important facts that can already be seen in this example.
+First, the fundamental reason why this matrix $K$ can be solved efficiently is
+the following property of exponentials
+\begin{eqnarray}
+    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
+    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
+    e^{-c\,(t_3 - t_1)}
+\end{eqnarray}
+and it is important to note that this property does not extend to other common
+covariance functions like the ``exponential-squared'' function
+\begin{eqnarray}
+    k(\tau) \propto e^{-c\,\tau^2} \quad.
+\end{eqnarray}
+Second, our derivation of the extended matrix requires that the data
+points be monotonically sorted.
+This is the motivation for our previous statement that this method can only be
+applied in one dimension since neither of these properties will be satisfied
+by multidimensional inputs.
+
+\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
+extended matrix formalism in practice.
+First, even if the covariance function is a mixture of exponentials, the
+extended matrix will still be banded with a bandwidth that scales linearly
+with the number of components, $J$.
+Second, \citet{Ambikasaran:2015} proved that the absolute value of the
+determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
+determinant of $K$.
+This means we can use this extended matrix formalism to compute the
+marginalized likelihood in $\mathcal{O}(N)$ operations.
+
+\subsection{The algorithm}
+
+In this section, we generalize the method from the previous section to the
+covariance function given by \eq{celerite-kernel}.
+This derivation follows \citet{Ambikasaran:2015} but it includes explicit
+treatment of complex parameters, and their complex conjugates.
+
+In the case of the full \celeriteterm\ covariance function
+(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
+analogy to the $u_n$ and $g_n$ that we introduced in the previous section
+\begin{eqnarray}\eqlabel{full-system-first}
+\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
+\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
+g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
+h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
+u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
+v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
+\end{eqnarray}
+with the boundary conditions
+\begin{eqnarray}
+    g_{0,j} = 0 \quad, \quad
+    h_{0,j} = 0 \quad, \quad
+    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
+    v_{N,j} = 0
+\end{eqnarray}
+for all $j$.
+Using these variables and some algebra, we find that the following expression
+\begin{eqnarray}\eqlabel{full-system-last}
+\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
++ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
++ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
+    &=& r_{\bvec{\theta},n}
+\end{eqnarray}
+is equivalent to the target matrix equation
+\begin{eqnarray}
+K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
+\end{eqnarray}
+if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
+$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
+\eq{full-system-first} through \eq{full-system-last} define a banded matrix
+equation in the ``extended'' space
+\begin{eqnarray}\eqlabel{kext-full}
+    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
+    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
+\end{eqnarray}
+and, as before, this can be used to solve
+for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
+operations.
+\Figure{matrix} shows a pictorial representation of the sparsity pattern of
+the extended matrix $K_\mathrm{ext}$.
+The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
+with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
+non-zero entries.
+Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
+$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
+as
+\begin{eqnarray}
+{\bvec{z}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
+    h_{N-1,j} & z_N
+\end{array}\right)
+\end{eqnarray}
+and
+\begin{eqnarray}
+{\bvec{r}_\mathrm{ext}} ^\T =
+\left(\begin{array}{cccccccccc}
+    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
+    0 & \cdots & 0 & r_{\bvec{\theta},N}
+\end{array}\right) \quad.
+\end{eqnarray}
+After constructing the extended matrix (using a compact storage format), the
+extended matrix can be factorized using a LU-decomposition\footnote{Even
+though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
+Cholesky solver cannot be used for increased efficiency.} routine optimized
+for band or sparse matrices.
+This decomposition can then be used to compute the determinant of $K$, solve
+$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
+likelihood in \eq{gp-likelihood}.
+
+In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
+case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
+will also be identically zero for all $n$.
+Removing these trivial rows from the extended matrix results in factor of two
+increase in the computational efficiency for these terms.
+In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
+$d_j=0$, and the general terms are called ``complex''.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
+\caption{A pictorial representation of the sparse extended matrix
+$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
+Each colored block corresponds to a non-zero entry in the matrix as described
+in the legend.
+    \figurelabel{matrix}}
+\end{center}
+\end{figure}
+
+\subsection{Implementation considerations \& scaling}
+
+The extended system defined in the previous section is sparse, with typically
+fewer than a few percent non-zero entries and band structure.
+In this section, we empirically investigate the performance and scaling
+of three different algorithms for solving this extended system:
+
+\begin{enumerate}
+
+\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
+    for banded matrices using Gaussian elimination \citep[we implemented this
+    method following][]{Press:1992, Press:2007},
+
+\item \texttt{lapack}: The general banded LU decomposition implementation from
+    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
+    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
+    \project{BLAS} routines,\footnote{The experiments in this paper use the
+    Intel Math Kernel Library (\project{MKL}):
+    \url{https://software.intel.com/en-us/intel-mkl}.} and
+
+\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
+    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
+    sparsity but not the band structure.
+
+\end{enumerate}
+
+The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
+because the dimension of the extended matrix scales as $N\,J$ and the
+bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
+\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
+\project{SuperLU} package \citep{Demmel:1999}.
+We find that, while the \texttt{vanilla} solver scales as expected, the
+\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
+offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
+
+The benchmark experiments shown here were performed on a MacBook Pro with two
+2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
+CPUs and running Ubuntu.
+\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
+with $N$ and $J$ using the \texttt{vanilla} solver.
+As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
+in $J$ for large $J$.
+\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
+the \texttt{lapack} and \texttt{sparse} solvers respectively.
+For each of these optimized solvers, the empirical scaling is
+$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
+Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
+efficient than the other algorithms.
+The real world performance of \celerite\ depends on the specific platform,
+hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
+found qualitatively similar results across popular platforms and
+state-of-the-art libraries.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
+\caption{\dfmtodo{Cholesky} A benchmark showing the computational scaling of \celerite\ using the
+    \texttt{vanilla} band solver as a function of the number of data points
+    and the number of terms.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
+    \figurelabel{benchmark}}
+\end{center}
+\end{figure}
+
+\section{Semi-separable operations}
 
@@
 \end{eqnarray}
-Note that rather than storing the full $N \times N$ covariance matrix, it
-is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
-as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
-point numbers.
-
-\subsection{Cholesky factorization}
-
-To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
-compute the log-determinant of $K$ and solve linear systems of the form
-$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
-semi-separable matrix.
-If $K$ is positive semi-definite, these can be computed using the Cholesky
-factorization of $K$:
-\begin{eqnarray}
-K = L\,D\,L^\T
-\end{eqnarray}
-where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
-diagonal matrix.
-We use the ansatz that $L$ has the form
-\begin{eqnarray}
-    L = I + \mathrm{tril} (U\,W^T)
-\end{eqnarray}
-where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
-$N \times 2\,J$ matrix.
-Combining this with \eq{semi-sep}, we find the equation
-\begin{eqnarray}\eqlabel{cholesky}
-A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
-&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
-    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
-&=& D + \mathrm{tril}(U\,W^\T)\,D
-    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
-&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
-\end{eqnarray}
-that we can solve for $W$.
-Setting each element on the left side of \eq{cholesky} equal to the
-corresponding element on the right side, we derive the following recursion
-relations
-\begin{eqnarray}
-S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
-D_{n,n} &=& A_{n,n} -
-    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
-    \nonumber\\
-W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
-    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
-    \eqlabel{algorithm}
-\end{eqnarray}
-where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
-$S_{1,j,k}=0$.
-The computational cost of one iteration of the update in \eq{algorithm} is
-$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
-observation~--~so the total cost of this factorization scales as
-$\mathcal{O}(N\,J^2)$.
-
-Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
-then these terms only add one to the rank of $U$ and $V$.  We use this fact
-to simplify the algebra when real components are included:  if $J_{real}$
-and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
-the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
-
-\subsection{Pre-conditioning}
-
-Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
-overflow and underflow issues in many realistic cases because both $U$ and $V$
-have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
-real number.
-This can be avoided with the appropriate reparameterization.
-We define the following pre-conditioned variables
-\begin{eqnarray}
-\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
-\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
-\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
-\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
-\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
-\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
-\end{eqnarray}
-and the set of variables
-\begin{eqnarray}
-\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
-\end{eqnarray}
-with the constraint
-\begin{eqnarray}
-\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
-\end{eqnarray}
-Using this parameterization, we find the following, numerically stable,
-algorithm
-\begin{eqnarray}
-    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
-    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
-D_{n,n} &=& A_{n,n} -
-    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
-    \nonumber\\
-\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
-    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
-\end{eqnarray}
-
-\subsection{Solving}
-
-After computing the Cholesky factorization of the matrix $K$, we can apply
-the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
-$\mathcal{O}(N)$ operations respectively.
-The log-determinant of $K$ is given by
-\begin{eqnarray}
-    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
-\end{eqnarray}
-where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
-
-The inverse of $K$ can be applied using the relationship
-\begin{eqnarray}
-\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
-\end{eqnarray}
-First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
-substitution algorithm
-\begin{eqnarray}
-    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
-    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
-    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
-\end{eqnarray}
-where $f_{0,j} = 0$ for all $j$.
-Using this result and the following backward substitution algorithm, we can
-compute $\bvec{z}$
-\begin{eqnarray}
-g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
-    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
-z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
-\end{eqnarray}
-where $g_{N+1,j} = 0$ for all $j$.
-The total cost of the forward and backward substitution passes scales as
-$\mathcal{O}(N\,J)$.
 
@@
 
-\section{celerite as a model of stellar variations}\sectlabel{sho}
-
-We now turn to a discussion of \celeriteterm\ as a model of astrophysical
-variability.
-A common concern in the context of GP modeling in astrophysics is the lack of
-physical motivation for the choice of kernel functions.
-Kernels are often chosen simply because they are popular with little
-consideration of the impact of this decision.
-While this isn't necessarily a problem for the inferred results, in this
-section we discuss an exact physical interpretation of the \celeriteterm\
-kernel that will be applicable to many astrophysical systems but especially
-the time-variability of stars.
-
-Many astronomical objects are variable on timescales determined by their
-physical structure.
-For phenomena such as stellar (asteroseismic) oscillations, variability is
-excited by noisy physical processes and grows most strongly at the
-characteristic timescale but is also damped due to dissipation in the system.
-These oscillations are strong at resonant frequencies determined by the
-internal stellar structure, which are both excited and damped by convective
-turbulence.
-
-To relate this physical picture to \celeriteterm, we consider the dynamics of a
-stochastically-driven damped simple harmonic oscillator (SHO).
-The differential equation for this system is
-\begin{equation}
-    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
-    + \omega_0^2\right]\, y(t) = \epsilon(t)
-\end{equation}
-where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
-quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
-force.
-If $\epsilon(t)$ is white noise, the PSD of this process is
-\citep{Anderson:1990}
-\begin{equation}\eqlabel{sho-psd}
-S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
-    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
-\end{equation}
-where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
-= \sqrt{2/\pi}\,S_0\,Q^2$.
-The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
-\begin{eqnarray}\eqlabel{sho-complex}
-a_j &=& S_0\,\omega_0\,Q \\
-b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
-c_j &=& \frac{\omega_0}{2\,Q}\\
-d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
-\end{eqnarray}
-for $Q \ge \frac{1}{2}$.
-For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of
-\celeriteterm\
-terms with parameters
-\begin{eqnarray}\eqlabel{sho-real}
-a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
-        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
-b_{j\pm} &=& 0 \nonumber\\
-    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
-    \nonumber\\
-d_{j\pm} &=& 0 \quad. \nonumber
-\end{eqnarray}
-
-For these definitions, the kernel is
-\begin{equation}\eqlabel{sho-kernel}
-k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_0) =
-S_0\,\omega_0\,Q\,e^{-\frac{\omega_0\,\tau}{2Q}}\,
-\begin{cases}
-    \cosh{(\eta\,\omega_0\,\tau)} +
-        \frac{1}{2\,\eta\,Q}\,\sinh{(\eta\,\omega_0\,\tau)}, & 0 < Q < 1/2\\
-    2\,(1+\omega_0\,\tau), & Q = 1/2\\
-    \cos{(\eta\,\omega_0\,\tau)} +
-        \frac{1}{2\,\eta\,Q} \sin{(\eta\,\omega_0\,\tau)},& 1/2 < Q\\
-\end{cases}
-\end{equation}
-where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
-We note that, because of the damping, the characteristic oscillation frequency
-in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
-the frequency of the undamped oscillator, $\omega_0$.
-
-The power spectrum in \eq{sho-psd} has several limits of physical interest:
-\begin{itemize}
-
-{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
-\begin{eqnarray}\eqlabel{granulation-psd}
-S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
-\end{eqnarray}
-This functional form is commonly used to model for the background granulation
-noise in asteoreseismic and helioseismic \citep{Harvey:1985, Michel:2009,
-Kallinger:2014}  analyses.
-The kernel for this value of $Q$ is
-\begin{equation}
-k(\tau) = S_0\,\omega_0\,e^{-\frac{1}{\sqrt{2}}\,\omega_0\,\tau}\,
-    \cos{\left(\frac{\omega_0\,\tau}{\sqrt{2}}-\frac{\pi}{4}\right)} \quad.
-\end{equation}}
-
-{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
-\begin{eqnarray}
-S(\omega) =
-    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
-\end{eqnarray}
-with the corresponding covariance function (using
-\eqalt{celerite-kernel} and \eqalt{sho-real})
-\begin{eqnarray}\eqlabel{approx-matern}
-k(\tau) &=& \lim_{f \to 0}\,
-    \frac{1}{2}\,S_0\,\omega_0\,
-    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
-          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
-    \right] \\
-&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
-\end{eqnarray}
-or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
-\begin{eqnarray}\eqlabel{approx-matern2}
-k(\tau) &=& \lim_{f \to 0}\,
-    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
-    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
-&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
-\end{eqnarray}
-This covariance function is also known as the Mat\'ern-3/2 function
-\citep{Rasmussen:2006}.
-This suggests that the Mat\'ern-3/2 covariance can be approximated using
-the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
-caution that this might lead to numerical issues for the solver.
-}
-
-{\item Finally, in the limit of large $Q$, the model approaches a high
-    quality oscillation with frequency $\omega_0$ and covariance function
-\begin{eqnarray}
-k(\tau) \approx
-    S_0\,\omega_0\,Q\,
-    \exp\left(-\frac{\omega_0\,\tau}{2\,Q}\right)\,
-    \cos\left(\omega_0\,\tau\right) \quad.
-\end{eqnarray}}
-
-\end{itemize}
-\Figure{sho} shows a plot of the PSD for these limits and several other values
-of $Q$.
-From this figure, it is clear that for $Q \le 1/2$, the model has no
-oscillatory behavior and that for large $Q$, the shape of the PSD near the
-peak frequency approaches a Lorentzian.
-
-These special cases demonstrate that the stochastically-driven simple harmonic
-oscillator provides a physically motivated model that is flexible enough to
-describe a wide range of stellar variations and we return to give some
-specific examples in \sect{examples-real}.
-Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
-model for asteroseismic oscillations.
-In practice, we take a sum over oscillators with different values of $Q$,
-$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
-of stellar time series.
-Since this kernel is exactly described by the exponential kernel, the
-likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
-measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
-described in the previous section.
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.95\textwidth]{figures/sho.pdf}
-\caption{(left) The power spectrum of a stochastically-driven simple harmonic
-    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
-    factor $Q$.
-    For comparison, the dashed line shows the Lorentzian function from
-    \eq{lorentz-psd} with $c_j = \omega_0/(2\,Q) = 1/20$ and normalized so that
-    $S(d_j)/S(0) = 100$.
-    (middle) The corresponding autocorrelation functions with the same colors.
-    \response{(right) Three realizations from each model in the same colors.}
-    \figurelabel{sho}}
-\end{center}
-\end{figure}
-
-\section{Examples}\sectlabel{examples}
-
-\response{
-To show the application of \celerite, we use it to perform posterior
-inference in two examples with realistic simulated data and three examples of
-real-world research problems.
-These examples have been chosen to justify the application of \celerite\ for a
-range of use cases, but they are by no means exhaustive.
-These examples are framed in the time domain with a clear bias in favor of
-large homogeneous photometric surveys, but \celerite\ can also be used for
-other one-dimensional problems like, for example, spectroscopy, where
-wavelength~--~instead of time~--~is the independent coordinate \citep[see][for
-a potential application]{Czekala:2017}.
-
-In the first example, we demonstrate that \celerite\ can be used to infer
-the power spectrum of a process when the data are generated from a
-\celeriteterm\ model.
-In the second example, we demonstrate that \celerite\ can be used as an
-effective model even if the true process cannot be represented in the space of
-allowed models.
-This is an interesting example because, when analyzing real data, we rarely
-have any fundamental reason to believe that the data were generated by a GP
-model with a specific kernel.
-Even in these cases, GPs can be useful effective models and \celerite\
-provides computational advantages over other GP methods.
-
-The next three examples show \celerite\ used to infer the properties of stars
-and exoplanets observed by NASA's \kepler\ Mission.
-Each of these examples touches on an active area of research so we limit our
-examples to be qualitative in nature and do not claim that \celerite\ is the
-optimal method, but we hope these examples encourage interested readers to
-investigate the applicability of \celerite\ to their research.
-
-In each example, the joint posterior probability density is given by
-\begin{eqnarray}\eqlabel{joint-post}
-p(\bvec{\theta},\,\bvec{\alpha}\,|\,\bvec{y},\,X) \propto
-p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})\,
-p(\bvec{\theta},\,\bvec{\alpha})
-\end{eqnarray}
-where $p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})$ is the GP
-likelihood defined in \eq{gp-likelihood} and $p(\bvec{\theta},\,\bvec{\alpha})$
-is the joint prior probability density for the parameters.
-We sample each posterior density using MCMC and investigate the performance of
-the model when used to interpret the physical processes that generated the
-data.
-The specific parameters and priors are discussed in detail in each section,
-but we generally assume separable uniform priors with plausibly broad support.
-}
-
-\subsection{Example 1: Recovery of a celerite process}\sectlabel{example1}
-
-In this first example, we simulate a dataset using a known \celeriteterm\
-process and fit it with \celerite\ to show that valid inferences can be made
-in this idealized case.
-\response{We simulated a small ($N = \exampleindata$) dataset using a GP model
-with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
-1\,\mathrm{ppm}^2$, $\omega_0 = e^2\,\mathrm{day}^{-1}$, and $Q = e^2$, where
-the units are arbitrarily chosen to simplify the discussion.
-We add further white noise with the amplitude $\sigma_n = 2.5\, \mathrm{ppm}$
-to each data point.
-The simulated data are plotted in the top left panel of
-\Figure{simulated-correct}.
-
-In this case, when simulating and fitting, we set the mean function
-$\mu_\bvec{\theta}$ to zero~--~this means that the parameter vector
-$\bvec{\theta}$ is empty~--~and the elements of the covariance matrix are
-given by \eq{sho-kernel} with three parameters $\bvec{\alpha} = (S_0,\,
-\omega_0,\,Q)$.
-We choose a proper separable prior for $\bvec{\alpha}$ with log-uniform
-densities for each parameter as listed in Table~\ref{tab:example-1-params}.
-
-To start, we estimate the maximum \emph{a posteriori} (MAP) parameters using
-the \project{L-BFGS-B} non-linear optimization routine \citep{Byrd:1995,
-Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
-The top left panel of \Figure{simulated-correct} shows the conditional mean
-and standard deviation of the MAP model over-plotted on the simulated data and
-the bottom panel shows the residuals away from this model.
-Since we are using the correct model to fit the data, it is reassuring that
-the residuals appear qualitatively uncorrelated in this figure.
-
-We then sample the joint posterior probability (\eqalt{joint-post}) using
-\emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
-We initialize \exampleinwalkers~walkers in a Gaussian ball with width
-$10^{-4}$ in each parameter around the MAP parameter vector, run
-\exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
-To assess convergence, we estimate the mean integrated autocorrelation time of
-the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
-chain results in \exampleineff~effective samples.
-
-Each sample in the chain corresponds to a model PSD and we compare this
-posterior constraint on the PSD to the true spectral density in the
-right-hand panel of \Figure{simulated-correct}.
-In this figure, the true PSD is plotted as a dashed black line and the
-numerical estimate of the posterior inference is shown as a blue contour
-indicating 68\% of the MCMC samples.
-It is clear from this figure that, as expected, the inference correctly
-reproduces the true PSD.
-}
-
-\input{figures/simulated/correct-params}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
-\caption{(top left) A simulated dataset (black error bars), and maximum
-    likelihood model (blue contours).
-    (bottom left) The residuals between the mean predictive model and the data
-    shown in the top left figure.
-    (right) The inferred PSD~--~the blue contours encompass 68\% of the
-    posterior mass~--~compared to the true PSD (dashed black line).
-    \figurelabel{simulated-correct}}
-\end{center}
-\end{figure}
-
-\subsection{Example 2: Inferences with the ``wrong'' model}\sectlabel{example2}
-
-In this example, we simulate a dataset using a known GP model with a kernel
-outside of the support of a \celeriteterm\ process.
-This means the true autocorrelation of the process can never be correctly
-represented by the model that we are using to fit, but we use this example
-to demonstrate that, at least in this case, valid inferences can still be made
-about the physical parameters of the model.
-
-The data are simulated from a quasiperiodic GP with the kernel
-\begin{eqnarray}\eqlabel{sim-wrong-true}
-k_\mathrm{true} (\tau) = A\,
-    \exp\left(-\frac{\tau^2}{2\,\lambda^2}\right)\,
-    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{true}}\right)
-\end{eqnarray}
-where $P_\mathrm{true}$ is the fundamental period of the process.
-This autocorrelation structure corresponds to the power spectrum
-\citep{Wilson:2013}
-\begin{eqnarray}
-S_\mathrm{true} (\omega) = \frac{\lambda\,A}{2}\,\left[
-    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega-
-        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right) +
-    \exp\left(-\frac{\lambda^2}{2}\,\left(\omega+
-        \frac{2\,\pi}{P_\mathrm{true}}\right)^2\right)
-\right]
-\end{eqnarray}
-which, for large values of $\omega$, falls off exponentially.
-When compared to \eq{celerite-psd}~--~which, for large $\omega$, goes as
-$\omega^{-4}$ at most~--~it is clear that a \celeriteterm\ model can never
-exactly reproduce the structure of this process.
-That being said, we demonstrate that robust inferences can be made about
-$P_\mathrm{true}$ even with this effective model.
-
-\response{We generate a realization of a GP model with the kernel given in
-\eq{sim-wrong-true} with $N = \exampleiindata$, $A = 1\,\mathrm{ppm}^2$,
-$\lambda = 5\,\mathrm{day}$, and $P_\mathrm{true} = 1\,\mathrm{day}$.
-We also add white noise with amplitude $\sigma = 0.5\,\mathrm{ppm}$ to each
-data point.
-The top left panel of \Figure{simulated-wrong} shows this simulated dataset.
-
-We then fit this simulated data using the product of two SHO terms
-(\eqalt{sho-kernel}) where one of the terms has $S_0 = 1$ and $Q =
-1/\sqrt{2}$ fixed.
-The kernel for this model is
-\begin{eqnarray}
-k(\tau) = k_\mathrm{SHO}(\tau;\,S_0,\,Q,\,\omega_1) \,
-    k_\mathrm{SHO}(\tau;\,S_0 = 1,\,Q = 1/\sqrt{2},\,\omega_2)
-\end{eqnarray}
-where $k_\mathrm{SHO}$ is defined in \eq{sho-kernel} and the period of the
-process is $P=2\,\pi/\omega_1$.
-We note that using \eq{product-rule}, the product of two \celeriteterm\ terms
-can also be expressed using \celerite.
-
-As in the previous example, we set the mean function to zero and can,
-therefore, omit the parameters $\bvec{\theta}$.
-Table~\ref{tab:example-2-params} lists the proper log-uniform priors that we
-choose for each parameter in $\bvec{\alpha} =
-(S_0,\,Q,\,\omega_1,\,\omega_2)$.
-These priors, together with the GP likelihood (\eqalt{gp-likelihood}) fully
-specify the posterior probability density.
-
-As above, we estimate the MAP parameters using \project{L-BFGS-B} and sample
-the posterior probability density using \emcee.
-The top left panel of \Figure{simulated-wrong} shows the conditional mean and
-standard deviation of the MAP model.
-The bottom left panel shows the residuals between the data and this MAP model
-and, even though this GP model is formally ``wrong'', there are no obvious
-correlations in these residuals.
-
-To perform posterior inference, we initialize \exampleiinwalkers~walkers by
-sampling from the 4-dimensional Gaussian centered on the MAP parameters with
-an isotropic standard deviation of $10^{-4}$.
-We then run \exampleiinburn~steps of burn-in and \exampleiinsteps~steps of
-MCMC.
-To estimate the number of effective samples, we estimate the integrated
-autocorrelation time of the chain for the parameter $\omega_1$~--~the
-parameter of primary interest~--~and find \exampleiineff~effective samples.
-
-For comparison, we run the same number of steps of MCMC to sample the
-``correct'' joint posterior density.
-For this reference inference, we use a GP likelihood with the kernel given by
-\eq{sim-wrong-true} and choose log-uniform priors on each of the
-three parameters $\ln A/\mathrm{ppm}^2 \sim \mathcal{U}(-10,\,10)$,
-$\ln \lambda/\mathrm{day} \sim \mathcal{U}(-10,\,10)$, and
-$\ln P_\mathrm{true}/\mathrm{day} \sim \mathcal{U}(-10,\,10)$.
-
-The marginalized posterior inferences of the characteristic period of the
-process are shown in the right panel of \Figure{simulated-wrong}.
-The inference using the correct model is shown as a dashed blue histogram and
-the inference made using the effective model is shown as a solid black
-histogram.
-These inferences are consistent with each other and with the true period used
-for the simulation (shown as a vertical gray line).
-This demonstrates that, in this case, \celerite\ can be used as a
-computationally efficient effective model and hints that this may be true in
-other problems as well.
-}
-
-\input{figures/simulated/wrong-qpo-params}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
-\caption{(top left) A simulated dataset (black error bars), and maximum
-    likelihood model (blue contours).
-    (bottom left) The residuals between the mean predictive model and the data
-    shown in the top left figure.
-    (right) The inferred period of the process. The true period is indicated
-    by the vertical orange line, the posterior inference using the correct
-    model is shown as the blue dashed histogram, and the inference made using
-    the ``wrong'' effective model is shown as the black histogram.
-    \figurelabel{simulated-wrong}}
-\end{center}
-\end{figure}
-
-\subsection{Example 3: Stellar rotation} \sectlabel{rotation}
-
-A source of variability that can be measured from time series measurements of
-stars is rotation.
-The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
-quasiperiodic variations in photometric or spectroscopic observations
-\citep{Dumusque:2014}.
-It has been demonstrated that for light curves with nearly uniform sampling,
-the empirical autocorrelation function provides a reliable estimate of the
-rotation period of a star \citep{Mcquillan:2013, Mcquillan:2014, Aigrain:2015}
-and that a GP model with a
-quasiperiodic covariance function can be used to make probabilistic
-measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
-The covariance function used for this type of analysis has the form
-\begin{eqnarray}\eqlabel{sine2}
-k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
-    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P_\mathrm{rot}} \right) \right)
-\end{eqnarray}
-where $P_\mathrm{rot}$ is the rotation period of the star.
-GP modeling with the same kernel function has been proposed as a method of
-measuring the mean periodicity in quasiperiodic photometric time series in
-general \citep{Wang:2012}.
-The key difference between \eq{sine2} and other quasiperiodic kernels is that
-it is positive for all values of $\tau$.
-We construct a simple \celeriteterm\ covariance function with similar
-properties as follows
-\begin{eqnarray}\eqlabel{rot-kernel}
-k(\tau) = \frac{B}{2+C}\,e^{-\tau/L}\,\left[
-    \cos\left(\frac{2\,\pi\,\tau}{P_\mathrm{rot}}\right) + (1 + C)
-\right]
-\end{eqnarray}
-for $B>0$, $C>0$, and $L>0$.
-The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
-but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
-as a drop-in replacement for a significant gain in computational efficiency.
-
-\response{GPs have been used to measure stellar rotation periods for
-individual datasets \citep[for example][]{Littlefair:2017}, but the
-computational cost of traditional GP methods has hindered the industrial
-application to existing surveys like \kepler\ with hundreds of thousands of
-targets.
-The increase in computational efficiency and scalability provided by
-\celerite\ opens the possibility of inferring rotation periods using GPs at
-scale of existing and forthcoming surveys like \kepler, \tess, and \lsst.
-}
-
-As a demonstration, we fit a \celeriteterm\ model with a kernel given
-by \eq{rot-kernel} to a \kepler\ light curve for the star KIC~1430163.
-This star has a published rotation period of $3.88 \pm 0.58\,\mathrm{days}$,
-measured using traditional periodogram and autocorrelation function approaches
-applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
-
-\response{We select about 180~days of contiguous observations of
-KIC~1430163 from \kepler.
-This dataset has \exampleiiindata\ measurements and, using a tuned linear
-algebra
-implementation\footnote{We use the Intel Math Kernel Library
-\url{https://software.intel.com/en-us/intel-mkl}}, a
-single evaluation of the likelihood requires over 8~seconds on a modern Intel
-CPU.
-This calculation, using \celerite\ with the model in \eq{rot-kernel}, only
-takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
-magnitude per model evaluation.
-
-We set the mean function $\mu_\bvec{\theta}$ to zero and the remaining
-parameters $\bvec{\alpha}$ and their priors are listed in
-Table~\ref{tab:example-3-params}.
-As with the earlier examples, we start by estimating the MAP parameters using
-\project{L-BFGS-B} and initialize \exampleiiinwalkers~walkers by sampling from
-an isotropic Gaussian with a standard deviation of $10^{-5}$ centered on the
-MAP parameters.
-The left panels of \Figure{rotation} show a subset of the data used in this
-example and the residuals away from the MAP predictive mean.
-
-We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
-of MCMC using \emcee.
-We estimate the integrated autocorrelation time of the chain for $\ln
-P_\mathrm{rot}$ and estimate that we have \exampleiiineff\ independent
-samples.
-These samples give a posterior constraint on the period of $P_\mathrm{rot} =
-\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
-in the right panel of \Figure{rotation}.
-This result is in good agreement with the literature value with smaller
-uncertainties.
-A detailed comparison of GP rotation period measurements and the traditional
-methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
-\etal, submitted) demonstrates that GP inferences are, at a population level,
-more reliable than other methods.
-}
-
-\input{figures/rotation/rotation-params}
-
-\begin{figure}[p]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
-\caption{Inferred constraints on a quasiperiodic GP model using the covariance
-    function in \eq{rot-kernel} and two quarters of \kepler\ data.
-(top left) The \kepler\ data (black points) and the maximum likelihood model
-    prediction (blue curve) for a 60~day subset of the data used.
-    The solid blue line shows the predictive mean and the blue contours show
-    the predictive standard deviation.
-(bottom left) The residuals between the mean predictive model and the data
-    shown in the top left figure.
-(right) The posterior constraint on the rotation period of KIC~1430163 using
-    the dataset and model from \Figure{rotation}.
-    The period is the parameter $P_\mathrm{rot}$ in \eq{rot-kernel} and this
-    figure shows the posterior distribution marginalized over all other
-    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
-    published rotation period made using the full
-    \kepler\ baseline shown as a vertical gray line \citep{Mathur:2014}.
-    \figurelabel{rotation}}
-\end{center}
-\end{figure}
-
-\subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}
-
-The asteroseismic oscillations of thousands of stars were measured using light
-curves from the \kepler\ Mission \citep{Gilliland:2010, Huber:2011,
-Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
-driver for many of the upcoming large scale photometric surveys
-\citep{Campante:2016, Rauer:2014, Gould:2015}.
-Most asteroseismic analyses have been limited to high signal-to-noise
-oscillations because the standard methods use statistics of the empirical
-periodogram and cannot formally propagate the measurement uncertainties to the
-constraints on physical parameters, instead relying on bootstrapped
-uncertainty estimators \citep{Huber:2009}.
-More sophisticated methods that compute the likelihood function in the time
-domain scale poorly to large survey datasets \citep{Brewer:2009,
-Corsaro:2014}.
-
-\celerite\ alleviates these problems by providing a physically motivated
-probabilistic model that can be evaluated efficiently even for large datasets.
-In practice, we model the star as a mixture of stochastically-driven simple
-harmonic oscillators where the amplitudes and frequencies of the oscillations
-are computed using a physical model, and evaluate the probability of the
-observed time series using a GP where the PSD is a sum of terms given by
-\eq{sho-psd}.
-This gives us a method for computing the likelihood function for the
-parameters of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta
-\nu$, or other more fundamental parameters) \emph{conditioned on the observed
-time series} in $\mathcal{O}(N)$ operations.
-In other words, \celerite\ provides a computationally efficient framework that
-can be combined with physically-motivated models of stars and numerical
-inference methods to make rigorous probabilistic measurements of asteroseismic
-parameters in the time domain.
-This has the potential to push asteroseismic analysis to lower signal-to-noise
-datasets and we hope to revisit this idea in a future paper.
-
-To demonstrate the method, we use a simple heuristic model where the PSD is
-given by a mixture of 8 components with amplitudes and frequencies specified
-by $\nu_\mathrm{max}$, $\Delta \nu$, and some nuisance parameters.
-The first term is used to capture the granulation ``background''
-\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
-$S_g$ and $\omega_g$.
-The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
-parameter shared between terms and the frequencies are given by
-\begin{eqnarray}
-\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
-\end{eqnarray}
-and the amplitudes are given by
-\begin{eqnarray}
-S_{0,\,j} =
-    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
-\end{eqnarray}
-where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
-are shared nuisance parameters.
-\response{Finally, we also fit for the amplitude of the white noise by adding
-a parameter $\sigma$ in quadrature with the uncertainties given for the
-\kepler\ observations.
-All of these parameters and their chosen priors are listed in
-Table~\ref{tab:example-4-params}.
-As before, these priors are all log-uniform except for $\epsilon$ where we use
-a zero-mean normal prior with a broad variance of $1~\mathrm{day}^2$ to break
-the degeneracy between $\nu_\mathrm{max}$ and $\epsilon$.
-}
-To build a more realistic model, this prescription could be extended to
-include more angular modes, or $\nu_\mathrm{max}$ and $\Delta \nu$ could be
-replaced by the fundamental physical parameters of the star.
-
-To demonstrate the applicability of this model, we apply it to infer the
-asteroseismic parameters of the giant star KIC~11615890, observed by the
-\kepler\ Mission.
-The goal of this example is to show that, even for a low signal-to-noise
-dataset with a short baseline, it is possible to infer asteroseismic
-parameters with formal uncertainties that are consistent with the parameters
-inferred with a much larger dataset.
-Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
-$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
-compare our results to the results inferred from the full 4 year baseline of
-the \kepler\ Mission.
-For KIC~11615890, the published asteroseismic parameters measured using several
-years of \kepler\ observations are \citep{Pinsonneault:2014}
-\begin{eqnarray}
-    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
-    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
-\end{eqnarray}
-
-Unlike typical giants, KIC~11615890 is a member of a class of stars where the
-dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
-in the core \citep{Stello:2016}.
-This makes this target simpler for the purposes of this demonstration because
-we can neglect the $\ell = 1$ modes and the model proposed above will be an
-effective model for the combined signal from the $\ell = 0$ and 2 modes.
-
-\response{
-For this demonstration, we randomly select a month-long segment of PDC
-\citep{Stumpe:2012, Smith:2012} \kepler\ data.
-Unlike the previous examples, the posterior distribution is sharply
-multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
-not practical.
-Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
-$\Delta\nu$ using only the month-long subset of data and following the
-standard procedure described by \citet{Huber:2009}.
-We then run set of \project{L-BFGS-B} optimizations with values of
-$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
-estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
-using the empirical relationship between these two quantities
-\citep{Stello:2009}.
-This initialization procedure is sufficient for this example, but the general
-application of this method will require a more sophisticated prescription.
-
-\Figure{astero} shows a 10~day subset of the dataset used for this example.
-The MAP model is overplotted on these data and the residuals away from the
-mean prediction of this model are shown in the bottom panel of
-\Figure{astero}.
-There is no obvious structure in these residuals, lending some credibility to
-the model specification.
-
-We initialize \exampleivnwalkers~walkers by sampling from an isotropic
-Gaussian centered on the MAP parameters (the full set of parameters and their
-priors are listed in Table~\ref{tab:example-4-params}), run
-\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
-\emcee.
-We estimate the mean autocorrelation time for the chains of
-$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
-samples from the marginalized posterior density.
-\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
-and $\Delta\nu$ compared to the results from the literature.
-This result is consistent within the published error bars and the posterior
-constraints are tighter than the published results.
-All asteroseismic analyses are known to have substantial systematic and
-method-dependent uncertainties \citep{Verner:2011} so further experiments
-would be needed to fully assess the reliability of this method.
-
-This model requires about 10~CPU minutes to run the MCMC to convergence.
-This is more computationally intensive than traditional methods of measuring
-asteroseismic oscillations, but it is much cheaper than the same analysis
-using a general direct GP solver.
-For comparison, we estimate that repeating this analysis using a general
-Cholesky factorization implemented as part of a tuned linear algebra
-library\footnote{We use the Intel Math Kernel Library
-\url{https://software.intel.com/en-us/intel-mkl}} would require about 15~CPU
-hours.
-An in-depth discussion of the benefits of rigorous probabilistic inference of
-asteroseismic parameters in the time domain is beyond the scope of this paper,
-but we hope to revisit this opportunity in the future.
-}
-
-\input{figures/astero/astero-params}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero/astero.pdf}
-\caption{
-\response{
-(top) The \kepler\ data (black points) and the maximum likelihood model
-    prediction (blue curve) for a 10~day subset of the month-long dataset that
-    was used for the fit.
-    The solid blue line shows the predictive mean and the blue contours show
-    the predictive standard deviation.
-(bottom) The residuals between the mean predictive model and the data
-    shown in the top figure.
-}
-    \figurelabel{astero}}
-\end{center}
-\end{figure}
-
-\begin{figure}[!p]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero/astero-corner.pdf}
-\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
-    from the inference shown in \Figure{astero} compared to the published
-    value (error bar) based on several years of \kepler\ observations
-    \citep{Pinsonneault:2014}.
-    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
-    regions in the marginalized planes and the histograms along the diagonal
-    show the marginalized posterior for each parameter.
-    \figurelabel{astero-corner}}
-\end{center}
-\end{figure}
-
-\begin{figure}[!p]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{figures/astero/astero-comp.pdf}
-\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
-posterior inference of the PSD as a mixture of stochastically-driven simple
-harmonic oscillators.
-(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
-    on the full four year baseline of the mission. The orange line shows a
-    smoothed periodogram and the blue line indicates the level of the
-    measurement uncertainties.
-(middle) The same periodogram computed using about a month of data.
-(bottom) The power spectrum inferred using the mixture of SHOs model described
-    in the text and only one month of \kepler\ data.
-    The black line shows the median of posterior PSD and the gray contours
-    show the 68\% credible region.
-    \figurelabel{astero}}
-\end{center}
-\end{figure}
-
-\subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}
-
-\response{
-This example is different from all the previous examples~--~both simulated and
-real~--~because, in this case, do not set the deterministic
-mean function $\mu_\bvec{\theta}$ to zero.
-Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
-physically significant function and the parameters are fundamental properties
-of the system.
-This is an example using real data~--~because we use the light curve from
-\sect{rotation}~--~but we multiply these data by a simulated transiting
-exoplanet model with known parameters.
-This allows us to show that we can recover the true parameters of the planet
-even when the transit signal is superimposed on the real variability of a
-star.
-}
-GP modeling has been used extensively for this purpose throughout the
-exoplanet literature \citep[for example][]{Dawson:2014, Barclay:2015,
-Evans:2015, Foreman-Mackey:2016b, Grunblatt:2016}.
-
-In \eq{gp-likelihood} the physical parameters of the exoplanet are called
-$\bvec{\theta}$ and, in this example, the mean function $\mu_\bvec{\theta}(t)$
-is a limb-darkened transit light curve \citep{Mandel:2002,
-Foreman-Mackey:2016a} and the parameters $\bvec{\theta}$ are the orbital
-period $P_\mathrm{orb}$, the transit duration $T$, the phase or epoch $t_0$,
-the impact parameter $b$, the radius of the planet in units of the stellar
-radius $R_P/R_\star$, the baseline relative flux of the light curve $f_0$, and
-two parameters describing the limb-darkening profile of the star
-\citep{Claret:2011, Kipping:2013}.
-As in \sect{rotation}, we model the stellar variability using a GP model
-with a kernel given by \eq{rot-kernel} and fit for the parameters of the
-exoplanet $\bvec{\theta}$ and the stellar variability $\bvec{\alpha}$
-simultaneously.
-\response{The full set of parameters $\bvec{\alpha}$ and $\bvec{\theta}$ are
-listed in Table~\ref{tab:example-5-params} along with their priors and the
-true values for $\bvec{\theta}$.}
-
-\response{ We take a 20~day segment of the \kepler\ light curve for
-KIC~1430163 ($N = \examplevndata$) and multiply it by a simulated transit
-model with the parameters listed in Table~\ref{tab:example-5-params}.
-Using these data, we maximize the joint posterior defined by the likelihood in
-\eq{gp-likelihood} and the priors in Table~\ref{tab:example-5-params} using
-\project{L-BFGS-B} for all the parameters $\bvec{\alpha}$ and $\bvec{\theta}$
-simultaneously.
-The top panel of \Figure{transit-ml} shows the data including the simulated
-transit as black points with the MAP model prediction over-plotted in blue.
-The bottom panel of \Figure{transit-ml} shows the ``de-trended'' light curve
-where the MAP model has been subtracted and the MAP mean model
-$\mu_\bvec{\theta}$ has been added back in.
-For comparison, the transit model is over-plotted in the bottom panel of
-\Figure{transit-ml} and we see no obvious correlations in the residuals.
-
-Sampling \examplevnwalkers~walkers from an isotropic Gaussian centered on the
-MAP parameters, we run \examplevnburn~steps of burn-in and
-\examplevnsteps~steps of production MCMC using \emcee.
-We estimate the integrated autocorrelation time for the $\ln(P_\mathrm{orb})$
-chain and find \examplevneff~effective samples across the full chain.
-\Figure{transit-corner} shows the marginalized posterior constraints on the
-physical properties of the planet compared to the true values.
-Even though the \celeriteterm\ representation of the stellar variability is
-only an effective model, the inferred distributions for the physical
-properties of the planet $\bvec{\theta}$ are consistent with the true values.
-This promising result suggests that \celerite\ can be used as an effective
-model for transit inference for a significant gain in computational
-tractability.
-Even for this example with small $N$, the cost of computing the likelihood
-using \celerite\ is nearly two orders of magnitude faster than the same
-computation using a general direct solver.
-}
-
-\input{figures/transit/transit-params-edited}
-
-\begin{figure}[!htbp]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/transit/transit-ml.pdf}
-    \caption{\emph{(top)} A month-long segment of \kepler\ light curve for
-    KIC~1430163 with a synthetic transit model injected (black points) and the
-    maximum likelihood model for the stellar variability
-    (blue line).
-    \emph{(bottom)} The maximum likelihood ``de-trending'' of the data in
-    the top panel.
-    In this panel, the maximum likelihood model for the stellar variability
-    has been subtracted to leave only the transits.
-    The de-trended light curve is shown by black error bars and the maximum
-    likelihood transit model is shown as a blue line.
-    \figurelabel{transit-ml}}
-\end{center}
-\end{figure}
-
-\begin{figure}[!p]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/transit/transit-corner.pdf}
-\caption{The marginalized posterior constraints on the physical parameters of
-    the planet transit in the light curve shown in the top panel of
-    \Figure{transit-ml}.
-    The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
-    regions in the marginalized planes and the histograms along the diagonal
-    show the marginalized posterior for each parameter.
-    The true values used in the simulation are indicated by blue lines.
-    For each parameter, the inference is consistent with the true value.
-    \figurelabel{transit-corner}}
-\end{center}
-\end{figure}
-
-\subsection{Summary}
-
-\response{
-In this section, we demonstrated five potential use cases for \celerite.
-These examples span a range of data sizes and model complexities, and the
-last three are all based on active areas of research with immediate real-world
-applicability.
-The sizes of these datasets are modest on the scale of some existing and
-forthcoming survey datasets because we designed the experiments to be easily
-reproducible but, even in these cases, the inferences made here would be
-intractable without substantial computational investment.
-
-Table~\ref{tab:example-stats} lists the specifications of each example.
-
-In Example~3, for example, we inferred the rotation period of a star observed
-by \kepler\ with a cost of about 4~CPU-minutes.
-By contrast, the same inference using a general but optimized Cholesky
-factorization routine would require nearly 400~CPU-hours.
-The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
-}
-
-\input{figures/example-stats}
-
-\section{Comparisons to other methods}\sectlabel{compare}
-
-There are other methods of scaling GP models to large datasets and in this
-section we draw comparisons between \celerite\ and other popular methods.
-Scalable GP methods tend to fall into two categories: approximate and
-restrictive.
-\celerite\ falls into the latter category because, while the method is exact,
-it requires a specific choice of stationary kernel function and it can only be
-used in one-dimension.
-
-\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}
-
-\response{
-The continuous autoregressive moving average (CARMA) models introduced into
-the astrophysics literature by \citet{Kelly:2014} share many features with
-\celerite.
-CARMA models are derived by solving a stochastic differential equation and,
-like \celerite, the likelihood function for a CARMA model conditioned on a
-one-dimensional dataset can be solved in $\mathcal{O}(N)$.
-The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
-\begin{eqnarray}\eqlabel{carma-kernel}
-k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
-\end{eqnarray}
-where
-\begin{eqnarray}\eqlabel{carma-coeff}
-A_j = \sigma^2 \,\frac{\left[\sum_{k=0}^K\beta_k\,{(r_j)}^k\right]\,
-    \left[\sum_{k=0}^K\beta_k\,{(-r_j)}^k\right]}
-    {-2\,\mathrm{Re}(r_j)\,\prod_{k=1,\,k \ne j}^{J}(r_k-r_j)({r_k}^*+r_j)}
-\end{eqnarray}
-and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
-the model.
-Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
-every CARMA model corresponds to an equivalent \celeriteterm\ model and the
-parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
-The inverse statement is not as simple.
-In practice, this means that \celerite\ could be trivially used to compute any
-CARMA model.
-Using the CARMA solver to compute a \celeriteterm\ model, however, requires
-solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
-While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
-the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
-is, in practice, somewhat faster and the memory requirements are smaller.
-One benefit of the \celeriteterm\ model is that it is parameterized in terms
-of the amplitudes directly.
-This means that the physical interpretation of these parameters is, in some
-cases, simpler.
-On the other hand, CARMA models are parameterized by modeling the power
-spectrum of the stochastic driving force and this might be the quantity of
-interest in other situations.
-Another benefit of the \celeriteterm\ framework is that, unlike
-filtering-based CARMA solvers, it directly solves matrix equations of the form
-${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
-This is a crucial feature when GP modeling is combined with linear models
-\citep[for example][]{Luger:2017}.
-}
-
-\subsection{Other scalable methods}
-
-%As discussed previously, the \celerite\ method is based upon a method
-%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
-%In the case where the kernel function has a single term of the form given in
-%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
-%and the Rybicki \& Press method.
-%Their method does not require solving the extended matrix because the matrix
-%inverse can be calculated analytically, but the method is made slightly more
-%complicated when white noise is included.
-%We implement their full method and find empirically that it is more
-%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
-%but we note that the requirement of a single, real, exponential kernel is
-%extremely restrictive and this model cannot describe quasi-periodic
-%variability.
-
-Another popular method uses the fact that, in the limit of evenly-spaced data
-and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
-\citep[for example][]{Dillon:2013}.
-There are exact methods for solving Toeplitz matrix equations that scale as
-$\mathcal{O}(N\,\log N)$ and methods for computing determinants exactly in
-$\mathcal{O}(N^2)$ or approximately in $\mathcal{O}(N\,\log N)$
-\citep{Wilson:2014}.
-The Toeplitz method is, in some ways, more flexible than \celerite\ because it
-can be used with any stationary kernel, but it requires uniformly spaced data
-and the scaling is worse than \celerite\ so it is less efficient when applied
-to large datasets.
-
-\citet{Carter:2009} improved the scaling of Toeplitz methods by introducing a
-wavelet-based method for computing a GP likelihood with $\mathcal{O}(N)$
-scaling.
-This method has been widely applied in the context of exoplanet transit
-characterization, but it requires evenly spaced observations and the power
-spectrum of the process must have the form $S(\omega)\propto \omega^{-1}$ to
-gain the computational advantage.
-This wavelet method has been demonstrated to improve parameter estimation for
-transiting exoplanets \citep{Carter:2009}, but these strict requirements make
-this method applicable for only a limited set of use cases.
-
-Another GP method that has been used extensively in astronomy is the
-hierarchical off-diagonal low rank \citep[HODLR,][]{Ambikasaran:2016} solver.
-This method exploits the fact that many commonly used kernel functions produce
-``smooth'' matrices to approximately compute the GP likelihood with the
-scaling $\mathcal{O}(N\,\log^2 N)$.
-This method has the advantage that, unlike \celerite, it can be used with any
-kernel function but, in practice, the cost can still prove to be prohibitively
-high for multi-dimensional inputs.
-The proportionality constant in the $N\,\log^2N$ scaling of the HODLR method
-is a function of the specific kernel and we find~--~using the \project{george}
-software package \citep{Foreman-Mackey:2014, Ambikasaran:2016}~--~that this
-scales approximately linearly with $J$, but it requires substantial overhead
-for small models.
-This means that the HODLR solver can \emph{approximately} evaluate
-\celeriteterm\
-models more efficiently than \celerite\ for large models $J \gg 10$ and
-small datasets where $N$ is smaller than a few thousand.
-
-The structured kernel interpolation \citep[SKI/KISS-GP][]{Wilson:2015}
-framework is another approximate method that can be used to scale GPs to large
-datasets.
-This method uses fast interpolation of the kernel into a space where Toeplitz
-or Kronecker structure can be used to scale inference and prediction.
-The SKI/KISS-GP framework can be applied to scale GP inference with a wide
-range of kernels, but the computational cost will depend on the specific
-dataset, model, and precision requirements for the approximation.
-The SKI method has an impressive $\mathcal{O}(1)$ cost for test-time
-predictions and it is interesting to consider how this could be applied to
-\celeriteterm\ models.
-
-Many other approximate methods for scaling GP inference exist \citep[see, for
-example,][and references therein]{Wilson:2015a} and we make no attempt to make
-our discussion exhaustive.
-The key takeaway here is that \celerite\ provides an \emph{exact} method for
-GP inference for a specific, but flexible, class of one-dimensional kernel
-functions.
-Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
-stochastically-driven, damped simple harmonic oscillators, they are a
-physically motivated choice of covariance function in many applications.
-
-\section{Summary}
-
-Gaussian Process models have been fruitfully applied to many problems in
-astronomical data analysis, but the fact that the computational cost scales as
-the cube of the number of data points has limited their use to relatively
-small datasets.
-With the linear scaling of \celerite\, we envision the application of Gaussian
-processes may be expanded to much larger datasets.
-Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
-number of components it is flexible enough to describe a wide range of
-astrophysical variability.
-In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
-oscillator matches simple models of astrophysical variability, and makes the
-parameterization interpretable in terms of resonant frequency, amplitude, and
-quality factor.
-% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
-A drawback of this method is its quadratic scaling with the number of
-terms $J$, but, in many cases, small values of $J$ are sufficient.
-
-Our background is in studying transiting exoplanets, a field which has only
-recently begun to adopt full covariance matrices in analyzing the noise in
-stellar light curves when detecting or characterizing transiting planets
-\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
-Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
-All of these analyses have been limited to small datasets or restrictive
-kernel choices.
-\celerite\ weakens these requirements by providing a scalable method for
-computing the likelihood and a physical motivation for the choice of kernel.
-\celerite\ can be used to model stellar oscillations using the relation to the
-mixture of stochastically-driven, damped simple harmonic oscillators.
-As higher signal-to-noise observations of transiting exoplanet systems are
-obtained, the effects of stellar variability will more dramatically impact the
-correct inference of planetary transit parameters.
-We expect that \celerite\ will be important for transit detection
-\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
-Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
-\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
-functions \citep{Knutson:2007, Zucker:2007}, and more.
-
-Beyond these applications to model stellar variability, the method is
-generally applicable to other one-dimensional GP models.
-Accreting black holes show time series which may be modeled using a GP
-\citep{Kelly:2014}; indeed, this was the motivation for the original technique
-developed by Rybicki \& Press \citep{Rybicki:1992, Rybicki:1995}.
-This approach may be broadly used for characterizing quasar variability
-\citep{MacLeod:2010}, measuring time lags with reverberation mapping
-\citep{Zu:2011, Pancoast:2014}, modeling time delays in multiply-imaged
-gravitationally-lensed systems \citep{Press:1998}, characterizing
-quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
-classification of variable objects \citep{Zinn:2016}.
-We expect that there are also applications beyond astronomy.
-
-The \celeriteterm\ formalism can also be used for power spectrum estimation and
-quantification of its uncertainties.
-In principle, a large number of \celeriteterm\ terms could be used to perform
-non-parametric probabilistic inference of the power spectrum despite
-unevenly-spaced data with heteroscedastic noise \citep[for
-example,][]{Wilson:2013, Kelly:2014}.
-This type of analysis will be limited by the quadratic scaling of \celerite\
-with the number of terms $J$, but this limits existing methods as well
-\citep[CARMA models,][]{Kelly:2014}.
-In general, we encourage the use of physically motivated models for parameter
-estimation instead of qualitative modeling of the power spectrum itself.
-
-There are many data analysis problems where \celerite\ will not be applicable.
-In particular, the restriction to one-dimensional problems is significant.
-There are many examples of multidimensional GP modeling the astrophysics
-literature \citep[recent examples from the field of exoplanet characterization
-include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
-be used to accelerate any of these analyses.
-It is plausible that an extension could be derived to tackle some
-multidimensional problems with the right structure~--~simultaneous parallel
-time series, for example~--~and we hope to revisit this possibility in future
-work.
-
-Alongside this paper, we have released a well-tested and documented open
-source software package that implements the method and all of the examples
-discussed in these pages.
-This software is available on GitHub
-\url{https://github.com/dfm/celerite}\footnote{This version of the paper was
-generated with git commit \texttt{\githash} (\gitdate).} and Zenodo
-\citep{Foreman-Mackey:2017}, and it is made available under the MIT license.
-
-\acknowledgments
-It is a pleasure to thank
-Megan Bedell,
-Ian Czekala,
-Will Farr,
-Sam Grunblatt,
-David W.\ Hogg,
-Dan Huber,
-Meredith Rawls,
-Dennis Stello,
-Jake VanderPlas, and
-Andrew Gordon Wilson
-for helpful discussions informing the ideas and code presented here.
-\response{We would also like to thank the anonymous referee for thorough and
-constructive feedback that greatly improved the paper.}
-
-This work was performed in part under contract with the Jet Propulsion
-Laboratory (JPL) funded by NASA through the Sagan Fellowship Program executed
-by the NASA Exoplanet Science Institute.
-EA acknowledges support from NASA grants NNX13AF20G, NNX13A124G, NNX13AF62G,
-from National Science Foundation (NSF) grant AST-1615315, and from
-NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
-by NASA under cooperative agreement NNH05ZDA001C.
-
-This research made use of the NASA \project{Astrophysics Data System} and the
-NASA Exoplanet Archive.
-The Exoplanet Archive is operated by the California Institute of Technology,
-under contract with NASA under the Exoplanet Exploration Program.
-
-This paper includes data collected by the \kepler\ Mission. Funding for the
-\kepler\ Mission is provided by the NASA Science Mission directorate.
-We are grateful to the entire \kepler\ team, past and present.
-These data were obtained from the Mikulski Archive for Space Telescopes
-(MAST).
-STScI is operated by the Association of Universities for Research in
-Astronomy, Inc., under NASA contract NAS5-26555.
-Support for MAST is provided by the NASA Office of Space Science via grant
-NNX13AC07G and by other grants and contracts.
-
-This research made use of Astropy, a community-developed core Python package
-for Astronomy \citep{Astropy-Collaboration:2013}.
-
-\facility{Kepler}
-\software{%
-     \project{AstroPy} \citep{Astropy-Collaboration:2013},
-     \project{corner.py} \citep{Foreman-Mackey:2016},
-     \project{Eigen} \citep{Guennebaud:2010},
-     \project{emcee} \citep{Foreman-Mackey:2013},
-     \project{george} \citep{Ambikasaran:2016},
-     \project{LAPACK} \citep{Anderson:1999},
-     \project{matplotlib} \citep{Hunter:2007},
-     \project{numpy} \citep{Van-Der-Walt:2011},
-     \project{transit} \citep{Foreman-Mackey:2016a},
-     \project{scipy} \citep{Jones:2001}.
-}
-
-\newpage
-\appendix
-
-\section{An alternative solver using the extended matrix formalism}
-
-\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
-elements given by \eq{kernel-simple}, could be computed
-efficiently by taking advantage of the structure of this covariance function.
-\citet{Ambikasaran:2015} generalized this computation to apply to the full
-mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
-efficient method for computing the determinant of $K$.
-
-\subsection{An example}
-
-To provide some insight for this method, we follow
-\citet{Ambikasaran:2015} and start by working through a simple example.
-In this case, we assume that we have three data points
-$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
-variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
-like to compute the likelihood of these data under a GP model with the
-covariance function
-\begin{eqnarray}
-k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
-\end{eqnarray}
-This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and it is the model studied by
-\citet{Rybicki:1995}.
-To demonstrate that the likelihood of this model can be computed in
-$\mathcal{O}(N)$, we write out the full system of
-equations that must be solved to apply the inverse of $K$ and compute
-the first term of \eq{gp-likelihood}.
-In matrix notation, this is
-\begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{z} &=& \bvec{y}, \\
-\begin{pmatrix}
-    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
-    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
-    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ z_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ y_2 \\ y_3
-\end{pmatrix}
-\end{eqnarray}
-where our goal is to solve for the unknown vector \bvec{z} for a given matrix
-$K$ and vector \bvec{y}.
-In \eq{impl-matrix}, we have assumed that the mean function is zero but a
-non-zero mean could be included by replacing \bvec{y} by
-$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
-Now, if we introduce the variables
-\begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
-\end{eqnarray}
-where $g_{0} = 0$, and
-\begin{eqnarray}\eqlabel{algo-first}
-    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
-\end{eqnarray}
-where $u_{N} = 0$,
-the system of equations can be rewritten as
-\begin{eqnarray}
-(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
-a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
-\end{eqnarray}
-Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
-as a matrix equation shows the benefit that this seemingly trivial
-reformulation provides:
-\begin{eqnarray}
-\begin{pmatrix}
-    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
-    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
-    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
-    0 & 0 & 0 & 0 & -1 & 0 & a \\
-    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
-\end{pmatrix}\nonumber
-\end{eqnarray}
-Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
-rewrite \eq{impl-matrix} as
-\begin{eqnarray}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
-\end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
-with, it is now sparse with banded structure that can be exploited to solve
-the system efficiently.
-In particular, sparse solvers are available that can perform a
-LU-decomposition of matrices like this in $\mathcal{O}(N)$
-operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
-general~--~and we can use these algorithms to solve our system
-exactly because the target vector $\bvec{z}$ is a subset of the elements of
-$\bvec{z}_\mathrm{ext}$.
-
-In the following section we discuss this method more generally, but it's
-worth noting a few important facts that can already be seen in this example.
-First, the fundamental reason why this matrix $K$ can be solved efficiently is
-the following property of exponentials
-\begin{eqnarray}
-    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
-    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
-    e^{-c\,(t_3 - t_1)}
-\end{eqnarray}
-and it is important to note that this property does not extend to other common
-covariance functions like the ``exponential-squared'' function
-\begin{eqnarray}
-    k(\tau) \propto e^{-c\,\tau^2} \quad.
-\end{eqnarray}
-Second, our derivation of the extended matrix requires that the data
-points be monotonically sorted.
-This is the motivation for our previous statement that this method can only be
-applied in one dimension since neither of these properties will be satisfied
-by multidimensional inputs.
-
-\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
-extended matrix formalism in practice.
-First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with a bandwidth that scales linearly
-with the number of components, $J$.
-Second, \citet{Ambikasaran:2015} proved that the absolute value of the
-determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
-determinant of $K$.
-This means we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(N)$ operations.
-
-\subsection{The algorithm}
-
-In this section, we generalize the method from the previous section to the
-covariance function given by \eq{celerite-kernel}.
-This derivation follows \citet{Ambikasaran:2015} but it includes explicit
-treatment of complex parameters, and their complex conjugates.
-
-In the case of the full \celeriteterm\ covariance function
-(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
-analogy to the $u_n$ and $g_n$ that we introduced in the previous section
-\begin{eqnarray}\eqlabel{full-system-first}
-\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
-\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
-g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
-h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
-u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
-v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
-\end{eqnarray}
-with the boundary conditions
-\begin{eqnarray}
-    g_{0,j} = 0 \quad, \quad
-    h_{0,j} = 0 \quad, \quad
-    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
-    v_{N,j} = 0
-\end{eqnarray}
-for all $j$.
-Using these variables and some algebra, we find that the following expression
-\begin{eqnarray}\eqlabel{full-system-last}
-\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
-+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
-+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
-    &=& r_{\bvec{\theta},n}
-\end{eqnarray}
-is equivalent to the target matrix equation
-\begin{eqnarray}
-K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
-\end{eqnarray}
-if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
-$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
-\eq{full-system-first} through \eq{full-system-last} define a banded matrix
-equation in the ``extended'' space
-\begin{eqnarray}\eqlabel{kext-full}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
-    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
-\end{eqnarray}
-and, as before, this can be used to solve
-for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
-operations.
-\Figure{matrix} shows a pictorial representation of the sparsity pattern of
-the extended matrix $K_\mathrm{ext}$.
-The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
-with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
-non-zero entries.
-Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
-$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
-as
-\begin{eqnarray}
-{\bvec{z}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
-    h_{N-1,j} & z_N
-\end{array}\right)
-\end{eqnarray}
-and
-\begin{eqnarray}
-{\bvec{r}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},N}
-\end{array}\right) \quad.
-\end{eqnarray}
-After constructing the extended matrix (using a compact storage format), the
-extended matrix can be factorized using a LU-decomposition\footnote{Even
-though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
-Cholesky solver cannot be used for increased efficiency.} routine optimized
-for band or sparse matrices.
-This decomposition can then be used to compute the determinant of $K$, solve
-$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
-likelihood in \eq{gp-likelihood}.
-
-In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
-case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
-will also be identically zero for all $n$.
-Removing these trivial rows from the extended matrix results in factor of two
-increase in the computational efficiency for these terms.
-In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
-$d_j=0$, and the general terms are called ``complex''.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
-\caption{A pictorial representation of the sparse extended matrix
-$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
-Each colored block corresponds to a non-zero entry in the matrix as described
-in the legend.
-    \figurelabel{matrix}}
-\end{center}
-\end{figure}
-
-\subsection{Implementation considerations \& scaling}
-
-The extended system defined in the previous section is sparse, with typically
-fewer than a few percent non-zero entries and band structure.
-In this section, we empirically investigate the performance and scaling
-of three different algorithms for solving this extended system:
-
-\begin{enumerate}
-
-\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
-    for banded matrices using Gaussian elimination \citep[we implemented this
-    method following][]{Press:1992, Press:2007},
-
-\item \texttt{lapack}: The general banded LU decomposition implementation from
-    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
-    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines,\footnote{The experiments in this paper use the
-    Intel Math Kernel Library (\project{MKL}):
-    \url{https://software.intel.com/en-us/intel-mkl}.} and
-
-\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
-    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
-    sparsity but not the band structure.
-
-\end{enumerate}
-
-The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
-because the dimension of the extended matrix scales as $N\,J$ and the
-bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
-\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
-\project{SuperLU} package \citep{Demmel:1999}.
-We find that, while the \texttt{vanilla} solver scales as expected, the
-\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
-offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
-
-The benchmark experiments shown here were performed on a MacBook Pro with two
-2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
-CPUs and running Ubuntu.
-\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
-with $N$ and $J$ using the \texttt{vanilla} solver.
-As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
-in $J$ for large $J$.
-\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
-the \texttt{lapack} and \texttt{sparse} solvers respectively.
-For each of these optimized solvers, the empirical scaling is
-$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
-Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
-efficient than the other algorithms.
-The real world performance of \celerite\ depends on the specific platform,
-hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
-found qualitatively similar results across popular platforms and
-state-of-the-art libraries.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
-\caption{\dfmtodo{Cholesky} A benchmark showing the computational scaling of \celerite\ using the
-    \texttt{vanilla} band solver as a function of the number of data points
-    and the number of terms.
-    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
-    matrix given by \eq{celerite-kernel} as a function of the number of data
-    points $N$.
-    The different lines show the cost for different numbers of terms $J$
-    increasing from bottom to top.
-    To guide the eye, the straight black line without points
-    shows linear scaling in $N$.
-    \emph{(right)} The same information plotted as a function of $J$ for
-    different values of $N$.
-    Each line shows the scaling for a specific value of $N$ increasing from
-    bottom to top.
-    The black line shows quadratic scaling in $J$.
-    \figurelabel{benchmark}}
-\end{center}
-\end{figure}
-
 \section{Ensuring positive definiteness}

### 2017-06-07 paper/ms.tex
moving sections around
paper edits

--- 2017-06-05 paper/ms.tex
@@
 overflow and underflow issues in many realistic cases because both $U$ and $V$
-have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
-real number.
-This can be avoided with the appropriate reparameterization.
+have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large(small)
+real number;  these factors can cancel in the end, but will not if over(under)flow
+is an issue.  This can be avoided with the appropriate reparameterization.
 We define the following pre-conditioned variables
@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that $\mathcal{Q} = -2 \mathcal{L}$
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $-2 \mathcal{L}$
 with $\mu_\bvec{\theta} =0$ is given by
@@
 \end{equation}
+We have yet to derive an expression for the variance.
 

### 2017-06-07 paper/ms.tex

--- 2017-06-07 paper/ms.tex
@@
 
-\section{Semi-separable matrix factorization}
-
-The relationship between the \celeriteterm\ model and semi-separable linear
-algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
-A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
-written as
-\begin{eqnarray}
-K_{n,m} = \left\{\begin{array}{ll}
-    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
-    A_{n,n} & \mathrm{if}\,m=n \\
-    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
-\end{array}\right.
-\end{eqnarray}
-for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
-N$-dimensional diagonal matrix $A$.
-This definition can be equivalently written as
-\begin{eqnarray}\eqlabel{semi-sep}
-K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
-\end{eqnarray}
-where the $\mathrm{tril}$ function extracts the strict lower triangular
-component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.  Note that the last two terms in this equation are
-transposes of one another.
-For the \celeriteterm\ model described in the previous section, the rank of
-the semi-separable matrix is $R = 2\,J$ and the components of the relevant
-matrices are
-\begin{eqnarray}
-U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
-    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
-U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
-    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
-V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
-V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
-A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
-\end{eqnarray}
-
-Note that rather than storing the full $N \times N$ covariance matrix, it
-is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
-as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
-point numbers.
-
-\subsection{Cholesky factorization}
-
-To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
-compute the log-determinant of $K$ and solve linear systems of the form
-$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
-semi-separable matrix.
-If $K$ is positive semi-definite, these can be computed using the Cholesky
-factorization of $K$:
-\begin{eqnarray}
-K = L\,D\,L^\T
-\end{eqnarray}
-where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
-diagonal matrix.
-We use the ansatz that $L$ has the form
-\begin{eqnarray}
-    L = I + \mathrm{tril} (U\,W^T)
-\end{eqnarray}
-where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
-$N \times 2\,J$ matrix.
-Combining this with \eq{semi-sep}, we find the equation
-\begin{eqnarray}\eqlabel{cholesky}
-A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
-&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
-    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
-&=& D + \mathrm{tril}(U\,W^\T)\,D
-    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
-&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
-\end{eqnarray}
-that we can solve for $W$.
-Setting each element on the left side of \eq{cholesky} equal to the
-corresponding element on the right side, we derive the following recursion
-relations
-\begin{eqnarray}
-S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
-D_{n,n} &=& A_{n,n} -
-    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
-    \nonumber\\
-W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
-    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
-    \eqlabel{algorithm}
-\end{eqnarray}
-where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
-$S_{1,j,k}=0$.
-The computational cost of one iteration of the update in \eq{algorithm} is
-$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
-observation~--~so the total cost of this factorization scales as
-$\mathcal{O}(N\,J^2)$.
-
-Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
-then these terms only add one to the rank of $U$ and $V$.  We use this fact
-to simplify the algebra when real components are included:  if $J_{real}$
-and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
-the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
-
-\subsection{Pre-conditioning}
-
-Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
-overflow and underflow issues in many realistic cases because both $U$ and $V$
-have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
-real number.
-This can be avoided with the appropriate reparameterization.
-We define the following pre-conditioned variables
-\begin{eqnarray}
-\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
-\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
-\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
-\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
-\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
-\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
-\end{eqnarray}
-and the set of variables
-\begin{eqnarray}
-\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
-\end{eqnarray}
-with the constraint
-\begin{eqnarray}
-\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
-\end{eqnarray}
-Using this parameterization, we find the following, numerically stable,
-algorithm
-\begin{eqnarray}
-    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
-    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
-D_{n,n} &=& A_{n,n} -
-    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
-    \nonumber\\
-\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
-    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
-\end{eqnarray}
-
-\subsection{Solving}
-
-After computing the Cholesky factorization of the matrix $K$, we can apply
-the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
-$\mathcal{O}(N)$ operations respectively.
-The log-determinant of $K$ is given by
-\begin{eqnarray}
-    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
-\end{eqnarray}
-where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
-
-The inverse of $K$ can be applied using the relationship
-\begin{eqnarray}
-\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
-\end{eqnarray}
-First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
-substitution algorithm
-\begin{eqnarray}
-    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
-    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
-    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
-\end{eqnarray}
-where $f_{0,j} = 0$ for all $j$.
-Using this result and the following backward substitution algorithm, we can
-compute $\bvec{z}$
-\begin{eqnarray}
-g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
-    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
-z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
-\end{eqnarray}
-where $g_{N+1,j} = 0$ for all $j$.
-The total cost of the forward and backward substitution passes scales as
-$\mathcal{O}(N\,J)$.
-
 \section{celerite as a model of stellar variations}\sectlabel{sho}
@@
 \end{figure}
+
+\section{Semi-separable matrix factorization}
+
+The relationship between the \celeriteterm\ model and semi-separable linear
+algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
+A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
+written as
+\begin{eqnarray}
+K_{n,m} = \left\{\begin{array}{ll}
+    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
+    A_{n,n} & \mathrm{if}\,m=n \\
+    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
+\end{array}\right.
+\end{eqnarray}
+for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
+N$-dimensional diagonal matrix $A$.
+This definition can be equivalently written as
+\begin{eqnarray}\eqlabel{semi-sep}
+K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+\end{eqnarray}
+where the $\mathrm{tril}$ function extracts the strict lower triangular
+component of its argument and $\mathrm{triu}$ is the equivalent strict upper
+triangular operation.  Note that the last two terms in this equation are
+transposes of one another.
+For the \celeriteterm\ model described in the previous section, the rank of
+the semi-separable matrix is $R = 2\,J$ and the components of the relevant
+matrices are
+\begin{eqnarray}
+U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
+    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
+U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
+    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
+V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
+\end{eqnarray}
+
+Note that rather than storing the full $N \times N$ covariance matrix, it
+is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
+as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
+point numbers.
+
+\subsection{Cholesky factorization}
+
+To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
+compute the log-determinant of $K$ and solve linear systems of the form
+$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
+semi-separable matrix.
+If $K$ is positive semi-definite, these can be computed using the Cholesky
+factorization of $K$:
+\begin{eqnarray}
+K = L\,D\,L^\T
+\end{eqnarray}
+where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
+diagonal matrix.
+We use the ansatz that $L$ has the form
+\begin{eqnarray}
+    L = I + \mathrm{tril} (U\,W^T)
+\end{eqnarray}
+where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
+$N \times 2\,J$ matrix.
+Combining this with \eq{semi-sep}, we find the equation
+\begin{eqnarray}\eqlabel{cholesky}
+A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
+&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
+    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
+&=& D + \mathrm{tril}(U\,W^\T)\,D
+    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
+&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
+\end{eqnarray}
+that we can solve for $W$.
+Setting each element on the left side of \eq{cholesky} equal to the
+corresponding element on the right side, we derive the following recursion
+relations
+\begin{eqnarray}
+S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
+    \nonumber\\
+W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
+    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
+    \eqlabel{algorithm}
+\end{eqnarray}
+where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
+$S_{1,j,k}=0$.
+The computational cost of one iteration of the update in \eq{algorithm} is
+$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
+observation~--~so the total cost of this factorization scales as
+$\mathcal{O}(N\,J^2)$.
+
+Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
+then these terms only add one to the rank of $U$ and $V$.  We use this fact
+to simplify the algebra when real components are included:  if $J_{real}$
+and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
+the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
+
+\subsection{Pre-conditioning}
+
+Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
+overflow and underflow issues in many realistic cases because both $U$ and $V$
+have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large(small)
+real number;  these factors can cancel in the end, but will not if over(under)flow
+is an issue.  This can be avoided with the appropriate reparameterization.
+We define the following pre-conditioned variables
+\begin{eqnarray}
+\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
+\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
+\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
+\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
+\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
+\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
+\end{eqnarray}
+and the set of variables
+\begin{eqnarray}
+\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
+\end{eqnarray}
+with the constraint
+\begin{eqnarray}
+\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
+\end{eqnarray}
+Using this parameterization, we find the following, numerically stable,
+algorithm
+\begin{eqnarray}
+    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
+    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
+D_{n,n} &=& A_{n,n} -
+    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
+    \nonumber\\
+\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
+    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
+\end{eqnarray}
+
+\subsection{Solving}
+
+After computing the Cholesky factorization of the matrix $K$, we can apply
+the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
+$\mathcal{O}(N)$ operations respectively.
+The log-determinant of $K$ is given by
+\begin{eqnarray}
+    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
+\end{eqnarray}
+where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
+
+The inverse of $K$ can be applied using the relationship
+\begin{eqnarray}
+\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
+substitution algorithm
+\begin{eqnarray}
+    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
+    \tilde{W}_{n-1,j}\,z_{n-1}^\prime\right] \nonumber\\
+    z_n^\prime &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{0,j} = 0$ for all $j$.
+Using this result and the following backward substitution algorithm, we can
+compute $\bvec{z}$
+\begin{eqnarray}
+g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
+    \tilde{U}_{n+1,j}\,z_{n+1}\right] \nonumber\\
+z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
+\end{eqnarray}
+where $g_{N+1,j} = 0$ for all $j$.
+The total cost of the forward and backward substitution passes scales as
+$\mathcal{O}(N\,J)$.
 

### 2017-06-07 paper/ms.tex
Corrected equation (B35); V -> V*; U -> U*

@@
 \begin{eqnarray}
-f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
+f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
 y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
@@
 for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
-\agoltodo{Check that these equations are in fact correct.}
 We expect that this simulation algorithm will be useful for test-driving Gaussian
@@
 Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
-X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
+X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i}^*,\\
 Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
-X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
+X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i}^*,
 \end{eqnarray}
 where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
-for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_m^*$.
+for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
 Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in

### 2017-06-08 paper/ms.tex
words about semiseparable matrices

--- 2017-06-07 paper/ms.tex
@@
 \response{Unlike earlier linear methods using Kalman filters \citep[for
-example,][]{Kelly:2014}, this novel algorithm exploits the semi-separable
+example,][]{Kelly:2014}, this novel algorithm exploits the semiseparable
 structure of a specific class of covariance matrices to directly factorize and
@@
 matrix when it is generated by a mixture of exponentials.
-\response{The semi-separable nature of these matrices was first recognized by
+\response{The semiseparable nature of these matrices was first recognized by
 \citet{Ambikasaran:2015}, building on intuition from a twenty year old paper
@@
 
-\section{The celerite model}
+\section{The celerite model}\sectlabel{celerite}
 
@@
 \response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
-kernel is semi-separable.}
+kernel is semiseparable.}
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
@@
 components in the mixture.
-\response{In the following section, we derive a novel method with the same
+\response{In \sect{factor}, we derive a novel method with the same
 $\mathcal{O}(N\,J^2)$ scaling that can be used when
@@
 physical motivation for the choice of kernel functions.
-Kernels are often chosen simply because they are popular with little
+Kernels are often chosen simply because they are popular, with little
 consideration of the impact of this decision.
-While this isn't necessarily a problem for the inferred results, in this
-section we discuss an exact physical interpretation of the \celeriteterm\
-kernel that will be applicable to many astrophysical systems but especially
-the time-variability of stars.
+While this doesn't necessarily impact the inferred results, in this section we
+discuss an exact physical interpretation of the \celeriteterm\ kernel that
+will be applicable to many astrophysical systems but especially the
+time-variability of stars.
 
@@
 describe a wide range of stellar variations and we return to give some
-specific examples in \sect{examples-real}.
+specific examples in \sect{examples}.
 Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
@@
 measurements in $\mathcal{O}(N)$ operations using the \celeriteterm\ method
-described in the previous section.
+described in the next section.
 
@@
 
-\section{Semi-separable matrix factorization}
-
-The relationship between the \celeriteterm\ model and semi-separable linear
-algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015}.
-A rank-$R$ semi-separable matrix is a matrix $K$ where the elements can be
-written as
+\section{Semiseparable matrices \& celerite}\sectlabel{factor}
+
+The previous two sections describe the \celeriteterm\ model and its physical
+interpretation.
+In this section, we derive a new scalable direct solver for covariance
+matrices generated by this model to compute the GP likelihood
+(\eqalt{gp-likelihood}) for large datasets.
+This method exploits the semiseparable structure of these matrices to compute
+their Cholesky factorization in $\mathcal{O}(N\,J^2)$ operations.
+
+The relationship between the \celeriteterm\ model and semiseparable linear
+algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
+on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
+\citet{Ambikasaran:2015} derived a scalable direct solver for any general
+semiseparable matrix and applied this method to \celeriteterm\ models with
+$b_j = 0$ and $d_j = 0$.
+In \app{extended}, we generalize this earlier method to include non-zero $b_j$
+and $d_j$.
+It turns out, however, that we can derive a higher performance algorithm by
+only considering positive semi-definite (PSD) matrices.
+In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
+PSD.
+
+To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
+elements can be written as
 \begin{eqnarray}
@@
 component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.  Note that the last two terms in this equation are
-transposes of one another.
-For the \celeriteterm\ model described in the previous section, the rank of
-the semi-separable matrix is $R = 2\,J$ and the components of the relevant
-matrices are
-\begin{eqnarray}
-U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
-    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
-U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
-    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
-V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
-V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
-A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
-\end{eqnarray}
-
-Note that rather than storing the full $N \times N$ covariance matrix, it
-is only necessary to store the $2J \times N$ matrices $U$ and $V$, as well
-as the diagonal terms, for a total storage volume of $(4J+1) \times N$ floating
-point numbers.
+triangular operation.
+In this representation, rather than storing the full $N \times N$ matrix $K$,
+it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
+$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
+floating point numbers.
+Given a semiseparable matrix of this form, it is known that matrix-vector
+products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
+operations and the interested reader is directed to \citet{Vandebril:2007} for
+more details about semiseparable matrices in general.
 
@@
 
-To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
-compute the log-determinant of $K$ and solve linear systems of the form
-$K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is a rank-$2\,J$
-semi-separable matrix.
-If $K$ is positive semi-definite, these can be computed using the Cholesky
-factorization of $K$:
+For our purposes, we want to compute the log-determinant of $K$ and solve
+linear systems of the form $K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is
+a rank-$R$ semiseparable PSD matrix.
+These can both be computed using the Cholesky factorization of $K$:
 \begin{eqnarray}
@@
 where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
-$N \times 2\,J$ matrix.
+$N \times R$ matrix.
 Combining this with \eq{semi-sep}, we find the equation
@@
 D_{n,n} &=& A_{n,n} -
-    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
+    \sum_{j=1}^{R}\sum_{k=1}^{R} U_{n,j}\,S_{n,j,k}\,U_{n,k}
     \nonumber\\
 W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
-    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
+    \sum_{k=1}^{R}U_{n,k}\,S_{n,j,k} \right]
     \eqlabel{algorithm}
 \end{eqnarray}
-where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
+where $S_n$ is a symmetric $R \times R$ matrix and every element
 $S_{1,j,k}=0$.
 The computational cost of one iteration of the update in \eq{algorithm} is
-$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
-observation~--~so the total cost of this factorization scales as
-$\mathcal{O}(N\,J^2)$.
-
-Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
-then these terms only add one to the rank of $U$ and $V$.  We use this fact
-to simplify the algebra when real components are included:  if $J_{real}$
-and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
-the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.
-
-\subsection{Pre-conditioning}
-
-Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
-overflow and underflow issues in many realistic cases because both $U$ and $V$
-have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large(small)
-real number;  these factors can cancel in the end, but will not if over(under)flow
-is an issue.  This can be avoided with the appropriate reparameterization.
+$\mathcal{O}(R^2)$ and this must be run $N$ times~--~once for each row~--~so
+the total cost of this factorization scales as $\mathcal{O}(N\,R^2)$.
+
+\subsection{Scalable factorization of the celerite model}
+
+The \celeriteterm\ model described in \sect{celerite} can be exactly
+represented as a rank $R = 2\,J$ semiseparable matrix\footnote{We note that if
+some \celeriteterm\ terms are real ($b_j=0$ and $d_j=0$) then these terms only
+add one to the rank.
+If a \celeriteterm\ model has $J_\mathrm{real}$ real and $J_\mathrm{complex}$
+general terms, the semiseparable rank is only $J_\mathrm{real} +
+2\,J_\mathrm{complex}$.}
+where the components of the relevant matrices are
+\begin{eqnarray}
+U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
+    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
+U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
+    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
+V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
+\end{eqnarray}
+Na\"ive implementation of the algorithm in \eq{algorithm} to this system will
+result in numerical overflow and underflow issues in many realistic cases
+because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
+can be any arbitrarily large real number.
+This issue was also noted by \citet{Ambikasaran:2015} and it can be avoided
+with a reparameterization similar to the one they use.
 We define the following pre-conditioned variables
@@
 Using this parameterization, we find the following, numerically stable,
-algorithm
+algorithm to compute the Cholesky factorization of a \celeriteterm\ model
 \begin{eqnarray}
@@
 \end{eqnarray}
+As before, a single iteration of this algorithm requires $\mathcal{O}(J^2)$
+operations and it is repeated $N$ times for a total scaling of
+$\mathcal{O}(N\,J^2)$.
+Furthermore, $\tilde{V}_{n,j}$ and $A_{n,n}$ can be updated in place and used
+to represent $\tilde{W}_{n,j}$ and $D_{n,n}$, giving a total memory footprint
+of $(6\,J + 1)\,N + J\,(J-1)/2$ floating point numbers.
 
@@
 
-\section{An alternative solver using the extended matrix formalism}
+\section{An alternative solver using the extended matrix
+formalism}\sectlabel{extended}
 
@@
 
-\section{Semi-separable operations}
-
-To help clarify these equations, we write out here an example of a semi-separable matrix with a
+\section{Semiseparable operations}
+
+To help clarify these equations, we write out here an example of a semiseparable matrix with a
 single \celeriteterm\ term with $c=d=1$: %and $N=3$
@@
 Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
-operations with a stable recursive algorithm using the semi-separable form:
+operations with a stable recursive algorithm using the semiseparable form:
 \begin{equation}
@@
 
-\section{Ensuring positive definiteness}
+\section{Ensuring positive definiteness}\sectlabel{psd}
 

### 2017-06-08 paper/ms.tex
words and edits about factorization and solving

@@
 
-\subsection{Cholesky factorization}
+\subsection{Cholesky factorization of PSD semiseparable matrices}
 
@@
 
-\subsection{Scalable factorization of the celerite model}
+After computing the Cholesky factorization of this matrix $K$, we can also
+apply its inverse and compute its log-determinant in $\mathcal{O}(N\,R)$ and
+$\mathcal{O}(N)$ operations respectively.
+The log-determinant of $K$ is given by
+\begin{eqnarray}
+    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
+\end{eqnarray}
+where, since $K$ is PSD, $D_{n,n}>0$ for all $n$.
+
+The inverse of $K$ can be applied using the relationship
+\begin{eqnarray}
+\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
+\end{eqnarray}
+First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using forward
+substitution
+\begin{eqnarray}\eqlabel{fwd}
+    f_{n,j} &=& f_{n-1,j} + W_{n-1,j}\,z_{n-1}^\prime \nonumber\\
+    z_n^\prime &=& y_n - \sum_{j=1}^{R} U_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{0,j} = 0$ for all $j$.
+Then, using this result and backward substitution, we compute $\bvec{z}$
+\begin{eqnarray}\eqlabel{bwd}
+g_{n,j} &=& g_{n+1,j} + U_{n+1,j}\,z_{n+1} \nonumber\\
+z_n &=& \frac{z_n^\prime}{D_{n,n}} - \sum_{j=1}^{R} W_{n,j}\,g_{n,j}
+\end{eqnarray}
+where $g_{N+1,j} = 0$ for all $j$.
+The total cost of the forward and backward substitution passes scales as
+$\mathcal{O}(N\,R)$.
+
+The algorithm derived here is a general method for factorizing, solving, and
+computing the determinant of PSD rank-$R$ semiseparable matrices.
+As we will show below, with some care, this method can be used to compute
+\eq{gp-likelihood} for \celeriteterm\ models.
+In practice, when $K$ is PSD and this method is applicable, we find that this
+method is about a factor of 20 faster than an optimized implementation of the
+earlier, more general method from \citet{Ambikasaran:2015}.
+
+\subsection{Scalable computation of celerite models}
 
@@
 add one to the rank.
-If a \celeriteterm\ model has $J_\mathrm{real}$ real and $J_\mathrm{complex}$
-general terms, the semiseparable rank is only $J_\mathrm{real} +
-2\,J_\mathrm{complex}$.}
+Therefore, if a \celeriteterm\ model has $J_\mathrm{real}$ real and
+$J_\mathrm{complex}$ general terms, the semiseparable rank is only
+$J_\mathrm{real} + 2\,J_\mathrm{complex}$.}
 where the components of the relevant matrices are
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{system}
 U_{n,2\,j-1} &=& a_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) +
-    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \\
+    b_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) \nonumber\\
 U_{n,2\,j} &=& a_j\,e^{-c_j\,t_n}\,\sin(d_j\,t_n) -
-    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \\
-V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \\
-V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \\
+    b_j\,e^{-c_j\,t_n}\,\cos(d_j\,t_n) \nonumber\\
+V_{m,2\,j-1} &=& e^{c_j\,t_m}\,\cos(d_j\,t_m) \nonumber\\
+V_{m,2\,j} &=& e^{c_j\,t_m}\,\sin(d_j\,t_m) \nonumber\\
 A_{n,n} &=& \sigma_n^2 + \sum_{j=1}^J a_j \quad.
 \end{eqnarray}
-Na\"ive implementation of the algorithm in \eq{algorithm} to this system will
-result in numerical overflow and underflow issues in many realistic cases
+A na\"ive implementation of the algorithm in \eq{algorithm} for this system
+will result in numerical overflow and underflow issues in many realistic cases
 because both $U$ and $V$ have factors of the form $e^{\pm c_j t}$, where $t$
 can be any arbitrarily large real number.
-This issue was also noted by \citet{Ambikasaran:2015} and it can be avoided
-with a reparameterization similar to the one they use.
+This issue can be avoided by reparameterizing the semiseparable representation
+from \eq{system}.
+%A similar issue was also noted by \citet{Ambikasaran:2015} and it can be
+%avoided using a reparameterization.
 We define the following pre-conditioned variables
-\begin{eqnarray}
-\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
-\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
-\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
-\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
-\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
+\begin{eqnarray}\eqlabel{reparam}
+\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \nonumber\\
+\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \nonumber\\
+\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \nonumber\\
+\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \nonumber\\
+\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \nonumber\\
 \tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
 \end{eqnarray}
-and the set of variables
+and the new set of variables
 \begin{eqnarray}
@@
 
-\subsection{Solving}
-
-After computing the Cholesky factorization of the matrix $K$, we can apply
-the inverse of $K$ and compute its log-determinant in $\mathcal{O}(N\,J)$ and
-$\mathcal{O}(N)$ operations respectively.
-The log-determinant of $K$ is given by
-\begin{eqnarray}
-    \ln \det K = \sum_{n=1}^N \ln D_{n,n}
-\end{eqnarray}
-where, if $K$ is positive semi-definite, $D_{n,n}>0$ for all $n$.
-
-The inverse of $K$ can be applied using the relationship
-\begin{eqnarray}
-\bvec{z} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
-\end{eqnarray}
-First, we solve for $\bvec{z}^\prime = L^{-1}\,\bvec{y}$ using the following forward
-substitution algorithm
+The log-determinant of $K$ can still be computed as in the previous section,
+but we must update the forward and backward substitution algorithms to account
+for the reparameterization in \eq{reparam}.
+The forward substitution with the reparameterized variables is
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $f_{0,j} = 0$ for all $j$.
-Using this result and the following backward substitution algorithm, we can
-compute $\bvec{z}$
+where $f_{0,j} = 0$ for all $j$, and the backward substitution algorithm
+becomes
 \begin{eqnarray}
@@
 where $g_{N+1,j} = 0$ for all $j$.
-The total cost of the forward and backward substitution passes scales as
-$\mathcal{O}(N\,J)$.
+The only differences, when compared to Equations~(\eqref{fwd})
+and~(\eqref{bwd}) is the multiplication by $\phi_{n,j}$ in the recursion
+relations for $f_{n,j}$ and $g_{n,j}$.
+As before, the computational cost of this solve scales as $\mathcal{O}(N\,J)$.
 

### 2017-06-08 paper/ms.tex
starting on performance section

@@
 
+\subsection{Performance}
+
+We have derived a novel method for computing the Cholesky factorization of the
+covariance matrices produced by \celeriteterm\ models.
+In this section, we demonstrate the scalability and numerical performance of
+this algorithm.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
+\caption{A benchmark showing the computational scaling of \celerite\ using the
+    Cholesky factorization routine for rank-$2\,J$ semiseparable matrices as
+    described in the text.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
+    For comparison, the computational cost for the same model using a general
+    Cholesky factorization routine implemented in the Intel MKL linear algebra
+    package.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
+    \figurelabel{benchmark}}
+\end{center}
+\end{figure}
+
 \section{Examples}\sectlabel{examples}
@@
 state-of-the-art libraries.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=\textwidth]{figures/benchmark_darwin.pdf}
-\caption{\dfmtodo{Cholesky} A benchmark showing the computational scaling of \celerite\ using the
-    \texttt{vanilla} band solver as a function of the number of data points
-    and the number of terms.
-    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
-    matrix given by \eq{celerite-kernel} as a function of the number of data
-    points $N$.
-    The different lines show the cost for different numbers of terms $J$
-    increasing from bottom to top.
-    To guide the eye, the straight black line without points
-    shows linear scaling in $N$.
-    \emph{(right)} The same information plotted as a function of $J$ for
-    different values of $N$.
-    Each line shows the scaling for a specific value of $N$ increasing from
-    bottom to top.
-    The black line shows quadratic scaling in $J$.
-    \figurelabel{benchmark}}
-\end{center}
-\end{figure}
 

### 2017-06-08 paper/figures/error.py

@@ -1,0 +1,76 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+from __future__ import division, print_function
+
+import h5py
+import numpy as np
+import matplotlib.pyplot as plt
+
+import celerite
+from celerite import terms, plot_setup
+
+np.random.seed(42)
+plot_setup.setup(auto=True)
+
+K = 10
+J = np.arange(2, 64, 8)
+N = 2**np.arange(6, 16)
+
+alpha_error = np.empty((K, len(J), len(N)))
+logdet_error = np.empty((K, len(J), len(N)))
+logdet_error[:, :, :] = np.nan
+
+for k in range(K):
+    t = np.sort(np.random.uniform(0, N.max() * 0.8, N.max()))
+    yerr = np.random.uniform(1.0, 1.5, len(t))
+
+    for ix, j in enumerate(J):
+        kernel = terms.RealTerm(np.random.uniform(-1, 1),
+                                np.random.uniform(-5, -1))
+        kernel += terms.RealTerm(np.random.uniform(-1, 1),
+                                 np.random.uniform(-5, -1))
+        while (len(kernel.coefficients[0]) + 2*len(kernel.coefficients[2]) <
+               2*j):
+            kernel += terms.SHOTerm(
+                log_S0=np.random.uniform(-1, 1),
+                log_omega0=np.random.uniform(-5, 0),
+                log_Q=np.random.uniform(0, 1),
+            )
+        kernel += terms.JitterTerm(np.random.uniform(-1, 1))
+        assert (
+            len(kernel.coefficients[0]) + 2*len(kernel.coefficients[2]) == 2*j
+        )
+
+        gp = celerite.GP(kernel)
+
+        for iy, n in enumerate(N):
+            gp.compute(t[:n], yerr[:n])
+            alpha_true = np.random.randn(n)
+            args = [kernel.jitter] + list(kernel.coefficients)
+            args += [t[:n], alpha_true]
+            y = gp.solver.dot(*args)[:, 0] + yerr[:n]**2 * alpha_true
+
+            alpha = gp.apply_inverse(y[:n])[:, 0]
+            logdet = gp.solver.log_determinant()
+            alpha_error[k, ix, iy] = np.max(np.abs(alpha-alpha_true))
+
+            if n <= 2048:
+                logdet0 = np.linalg.slogdet(gp.get_matrix())[1]
[…]

### 2017-06-08 paper/ms.tex
adding error plots to ms

@@
 this algorithm.
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/error/error.pdf}
+\caption{\emph{(left)} The maximum numerical error introduced by the algorithm
+    described in the text as a function of the number of \celeriteterm\ terms
+    $J$ and data points $N$.
+    Each line corresponds to a different $J$ and the error increases with $J$
+    as shown in the legend.
+    Each point in this figure is the average result across 10 systems with
+    randomly sampled parameters.
+\emph{(right)} The distribution of fractional error on the log determinant of
+    $K$ computed using the algorithm described in the text compared to the
+    algorithm implemented in \project{NumPy} \citep{Van-Der-Walt:2011} for the
+    same systems shown in the left-hand figure.
+    This histogram only includes comparisons for $N \le 2048$ because standard
+    methods become computationally expensive for larger systems.
+    \figurelabel{error}}
+\end{center}
+\end{figure}
 

### 2017-06-08 paper/ms.tex
words about performance

@@
 
-We have derived a novel method for computing the Cholesky factorization of the
-covariance matrices produced by \celeriteterm\ models.
-In this section, we demonstrate the scalability and numerical performance of
+We have derived a novel and scalable method for computing the Cholesky
+factorization of the covariance matrices produced by \celeriteterm\ models.
+In this section, we demonstrate the numerical and computational performance of
 this algorithm.
+
+First, we empirically confirm that the method solves systems of the form
+$K\,\bvec{z} = \bvec{y}$ to high numerical accuracy.
+The left panel of \Figure{error} shows the L-infinity norm for the residual
+$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
+terms $J$.
+For each pair of $N$ and $J$ we repeated the experiment 10 times with
+different randomly generated datasets and \celeriteterm\ parameters and
+averaged across experiments to remove some sampling noise from this figure.
+The numerical error introduced by our method increases weakly for larger
+datasets $\sim N^{0.15}$ and roughly linearly with $J$.
+
+For the experiments with small $N \le 2048$, we also compared the log
+determinant calculation to the result calculated using the general log
+determinant function provided by the \project{NumPy} project
+\citep{Van-Der-Walt:2011}, itself built on the \project{LAPACK}
+\citep{Anderson:1999} implementation in the Intel Math Kernel
+Library\footnote{\url{https://software.intel.com/en-us/intel-mkl}}.
+The right-hand panel of \Figure{error} shows a histogram of the fractional
+error in the log determinant calculation for each of the 480~experiments with
+$N \le 2048$.
+Across the board, these results agree within machine precision and the error
+only weakly increases with $N$ and $J$.
+
+We continue by measuring the real-world computational efficiency of this
+method as a function of $N$ and $J$.
+This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
+i5 processor, but find similar performance on other platforms.
+The \celerite\ implementation used for this test is the Python interface with
+a C++ back end, but there is little overhead introduced by Python so we find
+similar performance using implementations in C++ and Julia
+\citep{Bezanzon:2012}.
+\Figure{benchmark} shows the computational cost of evaluating
+\eq{gp-likelihood} using \celerite\ as a function of $N$ and $J$.
+The empirical scaling as a function of data size matches the theoretical
+scaling of $\mathcal{O}(N)$ for all values of $N$ and the scaling with the
+number of terms is the theoretical $\mathcal{O}(J^2)$ for $J \gtrsim 10$, with
+some overhead for smaller models.
+For comparison, the computational cost of the same computation using the Intel
+Math Kernel Library, a highly optimized general dense linear algebra package,
+is shown as a dashed line.
+\celerite\ is faster than the reference implementation for all but the
+smallest datasets with many terms.
 
@@
 
-\begin{figure}[p]
+\begin{figure}[htbp]
 \begin{center}
@@
      \project{george} \citep{Ambikasaran:2016},
+     \project{Julia} \citep{Bezanzon:2012},
      \project{LAPACK} \citep{Anderson:1999},

### 2017-06-08 paper/ms.tex
cleaning up appendix and removing extended

@@
 $b_j = 0$ and $d_j = 0$.
-In \app{extended}, we generalize this earlier method to include non-zero $b_j$
-and $d_j$.
-It turns out, however, that we can derive a higher performance algorithm by
-only considering positive semi-definite (PSD) matrices.
+During the preparation of this paper, we generalized this earlier method to
+include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
+performance algorithm by restricting our method to positive semi-definite
+(PSD) matrices.
 In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
@@
 
-\section{An alternative solver using the extended matrix
-formalism}\sectlabel{extended}
-
-\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
-elements given by \eq{kernel-simple}, could be computed
-efficiently by taking advantage of the structure of this covariance function.
-\citet{Ambikasaran:2015} generalized this computation to apply to the full
-mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
-efficient method for computing the determinant of $K$.
-
-\subsection{An example}
-
-To provide some insight for this method, we follow
-\citet{Ambikasaran:2015} and start by working through a simple example.
-In this case, we assume that we have three data points
-$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
-variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
-like to compute the likelihood of these data under a GP model with the
-covariance function
-\begin{eqnarray}
-k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
-\end{eqnarray}
-This is the result of setting $J=1$, $b=0$, and $d=0$ in
-\eq{celerite-kernel-complex} and it is the model studied by
-\citet{Rybicki:1995}.
-To demonstrate that the likelihood of this model can be computed in
-$\mathcal{O}(N)$, we write out the full system of
-equations that must be solved to apply the inverse of $K$ and compute
-the first term of \eq{gp-likelihood}.
-In matrix notation, this is
-\begin{eqnarray}\eqlabel{impl-matrix}
-K\,\bvec{z} &=& \bvec{y}, \\
-\begin{pmatrix}
-    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
-    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
-    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ z_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ y_2 \\ y_3
-\end{pmatrix}
-\end{eqnarray}
-where our goal is to solve for the unknown vector \bvec{z} for a given matrix
-$K$ and vector \bvec{y}.
-In \eq{impl-matrix}, we have assumed that the mean function is zero but a
-non-zero mean could be included by replacing \bvec{y} by
-$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
-Now, if we introduce the variables
-\begin{eqnarray}
-    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
-\end{eqnarray}
-where $g_{0} = 0$, and
-\begin{eqnarray}\eqlabel{algo-first}
-    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
-\end{eqnarray}
-where $u_{N} = 0$,
-the system of equations can be rewritten as
-\begin{eqnarray}
-(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
-a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
-a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
-\end{eqnarray}
-Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
-as a matrix equation shows the benefit that this seemingly trivial
-reformulation provides:
-\begin{eqnarray}
-\begin{pmatrix}
-    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
-    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
-    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
-    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
-    0 & 0 & 0 & 0 & -1 & 0 & a \\
-    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
-\end{pmatrix}\,
-\begin{pmatrix}
-    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
-\end{pmatrix} &=&
-\begin{pmatrix}
-    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
-\end{pmatrix}\nonumber
-\end{eqnarray}
-Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
-rewrite \eq{impl-matrix} as
-\begin{eqnarray}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
-\end{eqnarray}
-Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
-with, it is now sparse with banded structure that can be exploited to solve
-the system efficiently.
-In particular, sparse solvers are available that can perform a
-LU-decomposition of matrices like this in $\mathcal{O}(N)$
-operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
-general~--~and we can use these algorithms to solve our system
-exactly because the target vector $\bvec{z}$ is a subset of the elements of
-$\bvec{z}_\mathrm{ext}$.
-
-In the following section we discuss this method more generally, but it's
-worth noting a few important facts that can already be seen in this example.
-First, the fundamental reason why this matrix $K$ can be solved efficiently is
-the following property of exponentials
-\begin{eqnarray}
-    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
-    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
-    e^{-c\,(t_3 - t_1)}
-\end{eqnarray}
-and it is important to note that this property does not extend to other common
-covariance functions like the ``exponential-squared'' function
-\begin{eqnarray}
-    k(\tau) \propto e^{-c\,\tau^2} \quad.
-\end{eqnarray}
-Second, our derivation of the extended matrix requires that the data
-points be monotonically sorted.
-This is the motivation for our previous statement that this method can only be
-applied in one dimension since neither of these properties will be satisfied
-by multidimensional inputs.
-
-\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
-extended matrix formalism in practice.
-First, even if the covariance function is a mixture of exponentials, the
-extended matrix will still be banded with a bandwidth that scales linearly
-with the number of components, $J$.
-Second, \citet{Ambikasaran:2015} proved that the absolute value of the
-determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
-determinant of $K$.
-This means we can use this extended matrix formalism to compute the
-marginalized likelihood in $\mathcal{O}(N)$ operations.
-
-\subsection{The algorithm}
-
-In this section, we generalize the method from the previous section to the
-covariance function given by \eq{celerite-kernel}.
-This derivation follows \citet{Ambikasaran:2015} but it includes explicit
-treatment of complex parameters, and their complex conjugates.
-
-In the case of the full \celeriteterm\ covariance function
-(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
-analogy to the $u_n$ and $g_n$ that we introduced in the previous section
-\begin{eqnarray}\eqlabel{full-system-first}
-\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
-\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
-g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
-h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
-u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
-v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
-\end{eqnarray}
-with the boundary conditions
-\begin{eqnarray}
-    g_{0,j} = 0 \quad, \quad
-    h_{0,j} = 0 \quad, \quad
-    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
-    v_{N,j} = 0
-\end{eqnarray}
-for all $j$.
-Using these variables and some algebra, we find that the following expression
-\begin{eqnarray}\eqlabel{full-system-last}
-\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
-+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
-+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
-    &=& r_{\bvec{\theta},n}
-\end{eqnarray}
-is equivalent to the target matrix equation
-\begin{eqnarray}
-K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
-\end{eqnarray}
-if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
-$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
-\eq{full-system-first} through \eq{full-system-last} define a banded matrix
-equation in the ``extended'' space
-\begin{eqnarray}\eqlabel{kext-full}
-    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
-    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
-\end{eqnarray}
-and, as before, this can be used to solve
-for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
-operations.
-\Figure{matrix} shows a pictorial representation of the sparsity pattern of
-the extended matrix $K_\mathrm{ext}$.
-The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
-with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
-non-zero entries.
-Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
-$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
-as
-\begin{eqnarray}
-{\bvec{z}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
-    h_{N-1,j} & z_N
-\end{array}\right)
-\end{eqnarray}
-and
-\begin{eqnarray}
-{\bvec{r}_\mathrm{ext}} ^\T =
-\left(\begin{array}{cccccccccc}
-    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
-    0 & \cdots & 0 & r_{\bvec{\theta},N}
-\end{array}\right) \quad.
-\end{eqnarray}
-After constructing the extended matrix (using a compact storage format), the
-extended matrix can be factorized using a LU-decomposition\footnote{Even
-though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
-Cholesky solver cannot be used for increased efficiency.} routine optimized
-for band or sparse matrices.
-This decomposition can then be used to compute the determinant of $K$, solve
-$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
-likelihood in \eq{gp-likelihood}.
-
-In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
-case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
-will also be identically zero for all $n$.
-Removing these trivial rows from the extended matrix results in factor of two
-increase in the computational efficiency for these terms.
-In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
-$d_j=0$, and the general terms are called ``complex''.
-
-\begin{figure}[tp]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
-\caption{A pictorial representation of the sparse extended matrix
-$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
-Each colored block corresponds to a non-zero entry in the matrix as described
-in the legend.
-    \figurelabel{matrix}}
-\end{center}
-\end{figure}
-
-\subsection{Implementation considerations \& scaling}
-
-The extended system defined in the previous section is sparse, with typically
-fewer than a few percent non-zero entries and band structure.
-In this section, we empirically investigate the performance and scaling
-of three different algorithms for solving this extended system:
-
-\begin{enumerate}
-
-\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
-    for banded matrices using Gaussian elimination \citep[we implemented this
-    method following][]{Press:1992, Press:2007},
-
-\item \texttt{lapack}: The general banded LU decomposition implementation from
-    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
-    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
-    \project{BLAS} routines,\footnote{The experiments in this paper use the
-    Intel Math Kernel Library (\project{MKL}):
-    \url{https://software.intel.com/en-us/intel-mkl}.} and
-
-\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
-    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
-    sparsity but not the band structure.
-
-\end{enumerate}
-
-The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
-because the dimension of the extended matrix scales as $N\,J$ and the
-bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
-\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
-\project{SuperLU} package \citep{Demmel:1999}.
-We find that, while the \texttt{vanilla} solver scales as expected, the
-\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
-offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.
-
-The benchmark experiments shown here were performed on a MacBook Pro with two
-2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
-CPUs and running Ubuntu.
-\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
-with $N$ and $J$ using the \texttt{vanilla} solver.
-As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
-in $J$ for large $J$.
-\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
-the \texttt{lapack} and \texttt{sparse} solvers respectively.
-For each of these optimized solvers, the empirical scaling is
-$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
-Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
-efficient than the other algorithms.
-The real world performance of \celerite\ depends on the specific platform,
-hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
-found qualitatively similar results across popular platforms and
-state-of-the-art libraries.
+\section{Ensuring positive definiteness for celerite models}\sectlabel{psd}
+
+For a GP kernel to be valid, it must produce a positive definite covariance
+matrix for all input coordinates.
+For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
+Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
+Fourier transform of a positive finite measure.
+This means that the power spectrum of a positive definite kernel must be
+positive for all frequencies.
+This result is intuitive because, since the power spectrum of a process is
+defined as the expected squared amplitude of the Fourier transform of the time
+series, it must be non-negative.
+
+Using \eq{celerite-psd}, we find that for a single \celeriteterm\ term, this
+requirement is met when
+\begin{eqnarray}\eqlabel{psd-req}
+\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
+{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2} > 0 \quad.
+\end{eqnarray}
+The denominator is positive for all $c_j \ne 0$ and it can be shown that, when
+$c_j=0$, \eq{psd-req} is satisfied for all $\omega \ne d_j$ where the power is
+identically zero.
+Therefore, when $c_j \ne 0$, we require that the numerator is positive for all
+$\omega$.
+This requirement can also be written as
+\begin{eqnarray}
+a_j\,c_j > -b_j\,d_j \\
+a_j\,c_j > b_j\,d_j \quad.
+\end{eqnarray}
+Furthermore, we can see that $a_j$ must be positive since $k(0) = a_j$ should
+should be positive and, similarly, by requiring the covariance to be finite at
+infinite lag, we obtain the constraint $c_j \ge 0$.
+Combining these results, we find the constraint
+\begin{eqnarray}
+    |b_j\,d_j| < a_j\,c_j \quad.
+\end{eqnarray}
+
+\response{The constraint for $J > 1$ is more complicated so, in most cases, we
+require that each term is positive definite using this relationship and use
+the fact that a product of sum of positive definite kernels will also be
+positive definite \citep{Rasmussen:2006} to show that the full model is
+positive.}
+However, in the case of $J$ general \celeriteterm\ terms, we can check for
+negative values of the PSD by solving for the roots of the power spectrum;
+if there are any real, positive roots, then the power-spectrum goes negative
+(or zero), and thus does not represent a valid kernel.
+We rewrite the power spectrum, \eqalt{celerite-psd}), abbreviating with
+$z = \omega^2$:
+\begin{equation}
+S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
+\end{equation}
+where
+\begin{eqnarray}
+q_j &=& a_j\,c_j-b_j\,d_j\\
+r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
+s_j &=& 2(c_j^2-d_j^2)\\
+t_j &=& (c_j^2+d_j^2)^2.
+\end{eqnarray}
+The denominators of each term are positive, so we can multiply through by
+$\prod_j \left(z^2+s_jz + t_j\right)$ to find
+\begin{equation}
+Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
+    t_k\right) = 0\quad,
+\end{equation}
+which is a polynomial with order $2\,(J-1)+1$.
+With $J=2$, this yields a cubic equation whose roots can be obtained exactly.
+
+For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
+allows one to determine whether there are any real roots within the range
+$(0,\,\infty]$.
+We first construct $Q_0(z)$ and its derivative $Q_1(z) = {Q_0}^\prime(z)$,
+and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
+\begin{eqnarray}
+Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
+\end{eqnarray}
+where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
+dividing $p(z)$ by $q(z)$.
+
+We evaluate the coefficients of each of the polynomial in the series by
+evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
+of these polynomials evaluated at $z=0$.
+Likewise, we evaluate the coefficients of the largest order term in each
+polynomial that gives the sign of the polynomial as $z \rightarrow \infty$,
+$f_\infty$.
+With the sequence of coefficients $f_0$ and $f_\infty$, we then determine how
+many times the sign changes in each of these, where $\sigma(0)$ is the number
+of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
+at $z \rightarrow \infty$.
+The total number of real roots in the range
+$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.
+
+We have checked that this procedure works for a wide range of parameters, and
+we find that it robustly matches the number of positive real roots which we
+evaluated numerically.
+The advantage of this procedure is that it does not require computing the
+roots, but only carrying out algebraic manipulation of polynomials to
+determine the number of positive real roots.
+If a non-zero real root is found, the likelihood may be set to zero.
 
@@
 
-\section{Ensuring positive definiteness}\sectlabel{psd}
-
-For a GP kernel to be valid, it must produce a positive definite covariance
-matrix for all input coordinates.
-For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
-Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
-Fourier transform of a positive finite measure.
-This means that the power spectrum of a positive definite kernel must be
-positive for all frequencies.
-This result is intuitive because, since the power spectrum of a process is
-defined as the expected squared amplitude of the Fourier transform of the time
-series, it must be non-negative.
-
-Using \eq{celerite-psd}, we find that for a single \celeriteterm\ term, this
-requirement is met when
-\begin{eqnarray}\eqlabel{psd-req}
-\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
-{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2} > 0 \quad.
-\end{eqnarray}
-The denominator is positive for all $c_j \ne 0$ and it can be shown that, when
-$c_j=0$, \eq{psd-req} is satisfied for all $\omega \ne d_j$ where the power is
-identically zero.
-Therefore, when $c_j \ne 0$, we require that the numerator is positive for all
-$\omega$.
-This requirement can also be written as
-\begin{eqnarray}
-a_j\,c_j > -b_j\,d_j \\
-a_j\,c_j > b_j\,d_j \quad.
-\end{eqnarray}
-Furthermore, we can see that $a_j$ must be positive since $k(0) = a_j$ should
-should be positive and, similarly, by requiring the covariance to be finite at
-infinite lag, we obtain the constraint $c_j \ge 0$.
-Combining these results, we find the constraint
-\begin{eqnarray}
-    |b_j\,d_j| < a_j\,c_j \quad.
-\end{eqnarray}
-
-In the case of $J$ \celeriteterm\ terms, we can check for negative values of the
-PSD by solving for the roots of the power spectrum; if there are any real,
-positive roots, then the power-spectrum goes negative (or zero), and thus does not
-represent a valid kernel. We rewrite the power spectrum, \eqalt{celerite-psd}),
-abbreviating with $z = \omega^2$:
-\begin{equation}
-S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
-\end{equation}
-where
-\begin{eqnarray}
-q_j &=& a_j\,c_j-b_j\,d_j\\
-r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
-s_j &=& 2(c_j^2-d_j^2)\\
-t_j &=& (c_j^2+d_j^2)^2.
-\end{eqnarray}
-The denominators of each term are positive, so we can multiply through by
-$\prod_j \left(z^2+s_jz + t_j\right)$ to find
-\begin{equation}
-Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
-    t_k\right) = 0\quad,
-\end{equation}
-which is a polynomial with order $2\,(J-1)+1$.
-With $J=2$, this yields a cubic equation whose roots can be obtained exactly.
-
-For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
-allows one to determine whether there are any real roots within the range
-$(0,\,\infty]$.
-We first construct $Q_0(z)$ and its derivative $Q_1(z) = {Q_0}^\prime(z)$,
-and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
-\begin{eqnarray}
-Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
-\end{eqnarray}
-where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
-dividing $p(z)$ by $q(z)$.
-
-We evaluate the coefficients of each of the polynomial in the series by
-evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
-of these polynomials evaluated at $z=0$.
-Likewise, we evaluate the coefficients of the largest order term in each
-polynomial that gives the sign of the polynomial as $z \rightarrow \infty$,
-$f_\infty$.
-With the sequence of coefficients $f_0$ and $f_\infty$, we then determine how
-many times the sign changes in each of these, where $\sigma(0)$ is the number
-of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
-at $z \rightarrow \infty$.
-The total number of real roots in the range
-$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.
-
-We have checked that this procedure works for a wide range of parameters, and
-we find that it robustly matches the number of positive real roots which we
-evaluated numerically.
-The advantage of this procedure is that it does not require computing the
-roots, but only carrying out algebraic manipulation of polynomials to
-determine the number of positive real roots.
-If a non-zero real root is found, the likelihood may be set to zero.
-
 \newpage

### 2017-06-09 paper/ms.tex
Some minor edits.

@@
 component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.
+triangular operation.  An example of this type of matrix is given in appendix \ref{operations}.
 In this representation, rather than storing the full $N \times N$ matrix $K$,
@@
 takes $\sim 1.5\,\mathrm{ms}$~--~a speed-up of more than three orders of
-magnitude per model evaluation.
+magnitude per model evaluation. % :) closer to four orders of magnitude!
 
@@
 These samples give a posterior constraint on the period of $P_\mathrm{rot} =
-\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
+\rotationperiod$ days and the marginalized posterior distribution for $P$ is shown
 in the right panel of \Figure{rotation}.
@@
 \response{
-For this demonstration, we randomly select a month-long segment of PDC
+For this demonstration, we randomly select a month-long segment of PDC data
 \citep{Stumpe:2012, Smith:2012} \kepler\ data.
@@
 and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
-the model.
+the model; $\sigma$ is real, while the others are complex.
 Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
@@
 solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
-While the computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$,
-the Kalman filtering based method for solving CARMA models \citep{Kelly:2014}
-is, in practice, somewhat faster and the memory requirements are smaller.
-One benefit of the \celeriteterm\ model is that it is parameterized in terms
+The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
+the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
+however, in practice this method is somewhat slower.
+An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
 of the amplitudes directly.
@@
 stochastically-driven, damped simple harmonic oscillators, they are a
-physically motivated choice of covariance function in many applications.
+physically motivated choice of covariance function in many astronomical applications.
 
@@
 
-We evaluate the coefficients of each of the polynomial in the series by
+We evaluate the coefficients of each of the polynomials in the series by
 evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
@@
 
-\section{Semiseparable operations}
-
-To help clarify these equations, we write out here an example of a semiseparable matrix with a
-single \celeriteterm\ term with $c=d=1$: %and $N=3$
+\section{Semiseparable operations} \label{operations}
+
+To help clarify equations \ref{system}, we write out here an example of a semiseparable matrix with a
+single \celeriteterm\ term with $c=d=1$:
 \begin{eqnarray}\eqlabel{semi-separable-structure}
@@
 \begin{pmatrix}
-%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)} & {\scriptscriptstyle e^{-c\,\tau_{3,1}}\left(a\,\cos{(d \tau_{3,1})}+b\,\sin{(d \tau_{3,1})}\right)}\\
-%    {\scriptscriptstyle e^{-c\,\tau_{2,1}}\left(a\,\cos{(d \tau_{2,1})}+b\,\sin{(d \tau_{2,1})}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}\\
-%    {\scriptscriptstyle e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)}&{\scriptscriptstyle  e^{-c\,\tau_{3,2}}\left(a\,\cos{(d \tau_{3,2})}+b\,\sin{(d \tau_{3,2})}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
-%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\
-%    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
-%    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
     {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
@@
 \begin{pmatrix}
-%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
-%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
-%    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
     {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
@@
     {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
-%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & \ldots \\
-%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& \ldots \\
     \ldots & \ldots & \ldots & \ldots
@@
 &+&\mathrm{tril}\left[\begin{pmatrix}
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
     {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
@@
    \ldots & \ldots
-\end{pmatrix} %\right.\\
-%&\times &
+\end{pmatrix}
 \times
-%\left.
 \begin{pmatrix}
-% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} &{\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}}\\
-% {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
-% {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_3} \cos{t_3}}\\
-% {\scriptscriptstyle e^{t_1} \sin{t_1}} & {\scriptscriptstyle e^{t_2} \sin{t_2}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}
  {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
@@
 \mathrm{triu}\left[\begin{pmatrix}
-% {\scriptscriptstyle e^{c\,t_1} \cos{(d t_1)}} & {\scriptscriptstyle e^{c\,t_1} \sin{(d t_1)}} \\
-% {\scriptscriptstyle e^{c\,t_2} \cos{(d t_2)}} & {\scriptscriptstyle e^{c\,t_2} \sin{(d t_2)}}\\
-% {\scriptscriptstyle e^{c\,t_3} \cos{(d t_3)}} & {\scriptscriptstyle e^{c\,t_3} \sin{(d t_3)}}
  {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
@@
  \ldots & \ldots
-\end{pmatrix} % \right.\\
+\end{pmatrix} 
 \times
-%\left.
 \begin{pmatrix}
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\cos{(d t_1)}+b\,\sin{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\cos{(d t_2)}+b\,\sin{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\cos{(d t_3)}+b\,\sin{(d t_3)}\right)}\\
-%    {\scriptscriptstyle e^{-c\,t_1}\left(a\,\sin{(d t_1)}-b\,\cos{(d t_1)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_2}\left(a\,\sin{(d t_2)}-b\,\cos{(d t_2)}\right)}&
-%    {\scriptscriptstyle e^{-c\,t_3}\left(a\,\sin{(d t_3)}-b\,\cos{(d t_3)}\right)}
-%    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-%    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-%    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}\\
-%    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}&
-%    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}&
-%    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}
     {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
@@
 
+In addition to solving $K^{-1} {\bf y}$, several other operations can be
+carried out recursively in $\mathcal{O}(N)$, simulation, multiplication, 
+and prediction, which we describe in the following sections.
+
 \subsection{Simulation}
@@
 comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that $\mathcal{Q} = -2 \mathcal{L}$
+where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
 with $\mu_\bvec{\theta} =0$ is given by
@@
 The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$.
+which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
+diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ is the $\chi^2$ statistic.
 
@@
 We expect that this simulation algorithm will be useful for test-driving Gaussian
-process inference algorithms applied to large datasets; even simulation of
+process inference algorithms applied to large datasets as even simulation of
 Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
@@
 \begin{eqnarray}
-\bar f_* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
+\bar f^* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
 &=& K(\bvec{t^*},\bvec{t}) \bvec{z}

### 2017-06-09 paper/ms.tex
some edits
typo

@@
 their application has been limited to relatively small datasets.
-In this paper, we present a method for Gaussian Process modeling in
+In this paper, we present a novel method for Gaussian Process modeling in
 one-dimension where the computational requirements scale linearly with the
@@
 that it is effective in other cases.
-We present a mathematical description of the method, the details of the
-implementation, and a comparison to existing scalable Gaussian Process methods.
+We present a mathematical description of the method and compare it to existing
+scalable Gaussian Process methods.
 The method is flexible, fast, and most importantly, interpretable, with a wide
@@
 method in \project{C++}, \project{Python}, and \project{Julia}.
-A change.
 
@@
 consideration of the impact of this decision.
-While this doesn't necessarily impact the inferred results, in this section we
-discuss an exact physical interpretation of the \celeriteterm\ kernel that
-will be applicable to many astrophysical systems but especially the
-time-variability of stars.
+In this section, we discuss an exact physical interpretation of the
+\celeriteterm\ kernel that is applicable to many astrophysical systems, but
+especially the time-variability of stars.
 
@@
 where $\eta = \vert 1-(4\,Q^2)^{-1}\vert^{1/2}$.
-We note that, because of the damping, the characteristic oscillation frequency
-in this model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to
-the frequency of the undamped oscillator, $\omega_0$.
+Because of the damping, the characteristic oscillation frequency in this
+model, $d_j$, for any finite quality factor $Q > 1/2$, is not equal to the
+frequency of the undamped oscillator, $\omega_0$.
 
@@
 This functional form is commonly used to model for the background granulation
-noise in asteoreseismic and helioseismic \citep{Harvey:1985, Michel:2009,
-Kallinger:2014}  analyses.
+noise in asteoreseismic and helioseismic analyses \citep{Harvey:1985, Michel:2009,
+Kallinger:2014}.
 The kernel for this value of $Q$ is
@@
 \end{eqnarray}
-or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
+or, equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
 \begin{eqnarray}\eqlabel{approx-matern2}
@@
 on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
-\citet{Ambikasaran:2015} derived a scalable direct solver for any general
-semiseparable matrix and applied this method to \celeriteterm\ models with
-$b_j = 0$ and $d_j = 0$.
+Ambikasaran derived a scalable direct solver for any general semiseparable
+matrix and applied this method to \celeriteterm\ models with $b_j = 0$ and
+$d_j = 0$.
 During the preparation of this paper, we generalized this earlier method to
 include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
-performance algorithm by restricting our method to positive semi-definite
-(PSD) matrices.
+performance algorithm by restricting our method to positive definite matrices.
 In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
-PSD.
+positive.
 
@@
 
-\subsection{Cholesky factorization of PSD semiseparable matrices}
+\subsection{Cholesky factorization of positive definite semiseparable matrices}
 
@@
 linear systems of the form $K\,\bvec{z}=\bvec{y}$ for $\bvec{z}$, where $K$ is
-a rank-$R$ semiseparable PSD matrix.
+a rank-$R$ semiseparable positive definite matrix.
 These can both be computed using the Cholesky factorization of $K$:
@@
 \end{eqnarray}
-where, since $K$ is PSD, $D_{n,n}>0$ for all $n$.
+where, since $K$ is positive definite, $D_{n,n}>0$ for all $n$.
 
@@
 The algorithm derived here is a general method for factorizing, solving, and
-computing the determinant of PSD rank-$R$ semiseparable matrices.
-As we will show below, with some care, this method can be used to compute
+computing the determinant of positive definite rank-$R$ semiseparable
+matrices.
+As we will show below, this method can be used to compute
 \eq{gp-likelihood} for \celeriteterm\ models.
-In practice, when $K$ is PSD and this method is applicable, we find that this
-method is about a factor of 20 faster than an optimized implementation of the
-earlier, more general method from \citet{Ambikasaran:2015}.
+In practice, when $K$ is positive definite and this method is applicable, we
+find that this method is about a factor of 20 faster than an optimized
+implementation of the earlier, more general method from
+\citet{Ambikasaran:2015}.
 
@@
 For each pair of $N$ and $J$ we repeated the experiment 10 times with
-different randomly generated datasets and \celeriteterm\ parameters and
+different randomly generated datasets and \celeriteterm\ parameters, and
 averaged across experiments to remove some sampling noise from this figure.
@@
 This experiment was run on a 2013 MacBook Pro with a dual 2.6 GHz Intel Core
-i5 processor, but find similar performance on other platforms.
-The \celerite\ implementation used for this test is the Python interface with
-a C++ back end, but there is little overhead introduced by Python so we find
-similar performance using implementations in C++ and Julia
-\citep{Bezanzon:2012}.
+i5 processor, but we find similar performance on other platforms.
+The \celerite\ implementation used for this test is the \project{Python}
+interface with a \project{C++} back end, but there is little overhead
+introduced by \project{Python} so we find similar performance using
+implementations in \project{C++} and \project{Julia} \citep{Bezanzon:2012}.
 \Figure{benchmark} shows the computational cost of evaluating
@@
 \response{
-To show the application of \celerite, we use it to perform posterior
+To demonstrate the application of \celerite, we use it to perform posterior
 inference in two examples with realistic simulated data and three examples of
 real-world research problems.
-These examples have been chosen to justify the application of \celerite\ for a
-range of use cases, but they are by no means exhaustive.
+These examples have been chosen to justify the use of \celerite\ for a range
+of research problems, but they are by no means exhaustive.
 These examples are framed in the time domain with a clear bias in favor of
@@
 \emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
-We initialize \exampleinwalkers~walkers in a Gaussian ball with width
-$10^{-4}$ in each parameter around the MAP parameter vector, run
-\exampleinburn~steps of burn-in, and \exampleinsteps~steps of MCMC.
+We initialize \exampleinwalkers~walkers by sampling from three-dimensional
+isotropic Gaussian centered the MAP parameter vector and with a standard
+deviation of $10^{-4}$ in each dimension.
+We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
 To assess convergence, we estimate the mean integrated autocorrelation time of
@@
 In this figure, the true PSD is plotted as a dashed black line and the
-numerical estimate of the posterior inference is shown as a blue contour
-indicating 68\% of the MCMC samples.
+numerical estimate of the posterior constraint on the PSD is shown as a blue
+contour indicating 68\% of the MCMC samples.
 It is clear from this figure that, as expected, the inference correctly
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/rotation/rotation.pdf}
+\includegraphics[width=0.9\textwidth]{figures/rotation/rotation.pdf}
 \caption{Inferred constraints on a quasiperiodic GP model using the covariance
@@
 
-\response{
-For this demonstration, we randomly select a month-long segment of PDC data
-\citep{Stumpe:2012, Smith:2012} \kepler\ data.
+\response{For this demonstration, we randomly select a month-long segment of
+PDC \kepler\ data \citep{Stumpe:2012, Smith:2012}.
 Unlike the previous examples, the posterior distribution is sharply
@@
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/astero/astero.pdf}
+\includegraphics[width=0.5\textwidth]{figures/astero/astero.pdf}
 \caption{

### 2017-06-09 paper/ms.tex
a few more edits

@@
 \includegraphics[width=0.9\textwidth]{figures/simulated/correct.pdf}
-\caption{(top left) A simulated dataset (black error bars), and maximum
-    likelihood model (blue contours).
+\caption{(top left) A simulated dataset (black error bars), and MAP
+    model (blue contours).
     (bottom left) The residuals between the mean predictive model and the data
@@
 MCMC.
-To estimate the number of effective samples, we estimate the integrated
+To estimate the number of independent samples, we estimate the integrated
 autocorrelation time of the chain for the parameter $\omega_1$~--~the
@@
 \includegraphics[width=0.9\textwidth]{figures/simulated/wrong-qpo.pdf}
-\caption{(top left) A simulated dataset (black error bars), and maximum
-    likelihood model (blue contours).
+\caption{(top left) A simulated dataset (black error bars), and MAP
+    model (blue contours).
     (bottom left) The residuals between the mean predictive model and the data
@@
     function in \eq{rot-kernel} and two quarters of \kepler\ data.
-(top left) The \kepler\ data (black points) and the maximum likelihood model
+(top left) The \kepler\ data (black points) and the MAP model
     prediction (blue curve) for a 60~day subset of the data used.
@@
 oscillations because the standard methods use statistics of the empirical
-periodogram and cannot formally propagate the measurement uncertainties to the
-constraints on physical parameters, instead relying on bootstrapped
-uncertainty estimators \citep{Huber:2009}.
+periodogram.
+These methods cannot formally propagate the measurement uncertainties to the
+constraints on physical parameters and they instead rely on population-level
+bootstrap uncertainty estimates \citep{Huber:2009}.
 More sophisticated methods that compute the likelihood function in the time
@@
 standard procedure described by \citet{Huber:2009}.
-We then run set of \project{L-BFGS-B} optimizations with values of
+We then run a set of \project{L-BFGS-B} optimizations with values of
 $\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
@@
 method-dependent uncertainties \citep{Verner:2011} so further experiments
-would be needed to fully assess the reliability of this method.
+would be needed to fully assess the reliability of this specific model.
 
@@
 \response{
-(top) The \kepler\ data (black points) and the maximum likelihood model
+(top) The \kepler\ data (black points) and the MAP model
     prediction (blue curve) for a 10~day subset of the month-long dataset that
@@
     KIC~1430163 with a synthetic transit model injected (black points) and the
-    maximum likelihood model for the stellar variability
+    MAP model for the stellar variability
     (blue line).
-    \emph{(bottom)} The maximum likelihood ``de-trending'' of the data in
+    \emph{(bottom)} The MAP ``de-trending'' of the data in
     the top panel.
-    In this panel, the maximum likelihood model for the stellar variability
+    In this panel, the MAP model for the stellar variability
     has been subtracted to leave only the transits.
-    The de-trended light curve is shown by black error bars and the maximum
-    likelihood transit model is shown as a blue line.
+    The de-trended light curve is shown by black error bars and the MAP
+    transit model is shown as a blue line.
     \figurelabel{transit-ml}}

### 2017-06-09 paper/ms.tex
Some edits to allay referee's concerns

@@
 Gaussian Processes are a popular class of models used for this purpose but,
-since the computational cost scales as the cube of the number of data points,
+since the computational cost commonly scales as the cube of the number of data points,
 their application has been limited to relatively small datasets.
@@
 As we will discuss in the following pages, this choice of kernel function
-arises naturally in some physical systems and we demonstrate it can be used as
-an effective model in some other cases.
+arises naturally in physical systems and we demonstrate it can be used as
+an effective model in other cases.
 This method is especially appealing compared to other similar methods~--~we
@@
 GP models have been widely used across the physical sciences but their
-application is generally limited to small datasets because the cost of
+application is generally limited to small datasets, $N \approx 10^3$, because the cost of
 computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
@@
     show the 68\% credible region.
-    \figurelabel{astero}}
+    \figurelabel{astero_psd}}
 \end{center}
@@
 
-\section{Summary}
+\section{Summary and discussion}
 
 Gaussian Process models have been fruitfully applied to many problems in
-astronomical data analysis, but the fact that the computational cost scales as
-the cube of the number of data points has limited their use to relatively
+astronomical data analysis, but the fact that the computational cost generally
+scales as the cube of the number of data points has limited their use to relatively
 small datasets.
@@
 A drawback of this method is its quadratic scaling with the number of
-terms $J$, but, in many cases, small values of $J$ are sufficient.
+terms $J$, but, in some cases, small values of $J$ may suffice.
 
@@
 \celerite\ can be used to model stellar oscillations using the relation to the
-mixture of stochastically-driven, damped simple harmonic oscillators.
+mixture of stochastically-driven, damped simple harmonic oscillators, as
+we have demonstrated in this paper.
 As higher signal-to-noise observations of transiting exoplanet systems are
@@
 functions \citep{Knutson:2007, Zucker:2007}, and more.
+
+\subsection{Other applications and limitations}
 
@@
 work.
+
+\subsection{Code availability}
 
@@
 
-To help clarify equations \ref{system}, we write out here an example of a semiseparable matrix with a
+To help clarify equations \eq{system}, we write out here an example of a semiseparable matrix with a
 single \celeriteterm\ term with $c=d=1$:
@@
 \begin{eqnarray}
-f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
-f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
+f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
+f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
 y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
 \end{eqnarray}
-where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
+where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,2J$.  The recursion requires
 two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
-from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
-indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}
+from $n=2$ to $n=N$ to compute $f^-_{n,j}$.
 

### 2017-06-10 paper/ms.tex
adding carma benchmark figure

--- 2017-06-09 paper/ms.tex
@@
 These samples give a posterior constraint on the period of $P_\mathrm{rot} =
-\rotationperiod$ days and the marginalized posterior distribution for $P$ is shown
-in the right panel of \Figure{rotation}.
+\rotationperiod$ days and the marginalized posterior distribution for $P$ is
+shown in the right panel of \Figure{rotation}.
 This result is in good agreement with the literature value with smaller
@@
 more reliable than other methods.
+
+The total computational cost for this inference using \celerite\ is about
+4~CPU-minutes.
+By contrast, the same inference using a general but optimized Cholesky
+factorization routine would require nearly 400~CPU-hours.
+This speed-up enables probabilistic measurement of rotation periods using
+existing data from \project{K2} and forthcoming surveys like \tess\ and \lsst\
+where this inference will need to be run for at least hundreds of thousands of
+stars.
 }
@@
 
-\response{
-In this section, we demonstrated five potential use cases for \celerite.
-These examples span a range of data sizes and model complexities, and the
-last three are all based on active areas of research with immediate real-world
+\response{In the previous subsections, we demonstrate five potential use
+cases for \celerite.
+These examples span a range of data sizes and model complexities, and the last
+three are based on active areas of research with immediate real-world
 applicability.
@@
 Table~\ref{tab:example-stats} lists the specifications of each example.
-
-In Example~3, for example, we inferred the rotation period of a star observed
-by \kepler\ with a cost of about 4~CPU-minutes.
-By contrast, the same inference using a general but optimized Cholesky
-factorization routine would require nearly 400~CPU-hours.
-The upcoming \tess\ Mission will observe \dfmtodo{XXX} stars.
+In this table, we list the computational cost required for a single likelihood
+evaluation using a general Cholesky factorization\footnote{We use the Intel
+Math Kernel Library \url{https://software.intel.com/en-us/intel-mkl}}
+and the cost of computing the same likelihood using \celerite.
+Combining this with the total number of model evaluations required to run the
+MCMC algorithm to convergence (this number is also listed in
+Table~\ref{tab:example-stats}), we can estimate the total CPU time required in
+each case.
+While the exact computational requirements for any problem also depend on the
+choice of inference algorithm and the specific precision requirements, this
+table provides a rough estimate of what to expect for projects at these
+scales.
 }
@@
 
-\subsection{Kalman filtering \& CARMA models} \sectlabel{kalman}
-
-\response{
 The continuous autoregressive moving average (CARMA) models introduced into
@@
 \celerite.
-CARMA models are derived by solving a stochastic differential equation and,
-like \celerite, the likelihood function for a CARMA model conditioned on a
-one-dimensional dataset can be solved in $\mathcal{O}(N)$.
-The kernel function for a CARMA$(J,\,K)$ model is \citep{Kelly:2014}
+CARMA models are restricted to one-dimensional problems and the likelihood
+function for a CARMA model can be solved in $\mathcal{O}(N\,J^2)$ using a
+Kalman filter \citep{Kelly:2014}.
+The kernel function for a CARMA$(J,\,K)$ model is
 \begin{eqnarray}\eqlabel{carma-kernel}
@@
 \citep[for example][]{Luger:2017}.
-}
+
+\begin{figure}[tp]
+\begin{center}
+\includegraphics[width=\textwidth]{figures/benchmark_darwin_carma.pdf}
+\caption{The same as \Figure{benchmark}, but using an optimized
+    \project{C++} implementation of the Kalman filter method developed by
+    \citet{Kelly:2014}.
+    Comparing this figure to \Figure{benchmark}, we see that \celerite\ is
+    about an order of magnitude faster in all cases.
+    \emph{(left)} The cost of computing \eq{gp-likelihood} with a covariance
+    matrix given by \eq{celerite-kernel} as a function of the number of data
+    points $N$.
+    The different lines show the cost for different numbers of terms $J$
+    increasing from bottom to top.
+    To guide the eye, the straight black line without points
+    shows linear scaling in $N$.
+    \emph{(right)} The same information plotted as a function of $J$ for
+    different values of $N$.
+    Each line shows the scaling for a specific value of $N$ increasing from
+    bottom to top.
+    The black line shows quadratic scaling in $J$.
+    \figurelabel{benchmark-carma}}
+\end{center}
+\end{figure}
 

### 2017-06-11 paper/ms.tex
Minor edits; added discussion of power spectrum figure

--- 2017-06-09 paper/ms.tex
@@
 $\mathcal{O}(N\,J^2)$ scaling that can be used when
-$k_\bvec{\alpha}(\tau_{nm})$ is positive semi-definite.}
+$k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
 
@@
 In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
-positive.
+positive definite.
 
@@
 $J_\mathrm{complex}$ general terms, the semiseparable rank is only
-$J_\mathrm{real} + 2\,J_\mathrm{complex}$.}
+$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2J-J_{real}$.}
 where the components of the relevant matrices are
@@
 The left panel of \Figure{error} shows the L-infinity norm for the residual
-$\bvec{z} - K^{-1}\,\bvec{y}$ for a range of dataset sizes $N$ and numbers of
-terms $J$.
+$\bvec{z} - K^{-1}\,\bvec{y} = \bvec{z}-K^{-1}K\bvec{z}$ for a range of dataset sizes $N$ and numbers of terms $J$.
 For each pair of $N$ and $J$ we repeated the experiment 10 times with
@@
 introduced by \project{Python} so we find similar performance using
-implementations in \project{C++} and \project{Julia} \citep{Bezanzon:2012}.
+the implementation in \project{C++} and an independent implementation in
+\project{Julia} \citep{Bezanzon:2012}.
 \Figure{benchmark} shows the computational cost of evaluating
@@
 Math Kernel Library, a highly optimized general dense linear algebra package,
-is shown as a dashed line.
+is shown as a dashed line, which is independent of $J$.
 \celerite\ is faster than the reference implementation for all but the
-smallest datasets with many terms.
+smallest datasets with many terms, $J \ga 64$ for $N\approx 512$, 
+and $J$ increasing with larger $N$.
 
@@
 \emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
-We initialize \exampleinwalkers~walkers by sampling from three-dimensional
+We initialize \exampleinwalkers~walkers by sampling from a three-dimensional
 isotropic Gaussian centered the MAP parameter vector and with a standard
@@
 measured using traditional periodogram and autocorrelation function approaches
-applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}.
+applied to \kepler\ data from Quarters 0--16 \citep{Mathur:2014}, covering
+about four years.
 
@@
 constraints are tighter than the published results.
+\Figure{astero_psd} shows a comparison of
+the power spectrum estimated with a Lomb-Scargle periodogram 
+with the power spectrum inferred with \celerite.  Despite the much smaller 
+one-month sequence of data used in the \celerite\ analysis, the power spectrum 
+inferred with \celerite\ matches well the periodogram estimated from
+the full four years of Kepler data.
 All asteroseismic analyses are known to have substantial systematic and
@@
 
-\section{Summary and discussion}
+\section{Discussion}
 
@@
 which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
-diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ is the $\chi^2$ statistic.
+diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ becomes the $\chi^2$ statistic.
 
@@
 
-Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
+Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(RN)$
 operations with a stable recursive algorithm using the semiseparable form:
@@
 f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
-y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
-\end{eqnarray}
-where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,2J$.  The recursion requires
+y_n &=& A_{n,n} z_n + \sum_{j=1}^R \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
+\end{eqnarray}
+where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,R$.  The recursion requires
 two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
@@
 decomposition as described above.  The second step multiplication of the white-noise-free
-kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
+kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(RN)$ steps with this multiplication algorithm.
 Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,

### 2017-06-11 paper/ms.tex
edits

@@
 $J_\mathrm{complex}$ general terms, the semiseparable rank is only
-$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2J-J_{real}$.}
+$R=J_\mathrm{real} + 2\,J_\mathrm{complex}=2\,J-J_{real}$.}
 where the components of the relevant matrices are
@@
 constraints are tighter than the published results.
-\Figure{astero_psd} shows a comparison of
-the power spectrum estimated with a Lomb-Scargle periodogram 
-with the power spectrum inferred with \celerite.  Despite the much smaller 
-one-month sequence of data used in the \celerite\ analysis, the power spectrum 
-inferred with \celerite\ matches well the periodogram estimated from
-the full four years of Kepler data.
+The top two panels of \Figure{astero_psd} show the Lomb-Scargle periodogram
+\citep{VanderPlas:2017} estimator for the power spectrum based on the full
+4~years of \kepler\ data and the month-long subset used in our analysis.
+The bottom panel of \Figure{astero_psd} shows the posterior inference of the
+PSD using the \celerite\ and the model described here.
+Despite only using one month of data, the \celerite\ inference captures the
+salient features of the power spectrum and it is qualitatively consistent with
+the standard estimator applied to the full baseline.
 All asteroseismic analyses are known to have substantial systematic and
@@
 f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
-y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
-\end{eqnarray}
-for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
+y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2\,J} \tilde{U}_{n,k} f_{n,k}
+\end{eqnarray}
+for $n = 1$ to $N$ and $j=1$ to $2\,J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
 We expect that this simulation algorithm will be useful for test-driving Gaussian
@@
 \end{eqnarray}
-where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
+where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$
+to $2\,J$, and in $X^\pm_{m,n,i}$ the expressions
 for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
@@
 \begin{equation}
-\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
+\bar f^*_i = \sum_{i=1}^{2\,J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
 \end{equation}

### 2017-06-12 paper/ms.tex
snark about model selection

@@
 Gaussian Processes are a popular class of models used for this purpose but,
-since the computational cost commonly scales as the cube of the number of data points,
-their application has been limited to relatively small datasets.
+since the computational cost scales, in general, as the cube of the number of
+data points, their application has been limited to small datasets.
 In this paper, we present a novel method for Gaussian Process modeling in
@@
 motivation for and interpretation of this choice~--~but we also demonstrate
-that it is effective in other cases.
+that it can be a useful effective model in some other cases.
 We present a mathematical description of the method and compare it to existing
 scalable Gaussian Process methods.
-The method is flexible, fast, and most importantly, interpretable, with a wide
-range of potential applications within astronomical data analysis and beyond.
+The method is fast and interpretable, with a wide range of potential
+applications within astronomical data analysis and beyond.
 We provide well-tested and documented open-source implementations of this
@@
 to include the effects of correlated noise.
+\response{
 A point estimate for the values of the parameters $\bvec{\theta}$ and
@@
 maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
-$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.
-\response{Furthermore}, probabilistic constraints on $\bvec{\theta}$ and
-$\bvec{\alpha}$ can be obtained by multiplying the likelihood by a prior
-$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC)
-algorithm to sample the joint posterior probability density.
+$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.}
+\response{Furthermore, the uncertainties on $\bvec{\theta}$ and
+$\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
+$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov chain Monte Carlo (MCMC)
+algorithm to sample the joint posterior probability density.}
 
 GP models have been widely used across the physical sciences but their
-application is generally limited to small datasets, $N \approx 10^3$, because the cost of
-computing the inverse and determinant of a general matrix $K_\bvec{\alpha}$ is
-$\mathcal{O}(N^3)$.
+application is generally limited to small datasets, \response{$N \lesssim
+10^3$}, because the cost of computing the inverse and determinant of a general
+matrix $K_\bvec{\alpha}$ is $\mathcal{O}(N^3)$.
 In other words, this cost is proportional to the cube of the number of data
@@
 \end{eqnarray}
-Therefore, the method described in the following section can be used to
+Therefore, the method described in \sect{factor} can be used to
 perform scalable inference on large datasets for any model, where the kernel
@@
 
+\subsection{How to choose a celerite model}
+
+In any application of \celerite~--~or any other GP model, for that
+matter~--~one must choose a kernel.
+A detailed discussion of how to choose a kernel is beyond the scope of this
+paper and the interested reader is directed to Chapter~5 of
+\citet{Rasmussen:2006} a general introduction, but we include a summary and
+some suggestions specific to \celerite\ here.
+We argue that the most satisfying method of choosing a kernel is to use
+physical motivation and the discussion in \sect{sho} provides an opportunity
+for justified model specification when \celerite\ is applied to parameter
+estimation from stellar light curves.
+However, it is not always possible to rely on physics to specify the model
+and, in these cases, it is useful to have other tools at one's disposal.
+One option is to compute the evidence integral or Bayes factor for a set of
+models and choose the model with the highest evidence.
+Another option is to perform cross validation and select the model with the
+best predictive performance.
+For continuous autoregressive moving average models (we return to these in
+\sect{compare}), \citet{Kelly:2014} recommend the Bayesian Information
+Criterion (BIC) as a method for selecting the order of the model.
+This is a total hack, but I guess that you could do it if you wanted.
+
 \section{Examples}\sectlabel{examples}
@@
 
-The continuous autoregressive moving average (CARMA) models introduced into
-the astrophysics literature by \citet{Kelly:2014} share many features with
+The continuous autoregressive moving average (CARMA) models introduced to the
+astrophysics literature by \citet{Kelly:2014} share many features with
 \celerite.
@@
 solving \eq{carma-coeff} numerically for a given set of $\{A_j\}_{j=1}^J$.
+
+\response{
 The computational scaling of CARMA models is also $\mathcal{O}(N\,J^2)$ using
-the Kalman filtering based method for solving CARMA models \citep{Kelly:2014};
-however, in practice this method is somewhat slower.
-An additional benefit of the \celeriteterm\ model is that it is parameterized in terms
-of the amplitudes directly.
-This means that the physical interpretation of these parameters is, in some
-cases, simpler.
-On the other hand, CARMA models are parameterized by modeling the power
-spectrum of the stochastic driving force and this might be the quantity of
-interest in other situations.
-Another benefit of the \celeriteterm\ framework is that, unlike
-filtering-based CARMA solvers, it directly solves matrix equations of the form
-${K_\bvec{\alpha}}^{-1}\,b$ in $\mathcal{O}(N)$.
-This is a crucial feature when GP modeling is combined with linear models
+the Kalman filter method \citep{Kelly:2014}, but we find that, in practice,
+this method is more computationally expensive than \celerite, but it has a
+smaller memory footprint.
+\Figure{benchmark-carma} shows the scaling of the Kalman filter solver for the
+systems shown in \Figure{benchmark}.
+Comparing these figures, we find that the Kalman filter solver is slower than
+\celerite\ for all the systems we tested with an average difference of about
+an order of magnitude.
+
+Another benefit of \celerite\ compared to Kalman filter solvers is the fact
+that it is a direct solver for the relevant matrix equations instead of just
+an algorithm to compute \eq{gp-likelihood}.
+This means that the inverse covariance matrix $K^{-1}$ can be applied to
+general vectors and matrices in $\mathcal{O}(N)$, while reusing the
+factorization.
+This is a crucial feature when GP models are combined with linear regression
 \citep[for example][]{Luger:2017}.
+}
 
@@
 \includegraphics[width=\textwidth]{figures/benchmark_darwin_carma.pdf}
-\caption{The same as \Figure{benchmark}, but using an optimized
+\caption{\response{The same as \Figure{benchmark}, but using an optimized
     \project{C++} implementation of the Kalman filter method developed by
@@
     bottom to top.
-    The black line shows quadratic scaling in $J$.
+    The black line shows quadratic scaling in $J$.}
     \figurelabel{benchmark-carma}}
@@
 \end{figure}
-
-\subsection{Other scalable methods}
 
@@
 Another GP method that has been used extensively in astronomy is the
-hierarchical off-diagonal low rank \citep[HODLR,][]{Ambikasaran:2016} solver.
+hierarchical off-diagonal low rank (HODLR) solver \citep{Ambikasaran:2016}.
 This method exploits the fact that many commonly used kernel functions produce
@@
 scales approximately linearly with $J$, but it requires substantial overhead
-for small models.
-This means that the HODLR solver can \emph{approximately} evaluate
-\celeriteterm\
-models more efficiently than \celerite\ for large models $J \gg 10$ and
-small datasets where $N$ is smaller than a few thousand.
+making it much slower than \celerite\ for all the models we tested.
+For large $J \gtrsim 256$ and small $N \lesssim 1000$, we find that
+\project{george} can approximately evaluate \celeriteterm\ models with the
+same or less computational cost than \celerite, but that \celerite\ is
+faster~--~and exact~--~in all other parts of parameter space.
 
@@
 astrophysical variability.
-In fact, the relation of the \celeriteterm\ kernel to the damped, stochastically-driven harmonic
-oscillator matches simple models of astrophysical variability, and makes the
-parameterization interpretable in terms of resonant frequency, amplitude, and
-quality factor.
-% Siva: Removed major, since $J$ is most often small as stated in the second half of the sentence
+In fact, the relation of the \celeriteterm\ kernel to the damped,
+stochastically-driven harmonic oscillator matches simple models of
+astrophysical variability, and makes the parameterization interpretable in
+terms of resonant frequency, amplitude, and quality factor.
 A drawback of this method is its quadratic scaling with the number of

### 2017-06-12 paper/ms.tex
adding discussion of model selection and example

@@
 In any application of \celerite~--~or any other GP model, for that
-matter~--~one must choose a kernel.
-A detailed discussion of how to choose a kernel is beyond the scope of this
-paper and the interested reader is directed to Chapter~5 of
-\citet{Rasmussen:2006} a general introduction, but we include a summary and
-some suggestions specific to \celerite\ here.
-We argue that the most satisfying method of choosing a kernel is to use
-physical motivation and the discussion in \sect{sho} provides an opportunity
-for justified model specification when \celerite\ is applied to parameter
-estimation from stellar light curves.
-However, it is not always possible to rely on physics to specify the model
-and, in these cases, it is useful to have other tools at one's disposal.
-One option is to compute the evidence integral or Bayes factor for a set of
-models and choose the model with the highest evidence.
-Another option is to perform cross validation and select the model with the
-best predictive performance.
-For continuous autoregressive moving average models (we return to these in
-\sect{compare}), \citet{Kelly:2014} recommend the Bayesian Information
-Criterion (BIC) as a method for selecting the order of the model.
-This is a total hack, but I guess that you could do it if you wanted.
+matter~--~we must choose a kernel.
+A detailed discussion of model selection is beyond the scope of this paper and
+the interested reader is directed to Chapter~5 of \citet{Rasmussen:2006} an
+introduction in the context of GPs, but we include a summary and some
+suggestions specific to \celerite\ here.
+
+For many of the astrophysics problems that we will tackle using \celerite, it
+is possible to motivate the kernel choice physically.
+We give three examples of this in the next section and, in these cases, we
+have a physical description of the system that generated our data, we
+represent that system as a \celeriteterm\ model, and use \celerite\ for
+parameter estimation.
+For other projects, we might be interested in comparing different theories
+and, in those case, we would need to perform formal model selection.
+
+Sometimes, it is not possible to choose a physically motivated kernel when we
+use a GP as an effective model.
+In these cases, we recommend starting with a mixture of stochastically-driven
+damped SHOs (as discussed in \sect{sho}) and select the number of oscillators
+using Bayesian model comparison, cross-validation, or another model selection
+technique.
+Despite the fact that it is less flexible than a general \celeriteterm\
+kernel, we recommend the SHO model in most cases because it is 1-time mean
+square differentiable
+\begin{eqnarray}
+\left.\frac{\dd k_\mathrm{SHO}(\tau)}{\dd \tau}\right|_{\tau=0} = 0 \quad,
+\end{eqnarray}
+giving smoother functions than the general \celeriteterm\ model that is only
+mean square continuous.
+
+For a similar class of models~--~we return to these below in
+\sect{compare}~--~\citet{Kelly:2014} recommend using an ``information
+criterion'', such as the Akaike information criterion (AIC) or the Bayesian
+information criterion (BIC), as a model selection metric.
+These criteria are easy to implement and we use the BIC in \sect{example2}
+but, since are only formally valid under specific sets of assumptions, we
+cannot recommend their use in general.
 
@@
 
+We choose this functional form by comparing the Bayesian information criterion
+(BIC) for a set of different models.
+In each model, we take the sum or product of $J$ SHO terms, find the maximum
+likelihood model using \project{L-BFGS-B} and compute the BIC
+\citep{Schwarz:1978}
+\begin{eqnarray}
+\mathrm{BIC} = -2\,\mathcal{L}^* + K\,\log N
+\end{eqnarray}
+where $\mathcal{L}^*$ is the value of the likelihood at maximum, $K$ is the
+number of parameters, and $N$ is the number of data points.
+\Figure{bic} shows the value of the BIC for a set of products and sums of SHO
+terms and the model that we choose has the lowest BIC.
+
 As in the previous example, we set the mean function to zero and can,
@@
     \figurelabel{simulated-wrong}}
+\end{center}
+\end{figure}
+
+\begin{figure}[!htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{figures/simulated/bic.pdf}
+\caption{\response{
+    The Bayesian information criterion (BIC) evaluated for the data from
+    \Figure{simulated-wrong} and different kernel choices.
+    The $x$-axis shows the number of terms included in each model.
+    The blue circles show models where the terms have been summed and the
+    orange squares indicate that the model is a product of terms.
+    In the product models, the parameters $S_0$ and $Q$ are held fixed at $1$
+    and $1/\sqrt{2}$ respectively for all but the first term.
+    The dashed black line shows the minium value of the BIC and this
+    corresponds to the model chosen in the text.}
+    \figurelabel{bic}}
 \end{center}

### 2017-06-12 paper/response.txt
adding response

We would first like to thank the referee for their thorough and constructive
review. We have taken the feedback to heart and we think that the manuscript
has been greatly improved as a result.

Beyond addressing each of the referee's concerns in detail, we have also
derived a new algorithm for solving celerite models that is about a factor of
20 faster than the submitted version (and about and order of magnitude faster
than a reference implementation of a Kalman filter based method). We have
updated the manuscript to describe and test this new method in Section 5. We
hope that the referee will consider this updated method and welcome any
feedback.

The only other major structural change was to switch the order of the section
titled "celerite as a model of stellar variations" and the section describing
the algorithm. "Stellar variations" is now Section 4 and the implementation is
discussed in Section 5.

Below we address each of the referee's comments in detail:

## Major Issues

> Unclear model specification: A Bayesian model is composed of the likelihood
> function and joint prior distribution of the unknown parameters. The likelihood
> function of the unknown parameters, L(θ,α), for the proposed GP method is
> implicitly set to Equation (3). First of all, the components of the two
> unknown vectors, θ and α, are not the same for all of the examples, and thus
> the authors must define these clearly in each example. For example, two vectors
> are θ = (?, ?, ?) and α = (?, ?, ?) so that the likelihood function is uniquely
> determined. Next, the authors must clearly specify the joint prior distribution
> p(θ,α). If the authors use a Uniform prior distribution, please clearly specify
> the range of the Uniform distribution because there are uncountably many
> Uniform prior distributions depending on the range. Once the likelihood
> function and joint prior distribution are clearly specified, then the joint
> posterior distribution, p(θ, α | X, y), is uniquely determined and I cannot say
> that ‘your model specification is unclear.’ Please clearify your Bayesian
> models in all the numerical illustrations including the two simulations and
> three realistic data analyses. It may be possible that the full posterior
> distribution may contain unknown parameters more than θ and α in their
> numerical illustrations, e.g., nuisance parameters in Section 7.1. Again,
> please follow the principle rule by specifying the likelihood function of all
> the unknown parameters and their prior distributions. Please do not use any
> MCMC method before clearly specifying the full posterior distribution.
> Otherwise users do not know what the distribution of the posterior sample
> represents.

We have updated each example to include a list of each parameter and the chosen
priors. We also include a table for each example that lists this information.

> Posterior propriety: Posterior propriety means that your full posterior
> distribution has a finite integral,
> [...]
> This is the basis of all the Bayesian analyses because the fundamental Bayes
> rule holds if and only if posterior propriety holds. Treating X and y as the
> observed data, we can derive the Bayes
> rule as follows.
> [...]
> Without posterior propriety, the joint posterior distribution p(θ,α | X,y) is
> not guaranteed to be a probability distribution because the denominator in (2)
> blows up. Also, without posterior propriety, there is no way to know whether a
> Markov chain converges to the correct target (stationary) posterior
> distribution, p(θ, α | X, y). This is because a Markov chain is null recurrent
> for an improper posterior distribution and thus the stationary distribution of
> a Markov chain does not exist (see the article in footnote 1 for details). If
> the authors use any improper prior distributions, e.g., for a parameter θi,
> log(θi) ∼ Uniform(−∞, ∞) or θi ∼ Uniform(0, ∞), then the authors must mention
> whether posterior propriety holds before any MCMC method is used. (The authors
> may say that their MCMC method produces reasonable posterior samples of the
> unknown parameters, recovering their true values correctly. This is not about
> whether the MCMC method recovers the true values or not. This is about whether
> the Markov chain converges to the targeted stationary distribution. Your target
> posterior distribution may or may not contain the true values and recovering
> true values does not necessarily mean that your chain converges to the targeted
> posterior distribution. Converging to the stationary distribution and
> recovering true values are two different problems.) A well-known insidious
> property of posterior propriety is that the MCMC sampler may appear to work
> well even when the posterior distribution is improper, and thus practitioners
> do their sciences based on non-existent posterior distributions. There are only
> two ways to guarantee posterior propriety: (i) Using a jointly proper prior
> distribution, p(θ,α) dθdα < ∞, guarantees posterior propriety, which will
> become clear when the authors clarify their Bayesian models. (ii) Showing that
> (1) holds in a mathematically rigorous way.

In each example, we were already using (unspecified!) proper priors. We have
updated each example to include a discussion of the priors and these are fully
specified in the tables included with each example.

> Model checking: The authors provide a method for a model fitting but I do not
> know why I trust their estimates or posterior distributions. In each example,
> there should at least be graphical model checking, e.g., by plotting residuals
> with respect to an optimal model. This is one way to conduct posterior
> predictive check, i.e., generating simulated data using the proposed model
> given the posterior samples of the parameters and visually checking whether
> these simulated data match well with the observed data or whether the residuals
> do not have any patterns. Successful model checking could support an argument
> that, at least, the present model is not significantly misspecified, so that
> the uncertainty estimates can be trusted.

We have updated the figures in each example to include plots of the residuals
between the data and the maximum a posteriori model.

> Unknown data size: In Sections 6 and 7, the authors never specify the data size
> in each example. It is quite shocking to me that Section 7.1 even does not
> display any data at all. Section 6 does not seem to motivate the argued
> ‘scalability’ because the data size appears to be small in Figures 6 and 7. Are
> the authors motivated by a realistic problem to develop a scalable GP?  Why do
> the authors need a scalable GP? Do we really need the argued ‘scalability’ to
> fit the data used in each example? Are there any examples that motivate this
> research on the scalability of the GP with the specific covariance function?
> Throughout the manuscript, the authors argue that a conventional GP is not
> suitable for large data sets. Then please show me or readers a realistic
> example (evidence) where a conventional GP is almost impossible to implement
> while celerite works well. I feel that a conventional GP with the same
> covariance functions (but based on a matrix inversion, i.e., without the trick
> for the scalability) may work well at least for the simulation studies; I
> cannot judge whether my feeling is correct for the realistic examples because
> the authors do not clarify the data sizes. If my feeling is wrong, then please
> show me evidence.

We agree that this discussion was severely lacking from the submitted
manuscript and we have included a discussion of the need for scalable methods
in the introduction, in each example, and in a summary subsection at the end of
Section 6. We now fully specify the data sizes for each example and discuss the
practicality of applying general direct solvers in each case.

> Tightening descriptions is needed.
> - Abstract: “Gaussian Processes are a popular class of models used for this
>   purpose but, since the computational cost scales as the cube of the number of
>   data points, their application has been limited to relatively small
>   datasets.” Is it true for all GPs? This sentence gives me an impression that
>   all GPs scale as the cube of the number of data points. If I use a
>   Kalman-filtering-based GP, then what can the authors say? Please tighten the
>   sentence. The authors are posing a problem of GPs that involve a (large)
>   matrix inversion.
>
> - Abstract “but we also demonstrate that it is effective in many other cases.”,
>   Section 2 “improving this scaling in many circumstances” and “many
>   astronomical data analysis problems”: The authors provide three realistic
>   examples and in Section 9 (Summary) the authors say that the method is
>   expected to work well in many other situations. Expectations are
>   expectations. I am not sure whether the authors can use the word ‘many’ such
>   often based on three realistic examples. Do these three examples represent
>   various astronomical data analyses listed in Section 9?
>
> - Abstract: “flexible, fast, and most importantly, interpretable” The
>   covariance function for celerite is restrictive as the authors admit in
>   Section 8. I am not sure whether the word ‘flexible’ describes celerite in an
>   objective way.
>
> - Section 1: “However, there is no further constraint on the data or the
>   model.” This sentence is wrong because the authors mention that the
>   covariance function must be stationary on page 4.
>
> - Section 1: “As we will discuss in the following pages, this choice of kernel
>   function is not as restrictive as it first seems because it arises naturally
>   in some physical systems and it can be used as an effective model in many
>   other cases.” This sentence is vacuous due to the words, ‘seems’, ‘some’, and
>   ‘many’. As for the word ‘seems’, the covariance function that celerite allows
>   is restrictive at my first, second, and third glances and the authors also
>   admit it in Section 8. As for the word ‘some’, the three examples handles
>   three physical systems. As for the word ‘many’, in Section 9 (Summary) the
>   authors say that the method is expected to work well in many other
>   situations. But I keep feeling that this kind of argument, ‘in many other
>   cases’, based only on their expectations is appropriate.
>
> - Section 1: “GP models have been widely used across the physical sciences but
>   their application is generally limited to small datasets” How small is small
>   to the authors? A light curve with 999 data points? Please clarify the cases
>   where the authors think the data size is small so that the readers can judge
>   how small is small. Do the authors use a ‘large’ data set in the examples to
>   show that a typical GP with a matrix inversion never works while celerite
>   successfully works?

For each of the above, we agree and have address each concern.

> - Section 6: “Even in these cases, GPs can be useful
>   effective models and celerite provides computational advantages over other GP
>   methods.” This may imply that celerite has a computational advantage over any
>   GP methods. The authors mention in Section 8 that a Kalman-filtering is in
>   practice somewhat faster and the memory requirements are smaller. So the
>   sentence is not correct.

The current method is now faster than Kalman-filtering methods so this
statement is now true.

> - Section 7.1: “initialize our celerite model using
>   a grid search in the parameter space” and “This model requires about a minute
>   of computation time to find the maximum likelihood parameters and then it
>   takes about an hour to run the MCMC sampling to convergence on a dual-CPU
>   laptop.”: For the first sentence, I have never read in the previous sections
>   about how to initialize the model using a grid search in the parameter space.
>   What do the authors actually do? Why do the authors do the grid search? What
>   is the dimension of the parameter space and the grid? The second sentence is
>   also surprising because the method description in the previous sections never
>   says that the implementation procedure is actually a two-step procedure. I
>   guess that the authors use MLEs as initial values of their MCMC method.
>   Clearly specify the entire procedure including how to set up initial values
>   of the celerite model parameters for an MCMC in the section that describes
>   the method implementation. In the numerical illustrations, the authors are
>   not supposed to say anything new or unexplained about the implementation.

We have updated the discussion of each example to include a description of the
initialization procedure.

> - Section 7.1: “another GP solver” This phrase is broad. Which GP solver do the
>   authors indicate? Please be specific.

Fixed.

## Minor concerns

> Abstract: “a comparison to existing scalable Gaussian Process methods.” The
> authors clearly mention that any CARMA(p, q) model can be matched to their
> celerite model in Section 4.2. Then why not numerically compare the scaleable
> method of fitting CARMA(p, q) with celerite?

We have added an explicit numerical comparison between Kalman-filtering based
CARMA models and celerite. Figure 13 has been added to show the computational
scaling of the Kelly et al. (2014) solver and find that it is about a factor of
10 slower than celerite.

> Section 2: The authors say that the log likelihood function is Equation (3).
> However, Equation (3) is clearly a function of y given X, θ, and α.

We have fixed the sloppy notation by defining this function as L(θ,α).

> Until Section 4, I have thought that the authors may be using a
> Kalman-filtering for a specific form of a covariance matrix. Maybe it is my
> fault. When a scalable GP with O(n) is mentioned, the first thing I think about
> is a Kalman-filtering. What about mentioning that the proposed method is not a
> Kalman-filtering but is also scalable with O(n) in the early stage of the
> manuscript.

Good point. We have added this clarification to the introduction.

> Section 2: ‘using a non-linear optimization routine’ I am just curious about
> how Nocedal and Wright (2006) obtain the uncertainty of their point estimate
> because I believe that point estimation can be misleading without uncertainty
> quantification.

We absolutely agree! We have emphasized the fact that this estimate would only
be a point estimate.

> Section 2: Markov Chain Monte Carlo → Markov chain Monte Carlo. The word
> ‘chain’ is not a name of a person.

Fixed.

> The parameter J is an obvious tuning parameter of the proposed method. I was
> surprised when I read ‘but, in many cases, small values of J are sufficient” in
> Section 9. Maybe the authors think it is ok in an astronomical journal to give
> users a broad and unclear guideline on how to set the tuning parameter. This
> issue would be a major issue if it were for a publication in a statistical
> journal. Please keep in mind that without setting J no one can implement
> celerite. Imagine a user who wants to use celerite but do not know how to set
> J. Are the authors going to say ‘use a small value of J’? The current sentence
> sounds very irresponsible. This sentence gives me an impression that J = 1 must
> work well ‘in many cases,’ although I am not sure how many is many. The most
> important thing is to specify what made the authors believe that the small
> values of J would work well in many cases so that users can judge whether the
> small values of J are also appropriate for their sciences. Please show the
> readers the evidence for the argument or a clear procedure to determine J so
> that they can follow it for their data analyses. If the authors do not have any
> evidence or a procedure, i.e., the choices of J in the examples of the current
> manuscript are all ad-hoc, please do more research on it. Please do not use the
> word ‘small’ because this word means nothing to users. Also please specify the
> tuning procedure in the section regarding the implementation because without
> setting J no one can implement celerite. Finally, please clarify how the
> authors choose J in each of two simulation studies and three realistic data
> analyses, which will be helpful for users.

We respectfully disagree with the referee's interpretation of J as a "tuning
parameter" of the method and instead argue that this is just the general
problem of model selection when using GPs. Every GP analysis must specify a
kernel and celerite is no different. That being said, we agree with the
concerns about the vague statement about "small J" so we have removed this and
added a section (5.4) discussing options for model selection with some
celerite-specific recommendations.  The models in all examples except Example
2 are chosen using physical motivation and we have updated Example 2 to
include a discussion and a new figure to explain the procedure that we used to
select the model in that case.

> Section 3: The mixture idea shown in Equation (6) also appears in the
> literature. The discussion about this idea and the one in Ambikasaran (2015)
> may be interesting but can be distracting. Please give enough credit for the
> earlier work, if any.

Added.

> Section 3: Ornstein-Uhlenbeck process may need a citation.

Done.

> After Equation (19) in Section 4.1: There are not b and d in Equation (7).
> Do the authors mean bj and dj?

This section has been removed and replaced by our new derivation.

> Section 5: “This suggests that the Matern-3/2 covariance can be approximated
> using the celerite framework with a small value of f in Equation (54) but we
> caution that this might lead to numerical issues for the solver.” In the
> previous description, there is no approximation at all. Where does the
> approximation occur in the previous description? This sentence also sounds
> problematic. It indicates that some models definitely within the celerite
> framework may not be feasible because of numerical issues. Then the authors
> must specify feasible class of models among all the possible models within the
> celerite framework in order not to disappoint users.

We have expanded our discussion in this section and made the approximation
explicit. The numerical issue comes from the fact that the parameter b_j is
infinite for Q = 1/2. We now state this explicitly in the manuscript.

> Section 6.2: This is an impressive result. However, at the same time, I feel
> that this is a dangerous result because the proposed model does not reveal any
> evidence on a potential model mis-specification.

As with all the examples, we have added a graphical model check. As we discuss
in the text, any broad use of this method would require more research to
determine the detailed limits of applicability, but we think that that is
beyond the scope of this paper.

> Section 7.1: “is is” is a typo.

Fixed.

> Section 8: Comparison to a 20 year old method that the current method is
> exactly built upon? Without reading further, I can imagine that the proposed
> method is better.

We have removed this comparison.

> Section 8: Comparison to CARMA. The authors say that any CARMA model has a
> corresponding celerite model. This means that I can fit any CARMA model via
> Kalman-filtering, obtain posterior samples of the unknown parameters of the
> CARMA model, match these posterior samples of the CARMA parame- ters with
> corresponding celerite parameters, and take advantage of the better physical
> interpretability of the celerite parameters. Why should the readers use
> celerite at the beginning instead of using Kalman-filtering in fitting a CARMA
> model? I am not quite motivated at this point.

We have expanded our discussion in this section to provide further motivation.

> Section 9: The authors list all the possible cases where celerite may work
> well. I think this section (summary) is full of expectations (too many
> expectations for a scientific journal in my opinion). I want to understand why
> the authors expect celerite to work well in such many cases. If the authors
> have clear evidence for these cases, then what about making another section for
> potential applications? In my opinion if this section is for ‘Summary’, then it
> is not a good place to talk about things that have not explained in the
> previous sections.

We agree and we have changed the section title to "Discussion". We have removed
some speculation from this section and added a subsection where we explicitly
discuss some potential future applications.

### 2017-06-12 paper/ms.tex
adding response

@@
 with a small number of operations for each data point.
-Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
-mixtures of exponentials
+\response{This model has been generalized to arbitrary mixtures of
+exponentials \citep[for example,][]{Kelly:2011}
 \begin{eqnarray}
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
-    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
-\end{eqnarray}
-\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
-kernel is semiseparable.}
+    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm}) \quad.
+\end{eqnarray}
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
 evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
-components in the mixture.
-\response{In \sect{factor}, we derive a novel method with the same
-$\mathcal{O}(N\,J^2)$ scaling that can be used when
-$k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
+components in the mixture \citep{Kelly:2014,Ambikasaran:2015}.
+In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} derive a
+faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
+when $k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
 
@@
 a_j &=& S_0\,\omega_0\,Q \\
-b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
-c_j &=& \frac{\omega_0}{2\,Q}\\
-d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
+b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \nonumber\\
+c_j &=& \frac{\omega_0}{2\,Q} \nonumber\\
+d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad, \nonumber
 \end{eqnarray}
@@
 \citep{Rasmussen:2006}.
-This suggests that the Mat\'ern-3/2 covariance can be approximated using
-the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
-caution that this might lead to numerical issues for the solver.
-}
+\response{This model cannot be directly evaluated using \celerite\ because,
+as we can see from \eq{sho-complex}, the parameter $b_j$ is infinite for $Q =
+1/2$.
+The limit in \eq{approx-matern2}, however, suggests that an approximate
+Mat\'ern-3/2 covariance can be computed with \celerite\ by using a small value
+of $f$ in \eq{approx-matern2}.
+The required value of $f$ will depend on the specific dataset and the precision
+requirements for the inference, so we encourage users to test this
+approximation in the context of their research before using it.
+}}
 

### 2017-06-12 paper/ms.tex
author order

@@
 
+\edef \iis {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\iis}{Department of Computational and Data Sciences,
+                    Indian Institute of Science, Bangalore, India}
+
 \edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
@@
 
-\edef \iis {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\iis}{Department of Computational and Data Sciences,
-                    Indian Institute of Science, Bangalore, India}
-
 \author{%
@@
     Eric~Agol\altaffilmark{\uw,\guggenheim},
-    Ruth~Angus\altaffilmark{\simons,\columbia}, and
-    Sivaram~Ambikasaran\altaffilmark{\iis}
+    Sivaram~Ambikasaran\altaffilmark{\iis}, and
+    Ruth~Angus\altaffilmark{\simons,\columbia}
 }

### 2017-06-12 paper/ms.tex
edits to appendix

@@
 component of its argument and $\mathrm{triu}$ is the equivalent strict upper
-triangular operation.  An example of this type of matrix is given in appendix \ref{operations}.
+triangular operation.
 In this representation, rather than storing the full $N \times N$ matrix $K$,
@@
 \subsection{Cholesky factorization of positive definite semiseparable matrices}
+\sectlabel{cholesky}
 
@@
 
-\section{Semiseparable operations} \label{operations}
-
-To help clarify equations \eq{system}, we write out here an example of a semiseparable matrix with a
-single \celeriteterm\ term with $c=d=1$:
-\begin{eqnarray}\eqlabel{semi-separable-structure}
-K&=&%
-\begin{pmatrix}
-    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
-    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
-    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
-    \ldots & \ldots & \ldots & \ldots
-\end{pmatrix}\\
-&=&
-\begin{pmatrix}
-    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
-    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
-    \ldots & \ldots & \ldots & \ldots
-\end{pmatrix}\\
-&+&\mathrm{tril}\left[\begin{pmatrix}
-    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
-    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
-    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
-   \ldots & \ldots
-\end{pmatrix}
-\times
-\begin{pmatrix}
- {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
- {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
- \ldots & \ldots
-\end{pmatrix}^T\right]\\
-&+&
-\mathrm{triu}\left[\begin{pmatrix}
- {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
- {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
- {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
- \ldots & \ldots
-\end{pmatrix}
-\times
-\begin{pmatrix}
-    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
-    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
-    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
-    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
-    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
-    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
-   \ldots & \ldots
-\end{pmatrix}^T\right]\\
-&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
-\end{eqnarray}
-
-In addition to solving $K^{-1} {\bf y}$, several other operations can be
-carried out recursively in $\mathcal{O}(N)$, simulation, multiplication,
-and prediction, which we describe in the following sections.
+\response{
+\section{Semiseparable matrix operations} \sectlabel{operations}
+
+In this appendix, we summarize some algorithms for manipulating semiseparable
+matrices and discuss how these can be used with GP models and \celerite.
+
+\subsection{Multiplication and filtering}
+
+The product of a rank-$R$ semiseparable matrix with general vectors and
+matrices can be computed in $\mathcal{O}(N\,R)$ operations
+\citep{Vandebril:2007}.
+In the context of \celeriteterm\ models, na\"ive application of these methods
+will result in numerical overflow and underflow issues much like we found with
+the Cholesky factorization.
+To avoid these errors, we derive the following numerically stable algorithm
+for computing the dot product of a \celeriteterm\ covariance matrix
+\begin{eqnarray}
+\bvec{y} = K\,\bvec{z}
+\end{eqnarray}
+using the semiseparable representation for $K$ defined in \eqalt{reparam}:
+\begin{eqnarray}
+f^+_{n,j} &=& \phi_{n+1,j}\,\left(f^+_{n+1,j} + \tilde{U}_{n,j}\,z_{n+1}\right) \\
+f^-_{n,j} &=& \phi_{n,j}\,\left(f^-_{n-1,j} + \tilde{V}_{n-1,j}\,z_{n-1}\right) \\
+y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
+    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
+\end{eqnarray}
+where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
+This algorithm requires two sweeps~--~$n=1, \ldots, N$ to compute $f^-$ and
+$n=N,\ldots,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.
 
@@
 and multiplication are used.
-
-\subsection{Multiplication and filtering}
-
-Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(RN)$
-operations with a stable recursive algorithm using the semiseparable form:
-\begin{equation}
-\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
-\end{equation}
-The algorithm for multiplication is given by
-\begin{eqnarray}
-f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
-f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
-y_n &=& A_{n,n} z_n + \sum_{j=1}^R \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
-\end{eqnarray}
-where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,R$.  The recursion requires
-two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
-from $n=2$ to $n=N$ to compute $f^-_{n,j}$.
-
-Multiplication of the full covariance matrix is useful, for example, for isolating
-a particular component of a kernel, which amounts to a customized filter of a dataset.
-For example, a white-noise filter is given by:
-\begin{eqnarray}
-\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
-&=& K(\bvec{t},\bvec{t}) \bvec{z}
-\end{eqnarray}
-where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
-decomposition as described above.  The second step multiplication of the white-noise-free
-kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(RN)$ steps with this multiplication algorithm.
-Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
-but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.
 
@@
 %interspersed.
+}
 

### 2017-06-12 paper/ms.tex
fixed typo

--- 2017-06-12 paper/ms.tex
@@
 with a small number of operations for each data point.
-\response{This model has been generalized to arbitrary mixtures of
-exponentials \citep[for example,][]{Kelly:2011}
+Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
+mixtures of exponentials
 \begin{eqnarray}
 k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
-    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm}) \quad.
-\end{eqnarray}
+    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})
+\end{eqnarray}
+\response{by recognizing that the matrix $K_\bvec{\alpha}$ generated by this
+kernel is semiseparable.}
 In this case, the inverse is dense but \eq{gp-likelihood} can still be
 evaluated with a scaling of $\mathcal{O}(N\,J^2)$, where $J$ is the number of
-components in the mixture \citep{Kelly:2014,Ambikasaran:2015}.
-In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} derive a
-faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
-when $k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
+components in the mixture.
+\response{In \sect{factor}, we derive a novel method with the same
+$\mathcal{O}(N\,J^2)$ scaling that can be used when
+$k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
 
@@
 a_j &=& S_0\,\omega_0\,Q \\
-b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \nonumber\\
-c_j &=& \frac{\omega_0}{2\,Q} \nonumber\\
-d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad, \nonumber
+b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
+c_j &=& \frac{\omega_0}{2\,Q}\\
+d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
 \end{eqnarray}
@@
 \citep{Rasmussen:2006}.
-\response{This model cannot be directly evaluated using \celerite\ because,
-as we can see from \eq{sho-complex}, the parameter $b_j$ is infinite for $Q =
-1/2$.
-The limit in \eq{approx-matern2}, however, suggests that an approximate
-Mat\'ern-3/2 covariance can be computed with \celerite\ by using a small value
-of $f$ in \eq{approx-matern2}.
-The required value of $f$ will depend on the specific dataset and the precision
-requirements for the inference, so we encourage users to test this
-approximation in the context of their research before using it.
-}}
+This suggests that the Mat\'ern-3/2 covariance can be approximated using
+the \celeriteterm\ framework with a small value of $f$ in \eq{approx-matern2} but we
+caution that this might lead to numerical issues for the solver.
+}
 
@@
 These criteria are easy to implement and we use the BIC in \sect{example2}
-but, since are only formally valid under specific sets of assumptions, we
+but, since these are only formally valid under specific sets of assumptions, we
 cannot recommend their use in general.

### 2017-06-12 paper/ms.tex
nothing really

--- 2017-06-12 paper/ms.tex
@@
 where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
-This algorithm requires two sweeps~--~$n=1, \ldots, N$ to compute $f^-$ and
-$n=N,\ldots,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.
+This algorithm requires two sweeps~--~$n=2,\,\ldots,\,N$ to compute $f^-$ and
+$n=N-1,\,\ldots,\,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.
 

### 2017-06-12 paper/ms.tex
more edits to appendix

@@
 
-\subsection{Scalable computation of celerite models}
+\subsection{Scalable computation of celerite models}\sectlabel{scalable}
 
@@
 algorithm to compute the Cholesky factorization of a \celeriteterm\ model
-\begin{eqnarray}
+\begin{eqnarray}\eqlabel{cholesky-factor}
     S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
@@
 
-\newpage
+\vspace{5ex}
 \appendix
@@
 
-\subsection{Multiplication and filtering}
+\subsection{Multiplication}
 
@@
 \begin{eqnarray}
-f^+_{n,j} &=& \phi_{n+1,j}\,\left(f^+_{n+1,j} + \tilde{U}_{n,j}\,z_{n+1}\right) \\
-f^-_{n,j} &=& \phi_{n,j}\,\left(f^-_{n-1,j} + \tilde{V}_{n-1,j}\,z_{n-1}\right) \\
+f^+_{n,j} &=& \phi_{n+1,j}\,\left[f^+_{n+1,j} +
+    \tilde{U}_{n,j}\,z_{n+1}\right] \\
+f^-_{n,j} &=& \phi_{n,j}\,\left[f^-_{n-1,j} +
+    \tilde{V}_{n-1,j}\,z_{n-1}\right] \\
 y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
@@
 
-\subsection{Simulation}
-
-Simulation of a Gaussian process for large $N$ requires multiplying an
-$N$-dimensional vector of Gaussian normal deviates, $\bvec{q}$, by the square root
-of the covariance matrix:
-\begin{equation}
-\bvec{y} = K_\bvec{\alpha}^{1/2} \bvec{q},
+\subsection{Sampling data from a celerite process}
+
+To sample a dataset $\bvec{y}$ from a GP model for fixed parameters
+$\bvec{\theta}$ and $\bvec{\alpha}$, we must compute
+\begin{equation} \eqlabel{sim}
+\bvec{y} = \bvec{\mu}_\bvec{\theta}(\bvec{t})+
+    K_\bvec{\alpha}^{1/2}\,\bvec{q}
 \end{equation}
-where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
-comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
-where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the first term of $\mathcal{Q} = -2 \mathcal{L}$
-with $\mu_\bvec{\theta} =0$ is given by
-\begin{eqnarray}
-\mathcal{Q} &=&\bvec{y}^T K_\bvec{\alpha}^{-1} \bvec{y}\\
-&=& \bvec{y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} \bvec{y}\\
-&=& \bvec{q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} \bvec{q}\\
-&=& \bvec{q}^T \bvec{q}.
-\end{eqnarray}
-The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
-which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $\mathcal{Q} \sim \chi_N^2$; for
-diagonal $K_\bvec{\alpha}$, $\mathcal{Q}$ becomes the $\chi^2$ statistic.
-
-To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
-low-rank Cholesky decomposition first, generating the vector $\bvec{q}$ of random normal deviates, and then multiplying
-\begin{equation}
-\bvec{y} = L D^{1/2} \bvec{q}.
-\end{equation}
-This may be carried out in $\mathcal{O}(JN)$ operations with
-\begin{eqnarray}
-f_{n,j} &=& \phi_{n,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
-y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2\,J} \tilde{U}_{n,k} f_{n,k}
-\end{eqnarray}
-for $n = 1$ to $N$ and $j=1$ to $2\,J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
-We expect that this simulation algorithm will be useful for test-driving Gaussian
-process inference algorithms applied to large datasets as even simulation of
-Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition
-and multiplication are used.
-
-\subsection{Prediction and interpolation}
-
-%Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.
-With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $\mathcal{O}(N+M)$ operations.
-
-For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t^*} = \left\{t_{1}^*,t_{2}^*,...,t_{M}^*\right\}$, we need to apply the covariance between the measured and predicted times.
-The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
-\begin{eqnarray}
-\bar f^* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
-&=& K(\bvec{t^*},\bvec{t}) \bvec{z}
-\end{eqnarray}
-where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
-$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
-the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.
-
-Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
-\begin{eqnarray}
-\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
-&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
-\end{eqnarray}
-for $m =1$ to $M$.
+where $\bvec{\mu}_\bvec{\theta}(\bvec{t})$ is the mean function evaluated at
+the input coordinates $\bvec{t}$, $\bvec{q}$ is a vector of draws from the
+unit normal $q_i \sim \mathcal{N}(0,\,1)$, and
+\begin{eqnarray}
+K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \,
+    {\left(K_\bvec{\alpha}^{1/2}\right)}^T\quad.
+\end{eqnarray}
+For a general matrix $K_\bvec{\alpha}$, the cost of this operation scales as
+$\mathcal{O}(N^3)$ to compute $K_\bvec{\alpha}^{1/2}$ and $\mathcal{O}(N^2)$
+to compute the dot product.
+For \celeriteterm\ models, we use the semiseparable representation of the
+Cholesky factor that we derived in \sect{cholesky} to compute the dot product
+\eq{sim} in $\mathcal{O}(N\,J)$.
+Using the notation from \sect{scalable}, the algorithm to compute this dot
+product is
+\begin{eqnarray}
+f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} + \tilde{W}_{n-1,j}\,
+    \sqrt{D_{n-1,n-1}}\,q_{n-1}\right]\\
+y_n &=& \sqrt{D_{n,n}}\,q_n + \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
+\end{eqnarray}
+where $f_{1,j} = 0$ for all $j$.
+This scalable algorithm is useful for generating large simulated
+datasets for experiments with \celerite\ and for performing posterior
+predictive checks.
+
+\subsection{Interpolation \& extrapolation}
+
+The predictive distribution of a GP at a vector of $M$ input coordinates
+\begin{eqnarray}
+\bvec{y}^* = \left(\begin{array}{ccccc}
+    t_1^*\quad && \cdots\quad && t_M^*
+\end{array}\right)^\T
+\end{eqnarray}
+conditioned on a dataset $(\bvec{t},\,\bvec{y})$ and GP
+parameters $(\bvec{\theta},\,\bvec{\alpha})$, is a normal $\bvec{y}^* \sim
+\mathcal{N}(\bvec{\mu}^*,\,K^*)$ with mean
+\begin{eqnarray}\eqlabel{pred-mean}
+\bvec{\mu}^* &=& \bvec{\mu}_\bvec{\theta}(\bvec{t}^*) +
+    K(\bvec{t}^*,\,\bvec{t})\,{K(\bvec{t},\,\bvec{t})}^{-1}\,\left[\bvec{y} -
+    \bvec{\mu}_\bvec{\theta}(\bvec{t})\right]
+\end{eqnarray}
+and covariance
+\begin{eqnarray}
+K^* &=& K(\bvec{t}^*,\,\bvec{t}^*) -
+    K(\bvec{t}^*,\,\bvec{t})\,{K(\bvec{t},\,\bvec{t})}^{-1}\,
+    K(\bvec{t},\,\bvec{t}^*)
+\end{eqnarray}
+where $K(\bvec{v},\,\bvec{w})$ is the covariance matrix computed between
+\bvec{v} and \bvec{w} \citep{Rasmussen:2006}.
+
+Na\"ively, the computational cost of computing \eq{pred-mean} for a
+\celeriteterm\ model scales as $\mathcal{O}(N\,M)$ if we reuse the Cholesky
+factor derived in \sect{cholesky}.
+It is, however, possible to improve this scaling to $\mathcal{O}(n\,N +
+m\,M)$ where $n$ and $m$ are integer constants.
+To derive this, we expand \eq{pred-mean} using \eq{celerite-kernel}
+\begin{eqnarray}
+\mu^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
+    &=& \sum_{n=1}^N \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\,\left[
+        a_j\,\cos{(d_j\,\vert t^*_m-t_n\vert)} +
+        b_j\,\sin{(d_j\,\vert t^*_m-t_n\vert)}
+    \right] \,z_{n} \quad. \nonumber
+\end{eqnarray}
+where
+\begin{eqnarray}
+\bvec{z} = {K(\bvec{t},\,\bvec{t})}^{-1}\,\left[\bvec{y} -
+    \bvec{\mu}_\bvec{\theta}(\bvec{t})\right] \quad.
+\end{eqnarray}
 Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
@@
 \begin{eqnarray}
-\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
-&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
-\end{eqnarray}
-We can compute this recursively.  Defining
-\begin{eqnarray}
-Q^-_{n,i} &=& \left(Q^-_{n-1,i}+z_n\tilde V_{n,i}\right)e^{-c_j(t_{n+1}-t_{n})},\\
-X^-_{m,n,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i}^*,\\
-Q^+_{n,i} &=& \left(Q^+_{n+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_{n}-t_{n-1})},\\
-X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i}^*,
-\end{eqnarray}
-where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$
-to $2\,J$, and in $X^\pm_{m,n,i}$ the expressions
-for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluted at $t_m^*$.
-Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
-$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
-For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
-point the predicted mean can be computed from
+\mu^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0}
+    e^{-c_j\,(t^*_m-t_n)}\,\left[a_j\,\cos{(d_j\,(t^*_m-t_n))}
+        +b_j\,\sin{(d_j\,(t^*_m-t_n))}\right]\,z_n \\
+&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j\,(t_n-t^*_m)}\,\left[
+    a_j\,\cos{(d_j\,(t_n-t^*_m))}+b_j\,\sin{(d_j\,(t_n-t^*_m))}\right]\,z_n
+\quad. \nonumber
+\end{eqnarray}
+We compute this using two passes, a forward pass $n_0 = 1,\,\ldots,\,N$ and
+a backward pass $n_0 = N,\,\ldots,\,1$.
+Defining
+\begin{eqnarray}
+Q^-_{n,k} &=& \left[Q^-_{n-1,k}+z_n\,\tilde V_{n,k}\right]\,e^{-c_{k//2}\,(t_{n+1}-t_{n})},\\
+X^-_{m,n,k} &=& e^{-c_{k//2}\,(t_m^*-t_{n+1})}\,\tilde U_{m,k}^*,\\
+Q^+_{n,k} &=& \left[Q^+_{n+1,k}+z_n\,\tilde U_{n,k}\right]\,e^{-c_{k//2}(t_{n}-t_{n-1})},\\
+X^+_{m,n,k} &=& e^{-c_{k//2}\,(t_{n-1}-t_m^*)}\,\tilde V_{m,k}^* \quad,
+\end{eqnarray}
+where $t_0=t_1$, $t_{N+1}=t_N$, $Q^-_{0,k} = 0$ and $Q^+_{N+1,k}=0$ for $k=1$
+to $2\,J$, and in $X^\pm_{m,n,k}$ the expressions
+for $\tilde U_{m,i}^*$ and $\tilde V_{m,i}^*$ are evaluated at $t_m^*$.
+For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is
+reached, at which point the predicted mean can be computed from
 \begin{equation}
-\bar f^*_i = \sum_{i=1}^{2\,J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0+1,i}X^+_{m,n_0+1,i}\right],
+\mu^*_m = \sum_{k=1}^{2\,J}
+    \left[Q^-_{n_0,k}X^-_{m,n_0,k}+Q^+_{n_0+1,k}X^+_{m,n_0+1,k}\right] \quad.
 \end{equation}
-
-%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
-%\begin{eqnarray}
-%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
-%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
-%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
-%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
-%\end{eqnarray}
-%for $1 \le p,q \le J$,
-%where
-%\begin{equation}
-%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
-%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
-%\end{equation}
-%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
-%with the Cholesky decomposition.
-%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
-%\begin{equation}
-%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
-%\end{equation}
-%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
-%Note that this variance formula {\it only} works for prediction; we have
-%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
-%interspersed.
 }
 
-\newpage
 \bibliography{celerite}

### 2017-06-12 paper/ms.tex
a few edits

@@
 \tess\ \citep{Ricker:2014}, \lsst\ \citep{Ivezic:2008}, \project{WFIRST}
-\citep{Spergel:2015}, and PLATO \citep{Rauer:2014} will continue to produce
-similar or larger data volumes.}
+\citep{Spergel:2015}, and \project{PLATO} \citep{Rauer:2014} will continue to
+produce similar or larger data volumes.}
 
@@
 Another benefit of \celerite\ compared to Kalman filter solvers is the fact
-that it is a direct solver for the relevant matrix equations instead of just
-an algorithm to compute \eq{gp-likelihood}.
+that it is an algebraic solver for the relevant matrix equations instead of
+an algorithm to directly compute \eq{gp-likelihood}.
 This means that the inverse covariance matrix $K^{-1}$ can be applied to
@@
 
-%As discussed previously, the \celerite\ method is based upon a method
-%developed by Rybicki \& Press twenty years ago \citep{Rybicki:1995}.
-%In the case where the kernel function has a single term of the form given in
-%\eq{kernel-simple}, it is interesting to compare the performance of \celerite\
-%and the Rybicki \& Press method.
-%Their method does not require solving the extended matrix because the matrix
-%inverse can be calculated analytically, but the method is made slightly more
-%complicated when white noise is included.
-%We implement their full method and find empirically that it is more
-%computationally efficient by a factor of $\approx 2$ for large $N \gg 1000$,
-%but we note that the requirement of a single, real, exponential kernel is
-%extremely restrictive and this model cannot describe quasi-periodic
-%variability.
-
-Another popular method uses the fact that, in the limit of evenly-spaced data
-and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
+Another popular GP method uses the fact that, in the limit of evenly-spaced
+data and homoscedastic uncertainties, the covariance matrix is ``Toeplitz''
 \citep[for example][]{Dillon:2013}.
@@
 The key takeaway here is that \celerite\ provides an \emph{exact} method for
-GP inference for a specific, but flexible, class of one-dimensional kernel
-functions.
+GP inference for a specific class of one-dimensional kernel functions.
 Furthermore, since \celeriteterm\ models can be interpreted as a mixture of
@@
 astronomical data analysis, but the fact that the computational cost generally
-scales as the cube of the number of data points has limited their use to relatively
-small datasets.
-With the linear scaling of \celerite\, we envision the application of Gaussian
-processes may be expanded to much larger datasets.
+scales as the cube of the number of data points has limited their use to
+small datasets with $N \lesssim 1000$.
+With the linear scaling of \celerite\, we envision that the application of
+Gaussian processes will be expanded to the existing and forthcoming large
+astronomical time domain surveys such as \kepler\ \citep{Borucki:2010},
+\project{K2} \citep{Howell:2014}, \tess\ \citep{Ricker:2014}, \lsst\
+\citep{Ivezic:2008}, \project{WFIRST} \citep{Spergel:2015}, and
+\project{PLATO} \citep{Rauer:2014}.
 Despite the restrictive form of the \celeriteterm\ kernel, with a sufficient
@@
 terms of resonant frequency, amplitude, and quality factor.
-A drawback of this method is its quadratic scaling with the number of
-terms $J$, but, in some cases, small values of $J$ may suffice.
 
 Our background is in studying transiting exoplanets, a field which has only
-recently begun to adopt full covariance matrices in analyzing the noise in
-stellar light curves when detecting or characterizing transiting planets
-\citep[for example,][]{Carter:2009, Gibson:2012, Barclay:2015, Evans:2015,
-Aigrain:2016, Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
+recently begun to adopt GP methods for analyzing the noise in stellar light
+curves and radial velocity datasets when detecting or characterizing
+transiting planets \citep[for example,][]{Carter:2009, Gibson:2012,
+Haywood:2014, Barclay:2015, Evans:2015, Rajpaul:2015, Aigrain:2016,
+Foreman-Mackey:2016b, Grunblatt:2016, Luger:2016}.
 All of these analyses have been limited to small datasets or restrictive
-kernel choices.
-\celerite\ weakens these requirements by providing a scalable method for
-computing the likelihood and a physical motivation for the choice of kernel.
-\celerite\ can be used to model stellar oscillations using the relation to the
-mixture of stochastically-driven, damped simple harmonic oscillators, as
-we have demonstrated in this paper.
+kernel choices, but \celerite\ weakens these requirements by providing a
+scalable method for computing the likelihood and a physical motivation for the
+choice of kernel.
 As higher signal-to-noise observations of transiting exoplanet systems are
@@
 correct inference of planetary transit parameters.
-We expect that \celerite\ will be important for transit detection
-\citep{Pope:2016, Foreman-Mackey:2016b}, transit timing \citep{Agol:2005,
-Holman:2005}, transit spectroscopy \citep{Brown:2001}, Doppler beaming
-\citep{Loeb:2003, Zucker:2007}, tidal distortion \citep{Zucker:2007}, phase
-functions \citep{Knutson:2007, Zucker:2007}, and more.
+We predict that \celerite\ will be important for scaling methods of transit
+detection \citep{Pope:2016, Foreman-Mackey:2016b}, transit timing
+\citep{Agol:2005, Holman:2005}, transit spectroscopy \citep{Brown:2001},
+Doppler beaming \citep{Loeb:2003, Zucker:2007}, tidal distortion
+\citep{Zucker:2007}, and phase functions \citep{Knutson:2007, Zucker:2007} to
+the large astronomical time domain surveys of the future.
 
@@
 quantification of its uncertainties.
-In principle, a large number of \celeriteterm\ terms could be used to perform
-non-parametric probabilistic inference of the power spectrum despite
-unevenly-spaced data with heteroscedastic noise \citep[for
-example,][]{Wilson:2013, Kelly:2014}.
+A mixture of \celeriteterm\ terms can be used to perform non-parametric
+probabilistic inference of the power spectrum despite unevenly-spaced data
+with heteroscedastic noise \citep[see][for examples of this
+procedure]{Wilson:2013, Kelly:2014}.
 This type of analysis will be limited by the quadratic scaling of \celerite\
 with the number of terms $J$, but this limits existing methods as well
-\citep[CARMA models,][]{Kelly:2014}.
-In general, we encourage the use of physically motivated models for parameter
-estimation instead of qualitative modeling of the power spectrum itself.
-
-There are many data analysis problems where \celerite\ will not be applicable.
+\citep{Kelly:2014}.
+
+There are many data analysis problems where \celerite\ will not be immediately
+applicable.
 In particular, the restriction to one-dimensional problems is significant.
@@
 include][]{Haywood:2014, Rajpaul:2015, Aigrain:2016}, where \celerite\ cannot
-be used to accelerate any of these analyses.
-It is plausible that an extension could be derived to tackle some
+be used to speed up the analysis.
+It is plausible that an extension can be derived to tackle some
 multidimensional problems with the right structure~--~simultaneous parallel
@@
 In the context of \celeriteterm\ models, na\"ive application of these methods
-will result in numerical overflow and underflow issues much like we found with
-the Cholesky factorization.
+will result in numerical overflow and underflow issues, much like we found
+with the Cholesky factorization.
 To avoid these errors, we derive the following numerically stable algorithm

### 2017-06-14 paper/ms.tex
section 5 is new
missing caption

@@
 
+\response{
 \section{Semiseparable matrices \& celerite}\sectlabel{factor}
@@
     Cholesky factorization routine implemented in the Intel MKL linear algebra
-    package.
+    package is shown as the dashed black line.
     \emph{(right)} The same information plotted as a function of $J$ for
@@
 cannot recommend their use in general.
+}
 

### 2017-07-18 paper/ms.tex
typos and captions

@@
 scalable Gaussian Process methods.
-The method is fast and interpretable, with a wide range of potential
-applications within astronomical data analysis and beyond.
+The method is fast and interpretable, with a range of potential applications
+within astronomical data analysis and beyond.
 We provide well-tested and documented open-source implementations of this
@@
 characterization where GPs have become a model of choice.
-We are confident that the method described in the following pages will benefit
-this field, but we also expect that there will be applications in other
-branches of astrophysics and beyond.
+We are confident that this method will benefit this field, but we also expect
+that there will be applications in other branches of astrophysics and beyond.
 In \sect{examples}, we present applications of the method to research problems
@@
 
-In the following section, we present a method for improving the cubic scaling
-for some use cases.
+In this paper, we present a method for improving the cubic scaling for some
+use cases.
 We call our method and its implementations \celerite.\footnote{The name
@@
 components in the mixture \citep{Kelly:2014,Ambikasaran:2015}.
-In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} derive a
-faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
+In \sect{factor}, we build on previous work \citep{Ambikasaran:2015} to derive
+a faster method with the same $\mathcal{O}(N\,J^2)$ scaling that can be used
 when $k_\bvec{\alpha}(\tau_{nm})$ is positive definite.}
@@
 The details of this method and a few implementation considerations are
-outlined in the following section but we first discuss some
-properties of this covariance function.
+outlined in \sect{factor}, but we first discuss some properties of this
+covariance function.
 
@@
     (middle) The corresponding autocorrelation functions with the same colors.
-    \response{(right) Three realizations from each model in the same colors.}
+    \response{(right) Three realizations in arbitrary units from each model in
+    the same colors.}
     \figurelabel{sho}}
@@
 $N \le 2048$.
-Across the board, these results agree within machine precision and the error
-only weakly increases with $N$ and $J$.
+These results agree to about $10^{-15}$ and the error only weakly increases
+with $N$ and $J$.
 

### 2017-07-18 paper/ms.tex
more typos and refs

@@
 
-\begin{figure}[!htbp]
+\begin{figure}[!hptb]
 \begin{center}
@@
 A detailed discussion of model selection is beyond the scope of this paper and
-the interested reader is directed to Chapter~5 of \citet{Rasmussen:2006} an
-introduction in the context of GPs, but we include a summary and some
+the interested reader is directed to Chapter~5 of \citet{Rasmussen:2006} for
+an introduction in the context of GPs, but we include a summary and some
 suggestions specific to \celerite\ here.
@@
 Despite the fact that it is less flexible than a general \celeriteterm\
-kernel, we recommend the SHO model in most cases because it is 1-time mean
+kernel, we recommend the SHO model in most cases because it is once mean
 square differentiable
@@
 in this idealized case.
-\response{We simulated a small ($N = \exampleindata$) dataset using a GP model
+\response{We simulate a small ($N = \exampleindata$) dataset using a GP model
 with a SHO kernel (\eqalt{sho-kernel}) with parameters $S_0 =
@@
 quasiperiodic covariance function can be used to make probabilistic
-measurements even with sparsely sampled data (R.~Angus, \etal\ submitted).
+measurements even with sparsely sampled data \citep{Angus:2017}.
 The covariance function used for this type of analysis has the form
@@
 A detailed comparison of GP rotation period measurements and the traditional
-methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
-\etal, submitted) demonstrates that GP inferences are, at a population level,
-more reliable than other methods.
+methods is beyond the scope of this paper, but \citet{Angus:2017} demonstrate
+that GP inferences are, at a population level, more reliable than other
+methods.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[hptb]
 \begin{center}

### 2017-07-19 paper/ms.tex
page break

@@
 floating point numbers.
-Given a semiseparable matrix of this form, it is known that matrix-vector
-products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
-operations and the interested reader is directed to \citet{Vandebril:2007} for
-more details about semiseparable matrices in general.
+Given a semiseparable matrix of this form, the matrix-vector products and
+matrix-matrix products can be evaluated in $\mathcal{O}(N)$ operations.
+We include a summary of the key operations in \app{operations}, but the
+interested reader is directed to \citet{Vandebril:2007} for more details about
+semiseparable matrices in general.
 
@@
 
+\newpage
 \section{Examples}\sectlabel{examples}
@@
 
+\newpage
 \subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}
@@
 
+\newpage
 \subsection{Example 5: Exoplanet transit fitting}\sectlabel{transit}
@@
 In this table, we list the computational cost required for a single likelihood
-evaluation using a general Cholesky factorization\footnote{We use the Intel
-Math Kernel Library \url{https://software.intel.com/en-us/intel-mkl}}
-and the cost of computing the same likelihood using \celerite.
+evaluation using a general Cholesky factorization and the cost of computing
+the same likelihood using \celerite.
 Combining this with the total number of model evaluations required to run the

### 2017-07-19 paper/publish.py
different page breaks for pdflatex

@@ -29,11 +29,18 @@ def find_figures(tex):
     return re.findall("includegraphics(?:.*?){(.*)}", tex)
 
 def find_includes(tex):
-    return re.findall("\\include{(.*)}", tex)
+    return re.findall("include{(.*)}", tex)
+
+def find_inputs(tex):
+    return re.findall("input{(.*)}", tex)
 
 for fn in find_includes(tex):
     txt = open(fn + ".tex").read()
     tex = tex.replace("\\include{{{0}}}".format(fn), txt)
+
+for fn in find_inputs(tex):
+    txt = open(fn + ".tex").read()
+    tex = tex.replace("\\input{{{0}}}".format(fn), txt)
 
 for name, loc in map(rename, find_figures(tex)):
     shutil.copyfile(loc, os.path.join(TMPDIR, name))

### 2017-09-18 paper/ms.tex
different page breaks for pdflatex
removing response text
some responding to referee

@@
 
-\newcommand{\response}[1]{{\color{blue}#1}}
+\newcommand{\response}[1]{{#1}}
+\newcommand{\responseb}[1]{{\color{blue}{#1}}}
 
@@
 arises naturally in physical systems and we demonstrate it can be used as
-an effective model in other cases.
+an effective model\footnote{\responseb{Here and throughout, we use the term
+``effective model'' to refer to a model that we use to make inferences even
+though it is known to be wrong.}} in other cases.
 This method is especially appealing compared to other similar methods~--~we
@@
 $\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
-maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
-$\bvec{\alpha}$ using a non-linear optimization routine \citep{Nocedal:2006}.}
-\response{Furthermore, the uncertainties on $\bvec{\theta}$ and
-$\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
+maximizing \eq{gp-likelihood}~--~\responseb{or the likelihood times a prior for
+a Bayesian estimate}~---~with respect to $\bvec{\theta}$ and $\bvec{\alpha}$
+using a non-linear optimization routine \citep{Nocedal:2006}.}
+\responseb{Furthermore, the posterior probability density for $\bvec{\theta}$
+and $\bvec{\alpha}$ can be quantified by multiplying the likelihood by a prior
 $p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov chain Monte Carlo (MCMC)
@@
 \end{eqnarray}
-This model is often called an Ornstein--Uhlenbeck process \response{\citep[in
-reference to the classic paper,][]{Uhlenbeck:1930}} and this is the kernel
-that was studied by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
+This model is often called a \responseb{``damped random walk'' or}
+Ornstein--Uhlenbeck process \response{\citep[in reference to the classic
+paper,][]{Uhlenbeck:1930}} and this is the kernel that was studied by Rybicki
+\& Press \citep{Rybicki:1992,Rybicki:1995}.
 
@@
 \begin{eqnarray}\eqlabel{product-rule}
-k_j(\tau) \, k_k(\tau) =
+k_j(\tau) \, k_i(\tau) =
     e^{-\tilde{c}\,\tau}\,[
@@
 \begin{eqnarray}
-    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_k \pm b_j\,b_k) \\
-    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_k \mp a_j\,b_k) \\
-    \tilde{c} &=& c_j + c_k \\
-    \tilde{d}_{\pm} &=& d_j \mp d_k \quad.
+    \tilde{a}_{\pm} &=& \frac{1}{2}\,(a_j\,a_i \pm b_j\,b_i) \\
+    \tilde{b}_{\pm} &=& \frac{1}{2}\,(b_j\,a_i \mp a_j\,b_i) \\
+    \tilde{c} &=& c_j + c_i \\
+    \tilde{d}_{\pm} &=& d_j \mp d_i \quad.
 \end{eqnarray}
@@
 $d_j = 0$.
-During the preparation of this paper, we generalized this earlier method to
-include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
-performance algorithm by restricting our method to positive definite matrices.
-In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
+\responseb{The \citet{Ambikasaran:2015} method uses a factorization routine
+that can be applied to general semiseparable matrices.
+In other words, it can be applied to factorize matrices even if they are not
 positive definite.
+While preparing this paper, we generalized this method to include non-zero
+$b_j$ and $d_j$, but all valid GP covariance matrices are symmetric and
+positive definite so we can exploit this constraint to derive a Cholesky
+factorization algorithm.
+It turns out that this algorithm is more than an order of magnitude faster
+than the previously published method when applied to \celeriteterm\ models.
+In this and the following sections, we derive this novel factorization
+algorithm and, in \app{psd}, we discuss methods for ensuring that the
+covariance matrix for a \celeriteterm\ model is positive definite.}
 
@@
 
+\newpage
 \subsection{How to choose a celerite model}
@@
 criterion'', such as the Akaike information criterion (AIC) or the Bayesian
-information criterion (BIC), as a model selection metric.
-These criteria are easy to implement and we use the BIC in \sect{example2}
-but, since these are only formally valid under specific sets of assumptions, we
-cannot recommend their use in general.
-}
-
-\newpage
+information criterion (BIC), as a model selection metric.}
+\responseb{These criteria are easy to implement and we use the BIC in
+\sect{example2}, but each information criterion comes with caveats that should
+be considered in any application \citep[see, for example,][]{Ivezic:2014}.}
+
 \section{Examples}\sectlabel{examples}
@@
 optimal method, but we hope these examples encourage interested readers to
-investigate the applicability of \celerite\ to their research.
-
+investigate the applicability of \celerite\ to their research.}
+
+\response{
 In each example, the joint posterior probability density is given by
@@
 The specific parameters and priors are discussed in detail in each section,
-but we generally assume separable uniform priors with plausibly broad support.
+but we generally assume \responseb{independent} uniform priors with plausibly
+broad support.}
+
+\responseb{
+While we do perform a convergence analysis in each case by computing the
+autocorrelation function and integrated autocorrelation time for the target
+chain \citep{Sokal:1989, Goodman:2010}, the usual caveats associated with MCMC
+analysis apply when interpreting these results.
+For example, if there are multiple posterior modes that are not sampled by any
+chain, the autocorrelation analysis will not identify that the sampler has not
+converged.
+There are published sampling algorithms \citep{Gregory:2005, Liu:2008,
+Feroz:2009, Brewer:2011, Kelly:2014} that can, in some cases, be less
+sensitive to multi-modality than the sampler that we use
+\citep{Goodman:2010,Foreman-Mackey:2013}, but a comparison of sampling
+algorithms is beyond the scope of this paper and \celerite\ will also improve
+the computational efficiency of GP models when using the other inference
+algorithms.
 }
@@
 Zhu:1997} implemented by the \project{SciPy} project \citep{Jones:2001}.
-The top left panel of \Figure{simulated-correct} shows the conditional mean
-and standard deviation of the MAP model over-plotted on the simulated data and
-the bottom panel shows the residuals away from this model.
+The top left panel of \Figure{simulated-correct} shows the mean and standard
+deviation of the \responseb{predicted noise-free data distribution from the
+MAP model over-plotted as a blue contour} on the simulated data and the bottom
+panel shows the residuals away from this model.
 Since we are using the correct model to fit the data, it is reassuring that
@@
 \emcee\ \citep{Goodman:2010, Foreman-Mackey:2013}.
-We initialize \exampleinwalkers~walkers by sampling from a three-dimensional
-isotropic Gaussian centered the MAP parameter vector and with a standard
-deviation of $10^{-4}$ in each dimension.
+We initialize \exampleinwalkers~walkers\footnote{\responseb{``Walker'' is the
+name given by \citet{Goodman:2010} and \citet{Foreman-Mackey:2013} to each
+parallel MCMC chain in the ensemble sampling algorithm used in \emcee.}} by
+sampling from a three-dimensional isotropic Gaussian centered the MAP
+parameter vector and with a standard deviation of $10^{-4}$ in each dimension.
 We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
@@
 
-\newpage
 \subsection{Example 4: Asteroseismic oscillations}\sectlabel{astero}
@@
 
+\newpage
 \subsection{Code availability}
@@
 For stationary kernels, this is equivalent~--~by Bochner's theorem \citep[see
-Section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
+section 4.2.1 in][]{Rasmussen:2006}~--~to requiring that the kernel be the
 Fourier transform of a positive finite measure.

### 2017-09-18 paper/ms.tex
more response to referee

@@
 We then run \exampleinburn~steps of burn-in and \exampleinsteps~steps of MCMC.
-To assess convergence, we estimate the mean integrated autocorrelation time of
-the chain across parameters \citep{Sokal:1989, Goodman:2010} and find that the
-chain results in \exampleineff~effective samples.
+To assess convergence, we estimate the integrated autocorrelation time of the
+chain \responseb{averaged across parameters} \citep{Sokal:1989, Goodman:2010}
+and find that the chain results in \exampleineff~effective samples.
 
@@
     figure shows the posterior distribution marginalized over all other
-    nuisance parameters in \eq{rot-kernel}.  This is consistent with the
+    nuisance parameters in \eq{rot-kernel}.
+    \responseb{The 1-$\sigma$ error bar on this measurement is indicated with
+    vertical dashed lines.}
+    This result is consistent with the
     published rotation period made using the full
@@
 
-This model requires about 10~CPU minutes to run the MCMC to convergence.
+\responseb{For this example, about 10~CPU minutes were required to run the
+MCMC.}
 This is more computationally intensive than traditional methods of measuring
@@
     from the inference shown in \Figure{astero} compared to the published
-    value (error bar) based on several years of \kepler\ observations
-    \citep{Pinsonneault:2014}.
+    value (\responseb{blue error bars}) based on several years of \kepler\
+    observations \citep{Pinsonneault:2014}.
     The two-dimensional contours show the 0.5-, 1-, 1.5, and 2-sigma credible
@@
 This example is different from all the previous examples~--~both simulated and
-real~--~because, in this case, do not set the deterministic
-mean function $\mu_\bvec{\theta}$ to zero.
+real~--~because, in this case, do not set the mean function $\mu_\bvec{\theta}$
+to zero.
 Instead, we make inferences about $\mu_\bvec{\theta}$ because it is a
@@
 \begin{eqnarray}\eqlabel{carma-kernel}
-k_\mathrm{CARMA}(\tau) = \sum_{j=1}^J A_j\,\exp\,(r_j\,\tau)
+k_\mathrm{CARMA}(\tau) = \sum_{j=1}^{2\,J} A_j\,\exp\,(r_j\,\tau)
 \end{eqnarray}
@@
 \end{eqnarray}
-and $\sigma$, $\{r_j\}_{j=1}^J$ and $\{\beta_k\}_{k=1}^K$ are parameters of
-the model; $\sigma$ is real, while the others are complex.
+and $\sigma$, $\{r_j\}_{j=1}^{J}$ and $\{\beta_k\}_{k=1}^K$ are parameters
+of the model; $\sigma$ is real, while the others are complex.
 Comparing \eq{carma-kernel} to \eq{celerite-kernel-complex}, we can see that
 every CARMA model corresponds to an equivalent \celeriteterm\ model and the
-parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed analytically.
+parameters $a_j$, $b_j$, $c_j$, and $d_j$ can be easily computed
+analytically.\footnote{\responseb{Each CARMA term with complex parameters
+$A_j$ and $r_j$ corresponds to a \celeriteterm\ term with parameters $a_j =
+2\,\mathrm{Re}(A_j)$, $b_j = 2\,\mathrm{Im}(A_j)$, $c_j = -\mathrm{Re}(r_j)$,
+and $d_j = -\mathrm{Im}(r_j)$.}}
 The inverse statement is not as simple.

### 2017-09-18 paper/response.txt
why blue?

@@
 review. We have taken the feedback to heart and we think that the manuscript
-has been greatly improved as a result.
+has been greatly improved as a result. To simplify a second review, all major
+changes are colored blue.
 
@@
 updated the manuscript to describe and test this new method in Section 5. We
-hope that the referee will consider this updated method and welcome any
+hope that the referee will consider this updated method and we welcome any
 feedback.
@@
 
-For each of the above, we agree and have address each concern.
+For each of the above, we agree and have addressed each concern.
 

### 2017-09-19 paper/ms.tex
formatting
typo

@@
 
+\newpage
 \section{The celerite model}\sectlabel{celerite}
@@
 
+\newpage
 \subsection{Scalable computation of celerite models}\sectlabel{scalable}
@@
 parallel MCMC chain in the ensemble sampling algorithm used in \emcee.}} by
-sampling from a three-dimensional isotropic Gaussian centered the MAP
+sampling from a three-dimensional isotropic Gaussian centered the on MAP
 parameter vector and with a standard deviation of $10^{-4}$ in each dimension.
@@
 
-\newpage
 \subsection{Code availability}

### 2017-11-13 paper/ms.tex
layout
zenodo links

@@
 \newcommand{\response}[1]{{#1}}
-\newcommand{\responseb}[1]{{\color{blue}{#1}}}
+\newcommand{\responseb}[1]{{#1}}
 
@@
 
+\edef \cca {\arabic{affilcounter}}\stepcounter{affilcounter}
+\altaffiltext{\cca}{Center for Computational Astrophysics,
+                   Flatiron Institue, New York, NY}
+
 \edef \guggenheim {\arabic{affilcounter}}\stepcounter{affilcounter}
@@
 
-\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\simons}{Simons Fellow}
+%\edef \simons {\arabic{affilcounter}}\stepcounter{affilcounter}
+%\altaffiltext{\simons}{Simons Fellow}
 
 \edef \columbia {\arabic{affilcounter}}\stepcounter{affilcounter}
-\altaffiltext{\columbia}{Department of Astronomy, Columbia University,
+\altaffiltext{\columbia}{Simons Fellow;
+                         Department of Astronomy, Columbia University,
                          New York, NY}
@@
 \author{%
-    Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw},
+    Daniel~Foreman-Mackey\altaffilmark{\sagan,\uw,\cca},
     Eric~Agol\altaffilmark{\uw,\guggenheim},
     Sivaram~Ambikasaran\altaffilmark{\iis}, and
-    Ruth~Angus\altaffilmark{\simons,\columbia}
+    Ruth~Angus\altaffilmark{\columbia}
 }
@@
 and exoplanet transit fitting (\sect{transit}).
-Some readers might consider starting with these section for motivating
+Some readers might consider starting with this section for motivating
 examples before delving into the detailed derivations of the earlier sections.
@@
 These are available online at \project{GitHub}
-\url{https://github.com/dfm/celerite} and \project{Zenodo}
-\citep{Foreman-Mackey:2017}.
+\url{https://github.com/dfm/celerite} and
+\dataset[Zenodo]{https://doi.org/10.5281/zenodo.806847}.
 
@@
 
-\newpage
 \section{The celerite model}\sectlabel{celerite}
@@
 
-\newpage
 \subsection{Scalable computation of celerite models}\sectlabel{scalable}
@@
 
+\newpage
 \subsection{Example 3: Stellar rotation} \sectlabel{rotation}
@@
 \url{https://github.com/dfm/celerite}\footnote{This version of the paper was
-generated with git commit \texttt{\githash} (\gitdate).} and Zenodo
-\citep{Foreman-Mackey:2017}, and it is made available under the MIT license.
+generated with git commit \texttt{\githash} (\gitdate).} and
+\dataset[Zenodo]{https://doi.org/10.5281/zenodo.806847}, and it is made
+available under the MIT license.
 
